---
title: "Development review: partner-dependence diagnostics"
output:
  html_document:
    toc: true
---

This note reviews the cross-sectional Gaussian partner check using the fixed
simulated data shipped with `dyadMLM`. It is an end-to-end development review,
not a formal hypothesis-test calibration.

```{r setup, include=FALSE}
package_root <- normalizePath(file.path("..", ".."))
devtools::load_all(package_root, quiet = TRUE)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE
)

nsim <- 1000L

run_partner_check <- function(model, data, seed, role = NULL) {
  simulate_dyad_responses(model, nsim = nsim, seed = seed) |>
    check_partner_dependence(
      dyad = data$coupleID,
      role = role,
      plot = FALSE
    )
}
```

# Distinguishable dyads

The female-male composition has distinguishable roles. The intended model
allows separate role-specific spreads and their covariance. Two constrained
models then omit the partner correlation or impose an exchangeable covariance
structure.

```{r prepare-distinguishable}
distinguishable_data <- prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  model_types = "none",
  keep_compositions = "female-male",
  include_arbitrary_member_contrast = TRUE,
  seed = 12000L
)
```

```{r fit-distinguishable-models}
distinguishable_unstructured_model <- glmmTMB::glmmTMB(
  closeness ~
    0 + .is_female + .is_male +
    (0 + .is_female + .is_male | coupleID),
  dispformula = ~0,
  family = gaussian(),
  data = distinguishable_data
)

distinguishable_no_correlation_model <- glmmTMB::glmmTMB(
  closeness ~
    0 + .is_female + .is_male +
    (0 + .is_female + .is_male || coupleID),
  dispformula = ~0,
  family = gaussian(),
  data = distinguishable_data
)

distinguishable_exchangeable_model <- glmmTMB::glmmTMB(
  closeness ~
    0 + .is_female + .is_male +
    (1 | coupleID) +
    (0 + .member_contrast_arbitrary | coupleID),
  dispformula = ~0,
  family = gaussian(),
  data = distinguishable_data
)
```

```{r check-distinguishable-models, include=FALSE}
distinguishable_unstructured_check <- run_partner_check(
  distinguishable_unstructured_model,
  distinguishable_data,
  seed = 12001L,
  role = distinguishable_data$gender
)

distinguishable_no_correlation_check <- run_partner_check(
  distinguishable_no_correlation_model,
  distinguishable_data,
  seed = 12002L,
  role = distinguishable_data$gender
)

distinguishable_exchangeable_simulations <- simulate_dyad_responses(
  distinguishable_exchangeable_model,
  nsim = nsim,
  seed = 12003L
)

distinguishable_exchangeable_pooled_check <- check_partner_dependence(
  distinguishable_exchangeable_simulations,
  dyad = distinguishable_data$coupleID,
  plot = FALSE
)

distinguishable_exchangeable_gender_check <- check_partner_dependence(
  distinguishable_exchangeable_simulations,
  dyad = distinguishable_data$coupleID,
  role = distinguishable_data$gender,
  plot = FALSE
)
```

## Intended covariance model

```{r plot-distinguishable-unstructured, echo=FALSE}
plot(distinguishable_unstructured_check, ask = FALSE)
```

## Partner correlation omitted

```{r plot-distinguishable-no-correlation, echo=FALSE}
plot(distinguishable_no_correlation_check, ask = FALSE)
```

## Exchangeable covariance constraint

The pooled view asks whether the constrained model reproduces the overall
partner covariance. The gender-aware view asks whether pooling hides
role-specific differences. Both views use exactly the same simulations.

### Pooled view

```{r plot-distinguishable-exchangeable-pooled, echo=FALSE}
plot(distinguishable_exchangeable_pooled_check, ask = FALSE)
```

### Gender-aware view

```{r plot-distinguishable-exchangeable-gender, echo=FALSE}
plot(distinguishable_exchangeable_gender_check, ask = FALSE)
```

# Exchangeable dyads

The female-female composition was generated as exchangeable. Its intended
shared/difference model is compared with a model that omits all partner
dependence while retaining the marginal response spread.

```{r prepare-exchangeable}
exchangeable_data <- prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  model_types = "none",
  keep_compositions = "female-female",
  seed = 12000L
)
```

```{r fit-exchangeable-models}
exchangeable_model <- glmmTMB::glmmTMB(
  closeness ~
    1 +
    (1 | coupleID) +
    (0 + .member_contrast_arbitrary | coupleID),
  dispformula = ~0,
  family = gaussian(),
  data = exchangeable_data
)

exchangeable_independent_model <- glmmTMB::glmmTMB(
  closeness ~ 1,
  family = gaussian(),
  data = exchangeable_data
)
```

```{r check-exchangeable-models, include=FALSE}
exchangeable_check <- run_partner_check(
  exchangeable_model,
  exchangeable_data,
  seed = 12004L
)

exchangeable_independent_check <- run_partner_check(
  exchangeable_independent_model,
  exchangeable_data,
  seed = 12005L
)
```

## Intended covariance model

```{r plot-exchangeable, echo=FALSE}
plot(exchangeable_check, ask = FALSE)
```

## Partner dependence omitted

```{r plot-exchangeable-independent, echo=FALSE}
plot(exchangeable_independent_check, ask = FALSE)
```

# Findings

The table identifies which observed summaries fall outside the central 95%
interval of the corresponding replicated summaries. These intervals are
descriptive and are not hypothesis tests.

```{r summarize-findings, echo=FALSE}
partner_checks <- list(
  distinguishable_unstructured = distinguishable_unstructured_check,
  distinguishable_no_correlation = distinguishable_no_correlation_check,
  distinguishable_exchangeable_pooled =
    distinguishable_exchangeable_pooled_check,
  distinguishable_exchangeable_gender =
    distinguishable_exchangeable_gender_check,
  exchangeable_correct = exchangeable_check,
  exchangeable_independent = exchangeable_independent_check
)

findings <- do.call(
  rbind,
  lapply(names(partner_checks), function(check_name) {
    statistics <- partner_checks[[check_name]]$statistics_table
    outside_interval <-
      statistics$observed_value < statistics$replicated_lower |
      statistics$observed_value > statistics$replicated_upper

    data.frame(
      check = check_name,
      summaries_outside_interval = if (any(outside_interval)) {
        paste(statistics$label[outside_interval], collapse = "; ")
      } else {
        "None"
      },
      row.names = NULL
    )
  })
)

knitr::kable(findings)
```

The expected pattern is that the intended models remain compatible with the
observed summaries. Omitting partner correlation should affect the partner
correlation and its mean/difference representation. The gender-aware view of
the exchangeable constraint can reveal role asymmetry that is not apparent in
the pooled view.
