---
title: "Reference validation: exchangeable partner-dependence summaries"
output: html_document
---

This report checks the four exchangeable summaries against an independent
Woody–Sadler calculation and recorded Dingy output. It does not validate response
simulation, model centring, or predictive calibration.

```{r setup, include=FALSE}
package_root <- normalizePath(file.path("..", ".."))
devtools::load_all(package_root, quiet = TRUE)
knitr::opts_chunk$set(message = FALSE, warning = FALSE, collapse = TRUE,
                      comment = "#>")
tolerance <- 1e-12
```

## Reference calculation

For member values \(Y_{i1},Y_{i2}\), define the dyad average
\(M_i=(Y_{i1}+Y_{i2})/2\) and half-difference \(D_i=(Y_{i1}-Y_{i2})/2\).
The interchangeable-member decomposition uses
\(MS_B=2\operatorname{var}(M)\) and \(MS_W=2\operatorname{mean}(D^2)\).
The first divides by \(n-1\); the second divides by \(n\) and measures
half-differences about zero, making arbitrary within-dyad swaps irrelevant.

The function below was written independently from Woody and Sadler (2005),
[Structural Equation Models for Interchangeable Dyads](https://doi.org/10.1037/1082-989X.10.2.139).
No code or data were copied from the
[APA supplement](https://supp.apa.org/psycarticles/supplemental/1082-989X.10.2.139/1082-989X.10.2.139_supp.html).

```{r reference}
woody_sadler_reference <- function(member_1, member_2) {
  n <- length(member_1)
  stopifnot(n == length(member_2), n >= 2L)
  MS_between <- 2 * var((member_1 + member_2) / 2)
  MS_within <- sum((member_1 - member_2)^2) / (2 * n)
  member_variance <- (MS_between + MS_within) / 2
  partner_covariance <- (MS_between - MS_within) / 2
  c(
    MS_between = MS_between,
    MS_within = MS_within,
    exchangeable_member_sd = sqrt(member_variance),
    exchangeable_partner_correlation = partner_covariance / member_variance,
    dyad_mean_sd = sqrt(MS_between / 2),
    half_difference_rms = sqrt(MS_within / 2)
  )
}
```

## Numerical agreement and member swaps

The same seeded 48-dyad dataset underlies the calculation and the recorded
Dingy comparison. All assertions below use an absolute tolerance of `1e-12`.

```{r numerical-checks}
set.seed(20260826L)
n_dyads <- 48L
generated_means <- rnorm(n_dyads, mean = 2, sd = 1.1)
generated_half_differences <- rnorm(n_dyads, mean = 0, sd = 0.55)
validation_data <- data.frame(
  dyad = seq_len(n_dyads),
  member_1 = generated_means + generated_half_differences,
  member_2 = generated_means - generated_half_differences
)
stopifnot(nrow(validation_data) >= 40L, all(complete.cases(validation_data)))
actual <- dyadMLM:::calculate_partner_pair_statistics(
  validation_data$member_1, validation_data$member_2, role_specific = FALSE
)
reference <- woody_sadler_reference(
  validation_data$member_1, validation_data$member_2
)
M <- (validation_data$member_1 + validation_data$member_2) / 2
D <- (validation_data$member_1 - validation_data$member_2) / 2
member_variance <- actual[["exchangeable_member_sd"]]^2
partner_covariance <- actual[["exchangeable_partner_correlation"]] * member_variance
stopifnot(
  max(abs(actual - reference[names(actual)])) <= tolerance,
  abs(reference[["MS_between"]] - 2 * var(M)) <= tolerance,
  abs(reference[["MS_within"]] - 2 * mean(D^2)) <= tolerance,
  abs(member_variance - (var(M) + mean(D^2))) <= tolerance,
  abs(partner_covariance - (var(M) - mean(D^2))) <= tolerance
)

set.seed(20260827L)
rows_to_swap <- sort(sample(seq_len(n_dyads), size = 16L))
swapped <- validation_data
swapped[rows_to_swap, c("member_1", "member_2")] <-
  swapped[rows_to_swap, c("member_2", "member_1")]
swapped_D <- (swapped$member_1 - swapped$member_2) / 2
swapped_actual <- dyadMLM:::calculate_partner_pair_statistics(
  swapped$member_1, swapped$member_2, role_specific = FALSE
)
swapped_reference <- woody_sadler_reference(swapped$member_1, swapped$member_2)
stopifnot(
  max(abs(swapped_actual - actual)) <= tolerance,
  max(abs(swapped_reference - reference)) <= tolerance,
  max(abs(swapped_D[rows_to_swap] + D[rows_to_swap])) <= tolerance,
  max(abs(swapped_D^2 - D^2)) <= tolerance
)
cat("All reference, covariance-identity, and member-swap assertions passed.\n")
```

## Dingy cross-check

The live [Dingy](https://davidakenny.net/DyadR/DyadRweb.htm) run on 2026-08-26
reported lavaan 0.6-19. Its rounded estimates were variance **1.36**, covariance
**0.67**, and correlation **0.50**. The recorded nonindependence test,
\(\chi^2(1)=13.566, p<.001\), is context only, not one of these four summaries.

Dingy's ML convention divides the centred dyad-mean sum of squares by \(n\)
instead of \(n-1\). Reconstructing that convention must reproduce its displayed
estimates. This is a historical software cross-check, not a fresh Dingy run.

```{r dingy-check}
dingy_displayed <- c(
  common_member_variance = 1.36,
  cross_member_covariance = 0.67,
  partner_correlation = 0.50
)
ml_variance_of_M <- mean((M - mean(M))^2)
dingy_reconstruction <- c(
  common_member_variance = ml_variance_of_M + mean(D^2),
  cross_member_covariance = ml_variance_of_M - mean(D^2)
)
dingy_reconstruction <- c(
  dingy_reconstruction,
  partner_correlation = dingy_reconstruction[["cross_member_covariance"]] /
    dingy_reconstruction[["common_member_variance"]]
)
stopifnot(max(abs(round(dingy_reconstruction, 2) - dingy_displayed)) <= tolerance)
dingy_summaries <- c(
  sqrt(dingy_reconstruction[["common_member_variance"]]),
  dingy_reconstruction[["partner_correlation"]],
  sqrt(ml_variance_of_M),
  sqrt(mean(D^2))
)
knitr::kable(data.frame(
  statistic = names(actual),
  dyadMLM = unname(actual),
  Woody_Sadler = unname(reference[names(actual)]),
  absolute_difference = unname(abs(actual - reference[names(actual)])),
  Dingy_ML_reconstruction = dingy_summaries
), digits = 9)
```

The four package summaries match the independent reference. Arbitrary member
swaps preserve them. Dingy's small differences follow from ML scaling; both
conventions use the same half-difference mean square about zero.

To repeat the external check, upload the temporary CSV below, select `member_1`
and `member_2` as one mixed variable, treat positions as indistinguishable, and
request nonindependence analysis. Retain the tables and computer/lavaan output.

```{r dingy-upload}
dingy_csv <- file.path(tempdir(), "dyadMLM-partner-dependence-validation.csv")
write.csv(validation_data, dingy_csv, row.names = FALSE)
cat("Temporary Dingy upload file:", dingy_csv, "\n")
```
