---
title: "Outer simulation study: Gaussian partner-dependence checks"
output:
  html_document:
    toc: true
    code_folding: hide
params:
  outer_repetitions: 500
  inner_simulations: 1000
---

This study repeats **generate data → fit model → predictive check**. It compares
correct models with deliberately wrong covariance restrictions. An independent
known-truth control separates the ranking step from model fitting. These are
descriptive checks; observed positions are not p-values.

```{r setup, include=FALSE}
devtools::load_all(normalizePath(file.path("..", "..")), quiet = TRUE)
library(dplyr)
knitr::opts_chunk$set(fig.width = 8, fig.height = 5, message = FALSE, warning = FALSE)
n_outer <- as.integer(params$outer_repetitions)
n_inner <- as.integer(params$inner_simulations)
stopifnot(n_outer >= 1L, n_inner >= 20L)
sample_sizes <- c(50L, 200L)
base_seed <- 20260826L
is_pilot <- n_outer < 500L || n_inner < 1000L
```

# Design

Each population has a fixed predictor design; new responses are generated for
every repetition. Correct and restricted models use the same datasets. Defaults
are 500 repetitions × 1,000 reference simulations × 50/200 dyads: 7,000 fits.
At 500 repetitions, a rate has Monte Carlo SE of about 1 percentage point near
5%, and at most 2.2 points near 50%. The large restrictions below assess whether
the intended problems are detectable, not sensitivity to small departures.

```{r scenarios}
truths <- tibble::tribble(
  ~truth, ~sd_1, ~sd_2, ~rho, ~mean_1, ~mean_2, ~role_aware,
  "positive",            1,   1,  0.5,    0,   0, FALSE,
  "zero",                1,   1,    0,    0,   0, FALSE,
  "negative",            1,   1, -0.5,    0,   0, FALSE,
  "distinguishable",    0.8, 1.4,  0.4, -0.3, 0.3, TRUE
)
scenarios <- tibble::tribble(
  ~scenario, ~truth, ~fitted_model, ~model_correct,
  "positive_correct", "positive", "exchangeable", TRUE,
  "positive_independent", "positive", "independent", FALSE,
  "zero_correct", "zero", "independent", TRUE,
  "negative_correct", "negative", "exchangeable", TRUE,
  "negative_independent", "negative", "independent", FALSE,
  "distinguishable_correct", "distinguishable", "unstructured", TRUE,
  "distinguishable_exchangeable", "distinguishable", "exchangeable_roles", FALSE
)
knitr::kable(left_join(scenarios, truths, by = "truth"))
```

# Generate, fit, check

Data come directly from a bivariate Gaussian distribution, independently of
`glmmTMB`. The predictor combines shared and member-specific variation; its
population slope is 0.6. Shared and member-sign random effects allow either sign
of correlation with equal role variances. The unstructured model allows unequal
role variances. Role means are estimated separately for distinguishable dyads.

```{r functions}
make_design <- function(n_dyads, seed) {
  set.seed(seed)
  shared_x <- rep(rnorm(n_dyads, sd = 0.6), each = 2L)
  member_x <- rnorm(2L * n_dyads, sd = 0.8)
  data.frame(
    dyad = factor(rep(seq_len(n_dyads), each = 2L)),
    role = factor(rep(c("member_1", "member_2"), n_dyads),
                  levels = c("member_1", "member_2")),
    member_sign = rep(c(-1, 1), n_dyads), x = shared_x + member_x
  )
}

draw_data <- function(design, truth, seed) {
  set.seed(seed)
  n <- nlevels(design$dyad)
  z_1 <- rnorm(n)
  z_2 <- rnorm(n)
  error_1 <- truth$sd_1 * z_1
  error_2 <- truth$sd_2 * (truth$rho * z_1 + sqrt(1 - truth$rho^2) * z_2)
  design$expected_mean <- rep(c(truth$mean_1, truth$mean_2), n) + 0.6 * design$x
  design$y <- design$expected_mean + as.vector(rbind(error_1, error_2))
  design
}

fit_model <- function(data, model_name) {
  formula <- switch(model_name,
    exchangeable = y ~ 1 + x + (1 | dyad) + (0 + member_sign | dyad),
    independent = y ~ 1 + x,
    unstructured = y ~ 0 + role + x + (0 + role | dyad),
    exchangeable_roles = y ~ 0 + role + x + (1 | dyad) + (0 + member_sign | dyad),
    stop("Unknown model: ", model_name)
  )
  glmmTMB::glmmTMB(formula, data = data, family = gaussian(),
                  dispformula = if (model_name == "independent") ~1 else ~0)
}

extract_results <- function(check, scenario, truth, n_dyads, outer_rep, method) {
  check$statistics_table |>
    filter(parameterization == "member") |>
    transmute(
      scenario = scenario, truth = truth, n_dyads = n_dyads,
      outer_rep = outer_rep, method = method, statistic_name,
      observed_value, replicated_median, replicated_lower, replicated_upper,
      observed_quantile,
      outside_95 = observed_value < replicated_lower | observed_value > replicated_upper,
      direction = case_when(observed_value < replicated_lower ~ "below",
                            observed_value > replicated_upper ~ "above", TRUE ~ "inside")
    )
}

run_oracle <- function(data, truth, n_dyads, outer_rep, nsim, seed) {
  set.seed(seed)
  z_1 <- matrix(rnorm(n_dyads * nsim), nrow = n_dyads)
  z_2 <- matrix(rnorm(n_dyads * nsim), nrow = n_dyads)
  error_1 <- truth$sd_1 * z_1
  error_2 <- truth$sd_2 * (truth$rho * z_1 + sqrt(1 - truth$rho^2) * z_2)
  simulated <- matrix(NA_real_, nrow = nsim, ncol = 2L * n_dyads)
  simulated[, seq.int(1L, 2L * n_dyads, 2L)] <-
    t(error_1 + data$expected_mean[data$role == "member_1"])
  simulated[, seq.int(2L, 2L * n_dyads, 2L)] <-
    t(error_2 + data$expected_mean[data$role == "member_2"])
  simulations <- structure(list(
    observed_response = data$y, simulated_responses = simulated,
    predicted_response = data$expected_mean, model_frame = data.frame(y = data$y)
  ), class = c("dyadMLM_response_simulations", "list"), dyadMLM = list(
    backend = "glmmTMB", family = "gaussian", link = "identity",
    reference = "known-parameter oracle", random_effects = "known covariance",
    parameter_uncertainty = "not applicable", seed = seed
  ))
  check <- check_partner_dependence(simulations, dyad = data$dyad, plot = FALSE)
  extract_results(check, "positive_oracle", truth$truth, n_dyads, outer_rep, "oracle")
}

run_fitted_check <- function(data, scenario, role_aware, n_dyads, outer_rep, nsim, seed) {
  warnings <- character()
  scenario_name <- scenario$scenario
  truth_name <- scenario$truth
  status <- tibble::tibble(
    scenario = scenario_name, truth = truth_name, n_dyads = n_dyads,
    outer_rep = outer_rep, status = "fit_error", convergence = NA_integer_,
    pd_hess = NA, warnings = "", error = NA_character_
  )
  statistics <- tryCatch(withCallingHandlers({
    fit <- fit_model(data, scenario$fitted_model)
    status$convergence <- fit$fit$convergence
    status$pd_hess <- fit$sdr$pdHess
    if (!isTRUE(status$convergence == 0L) || !isTRUE(status$pd_hess)) {
      status$status <- "fit_problem"
      NULL
    } else {
      status$status <- "check_error"
      simulations <- simulate_dyad_responses(fit, nsim = nsim, seed = seed)
      check <- check_partner_dependence(
        simulations, dyad = data$dyad,
        role = if (role_aware) data$role else NULL, plot = FALSE
      )
      stopifnot(check$n_pairs == n_dyads)
      if (role_aware) stopifnot(identical(check$role_order, c("member_1", "member_2")))
      result <- extract_results(check, scenario$scenario, scenario$truth,
                                n_dyads, outer_rep, "fitted plug-in")
      status$status <- "success"
      result
    }
  }, warning = function(w) {
    warnings <<- c(warnings, conditionMessage(w))
    invokeRestart("muffleWarning")
  }), error = function(e) {
    status$error <<- conditionMessage(e)
    NULL
  })
  status$warnings <- paste(unique(warnings), collapse = " | ")
  list(status = status, statistics = statistics)
}
```

The oracle uses true means and covariance for the positive-correlation population.
Observed and reference datasets are then exchangeable; positions should be roughly
uniform. Fitted checks instead use estimated parameters. Fit/check errors,
convergence codes, Hessian problems, and warnings are recorded without retries.

```{r run-study}
designs <- lapply(seq_along(sample_sizes), function(i) {
  make_design(sample_sizes[[i]], seed = base_seed + 1000L * i)
})
statuses <- statistics <- list()
for (size_i in seq_along(sample_sizes)) {
  n_dyads <- sample_sizes[[size_i]]
  for (truth_i in seq_len(nrow(truths))) {
    truth <- truths[truth_i, ]
    truth_scenarios <- scenarios[scenarios$truth == truth$truth, ]
    for (outer_rep in seq_len(n_outer)) {
      data_seed <- base_seed + 100000L * truth_i + 10000L * size_i + outer_rep
      data <- draw_data(designs[[size_i]], truth, seed = data_seed)
      if (truth$truth == "positive") {
        statistics[[length(statistics) + 1L]] <- run_oracle(
          data, truth, n_dyads, outer_rep, n_inner, seed = data_seed + 20000000L
        )
      }
      for (scenario_i in seq_len(nrow(truth_scenarios))) {
        scenario <- truth_scenarios[scenario_i, ]
        global_i <- match(scenario$scenario, scenarios$scenario)
        result <- run_fitted_check(data, scenario, truth$role_aware,
          n_dyads, outer_rep, n_inner, seed = data_seed + 10000000L + 1000L * global_i)
        statuses[[length(statuses) + 1L]] <- result$status
        if (!is.null(result$statistics)) statistics[[length(statistics) + 1L]] <- result$statistics
      }
    }
  }
}
run_status <- bind_rows(statuses)
outer_results <- bind_rows(statistics)
```

# Results

Rates use only successful fits/checks and the member parameterization. The
mean/difference parameterization carries equivalent covariance information and
is not counted as additional evidence. MCSE is the plug-in estimate; zero MCSE
at an observed rate of 0 or 1 does not establish an exact population rate.

```{r completion, echo=FALSE}
completion <- run_status |>
  group_by(scenario, n_dyads) |>
  summarise(attempted = n(), success_rate = mean(status == "success"),
    warning_runs = sum(nzchar(warnings)), fit_problems = sum(status == "fit_problem"),
    fit_errors = sum(status == "fit_error"), check_errors = sum(status == "check_error"),
    .groups = "drop")
knitr::kable(completion, digits = 3)
```

## Known truth and correct fits

The oracle should put about 2.5% in each tail, with position SD near
`r sprintf("%.3f", 1 / sqrt(12))`. Flexible fitted models can cluster near the
middle because they estimated the same features being checked. Independence
fits estimate variation but fix correlation, leaving correlation informative.

```{r rates, echo=FALSE}
rates <- outer_results |>
  group_by(method, scenario, n_dyads, statistic_name) |>
  summarise(repetitions = n(), outside_count = sum(outside_95),
    below = mean(direction == "below"), outside = mean(outside_95),
    above = mean(direction == "above"), mean_position = mean(observed_quantile),
    sd_position = sd(observed_quantile), .groups = "drop") |>
  mutate(mcse = sqrt(outside * (1 - outside) / repetitions))
oracle_rates <- filter(rates, method == "oracle")
correct_rates <- filter(rates, method == "fitted plug-in",
                         scenario %in% scenarios$scenario[scenarios$model_correct])
knitr::kable(oracle_rates, digits = 3)
knitr::kable(correct_rates, digits = 3)
```

```{r position-plot, echo=FALSE}
plot_data <- filter(outer_results, n_dyads == max(sample_sizes),
                    statistic_name == "exchangeable_partner_correlation")
old_par <- par(mfrow = c(1, 2))
for (shown in c("positive_oracle", "positive_correct")) {
  plot(ecdf(plot_data$observed_quantile[plot_data$scenario == shown]),
       main = if (shown == "positive_oracle") "Oracle" else "Correct fitted model",
       xlab = "Observed position", ylab = "Cumulative proportion",
       xlim = c(0, 1), ylim = c(0, 1))
  abline(a = 0, b = 1, lty = 2, col = "grey40")
}
par(old_par)
```

## Wrong restrictions: paired comparisons

Directions are specified in advance: omitting positive/negative correlation
should put it above/below the reference; imposing equal role spreads should put
the smaller/larger SD below/above it. Each paired gain compares both fits on the
same dataset, conditional on both succeeding.

```{r paired-comparisons, echo=FALSE}
targets <- tibble::tribble(
  ~constrained, ~correct, ~statistic_name, ~expected,
  "positive_independent", "positive_correct", "exchangeable_partner_correlation", "above",
  "negative_independent", "negative_correct", "exchangeable_partner_correlation", "below",
  "distinguishable_exchangeable", "distinguishable_correct", "role_1_sd", "below",
  "distinguishable_exchangeable", "distinguishable_correct", "role_2_sd", "above"
)
fitted_results <- outer_results |>
  filter(method == "fitted plug-in") |>
  select(scenario, truth, n_dyads, outer_rep, statistic_name, direction)
paired <- targets |>
  inner_join(fitted_results, by = c("constrained" = "scenario", "statistic_name")) |>
  inner_join(fitted_results,
    by = c("correct" = "scenario", "truth", "n_dyads", "outer_rep", "statistic_name"),
    suffix = c("_constrained", "_correct")) |>
  mutate(constrained_flag = direction_constrained == expected,
    correct_flag = direction_correct == expected,
    gain = as.integer(constrained_flag) - as.integer(correct_flag))
paired_summary <- paired |>
  group_by(constrained, n_dyads, statistic_name, expected) |>
  summarise(paired_runs = n(), correct_rate = mean(correct_flag),
    constrained_rate = mean(constrained_flag), paired_gain = mean(gain),
    gain_mcse = sd(gain) / sqrt(n()), .groups = "drop")
knitr::kable(paired_summary, digits = 3)
```

# Interpretation and reproducibility

```{r pilot-note, echo=FALSE, results='asis'}
if (is_pilot) cat("**Pilot run:** these rates are too imprecise for validation conclusions.\n")
```

Read the oracle ranks first, then fitted-model concentration and paired detection
rates. Agreement for flexible fits is expected and alone provides little evidence
of fit. This study concerns large covariance restrictions in cross-sectional
Gaussian models; it does not establish sensitivity to smaller problems or
performance for non-Gaussian or longitudinal models.

The default run is the full study. For a quick execution check, override its
parameters (this does not replace the full study):

```r
rmarkdown::render(
  "dev/diagnostic_checks/partner-dependence-outer-simulation-study.Rmd",
  params = list(outer_repetitions = 20, inner_simulations = 99)
)
```

```{r session-info}
sessionInfo()
```
