---
title: "Worked example: checking partner dependence"
output:
  html_document:
    toc: true
---

Does a model reproduce how much partners' responses vary and how strongly they
are related? This example compares flexible and restricted Gaussian models
using the simulated cross-sectional data shipped with `dyadMLM`. See the
[quick start](README.md#quick-start) to run a check on your own fitted model.

Each histogram shows a summary across datasets created by
`simulate_dyad_responses()`, with the observed value marked by
`check_partner_dependence()`. The checks below remove the fitted mean pattern
from observed and simulated responses in the same way. Values near an edge of
the simulated range suggest a feature to investigate. The middle 95% intervals
and observed positions are descriptive, not formal tests or p-values.

```{r setup, include=FALSE}
package_root <- normalizePath(file.path("..", ".."))
devtools::load_all(package_root, quiet = TRUE)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE
)

nsim <- 1000L

run_partner_check <- function(model, data, seed, role = NULL) {
  simulate_dyad_responses(model, nsim = nsim, seed = seed) |>
    check_partner_dependence(
      dyad = data$coupleID,
      role = role,
      plot = FALSE
    )
}
```

# Distinguishable dyads

The female-male dyads have meaningful member roles. The flexible model allows
each role's responses to vary by a different amount and estimates the partner
correlation. Two restricted models either omit that correlation or require
equal variation for both roles.

```{r prepare-distinguishable}
distinguishable_data <- prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  model_types = "none",
  keep_compositions = "female-male",
  include_arbitrary_member_contrast = TRUE,
  seed = 12000L
)
```

```{r fit-distinguishable-models}
distinguishable_unstructured_model <- glmmTMB::glmmTMB(
  closeness ~
    0 + .is_female + .is_male +
    (0 + .is_female + .is_male | coupleID),
  dispformula = ~0,
  family = gaussian(),
  data = distinguishable_data
)

distinguishable_no_correlation_model <- glmmTMB::glmmTMB(
  closeness ~
    0 + .is_female + .is_male +
    (0 + .is_female + .is_male || coupleID),
  dispformula = ~0,
  family = gaussian(),
  data = distinguishable_data
)

distinguishable_exchangeable_model <- glmmTMB::glmmTMB(
  closeness ~
    0 + .is_female + .is_male +
    (1 | coupleID) +
    (0 + .member_contrast_arbitrary | coupleID),
  dispformula = ~0,
  family = gaussian(),
  data = distinguishable_data
)
```

```{r check-distinguishable-models, include=FALSE}
distinguishable_unstructured_check <- run_partner_check(
  distinguishable_unstructured_model,
  distinguishable_data,
  seed = 12001L,
  role = distinguishable_data$gender
)

distinguishable_no_correlation_check <- run_partner_check(
  distinguishable_no_correlation_model,
  distinguishable_data,
  seed = 12002L,
  role = distinguishable_data$gender
)

distinguishable_exchangeable_simulations <- simulate_dyad_responses(
  distinguishable_exchangeable_model,
  nsim = nsim,
  seed = 12003L
)

distinguishable_exchangeable_pooled_check <- check_partner_dependence(
  distinguishable_exchangeable_simulations,
  dyad = distinguishable_data$coupleID,
  plot = FALSE
)

distinguishable_exchangeable_gender_check <- check_partner_dependence(
  distinguishable_exchangeable_simulations,
  dyad = distinguishable_data$coupleID,
  role = distinguishable_data$gender,
  plot = FALSE
)
```

## Flexible partner variation and correlation

```{r plot-distinguishable-unstructured, echo=FALSE}
plot(distinguishable_unstructured_check, ask = FALSE)
```

## Partner correlation omitted

```{r plot-distinguishable-no-correlation, echo=FALSE}
plot(distinguishable_no_correlation_check, ask = FALSE)
```

## Equal variation for both roles

The pooled view treats members as interchangeable and reports their common
variation and partner correlation. The gender-aware view asks whether pooling
hides role differences. Both views use exactly the same simulations.

### Pooled view

```{r plot-distinguishable-exchangeable-pooled, echo=FALSE}
plot(distinguishable_exchangeable_pooled_check, ask = FALSE)
```

### Gender-aware view

```{r plot-distinguishable-exchangeable-gender, echo=FALSE}
plot(distinguishable_exchangeable_gender_check, ask = FALSE)
```

# Exchangeable dyads

The female-female dyads were generated with interchangeable members. Their
model allows partner correlation and a common amount of response variation.
The comparison model retains response variation but assumes no partner
correlation.

```{r prepare-exchangeable}
exchangeable_data <- prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  model_types = "none",
  keep_compositions = "female-female",
  seed = 12000L
)
```

```{r fit-exchangeable-models}
exchangeable_model <- glmmTMB::glmmTMB(
  closeness ~
    1 +
    (1 | coupleID) +
    (0 + .member_contrast_arbitrary | coupleID),
  dispformula = ~0,
  family = gaussian(),
  data = exchangeable_data
)

exchangeable_independent_model <- glmmTMB::glmmTMB(
  closeness ~ 1,
  family = gaussian(),
  data = exchangeable_data
)
```

```{r check-exchangeable-models, include=FALSE}
exchangeable_check <- run_partner_check(
  exchangeable_model,
  exchangeable_data,
  seed = 12004L
)

exchangeable_independent_check <- run_partner_check(
  exchangeable_independent_model,
  exchangeable_data,
  seed = 12005L
)
```

## Flexible partner variation and correlation

```{r plot-exchangeable, echo=FALSE}
plot(exchangeable_check, ask = FALSE)
```

## Partner dependence omitted

```{r plot-exchangeable-independent, echo=FALSE}
plot(exchangeable_independent_check, ask = FALSE)
```

# Findings

The table lists observed summaries outside the middle 95% of simulated values.
These are prompts for investigation, not failed hypothesis tests.

```{r summarize-findings, echo=FALSE}
partner_checks <- list(
  distinguishable_unstructured = distinguishable_unstructured_check,
  distinguishable_no_correlation = distinguishable_no_correlation_check,
  distinguishable_exchangeable_pooled =
    distinguishable_exchangeable_pooled_check,
  distinguishable_exchangeable_gender =
    distinguishable_exchangeable_gender_check,
  exchangeable_correct = exchangeable_check,
  exchangeable_independent = exchangeable_independent_check
)

findings <- do.call(
  rbind,
  lapply(names(partner_checks), function(check_name) {
    statistics <- partner_checks[[check_name]]$statistics_table
    outside_interval <-
      statistics$observed_value < statistics$replicated_lower |
      statistics$observed_value > statistics$replicated_upper

    data.frame(
      check = check_name,
      summaries_outside_interval = if (any(outside_interval)) {
        paste(statistics$label[outside_interval], collapse = "; ")
      } else {
        "None"
      },
      row.names = NULL
    )
  })
)

knitr::kable(findings)
```

The flexible model is expected to reproduce features that it learned from the
same data. A close match there therefore adds little new evidence about model
fit. The deliberately simpler models provide the more informative comparisons:
leaving out partner correlation should make the correlation summaries stand
out, while forcing both roles to vary equally can reveal a role difference that
the pooled view hides.

# Technical and development notes

Each check uses 1,000 simulated datasets, keeping the fitted parameters fixed
while drawing new random effects and observation errors. The model is not
refitted, and parameter uncertainty is excluded. This document is a worked
development review, not a calibration study. See the
[technical guide](README.md#technical-details) for implementation details and
links to the separate validation studies.
