---
title: "Partner-dependence diagnostics (exploratory vignette draft)"
output: html_document
---

This exploratory note keeps the fuller model sequence as a possible future
vignette draft. See the shorter
[development review](partner-dependence-review.Rmd) for the current feature.

# Distinguishable dyads

## Prep data

```{r}
library(dyadMLM)

diagnostic_data <- prepare_dyad_data(
    dyads_cross,
    dyad = coupleID,
    member = personID,
    role = gender,
    model_types = "none",
    keep_compositions = "female-male",
    seed = 13000L
)
```


## Correct distinguishable model

```{r}
m_c <- glmmTMB::glmmTMB(
    closeness ~
        0 + .is_female + .is_male + (0 + .is_female + .is_male | coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
simulate_dyad_responses(m_c, nsim = 5000, seed = 13001L) |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID, 
    role = diagnostic_data$gender)
```


## Constrained models

```{r}
m_w1 <- glmmTMB::glmmTMB(
    closeness ~
        0 + .is_female + .is_male + (1 | coupleID),
    dispformula = ~ 1,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
simulate_dyad_responses(m_w1, nsim = 5000, seed = 13002L) |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID, 
    role = diagnostic_data$gender)
```



```{r}
m_w2 <- glmmTMB::glmmTMB(
    closeness ~
        0 + .is_female + .is_male + (1 | coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
simulate_dyad_responses(m_w2, nsim = 5000, seed = 13003L) |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID, 
    role = diagnostic_data$gender)
```


```{r}
m_w3 <- glmmTMB::glmmTMB(
    closeness ~
        0 + .is_female + .is_male + (0 + .is_female + .is_male || coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
simulate_dyad_responses(m_w3, nsim = 5000, seed = 13004L) |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID, 
    role = diagnostic_data$gender)
```

```{r}
compare_nested_models(
  m_w3,
  m_c
)
```

# Exchangeability sensitivity

Here, the female-male data are deliberately treated as exchangeable. The
pooled check assesses the imposed covariance structure; the gender-aware check
asks whether that pooling hides role-specific differences.

## Prep data

```{r}
diagnostic_data <- prepare_dyad_data(
    dyads_cross,
    dyad = coupleID,
    member = personID,
    role = gender,
    model_types = "none",
    keep_compositions = "female-male",
    set_exchangeable_compositions = "female-male",
    seed = 13000L
)
```


## Exchangeability-constrained model

```{r}
e_c <- glmmTMB::glmmTMB(
    closeness ~
        1 + 
      
      (1 | coupleID) + 
      (0 + .member_contrast_arbitrary | coupleID),
    
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
e_c_simulations <- simulate_dyad_responses(
  e_c,
  nsim = 5000,
  seed = 13005L
)

e_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID) 
```

## Inspect the same model by gender

```{r}
e_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender) 
```


## Additional constrained models

```{r}
e_w1 <- glmmTMB::glmmTMB(
    closeness ~
        1,
    dispformula = ~ 1,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
simulate_dyad_responses(e_w1, nsim = 5000, seed = 13006L) |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID) 
```



```{r}
e_w2 <- glmmTMB::glmmTMB(
    closeness ~
        1 + (1 | coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
simulate_dyad_responses(e_w2, nsim = 5000, seed = 13007L) |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID) 
```


```{r}
e_w3 <- glmmTMB::glmmTMB(
    closeness ~
        1 + (1 | coupleID/personID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
simulate_dyad_responses(e_w3, nsim = 5000, seed = 13008L) |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID)
```

```{r}
compare_nested_models(
  e_w2,
  e_c
)
```
