---
title: "Checking partner dependence"
output: html_document
---

Does a fitted model reproduce how much partners' responses vary and how
strongly they are related? `dyadMLM` compares these features in the observed
data with the same features in datasets simulated from the model.

```{r setup, include=FALSE}
devtools::load_all(file.path("..", ".."), quiet = TRUE)
knitr::opts_chunk$set(
  message = FALSE, fig.width = 10, fig.height = 6.5, out.width = "100%"
)
```

## Fit a model

Prepare actor and partner predictors for the female-male dyads. Fit an APIM
that allows different effects and response variances for women and men,
but assumes no remaining correlation between partners.

```{r fit}
data <- prepare_dyad_data(
  dyads_cross, dyad = coupleID, member = personID, role = gender,
  predictors = provided_support, model_types = "apim",
  keep_compositions = "female-male"
)
model <- glmmTMB::glmmTMB(
  closeness ~ gender * (.provided_support_actor + .provided_support_partner) +
    diag(0 + gender | coupleID),
  dispformula = ~0, family = gaussian(), data = data
)
```

## Simulate, check, and plot

Simulate new responses and compare their standard deviations (SDs) and
correlations with the observed values. By default, the same model predictions
are subtracted from every dataset. This focuses the check on variation and
correlation remaining after the fixed effects.

```{r check}
simulations <- simulate_dyad_responses(model, seed = 123)
check <- check_partner_dependence(
  simulations, dyad = coupleID, role = gender, panels = TRUE
)
```

The red line is the observed value; dashed lines mark the middle 95% of
simulated values. Here, the role SDs are within their ranges, but the partner correlation
exceeds its upper limit. These ranges and observed positions are descriptive, not
confidence intervals or p-values. Use
`plot(check, panels = TRUE)` to show the panel again without repeating simulations.

The check is most useful for examining restrictions such as the omitted
partner correlation above. A flexible covariance model will usually reproduce
features it estimated from the same data; close agreement alone does not
establish good fit.

## Allow partner correlation

Replace the diagonal block with an unstructured block, keeping the fixed
effects unchanged. The same checking steps apply.

```{r flexible-model}
flexible <- update(model, formula =
  closeness ~ gender * (.provided_support_actor + .provided_support_partner) +
    us(0 + gender | coupleID)
)
flexible_simulations <- simulate_dyad_responses(flexible, seed = 123)
flexible_check <- check_partner_dependence(
  flexible_simulations, dyad = coupleID, role = gender, panels = TRUE
)
```

The observed partner correlation now falls within the simulated range: allowing
partner correlation resolves the mismatch seen above.

## How much dependence can the check reveal?

These examples compare checks that subtract the fitted fixed-effect predictions
with checks that use raw outcomes. Both use the same generated datasets, fitted
models, and reference simulations. The fitted mean models include actor and partner
effects but omit the remaining partner dependence.

The curves target residual correlations of 0.10, 0.30, and 0.50 after accounting for
the true mean model. Each figure reports the number of generated datasets per
condition and reference simulations per fit. Detection requires the observed
correlation to exceed the upper limit of the middle 95% simulated range, matching
the positive dependence omitted from the model. Shading and bars show how precisely
these percentages are estimated from the generated datasets.

```{r selected-family-plots, echo=FALSE, results='asis'}
selected_families <- c(
  "Gaussian" = "gaussian", "Negative binomial (NB2)" = "nbinom2",
  "Ordinal" = "ordinal", "Lognormal" = "lognormal",
  "Zero-inflated Poisson" = "zi_poisson", "Hurdle negative binomial (NB2)" = "hurdle_nbinom2"
)
figure_files <- file.path("figures", paste0("family-comparison-", selected_families, ".png"))
available_figures <- file.exists(figure_files)
if (!all(available_figures)) {
  cat("*Study in progress: ", sum(available_figures), " of ", length(selected_families),
      " selected family results are available.*\n\n", sep = "")
}
for (figure_index in which(available_figures)) {
  cat("\n\n![", names(selected_families)[figure_index],
      ": detection of omitted partner correlation by number of dyads](",
      figure_files[figure_index], ")\n\n", sep = "")
}
```

In the Gaussian example, the default check detected a residual correlation of
0.30 in about 48% of datasets with 40 dyads and 86% with 100 dyads. A mismatch
can therefore remain unnoticed, especially in smaller samples.

For ordinal outcomes, the check compares category scores 1, 2, and so on;
these are not distances on an underlying continuous scale. Zero-inflated and
hurdle examples check the combined outcome, including zeros.
The [full comparison](simulation-studies/family-comparison.html) reports available
results by family, model settings, false alarms, and fitting failures. Read detection
rates alongside the false-alarm plots: how often a mismatch is flagged when there
is no residual partner correlation.

These curves describe the illustrated settings, rather than a general power
calculation. Sensitivity also depends on the outcome distribution, predictors, and
fitted model. A 95% simulated range does not guarantee a 5% false-positive rate,
and agreement does not establish that omitted dependence is negligible.

## Reuse the simulations

Use `response = "raw"` to compare responses without subtracting predictions.
Each call shows both member summaries and dyad means and half-differences,
which express the same covariance information.

```{r other-views, eval=FALSE}
raw_check <- check_partner_dependence(
  simulations, dyad = coupleID, role = gender, response = "raw", panels = TRUE
)
```

Supply `role` whenever the member distinction is meaningful. For substantively
interchangeable members, omit it: the check then reports a common member SD
and a partner correlation unchanged by arbitrary within-dyad member swaps.

With mixed dyad compositions, supplied roles produce one check per observed
role pair. Same-role pairs use exchangeable summaries. Each composition has
its own figure and pair count; `panels = FALSE` shows individual plots instead.
All checks use the same full-model simulations. A composition with fewer than
three usable pairs is listed but not checked.

## Requirements and technical details

See `?simulate_dyad_responses` for supported families. Prediction and simulation
use the model's fitted link. The fitted model must produce finite predictions
and responses.

In the Gaussian examples, `diag()` models separate role variances; `us()` also
allows partner covariance. `dispformula = ~0` fixes observation-level residual
variance near zero so the dyad block represents the residual covariance.
This model specification is Gaussian-specific.

Every simulation redraws all random effects and responses while keeping fitted
parameters and predictors fixed. The default check subtracts the same
prediction with random effects set to zero from each dataset. With nonlinear
links, this subtraction does not separate response variation into fixed-effect
and residual components.

Use unweighted cross-sectional data, with at least
three complete dyads and at most two fitted rows per dyad. Missing identifiers
and incomplete pairs are omitted and counted. If an identifier is absent from
the model frame, supply the data frame used to fit the model with `data`.
Rows excluded during fitting are handled automatically.

Sparse outcomes can produce undefined correlations. The check warns and uses
only defined simulations for each reference; plot labels show their counts.
Interpret these conditional references alongside the counts. An undefined observed
statistic or entirely undefined reference causes an error.

See `?simulate_dyad_responses` and `?check_partner_dependence` for the full
interface. The [development guide](README.md) links the separate mathematical
validation and simulation study.
