---
title: "Reference validation: exchangeable partner-dependence summaries"
output:
  html_document:
    toc: true
---

This report answers one narrow question: does
`calculate_partner_response_statistics()` correctly summarize partner
dependence for exchangeable dyads?

It checks the calculation against the equations in Woody and Sadler (2005)
and against David Kenny's Dingy application. It does **not** validate the
simulation-based predictive comparison, model centring, or calibration used
elsewhere in `dyadMLM`.

```{r setup, include=FALSE}
package_root <- normalizePath(file.path("..", ".."))
devtools::load_all(package_root, quiet = TRUE)

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)

tolerance <- 1e-12
```

# What is being checked?

For each dyad, the two member values are rewritten as:

\[
M_i = \frac{Y_{i1} + Y_{i2}}{2}, \qquad
D_i = \frac{Y_{i1} - Y_{i2}}{2}.
\]

In plain language, \(M_i\) is the dyad average and \(D_i\) is half of the gap
between the two members. When members are interchangeable, it is arbitrary
which member is written first. The sign of \(D_i\) is therefore arbitrary, and
its size is measured relative to zero rather than relative to its sample mean.

The reference calculation was written independently of the `dyadMLM`
implementation, directly from the published equations in Woody and Sadler
(2005), *Structural Equation Models for Interchangeable Dyads:
Being the Same Makes a Difference*, *Psychological Methods, 10*(2), 139--158
([article](https://doi.org/10.1037/1082-989X.10.2.139);
[APA supplement](https://supp.apa.org/psycarticles/supplemental/1082-989X.10.2.139/1082-989X.10.2.139_supp.html)).
No code or data were copied from the supplement.
The external comparison uses Kenny's
[Dingy application and documentation](https://davidakenny.net/DyadR/DyadRweb.htm).

# Make a small reproducible dataset

The data contain 48 complete dyads. They are built from a dyad average and a
half-difference so that both parts of the calculation are easy to see.

```{r validation-data}
set.seed(20260826L)

n_dyads <- 48L
generated_means <- rnorm(n_dyads, mean = 2, sd = 1.1)
generated_half_differences <- rnorm(n_dyads, mean = 0, sd = 0.55)

validation_data <- data.frame(
  dyad = seq_len(n_dyads),
  member_1 = generated_means + generated_half_differences,
  member_2 = generated_means - generated_half_differences
)

stopifnot(
  nrow(validation_data) >= 40L,
  all(complete.cases(validation_data))
)

dingy_csv <- file.path(
  tempdir(),
  "dyadMLM-partner-dependence-validation.csv"
)
write.csv(validation_data, dingy_csv, row.names = FALSE)

cat("Optional Dingy upload file:", dingy_csv, "\n")
head(validation_data)
```

The CSV is temporary and is not saved in the repository.

# Run the current dyadMLM calculation

The helper expects the member values in two stacked blocks and an index showing
which two rows belong to the same dyad.

```{r dyadmlm-calculation}
response_values <- c(
  validation_data$member_1,
  validation_data$member_2
)

paired_rows <- cbind(
  seq_len(n_dyads),
  n_dyads + seq_len(n_dyads)
)

dyadmlm_result <-
  dyadMLM:::calculate_partner_response_statistics(
    response_values,
    paired_rows,
    use_role_specific_statistics = FALSE
  )

dyadmlm_result
```

# Recalculate the same quantities from the published equations

This small function is independent of `dyadMLM`. Each calculation is written
out directly.

```{r woody-sadler-reference}
woody_sadler_reference <- function(member_1, member_2) {
  n <- length(member_1)
  stopifnot(n == length(member_2), n >= 2L)

  M <- (member_1 + member_2) / 2
  D <- (member_1 - member_2) / 2

  MS_between <- 2 * var(M)
  MS_within <- sum((member_1 - member_2)^2) / (2 * n)

  member_variance <- (MS_between + MS_within) / 2
  partner_covariance <- (MS_between - MS_within) / 2

  c(
    MS_between = MS_between,
    MS_within = MS_within,
    exchangeable_member_sd = sqrt(member_variance),
    exchangeable_partner_correlation =
      partner_covariance / member_variance,
    dyad_mean_sd = sqrt(MS_between / 2),
    half_difference_rms = sqrt(MS_within / 2)
  )
}

reference_calculation <- woody_sadler_reference(
  validation_data$member_1,
  validation_data$member_2
)

reference_result <- reference_calculation[names(dyadmlm_result)]

absolute_difference <- abs(dyadmlm_result - reference_result)
stopifnot(max(absolute_difference) <= tolerance)

cat("Largest absolute difference:", max(absolute_difference), "\n")
```

The following direct check makes the two key identities explicit.

```{r moment-identities}
M <- (validation_data$member_1 + validation_data$member_2) / 2
D <- (validation_data$member_1 - validation_data$member_2) / 2

identity_check <- data.frame(
  quantity = c("MS_between", "MS_within"),
  reference_value = unname(reference_calculation[c(
    "MS_between", "MS_within"
  )]),
  direct_value = c(
    2 * var(M),
    2 * mean(D^2)
  )
)

stopifnot(
  max(abs(identity_check$reference_value - identity_check$direct_value)) <=
    tolerance
)

identity_check
```

Thus, \(MS_{between}=2\operatorname{var}(M)\) and
\(MS_{within}=2\operatorname{mean}(D^2)\). There is one intentional
finite-sample difference between them: `var(M)` divides by \(n-1\), while
`mean(D^2)` divides by \(n\) and measures the half-differences about zero.

# Check that member order does not matter

We now swap the two member columns for the same reproducible set of 16 dyads.

```{r member-order-invariance}
set.seed(20260827L)
rows_to_swap <- sort(sample(seq_len(n_dyads), size = 16L))

swapped_data <- validation_data
old_member_1 <- swapped_data$member_1[rows_to_swap]
swapped_data$member_1[rows_to_swap] <-
  swapped_data$member_2[rows_to_swap]
swapped_data$member_2[rows_to_swap] <- old_member_1

swapped_D <- (swapped_data$member_1 - swapped_data$member_2) / 2

swapped_dyadmlm_result <-
  dyadMLM:::calculate_partner_response_statistics(
    c(swapped_data$member_1, swapped_data$member_2),
    paired_rows,
    use_role_specific_statistics = FALSE
  )

swapped_reference <- woody_sadler_reference(
  swapped_data$member_1,
  swapped_data$member_2
)[names(dyadmlm_result)]

stopifnot(
  max(abs(swapped_dyadmlm_result - dyadmlm_result)) <= tolerance,
  max(abs(swapped_reference - reference_result)) <= tolerance,
  max(abs(swapped_D[rows_to_swap] + D[rows_to_swap])) <= tolerance,
  max(abs(swapped_D^2 - D^2)) <= tolerance
)

data.frame(
  statistic = names(dyadmlm_result),
  before_swap = unname(dyadmlm_result),
  after_swap = unname(swapped_dyadmlm_result),
  absolute_difference = unname(abs(
    swapped_dyadmlm_result - dyadmlm_result
  ))
)
```

The swaps reverse the sign of \(D\) for those dyads but leave \(D^2\), and
therefore all four summaries, unchanged.

# Compare with Dingy

To repeat the external check in Dingy:

1. Upload the temporary CSV printed above.
2. Use `member_1` and `member_2` as the two versions of one mixed variable.
3. Treat the two positions as indistinguishable and request the
   nonindependence analysis.
4. Retain the tables and computer/lavaan output.

The following values were transcribed from the live Dingy run on 2026-08-26.
Dingy displayed lavaan 0.6-19. Its parameter estimates were rounded to two
decimal places and its chi-square statistic to three decimal places.

```{r dingy-values}
# Source: live Dingy web run with the generated CSV on 2026-08-26.
dingy_values <- c(
  common_member_variance = 1.36,
  cross_member_covariance = 0.67,
  partner_correlation = 0.50,
  nonindependence_chisq = 13.566,
  nonindependence_df = 1,
  nonindependence_p_upper_bound = 0.001
)

dingy_values
```

Dingy uses maximum-likelihood (ML) estimation. For this model, its moment
calculation can be reproduced directly in R. The important difference is that
ML divides the centred dyad-mean sum of squares by \(n\), whereas the
Woody--Sadler calculation above uses the usual sample variance and divides by
\(n-1\).

```{r reproduce-dingy-convention}
ml_variance_of_M <- mean((M - mean(M))^2)
mean_squared_D_about_zero <- mean(D^2)

dingy_reconstruction <- c(
  common_member_variance =
    ml_variance_of_M + mean_squared_D_about_zero,
  cross_member_covariance =
    ml_variance_of_M - mean_squared_D_about_zero
)

dingy_reconstruction <- c(
  dingy_reconstruction,
  partner_correlation =
    dingy_reconstruction[["cross_member_covariance"]] /
      dingy_reconstruction[["common_member_variance"]]
)

displayed_names <- c(
  "common_member_variance",
  "cross_member_covariance",
  "partner_correlation"
)

stopifnot(
  max(abs(
    round(dingy_reconstruction[displayed_names], 2) -
      dingy_values[displayed_names]
  )) <= tolerance
)

data.frame(
  quantity = displayed_names,
  Dingy_displayed = unname(dingy_values[displayed_names]),
  R_reconstruction = unname(dingy_reconstruction[displayed_names])
)
```

The nonindependence test is recorded for context only; it is not one of the
four `dyadMLM` summary statistics. Dingy also detected a difference between
the two sample position means in this random dataset. Because the positions
were assigned arbitrarily, that chance imbalance has no substantive meaning
for this validation.

# Results

For the table below, the Dingy column uses the full-precision R reconstruction
of Dingy's ML convention. This avoids treating Dingy's displayed rounding as a
real difference.

```{r final-results}
dingy_result <- c(
  exchangeable_member_sd =
    sqrt(dingy_reconstruction[["common_member_variance"]]),
  exchangeable_partner_correlation =
    dingy_reconstruction[["partner_correlation"]],
  dyad_mean_sd = sqrt(
    (
      dingy_reconstruction[["common_member_variance"]] +
        dingy_reconstruction[["cross_member_covariance"]]
    ) / 2
  ),
  half_difference_rms = sqrt(
    (
      dingy_reconstruction[["common_member_variance"]] -
        dingy_reconstruction[["cross_member_covariance"]]
    ) / 2
  )
)

interpretation <- c(
  "Exact; Dingy uses ML scaling.",
  "Exact; Dingy uses ML-scaled moments.",
  "Exact; Dingy divides by n rather than n - 1.",
  "Exact; Dingy uses the same about-zero calculation."
)

results <- data.frame(
  statistic = names(dyadmlm_result),
  dyadMLM = unname(dyadmlm_result),
  Woody_Sadler = unname(reference_result),
  Dingy_ML_reconstruction = unname(dingy_result),
  absolute_difference = unname(absolute_difference),
  interpretation = interpretation,
  check.names = FALSE
)

knitr::kable(results, digits = 9)
```

# Conclusion

1. **The dyadMLM calculation is correct.** All four results exactly reproduce
   the independent Woody--Sadler calculation within numerical tolerance.
2. **Member order does not matter.** Swapping arbitrary member labels changes
   the direction of some half-differences, but none of the four results.
3. **Dingy supports the same decomposition.** Its displayed partner
   correlation is 0.50, compared with 0.504 for `dyadMLM`. The small difference
   is fully explained by Dingy's ML division by \(n\) instead of \(n-1\); it is
   not a problem in `dyadMLM`. The half-difference result is the same because
   both methods measure it about zero and divide by \(n\).

These conclusions apply only to this deterministic summary calculation, not
to the larger simulation-based diagnostic in which it is used.
