---
title: "Checking partner dependence"
output: html_document
---

Does a fitted model reproduce how much partners' responses vary and how
strongly they are related? `dyadMLM` compares these features in the observed
data with the same features in datasets simulated from the model.

```{r setup, include=FALSE}
devtools::load_all(file.path("..", ".."), quiet = TRUE)
knitr::opts_chunk$set(
  message = FALSE, fig.width = 8, fig.height = 9, out.width = "100%"
)
```

## Fit a model

Prepare actor and partner predictors for the female-male dyads. Fit an APIM
that allows different effects and response variances for women and men,
but assumes no remaining correlation between partners.

```{r fit}
data <- prepare_dyad_data(
  dyads_cross, dyad = coupleID, member = personID, role = gender,
  predictors = provided_support, model_types = "apim",
  keep_compositions = "female-male"
)
model <- glmmTMB::glmmTMB(
  closeness ~ gender * (.provided_support_actor + .provided_support_partner) +
    diag(0 + gender | coupleID),
  dispformula = ~0, family = gaussian(), data = data
)
```

## Simulate, check, and plot

Simulate new responses and compare their standard deviations (SDs) and
correlations with the observed values. By default, the same model prediction
is subtracted from every dataset before comparison; details follow below.

```{r check}
par(mfcol = c(3, 2), mar = c(5.1, 4.1, 2.5, 1), cex = 0.66, cex.main = 0.9)
simulations <- simulate_dyad_responses(model, nsim = 1000, seed = 123)
check <- check_partner_dependence(
  simulations, dyad = coupleID, role = gender, ask = FALSE
)
```

The red line is the observed value; dashed lines mark the middle 95% of
simulated values. Here, the role SDs are within their ranges, but the partner correlation
exceeds its upper limit. These ranges and observed positions are descriptive, not
confidence intervals or p-values. Reuse the same `par()` layout before calling
`plot(check, ask = FALSE)` to show the panel again without repeating simulations.

The check is most useful for examining restrictions such as the omitted
partner correlation above. A flexible covariance model will usually reproduce
features it estimated from the same data; close agreement alone does not
establish good fit.

## Allow partner correlation

Replace the diagonal block with an unstructured block, keeping the fixed
effects unchanged. The same checking steps apply.

```{r flexible-model}
par(mfcol = c(3, 2), mar = c(5.1, 4.1, 2.5, 1), cex = 0.66, cex.main = 0.9)
flexible <- update(model, formula =
  closeness ~ gender * (.provided_support_actor + .provided_support_partner) +
    us(0 + gender | coupleID)
)
flexible_simulations <- simulate_dyad_responses(flexible, seed = 123)
flexible_check <- check_partner_dependence(
  flexible_simulations, dyad = coupleID, role = gender, ask = FALSE
)
```

## Reuse the simulations

Use `response = "raw"` to compare responses without subtracting predictions.
Each call shows both member summaries and dyad means and half-differences,
which express the same covariance information.

```{r other-views, eval=FALSE}
par(mfcol = c(3, 2), mar = c(5.1, 4.1, 2.5, 1), cex = 0.66, cex.main = 0.9)
raw_check <- check_partner_dependence(
  simulations, dyad = coupleID, role = gender, response = "raw", ask = FALSE
)
```

Supply `role` whenever the member distinction is meaningful. For substantively
interchangeable members, omit it: the check then reports a common member SD
and a partner correlation unchanged by arbitrary within-dyad member swaps.
Use `par(mfcol = c(2, 2))` to arrange its four plots in a panel.

## Requirements and technical details

The simulation and check calls also support Poisson, NB1, NB2, Tweedie, Gamma,
and beta models. They use the model's fitted link for prediction and simulation.
Use an appropriate fitted model that produces finite predictions and responses.

In the Gaussian examples, `diag()` models separate role variances; `us()` also
allows partner covariance. `dispformula = ~0` fixes observation-level residual
variance near zero so the dyad block represents the residual covariance.
This model specification is Gaussian-specific.

Every simulation redraws all random effects and responses while keeping fitted
parameters and predictors fixed. The default check subtracts the same
prediction with random effects set to zero from each dataset. Under nonlinear
links, this is a common transformation, not a residual covariance decomposition.

Use unweighted cross-sectional data without zero inflation, with at least
three complete dyads and at most two fitted rows per dyad. Missing identifiers
and incomplete pairs are omitted and counted. Identifier columns must be in
the fitted model frame; external vectors must match its retained row order.

Sparse outcomes can produce undefined correlations. The check warns and uses
only defined simulations for each reference; plot labels show their counts.
Interpret these conditional references alongside the counts. An undefined observed
statistic or entirely undefined reference causes an error.

See `?simulate_dyad_responses` and `?check_partner_dependence` for the full
interface. The [development guide](README.md) links the separate mathematical
validation and simulation study.
