---
title: "Extended cross-sectional Gaussian partner-dependence example"
output: html_document
---

This extended development example exercises the cross-sectional Gaussian
`glmmTMB` feature across a fuller model sequence. See the shorter
[development review](partner-dependence-review.Rmd) for the primary review
artifact.

# When is this check most informative?

The check is most useful when a model makes a clear simplifying assumption
that could be wrong. For example, it can show whether assuming that two
different roles vary by the same amount hides an important role difference. If
the model was allowed to learn the same amount of variation and the same
partner relationship freely from the data, a close match is expected. It mainly
shows that the model can reproduce what it already learned; it provides little
new evidence that the model describes the data well.

The [repeated simulation study](partner-dependence-outer-simulation-study.Rmd)
gave encouraging initial results. For the deliberately large problems studied,
the observed pattern stood out from the model's simulations in 84--97% of
datasets with 50 dyads and in every dataset with 200 dyads. All 7,000
fitted-model runs completed successfully. However, flexible models that had
already learned the same pattern nearly always looked reassuring. A close
match therefore does not by itself show that the model fits well, and smaller
problems have not yet been studied.

Similar checks may become useful when a generalized model needs simpler
assumptions about the partner pattern or models response variation separately.
These are future use cases: the implementation shown here currently supports
only cross-sectional Gaussian identity-link models.

```{r setup, include=FALSE}
package_root <- normalizePath(file.path("..", ".."))
devtools::load_all(package_root, quiet = TRUE)

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```

# Distinguishable dyads

## Prep data

```{r}
diagnostic_data <- prepare_dyad_data(
    dyads_cross,
    dyad = coupleID,
    member = personID,
    role = gender,
    predictors = provided_support,
    model_types = "apim",
    add_apim_gmc_predictors = TRUE,
    keep_compositions = "female-male",
    seed = 13000L
)
```


## Flexible role-specific covariance model

```{r}
m_c <- glmmTMB::glmmTMB(
    closeness ~
        0 +
        .is_female +
        .is_male +
        .is_female:.provided_support_gmc_actor +
        .is_male:.provided_support_gmc_actor +
        .is_female:.provided_support_gmc_partner +
        .is_male:.provided_support_gmc_partner +
        (0 + .is_female + .is_male | coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

The actor and partner fixed effects make the fitted means vary across
observations. The centred check removes this fitted pattern, whereas the raw
check retains it. Both checks use the same simulated datasets.

### Model-centred responses

```{r}
m_c_simulations <- simulate_dyad_responses(
  m_c,
  nsim = 5000,
  seed = 13001L
)

m_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "model-centred"
  )
```

### Raw responses

```{r}
m_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "raw"
  )
```


## Constrained models

```{r}
m_w1 <- glmmTMB::glmmTMB(
    closeness ~
        0 +
        .is_female +
        .is_male +
        .is_female:.provided_support_gmc_actor +
        .is_male:.provided_support_gmc_actor +
        .is_female:.provided_support_gmc_partner +
        .is_male:.provided_support_gmc_partner +
        (1 | coupleID),
    dispformula = ~ 1,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
m_w1_simulations <- simulate_dyad_responses(
  m_w1,
  nsim = 5000,
  seed = 13002L
)

# Model-centred responses
m_w1_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "model-centred"
  )
```


```{r}
# Raw responses
m_w1_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "raw"
  )
```

In these data, the check did not show a clear mismatch for this simpler
covariance structure. It retains the role-specific fixed effects but constrains
the remaining covariance to equal marginal variances and nonnegative compound
symmetry. The simpler model has slightly lower AIC, and the likelihood-ratio
comparison below provides no clear evidence that the more flexible
role-specific covariance improves fit. This is not evidence that partners are
interchangeable or that the two covariance structures are equivalent.

```{r}
compare_nested_models(
  m_w1,
  m_c
)
```


```{r}
m_w2 <- glmmTMB::glmmTMB(
    closeness ~
        0 +
        .is_female +
        .is_male +
        .is_female:.provided_support_gmc_actor +
        .is_male:.provided_support_gmc_actor +
        .is_female:.provided_support_gmc_partner +
        .is_male:.provided_support_gmc_partner +
        (1 | coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
m_w2_simulations <- simulate_dyad_responses(
  m_w2,
  nsim = 5000,
  seed = 13003L
)

# Model-centred responses
m_w2_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "model-centred"
  )
```


```{r}
# Raw responses
m_w2_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "raw"
  )
```


```{r}
m_w3 <- glmmTMB::glmmTMB(
    closeness ~
        0 +
        .is_female +
        .is_male +
        .is_female:.provided_support_gmc_actor +
        .is_male:.provided_support_gmc_actor +
        .is_female:.provided_support_gmc_partner +
        .is_male:.provided_support_gmc_partner +
        (0 + .is_female + .is_male || coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
m_w3_simulations <- simulate_dyad_responses(
  m_w3,
  nsim = 5000,
  seed = 13004L
)

# Model-centred responses
m_w3_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "model-centred"
  )
```


```{r}
# Raw responses
m_w3_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "raw"
  )
```

# Sensitivity analysis: treating partners as exchangeable

Here, the female-male data are deliberately treated as exchangeable. The
pooled check assesses the imposed covariance structure; the gender-aware check
asks whether that pooling hides role-specific differences.

These models have only a constant fixed-effect prediction. Their raw and
centred checks should therefore match; both are shown for completeness.

## Prep data

```{r}
diagnostic_data <- prepare_dyad_data(
    dyads_cross,
    dyad = coupleID,
    member = personID,
    role = gender,
    model_types = "none",
    keep_compositions = "female-male",
    set_exchangeable_compositions = "female-male",
    seed = 13000L
)
```


## Exchangeable mean/difference model

```{r}
e_c <- glmmTMB::glmmTMB(
    closeness ~
        1 +

      (1 | coupleID) +
      (0 + .member_contrast_arbitrary | coupleID),

    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
e_c_simulations <- simulate_dyad_responses(
  e_c,
  nsim = 5000,
  seed = 13005L
)

# Model-centred responses
e_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "model-centred"
  )
```


```{r}
# Raw responses
e_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "raw"
  )
```

## Check the same model with gender labels

```{r}
# Model-centred responses
e_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "model-centred"
  )
```


```{r}
# Raw responses
e_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "raw"
  )
```

The pooled check can look adequate while the gender-aware check reveals omitted
role differences. However, this sensitivity model also removes role-specific
fixed effects, so a discrepancy cannot be attributed to the covariance
structure alone.

## More restrictive exchangeable models

```{r}
e_w1 <- glmmTMB::glmmTMB(
    closeness ~
        1,
    dispformula = ~ 1,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
e_w1_simulations <- simulate_dyad_responses(
  e_w1,
  nsim = 5000,
  seed = 13006L
)

# Model-centred responses
e_w1_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "model-centred"
  )
```


```{r}
# Raw responses
e_w1_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "raw"
  )
```

```{r}
e_w2 <- glmmTMB::glmmTMB(
    closeness ~
        1 + (1 | coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
e_w2_simulations <- simulate_dyad_responses(
  e_w2,
  nsim = 5000,
  seed = 13007L
)

# Model-centred responses
e_w2_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "model-centred"
  )
```


```{r}
# Raw responses
e_w2_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "raw"
  )
```


```{r}
e_w3 <- glmmTMB::glmmTMB(
    closeness ~
        1 + (1 | coupleID/personID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
e_w3_simulations <- simulate_dyad_responses(
  e_w3,
  nsim = 5000,
  seed = 13008L
)

# Model-centred responses
e_w3_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "model-centred"
  )
```


```{r}
# Raw responses
e_w3_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "raw"
  )
```
