---
title: "Outer simulation study: Gaussian partner-dependence checks"
output:
  html_document:
    toc: true
    number_sections: true
    code_folding: hide
params:
  outer_repetitions: 500
  inner_simulations: 1000
---

This development study repeats the complete workflow many times:

```text
known population -> new dataset -> fitted model -> predictive check
```

It asks two simple questions: Does the check behave sensibly when the model
matches the data, and can it reveal deliberately wrong assumptions about the
partner relationship or differences between roles? The partner-summary
calculations are validated separately. The results here are descriptive, not a
formal hypothesis test.

```{r setup, include=FALSE}
package_root <- normalizePath(file.path("..", ".."))
devtools::load_all(package_root, quiet = TRUE)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)

n_outer <- as.integer(params$outer_repetitions)
n_inner <- as.integer(params$inner_simulations)
stopifnot(n_outer >= 1L, n_inner >= 20L)

sample_sizes <- c(50L, 200L)
base_seed <- 20260826L
is_pilot <- n_outer < 500L || n_inner < 1000L
tail_rate_mcse <- sqrt(0.05 * 0.95 / n_outer)
largest_rate_mcse <- 0.5 / sqrt(n_outer)
```

# Design

Every outer repetition generates new responses and refits each candidate model.
The predictor design is held fixed, matching the check's
conditional-on-design target. Correct and restricted models are fitted to the
same generated datasets, allowing paired comparisons.

The default report uses `r n_outer` outer repetitions, `r n_inner` inner
simulations, and `r paste(sample_sizes, collapse = " and ")` dyads. The four
known populations and seven fitted-model conditions are shown below. With this
number of outer repetitions, the Monte Carlo standard error is about
`r sprintf("%.1f", 100 * tail_rate_mcse)` percentage points near a 5% rate and
at most `r sprintf("%.1f", 100 * largest_rate_mcse)` points near a 50% rate.

```{r scenarios}
truths <- tibble::tribble(
  ~truth, ~population, ~sd_1, ~sd_2, ~rho, ~mean_1, ~mean_2, ~role_aware,
  "positive", "Exchangeable: SD = 1, correlation = 0.5",
    1, 1, 0.5, 0, 0, FALSE,
  "zero", "Exchangeable: SD = 1, correlation = 0",
    1, 1, 0, 0, 0, FALSE,
  "negative", "Exchangeable: SD = 1, correlation = -0.5",
    1, 1, -0.5, 0, 0, FALSE,
  "distinguishable", "Distinguishable: SDs = 0.8/1.4, correlation = 0.4",
    0.8, 1.4, 0.4, -0.3, 0.3, TRUE
)

scenarios <- tibble::tribble(
  ~scenario, ~truth, ~fitted_model, ~description, ~model_correct,
  "positive_correct", "positive", "exchangeable",
    "Correct exchangeable covariance", TRUE,
  "positive_independent", "positive", "independent",
    "Positive dependence omitted", FALSE,
  "zero_correct", "zero", "independent",
    "Correct independence model", TRUE,
  "negative_correct", "negative", "exchangeable",
    "Correct exchangeable covariance", TRUE,
  "negative_independent", "negative", "independent",
    "Negative dependence omitted", FALSE,
  "distinguishable_correct", "distinguishable", "unstructured",
    "Correct role-specific covariance", TRUE,
  "distinguishable_exchangeable", "distinguishable", "exchangeable_roles",
    "Equal role spreads imposed", FALSE
)

scenarios |>
  dplyr::left_join(truths[c("truth", "population")], by = "truth") |>
  dplyr::select(
    scenario = .data$scenario,
    population = .data$population,
    fitted_model = .data$description
  ) |>
  knitr::kable()
```

The deliberately large, clear problems are intentional. This first study asks
whether the check can reveal the restrictions it was designed to inspect. It
does not yet show how sensitive the check is to smaller problems.

# Data and models

Responses are generated directly from a bivariate Gaussian distribution,
independently of `glmmTMB`. The predictor combines dyad-shared and
member-specific variation and has a population slope of 0.6.

```{r data-and-model-functions}
make_design <- function(n_dyads, seed) {
  set.seed(seed)
  shared_x <- rep(stats::rnorm(n_dyads, sd = 0.6), each = 2L)
  member_x <- stats::rnorm(2L * n_dyads, sd = 0.8)

  data.frame(
    dyad = factor(rep(seq_len(n_dyads), each = 2L)),
    role = factor(
      rep(c("member_1", "member_2"), n_dyads),
      levels = c("member_1", "member_2")
    ),
    member_sign = rep(c(-1, 1), n_dyads),
    x = shared_x + member_x
  )
}

draw_data <- function(design, truth, seed) {
  set.seed(seed)
  n_dyads <- nlevels(design$dyad)
  z_1 <- stats::rnorm(n_dyads)
  z_2 <- stats::rnorm(n_dyads)

  error_1 <- truth$sd_1 * z_1
  error_2 <- truth$sd_2 * (
    truth$rho * z_1 + sqrt(1 - truth$rho^2) * z_2
  )
  mean_1 <- truth$mean_1 + 0.6 * design$x[design$role == "member_1"]
  mean_2 <- truth$mean_2 + 0.6 * design$x[design$role == "member_2"]

  data <- design
  data$expected_mean <- as.vector(rbind(mean_1, mean_2))
  data$y <- data$expected_mean + as.vector(rbind(error_1, error_2))
  data
}

fit_model <- function(data, model_name) {
  switch(
    model_name,
    exchangeable = glmmTMB::glmmTMB(
      y ~ 1 + x + (1 | dyad) + (0 + member_sign | dyad),
      dispformula = ~0,
      family = gaussian(),
      data = data
    ),
    independent = glmmTMB::glmmTMB(
      y ~ 1 + x,
      family = gaussian(),
      data = data
    ),
    unstructured = glmmTMB::glmmTMB(
      y ~ 0 + role + x + (0 + role | dyad),
      dispformula = ~0,
      family = gaussian(),
      data = data
    ),
    exchangeable_roles = glmmTMB::glmmTMB(
      y ~ 0 + role + x + (1 | dyad) + (0 + member_sign | dyad),
      dispformula = ~0,
      family = gaussian(),
      data = data
    ),
    stop("Unknown model: ", model_name, call. = FALSE)
  )
}

extract_results <- function(check, scenario, truth, n_dyads, outer_rep,
                            method) {
  result <- check$statistics_table |>
    dplyr::filter(.data$parameterization == "member") |>
    dplyr::mutate(
      outside_95 =
        .data$observed_value < .data$replicated_lower |
        .data$observed_value > .data$replicated_upper,
      direction = dplyr::case_when(
        .data$observed_value < .data$replicated_lower ~ "below",
        .data$observed_value > .data$replicated_upper ~ "above",
        TRUE ~ "inside"
      ),
      scenario = scenario,
      truth = truth,
      n_dyads = n_dyads,
      outer_rep = outer_rep,
      method = method
    )

  result |>
    dplyr::select(
      .data$scenario, .data$truth, .data$n_dyads, .data$outer_rep,
      .data$method, .data$statistic_name,
      .data$observed_value, .data$replicated_median,
      .data$replicated_lower, .data$replicated_upper,
      .data$observed_quantile, .data$outside_95, .data$direction
    )
}
```

The exchangeable fit uses shared and member-sign random effects. Their summed
and differenced variances span equal member variances with either positive or
negative partner correlation. The role-specific fit estimates both member
variances and their covariance directly.

# Control using the known truth

For one correct population, this control draws comparison datasets using the
true values that generated the data rather than values estimated by a fitted
model. The observed and comparison datasets should then behave alike, so their
positions should spread roughly evenly from 0 to 1. This checks the ranking
step separately from model fitting.

```{r oracle-function}
run_oracle <- function(data, truth, n_dyads, outer_rep, nsim, seed) {
  set.seed(seed)
  z_1 <- matrix(stats::rnorm(n_dyads * nsim), nrow = n_dyads)
  z_2 <- matrix(stats::rnorm(n_dyads * nsim), nrow = n_dyads)
  error_1 <- truth$sd_1 * z_1
  error_2 <- truth$sd_2 * (
    truth$rho * z_1 + sqrt(1 - truth$rho^2) * z_2
  )

  simulated <- matrix(NA_real_, nrow = nsim, ncol = 2L * n_dyads)
  simulated[, seq.int(1L, 2L * n_dyads, 2L)] <- t(
    error_1 + data$expected_mean[data$role == "member_1"]
  )
  simulated[, seq.int(2L, 2L * n_dyads, 2L)] <- t(
    error_2 + data$expected_mean[data$role == "member_2"]
  )

  simulations <- list(
    observed_response = data$y,
    simulated_responses = simulated,
    response_center = data$expected_mean,
    model_frame = data.frame(y = data$y),
    backend = "glmmTMB",
    family = "gaussian",
    link = "identity",
    reference = "known-parameter oracle",
    random_effects = "known covariance",
    parameter_uncertainty = "not applicable",
    nsim = nsim,
    seed = seed
  )
  class(simulations) <- c("dyadMLM_response_simulations", "list")

  check <- check_partner_dependence(
    simulations,
    dyad = data$dyad,
    plot = FALSE
  )
  extract_results(
    check,
    scenario = "positive_oracle",
    truth = truth$truth,
    n_dyads,
    outer_rep,
    method = "oracle"
  )
}
```

# Fitted-model check

Fit errors, convergence problems, non-positive-definite Hessians, and check
errors are recorded rather than retried or silently discarded.

```{r fitted-check-function}
run_fitted_check <- function(data, scenario, role_aware, n_dyads,
                             outer_rep, nsim, seed) {
  scenario_name <- scenario$scenario
  truth_name <- scenario$truth
  fit_warnings <- character()
  fit <- tryCatch(
    withCallingHandlers(
      fit_model(data, scenario$fitted_model),
      warning = function(warning) {
        fit_warnings <<- c(fit_warnings, conditionMessage(warning))
        invokeRestart("muffleWarning")
      }
    ),
    error = function(error) error
  )

  status <- tibble::tibble(
    scenario = scenario_name,
    truth = truth_name,
    n_dyads = n_dyads,
    outer_rep = outer_rep,
    status = NA_character_,
    convergence = NA_integer_,
    pd_hess = NA,
    warnings = paste(unique(fit_warnings), collapse = " | "),
    error = NA_character_
  )

  if (inherits(fit, "error")) {
    status$status <- "fit_error"
    status$error <- conditionMessage(fit)
    return(list(status = status, statistics = NULL))
  }

  status$convergence <- fit$fit$convergence
  status$pd_hess <- fit$sdr$pdHess
  if (!isTRUE(status$convergence == 0L) || !isTRUE(status$pd_hess)) {
    status$status <- "fit_problem"
    return(list(status = status, statistics = NULL))
  }

  check_warnings <- character()
  check <- tryCatch({
    withCallingHandlers({
      simulations <- simulate_dyad_responses(fit, nsim = nsim, seed = seed)
      if (role_aware) {
        check_partner_dependence(
          simulations,
          dyad = data$dyad,
          role = data$role,
          plot = FALSE
        )
      } else {
        check_partner_dependence(simulations, dyad = data$dyad, plot = FALSE)
      }
    }, warning = function(warning) {
      check_warnings <<- c(check_warnings, conditionMessage(warning))
      invokeRestart("muffleWarning")
    })
  }, error = function(error) error)
  status$warnings <- paste(
    unique(c(fit_warnings, check_warnings)),
    collapse = " | "
  )

  if (inherits(check, "error")) {
    status$status <- "check_error"
    status$error <- conditionMessage(check)
    return(list(status = status, statistics = NULL))
  }

  stopifnot(check$n_pairs == n_dyads)
  if (role_aware) {
    stopifnot(identical(check$role_order, c("member_1", "member_2")))
  }

  status$status <- "success"
  list(
    status = status,
    statistics = extract_results(
      check,
      scenario$scenario,
      scenario$truth,
      n_dyads,
      outer_rep,
      method = "fitted plug-in"
    )
  )
}
```

# Outer loop

The loop is kept explicit: generate one dataset, run its oracle control where
applicable, and fit every candidate model belonging to that population.

```{r run-study}
designs <- lapply(seq_along(sample_sizes), function(i) {
  make_design(sample_sizes[[i]], seed = base_seed + 1000L * i)
})

statuses <- list()
statistics <- list()
status_i <- 0L
statistics_i <- 0L

for (size_i in seq_along(sample_sizes)) {
  n_dyads <- sample_sizes[[size_i]]

  for (truth_i in seq_len(nrow(truths))) {
    truth <- truths[truth_i, ]
    truth_scenarios <- scenarios[scenarios$truth == truth$truth, ]

    for (outer_rep in seq_len(n_outer)) {
      data_seed <- base_seed + 100000L * truth_i +
        10000L * size_i + outer_rep
      data <- draw_data(designs[[size_i]], truth, seed = data_seed)

      if (truth$truth == "positive") {
        statistics_i <- statistics_i + 1L
        statistics[[statistics_i]] <- run_oracle(
          data, truth, n_dyads, outer_rep, n_inner,
          seed = data_seed + 20000000L
        )
      }

      for (scenario_i in seq_len(nrow(truth_scenarios))) {
        scenario <- truth_scenarios[scenario_i, ]
        global_i <- match(scenario$scenario, scenarios$scenario)
        result <- run_fitted_check(
          data,
          scenario,
          truth$role_aware,
          n_dyads,
          outer_rep,
          n_inner,
          seed = data_seed + 10000000L + 1000L * global_i
        )

        status_i <- status_i + 1L
        statuses[[status_i]] <- result$status
        if (!is.null(result$statistics)) {
          statistics_i <- statistics_i + 1L
          statistics[[statistics_i]] <- result$statistics
        }
      }
    }
  }
}

run_status <- dplyr::bind_rows(statuses)
outer_results <- dplyr::bind_rows(statistics)
```

# Results

## Completion

```{r completion, echo=FALSE}
completion <- run_status |>
  dplyr::group_by(.data$scenario, .data$n_dyads) |>
  dplyr::summarise(
    attempted = dplyr::n(),
    success_rate = mean(.data$status == "success"),
    warning_runs = sum(nzchar(.data$warnings)),
    fit_problems = sum(.data$status == "fit_problem"),
    fit_errors = sum(.data$status == "fit_error"),
    check_errors = sum(.data$status == "check_error"),
    .groups = "drop"
  )

knitr::kable(completion, digits = 3)
```

All fitted-model rates below are conditional on a successful fit and check.

## Oracle and correct models

Only the member parameterization is used as primary evidence. The
mean/difference view contains equivalent covariance information and is not
counted as another set of checks.

```{r rates, include=FALSE}
rates <- outer_results |>
  dplyr::group_by(
    .data$method,
    .data$scenario,
    .data$n_dyads,
    .data$statistic_name
  ) |>
  dplyr::summarise(
    repetitions = dplyr::n(),
    outside_count = sum(.data$outside_95),
    below = mean(.data$direction == "below"),
    outside = mean(.data$outside_95),
    above = mean(.data$direction == "above"),
    mean_position = mean(.data$observed_quantile),
    sd_position = stats::sd(.data$observed_quantile),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    mcse = sqrt(.data$outside * (1 - .data$outside) / .data$repetitions)
  )

oracle_rates <- rates |>
  dplyr::filter(.data$method == "oracle")

correct_scenarios <- scenarios$scenario[scenarios$model_correct]
correct_rates <- rates |>
  dplyr::filter(
    .data$method == "fitted plug-in",
    .data$scenario %in% correct_scenarios
  )
```

In the control using the known truth, about 2.5% of observed positions should
fall in each tail, and the position SD should be near
`r sprintf("%.3f", 1 / sqrt(12))`. The finite number of repeated datasets and
simulations adds some random simulation uncertainty.

```{r oracle-table, echo=FALSE}
knitr::kable(oracle_rates, digits = 3)
```

For flexible fitted models, positions can cluster near the middle because the
model estimated the same amounts of variation and the same partner relationship
from the data. A close match is therefore expected and adds little new
information. The independence model is a useful comparison: it estimates the
amount of variation but forces the partner relationship to zero, so its
correlation result can still reveal a problem.

```{r correct-table, echo=FALSE}
knitr::kable(correct_rates, digits = 3)
```

The displayed MCSE is the usual plug-in estimate. A value of zero when every
run is inside or every run is flagged does not prove that the underlying rate
is exactly zero or one.

```{r position-plot, echo=FALSE}
plot_data <- outer_results |>
  dplyr::filter(
    .data$n_dyads == max(sample_sizes),
    .data$statistic_name == "exchangeable_partner_correlation",
    .data$scenario %in% c("positive_oracle", "positive_correct")
  )

old_par <- graphics::par(mfrow = c(1, 2))
for (shown in c("positive_oracle", "positive_correct")) {
  graphics::plot(
    stats::ecdf(plot_data$observed_quantile[plot_data$scenario == shown]),
    main = if (shown == "positive_oracle") "Oracle" else "Correct fitted model",
    xlab = "Observed position",
    ylab = "Cumulative proportion",
    xlim = c(0, 1),
    ylim = c(0, 1)
  )
  graphics::abline(a = 0, b = 1, lty = 2, col = "grey40")
}
graphics::par(old_par)
```

## Deliberately restricted models

The expected directions were specified in advance: omitted positive
correlation should appear above the reference, omitted negative correlation
below it, and incorrectly equal role spreads should place the smaller observed
SD below and the larger observed SD above the reference.

```{r paired-comparisons, echo=FALSE}
targets <- tibble::tribble(
  ~constrained, ~correct, ~statistic_name, ~expected,
  "positive_independent", "positive_correct",
    "exchangeable_partner_correlation", "above",
  "negative_independent", "negative_correct",
    "exchangeable_partner_correlation", "below",
  "distinguishable_exchangeable", "distinguishable_correct",
    "role_1_sd", "below",
  "distinguishable_exchangeable", "distinguishable_correct",
    "role_2_sd", "above"
)

fitted_results <- outer_results |>
  dplyr::filter(.data$method == "fitted plug-in")

constrained_flags <- targets |>
  dplyr::inner_join(
    fitted_results,
    by = c("constrained" = "scenario", "statistic_name")
  ) |>
  dplyr::transmute(
    .data$truth, .data$n_dyads, .data$outer_rep,
    .data$constrained, .data$correct, .data$statistic_name, .data$expected,
    constrained_flag = .data$direction == .data$expected
  )

correct_flags <- targets |>
  dplyr::inner_join(
    fitted_results,
    by = c("correct" = "scenario", "statistic_name")
  ) |>
  dplyr::transmute(
    .data$truth, .data$n_dyads, .data$outer_rep,
    .data$constrained, .data$correct, .data$statistic_name, .data$expected,
    correct_flag = .data$direction == .data$expected
  )

paired <- dplyr::inner_join(
  constrained_flags,
  correct_flags,
  by = c(
    "truth", "n_dyads", "outer_rep", "constrained", "correct",
    "statistic_name", "expected"
  )
) |>
  dplyr::mutate(
    gain = as.integer(.data$constrained_flag) -
      as.integer(.data$correct_flag)
  )

paired_summary <- paired |>
  dplyr::group_by(
    .data$constrained,
    .data$n_dyads,
    .data$statistic_name,
    .data$expected
  ) |>
  dplyr::summarise(
    paired_runs = dplyr::n(),
    correct_rate = mean(.data$correct_flag),
    constrained_rate = mean(.data$constrained_flag),
    paired_gain = mean(.data$gain),
    gain_mcse = stats::sd(.data$gain) / sqrt(dplyr::n()),
    .groups = "drop"
  )

knitr::kable(paired_summary, digits = 3)
```

Each correct and restricted model used the same generated dataset. The paired
gain therefore shows what changed when the wrong restriction was added, rather
than mixing that change with differences between datasets.

# Conclusion

```{r conclusion, echo=FALSE, results='asis'}
if (is_pilot) {
  cat(
    "This is a **pilot render**. It verifies that the complete study runs, ",
    "but its rates are too imprecise for validation conclusions. Re-render ",
    "with the default 500 outer repetitions and 1,000 inner simulations.\n"
  )
} else {
  oracle_range <- range(oracle_rates$outside)
  oracle_max_mcse <- max(oracle_rates$mcse)
  oracle_sd_range <- range(oracle_rates$sd_position)
  flexible_correct <- correct_rates |>
    dplyr::filter(.data$scenario != "zero_correct")
  flexible_range <- range(flexible_correct$outside)
  flexible_sd_range <- range(flexible_correct$sd_position)
  small_sample <- paired_summary |>
    dplyr::filter(.data$n_dyads == min(sample_sizes))
  large_sample <- paired_summary |>
    dplyr::filter(.data$n_dyads == max(sample_sizes))
  minimum_success <- min(completion$success_rate)

  cat(sprintf(
    paste0(
      "Using the known population values as the reference, %.1f%% to %.1f%% ",
      "of observed summaries fell outside the middle 95%%. Their position SDs ",
      "ranged from %.3f to %.3f, close to the expected value of %.3f. For the ",
      "flexible correct fits, %.1f%% to %.1f%% fell outside and the position SDs ",
      "were only %.3f to %.3f. In plain language, these fitted-model results ",
      "were tightly grouped near the middle. That close match is expected ",
      "because the model first learned the same features from the data; it does ",
      "not by itself show that the model describes the data well. The largest ",
      "estimated simulation uncertainty for the first set of rates was %.1f ",
      "percentage points.\n\n"
    ),
    100 * oracle_range[[1L]], 100 * oracle_range[[2L]],
    oracle_sd_range[[1L]], oracle_sd_range[[2L]],
    1 / sqrt(12),
    100 * flexible_range[[1L]], 100 * flexible_range[[2L]],
    flexible_sd_range[[1L]], flexible_sd_range[[2L]],
    100 * oracle_max_mcse
  ))

  cat(sprintf(
    paste0(
      "With %d dyads, the deliberately wrong assumptions placed the targeted ",
      "summary outside the reference interval in the expected direction in ",
      "%.1f%% to %.1f%% of datasets. With %d dyads, this happened in %.1f%% to ",
      "%.1f%% of datasets. The lowest cell-specific fit-and-check success rate ",
      "was %.1f%%.\n\n"
    ),
    min(sample_sizes),
    100 * min(small_sample$constrained_rate),
    100 * max(small_sample$constrained_rate),
    max(sample_sizes),
    100 * min(large_sample$constrained_rate),
    100 * max(large_sample$constrained_rate),
    100 * minimum_success
  ))

  cat(
    "For the large, clear problems studied here, the check was useful when the ",
    "model imposed a simplifying assumption, including forcing different roles ",
    "to vary by the same amount. This is initial evidence that the check can ",
    "reveal such problems. It does not yet show how well the check finds small ",
    "problems or how it behaves for generalized or longitudinal models. Observed ",
    "positions remain descriptive; they are not p-values.\n"
  )
}
```

# Reproducibility

Everything needed to run and interpret the study is in this file. A quick code
check can override the YAML defaults without editing the report:

```r
rmarkdown::render(
  "dev/diagnostic_checks/partner-dependence-outer-simulation-study.Rmd",
  params = list(outer_repetitions = 20, inner_simulations = 99)
)
```

```{r session-info}
sessionInfo()
```
