---
title: "Extended cross-sectional Gaussian partner-dependence example"
output: html_document
---

This example asks whether fitted models reproduce how much partners' responses
vary and how strongly they are related. It walks through flexible and
restricted models using the same cross-sectional data. For a shorter worked
comparison, see the [development review](partner-dependence-review.Rmd).
The current feature supports Gaussian identity-link `glmmTMB` models.

# How to read the checks

`simulate_dyad_responses()` creates reusable datasets from a fitted model.
`check_partner_dependence()` then compares the observed partner summaries with
the same summaries in those datasets. Each histogram shows what the model
generated, with the observed value and middle 95% interval marked. An observed
value near an edge is a reason to investigate; the interval and observed
position are descriptive, not formal tests or p-values.

By default, the check removes the same fitted mean pattern from observed and
simulated responses. The raw view retains that pattern. Supply `role` for
meaningful member roles; omit it only when treating members as interchangeable.
Both choices are illustrated below.

The check is most informative when examining a simplifying assumption, such
as equal variation for both roles. A model that freely estimated the same
pattern will usually reproduce it. A close match therefore provides little
new evidence that such a model describes the data well.

```{r setup, include=FALSE}
package_root <- normalizePath(file.path("..", ".."))
devtools::load_all(package_root, quiet = TRUE)

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```

# Distinguishable dyads

## Prep data

```{r}
diagnostic_data <- prepare_dyad_data(
    dyads_cross,
    dyad = coupleID,
    member = personID,
    role = gender,
    predictors = provided_support,
    model_types = "apim",
    add_apim_gmc_predictors = TRUE,
    keep_compositions = "female-male",
    seed = 13000L
)
```


## Flexible role-specific covariance model

```{r}
m_c <- glmmTMB::glmmTMB(
    closeness ~
        0 +
        .is_female +
        .is_male +
        .is_female:.provided_support_gmc_actor +
        .is_male:.provided_support_gmc_actor +
        .is_female:.provided_support_gmc_partner +
        .is_male:.provided_support_gmc_partner +
        (0 + .is_female + .is_male | coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

The actor and partner fixed effects make the fitted means vary across
observations. The centred check removes this fitted pattern, whereas the raw
check retains it. Both checks use the same simulated datasets.

### Model-centred responses

```{r}
m_c_simulations <- simulate_dyad_responses(
  m_c,
  nsim = 5000,
  seed = 13001L
)

m_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "model-centred"
  )
```

### Raw responses

```{r}
m_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "raw"
  )
```


## Constrained models

```{r}
m_w1 <- glmmTMB::glmmTMB(
    closeness ~
        0 +
        .is_female +
        .is_male +
        .is_female:.provided_support_gmc_actor +
        .is_male:.provided_support_gmc_actor +
        .is_female:.provided_support_gmc_partner +
        .is_male:.provided_support_gmc_partner +
        (1 | coupleID),
    dispformula = ~ 1,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
m_w1_simulations <- simulate_dyad_responses(
  m_w1,
  nsim = 5000,
  seed = 13002L
)

# Model-centred responses
m_w1_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "model-centred"
  )
```


```{r}
# Raw responses
m_w1_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "raw"
  )
```

In these data, the check did not show a clear mismatch for this simpler
covariance structure. It retains the role-specific fixed effects but constrains
the remaining covariance to equal marginal variances and nonnegative compound
symmetry. The simpler model has slightly lower AIC, and the likelihood-ratio
comparison below provides no clear evidence that the more flexible
role-specific covariance improves fit. This is not evidence that partners are
interchangeable or that the two covariance structures are equivalent.

```{r}
compare_nested_models(
  m_w1,
  m_c
)
```


```{r}
m_w2 <- glmmTMB::glmmTMB(
    closeness ~
        0 +
        .is_female +
        .is_male +
        .is_female:.provided_support_gmc_actor +
        .is_male:.provided_support_gmc_actor +
        .is_female:.provided_support_gmc_partner +
        .is_male:.provided_support_gmc_partner +
        (1 | coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
m_w2_simulations <- simulate_dyad_responses(
  m_w2,
  nsim = 5000,
  seed = 13003L
)

# Model-centred responses
m_w2_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "model-centred"
  )
```


```{r}
# Raw responses
m_w2_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "raw"
  )
```


```{r}
m_w3 <- glmmTMB::glmmTMB(
    closeness ~
        0 +
        .is_female +
        .is_male +
        .is_female:.provided_support_gmc_actor +
        .is_male:.provided_support_gmc_actor +
        .is_female:.provided_support_gmc_partner +
        .is_male:.provided_support_gmc_partner +
        (0 + .is_female + .is_male || coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
m_w3_simulations <- simulate_dyad_responses(
  m_w3,
  nsim = 5000,
  seed = 13004L
)

# Model-centred responses
m_w3_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "model-centred"
  )
```


```{r}
# Raw responses
m_w3_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "raw"
  )
```

# Sensitivity analysis: treating partners as exchangeable

Here, the female-male data are deliberately treated as exchangeable. The
pooled check assesses the imposed covariance structure; the gender-aware check
asks whether that pooling hides role-specific differences.

These models have only a constant fixed-effect prediction. Their raw and
centred checks should therefore match; both are shown for completeness.

## Prep data

```{r}
diagnostic_data <- prepare_dyad_data(
    dyads_cross,
    dyad = coupleID,
    member = personID,
    role = gender,
    model_types = "none",
    keep_compositions = "female-male",
    set_exchangeable_compositions = "female-male",
    seed = 13000L
)
```


## Exchangeable mean/difference model

```{r}
e_c <- glmmTMB::glmmTMB(
    closeness ~
        1 +

      (1 | coupleID) +
      (0 + .member_contrast_arbitrary | coupleID),

    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
e_c_simulations <- simulate_dyad_responses(
  e_c,
  nsim = 5000,
  seed = 13005L
)

# Model-centred responses
e_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "model-centred"
  )
```


```{r}
# Raw responses
e_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "raw"
  )
```

## Check the same model with gender labels

```{r}
# Model-centred responses
e_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "model-centred"
  )
```


```{r}
# Raw responses
e_c_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    role = diagnostic_data$gender,
    response = "raw"
  )
```

The pooled check can look adequate while the gender-aware check reveals omitted
role differences. However, this sensitivity model also removes role-specific
fixed effects, so a discrepancy cannot be attributed to the covariance
structure alone.

## More restrictive exchangeable models

```{r}
e_w1 <- glmmTMB::glmmTMB(
    closeness ~
        1,
    dispformula = ~ 1,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
e_w1_simulations <- simulate_dyad_responses(
  e_w1,
  nsim = 5000,
  seed = 13006L
)

# Model-centred responses
e_w1_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "model-centred"
  )
```


```{r}
# Raw responses
e_w1_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "raw"
  )
```

```{r}
e_w2 <- glmmTMB::glmmTMB(
    closeness ~
        1 + (1 | coupleID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
e_w2_simulations <- simulate_dyad_responses(
  e_w2,
  nsim = 5000,
  seed = 13007L
)

# Model-centred responses
e_w2_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "model-centred"
  )
```


```{r}
# Raw responses
e_w2_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "raw"
  )
```


```{r}
e_w3 <- glmmTMB::glmmTMB(
    closeness ~
        1 + (1 | coupleID/personID),
    dispformula = ~ 0,
    family = gaussian(),
    data = diagnostic_data
)
```

```{r}
e_w3_simulations <- simulate_dyad_responses(
  e_w3,
  nsim = 5000,
  seed = 13008L
)

# Model-centred responses
e_w3_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "model-centred"
  )
```


```{r}
# Raw responses
e_w3_simulations |>
  check_partner_dependence(
    dyad = diagnostic_data$coupleID,
    response = "raw"
  )
```

# Development validation

The [repeated simulation study](partner-dependence-outer-simulation-study.Rmd)
gave encouraging initial results. For the deliberately large problems studied,
the observed pattern stood out from the model's simulations in 84--97% of
datasets with 50 dyads and in every dataset with 200 dyads. All 7,000
fitted-model runs completed successfully. Flexible models that had already
learned the same pattern nearly always looked reassuring; smaller problems
have not yet been studied.

Non-Gaussian models and further diagnostics are future work. See the
[technical and development guide](README.md#technical-details) for simulation
details, object shapes, validation reports, and the follow-up plan.
