---
title: "Tweedie outcomes checked with a Gaussian model"
output: html_document
---

```{r tweedie-setup, include=FALSE}
pkgload::load_all(file.path("..", ".."), quiet = TRUE)
knitr::opts_chunk$set(
  message = FALSE, dev = "svg", fig.width = 12, fig.height = 14,
  out.width = "100%",
  fig.path = "distribution-diagnostics/results/tweedie-gaussian/"
)
```

## A model can reproduce correlation but miss other features

These simulated Tweedie outcomes are nonnegative and skewed, with some zeros
and different variability for women and men. We deliberately fit an exchangeable
Gaussian model: the same effects and variance for both roles. Its shared dyad
intercept allows positive partner correlation; there is no member-difference
random effect, but observation-level variance is still estimated.

Every code chunk below runs when this document is rendered. Each check is
followed by all the complete panels it produces, in their original order.

### Fit the model and simulate once

Run from `dev/diagnostic_checks/` to read the saved dataset of 120 female-male
dyads. All checks reuse the same 2,000 simulated datasets.

```{r tweedie-fit}
library(dyadMLM)
tweedie_data <- read.csv(
  "distribution-diagnostics/results/tweedie-gaussian/data.csv",
  colClasses = c(role = "factor")
)
tweedie_data$dyad <- factor(tweedie_data$dyad)
gaussian_model <- glmmTMB::glmmTMB(
  outcome ~ predictor + (1 | dyad),
  family = gaussian(), dispformula = ~1, data = tweedie_data
)
gaussian_simulations <- simulate_dyad_responses(
  gaussian_model, nsim = 2000, seed = 100105
)
```

Supply the fitting data because `role` is absent from the model frame. Roles
split the plots without changing the fitted model's assumptions.

### Check residuals

The default panel contains all six residual checks. `predictors` adds a page
for the named predictor, giving two complete panels here.

```{r residual, fig.height=20, fig.alt="Complete panels produced by check_residuals"}
check_residuals(
  gaussian_simulations, dyad = dyad, role = role, data = tweedie_data,
  predictors = "predictor",
  seed = 123, ask = FALSE
)
```

The QQ curves and PIT histogram reveal a distribution mismatch. Five male
outcomes fall outside all their reference simulations, compared with a usual
simulated count of zero to one. *Predicted outcome* means the model prediction
with random effects set to zero. The separate predictor page uses the supplied
predictor values. Red quartiles should roughly follow the horizontal lines at
0.25, 0.50 and 0.75, usually within their matching blue bands.

### Check outcomes

`check_zeros = TRUE` includes zero counts. The call produces one complete panel:
the raw outcome distribution, variability and largest deviations after
subtracting predictions, and zero counts.

```{r outcome, fig.alt="Complete panels produced by check_outcomes"}
check_outcomes(
  gaussian_simulations, dyad = dyad, role = role, data = tweedie_data,
  check_zeros = TRUE, ask = FALSE
)
```

The Gaussian model generates negative values, misses the male zeros, and
understates the largest deviations from predictions. Its common variance is
too large for female responses and too small for male responses.

### Check partner dependence

The default check subtracts the same predictions from observed and simulated
outcomes. It produces one complete panel for this composition.

```{r partner, fig.height=8, fig.alt="Complete panel produced by check_partner_dependence"}
check_partner_dependence(
  gaussian_simulations, dyad = dyad, role = role,
  data = tweedie_data, ask = FALSE
)
```

The role SDs and mean–difference correlation reveal unequal variability, while
the partner correlation agrees reasonably well. Omitting the difference random
effect does not by itself force a mismatch: the shared intercept and
observation-level noise can reproduce positive partner correlation.

For comparison, request the same panel for raw outcomes:

```{r partner-raw, fig.height=8, fig.alt="Complete raw-outcome panel produced by check_partner_dependence"}
check_partner_dependence(
  gaussian_simulations, dyad = dyad, role = role, data = tweedie_data,
  response = "raw", ask = FALSE
)
```

This includes variation explained by the predictor; the role-variance mismatch
remains visible. Several assumptions are wrong here, so a failed check does not
identify a single cause. Blue ranges describe the middle 95% at each comparison,
not across all plots. These are descriptive checks, not significance tests.
