---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# interdep

<!-- badges: start -->
<!-- badges: end -->

`interdep` helps researchers prepare cross-sectional and intensive longitudinal
dyadic data for statistical models.

It supports dyadic datasets with distinguishable dyads, exchangeable dyads, or
mixed dyad compositions in the same data, such as female-male, female-female,
and male-male couples. Mixed dyad compositions are supported for
Actor-Partner Interdependence Model (APIM) preparation. Current
Dyad-Individual Model (DIM) and undirected Dyadic Score Model (DSM) helpers
require one exchangeable dyad composition.

Start with the [Getting Started](vignettes/getting-started.Rmd) vignette for
the main data-preparation workflow. The package documentation is organized
around the main modeling tasks:

| Vignette | Focus |
|---|---|
| [Getting Started](vignettes/getting-started.Rmd) | Data structure, validation, dyad compositions, generated columns, and basic preparation |
| APIM | Cross-sectional APIM preparation for distinguishable, exchangeable, and mixed dyad compositions |
| Intensive Longitudinal APIM | Temporal predictor decomposition and intensive longitudinal APIM preparation |
| [Dyad-Individual Model](vignettes/Dyad-Individual-Model.Rmd) | DIM predictor construction and APIM-DIM equivalence for exchangeable dyads |
| Undirected Dyadic Score Model | Undirected DSM outcome and predictor construction |


## Installation

You can install the development version of `interdep` from GitHub with:

``` r
# install.packages("pak")
pak::pak("Pascal-Kueng/interdep")
```

## Simple Cross-Sectional Example

Prepare distinguishable dyads for a cross-sectional APIM:

```{r cross-sectional-prep}
library(interdep)

prepared_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  model_type = "apim"
)

print(prepared_data)
```

The prepared data contains the composition indicators and APIM actor/partner
predictor columns used in the model formulas below.

One possible APIM formula is:

```{r cross-sectional-apim, eval = FALSE}
simple_apim <- glmmTMB::glmmTMB(
  satisfaction ~ 

    # Gender-specific intercepts
    0 + .i_is_female_x_male_female + .i_is_female_x_male_male +

    # Gender-specific actor effects
    .i_communication_raw_actor:.i_is_female_x_male_female +
    .i_communication_raw_actor:.i_is_female_x_male_male +

    # Gender-specific partner effects
    .i_communication_raw_partner:.i_is_female_x_male_female +
    .i_communication_raw_partner:.i_is_female_x_male_male +

    # Dyad-level random effects represent the partner residual
    # variance-covariance structure
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID),

  # With the residual variance-covariance represented by the dyad-level
  # random effects above, the Gaussian residual dispersion is fixed near zero.
  dispformula = ~ 0,
  family = gaussian(),
  data = prepared_data
)
```


## Citation

If you use `interdep`, please cite the package directly:

```bibtex
@Manual{interdep,
  title = {interdep: Tools for Cross-Sectional and Intensive Longitudinal Dyadic Data},
  author = {Pascal Küng},
  year = {2026},
  note = {R package version 0.0.0.9000},
  url = {https://github.com/Pascal-Kueng/interdep},
}
```

An archival DOI will be added once the package has a public release archive.
