---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# interdep

<!-- badges: start -->
<!-- badges: end -->

`interdep` is being developed for dyadic data analysis. It currently provides a
basic preparation function for validating long-format dyadic data and deriving
dyad-composition metadata.

## Installation

You can install the development version of `interdep` from GitHub with:

``` r
# install.packages("pak")
pak::pak("Pascal-Kueng/interdep")
```

## Example

```{r example}
library(interdep)

prepared <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender
)

attr(prepared, "interdep")$dyad_compositions
```

Use `incomplete_dyads` and `missing_role` to choose whether incomplete dyads or
unresolved role information should error, be dropped, or be kept with unknown
composition labels. See the vignette for examples.
