---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# interdep

<!-- badges: start -->
<!-- badges: end -->

`interdep` helps researchers prepare cross-sectional and intensive longitudinal
dyadic data for statistical models.

It supports common dyadic studies with one kind of dyad, and it also handles
studies where different kinds of dyads appear in the same dataset, such as
female-male, female-female, and male-male couples.

The vignette "Preparing dyadic data with interdep" demonstrates how prepared
data can be used to estimate dyadic models, including generalized intensive
longitudinal APIMs with multiple dyad types in a single model.

## Citation

If you use `interdep`, please cite the package directly:

```bibtex
@Manual{interdep,
  title = {interdep: Tools for Cross-Sectional and Intensive Longitudinal Dyadic Data},
  author = {Pascal Küng},
  year = {2026},
  note = {R package version 0.0.0.9000},
  url = {https://github.com/Pascal-Kueng/interdep},
}
```

An archival DOI will be added once the package has a public release archive.

## Installation

You can install the development version of `interdep` from GitHub with:

``` r
# install.packages("pak")
pak::pak("Pascal-Kueng/interdep")
```

## Example

```{r example}
library(interdep)

prepared <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender
)

prepared
```

The prepared data includes `.i_composition_role`, formula-friendly `.i_is_*`
indicators, and `.i_diff` / `.i_diff_*` columns for exchangeable dyads.

Use `incomplete_dyads` and `missing_role` to choose whether incomplete dyads or
unresolved role information should error or be dropped. See the vignette for
examples.
