---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dyadMLM

<!-- badges: start -->
[![R-CMD-check](https://github.com/Pascal-Kueng/dyadMLM/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Pascal-Kueng/dyadMLM/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/Pascal-Kueng/dyadMLM/graph/badge.svg)](https://app.codecov.io/gh/Pascal-Kueng/dyadMLM)
<!-- badges: end -->

`dyadMLM` provides tools for dyadic multilevel modeling with linear and
generalized linear mixed-effects models. It validates and prepares
cross-sectional and intensive longitudinal dyadic data.

It supports distinguishable and exchangeable dyads and can also prepare
datasets containing multiple dyad compositions. It creates composition-aware,
model-ready columns for dyadic multilevel model parameterizations such as the
Actor-Partner Interdependence Model (APIM), Dyad-Individual Model (DIM), and
Dyadic Score Model (DSM).

Selected post-estimation tools compare compatible fitted models and
back-transform exchangeable random-effect covariance structures into
member-level quantities.

Start with the vignettes, or scroll down for a quick-start.

| Vignette | Focus |
|---|---|
| [Getting Started](https://pascal-kueng.github.io/dyadMLM/articles/getting-started.html) | Data structure, validation, dyad compositions, generated columns, and basic preparation |
| [Actor-Partner Interdependence Model](https://pascal-kueng.github.io/dyadMLM/articles/apim.html) | APIM preparation and formulas for distinguishable and exchangeable dyads in cross-sectional and intensive longitudinal data |
| [Dyad-Individual Model](https://pascal-kueng.github.io/dyadMLM/articles/dim.html) | DIM predictor construction, formulas, and an interactive demonstration of APIM-DIM equivalence for exchangeable dyads |
| [Dyadic Score Model](https://pascal-kueng.github.io/dyadMLM/articles/dsm.html) | DSM predictor-score and contrast construction, formulas, and the relationship between the DSM and APIM for distinguishable dyads |

For an in-depth tutorial covering data preparation, model fitting, diagnostics,
and assumption checks, see
[Distinguishable and Exchangeable Dyads: Bayesian Multilevel Modelling](https://pascal-kueng.github.io/05DyadicDataAnalysis/).
It uses `dyadMLM` for cross-sectional and intensive longitudinal APIM and DIM
workflows, with models fitted primarily using `brms`
([source](https://github.com/Pascal-Kueng/05DyadicDataAnalysis),
[DOI](https://doi.org/10.5281/zenodo.17400655)).


## Installation

You can install the released version of `dyadMLM` from CRAN with:

``` r
install.packages("dyadMLM")
```

You can install the development version from GitHub with:

``` r
# install.packages("pak")
pak::pak("Pascal-Kueng/dyadMLM")
```

## Simple Cross-Sectional Example

Prepare distinguishable dyads for a cross-sectional APIM:

```{r cross-sectional-prep}
library(dyadMLM)

prepared_data <- prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  predictors = provided_support,
  model_types = "apim",
  # All three observed compositions in `dyads_cross` are detected and retained by
  # default. This example focuses on `female-male` dyads, so we restrict the
  # analysis here.
  keep_compositions = "female-male"
)

print(prepared_data, n = 4)
```

The prepared data contains the composition indicators and APIM actor/partner
predictor columns used in the model formulas below.

One simple distinguishable APIM formula is:

```{r cross-sectional-apim, eval = FALSE}
simple_apim <- glmmTMB::glmmTMB(
  closeness ~

    # Gender-specific intercepts
    0 + .dy_is_female_x_male_female + .dy_is_female_x_male_male +

    # Gender-specific actor effects
    .dy_provided_support_actor:.dy_is_female_x_male_female +
    .dy_provided_support_actor:.dy_is_female_x_male_male +

    # Gender-specific partner effects
    .dy_provided_support_partner:.dy_is_female_x_male_female +
    .dy_provided_support_partner:.dy_is_female_x_male_male +

    # Dyad-level random effects represent the two members'
    # residual covariance structure
    us(0 + .dy_is_female_x_male_female + .dy_is_female_x_male_male | coupleID),

  # With the residual covariance represented by the dyad-level
  # random effects above, the Gaussian residual dispersion is fixed near zero.
  dispformula = ~ 0,
  family = gaussian(),
  data = prepared_data
)
```

## Citation

If you use `dyadMLM`, please cite the package directly:

```bibtex
@Manual{dyadMLM,
  title = {dyadMLM: Tools for Dyadic Multilevel Models},
  author = {Pascal Küng},
  year = {2026},
  note = {R package version 0.1.0},
  doi = {10.5281/zenodo.21481720},
  url = {https://github.com/Pascal-Kueng/dyadMLM},
}
```

The package uses the concept DOI `10.5281/zenodo.21481720` across releases.

---

**Continue** with the [Getting Started Vignette](https://pascal-kueng.github.io/dyadMLM/articles/getting-started.html).

Or go directly to a model-specific vignette:

- [Actor-Partner Interdependence Model (APIM) vignette](https://pascal-kueng.github.io/dyadMLM/articles/apim.html),
- [Dyad-Individual Model vignette](https://pascal-kueng.github.io/dyadMLM/articles/dim.html), or
- [Dyadic Score Model vignette](https://pascal-kueng.github.io/dyadMLM/articles/dsm.html).
