---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# interdep

<!-- badges: start -->
[![R-CMD-check](https://github.com/Pascal-Kueng/interdep/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Pascal-Kueng/interdep/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`interdep` helps researchers prepare cross-sectional and intensive longitudinal
dyadic data for multilevel models, including generalized multilevel models.

It supports dyadic datasets with distinguishable dyads, exchangeable dyads, or
mixed dyad types in the same data, such as female-male, female-female,
and male-male couples. It creates composition-aware, model-ready columns for
dyadic multilevel model parameterizations such as the Actor-Partner
Interdependence Model (APIM), Dyad-Individual Model (DIM), and undirected
Dyadic Score Model (DSM). Current DIM and undirected DSM helpers require one
exchangeable dyad composition.

Start with the vignettes, or scroll down for a quick-start.

| Vignette | Focus |
|---|---|
| [Getting Started](https://pascal-kueng.github.io/interdep/articles/getting-started.html) | Data structure, validation, dyad compositions, generated columns, and basic preparation |
| [Actor-Partner Interdependence Model](https://pascal-kueng.github.io/interdep/articles/apim.html) | APIM preparation and formulas for distinguishable, exchangeable, generalized, and intensive longitudinal dyads |
| [APIMs with Mixed Dyad Compositions](https://pascal-kueng.github.io/interdep/articles/mixed-apim.html) | APIMs that combine distinguishable and exchangeable dyad compositions in one analysis |
| [Dyad-Individual Model](https://pascal-kueng.github.io/interdep/articles/dim.html) | DIM predictor construction and formulas and interactive demonstration of APIM-DIM equivalence for exchangeable dyads |
| [Undirected Dyadic Score Model](https://pascal-kueng.github.io/interdep/articles/undirected-dsm.html) | Undirected DSM outcome and predictor construction and formulas |

For an in-depth tutorial covering data preparation, model fitting, diagnostics,
and assumption checks, see
[Distinguishable and Exchangeable Dyads: Bayesian Multilevel Modelling](https://pascal-kueng.github.io/05DyadicDataAnalysis/).
It uses `interdep` for cross-sectional and intensive longitudinal APIM and DIM
workflows, with models fitted primarily using `brms`
([source](https://github.com/Pascal-Kueng/05DyadicDataAnalysis),
[DOI](https://doi.org/10.5281/zenodo.17400655)).

The tutorial does not cover models that combine distinguishable and exchangeable
dyads in a single analysis. For that workflow, see the
[Mixed-Composition APIM vignette](https://pascal-kueng.github.io/interdep/articles/mixed-apim.html).


## Installation

You can install the development version of `interdep` from GitHub with:

``` r
# install.packages("pak")
pak::pak("Pascal-Kueng/interdep")
```

## Simple Cross-Sectional Example

Prepare distinguishable dyads for a cross-sectional APIM:

```{r cross-sectional-prep}
library(interdep)

prepared_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  model_type = "apim"
)

print(prepared_data)
```

The prepared data contains the composition indicators and APIM actor/partner
predictor columns used in the model formulas below.

One possible APIM formula is:

```{r cross-sectional-apim, eval = FALSE}
simple_apim <- glmmTMB::glmmTMB(
  satisfaction ~ 

    # Gender-specific intercepts
    0 + .i_is_female_x_male_female + .i_is_female_x_male_male +

    # Gender-specific actor effects
    .i_communication_raw_actor:.i_is_female_x_male_female +
    .i_communication_raw_actor:.i_is_female_x_male_male +

    # Gender-specific partner effects
    .i_communication_raw_partner:.i_is_female_x_male_female +
    .i_communication_raw_partner:.i_is_female_x_male_male +

    # Dyad-level random effects represent the partner residual
    # variance-covariance structure
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID),

  # With the residual variance-covariance represented by the dyad-level
  # random effects above, the Gaussian residual dispersion is fixed near zero.
  dispformula = ~ 0,
  family = gaussian(),
  data = prepared_data
)
```


## Citation

If you use `interdep`, please cite the package directly:

```bibtex
@Manual{interdep,
  title = {interdep: Prepare Dyadic Data for Multilevel Models},
  author = {Pascal Küng},
  year = {2026},
  note = {R package version 0.0.0.9000},
  url = {https://github.com/Pascal-Kueng/interdep},
}
```

An archival DOI will be added once the package has a public release archive.
