---
title: "Partner-dependence checks across outcome families"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
params:
  results_directory: "results/family-comparison/500-datasets-1000-draws"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 13, fig.height = 6, dpi = 180, fig.retina = 1, out.width = "100%")
source("plot-family-comparison.R")
source("summarise-family-comparison.R")
results_directory <- params$results_directory
settings <- read.csv(file.path(results_directory, "settings.csv"))
saved_summary <- read.csv(file.path(results_directory, "summary.csv"))
calibration <- read.csv(file.path(results_directory, "calibration.csv"))
# Recount saved observations and limits, including results from an older running script.
summary <- lapply(settings$family, function(family_name) {
  statistics <- read.csv(file.path(results_directory, family_name, "statistics.csv"))
  completion <- saved_summary |>
    dplyr::filter(family == family_name) |>
    dplyr::select(family, condition, n_dyads, target_correlation, attempted,
      successful, with_warnings, check_errors) |>
    dplyr::distinct()
  summarise_family_comparison(statistics, completion)$summary
}) |> dplyr::bind_rows()
vignette_families <- c("gaussian", "nbinom2", "ordinal", "lognormal",
                      "zi_poisson", "hurdle_nbinom2")
# Refresh completed families from the default study while the remaining families run.
update_vignette_figures <-
  all(settings$datasets_per_condition == 500L & settings$reference_draws == 1000L) &&
  all(summary$attempted == 500L)
if (update_vignette_figures) dir.create("../figures", showWarnings = FALSE)
```

## What these plots compare

```{r progress, results='asis'}
if (nrow(settings) < 22L) {
  cat("**Study in progress:** Results for ", nrow(settings),
      " of 22 family/model settings are available. The remaining results are pending.\n\n", sep = "")
}
```

The models include actor and partner effects but omit any remaining partner
dependence. We compare checks that subtract fixed-effect predictions with checks
of raw outcomes, using the **same datasets, fitted models, and reference simulations**.

- **Coloured lines:** how often the check detects omitted positive correlation.
  Detection means the observed correlation exceeds the upper limit of the model's
  middle 95% simulated range. Opposite-direction flags are listed separately.
- **Black lines:** false alarms when residual correlation is zero. Either tail
  counts. Solid and dotted lines distinguish the methods; the grey dashed 5% line is a
  benchmark, not a guaranteed rate.

Read both together: a method that detects more real mismatches may also raise more
false alarms. These results describe the settings studied, not general power for
each family. Bars and bands show uncertainty from using a finite number of study
datasets.

Ordinal checks use category scores 1, 2, and so on. Zero-inflated and hurdle checks
use the combined outcome, including zeros.

<details>
<summary>Study details</summary>

Predictors are standard normal with partner correlation 0.3. Actor and partner
effects are 0.5 and 0.3 on the model's link scale. Correlated normal draws are
transformed to each outcome distribution (a Gaussian copula), with dependence
adjusted to give residual correlations of 0, 0.10, 0.30, and 0.50 after subtracting
the true means given the predictors. Each family section reports its calibration method and checks.

Reference simulations keep fitted parameters and predictors fixed; models are not
refitted. Their number controls the precision of the 2.5th and 97.5th percentile
limits. The number of study datasets controls the precision of the plotted rates;
bars and bands are 95% Wilson intervals for this Monte Carlo uncertainty.

Rates use successful fits and defined correlations. Failed fits and checks are
excluded without retry, as are ordinal datasets missing a category. Undefined
simulated correlations are omitted from reference ranges. Reduced check counts
and exclusion reasons are listed below each family's plots.

The sample-size axis is linear through 500 dyads, with a shortened gap before
1,000. This report redraws saved results without fitting or simulating. See the
[study code](compare-centering.R), [plot code](plot-family-comparison.R), and
[study notes](README.md).

</details>

```{r family-results, results='asis'}
for (family_index in seq_len(nrow(settings))) {
  family_settings <- settings[family_index, ]
  family_summary <- summary[summary$family == family_settings$family, ]
  family_calibration <- calibration[calibration$family == family_settings$family, ]
  cat("\n\n## ", family_settings$label, "\n\n", sep = "")
  cat(family_settings$datasets_per_condition, " generated datasets per condition; ",
      format(family_settings$reference_draws, big.mark = ","),
      " reference simulations per fitted model.\n\n", sep = "")
  figure <- plot_family_comparison(family_summary, family_settings$label,
    family_settings$datasets_per_condition, family_settings$reference_draws)
  print(figure)
  if (update_vignette_figures && family_settings$family %in% vignette_families) {
    ggplot2::ggsave(file.path("../figures", paste0("family-comparison-", family_settings$family, ".png")),
      figure, width = 13, height = 6, dpi = 180)
  }
  print(plot_family_false_alarms(family_summary, family_settings$label,
    family_settings$datasets_per_condition, family_settings$reference_draws))

  cat("<details>\n<summary>Model settings and checks</summary>\n\n")
  cat("**Fitted model:** `", family_settings$formula, "`.\n\n", sep = "")
  cat("**Mean:** ", family_settings$mean_settings, "\n\n", sep = "")
  cat("**Dispersion:** ", family_settings$dispersion_settings, "\n\n", sep = "")
  if (!is.na(family_settings$zero_settings) && nzchar(family_settings$zero_settings)) {
    cat("**Zero component:** ", family_settings$zero_settings, "\n\n", sep = "")
  }

  cat("**Population correlations**\n\n")
  cat(paste(unique(family_calibration$method), collapse = "; "), ".\n\n", sep = "")
  print(knitr::kable(family_calibration[c("target_correlation", "measured_residual_correlation", "verification_mcse")],
    col.names = c("Target residual correlation", "Verified correlation", "Monte Carlo SE"),
    digits = 3, row.names = FALSE))

  cat("\n\n**Opposite-direction flags with positive residual correlation**\n\n")
  opposite_direction <- family_summary |>
    dplyr::filter(target_correlation > 0) |>
    dplyr::group_by(response) |>
    dplyr::summarise(checked = sum(checked),
      opposite_direction = sum(opposite_direction), .groups = "drop")
  print(knitr::kable(opposite_direction,
    col.names = c("Method", "Datasets checked", "Opposite-direction flags")))

  cat("\n\n**Either-tail flags with zero residual correlation**\n\n")
  baseline <- family_summary[family_summary$target_correlation == 0,
    c("n_dyads", "response", "checked", "frequency")]
  baseline$frequency <- ifelse(is.finite(baseline$frequency),
    sprintf("%.1f%%", 100 * baseline$frequency), "Not checked")
  print(knitr::kable(baseline,
    col.names = c("Dyads", "Method", "Datasets checked", "Flagged")))

  # Fit counts repeat for the two methods; count them only once.
  fit_counts <- family_summary[family_summary$response == "model-centred", ]
  cat("\n\n", sum(fit_counts$successful), " of ", sum(fit_counts$attempted),
      " datasets produced reference simulations.\n\n", sep = "")
  excluded_fits <- read.csv(file.path(results_directory, family_settings$family, "fits.csv")) |>
    dplyr::filter(status != "success") |>
    dplyr::count(status, name = "Datasets")
  if (nrow(excluded_fits)) {
    excluded_fits$status <- c(fit_error = "Model fitting failed",
      fit_problem = "Convergence or Hessian problem", check_error = "Check failed",
      study_exclusion = "Ordinal category absent")[excluded_fits$status]
    print(knitr::kable(excluded_fits, col.names = c("Exclusion reason", "Datasets")))
  }
  reduced_counts <- family_summary |>
    dplyr::filter(checked < attempted) |>
    dplyr::select(n_dyads, target_correlation, response, attempted, checked)
  if (nrow(reduced_counts)) {
    cat("\n\n**Conditions with fewer usable checks**\n\n")
    print(knitr::kable(reduced_counts,
      col.names = c("Dyads", "Residual correlation", "Method", "Generated", "Checked")))
  }
  cat("\n\nCounts for every plotted point are saved in `summary.csv`.\n\n</details>\n\n")
}
```
