---
title: "Explaining Interdependence in the Actor–Partner Interdependence Model"
subtitle: "A visual paper outline and covariance-decomposition prototype"
author: "Pascal Küng"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_caption: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  dpi = 144
)

helper_candidates <- c(
  "../vignettes/diagram-helpers.Rinc",
  "vignettes/diagram-helpers.Rinc",
  "../../vignettes/diagram-helpers.Rinc"
)
helper_file <- helper_candidates[file.exists(helper_candidates)][1]

if (!is.na(helper_file)) {
  sys.source(helper_file, envir = knitr::knit_global())
  diagram_device <- .vignette_diagram_device()
  effect_colours <- .interdep_effect_colors(c("actor", "partner"))
} else {
  diagram_device <- "png"
  effect_colours <- c(actor = "#1769AA", partner = "#7B3FA1")
}

knitr::opts_chunk$set(dev = diagram_device, fig.ext = diagram_device)

colours <- c(
  actor = unname(effect_colours[["actor"]]),
  partner = unname(effect_colours[["partner"]]),
  mixed_1 = "#3F5EAE",
  mixed_2 = "#6250A7",
  residual = "#546E7A",
  covariance = "#334155",
  ink = "#263238",
  muted = "#64748B",
  faint = "#DCE3EA",
  surface = "#F8FAFC",
  white = "#FFFFFF"
)

fade <- function(colour, alpha = 0.22) {
  grDevices::adjustcolor(colour, alpha.f = alpha)
}

.paper_math_label <- function(label) {
  symbols <- c("X", "Y", "a", "p", "sigma", "psi", "epsilon")

  format_node <- function(node) {
    if (is.name(node) && as.character(node) %in% symbols) {
      return(call("italic", node))
    }
    if (!is.call(node)) {
      return(node)
    }

    head <- as.character(node[[1L]])
    if (head %in% c("italic", "plain")) {
      return(node)
    }

    as.call(lapply(node, format_node))
  }

  if (is.expression(label)) {
    return(as.expression(lapply(as.list(label), format_node)))
  }
  if (is.language(label)) {
    return(format_node(label))
  }
  label
}

draw_box <- function(x, y, label, width = 0.18, height = 0.105,
                     fill = colours[["white"]], border = colours[["ink"]],
                     cex = 1.2, lwd = 2) {
  graphics::rect(
    x - width / 2, y - height / 2,
    x + width / 2, y + height / 2,
    col = fill, border = border, lwd = lwd
  )
  graphics::text(
    x, y, .paper_math_label(label), cex = cex, col = colours[["ink"]]
  )
}

draw_circle <- function(x, y, label, radius = 0.043,
                        fill = colours[["white"]],
                        border = colours[["ink"]], cex = 1.05, lwd = 2) {
  theta <- seq(0, 2 * pi, length.out = 100)
  graphics::polygon(
    x + radius * cos(theta), y + radius * sin(theta),
    col = fill, border = border, lwd = lwd
  )
  graphics::text(
    x, y, .paper_math_label(label), cex = cex, col = colours[["ink"]]
  )
}

draw_path <- function(x0, y0, x1, y1, label = NULL,
                      colour = colours[["faint"]], lwd = 2.3,
                      label_x = mean(c(x0, x1)),
                      label_y = mean(c(y0, y1)),
                      label_cex = 0.9) {
  graphics::arrows(
    x0, y0, x1, y1,
    length = 0.075, angle = 18, lwd = lwd, col = colour
  )
  if (!is.null(label)) {
    graphics::text(
      label_x, label_y, .paper_math_label(label),
      cex = label_cex, col = colour, font = 2
    )
  }
}

draw_covariance <- function(x0, y0, x1, y1, label = NULL,
                            colour = colours[["faint"]], lwd = 2.1,
                            label_y = y0 + 0.045, label_cex = 0.82) {
  graphics::arrows(
    x0, y0, x1, y1,
    code = 3, length = 0.06, angle = 18, lwd = lwd, col = colour
  )
  if (!is.null(label)) {
    graphics::text(
      mean(c(x0, x1)), label_y, .paper_math_label(label),
      cex = label_cex, col = colour
    )
  }
}

component_state <- function(component) {
  all_on <- identical(component, "TOTAL")
  c(
    actor_1 = all_on || component %in% c("AA", "AP"),
    actor_2 = all_on || component %in% c("AA", "PA"),
    partner_1 = all_on || component %in% c("PA", "PP"),
    partner_2 = all_on || component %in% c("AP", "PP"),
    predictor_covariance = all_on || component %in% c("AA", "PP"),
    residual_covariance = all_on || component == "R"
  )
}

on_colour <- function(on, colour) {
  if (isTRUE(on)) colour else colours[["faint"]]
}

draw_apim_paths <- function(component = "TOTAL", panel_title = NULL,
                            show_formula = FALSE) {
  state <- component_state(component)

  graphics::plot.new()
  graphics::plot.window(xlim = c(0, 1), ylim = c(0, 1), asp = 1)
  graphics::rect(0.01, 0.01, 0.99, 0.99,
                 col = colours[["surface"]], border = NA)

  # Predictor covariance and the four directed APIM paths.
  draw_covariance(
    0.29, 0.84, 0.71, 0.84,
    expression(sigma[X[12]]),
    on_colour(state[["predictor_covariance"]], colours[["covariance"]]),
    label_y = 0.905
  )

  draw_path(
    0.20, 0.735, 0.20, 0.365, expression(a[1]),
    on_colour(state[["actor_1"]], colours[["actor"]]),
    label_x = 0.155, label_y = 0.55
  )
  draw_path(
    0.80, 0.735, 0.80, 0.365, expression(a[2]),
    on_colour(state[["actor_2"]], colours[["actor"]]),
    label_x = 0.845, label_y = 0.55
  )
  draw_path(
    0.73, 0.755, 0.27, 0.345, expression(p[1]),
    on_colour(state[["partner_1"]], colours[["partner"]]),
    label_x = 0.57, label_y = 0.53
  )
  draw_path(
    0.27, 0.755, 0.73, 0.345, expression(p[2]),
    on_colour(state[["partner_2"]], colours[["partner"]]),
    label_x = 0.43, label_y = 0.43
  )

  # Residual paths and residual covariance.
  residual_node_colour <- on_colour(
    state[["residual_covariance"]], colours[["residual"]]
  )
  draw_path(0.20, 0.115, 0.20, 0.245, NULL, residual_node_colour, lwd = 2)
  draw_path(0.80, 0.115, 0.80, 0.245, NULL, residual_node_colour, lwd = 2)
  draw_covariance(
    0.245, 0.09, 0.755, 0.09, expression(psi[12]),
    residual_node_colour, label_y = 0.035
  )

  # Nodes are drawn last, so crossing paths disappear behind them.
  draw_box(0.20, 0.79, expression(X[1]))
  draw_box(0.80, 0.79, expression(X[2]))
  draw_box(0.20, 0.31, expression(Y[1]))
  draw_box(0.80, 0.31, expression(Y[2]))
  draw_circle(0.20, 0.09, expression(epsilon[1]), border = residual_node_colour)
  draw_circle(0.80, 0.09, expression(epsilon[2]), border = residual_node_colour)

  if (!is.null(panel_title)) {
    graphics::mtext(panel_title, side = 3, line = -1.2,
                    cex = 1.05, font = 2, col = colours[["ink"]])
  }

  if (isTRUE(show_formula)) {
    formulas <- c(
      AA = "a1 a2 Cov(X1, X2)",
      AP = "a1 p2 Var(X1)",
      PA = "p1 a2 Var(X2)",
      PP = "p1 p2 Cov(X1, X2)",
      R = "Cov(e1, e2)",
      TOTAL = "all five paths"
    )
    graphics::mtext(formulas[[component]], side = 1, line = -1.3,
                    cex = 0.82, col = colours[["muted"]])
  }
}

component_colours <- c(
  "Actor–actor" = colours[["actor"]],
  "Actor–partner" = colours[["mixed_1"]],
  "Partner–actor" = colours[["mixed_2"]],
  "Partner–partner" = colours[["partner"]],
  "Residual" = colours[["residual"]]
)
```

> **Working claim.** An APIM should not only *account for* interdependence. It
> can also explain where the covariance between partners' outcomes comes from.

This document first develops the argument as it could appear in the paper and
checks the underlying path tracing visually and numerically. The final section
collects the proposed contribution, study plan, manuscript structure, open
questions, and next steps.

# The idea in one paragraph

The APIM is commonly introduced through actor effects, partner effects, and a
residual correlation. That presentation says how predictors relate to each
person's outcome, but it does not by itself answer a natural dyadic question:

> Why are the two partners' outcomes correlated, and how much of that
> interdependence is associated with each modeled route?

For one predictor construct measured on both partners, standard path tracing
gives an exact additive answer under the usual APIM assumption that predictors
are uncorrelated with the outcome residuals. Outcome covariance can travel through four
predictor-side path pairs: actor–actor, actor–partner, partner–actor, and
partner–partner. A fifth component is the remaining residual covariance. These
five signed quantities sum exactly to the outcome covariance implied by the
model. Dividing every component by the same two model-implied outcome standard
deviations makes them sum to the model-implied outcome correlation.

# The complete one-construct APIM

For distinguishable members 1 and 2, write

$$
Y_1=a_1X_1+p_1X_2+\varepsilon_1,
$$

$$
Y_2=a_2X_2+p_2X_1+\varepsilon_2.
$$

Intercepts are omitted because they do not affect variances or covariances.
The subscript on an actor or partner effect identifies the **outcome member**:
$p_1$ is the effect of member 2's predictor on member 1's outcome; $p_2$ is the
effect of member 1's predictor on member 2's outcome.

Unless stated otherwise, $a_1,a_2,p_1,p_2$ are unstandardized coefficients and
all predictor moments are computed on the matching original scales. A fully
standardized presentation is possible, but coefficients, predictor moments,
residual covariance, and the outcome denominator must then be transformed
consistently.

```{r complete-apim, echo=FALSE, fig.width=8.8, fig.height=6.2, out.width="82%", fig.cap="Complete distinguishable APIM used for the decomposition. Blue paths are actor effects, purple paths are partner effects, the upper double-headed arrow is predictor covariance, and the lower double-headed arrow is residual outcome covariance.", fig.alt="Path diagram with predictors X1 and X2, outcomes Y1 and Y2, blue actor paths, purple crossed partner paths, covariance between predictors, and covariance between outcome residuals."}
draw_apim_paths("TOTAL")
```

## Assumptions and what "exact" means

The proposed paper concerns independent dyads, continuous outcomes, and a
linear, fixed-slope, cross-sectional APIM. The variables are viewed as random
over a target population of dyads, so the result below decomposes **marginal
across-dyad covariance**. Intercepts are omitted and all variances and
covariances are taken around the appropriate role-specific means.

Assume finite second moments and

$$
\operatorname{E}(\boldsymbol\varepsilon\mid X_1,X_2)=\boldsymbol 0,
\qquad
\operatorname{Cov}(\boldsymbol\varepsilon\mid X_1,X_2)
=\boldsymbol\Psi,
$$

which implies that each predictor is uncorrelated with each outcome residual:

$$
\operatorname{Cov}(X_j,\varepsilon_k)=0
\qquad\text{for }j,k\in\{1,2\}.
$$

This is the usual exogeneity or residual-orthogonality assumption in the APIM.
It is part of the statistical model, not an identity that holds for any four
variables. Normality is not required for the covariance algebra. If
predictor–residual covariances are allowed, four additional terms enter the
outcome covariance and the five-part decomposition is incomplete.

Conditional on the observed predictors, the fixed-slope model implies
$\operatorname{Cov}(Y_1,Y_2\mid X_1,X_2)=\psi_{12}$. The predictor-side routes
appear when moving to marginal covariance across dyads whose predictor values
vary. Random slopes, nonlinear links, and longitudinal carryover paths create
additional terms and are outside the first paper's scope.

The derivation itself is short. By the linearity of covariance,

$$
\begin{aligned}
\operatorname{Cov}(Y_1,Y_2)
={}&\operatorname{Cov}\!\left(
a_1X_1+p_1X_2+\varepsilon_1,
a_2X_2+p_2X_1+\varepsilon_2
\right).
\end{aligned}
$$

Expanding this expression and applying the assumption above gives

$$
\begin{aligned}
\operatorname{Cov}(Y_1,Y_2)
={}&
\underbrace{a_1a_2\operatorname{Cov}(X_1,X_2)}_{\text{actor--actor}}
+\underbrace{a_1p_2\operatorname{Var}(X_1)}_{\text{actor--partner}}\\
&+\underbrace{p_1a_2\operatorname{Var}(X_2)}_{\text{partner--actor}}
+\underbrace{p_1p_2\operatorname{Cov}(X_1,X_2)}_{\text{partner--partner}}
+\underbrace{\operatorname{Cov}(\varepsilon_1,\varepsilon_2)}_{\text{residual}}.
\end{aligned}
$$

The four predictor–residual terms that disappear are
$a_1\operatorname{Cov}(X_1,\varepsilon_2)$,
$p_1\operatorname{Cov}(X_2,\varepsilon_2)$,
$a_2\operatorname{Cov}(X_2,\varepsilon_1)$, and
$p_2\operatorname{Cov}(X_1,\varepsilon_1)$. No more elaborate proof is needed:
the five remaining terms are precisely the five covariance-carrying routes in
the diagram.

# The five covariance-carrying routes

```{r path-components, echo=FALSE, fig.width=12, fig.height=8.1, out.width="100%", fig.cap="The five covariance-carrying routes in a distinguishable APIM with one predictor construct measured on both partners. Pale paths are present in the APIM but do not belong to the highlighted component. The final panel restores all paths.", fig.alt="Six small APIM path diagrams. Five panels separately highlight actor-actor, actor-partner, partner-actor, partner-partner, and residual covariance routes. A sixth panel shows their sum."}
old_par <- graphics::par(
  mfrow = c(2, 3),
  mar = c(0.8, 0.5, 1.5, 0.5),
  oma = c(0, 0, 0.4, 0),
  xpd = NA
)
draw_apim_paths("AA", "1. Actor–actor", show_formula = TRUE)
draw_apim_paths("AP", "2. Actor–partner", show_formula = TRUE)
draw_apim_paths("PA", "3. Partner–actor", show_formula = TRUE)
draw_apim_paths("PP", "4. Partner–partner", show_formula = TRUE)
draw_apim_paths("R", "5. Residual", show_formula = TRUE)
draw_apim_paths("TOTAL", "Their exact sum", show_formula = TRUE)
graphics::par(old_par)
```

The four predictor-side combinations form a useful $2\times2$ table:

| Path into $Y_1$ | Path into $Y_2$ | Shared predictor quantity | Contribution |
|:--|:--|:--|:--|
| actor $a_1$ | actor $a_2$ | $\operatorname{Cov}(X_1,X_2)$ | $C_{\mathrm{AA}}=a_1a_2\operatorname{Cov}(X_1,X_2)$ |
| actor $a_1$ | partner $p_2$ | $\operatorname{Var}(X_1)$ | $C_{\mathrm{AP}}=a_1p_2\operatorname{Var}(X_1)$ |
| partner $p_1$ | actor $a_2$ | $\operatorname{Var}(X_2)$ | $C_{\mathrm{PA}}=p_1a_2\operatorname{Var}(X_2)$ |
| partner $p_1$ | partner $p_2$ | $\operatorname{Cov}(X_1,X_2)$ | $C_{\mathrm{PP}}=p_1p_2\operatorname{Cov}(X_1,X_2)$ |

The residual component is

$$
C_{\mathrm{R}}=\operatorname{Cov}(\varepsilon_1,\varepsilon_2).
$$

Therefore,

$$
\operatorname{Cov}(Y_1,Y_2)
=C_{\mathrm{AA}}+C_{\mathrm{AP}}+C_{\mathrm{PA}}+C_{\mathrm{PP}}+C_{\mathrm{R}}.
$$

## Plain-language interpretation

- **Actor–actor:** partners resemble each other on $X$, and each person's own
  $X$ predicts their own $Y$.
- **Actor–partner:** member 1's $X$ reaches both outcomes—through an actor path
  into $Y_1$ and a partner path into $Y_2$.
- **Partner–actor:** member 2's $X$ reaches both outcomes in the analogous way.
- **Partner–partner:** partners resemble each other on $X$, and each predictor
  crosses to the other member's outcome.
- **Residual:** outcome interdependence that remains after the modeled
  predictor routes.

## Female-driven and male-driven interdependence

For a female--male dyad, let member 1 be the female partner and member 2 the
male partner. The two routes carried by a single member's predictor variance
then have an especially intuitive interpretation:

$$
\underbrace{a_1p_2\operatorname{Var}(X_1)}_{\text{female-driven}}
\qquad\text{and}\qquad
\underbrace{p_1a_2\operatorname{Var}(X_2)}_{\text{male-driven}}.
$$

The first is female-driven because the female partner's predictor reaches both
outcomes: through her actor path into her own outcome and through her partner
path into the male partner's outcome. The second is male-driven in the
corresponding way. Here, "driven" describes a model-implied covariance route;
it is causal only when the design and assumptions support a causal
interpretation.

This yields two broader substantive families:

$$
\underbrace{C_{\mathrm{AP}}+C_{\mathrm{PA}}}_{\text{predictor-transmission routes}}
\qquad\text{and}\qquad
\underbrace{C_{\mathrm{AA}}+C_{\mathrm{PP}}}_{\text{cross-member predictor-covariance routes}}.
$$

"Predictor transmission" is an algebraic route label, not by itself a claim
that one partner causally transmits something to the other. The second family
can be called **predictor similarity** when
$\operatorname{Cov}(X_1,X_2)>0$; a negative predictor covariance instead
represents dissimilarity.

The phrase "actor-driven covariance" by itself is potentially misleading. One
person's predictor variance creates covariance between the two outcomes only
when that predictor reaches **both** outcomes—once through an actor path and
once through a partner path. Actor and partner paths work in pairs.

# From covariance to correlation points

The five components above are in the original units of $Y_1$ times the units
of $Y_2$. That scale is exact but often difficult to interpret. To express each
component on the familiar correlation scale, define the model-implied
denominator

$$
s_Y=\operatorname{SD}(Y_1)\operatorname{SD}(Y_2).
$$

For every component $C_k$, define a correlation-point contribution

$$
c_k=\frac{C_k}{s_Y}.
$$

These are signed additive contributions on the outcome-correlation scale. An
individual $c_k$ is not itself a correlation, a proportion, or a variance
share, and it need not lie between $-1$ and $1$. Percentages are especially
misleading when positive and negative routes cancel: they can be negative,
exceed 100%, or become unstable when the total covariance is near zero.

Because every component uses the same denominator,

$$
\rho^{\mathrm{model}}_{Y_1Y_2}
=c_{\mathrm{AA}}+c_{\mathrm{AP}}+c_{\mathrm{PA}}+c_{\mathrm{PP}}+c_{\mathrm{R}}.
$$

The required total outcome variances are

$$
\begin{aligned}
\operatorname{Var}(Y_1)
={}&a_1^2\operatorname{Var}(X_1)
+p_1^2\operatorname{Var}(X_2)
+2a_1p_1\operatorname{Cov}(X_1,X_2)
+\operatorname{Var}(\varepsilon_1),\\
\operatorname{Var}(Y_2)
={}&a_2^2\operatorname{Var}(X_2)
+p_2^2\operatorname{Var}(X_1)
+2a_2p_2\operatorname{Cov}(X_1,X_2)
+\operatorname{Var}(\varepsilon_2).
\end{aligned}
$$

The observed sample correlation, $r_{Y_1Y_2}$, should be reported alongside
the model-implied correlation as a descriptive reference. They are conceptually
different quantities and are equal only when the fitted model reproduces the
relevant sample moments. The five components should therefore not be rescaled
afterward merely to force them to sum to the observed correlation.

> **Important distinction.** The residual correlation
> $\operatorname{Cor}(\varepsilon_1,\varepsilon_2)$ divides residual covariance
> by the two **residual** standard deviations. The residual contribution $c_{\mathrm{R}}$
> divides the same covariance by the two **total model-implied outcome** standard
> deviations. Only $c_{\mathrm{R}}$ is on the same scale as the other pieces of
> the model-implied outcome correlation.

# Worked numerical example

Suppose

$$
\operatorname{Var}(X_1)=\operatorname{Var}(X_2)=1,
\qquad
\operatorname{Cov}(X_1,X_2)=0.40,
$$

$$
a_1=0.50,\quad a_2=0.40,\quad p_1=0.20,\quad p_2=0.30,
$$

and

$$
\boldsymbol\Psi=
\begin{pmatrix}
0.50 & 0.15\\
0.15 & 0.55
\end{pmatrix}.
$$

```{r calculate-example, include=FALSE}
example_parameters <- list(
  var_x1 = 1,
  var_x2 = 1,
  cov_x12 = 0.40,
  a1 = 0.50,
  a2 = 0.40,
  p1 = 0.20,
  p2 = 0.30,
  psi11 = 0.50,
  psi22 = 0.55,
  psi12 = 0.15
)

with(example_parameters, {
  example_components <<- c(
    "Actor–actor" = a1 * a2 * cov_x12,
    "Actor–partner" = a1 * p2 * var_x1,
    "Partner–actor" = p1 * a2 * var_x2,
    "Partner–partner" = p1 * p2 * cov_x12,
    "Residual" = psi12
  )

  example_var_y1 <<-
    a1^2 * var_x1 + p1^2 * var_x2 +
    2 * a1 * p1 * cov_x12 + psi11
  example_var_y2 <<-
    a2^2 * var_x2 + p2^2 * var_x1 +
    2 * a2 * p2 * cov_x12 + psi22
})

example_cov_y12 <- sum(example_components)
example_denominator <- sqrt(example_var_y1 * example_var_y2)
example_correlation_components <- example_components / example_denominator
example_rho_y12 <- sum(example_correlation_components)
example_residual_rho <- with(
  example_parameters,
  psi12 / sqrt(psi11 * psi22)
)

example_table <- data.frame(
  Route = names(example_components),
  `Covariance contribution` = unname(example_components),
  `Correlation-point contribution` = unname(example_correlation_components),
  check.names = FALSE
)
```

```{r example-table, echo=FALSE}
knitr::kable(
  example_table,
  digits = 3,
  align = c("l", "r", "r")
)
```

```{r example-contributions, echo=FALSE, fig.width=9.4, fig.height=5.4, out.width="92%", fig.cap="Waterfall decomposition of the model-implied outcome correlation. Each route starts where the preceding route ended; the five signed increments arrive exactly at the model-implied total.", fig.alt="Waterfall chart in which actor-actor, actor-partner, partner-actor, partner-partner, and residual correlation-point contributions accumulate to the model-implied outcome correlation."}
old_par <- graphics::par(mar = c(7.0, 4.8, 1.2, 1.0), xpd = NA)
increments <- unname(example_correlation_components)
starts <- c(0, head(cumsum(increments), -1))
ends <- cumsum(increments)
x <- seq_along(increments)

graphics::plot(
  NA,
  xlim = c(0.45, length(increments) + 0.75),
  ylim = c(0, example_rho_y12 * 1.16),
  xaxt = "n",
  xlab = "",
  ylab = "Outcome-correlation points",
  bty = "l"
)

for (index in seq_along(increments)) {
  graphics::rect(
    x[index] - 0.34,
    min(starts[index], ends[index]),
    x[index] + 0.34,
    max(starts[index], ends[index]),
    col = component_colours[index],
    border = NA
  )
  if (index < length(increments)) {
    graphics::segments(
      x[index] + 0.34, ends[index],
      x[index + 1] - 0.34, ends[index],
      col = colours[["muted"]], lty = 3, lwd = 1.4
    )
  }
  graphics::text(
    x[index], ends[index] + example_rho_y12 * 0.025,
    sprintf("%+.3f", increments[index]),
    cex = 0.82, col = colours[["ink"]]
  )
}

graphics::axis(
  1,
  at = x,
  labels = names(example_correlation_components),
  las = 2,
  tick = FALSE,
  line = -0.3,
  cex.axis = 0.85
)
graphics::abline(h = example_rho_y12, lty = 2, lwd = 2.1,
                 col = colours[["covariance"]])
graphics::text(
  0.55, example_rho_y12 + example_rho_y12 * 0.025,
  sprintf("total ρ = %.3f", example_rho_y12),
  pos = 4, offset = 0.25, cex = 0.92,
  col = colours[["covariance"]], font = 2
)
graphics::par(old_par)
```

Here,

$$
\operatorname{Cov}(Y_1,Y_2)=`r sprintf('%.3f', example_cov_y12)`,
\qquad
\rho^{\mathrm{model}}_{Y_1Y_2}=`r sprintf('%.3f', example_rho_y12)`.
$$

The residual correlation is
`r sprintf('%.3f', example_residual_rho)`, but its contribution to the total
outcome correlation is only
`r sprintf('%.3f', unname(example_correlation_components[['Residual']]))`.
These numbers answer different questions.

# Same interdependence, different route profiles

The decomposition is useful because the same total partner correlation can
hide very different model-implied route profiles. The second hypothetical
model below is chosen to have the **same** outcome variances and covariance as the first model, but
stronger actor paths, weaker partner paths, and more residual covariance.

```{r same-total-calculation, include=FALSE}
scenario_a <- example_components

scenario_b_parameters <- list(
  var_x1 = 1,
  var_x2 = 1,
  cov_x12 = 0.40,
  a1 = 0.70,
  a2 = 0.60,
  p1 = 0.05,
  p2 = 0.05,
  psi12 = 0.25
)

with(scenario_b_parameters, {
  scenario_b <<- c(
    "Actor–actor" = a1 * a2 * cov_x12,
    "Actor–partner" = a1 * p2 * var_x1,
    "Partner–actor" = p1 * a2 * var_x2,
    "Partner–partner" = p1 * p2 * cov_x12,
    "Residual" = psi12
  )
})

# The residual variances in scenario B can be set to preserve both total
# outcome variances from scenario A; they are positive and Psi remains valid.
scenario_b_psi11 <- example_var_y1 - with(
  scenario_b_parameters,
  a1^2 * var_x1 + p1^2 * var_x2 + 2 * a1 * p1 * cov_x12
)
scenario_b_psi22 <- example_var_y2 - with(
  scenario_b_parameters,
  a2^2 * var_x2 + p2^2 * var_x1 + 2 * a2 * p2 * cov_x12
)

stopifnot(
  isTRUE(all.equal(sum(scenario_a), sum(scenario_b))),
  scenario_b_psi11 > 0,
  scenario_b_psi22 > 0,
  scenario_b_psi11 * scenario_b_psi22 >
    scenario_b_parameters$psi12^2
)

scenario_matrix <- rbind(
  "A: stronger transmission" = scenario_a,
  "B: similarity + residual" = scenario_b
) / example_denominator
```

```{r same-total-figure, echo=FALSE, fig.width=10, fig.height=5.2, out.width="96%", fig.cap="Two APIMs with the same model-implied outcome correlation but different route profiles. Each horizontal stack has exactly the same length.", fig.alt="Two horizontal stacked bars of identical total length. Scenario A has larger actor-partner and partner-actor sections; scenario B has larger actor-actor and residual sections."}
old_par <- graphics::par(mar = c(7.2, 10.2, 1.0, 1.0), xpd = NA)
graphics::barplot(
  t(scenario_matrix),
  horiz = TRUE,
  beside = FALSE,
  col = component_colours,
  border = colours[["white"]],
  names.arg = rownames(scenario_matrix),
  las = 1,
  xlim = c(0, example_rho_y12 * 1.04),
  xlab = "Outcome-correlation points",
  cex.names = 0.9
)
graphics::legend(
  "bottom", inset = c(0, -0.36),
  legend = names(component_colours),
  fill = component_colours, border = NA,
  horiz = TRUE, bty = "n", cex = 0.78, xpd = NA
)
graphics::par(old_par)
```

This is the paper's most intuitive motivation: reporting only the total or
residual correlation cannot distinguish these model-implied route profiles.

# From a fitted APIM to estimated contributions

The formulas define population or model-implied quantities. In an empirical
analysis, estimate them by plugging the fitted APIM coefficients and predictor
moments into the same expressions. Let
$\widehat\sigma_{X1}^2$, $\widehat\sigma_{X2}^2$, and
$\widehat\sigma_{X12}$ denote the estimated predictor variances and covariance,
and let $\widehat\psi_{12}$ denote the fitted residual covariance. Then

$$
\begin{aligned}
\widehat C_{\mathrm{AA}}
&=\widehat a_1\widehat a_2\widehat\sigma_{X12}, &
\widehat C_{\mathrm{AP}}
&=\widehat a_1\widehat p_2\widehat\sigma_{X1}^2,\\
\widehat C_{\mathrm{PA}}
&=\widehat p_1\widehat a_2\widehat\sigma_{X2}^2, &
\widehat C_{\mathrm{PP}}
&=\widehat p_1\widehat p_2\widehat\sigma_{X12},\\
\widehat C_{\mathrm{R}}
&=\widehat\psi_{12}.&&
\end{aligned}
$$

The estimated model-implied outcome variances supply the denominator for the
correlation-point contributions. For example,

$$
\widehat c_{\mathrm{AP}}
=\frac{\widehat C_{\mathrm{AP}}}
{\sqrt{\widehat{\operatorname{Var}}(Y_1)
       \widehat{\operatorname{Var}}(Y_2)}}.
$$

In an SEM, predictor moments may be estimated as part of the joint model or
supplied from the observed exogenous variables, depending on how the model is
specified. In a multilevel or regression implementation that conditions on the
predictors, use their empirical dyad-level variance–covariance matrix. In both
cases, the reported decomposition should state where those moments came from.

Likewise, $\widehat\psi_{12}$ must represent the **total model-implied
covariance remaining between the two members** after the fixed predictor paths.
Depending on the mixed-model parameterization, that quantity may be represented
by a shared dyad effect, a covariance between member-specific random effects,
a residual covariance, or a sum of such components. It should not be replaced
uncritically by whichever single variance component is easiest to extract.

## Uncertainty without complicated mathematics

A dyad bootstrap provides a direct uncertainty procedure and avoids deriving a
large delta-method covariance matrix:

1. Resample whole dyads with replacement, keeping both partners together and
   preserving their role labels.
2. Refit the APIM in the bootstrap sample.
3. Recompute the predictor moments, all five covariance components, the
   model-implied outcome variances, and the correlation-point contributions.
4. Repeat and form intervals from the bootstrap distribution of each signed
   component.

Resampling whole dyads preserves the dependence that the model is designed to
represent. Recomputing both the fitted paths and predictor moments carries both
sources of sampling uncertainty forward. The analysis plan should prespecify
the interval type and how nonconverged bootstrap fits are handled.

Within the fitted predictor model, the exact identity is

$$
\Sigma^{\mathrm{model}}_{Y,12}-\Psi_{12}
=C_{\mathrm{AA}}+C_{\mathrm{AP}}+C_{\mathrm{PA}}+C_{\mathrm{PP}}.
$$

An intercept-only model attributes all model-implied outcome covariance to its
residual covariance. However, the reduction in residual covariance between two
separately fitted models equals the predictor-side sum above only when both
models reproduce the same marginal outcome covariance under compatible
estimation and sample-moment conventions. The within-model identity is the
safer primary result.

# Matrix form and extension to several predictors

The scalar derivation is easiest to teach, but the same result can be written
compactly in matrix form. For the one-construct APIM, let

$$
\boldsymbol Y=
\begin{pmatrix}Y_1\\Y_2\end{pmatrix},
\qquad
\boldsymbol X=
\begin{pmatrix}X_1\\X_2\end{pmatrix},
\qquad
\boldsymbol B=
\begin{pmatrix}
a_1 & p_1\\
p_2 & a_2
\end{pmatrix},
$$

and

$$
\boldsymbol\Sigma_X=
\begin{pmatrix}
\sigma_{X1}^2 & \sigma_{X12}\\
\sigma_{X12} & \sigma_{X2}^2
\end{pmatrix},
\qquad
\boldsymbol\Psi=
\begin{pmatrix}
\psi_{11} & \psi_{12}\\
\psi_{12} & \psi_{22}
\end{pmatrix}.
$$

With $\boldsymbol Y=\boldsymbol B\boldsymbol X+\boldsymbol\varepsilon$ and
$\operatorname{Cov}(\boldsymbol X,\boldsymbol\varepsilon)=\boldsymbol 0$,

$$
\boldsymbol\Sigma_Y
=\boldsymbol B\boldsymbol\Sigma_X\boldsymbol B^\mathsf{T}
+\boldsymbol\Psi.
$$

For the covariance between the two outcomes,

$$
\operatorname{Cov}(Y_1,Y_2)
=\sum_r\sum_s B_{1r}\Sigma_{X,rs}B_{2s}+\Psi_{12}.
$$

Thus, for a linear APIM with exogenous predictors, every contribution combines
one coefficient into $Y_1$, one predictor variance or covariance, and one
coefficient into $Y_2$.

However, several predictors introduce cross-predictor covariance terms. Their
allocation to a named construct may depend on how predictors are grouped or
parameterized. The first paper should therefore make the single dyadic
predictor pair the transparent core result, then present the matrix expression
and a prespecified grouping rule as the extension.

# Brief closing extension: exchangeable dyads

The paper's main case is cross-sectional and distinguishable: member 1 is the
female partner and member 2 is the male partner, so the female-driven and
male-driven components can differ. A short closing section can show that the
same logic also applies to exchangeable dyads without turning exchangeability
into a second focus of the paper.

For exchangeable dyads, the member labels are arbitrary. The
covariance-relevant constraints are

$$
a_1=a_2=a,
\qquad
p_1=p_2=p,
$$

$$
\operatorname{Var}(X_1)=\operatorname{Var}(X_2)=\sigma_X^2,
\qquad
\operatorname{Cov}(X_1,X_2)=\sigma_{X12}.
$$

These are only the constraints needed to simplify the covariance components.
Second-order exchangeability also requires equal role-specific predictor
means, equal outcome intercepts, and equal residual variances. A full
distributional claim additionally requires a joint distribution that is
unchanged when the two member labels are swapped.

Keeping every path-tracing component visible gives

$$
\begin{aligned}
\operatorname{Cov}(Y_1,Y_2)
={}&
\underbrace{a^2\sigma_{X12}}_{\text{actor--actor}}
+\underbrace{ap\sigma_X^2}_{\text{member 1 driven}}\\
&+\underbrace{pa\sigma_X^2}_{\text{member 2 driven}}
+\underbrace{p^2\sigma_{X12}}_{\text{partner--partner}}
+\underbrace{\psi_{12}}_{\text{residual}}.
\end{aligned}
$$

The two member-driven contributions are now necessarily equal. Because
"member 1" and "member 2" are arbitrary seats, neither component has a stable
substantive identity. They should be combined:

$$
\underbrace{ap\sigma_X^2+ap\sigma_X^2}_{\text{member-driven transmission}}
=2ap\sigma_X^2.
$$

The complete compact form is therefore

$$
\operatorname{Cov}(Y_1,Y_2)
=\underbrace{2ap\sigma_X^2}_{\text{member-driven transmission}}
+\underbrace{(a^2+p^2)\sigma_{X12}}_{\text{cross-member predictor covariance}}
+\underbrace{\psi_{12}}_{\text{residual interdependence}}.
$$

Thus, exchangeability does not change the path-tracing logic. It changes what
can be named and compared: an exchangeable APIM estimates one symmetric
member-driven contribution, not separate female-driven and male-driven
contributions. Female–male dyads remain substantively distinguishable even if
some paths are constrained to equality. Such constraints test or impose
symmetry within a distinguishable model; they do not make the member labels
exchangeable. Under the full exchangeability constraints above, equality of
the two member-driven components follows from the model structure rather than
from evidence favoring one label assignment over another.

# Plan for the paper

The sections above form a teaching-oriented prototype of the paper's argument.
This final section records the intended scope, contribution, evidence, and
remaining work.

## Working scope

The first paper should focus on:

- cross-sectional, independent, distinguishable dyads;
- female and male partners as the motivating distinguishable roles;
- one predictor construct measured on both partners;
- continuous outcomes and a linear, fixed-slope APIM;
- marginal, model-implied outcome covariance and correlation;
- descriptive statistical attribution rather than automatic causal claims.

Several predictors and exchangeable dyads can appear as compact extensions.
Longitudinal paths, random slopes, nonlinear outcome models, and formal causal
identification should remain outside the first paper.

## Central claim and contributions

The central claim is that dyadic interdependence need not be treated only as a
nuisance to accommodate. Within a fitted APIM, the model-implied covariance
between partners' outcomes can be decomposed into interpretable, signed route
contributions.

The paper should contribute the following package:

1. Present the exact five-part decomposition for a distinguishable APIM in one
   diagram, one table, and a short covariance expansion.
2. Name and interpret the female-driven and male-driven components while
   retaining the complete actor–actor, actor–partner, partner–actor,
   partner–partner, and residual breakdown.
3. Distinguish predictor-transmission routes from cross-member
   predictor-covariance routes, without treating either label as automatically
   causal.
4. Put all five components on a common signed correlation-point scale and show
   why they are not percentages or separate correlations.
5. Explain why the residual correlation is not the residual contribution to
   the total outcome correlation.
6. Provide a plug-in estimator, dyad-bootstrap uncertainty intervals, and a
   reproducible table plus waterfall visualization.
7. Demonstrate that the same total interdependence can have different
   model-implied route profiles.
8. End with the compact exchangeable special case, where the two member-driven
   terms become equal and lose their role-specific interpretation.

## Novelty boundary and literature work

The covariance identity follows from standard path tracing and covariance
algebra; it should not be claimed as a newly discovered theorem. Existing APIM
teaching material contains versions of the standardized expression, and prior
dyadic work, including work by Bolger and Shrout, treats observed dyadic
covariance as potentially arising from interpersonal influence or shared
events.

The potentially publishable contribution is the coherent framework around the
identity: treating total interdependence as an estimand to explain, naming all
routes, placing them on one additive scale, quantifying uncertainty, providing
visualizations and software, and showing why equal total correlations need not
have equal explanations.

Before making a novelty claim, the literature review should establish:

- where the full distinguishable five-term expression has appeared;
- whether anyone has already estimated uncertainty for the individual terms;
- whether outcome-correlation-point contributions have been proposed;
- whether female-driven and male-driven interdependence have been named or
  compared explicitly;
- how previous work distinguishes residual covariance, explained covariance,
  and total interdependence.

The literature review should also establish how commonly APIM applications
report or decompose total outcome interdependence rather than only individual
paths and residual correlation.

## Mathematics in the manuscript

The main text needs only:

1. the two APIM equations;
2. the exogeneity and constant-residual-covariance assumptions;
3. the short covariance expansion producing the five terms;
4. the model-implied outcome-variance denominator;
5. the plug-in estimates and dyad-bootstrap recipe.

The expanded predictor–residual cross-terms, matrix generalization, and full
bootstrap implementation can go in an appendix or supplement. Delta-method
gradients, asymptotic proofs, elaborate matrix calculus, and longitudinal
derivations are not needed for the first paper.

## Proposed evidence

### Analytical and visual demonstration

- Derive the five components directly and by path tracing.
- Show how female-driven and male-driven covariance arise because one member's
  predictor reaches both outcomes.
- Contrast covariance contributions, correlation-point contributions,
  residual covariance, and residual correlation.
- Use the waterfall and same-total/different-profile figures as the primary
  intuitive demonstrations.

### Simulation study

The algebraic identity does not need simulation. Simulation is useful only to
study estimation and uncertainty. Vary sample size, predictor covariance,
unequal predictor variances, female–male coefficient differences, path signs,
and residual covariance. Evaluate:

- finite-sample bias of each plug-in component;
- coverage and width of dyad-bootstrap intervals;
- behavior when positive and negative components cancel;
- stability when total covariance is near zero;
- convergence and interval behavior in smaller samples.

Signed covariance and correlation-point contributions should be the primary
estimands. Percentage shares should not be a primary analysis.

### Empirical demonstration

Use one substantively clear predictor construct and outcome. Report:

1. raw partner outcome covariance and correlation;
2. the fitted model-implied covariance and correlation;
3. all five route contributions in covariance units;
4. all five correlation-point contributions;
5. the female-driven and male-driven comparison;
6. grouped predictor-transmission and predictor-covariance routes;
7. dyad-bootstrap uncertainty intervals;
8. one path diagram, one results table, and one signed waterfall plot.

The strongest example would compare two outcomes, predictors, or substantively
meaningful groups with similar total interdependence but different route
profiles.

## Planned manuscript structure

1. **The missing question:** APIMs quantify actor and partner paths, but what
   explains the covariance between the two outcomes?
2. **One picture, five routes:** introduce the full distinguishable APIM and
   the visual path tracing.
3. **Exact decomposition:** state assumptions, show the short derivation, and
   interpret female-driven, male-driven, predictor-covariance, and residual
   routes.
4. **A common reporting scale:** introduce signed correlation-point
   contributions and distinguish them from residual correlation.
5. **Estimation and uncertainty:** define plug-in estimates and the dyad
   bootstrap.
6. **Simulation:** assess finite-sample behavior and interval coverage.
7. **Empirical example:** show how similar total interdependence can have
   different route profiles.
8. **Extensions and limits:** several predictors, the brief exchangeable
   special case, and causal caution.
9. **Discussion:** reporting recommendations, substantive uses, and future
   longitudinal work.

## Planned displays

- **Figure 1:** complete distinguishable APIM.
- **Figure 2:** the five covariance-carrying routes plus their exact sum.
- **Figure 3:** signed waterfall decomposition of one outcome correlation.
- **Figure 4:** equal total interdependence with different route profiles.
- **Table 1:** formulas, substantive interpretations, and grouped route family.
- **Table 2:** empirical estimates with dyad-bootstrap intervals.

## Candidate titles

**Preferred**

> Explaining Interdependence in the Actor–Partner Interdependence Model:  
> Decomposing Dyadic Outcome Covariance

Alternatives:

- *Where Does Dyadic Interdependence Come From? A Covariance Decomposition for
  the APIM*
- *Beyond the Residual Correlation: Explaining Interdependence in the APIM*
- *From Actor and Partner Paths to Dyadic Interdependence*
- *Partitioning Interdependence in the Actor–Partner Interdependence Model*

## What to verify with Niall

The remembered scope is a cross-sectional APIM for distinguishable partners,
with particular interest in female-driven and male-driven interdependence. A
concise check would be:

> I reconstructed five sources of model-implied APIM outcome covariance:
> actor–actor and partner–partner routes carried by predictor covariance;
> actor–partner and partner–actor routes carried by each member's predictor
> variance; and residual covariance. For female–male dyads, the actor–partner
> term is female-driven and the partner–actor term is male-driven. Putting all
> five over the same model-implied outcome standard deviations makes them sum
> exactly to the model-implied outcome correlation. Is this the set of
> combinations you had in mind?

Further questions:

- Did he intend all four predictor-side components or mainly the female-driven
  and male-driven terms?
- Was covariance, correlation, or reduction in residual covariance his primary
  reporting target?
- Does he know an earlier derivation that should be the acknowledged starting
  point?
- Does he have a preferred empirical example or dataset?

## Immediate next steps

1. Show Niall the six-panel path figure and the concise question above.
2. Confirm intellectual provenance and naming before presenting a novelty
   claim.
3. Conduct the focused literature review described above.
4. Prototype an internal function that extracts model coefficients, predictor
   moments, total residual covariance, and all five contributions.
5. Test the decomposition and extraction logic on the workshop APIM example.
6. Choose an empirical example with a clear substantive story.
7. Run a small bootstrap prototype before designing the full simulation.
