---
title: "Actor-Partner Interdependence Model (APIM)"
author: "Pascal Küng"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Actor-Partner Interdependence Model (APIM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

diagram_helper <- if (file.exists("diagram-helpers.Rinc")) {
  "diagram-helpers.Rinc"
} else {
  file.path("vignettes", "diagram-helpers.Rinc")
}
sys.source(diagram_helper, envir = knitr::knit_global())
```

```{r setup}
library(interdep)
```

This vignette focuses on the cross-sectional and intensive longitudinal 
Actor-Partner Interdependence model for distinguishable and exchangeable dyads. 

For the main data requirements and validation workflow of the `interdep` package, start
with the [Getting Started vignette](getting-started.html). For APIMs that
combine distinguishable and exchangeable dyad compositions, see the
[Mixed-Composition APIM vignette](mixed-apim.html). For DIM predictors and their
equivalence to APIM effects in exchangeable dyads, see the
[Dyad-Individual Model vignette](dim.html). For DSM predictor scores and their
relationship to APIM effects in distinguishable dyads, see the
[Dyadic Score Model vignette](dsm.html).

> This vignette is under construction and for now only contains a few preliminary example models. Please check back soon!

## The cross sectional Gaussian distinguishable APIM

```{r distinguishable-apim-diagram, echo=FALSE, dev="svglite", fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Conceptual cross-sectional APIM for distinguishable female-male dyads. Actor and partner effects can differ by the role of the outcome member, and the two outcome residuals covary within dyads.", fig.alt="Path diagram for a distinguishable cross-sectional APIM. Female and male predictors each have an actor path to their own outcome and a partner path to the other member's outcome. The female and male outcome residuals covary."}
draw_apim_diagram("distinguishable")
```

Here, $a_f$ and $a_m$ are the actor effects on the female and male outcomes,
whereas $p_f$ and $p_m$ are the corresponding partner effects. All four paths
can differ in a distinguishable APIM.

### Residual random effects structure

### Interpretation

## The cross-sectional Gaussian exchangeable APIM

```{r exchangeable-apim-diagram, echo=FALSE, dev="svglite", fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Conceptual cross-sectional APIM for exchangeable dyads. The two members share one actor effect and one partner effect; their outcome residuals have equal variances and covary within dyads.", fig.alt="Path diagram for an exchangeable cross-sectional APIM. Each member's predictor has the same actor effect on their own outcome and the same partner effect on the other member's outcome. The two outcome residuals have equal variances and covary."}
draw_apim_diagram("exchangeable")
```

Because the member labels are arbitrary, swapping members 1 and 2 does not
change the model. The two actor paths therefore share the coefficient $a$, and
the two partner paths share the coefficient $p$. The two residual variances are
also constrained to be equal, while their covariance is estimated.

### Assumptions

### Residual random effects structure - and back-transformation

For an exchangeable dyad, `interdep` generates an arbitrary member-difference
column, named `.i_diff_*`, that is `+1` for one member and `-1` for the other.
The exchangeable residual structure is represented by two separate
random-effects terms: a shared dyad random intercept and a random coefficient
for this difference column [@delrosarioPracticalGuideSpecifying2025]. If their
random effects are denoted by $u_{Mj}$ and $u_{Dj}$, the two members receive

$$
r_{1j} = u_{Mj} + u_{Dj}, \qquad
r_{2j} = u_{Mj} - u_{Dj}.
$$

Because the shared and difference effects are fitted in separate random-effects
terms, they are uncorrelated. Writing their variances as $\sigma_M^2$ and
$\sigma_D^2$ gives

$$
\begin{pmatrix}
u_{Mj} \\
u_{Dj}
\end{pmatrix}
\sim
\mathcal{N}
\left[
\begin{pmatrix}
0 \\
0
\end{pmatrix},
\begin{pmatrix}
\sigma_M^2 & 0 \\
0 & \sigma_D^2
\end{pmatrix}
\right].
$$

The implied member-level residual covariance matrix is therefore

$$
\Sigma =
\begin{pmatrix}
\sigma_M^2 + \sigma_D^2 & \sigma_M^2 - \sigma_D^2 \\
\sigma_M^2 - \sigma_D^2 & \sigma_M^2 + \sigma_D^2
\end{pmatrix}.
$$

Thus, both members have the same residual variance
$V = \sigma_M^2 + \sigma_D^2$, their covariance is
$C = \sigma_M^2 - \sigma_D^2$, and their residual correlation is $C/V$.
Conversely, the shared and difference variances can be recovered from a
member-level variance and covariance as

$$
\sigma_M^2 = \frac{V + C}{2}, \qquad
\sigma_D^2 = \frac{V - C}{2}.
$$

Reversing the arbitrary `+1/-1` assignment changes the sign of $u_{Dj}$ but not
its variance or the implied member-level covariance matrix. In longitudinal
models, the same transformation applies separately to stable dyad-level and
same-occasion dyad-level covariance blocks.

### Interpretation

## Testing distinguishability

Aside from using a Wald test on the first model, nested model comparisons
require models fit to the same prepared data object. Helpers for creating those
constrained columns are planned.



## Intensive longitudinal APIMs

### Concurrent ILD Gaussian APIM for distinguishable dyads

Observed person means used to construct the between-person (`cbp`) predictors
can be unreliable when each member contributes few occasions, which can bias
between-person estimates [@gottfredsonStraightforwardApproachCoping2019].

Example model specification: 

```{r fit-ild-distinguishable, eval = FALSE}

ild_distinguishable_model <- glmmTMB(
  closeness ~ 0 + 
    
    .i_is_female_x_male_female + 
    .i_is_female_x_male_male + 
    
    # Gender specific time trends
    .i_is_female_x_male_female:diaryday + 
    .i_is_female_x_male_male:diaryday +
    
    # Gender-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +

    # Gender-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    # Gender-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +

    # Gender-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    # random effects for stable non-independence (means)
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID)  +

    # Same-day residual covariance
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID:diaryday) 

  , dispformula = ~ 0  
  , family = gaussian()
  , data = ild_distinguishable_data
)

summary(ild_distinguishable_model)

```


### Concurrent ILD Gaussian APIM for exchangeable dyads


### Current limitations of dyadic ILD designs in R

#### Dynamic models


---

From here, choose the model-specific vignette that matches the research
question:

- [Mixed-Composition APIM vignette](mixed-apim.html) for analyses combining
  distinguishable and exchangeable dyad compositions;
- [Dyad-Individual Model vignette](dim.html) for the exchangeable DIM
  parameterization; or
- [Dyadic Score Model vignette](dsm.html) for the distinguishable DSM
  parameterization.

Or return to the [Overview](index.html).

A vignette with non-Gaussian generalized APIM examples is planned. 


## References

::: {#refs}
:::
