---
title: "Actor-Partner Interdependence Model (APIM)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Actor-Partner Interdependence Model (APIM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```

This vignette focuses on APIM-style actor and partner effect preparation and
model formulas. For the main data requirements and validation workflow, start
with the [Getting Started vignette](getting-started.html). For DIM predictors
and their equivalence to APIM effects in exchangeable dyads, see the
[Dyad-Individual Model vignette](dim.html). For undirected dyadic score
outcomes, see the [Undirected Dyadic Score Model vignette](undirected-dsm.html).



### Test distinguishability

Aside from using a Wald test on the first model, nested model comparisons
require models fit to the same prepared data object. Helpers for creating those
constrained columns are planned.



## Semi-continuous cross-sectional data

`example_dyadic_crosssectional_tweedie` has the same dyadic structure as before, but the
outcome has exact zeros and positive skewed values, as in
some physical activity outcomes, in contrast to the Gaussian distribution from the 
previous example. 

```{r tweedie-raw}
print(head(example_dyadic_crosssectional_tweedie))
hist(example_dyadic_crosssectional_tweedie$physical_activity, breaks = 20)
```

Validation and modeling work the same way because the dyadic structure is the same.
The random-effect interpretation is different from the Gaussian case: Tweedie
random effects are latent effects on the log-mean scale, while the
observation-level Tweedie variance remains part of the model.

### Distinguishable Tweedie APIM

```{r prepare-tweedie-distinguishable}
tweedie_distinguishable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_tweedie,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = motivation,
  # when no model_type is specified, apim is the default.
  seed = 123
)

print(tweedie_distinguishable_data)
summary(tweedie_distinguishable_data)
```


```{r fit-tweedie-distinguishable, eval = FALSE}

tweedie_distinguishable_model <- glmmTMB(
  physical_activity ~ 
    # remove standard intercept and model separate intercepts for male and female
    0 + .i_is_female_x_male_female + .i_is_female_x_male_male + 
    
    # gender-specific slopes for motivation actor effect
    .i_is_female_x_male_female:.i_motivation_raw_actor + .i_is_female_x_male_male:.i_motivation_raw_actor +
    
    # gender-specific slopes for motivation partner effect
    .i_is_female_x_male_female:.i_motivation_raw_partner + .i_is_female_x_male_male:.i_motivation_raw_partner +
    
    # keep a simple couple-level latent effect for stable non-independence
    # important limitation: this can only induce positive partner dependence
    (1 | coupleID) 
  
  # allow role-specific Tweedie dispersion
  , dispformula = ~ 0 + .i_is_female_x_male_female + .i_is_female_x_male_male
  , family = tweedie()
  , data = tweedie_distinguishable_data
)

summary(tweedie_distinguishable_model)

```

### Exchangeable Tweedie APIM

```{r prepare-tweedie-exchangeable}
tweedie_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_tweedie,
  group = coupleID,
  member = personID,
  # role = gender,
  predictors = motivation,
  seed = 123
)

print(tweedie_exchangeable_data)
summary(tweedie_exchangeable_data)
```

```{r fit-tweedie-exchangeable, eval = FALSE}

tweedie_exchangeable_model <- glmmTMB(
  physical_activity ~ 
    # pooled intercept 
    1 + 
    
    # pooled actor slope for motivation
    .i_motivation_raw_actor +
    
    # pooled partner slope for motivation
    .i_motivation_raw_partner +
    
    # exchangeable latent dyad block on the log-mean scale
    (1 | coupleID) + (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID)
    
  # estimate a single pooled Tweedie dispersion parameter
  , dispformula = ~ 1
  , family = tweedie()
  , data = tweedie_exchangeable_data
)

summary(tweedie_exchangeable_model)

```


## ILD

Example model specification: 

```{r fit-ild-distinguishable, eval = FALSE}

ild_distinguishable_model <- glmmTMB(
  closeness ~ 0 + 
    
    .i_is_female_x_male_female + 
    .i_is_female_x_male_male + 
    
    # Gender specific time trends
    .i_is_female_x_male_female:diaryday + 
    .i_is_female_x_male_male:diaryday +
    
    # Gender-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +

    # Gender-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    # Gender-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +

    # Gender-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    # random effects for stable non-independence (means)
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID)  +

    # Same-day residual covariance
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID:diaryday) 

  , dispformula = ~ 0  
  , family = gaussian()
  , data = ild_distinguishable_data
)

summary(ild_distinguishable_model)

```


## Semi-continuous intensive longitudinal dyadic data

`example_dyadic_ILD_tweedie` has the same intensive longitudinal dyadic
structure as `example_dyadic_ILD`, but the outcome is semi-continuous.

```{r ild-tweedie-raw}
print(head(example_dyadic_ILD_tweedie, n = 26), n = 26)
hist(example_dyadic_ILD_tweedie$physical_activity, breaks = 20)
```

### Distinguishable Tweedie ILD APIM

```{r prepare-ild-tweedie-distinguishable}
ild_tweedie_distinguishable_data <- prepare_interdep_data(
  example_dyadic_ILD_tweedie,
  group = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(ild_tweedie_distinguishable_data)
summary(ild_tweedie_distinguishable_data)
```

For Tweedie models, the random-effect blocks are latent effects on the log-mean
scale. They can induce partner dependence, but they are not residual covariance
structures in the same direct sense as in Gaussian models, because the Tweedie
observation-level variance remains part of the model.

The first model uses role-specific stable couple effects and a simple same-day
shared latent shock. This is the easier teaching model: it keeps role-specific
Tweedie dispersion, but the same-day latent shock can only induce positive
partner dependence.

```{r fit-ild-tweedie-distinguishable-shared-day, eval = FALSE}

ild_tweedie_distinguishable_shared_day_model <- glmmTMB(
  physical_activity ~ 0 +

    .i_is_female_x_male_female + .i_is_female_x_male_male +

    .i_is_female_x_male_female:diaryday + .i_is_female_x_male_male:diaryday +

    # Gender-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +

    # Gender-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +

    # Gender-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +

    # Gender-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +

    # random effects for stable non-independence (means)
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # same-day shared latent shock; positive dependence only
    (1 | coupleID:diaryday)

  , dispformula = ~ 0 + .i_is_female_x_male_female + .i_is_female_x_male_male
  , family = tweedie()
  , data = ild_tweedie_distinguishable_data
)

summary(ild_tweedie_distinguishable_shared_day_model)

```

The second model uses a role-specific same-day latent covariance block. This is
closer to the Gaussian ILD covariance structure and can represent positive or
negative same-day partner dependence. In this ILD example, the repeated paired
occasions provide enough information to also estimate role-specific Tweedie
dispersion.

```{r fit-ild-tweedie-distinguishable-latent-day-cov, eval = FALSE}

ild_tweedie_distinguishable_latent_day_cov_model <- glmmTMB(
  physical_activity ~ 0 +

    .i_is_female_x_male_female + .i_is_female_x_male_male +

    .i_is_female_x_male_female:diaryday + .i_is_female_x_male_male:diaryday +

    # Gender-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +

    # Gender-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +

    # Gender-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +

    # Gender-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +

    # random effects for stable non-independence (means)
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # same-day role-specific latent covariance; positive or negative dependence
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID:diaryday)

  , dispformula = ~ 0 + .i_is_female_x_male_female + .i_is_female_x_male_male
  # in case of non-convergence, a first simplification could be to remove the
  # role-specific dispersion formula
  , family = tweedie()
  , data = ild_tweedie_distinguishable_data
)

summary(ild_tweedie_distinguishable_latent_day_cov_model)

```

The fuller same-day latent covariance structure is much more fragile in the
cross-sectional case because each dyad contributes only one paired occasion. The
same pair of observations must inform the fixed effects, Tweedie dispersion and
power, and the dyadic dependence structure. In ILD data, each dyad contributes
many paired occasions, so the stable couple-level block and the same-day
occasion-level block are informed by repeated within-dyad patterns over time.

### Exchangeable Tweedie ILD APIM

```{r prepare-ild-tweedie-exchangeable}
ild_tweedie_exchangeable_data <- prepare_interdep_data(
  example_dyadic_ILD_tweedie,
  group = coupleID,
  member = personID,
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(ild_tweedie_exchangeable_data)
summary(ild_tweedie_exchangeable_data)
```

This is the exchangeable analogue of the fuller Tweedie ILD model above. The
sum-diff random-effect blocks provide stable and same-day exchangeable latent
covariance structures. The Tweedie dispersion stays pooled because the
sum-diff signs identify exchangeable positions, not substantive roles.

```{r fit-ild-tweedie-exchangeable, eval = FALSE}

ild_tweedie_exchangeable_model <- glmmTMB(
  physical_activity ~
    1 +

    diaryday +

    # Pooled within-person actor and partner effects
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Pooled between-person actor and partner effects
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # stable exchangeable latent covariance
    (1 | coupleID) + (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID) +
  
    # same-day exchangeable latent covariance
    (1 | coupleID:diaryday) + (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  # pooled dispersion; exchangeable positions should not define dispersion differences
  , dispformula = ~ 1
  , family = tweedie()
  , data = ild_tweedie_exchangeable_data
)

summary(ild_tweedie_exchangeable_model)

```


# Cross-sectional mixed-dyads APIM

The data were simulated with the following fixed effects and residual covariance
parameters. For exchangeable dyads, `sum_variance` and `diff_variance` imply the
partner correlation.

```{r mixed-cross-sectional-targets, echo=FALSE}
mixed_cross_targets <- data.frame(
  block = c(
    "female_x_male",
    "female_x_male",
    "female_x_male",
    "female_x_female",
    "female_x_female",
    "female_x_female",
    "male_x_male",
    "male_x_male",
    "male_x_male"
  ),
  parameter = c(
    "female_mean",
    "male_mean",
    "correlation",
    "mean",
    "sum_variance",
    "diff_variance",
    "mean",
    "sum_variance",
    "diff_variance"
  ),
  value = c(
    5.50,
    4.50,
    -0.30,
    5.80,
    0.675,
    0.325,
    4.20,
    0.63375,
    1.05625
  )
)

mixed_cross_targets
```

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data)
```

```{r fit-mixed-cross-sectional-gaussian, eval = FALSE}

mixed_cross_gaussian_model <- glmmTMB(
  satisfaction ~ 0 +

    .i_is_female_x_male_female + .i_is_female_x_male_male +
    .i_is_female_x_female + .i_is_male_x_male +

    # distinguishable female-male residual covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # exchangeable female-female residual covariance via sum-diff
    (0 + .i_is_female_x_female | coupleID) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID) +

    # exchangeable male-male residual covariance via sum-diff
    (0 + .i_is_male_x_male | coupleID) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_cross_data
)

summary(mixed_cross_gaussian_model)

```

### Single versus separate mixed-composition fits

The mixed-composition models in this vignette fit all dyad compositions in one
model call. That is useful when the goal is to
compare effects across compositions, test equality constraints, keep one
model-based covariance matrix for those comparisons, or intentionally share
parameters such as a common Tweedie power parameter.

A fixed-effect formula with all compositions in one model call does not, by
itself, create partial pooling across dyad compositions. The
composition-specific intercepts and slopes above are ordinary fixed effects, so
estimates for female-female or male-male dyads are not automatically shrunk
toward estimates from the other dyad types. If every mean, variance, and
covariance parameter is composition-specific, the likelihood largely factorizes
by composition. In the cross-sectional Gaussian example, the single
mixed-composition fit and three separate composition-specific fits give the same
log-likelihood and fixed-effect estimates up to numerical tolerance.

Partial pooling requires a different model specification, such as a common effect
plus composition deviations with a hierarchical prior, or a random dyad-type
effect. With only a few dyad types, frequentist random dyad-type variance
components are usually weakly identified. Complete pooling, where a slope is
constrained to be identical across compositions, can be more stable and more
powerful when the constraint is substantively correct, but it is biased when the
dyad compositions truly differ.


## Mixed-composition intensive longitudinal Gaussian model

`example_dyadic_ILD_mixed` contains the same three dyad compositions as the
mixed-composition cross-sectional example, but each dyad contributes repeated paired
observations over `diaryday`.

To prepare the data: 

```{r prepare-mixed-ild}
mixed_ild_data <- prepare_interdep_data(
  example_dyadic_ILD_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(mixed_ild_data)
```


This first mixed-composition ILD model includes composition-specific intercepts, fixed
time slopes, and within-person and between-person actor and partner effects.
Random time slopes are omitted; with separate stable and same-day dyadic
covariance blocks, random time slopes add many weakly identified parameters and
are a likely first source of convergence problems.

This maximal covariance model is a demanding all-in-one likelihood. In this
simulated example, separate composition-specific fits recover the same
fixed-effect solution cleanly, while the all-in-one mixed-composition fit can
report false convergence even when it reaches the same likelihood and has a
positive definite Hessian.
Treat convergence diagnostics, Hessian diagnostics, and comparison with simpler
or separate fits as part of the workflow. If the all-in-one fit is not clean, fit
the composition-specific models separately for estimation, or simplify the
same-day and stable covariance blocks before adding more random effects.

```{r fit-mixed-ild-gaussian, eval = FALSE}

mixed_ild_gaussian_model <- glmmTMB(
  closeness ~ 0 +

    # Composition-specific intercepts
    .i_is_female_x_male_female + .i_is_female_x_male_male +
    
    .i_is_female_x_female + 
    .i_is_male_x_male +

    # Composition-specific time trends
    .i_is_female_x_male_female:diaryday +
    .i_is_female_x_male_male:diaryday +
    
    .i_is_female_x_female:diaryday +
    
    .i_is_male_x_male:diaryday +

    # Composition-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +
    
    .i_is_female_x_female:.i_provided_support_cwp_actor +
    
    .i_is_male_x_male:.i_provided_support_cwp_actor +

    # Composition-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    .i_is_female_x_female:.i_provided_support_cwp_partner +
    
    .i_is_male_x_male:.i_provided_support_cwp_partner +

    # Composition-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +
    
    .i_is_female_x_female:.i_provided_support_cbp_actor +
    
    .i_is_male_x_male:.i_provided_support_cbp_actor +

    # Composition-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    .i_is_female_x_female:.i_provided_support_cbp_partner +
    
    .i_is_male_x_male:.i_provided_support_cbp_partner +

    # stable dyad-level covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +
    
    (0 + .i_is_female_x_female | coupleID) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID) +
    
    (0 + .i_is_male_x_male | coupleID) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID) +

    # same-day covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID:diaryday) +
    
    (0 + .i_is_female_x_female | coupleID:diaryday) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID:diaryday) +
    
    (0 + .i_is_male_x_male | coupleID:diaryday) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_ild_data
)

summary(mixed_ild_gaussian_model)

```

## Mixed-composition intensive longitudinal Tweedie model

`example_dyadic_ILD_mixed_tweedie` has the same mixed-composition intensive
longitudinal dyadic structure, but the outcome is semi-continuous.

```{r mixed-ild-tweedie-raw}
print(head(example_dyadic_ILD_mixed_tweedie, n = 26), n = 26)
hist(example_dyadic_ILD_mixed_tweedie$physical_activity, breaks = 20)
```

```{r prepare-mixed-ild-tweedie}
mixed_ild_tweedie_data <- prepare_interdep_data(
  example_dyadic_ILD_mixed_tweedie,
  group = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(mixed_ild_tweedie_data)
summary(mixed_ild_tweedie_data)
```

As in the Gaussian mixed-composition ILD model, the mean model contains
composition-specific intercepts, fixed time slopes, and within-person and
between-person actor and partner effects. The dyadic random-effect blocks are
latent effects on the log-mean scale, not residual covariance parameters in the
Gaussian sense. This is the most fragile example in the vignette because it
combines composition-specific mean effects, composition-specific latent dyadic
dependence, composition-specific dispersion, and a common Tweedie power
parameter. A practical workflow is to fit separate Gaussian or simpler Tweedie
models first, then add shared-parameter constraints only when they answer a
specific comparison question.

```{r fit-mixed-ild-tweedie, eval = FALSE}

mixed_ild_tweedie_model <- glmmTMB(
  physical_activity ~ 0 +

    # Intercepts for each composition_role
    .i_is_female_x_male_female + 
    .i_is_female_x_male_male +
    
    .i_is_female_x_female + 
    
    .i_is_male_x_male +

    # C-specific specific time trends
    .i_is_female_x_male_female:diaryday + 
    .i_is_female_x_male_male:diaryday +
    
    .i_is_female_x_female:diaryday +
    
    .i_is_male_x_male:diaryday +

    # Composition-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +
    
    .i_is_female_x_female:.i_provided_support_cwp_actor +
    
    .i_is_male_x_male:.i_provided_support_cwp_actor +

    # Composition-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    .i_is_female_x_female:.i_provided_support_cwp_partner +
    
    .i_is_male_x_male:.i_provided_support_cwp_partner +

    # Composition-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +
    
    .i_is_female_x_female:.i_provided_support_cbp_actor +
    
    .i_is_male_x_male:.i_provided_support_cbp_actor +

    # Composition-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    .i_is_female_x_female:.i_provided_support_cbp_partner +
    
    .i_is_male_x_male:.i_provided_support_cbp_partner +

    # stable dyad-level latent covariance
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +
    
    (0 + .i_is_female_x_female | coupleID) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID) +
    
    (0 + .i_is_male_x_male | coupleID) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID) +

    # same-day latent covariance
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID:diaryday) +
    
    (0 + .i_is_female_x_female | coupleID:diaryday) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID:diaryday) +
    
    (0 + .i_is_male_x_male | coupleID:diaryday) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID:diaryday)

    # recommended first simplification in convergence issues is to set dispfomula = ~ 1
  , dispformula = ~ 0 +
      .i_is_female_x_male_female + .i_is_female_x_male_male +
      .i_is_female_x_female + .i_is_male_x_male
  , family = tweedie()
  , data = mixed_ild_tweedie_data
)

summary(mixed_ild_tweedie_model)

```

For an in-depth tutorial covering data preparation, model fitting, diagnostics,
and assumption checks, see
[Distinguishable and Exchangeable Dyads: Bayesian Multilevel Modelling](https://pascal-kueng.github.io/05DyadicDataAnalysis/).
It uses `interdep` for cross-sectional and intensive longitudinal APIM and DIM
workflows, with models fitted primarily using `brms`
([source](https://github.com/Pascal-Kueng/05DyadicDataAnalysis),
[DOI](https://doi.org/10.5281/zenodo.17400655)).
