---
title: "Actor-Partner Interdependence Model (APIM)"
author: "Pascal Küng"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Actor-Partner Interdependence Model (APIM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```

This vignette focuses on the cross-sectional and intensive longitudinal 
Actor-Partner Interdependence model for distinguishable and exchangeable dyads. 

For the main data requirements and validation workflow of the `interdep` package, start
with the [Getting Started vignette](getting-started.html). For APIMs that
combine distinguishable and exchangeable dyad compositions, see the
[Mixed-Composition APIM vignette](mixed-apim.html). For DIM predictors and their
equivalence to APIM effects in exchangeable dyads, see the
[Dyad-Individual Model vignette](dim.html). For DSM predictor scores and their
relationship to APIM effects in distinguishable dyads, see the
[Dyadic Score Model vignette](dsm.html).

> This vignette is under construction and for now only contains a few preliminary example models. Please check back soon!

## The cross sectional Gaussian distinguishable APIM

### Residual random effects structure

### Interpretation

## The cross-sectional Gaussian exchangeable APIM

### Assumptions

### Residual random effects structure - and back-transformation

### Interpretation

## Testing distinguishability

Aside from using a Wald test on the first model, nested model comparisons
require models fit to the same prepared data object. Helpers for creating those
constrained columns are planned.



## Intensive longitudinal APIMs

### Concurrent ILD Gaussian APIM for distinguishable dyads

Example model specification: 

```{r fit-ild-distinguishable, eval = FALSE}

ild_distinguishable_model <- glmmTMB(
  closeness ~ 0 + 
    
    .i_is_female_x_male_female + 
    .i_is_female_x_male_male + 
    
    # Gender specific time trends
    .i_is_female_x_male_female:diaryday + 
    .i_is_female_x_male_male:diaryday +
    
    # Gender-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +

    # Gender-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    # Gender-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +

    # Gender-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    # random effects for stable non-independence (means)
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID)  +

    # Same-day residual covariance
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID:diaryday) 

  , dispformula = ~ 0  
  , family = gaussian()
  , data = ild_distinguishable_data
)

summary(ild_distinguishable_model)

```


### Concurrent ILD Gaussian APIM for exchangeable dyads


### Current limitations of dyadic ILD designs in R

#### Dynamic models


---

**Continue** with the [Mixed-Composition APIM vignette](mixed-apim.html),

refer to the: 

- [Dyad-Individual Model vignette](dim.html),
- [Dyadic Score Model vignette](dsm.html),

or return to the [Overview](index.html).

A vignette with non-Gaussian generalized APIM examples is planned. 
