---
title: "Actor-Partner Interdependence Model (APIM)"
author: "Pascal Küng"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Actor-Partner Interdependence Model (APIM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

diagram_helper <- if (file.exists("diagram-helpers.Rinc")) {
  "diagram-helpers.Rinc"
} else {
  file.path("vignettes", "diagram-helpers.Rinc")
}
sys.source(diagram_helper, envir = knitr::knit_global())

diagram_device <- if (identical(Sys.getenv("IN_PKGDOWN"), "true")) {
  "ragg_png"
} else {
  "svglite"
}
```

```{r setup}
library(interdep)
```

This vignette focuses on the cross-sectional and intensive longitudinal 
Actor-Partner Interdependence model for distinguishable and exchangeable dyads. 

For the main data requirements and validation workflow of the `interdep` package, start
with the [Getting Started vignette](getting-started.html). For APIMs that
combine distinguishable and exchangeable dyad compositions, see the
[Mixed-Composition APIM vignette](mixed-apim.html). For DIM predictors and their
equivalence to APIM effects in exchangeable dyads, see the
[Dyad-Individual Model vignette](dim.html). For DSM predictor scores and their
relationship to APIM effects in distinguishable dyads, see the
[Dyadic Score Model vignette](dsm.html).

> This vignette is under construction and for now only contains a few preliminary example models. Please check back soon!

## The cross sectional Gaussian distinguishable APIM

```{r distinguishable-apim-diagram, echo=FALSE, dev=diagram_device, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Conceptual cross-sectional APIM for distinguishable female-male dyads. Actor and partner effects can differ by the role of the outcome member, and the two outcome residuals covary within dyads.", fig.alt="Path diagram for a distinguishable cross-sectional APIM. Female and male predictors each have an actor path to their own outcome and a partner path to the other member's outcome. The female and male outcome residuals covary."}
draw_apim_diagram("distinguishable")
```

Here, $a_f$ and $a_m$ are the actor effects on the female and male outcomes,
whereas $p_f$ and $p_m$ are the corresponding partner effects. All four paths
can differ in a distinguishable APIM.

We first prepare the example data with `prepare_interdep_data()`:

```{r prepare-cross-distinguishable-apim}
apim_distinguishable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  model_type = "apim"
)
```

The generated `.i_*` columns can be used directly in the model formula. Here is
a simple example:

```{r fit-cross-distinguishable-gaussian}
apim_distinguishable_model <- glmmTMB::glmmTMB(
  satisfaction ~

    # Gender-specific intercepts
    0 +
    .i_is_female_x_male_female +
    .i_is_female_x_male_male +

    # Gender-specific actor effects
    .i_is_female_x_male_female:.i_communication_actor +
    .i_is_female_x_male_male:.i_communication_actor +

    # Gender-specific partner effects
    .i_is_female_x_male_female:.i_communication_partner +
    .i_is_female_x_male_male:.i_communication_partner +

    # Dyad-level unstructured random effects represent the two partner
    # residual variances and their covariance when dispformula = ~ 0.
    # This is glmmTMB-specific syntax! lme4 and brms use different syntax.
    us(0 +
         .i_is_female_x_male_female +
         .i_is_female_x_male_male
       | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = apim_distinguishable_data
)
```

### Residual random-effects structure {#distinguishable-residual-structure}

For a distinguishable female-male dyad, the two members can have different
residual variances. Their residual covariance matrix is

$$
\Sigma_{FM} =
\begin{pmatrix}
\sigma_{FM,F}^2 & c_{FM} \\
c_{FM} & \sigma_{FM,M}^2
\end{pmatrix}.
$$

With one outcome row per member, this structure is estimated with an
unstructured random-effects block such as
`us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID)`
and `dispformula = ~ 0`. The two indicator columns select the female and male
rows, respectively. The `us()` block estimates both variances and their
covariance.

### Interpretation

## The cross-sectional Gaussian exchangeable APIM

```{r exchangeable-apim-diagram, echo=FALSE, dev=diagram_device, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Conceptual cross-sectional APIM for exchangeable dyads. The two members share one actor effect and one partner effect; their outcome residuals have equal variances and covary within dyads.", fig.alt="Path diagram for an exchangeable cross-sectional APIM. Each member's predictor has the same actor effect on their own outcome and the same partner effect on the other member's outcome. The two outcome residuals have equal variances and covary."}
draw_apim_diagram("exchangeable")
```

Because the member labels are arbitrary, swapping members 1 and 2 does not
change the model. The two actor paths therefore share the coefficient $a$, and
the two partner paths share the coefficient $p$. The two residual variances are
also constrained to be equal, while their covariance is estimated.

### Assumptions

### Residual random-effects structure and back-transformation {#exchangeable-residual-structure}

For an exchangeable dyad, `interdep` generates an arbitrary member-difference
column, named `.i_diff_*`, that is `+1` for one member and `-1` for the other.
The exchangeable residual structure is represented by two separate
random-effects terms: a shared dyad random intercept and a random coefficient
for this difference column [@delrosarioPracticalGuideSpecifying2025]. If their
random effects are denoted by $u_{Mj}$ and $u_{Dj}$, the two members receive

$$
r_{1j} = u_{Mj} + u_{Dj}, \qquad
r_{2j} = u_{Mj} - u_{Dj}.
$$

Because the shared and difference effects are fitted in separate random-effects
terms, they are uncorrelated. Writing their variances as $\sigma_M^2$ and
$\sigma_D^2$ gives

$$
\begin{pmatrix}
u_{Mj} \\
u_{Dj}
\end{pmatrix}
\sim
\mathcal{N}
\left[
\begin{pmatrix}
0 \\
0
\end{pmatrix},
\begin{pmatrix}
\sigma_M^2 & 0 \\
0 & \sigma_D^2
\end{pmatrix}
\right].
$$

The implied member-level residual covariance matrix is therefore

$$
\Sigma =
\begin{pmatrix}
\sigma_M^2 + \sigma_D^2 & \sigma_M^2 - \sigma_D^2 \\
\sigma_M^2 - \sigma_D^2 & \sigma_M^2 + \sigma_D^2
\end{pmatrix}.
$$

Thus, both members have the same residual variance
$V = \sigma_M^2 + \sigma_D^2$, their covariance is
$C = \sigma_M^2 - \sigma_D^2$, and their residual correlation is $C/V$.
Conversely, the shared and difference variances can be recovered from a
member-level variance and covariance as

$$
\sigma_M^2 = \frac{V + C}{2}, \qquad
\sigma_D^2 = \frac{V - C}{2}.
$$

For example, a random-intercept variance of 1.2 and a difference variance of
0.3 imply a member variance of 1.5 and a covariance between partners of 0.9.

Reversing the arbitrary `+1/-1` assignment changes the sign of $u_{Dj}$ but not
its variance or the implied member-level covariance matrix. In longitudinal
models, the same transformation applies separately to stable dyad-level and
same-occasion dyad-level covariance blocks.

### Interpretation

## Testing distinguishability {#testing-distinguishability}

Distinguishability can be evaluated by comparing a full model in which the two
roles may differ with a restricted exchangeable model. This comparison tests
the imposed equality constraints jointly. Here, they concern the fixed
intercepts, actor effects, partner effects, and residual variances.

The two parameterizations require different generated columns. The full
distinguishable model was fitted above, so we now prepare the same original
observations as exchangeable:

```{r prepare-cross-distinguishability}
apim_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  set_exchangeable_compositions = "female-male"
)
```

The exchangeable model constrains the corresponding fixed effects and residual
variances to be equal across members:

```{r fit-cross-distinguishability}
apim_exchangeable_model <- glmmTMB::glmmTMB(
  satisfaction ~ 0 +
    .i_is_female_x_male +
    .i_communication_actor +
    .i_communication_partner +
    (0 + .i_is_female_x_male | coupleID) +
    (0 + .i_diff_female_x_male_arbitrary | coupleID),
  dispformula = ~ 0,
  family = gaussian(),
  data = apim_exchangeable_data
)
```

`compare_interdep_models()` verifies that both models use equivalent original
observations before performing the likelihood-ratio test:

```{r compare-cross-distinguishability}
compare_interdep_models(
  restricted = apim_exchangeable_model,
  full = apim_distinguishable_model
)
```

The test provides evidence against all restrictions jointly, but it does not
show which parameter differs. The helper also cannot determine whether the
models are mathematically nested; that remains a modeling requirement. The
usual chi-squared reference distribution may be unreliable when a tested
variance parameter lies on its boundary.



## Intensive longitudinal APIMs

### Concurrent ILD Gaussian APIM for distinguishable dyads

Observed person means used to construct the between-person (`cbp`) predictors
can be unreliable when each member contributes few occasions, which can bias
between-person estimates [@gottfredsonStraightforwardApproachCoping2019].

Example model specification: 

```{r fit-ild-distinguishable, eval = FALSE}

ild_distinguishable_model <- glmmTMB(
  closeness ~ 0 + 
    
    .i_is_female_x_male_female + 
    .i_is_female_x_male_male + 
    
    # Gender specific time trends
    .i_is_female_x_male_female:diaryday + 
    .i_is_female_x_male_male:diaryday +
    
    # Gender-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +

    # Gender-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    # Gender-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +

    # Gender-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    # random effects for stable non-independence (means)
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID)  +

    # Same-day residual covariance
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID:diaryday) 

  , dispformula = ~ 0  
  , family = gaussian()
  , data = ild_distinguishable_data
)

summary(ild_distinguishable_model)

```


### Concurrent ILD Gaussian APIM for exchangeable dyads

Following del Rosario and West, the stable dyad covariance can be represented
using sum-and-difference random effects
[@delrosarioPracticalGuideSpecifying2025]. In longitudinal Gaussian APIMs fitted
with `glmmTMB`, the same parameterization can be extended to the dyad-occasion
level to represent same-occasion residual dependence.


### Current limitations of dyadic ILD designs in R

Concurrent dyadic ILD models can adjust for a time trend and account for stable
dyadic dependence and same-occasion partner dependence. They do not model
residual serial dependence, however, and therefore estimate concurrent
associations under the assumption that residuals from different days are
independent.

If the goal is to retain these concurrent APIM associations while accounting
for serial dependence, the closest extension is a dyadic residual dynamic
structural equation model (RDSEM). It keeps the concurrent regression separate
from a VAR model for its residuals [@asparouhovComparisonModelsAnalysis2020;
@mcneishPrimerTwoLevel2020]. This residual-VAR structure is not directly
available through the `glmmTMB` interface used here, and open-source support for
dyadic dynamic models remains very limited
[@delrosarioPracticalGuideSpecifying2025]. Within open-source R, such a model
generally requires custom TMB or Stan code.

#### Dynamic models {#dynamic-models}

A **practical alternative** is a model with lagged outcomes, especially when
carryover or temporal dynamics are part of the research question
[@gistelinckMultilevelAutoregressiveDyadic2020]. In such a model, the
interpretation changes. All APIM predictor effects then describe associations
conditional on the members' prior outcomes.

By adding the outcome to `predictors` and selecting it with `lag_predictors`,
`prepare_interdep_data()` returns lag-1 raw and within-person scores alongside
the contemporaneous scores. Between-person scores are not lagged because they
describe stable differences between members.

For this example, we obtain the lagged actor and partner outcome columns through
the `lag_predictors` argument:

```{r prepare-ild-apim-dynamic-model}
ild_apim_data_dynamic <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  time = diaryday,
  predictors = closeness,
  lag_predictors = closeness,
  model_type = "apim",
  seed = 123
)
```

This returns all necessary variables for either choice, including lag-1 raw
and within-person closeness scores. Lags are matched at exactly
`diaryday - 1`, so omitted diary days are not bridged.

A simple fixed-slope dyadic stability and influence model
[@delrosarioPracticalGuideSpecifying2025]:

```{r fit-ild-stability-influence}
stability_influence <- glmmTMB::glmmTMB(
  closeness ~ 1 +

    # Stability (actor effect across time)
    .i_closeness_actor_lag1 +

    # Influence (partner effect across time)
    .i_closeness_partner_lag1 +

    # Linear time trend
    diaryday +

    # Stable exchangeable dyad-level covariance
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID) +

    # Same-day exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_apim_data_dynamic
)

summary(stability_influence)
```

This model can be extended with contemporaneous actor and partner predictor
associations and other lagged predictors. Time-varying predictors should usually
be separated into within-person and between-person components. Their
contemporaneous coefficients are then conditional on both partners' prior
outcomes.

The model above uses raw lagged outcomes. The corresponding
within-person-centered lagged terms are `.i_closeness_cwp_actor_lag1` and
`.i_closeness_cwp_partner_lag1`.

Whether to use a raw or within-person-centered lagged outcome depends on the
research question and the data. Person-mean centering the outcome lag can bias
the average carryover estimate downward [@hamakerCenterNotCenter2015]. This
downward bias is known as Nickell bias [@nickellBiasesDynamicModels1981]. A raw
lag avoids this centering bias, but a standard random-intercept model can still
be biased because it assumes that the lag is unrelated to stable member levels
[@gistelinckMultilevelAutoregressiveSmall2021].

Both problems matter most when there are few occasions. In one simple
simulation, most approaches other than person-mean centering performed
acceptably from about ten occasions, but this is not a universal cutoff
[@gistelinckMultilevelAutoregressiveSmall2021]. For shorter panels, consider an
LD-APIM, which lets the first outcomes relate to the members' stable levels in a
wide-format SEM [@gistelinckMultilevelAutoregressiveDyadic2020].

These manifest-lag models are not equivalent to Mplus DSEM, which uses latent
person-mean centering by default and can estimate multivariate or residual
dynamics jointly [@mcneishPrimerTwoLevel2020]. For very short panels, however,
default DSEM may still be biased or unstable, so the LD-APIM recommendation
above may be preferable [@gistelinckMultilevelAutoregressiveSmall2021].

These lagged-outcome issues are separate from the earlier concern about
unreliable person means used to construct `cbp` predictors with few occasions.


---

From here, choose the model-specific vignette that matches the research
question:

- [Mixed-Composition APIM vignette](mixed-apim.html) for analyses combining
  distinguishable and exchangeable dyad compositions;
- [Dyad-Individual Model vignette](dim.html) for the exchangeable DIM
  parameterization; or
- [Dyadic Score Model vignette](dsm.html) for the distinguishable DSM
  parameterization.

Or return to the [Overview](index.html).

A vignette with non-Gaussian generalized APIM examples is planned. 


## References

::: {#refs}
:::
