---
title: "Actor-Partner Interdependence Model (APIM)"
author: "Pascal Küng"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Actor-Partner Interdependence Model (APIM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```

This vignette focuses on the cross-sectional and intensive longitudinal 
Actor-Partner Interdependence model for distinguishable and exchangeable dyads. 

For the main data requirements and validation workflow of the `interdep` package, start
with the [Getting Started vignette](getting-started.html). For APIMs that
combine distinguishable and exchangeable dyad compositions, see the
[Mixed-Composition APIM vignette](mixed-apim.html). For DIM predictors and their
equivalence to APIM effects in exchangeable dyads, see the
[Dyad-Individual Model vignette](dim.html). For DSM predictor scores and their
relationship to APIM effects in distinguishable dyads, see the
[Dyadic Score Model vignette](dsm.html).

> This vignette is under construction and for now only contains a few preliminary example models. Please check back soon!


### Test distinguishability

Aside from using a Wald test on the first model, nested model comparisons
require models fit to the same prepared data object. Helpers for creating those
constrained columns are planned.



## Semi-continuous cross-sectional data

`example_dyadic_crosssectional_tweedie` has the same dyadic structure as before, but the
outcome has exact zeros and positive skewed values, as in
some physical activity outcomes, in contrast to the Gaussian distribution from the 
previous example. 

```{r tweedie-raw}
print(head(example_dyadic_crosssectional_tweedie))
hist(example_dyadic_crosssectional_tweedie$physical_activity, breaks = 20)
```

Validation and modeling work the same way because the dyadic structure is the same.
The random-effect interpretation is different from the Gaussian case: Tweedie
random effects are latent effects on the log-mean scale, while the
observation-level Tweedie variance remains part of the model.

### Distinguishable Tweedie APIM

```{r prepare-tweedie-distinguishable}
tweedie_distinguishable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_tweedie,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = motivation,
  # when no model_type is specified, apim is the default.
  seed = 123
)

print(tweedie_distinguishable_data)
summary(tweedie_distinguishable_data)
```


```{r fit-tweedie-distinguishable, eval = FALSE}

tweedie_distinguishable_model <- glmmTMB(
  physical_activity ~ 
    # remove standard intercept and model separate intercepts for male and female
    0 + .i_is_female_x_male_female + .i_is_female_x_male_male + 
    
    # gender-specific slopes for motivation actor effect
    .i_is_female_x_male_female:.i_motivation_actor + .i_is_female_x_male_male:.i_motivation_actor +
    
    # gender-specific slopes for motivation partner effect
    .i_is_female_x_male_female:.i_motivation_partner + .i_is_female_x_male_male:.i_motivation_partner +
    
    # keep a simple couple-level latent effect for stable non-independence
    # important limitation: this can only induce positive partner dependence
    (1 | coupleID) 
  
  # allow role-specific Tweedie dispersion
  , dispformula = ~ 0 + .i_is_female_x_male_female + .i_is_female_x_male_male
  , family = tweedie()
  , data = tweedie_distinguishable_data
)

summary(tweedie_distinguishable_model)

```

### Exchangeable Tweedie APIM

```{r prepare-tweedie-exchangeable}
tweedie_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_tweedie,
  group = coupleID,
  member = personID,
  # role = gender,
  predictors = motivation,
  seed = 123
)

print(tweedie_exchangeable_data)
summary(tweedie_exchangeable_data)
```

```{r fit-tweedie-exchangeable, eval = FALSE}

tweedie_exchangeable_model <- glmmTMB(
  physical_activity ~ 
    # pooled intercept 
    1 + 
    
    # pooled actor slope for motivation
    .i_motivation_actor +
    
    # pooled partner slope for motivation
    .i_motivation_partner +
    
    # exchangeable latent dyad block on the log-mean scale
    (1 | coupleID) + (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID)
    
  # estimate a single pooled Tweedie dispersion parameter
  , dispformula = ~ 1
  , family = tweedie()
  , data = tweedie_exchangeable_data
)

summary(tweedie_exchangeable_model)

```


## ILD

Example model specification: 

```{r fit-ild-distinguishable, eval = FALSE}

ild_distinguishable_model <- glmmTMB(
  closeness ~ 0 + 
    
    .i_is_female_x_male_female + 
    .i_is_female_x_male_male + 
    
    # Gender specific time trends
    .i_is_female_x_male_female:diaryday + 
    .i_is_female_x_male_male:diaryday +
    
    # Gender-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +

    # Gender-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    # Gender-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +

    # Gender-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    # random effects for stable non-independence (means)
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID)  +

    # Same-day residual covariance
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID:diaryday) 

  , dispformula = ~ 0  
  , family = gaussian()
  , data = ild_distinguishable_data
)

summary(ild_distinguishable_model)

```


## Semi-continuous intensive longitudinal dyadic data

`example_dyadic_ILD_tweedie` has the same intensive longitudinal dyadic
structure as `example_dyadic_ILD`, but the outcome is semi-continuous.

```{r ild-tweedie-raw}
print(head(example_dyadic_ILD_tweedie, n = 26), n = 26)
hist(example_dyadic_ILD_tweedie$physical_activity, breaks = 20)
```

### Distinguishable Tweedie ILD APIM

```{r prepare-ild-tweedie-distinguishable}
ild_tweedie_distinguishable_data <- prepare_interdep_data(
  example_dyadic_ILD_tweedie,
  group = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(ild_tweedie_distinguishable_data)
summary(ild_tweedie_distinguishable_data)
```

For Tweedie models, the random-effect blocks are latent effects on the log-mean
scale. They can induce partner dependence, but they are not residual covariance
structures in the same direct sense as in Gaussian models, because the Tweedie
observation-level variance remains part of the model.

The first model uses role-specific stable couple effects and a simple same-day
shared latent shock. This is the easier teaching model: it keeps role-specific
Tweedie dispersion, but the same-day latent shock can only induce positive
partner dependence.

```{r fit-ild-tweedie-distinguishable-shared-day, eval = FALSE}

ild_tweedie_distinguishable_shared_day_model <- glmmTMB(
  physical_activity ~ 0 +

    .i_is_female_x_male_female + .i_is_female_x_male_male +

    .i_is_female_x_male_female:diaryday + .i_is_female_x_male_male:diaryday +

    # Gender-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +

    # Gender-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +

    # Gender-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +

    # Gender-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +

    # random effects for stable non-independence (means)
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # same-day shared latent shock; positive dependence only
    (1 | coupleID:diaryday)

  , dispformula = ~ 0 + .i_is_female_x_male_female + .i_is_female_x_male_male
  , family = tweedie()
  , data = ild_tweedie_distinguishable_data
)

summary(ild_tweedie_distinguishable_shared_day_model)

```

The second model uses a role-specific same-day latent covariance block. This is
closer to the Gaussian ILD covariance structure and can represent positive or
negative same-day partner dependence. In this ILD example, the repeated paired
occasions provide enough information to also estimate role-specific Tweedie
dispersion.

```{r fit-ild-tweedie-distinguishable-latent-day-cov, eval = FALSE}

ild_tweedie_distinguishable_latent_day_cov_model <- glmmTMB(
  physical_activity ~ 0 +

    .i_is_female_x_male_female + .i_is_female_x_male_male +

    .i_is_female_x_male_female:diaryday + .i_is_female_x_male_male:diaryday +

    # Gender-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +

    # Gender-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +

    # Gender-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +

    # Gender-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +

    # random effects for stable non-independence (means)
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # same-day role-specific latent covariance; positive or negative dependence
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID:diaryday)

  , dispformula = ~ 0 + .i_is_female_x_male_female + .i_is_female_x_male_male
  # in case of non-convergence, a first simplification could be to remove the
  # role-specific dispersion formula
  , family = tweedie()
  , data = ild_tweedie_distinguishable_data
)

summary(ild_tweedie_distinguishable_latent_day_cov_model)

```

The fuller same-day latent covariance structure is much more fragile in the
cross-sectional case because each dyad contributes only one paired occasion. The
same pair of observations must inform the fixed effects, Tweedie dispersion and
power, and the dyadic dependence structure. In ILD data, each dyad contributes
many paired occasions, so the stable couple-level block and the same-day
occasion-level block are informed by repeated within-dyad patterns over time.

### Exchangeable Tweedie ILD APIM

```{r prepare-ild-tweedie-exchangeable}
ild_tweedie_exchangeable_data <- prepare_interdep_data(
  example_dyadic_ILD_tweedie,
  group = coupleID,
  member = personID,
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(ild_tweedie_exchangeable_data)
summary(ild_tweedie_exchangeable_data)
```

This is the exchangeable analogue of the fuller Tweedie ILD model above. The
sum-diff random-effect blocks provide stable and same-day exchangeable latent
covariance structures. The Tweedie dispersion stays pooled because the
sum-diff signs identify exchangeable positions, not substantive roles.

```{r fit-ild-tweedie-exchangeable, eval = FALSE}

ild_tweedie_exchangeable_model <- glmmTMB(
  physical_activity ~
    1 +

    diaryday +

    # Pooled within-person actor and partner effects
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Pooled between-person actor and partner effects
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # stable exchangeable latent covariance
    (1 | coupleID) + (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID) +
  
    # same-day exchangeable latent covariance
    (1 | coupleID:diaryday) + (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  # pooled dispersion; exchangeable positions should not define dispersion differences
  , dispformula = ~ 1
  , family = tweedie()
  , data = ild_tweedie_exchangeable_data
)

summary(ild_tweedie_exchangeable_model)

```

## Back transforming the variance-covariance matrix from Idiff to actor-partner

-> should be the next model function?

---

**Continue** with the [Mixed-Composition APIM vignette](mixed-apim.html),

refer to the: 

- [Dyad-Individual Model vignette](dim.html),
- [Dyadic Score Model vignette](dsm.html),

or return to the [Overview](index.html).
