---
title: "Actor-Partner Interdependence Model (APIM)"
author: "Pascal Küng"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Actor-Partner Interdependence Model (APIM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

diagram_helper <- if (file.exists("diagram-helpers.Rinc")) {
  "diagram-helpers.Rinc"
} else {
  file.path("vignettes", "diagram-helpers.Rinc")
}
sys.source(diagram_helper, envir = knitr::knit_global())

diagram_device <- .vignette_diagram_device()
knitr::opts_chunk$set(dev = diagram_device, fig.ext = diagram_device)
```

```{r setup}
library(interdep)
```


> This vignette is under construction and preliminary. Please check back soon!


This vignette focuses on the Gaussian cross-sectional and intensive longitudinal 
Actor-Partner Interdependence model for distinguishable and exchangeable dyads. 

For the broader package workflow and an overview of the available
model-specific vignettes, including the
[Mixed-Composition APIM](mixed-apim.html),
[Dyad-Individual Model](dim.html), and
[Dyadic Score Model](dsm.html), see the [Overview](index.html).

A vignette for non-Gaussian generalized models is planned.


# Cross sectional APIMs

## The distinguishable APIM

A conceptual example for distinguishable female-male dyads:

```{r distinguishable-apim-diagram, echo=FALSE, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Conceptual cross-sectional APIM for distinguishable female-male dyads. Intercepts $b_\mathrm{0}$, actor effects $a$, and partner effects $p$ can differ by the role of the outcome member (F and M), and the two outcome residuals covary within dyads.", fig.alt="Path diagram for a distinguishable cross-sectional APIM. Female and male outcomes have separate intercepts. Female and male predictors each have an actor path to their own outcome and a partner path to the other member's outcome. The female and male outcome residuals covary."}
draw_apim_diagram("distinguishable")
```

For univariate MLM software like `glmmTMB`, this model is fitted in long format 
with one outcome row per member, which can be visualized as:

```{r distinguishable-apim-member-diagram, echo=FALSE, fig.width=9, fig.height=5.2, out.width="100%", fig.cap="Individual-level representation of the distinguishable cross-sectional APIM used for the long-format multilevel model. For the female outcome, the female predictor is the actor predictor and the male predictor is the partner predictor; these roles reverse for the male outcome. Intercepts, actor coefficients, and partner coefficients may differ by outcome role, and the two member residuals may have different variances and covary.", fig.alt="Two-panel path diagram for a distinguishable female-male APIM. The female and male outcomes have separate intercepts. In the female outcome panel, female X is the actor predictor and male X is the partner predictor of female Y, with coefficients a F and p F. In the male outcome panel, male X is the actor predictor and female X is the partner predictor of male Y, with coefficients a M and p M. The female and male outcome residuals covary."}
draw_apim_member_diagram("distinguishable")
```


### Residual random-effects structure {#distinguishable-residual-structure}

For a distinguishable female-male dyad, the two members can have different
residual variances. The within-dyad residual covariance block (shared across dyads) is:

$$
\operatorname{Cov}
\begin{pmatrix}
\epsilon_{Fi} \\
\epsilon_{Mi}
\end{pmatrix}
= \boldsymbol{\Sigma}_{\epsilon}
= \begin{bmatrix}
\sigma_{\epsilon_F}^{2}
& \rho_{\epsilon_F\epsilon_M}\sigma_{\epsilon_F}\sigma_{\epsilon_M} \\
\rho_{\epsilon_F\epsilon_M}\sigma_{\epsilon_F}\sigma_{\epsilon_M}
& \sigma_{\epsilon_M}^{2}
\end{bmatrix}
$$

And the full residual covariance matrix for all dyads (first three shown) is then block-diagonal:

$$
\boldsymbol{\Sigma}_{\mathrm{model}}
= \begin{bmatrix}
\boldsymbol{\Sigma}_{\epsilon} & \boldsymbol{0} & \boldsymbol{0} & \cdots \\
\boldsymbol{0} & \boldsymbol{\Sigma}_{\epsilon} & \boldsymbol{0} & \cdots \\
\boldsymbol{0} & \boldsymbol{0} & \boldsymbol{\Sigma}_{\epsilon} & \cdots \\
\vdots & \vdots & \vdots & \ddots
\end{bmatrix}
$$


This structure is estimated with an unstructured random-effects block such as
`us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID)`
and `dispformula = ~ 0`. 

<!-- TODO: what is the best order? This shows the RE before we prepped the data.
But if we move this whole stuff after data-prep, it interrupts before interpretation. 


TODO: should we discuss constraints briefly too, since we do it for exchnageable dyads? 
Like removeing the correlation, or enforcing equal variances as well (homcs()), 
if the model does not converge etc.?
-->

### Fitting the distinguishable APIM with glmmTMB

We first prepare the example data with `interdep::prepare_interdep_data()`:

```{r prepare-cross-distinguishable-apim}
apim_distinguishable_data <- interdep::prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  model_type = "apim"
)

print(apim_distinguishable_data, n=4)

```

The generated `.i_*` columns can be used directly in the model formula. Here is
a simple example:

```{r fit-cross-distinguishable-gaussian}
apim_distinguishable_model <- glmmTMB::glmmTMB(
  satisfaction ~

    # Gender-specific intercepts
    0 +
    .i_is_female_x_male_female +
    .i_is_female_x_male_male +

    # Gender-specific actor effects
    .i_is_female_x_male_female:.i_communication_actor +
    .i_is_female_x_male_male:.i_communication_actor +

    # Gender-specific partner effects
    .i_is_female_x_male_female:.i_communication_partner +
    .i_is_female_x_male_male:.i_communication_partner +

    # Dyad-level unstructured random effects represent the two partner
    # residual variances and their covariance when dispformula = ~ 0.
    # This is glmmTMB-specific syntax! `brms` uses different syntax.
    us(0 +
         .i_is_female_x_male_female +
         .i_is_female_x_male_male
       | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = apim_distinguishable_data
)

summary(apim_distinguishable_model)
```



```{r prepare-fitted-distinguishable-apim-diagram, include=FALSE}
apim_distinguishable_diagram_values <- .extract_apim_diagram_values(
  apim_distinguishable_model,
  type = "distinguishable"
)
apim_distinguishable_diagram_estimates <-
  apim_distinguishable_diagram_values$estimates
apim_distinguishable_diagram_residuals <-
  apim_distinguishable_diagram_values$residuals

apim_distinguishable_fitted_alt <- sprintf(
  paste(
    "Fitted distinguishable APIM. Female and male intercepts %.2f and %.2f;",
    "actor effects %.2f and %.2f; partner effects %.2f and %.2f;",
    "residual SDs %.2f and %.2f, with correlation %.2f."
  ),
  apim_distinguishable_diagram_estimates[["intercept_female"]],
  apim_distinguishable_diagram_estimates[["intercept_male"]],
  apim_distinguishable_diagram_estimates[["actor_female"]],
  apim_distinguishable_diagram_estimates[["actor_male"]],
  apim_distinguishable_diagram_estimates[["partner_female"]],
  apim_distinguishable_diagram_estimates[["partner_male"]],
  apim_distinguishable_diagram_residuals[["sd_female"]],
  apim_distinguishable_diagram_residuals[["sd_male"]],
  apim_distinguishable_diagram_residuals[["correlation"]]
)
```

The estimated coefficients map as follows: 

```{r fitted-distinguishable-apim-diagram, echo=FALSE, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Fitted cross-sectional distinguishable APIM for the example data. Fixed effects, residual standard deviations, and the residual correlation are extracted from the fitted model.", fig.alt=apim_distinguishable_fitted_alt}
draw_apim_diagram(
  "distinguishable",
  model = apim_distinguishable_model,
  labels = c(predictor = "Communication", outcome = "Satisfaction")
)
```

<!-- TODO: add a brief interpretation section? 

### Interpretation

Maybe like on:
https://quantdev.ssri.psu.edu/tutorials/apa-ati-intensive-longitudinal-data-session-m-dyadic-multilevel-model

-->

## The exchangeable APIM

Conceptually, the exchangeable APIM constrains several of the effects to be equal: 

```{r exchangeable-apim-diagram, echo=FALSE, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Conceptual cross-sectional APIM for exchangeable dyads. The two members share one intercept, one actor effect, and one partner effect. Their outcome residuals have equal variances, yet still covary within dyads.", fig.alt="Path diagram for an exchangeable cross-sectional APIM. Both outcomes have the same intercept. Each member's predictor has the same actor effect on their own outcome and the same partner effect on the other member's outcome. The two outcome residuals have equal variances and covary."}
draw_apim_diagram("exchangeable")
```

Because the member labels are arbitrary, swapping members 1 and 2 does **not**
change the model. 

To estimate this model in a univariate MLM framework, we can draw a conceptual
diagram as such: 

```{r exchangeable-apim-member-diagram, echo=FALSE, fig.width=9, fig.height=5.2, out.width="100%", fig.cap="Individual-level representation of the exchangeable cross-sectional APIM used for the long-format multilevel model. Both members share the same intercept. Each member's own predictor has the shared actor effect, and the other member's predictor has the shared partner effect. The two residual variances are equal and the residuals may covary.", fig.alt="Two-panel path diagram for an exchangeable APIM. Both outcomes have the same intercept. For arbitrary member 1, X 1 is the actor predictor and X 2 is the partner predictor of Y 1. For arbitrary member 2, X 2 is the actor predictor and X 1 is the partner predictor of Y 2. Both panels use the same actor coefficient a and partner coefficient p, and their outcome residuals covary."}
draw_apim_member_diagram("exchangeable")
```


### Modeling the residual random-effects structure {#exchangeable-residual-structure}

Currently, glmmTMB and brms allow us to specify the relevant residual correlation structure, 
but any additional random effects structure that includes more than just a random intercept presents a problem:

For instance, in glmmTMB the expression `homcs(0 + arbitrary_member_1 + arbitrary_member_2 | coupleID)` 
would achieve in setting both residual variances equal while still estimating a correlation between
the residuals. However, if we want to model intensive longitudinal data, like we do below, 
using `homcs(0 + arbitrary_member_1 + arbitrary_member_2 + time:arbitrary_member1 + arbitrary_member_2 | coupleID)` 
would impose a single variance shared by both intercepts *and* both time slopes. 
`homcs(0 + arbitrary_member_1 + arbitrary_member_2)  + homcs(time:arbitrary_member1 + arbitrary_member_2 | coupleID)` 
would produce the correct variance restrictions, but does then not model a correlation between slopes and intercepts
anymore. 

Therefore, we introduce a solution that works for residuals and other random effect terms regardless of slopes. 
Following @delrosarioPracticalGuideSpecifying2025, `interdep::prepare_interdep_data()` generates an 
arbitrary member-difference column, named `.i_diff_*`, that is `+1` for one member and `-1` for the other.
The exchangeable residual structure is represented by two separate
random-effects terms: a shared dyad random intercept and a random coefficient
for this difference column. 

<!-- TODO: do we need a brief explanation of why this is possible? -->

We will now fit the model and then use the `interdep::exchangeable_rescov()` 
function that "rotates" the structure back to the often more interpretable 
actor-partner residual-covariance matrix we are used to. 



### Fitting the exchangeable APIM with glmmTMB

<!-- TODO: insert the model fitting here, and then the applied back-transform, and only then explain
all the back-transform math? maybe here just say that this is discussed below, alongside constraints? -->

We use the same dataset as before, but do not distinguish males and females, 
so we can test distinguishability later. Another option here would be to omit
roles, especially if there is no clear role in the data.

```{r prepare-cross-distinguishability}
apim_exchangeable_data <- interdep::prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  set_exchangeable_compositions = "female-male"
)

print(apim_exchangeable_data, n = 4)
```

We then use the columns to fit the model as follows: 

```{r fit-cross-distinguishability}
apim_exchangeable_model <- glmmTMB::glmmTMB(
  satisfaction ~ 
    
    # Pooled single intercept
    1 +
    
    # Pooled single actor and partner effects
    .i_communication_actor +
    .i_communication_partner +
    
    # Residual variance covariance matrix via the shared/difference
    # specification in two uncorrelated blocks
    us(1 | coupleID) +
    us(0 + .i_diff_female_x_male_arbitrary | coupleID),
  dispformula = ~ 0,
  family = gaussian(),
  data = apim_exchangeable_data
)

summary(apim_exchangeable_model)
```


```{r prepare-fitted-exchangeable-apim-diagram, include=FALSE}
apim_exchangeable_diagram_values <- .extract_apim_diagram_values(
  apim_exchangeable_model,
  type = "exchangeable"
)
apim_exchangeable_diagram_estimates <-
  apim_exchangeable_diagram_values$estimates
apim_exchangeable_diagram_residuals <-
  apim_exchangeable_diagram_values$residuals

apim_exchangeable_fitted_alt <- sprintf(
  paste(
    "Fitted exchangeable APIM. Intercept %.2f, actor effect %.2f, partner",
    "effect %.2f, common residual SD %.2f, and residual correlation %.2f."
  ),
  apim_exchangeable_diagram_estimates[["intercept"]],
  apim_exchangeable_diagram_estimates[["actor"]],
  apim_exchangeable_diagram_estimates[["partner"]],
  apim_exchangeable_diagram_residuals[["sd"]],
  apim_exchangeable_diagram_residuals[["correlation"]]
)
```

We use the `interdep::exchangeable_rescov()` to recover the interpretable 
variance-covariance matrix: 

```{r}
backtransformed <- interdep::exchangeable_rescov(apim_exchangeable_model)

# residual variance-covariance matrix
backtransformed$varcov

# residual standard deviation and correlation matrix
backtransformed$sdcor
```


The output can now be mapped as follows:

```{r fitted-exchangeable-apim-diagram, echo=FALSE, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Fitted cross-sectional exchangeable APIM for the example data. The common member residual standard deviation and residual correlation are back-transformed from the fitted mean and difference components.", fig.alt=apim_exchangeable_fitted_alt}
draw_apim_diagram(
  "exchangeable",
  model = apim_exchangeable_model,
  labels = c(predictor = "Communication", outcome = "Satisfaction")
)
```

<!-- TODO: do we need a brief interpretation section? -->

#### Fitted constraints and omitted blocks {#fitted-constraints-and-omitted-blocks}

The function can also recover the member-level implication of constraints that
were imposed when fitting the model. For example, we can omit the entire shared block:

```{r fit-cross-distinguishability}
apim_exchangeable_model_no_shared <- glmmTMB::glmmTMB(
  satisfaction ~  1 +
    .i_communication_actor +
    .i_communication_partner +
    
    # Residual variance covariance matrix
    # omitting the us(1 | coupleID) block
    us(0 + .i_diff_female_x_male_arbitrary | coupleID),
  dispformula = ~ 0,
  family = gaussian(),
  data = apim_exchangeable_data
)

summary(apim_exchangeable_model)
```

<!-- TODO: what does this constraint imply exactly? -->

Since matching is now not automatically possible anymore, we need to tell the
function where both blocks are. Since we completely omitted one block, we tell the 
function:

```{r}
backtransformed <- interdep::exchangeable_rescov(
  apim_exchangeable_model_no_shared, 
  pairs = list(
    shared = NULL,
    difference = "us(0 + .i_diff_female_x_male_arbitrary | coupleID)",
    difference_indicator =".i_diff_female_x_male_arbitrary"
  )
)

# residual variance-covariance matrix
backtransformed$varcov

# residual standard deviation and correlation matrix
backtransformed$sdcor
```

If we fit a model that allows us to also specify higher-level random effect terms,
such as an intensive longitudinal model, we can also omit individual terms from 
any of the two blocks to allow for several restraints. Speficications such 
as `diag(1 + x1 + x2 + x3 | coupleID) + homcs(0 + idiff + x3:idiff | coupleID)` 
are possible, for instance. <!-- TODO: test if it really is -->


### Interpretation

## Testing distinguishability {#testing-distinguishability}

Distinguishability can be evaluated by comparing a full model in which the two
roles may differ with a restricted exchangeable model. This comparison tests
the imposed equality constraints jointly. Here, they concern the fixed
intercepts, actor effects, partner effects, and residual variances.

The two parameterizations require different generated columns. The full
distinguishable model was fitted above, so we now prepare the same original
observations as exchangeable:


[`interdep::compare_interdep_models()`](../reference/compare_interdep_models.html)
verifies that both models use equivalent original
observations before performing the likelihood-ratio test:

```{r compare-cross-distinguishability}
interdep::compare_interdep_models(
  apim_exchangeable_model,
  apim_distinguishable_model
)
```

The test provides evidence against all restrictions jointly, but it does not
show which parameter differs. The helper also cannot determine whether the
models are mathematically nested; that remains a modeling requirement. The
usual chi-squared reference distribution may be unreliable when a tested
variance parameter lies on its boundary.


## Intensive longitudinal APIMs

### Concurrent ILD Gaussian APIM for distinguishable dyads

Observed person means used to construct the between-person (`cbp`) predictors
can be unreliable when each member contributes few occasions, which can bias
between-person estimates [@gottfredsonStraightforwardApproachCoping2019].

Example model specification: 

```{r fit-ild-distinguishable, eval = FALSE}

ild_distinguishable_model <- glmmTMB(
  closeness ~ 0 + 
    
    .i_is_female_x_male_female + 
    .i_is_female_x_male_male + 
    
    # Gender specific time trends
    .i_is_female_x_male_female:diaryday + 
    .i_is_female_x_male_male:diaryday +
    
    # Gender-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +

    # Gender-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    # Gender-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +

    # Gender-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    # random effects for stable non-independence (means)
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID)  +

    # Same-day residual covariance
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID:diaryday) 

  , dispformula = ~ 0  
  , family = gaussian()
  , data = ild_distinguishable_data
)

summary(ild_distinguishable_model)

```


### Concurrent ILD Gaussian APIM for exchangeable dyads

Following del Rosario and West, the stable dyad covariance can be represented
using sum-and-difference random effects
[@delrosarioPracticalGuideSpecifying2025]. In longitudinal Gaussian APIMs fitted
with `glmmTMB`, the same parameterization can be extended to the dyad-occasion
level to represent same-occasion residual dependence.

#### Extension to exchangeable random slopes {#exchangeable-random-slope-back-transformation}

The same shared/difference back-transformation applies to random slopes
[@delrosarioPracticalGuideSpecifying2025]. For example, let
$u_{\mathrm{actor},j}$ denote the shared actor random slope for dyad $j$ and
$\widetilde{u}_{\mathrm{actor},j}$ the corresponding `.i_diff_*` random slope.
The tilde marks random coefficients from the member-difference block. The actor
slopes for the members assigned `+1` and `-1` are

$$
u_{\mathrm{actor},1j}
= u_{\mathrm{actor},j} + \widetilde{u}_{\mathrm{actor},j},
\qquad
u_{\mathrm{actor},2j}
= u_{\mathrm{actor},j} - \widetilde{u}_{\mathrm{actor},j}.
$$

Because the shared and `.i_diff_*` blocks are fitted as separate random-effects
terms, they are independent. Therefore,

$$
\operatorname{Var}(u_{\mathrm{actor},1j})
= \operatorname{Var}(u_{\mathrm{actor},2j})
= \operatorname{Var}(u_{\mathrm{actor},j})
+ \operatorname{Var}(\widetilde{u}_{\mathrm{actor},j}),
$$

and

$$
\operatorname{Cov}(u_{\mathrm{actor},1j}, u_{\mathrm{actor},2j})
= \operatorname{Var}(u_{\mathrm{actor},j})
- \operatorname{Var}(\widetilde{u}_{\mathrm{actor},j}).
$$

The same calculation applies to the partner slopes and random intercepts. Any
covariances among the random intercept, actor slope, and partner slope can be
back-transformed in the same way.


### Current limitations of dyadic ILD designs in R

Concurrent dyadic ILD models can adjust for a time trend and account for stable
dyadic dependence and same-occasion partner dependence. They do not model
residual serial dependence, however, and therefore estimate concurrent
associations under the assumption that residuals from different days are
independent.

If the goal is to retain these concurrent APIM associations while accounting
for serial dependence, the closest extension is a dyadic residual dynamic
structural equation model (RDSEM). It keeps the concurrent regression separate
from a VAR model for its residuals [@asparouhovComparisonModelsAnalysis2020;
@mcneishPrimerTwoLevel2020]. This residual-VAR structure is not directly
available through the `glmmTMB` interface used here, and open-source support for
dyadic dynamic models remains very limited
[@delrosarioPracticalGuideSpecifying2025]. Within open-source R, such a model
generally requires custom TMB or Stan code.

#### Dynamic models {#dynamic-models}

A **practical alternative** is a model with lagged outcomes, especially when
carryover or temporal dynamics are part of the research question
[@gistelinckMultilevelAutoregressiveDyadic2020]. In such a model, the
interpretation changes. All APIM predictor effects then describe associations
conditional on the members' prior outcomes.

By adding the outcome to `predictors` and selecting it with `lag_predictors`,
`interdep::prepare_interdep_data()` returns lag-1 raw and within-person scores alongside
the contemporaneous scores. Between-person scores are not lagged because they
describe stable differences between members.

For this example, we obtain the lagged actor and partner outcome columns through
the `lag_predictors` argument:

```{r prepare-ild-apim-dynamic-model}
ild_apim_data_dynamic <- interdep::prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  time = diaryday,
  predictors = closeness,
  lag_predictors = closeness,
  model_type = "apim",
  seed = 123
)
```

This returns all necessary variables for either choice, including lag-1 raw
and within-person closeness scores. Lags are matched at exactly
`diaryday - 1`, so omitted diary days are not bridged.

A simple fixed-slope dyadic stability and influence model
[@delrosarioPracticalGuideSpecifying2025]:

```{r fit-ild-stability-influence}
stability_influence <- glmmTMB::glmmTMB(
  closeness ~ 1 +

    # Stability (actor effect across time)
    .i_closeness_actor_lag1 +

    # Influence (partner effect across time)
    .i_closeness_partner_lag1 +

    # Linear time trend
    diaryday +

    # Stable exchangeable dyad-level covariance
    us(1 | coupleID) +
    us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID) +

    # Same-day exchangeable dyad-level covariance
    us(1 | coupleID:diaryday) +
    us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_apim_data_dynamic
)

summary(stability_influence)
```

This model can be extended with contemporaneous actor and partner predictor
associations and other lagged predictors. Time-varying predictors should usually
be separated into within-person and between-person components. Their
contemporaneous coefficients are then conditional on both partners' prior
outcomes.

The model above uses raw lagged outcomes. The corresponding
within-person-centered lagged terms are `.i_closeness_cwp_actor_lag1` and
`.i_closeness_cwp_partner_lag1`.

Whether to use a raw or within-person-centered lagged outcome depends on the
research question and the data. Person-mean centering the outcome lag can bias
the average carryover estimate downward [@hamakerCenterNotCenter2015]. This
downward bias is known as Nickell bias [@nickellBiasesDynamicModels1981]. A raw
lag avoids this centering bias, but a standard random-intercept model can still
be biased because it assumes that the lag is unrelated to stable member levels
[@gistelinckMultilevelAutoregressiveSmall2021].

Both problems matter most when there are few occasions. In one simple
simulation, most approaches other than person-mean centering performed
acceptably from about ten occasions, but this is not a universal cutoff
[@gistelinckMultilevelAutoregressiveSmall2021]. For shorter panels, consider an
LD-APIM, which lets the first outcomes relate to the members' stable levels in a
wide-format SEM [@gistelinckMultilevelAutoregressiveDyadic2020].

These manifest-lag models are not equivalent to Mplus DSEM, which uses latent
person-mean centering by default and can estimate multivariate or residual
dynamics jointly [@mcneishPrimerTwoLevel2020]. For very short panels, however,
default DSEM may still be biased or unstable, so the LD-APIM recommendation
above may be preferable [@gistelinckMultilevelAutoregressiveSmall2021].

These lagged-outcome issues are separate from the earlier concern about
unreliable person means used to construct `cbp` predictors with few occasions.


---

From here, choose the model-specific vignette that matches the research
question:

- [Mixed-Composition APIM vignette](mixed-apim.html) for analyses combining
  distinguishable and exchangeable dyad compositions;
- [Dyad-Individual Model vignette](dim.html) for the exchangeable DIM
  parameterization; or
- [Dyadic Score Model vignette](dsm.html) for the distinguishable DSM
  parameterization.

Or return to the [Overview](index.html).

A vignette with non-Gaussian generalized APIM examples is planned. 


## References

::: {#refs}
:::
