---
title: "Dyad-Individual Model"
description: >
  Construction and interpretation of DIM predictors, including their
  relationship to APIM effects in dyadic multilevel models for exchangeable
  dyads.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dyad-Individual Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

This vignette focuses on the Dyad-Individual Model (DIM) for dyadic multilevel
models and its relationship to the Actor-Partner Interdependence Model (APIM)
parameterizations.

For broader guidance on using this package for data preparation for various
dyadic model types, especially APIMs with multiple dyad types, generalized outcomes,
intensive longitudinal models, and optimizer choices, see the
[Getting Started vignette](getting-started.html) and the
[Actor-Partner Interdependence Model vignette](apim.html). For undirected
dyadic score outcomes, see the
[Undirected Dyadic Score Model vignette](undirected-dsm.html).

## Cross-Sectional Gaussian DIM

The current DIM implementation prepares undirected DIM predictors. This means
that the dyad members are treated as exchangeable. One way to achieve this is
to omit `role` from `prepare_interdep_data`, which then assumes all dyads
are the same type of exchangeable dyads.

```{r prepare-cross-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  predictors = communication,
  # Create both APIM and DIM columns for comparison.
  model_type = c("apim", "dim"),
  seed = 123
)

# printing the first couple
print(cross_exchangeable_data, n = 20)
```

Passing a `role` is also possible when it leads to exactly one
exchangeable composition (e.g., only female-female dyads).

Other datasets need a preparation step that produces exactly one exchangeable
composition before DIM columns can be constructed. In a dataset with mixed dyad
types, the right preparation depends on the analysis goal.

For instance, we can filter dyads and keep only male-male and female-female
dyads. Because DIM currently supports a single type of exchangeable dyad, we
would pool those like this:

```{r prepare-cross-filter-pool-same-sex}
cross_same_sex_pooled_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = satisfaction,
  model_type = c("apim", "dim"),
  # removing "male-female" dyads
  include_compositions = c("male-male", "female-female"),
  pool_compositions = list(
    "same-sex-couples" = c("male-male", "female-female")
  ),
  seed = 123
)

print(cross_same_sex_pooled_data, n = 4)
```

If we want to include, for instance, only male-female couples and treat those
as exchangeable, we can do:

```{r prepare-cross-filter-set-male-female-exchangeable}
cross_male_female_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = satisfaction,
  model_type = c("apim", "dim"),
  include_compositions = "male-female",
  set_exchangeable_compositions = "male-female",
  seed = 123
)

print(cross_male_female_exchangeable_data, n = 4)
```

This allows full control over which dyads are included, how they are treated,
and which are pooled.

### Example DIM Model

The variables that enter the DIM in the fixed effects include:

1. A dyad-mean variable that is grand-mean centered. This describes a couple's
shared level compared to all other couples.
2. A within-dyad variable describing each partner's deviation from the couple
mean, equivalently half the signed difference between partners.

The random-effects structure is equivalent to an APIM for exchangeable dyads,
because the APIM and the DIM are just reparametrizations of the same model. The
random effects thus contain a dyad-level intercept and a within-dyad deviation
indexed by `.i_diff_assumed_exchangeable`. In glmmTMB, with `dispformula = ~ 0`,
these dyad-level random effects model the Gaussian residual covariance.

```{r fit-cross-dim, eval=has_glmmTMB}

dim_1 <- glmmTMB::glmmTMB(
  satisfaction ~

    # Pooled fixed intercept
    1 +

    # Between-couple effect
    .i_communication_raw_dyad_mean_gmc +

    # Within-couple effect
    .i_communication_raw_within_dyad_deviation +

    # Residual Gaussian variance-covariance
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable | coupleID)
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

summary(dim_1)
```

```{r glmmTMB-missing, eval=!has_glmmTMB}
message("Install glmmTMB to run the fitted-model examples in this vignette.")
```

The same model can be written in APIM form. Since we have requested both sets of
variables from `prepare_interdep_data`, we can fit one directly. For more
guidance on APIM specifications and different models, see the
[Actor-Partner Interdependence Model vignette](apim.html).

```{r fit-cross-apim, eval=has_glmmTMB}

apim_1 <- glmmTMB::glmmTMB(
  satisfaction ~ 1 +
    .i_communication_raw_actor + .i_communication_raw_partner +
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable | coupleID)
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

summary(apim_1)
```

The two models have identical fit statistics:

```{r compare-cross-fit, eval=has_glmmTMB}
data.frame(
  model = c("DIM", "APIM"),
  AIC = c(AIC(dim_1), AIC(apim_1)),
  BIC = c(BIC(dim_1), BIC(apim_1)),
  logLik = c(as.numeric(logLik(dim_1)), as.numeric(logLik(apim_1)))
)
```

This demonstrates that the same statistical model is being estimated with different
parameterizations and coefficient interpretations.

The coefficients relate as follows:

\[
\beta_{\text{dyad mean}} =
\beta_{\text{actor}} + \beta_{\text{partner}}
\]

and

\[
\beta_{\text{within-dyad deviation}} =
\beta_{\text{actor}} - \beta_{\text{partner}}
\]

In this example:

```{r compare-cross-coefficients, eval=has_glmmTMB}
apim_coef <- glmmTMB::fixef(apim_1)$cond
dim_coef <- glmmTMB::fixef(dim_1)$cond

b_actor <- round(apim_coef[[".i_communication_raw_actor"]], 3)
b_partner <- round(apim_coef[[".i_communication_raw_partner"]], 3)

b_dyad_mean <- round(dim_coef[[".i_communication_raw_dyad_mean_gmc"]], 3)
b_within_dyad <- round(dim_coef[[".i_communication_raw_within_dyad_deviation"]], 3)

cat("From APIM model:\n")
cat("  actor effect:                  ",  b_actor, "\n")
cat("  partner effect:                ", b_partner, "\n\n")

cat("  DIM transformation:\n")
cat("  actor effect + partner effect: ", b_actor + b_partner, "\n")
cat("  actor effect - partner effect: ", b_actor - b_partner, "\n\n")

cat("From DIM model:\n")
cat("  dyad-mean effect:              ", b_dyad_mean, "\n")
cat("  within-dyad-deviation effect:  ", b_within_dyad, "\n")
```

## Intensive longitudinal DIM

For longitudinal DIM, predictors are decomposed into within-person and
between-person components with `time_2l`.
`temporal_predictor_decomposition = "none"` is currently rejected for DIM and
undirected DSM longitudinal predictor construction.

```{r prepare-ild-exchangeable}
ild_exchangeable_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  time = diaryday,
  predictors = provided_support,
  model_type = c("apim", "dim"),
  seed = 123
)

print(ild_exchangeable_data)
```

```{r fit-ild-dim, eval = has_glmmTMB}

dim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person DIM
    .i_provided_support_cwp_dyad_mean +
    .i_provided_support_cwp_within_dyad_deviation +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_deviation +

    # Stable exchangeable dyad-level covariance
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable | coupleID) +

    # Residual (same-day) exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

summary(dim_ILD)

```

The equivalent APIM uses actor and partner effects on both levels:

```{r fit-ild-apim, eval = has_glmmTMB}

apim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person APIM
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # Stable exchangeable dyad-level covariance
    (1 | coupleID)  + (0 + .i_diff_assumed_exchangeable | coupleID) +

    # Residual (same-day) exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

summary(apim_ILD)

```

The two ILD models again have identical fit statistics:

```{r compare-ild-fit, eval = has_glmmTMB}
data.frame(
  model = c("DIM", "APIM"),
  AIC = c(AIC(dim_ILD), AIC(apim_ILD)),
  BIC = c(BIC(dim_ILD), BIC(apim_ILD)),
  logLik = c(as.numeric(logLik(dim_ILD)), as.numeric(logLik(apim_ILD)))
)
```

The equivalence holds separately for the within-person (`cwp`) and
between-person (`cbp`) predictor components. For the within-person component:

\[
\beta_{\text{cwp dyad mean}} =
\beta_{\text{cwp actor}} + \beta_{\text{cwp partner}}
\]

\[
\beta_{\text{cwp within-dyad deviation}} =
\beta_{\text{cwp actor}} - \beta_{\text{cwp partner}}
\]

For the between-person component:

\[
\beta_{\text{cbp dyad mean}} =
\beta_{\text{cbp actor}} + \beta_{\text{cbp partner}}
\]

\[
\beta_{\text{cbp within-dyad deviation}} =
\beta_{\text{cbp actor}} - \beta_{\text{cbp partner}}
\]

This also means that an APIM parameterization can be used on one level and a
DIM parameterization on the other. For example:

```{r fit-ild-apim-within-dim-between, eval = has_glmmTMB}

apim_dim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_deviation +

    # Stable exchangeable dyad-level covariance
    (1 | coupleID)  + (0 + .i_diff_assumed_exchangeable | coupleID) +

    # Same-day exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

summary(apim_dim_ILD)

```

This mixed parameterization still estimates the same model:

```{r compare-ild-all-fit, eval = has_glmmTMB}
data.frame(
  model = c("DIM within / DIM between", "APIM within / APIM between", "APIM within / DIM between"),
  AIC = c(AIC(dim_ILD), AIC(apim_ILD), AIC(apim_dim_ILD)),
  BIC = c(BIC(dim_ILD), BIC(apim_ILD), BIC(apim_dim_ILD)),
  logLik = c(
    as.numeric(logLik(dim_ILD)),
    as.numeric(logLik(apim_ILD)),
    as.numeric(logLik(apim_dim_ILD))
  )
)
```

## Including random slopes

Random-slope DIM and APIM parameterizations can be written analogously by adding
the corresponding within-person effects to the stable dyad-level random-effect
blocks. In the example data below these larger models do not converge cleanly,
so the chunks are not evaluated and should absolutely **not be interpreted**.
This structure can be reasonable if convergence diagnostics are clean.

In the APIM:

```{r fit-ild-apim-random, eval = FALSE}

apim_ILD_random <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person APIM
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # Stable dyad-level covariance with within-person random slopes
    (1 +
         .i_provided_support_cwp_actor +
         .i_provided_support_cwp_partner
       | coupleID)  +
    (0 +
         .i_diff_assumed_exchangeable +
         .i_diff_assumed_exchangeable:.i_provided_support_cwp_actor +
         .i_diff_assumed_exchangeable:.i_provided_support_cwp_partner
       | coupleID) +

    # Same-day exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

In the DIM:

```{r fit-ild-dim-random, eval = FALSE}

dim_ILD_random <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person DIM
    .i_provided_support_cwp_dyad_mean +
    .i_provided_support_cwp_within_dyad_deviation +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_deviation +

    # Stable dyad-level covariance with within-person random slopes
    (1 +
       .i_provided_support_cwp_dyad_mean +
       .i_provided_support_cwp_within_dyad_deviation
     | coupleID)  +
    (0 +
       .i_diff_assumed_exchangeable +
       .i_diff_assumed_exchangeable:.i_provided_support_cwp_dyad_mean +
       .i_diff_assumed_exchangeable:.i_provided_support_cwp_within_dyad_deviation
     | coupleID) +

    # Same-day exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

For APIM formulas with distinguishable, exchangeable, generalized, and
mixed-composition dyads, see the
[Actor-Partner Interdependence Model vignette](apim.html). For undirected dyadic
score outcomes, see the
[Undirected Dyadic Score Model vignette](undirected-dsm.html). For the broader
data-preparation workflow, return to the
[Getting Started vignette](getting-started.html).
