---
title: "Dyad-Individual Model"
author: "Pascal Küng"
description: >
  Construction and interpretation of DIM predictors, including their
  relationship to APIM effects in dyadic multilevel models for exchangeable
  dyads.
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dyad-Individual Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

diagram_helper <- if (file.exists("diagram-helpers.Rinc")) {
  "diagram-helpers.Rinc"
} else {
  file.path("vignettes", "diagram-helpers.Rinc")
}
sys.source(diagram_helper, envir = knitr::knit_global())
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
dim_fitted_alt <- "Fitted DIM diagram unavailable."
```

This vignette focuses on the Dyad-Individual Model (DIM) for dyadic multilevel
models and its relationship to the exchangeable Actor-Partner Interdependence
Model (APIM). The DIM separates a predictor into the dyad's shared level and
each member's deviation from that level. While the APIM expresses effects in
terms of two interdependent individuals, the DIM expresses them in terms of the
dyad's shared level and the contrast between partners.

Under exchangeability constraints, the reduced, label-invariant DSM is also
equivalent to the DIM, as discussed below. For the broader package workflow and
an overview of the available model-specific vignettes, including the
[Actor-Partner Interdependence Model](apim.html),
[Mixed-Composition APIM](mixed-apim.html), and
[Dyadic Score Model](dsm.html), see the [Overview](index.html).

## Cross-Sectional Gaussian DIM

The current DIM implementation needs one exchangeable dyad composition.
Exchangeability means that swapping the two member labels does not
change the model [@kenny2006dyadic]. Whether roles can and should be treated as
exchangeable is a substantive assumption (see [Testing distinguishability in the APIM vignette](apim.html#testing-distinguishability)).

One way to make this assumption in `interdep` at the data-preparation step is
to omit `role` from `prepare_interdep_data()`. This treats all dyads as the
same exchangeable composition. Passing a `role` is also possible when it leads
to exactly one exchangeable composition (e.g., only female-female dyads).
Otherwise, refer to the [Getting Started vignette](getting-started.html) for how
to filter, pool, and constrain dyad compositions to obtain a single exchangeable
dyad composition.

```{r prepare-cross-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  predictors = communication,
  # Create both APIM and DIM columns for comparison.
  model_type = c("apim", "dim"),
  seed = 123
)

# Print the first two dyads.
print(cross_exchangeable_data, n = 4)
```

For the exchangeable random-effects specification,
`prepare_interdep_data()` creates a member-difference contrast `.i_diff_*`,
coded as `+1` for one partner and `-1` for the other. Because these member labels
are arbitrary, setting `seed` makes their assignment reproducible.

### Example DIM Model

For member \(i \in \{1, 2\}\) of dyad \(j\), define the predictor mean and
member deviation as:

$$
\bar{x}_j = \frac{x_{1j} + x_{2j}}{2}, \qquad
x_{\mathrm{dev},ij} = x_{ij} - \bar{x}_j.
$$

The deviations of the two partners have equal magnitude and opposite signs:
\(x_{\mathrm{dev},1j} = -x_{\mathrm{dev},2j}\). Outcome means and deviations
are defined analogously. The variables that enter the DIM fixed effects then
separate two associations:

1. The **between-dyad effect**, $b_{\mathrm{mean}}$: whether dyads with a
higher average predictor value than other dyads also have a higher average
outcome value.
2. The **within-dyad effect**, $b_{\mathrm{dev}}$: whether the member who is
above the dyad's predictor mean is also above the dyad's outcome mean. With
two members, each deviation is half the corresponding signed partner difference.

The first diagram emphasizes this between- and within-dyad decomposition. Its
upper row describes differences between dyads. Its lower row describes the
corresponding deviations within a dyad for either member.
Switching members reverses both deviations and therefore leaves the pooled
$b_{\mathrm{dev}}$ unchanged.

```{r conceptual-dim-diagram, echo=FALSE, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Between- and within-dyad effects representation of the cross-sectional DIM. The uncorrelated $r_m$ and $r_{d,i}$ in the conceptual representation do not imply that the member residuals are independent, instead, both members information is contained in each of these residuals.", fig.alt="Path diagram separating the between- and within-dyad associations in a cross-sectional Dyad-Individual Model. The centered predictor mean predicts the outcome mean with coefficient b mean. Member i's predictor deviation predicts the same member's outcome deviation with coefficient b dev. There are no cross-paths, and only the outcome-mean equation has intercept b zero."}
draw_dim_diagram()
```

Uncorrelated $r_m$ and $r_{d,i}$ in the conceptual representation do not imply
that the member residuals are independent

The second diagram translates the same decomposition to the individual-member
rows used by the long-format multilevel model. Each member's outcome is predicted
by the dyad's shared predictor mean and that member's own predictor deviation
from that shared mean. Both members share the same two coefficients, so estimation
pools information across members under the exchangeability assumptions.

```{r conceptual-dim-member-diagram, echo=FALSE, fig.width=9, fig.height=5.2, out.width="100%", fig.cap="Individual-level representation of the cross-sectional DIM used for the long-format multilevel model.", fig.alt="Path diagram for two arbitrarily labelled members of an exchangeable dyad. For each member, the shared, grand-mean-centered dyad predictor mean and the member's own predictor deviation predict the individual outcome. Both members share the same between- and within-dyad coefficients and residual standard deviation, while their outcome residuals are correlated."}
draw_dim_member_diagram()
```

The resulting estimated fixed effects are a reparameterization of the APIM actor
and partner effects [@Bolger2025Unified]. And just like the exchangeable APIM,
the random-effects structure comprises a dyad-level intercept
and a dyad-level difference contrast indexed by `.i_diff_assumed_exchangeable_arbitrary`.
In `glmmTMB`, with `dispformula = ~ 0`, these random effects represent the two
members' Gaussian residual variance and covariance.

The intercept and difference contrast are specified as separate random-effects
terms, which constrains their correlation to zero, preserving exchangeability
[@delrosarioPracticalGuideSpecifying2025].

The full model can be estimated as:

```{r fit-cross-dim, eval=has_glmmTMB}

dim_1 <- glmmTMB::glmmTMB(
  satisfaction ~

    # Pooled fixed intercept
    1 +

    # Between-couple effect
    .i_communication_dyad_mean_gmc +

    # Member-deviation effect
    .i_communication_within_dyad_dev +

    # Residual Gaussian covariance structure
    us(1 | coupleID) +
    us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID)
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

summary(dim_1)
```

The same mean-and-deviation diagram can now be labelled with the estimated
fixed effects and residual-component standard deviations:

```{r prepare-fitted-dim-diagram, include=FALSE, eval=has_glmmTMB}
dim_fixed_effects <- glmmTMB::fixef(dim_1)$cond
dim_random_components <- glmmTMB::VarCorr(dim_1)$cond

dim_diagram_estimates <- c(
  b0 = dim_fixed_effects[["(Intercept)"]],
  b_mean = dim_fixed_effects[[".i_communication_dyad_mean_gmc"]],
  b_dev = dim_fixed_effects[[".i_communication_within_dyad_dev"]]
)
dim_diagram_residuals <- c(
  sd_mean = .extract_diagram_random_sd(
    dim_random_components,
    "(Intercept)"
  ),
  sd_difference = .extract_diagram_random_sd(
    dim_random_components,
    ".i_diff_assumed_exchangeable_arbitrary"
  )
)

dim_fitted_alt <- sprintf(
  paste(
    "Fitted DIM. Intercept %.2f, between-couple effect %.2f, within-couple",
    "effect %.2f, and mean/deviation residual SDs %.2f and %.2f; their",
    "correlation is fixed at zero."
  ),
  dim_diagram_estimates[["b0"]],
  dim_diagram_estimates[["b_mean"]],
  dim_diagram_estimates[["b_dev"]],
  dim_diagram_residuals[["sd_mean"]],
  dim_diagram_residuals[["sd_difference"]]
)
```

```{r fitted-dim-diagram, echo=FALSE, eval=has_glmmTMB, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Estimated fixed effects and residual-component standard deviations from the cross-sectional Gaussian DIM in its mean-and-deviation representation. The intercept belongs to the outcome-mean equation.", fig.alt=dim_fitted_alt}
draw_dim_diagram(
  estimates = dim_diagram_estimates,
  residuals = dim_diagram_residuals,
  effect_unit = "couple"
)
```

Under these exchangeability constraints, the Gaussian DIM is algebraically
equivalent to the reduced, label-invariant Dyadic Score Model (DSM)
[@iidaModelsInterdependentIndividuals2018]. Compared to a full DSM, the DIM's
exchangeability constraints fix the outcome-deviation intercept and both
cross-paths to zero and constrain the mean and deviation residual components to
be uncorrelated (see the diagrams here and compare them with the
[conceptual DSM diagrams](dsm.html#cross-sectional-gaussian-dsm)).

### Model interpretation

Therefore, each coefficient has both an individual-member interpretation and an
equivalent couple mean/difference interpretation.

In this Gaussian model, fixed coefficients are
interpreted in units of the outcome, e.g., "satisfaction":

- The intercept (about 5.04) is the expected satisfaction of either member,
  and therefore the expected couple-average satisfaction, when both members'
  communication equals the sample grand mean.

- The dyad-mean estimate (about 2.00) means that, comparing couples with the
  same communication difference between partners, a one-point higher
  couple-average communication level is associated with a 2.00-point higher
  expected couple-average satisfaction. Equivalently, each member's expected
  satisfaction is 2.00 points higher.

- The member-deviation estimate (about 1.52) means that a one-point
  difference in communication between partners is associated with a 1.52-point
  difference in their expected satisfaction, holding their average communication
  constant. In member terms, suppose one member is 0.5 points above the dyad
  mean and the other is 0.5 points below it. Their expected satisfaction is then
  0.76 points above and below the couple's predicted mean, respectively, so they
  are expected to differ by 1.52 points in satisfaction.

Because the Gaussian DIM is the exchangeability-constrained version of the full
DSM, exchangeability can also be tested by comparing these nested models. This
is equivalent to the comparison shown in [Testing distinguishability in the APIM vignette](apim.html#testing-distinguishability).

### Demonstrating model equivalence to APIM

The same model can be written in APIM form. Since we have requested both sets of
variables from `prepare_interdep_data()`, we can fit one directly. For more
guidance on APIM specifications and different models, see the
[Actor-Partner Interdependence Model vignette](apim.html).

```{r fit-cross-apim, eval = has_glmmTMB}

apim_1 <- glmmTMB::glmmTMB(
  satisfaction ~ 1 +

    # Fixed effects APIM
    .i_communication_actor + .i_communication_partner +

    # Since both models are equivalent, the same random-effects structure
    # can be used. See the APIM vignette to learn how to back-transform
    # these blocks to a full actor-partner covariance matrix.
    us(1 | coupleID) +
    us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID)
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

```

The two models have identical fit statistics:

```{r compare-cross-fit, eval=has_glmmTMB}
data.frame(
  model = c("DIM", "APIM"),
  AIC = c(AIC(dim_1), AIC(apim_1)),
  BIC = c(BIC(dim_1), BIC(apim_1)),
  logLik = c(as.numeric(logLik(dim_1)), as.numeric(logLik(apim_1)))
)
```

This demonstrates that the same statistical model is being estimated with different
parameterizations and coefficient interpretations.

Once APIM estimates are present, one can easily obtain DIM estimates, and
the other way around. Let $b_{\mathrm{actor}}$ and $b_{\mathrm{partner}}$
denote the APIM actor and partner slopes, and let $b_{\mathrm{mean}}$ and
$b_{\mathrm{dev}}$ denote the DIM dyad-mean and member-deviation slopes.
They relate as follows:

$$
b_{\mathrm{mean}} = b_{\mathrm{actor}} + b_{\mathrm{partner}}
$$

and

$$
b_{\mathrm{dev}} = b_{\mathrm{actor}} - b_{\mathrm{partner}}
$$

Conversely:

$$
b_{\mathrm{actor}}
= \frac{b_{\mathrm{mean}} + b_{\mathrm{dev}}}{2}
$$

and

$$
b_{\mathrm{partner}}
= \frac{b_{\mathrm{mean}} - b_{\mathrm{dev}}}{2}
$$


In this example we can see that the transformations work:

```{r compare-cross-coefficients, echo = TRUE, eval = has_glmmTMB}
apim_coef <- glmmTMB::fixef(apim_1)$cond
dim_coef <- glmmTMB::fixef(dim_1)$cond

b_actor <- apim_coef[[".i_communication_actor"]]
b_partner <- apim_coef[[".i_communication_partner"]]

b_mean <- dim_coef[[".i_communication_dyad_mean_gmc"]]
b_dev <- dim_coef[[".i_communication_within_dyad_dev"]]


cat("From APIM model:\n",
     "  actor effect:                  ", round(b_actor, 3), "\n",
     "  partner effect:                ", round(b_partner, 3), "\n\n",

     "DIM transformation:\n",
     "  b_mean = b_actor + b_partner:  ", round(b_actor + b_partner, 3), "\n",
     "  b_dev = b_actor - b_partner:   ", round(b_actor - b_partner, 3), "\n\n",

     "From DIM model:\n",
     "  dyad-mean effect:              ", round(b_mean, 3), "\n",
     "  member-deviation effect:       ", round(b_dev, 3), "\n"
)

```

```{r prepare-dim-grid-values, include=FALSE}
dim_grid_slopes <- c(
  actor = NA_real_,
  partner = NA_real_,
  mean = NA_real_,
  within = NA_real_
)

if (has_glmmTMB) {
  dim_grid_slopes[] <- unname(c(b_actor, b_partner, b_mean, b_dev))
}

dim_grid_js_number <- function(name) {
  value <- dim_grid_slopes[[name]]
  if (!is.finite(value)) {
    return("null")
  }
  formatC(value, digits = 17, format = "g", decimal.mark = ".")
}
```

The DIM and APIM intercepts are not expected to be equal because the DIM dyad
mean is grand-mean centered, whereas the APIM predictors retain their original
scale.


### Why Are These Models Equivalent? Exploring the Reparameterization

An intuitive way to think about this is:

- When the dyad mean goes up by 1 unit while the difference between partners
  remains stable, both partners' values must go up by 1. Both the actor and
  partner effects therefore contribute, which is why the dyad-mean effect is
  the actor effect + the partner effect.

- When a person's deviation from the dyad mean goes up by 1 unit while the dyad
  mean remains constant, the other partner's value must go **down** by 1 unit.
  The actor value therefore changes by +1 and the partner value by -1, which is
  why the member-deviation effect is the actor effect - the partner effect.

The grid below shows the same predictor values in both coordinate systems. The
horizontal and vertical axes are actor and partner values centered at the
sample grand mean. The diagonal axes are their dyad mean and within-dyad
deviation.

The displayed slope values are read directly from the fitted APIM and DIM
models above. They illustrate that the two forms make the same change in the
linear predictor relative to the grand-mean reference. The intercept is omitted
from both displayed equations.

```{=html}
<style>
.interdep-dim-grid {
  --idg-actor: #1769aa;
  --idg-partner: #7b3fa1;
  --idg-mean: #c43c35;
  --idg-within: #8a6500;
  --idg-border: var(--bs-border-color, #c9ced6);
  --idg-surface: var(--bs-tertiary-bg, #f5f6f8);
  --idg-ink: var(--bs-body-color, #202428);
  box-sizing: border-box;
  display: grid;
  grid-template-columns: minmax(15rem, 19rem) minmax(0, 1fr);
  gap: 1.25rem;
  margin: 1.25rem 0 2rem;
  padding: 1rem;
  color: var(--idg-ink);
  background: var(--idg-surface);
  border: 1px solid var(--idg-border);
  border-radius: 8px;
}
.interdep-dim-grid *,
.interdep-dim-grid *::before,
.interdep-dim-grid *::after {
  box-sizing: border-box;
}
.idg-controls,
.idg-visual,
.idg-control-group {
  display: grid;
  align-content: start;
  gap: 0.9rem;
  min-width: 0;
}
.idg-toolbar,
.idg-group-title,
.idg-control-head,
.idg-demo-row {
  display: flex;
  align-items: center;
  justify-content: space-between;
  gap: 0.75rem;
}
.idg-toolbar {
  padding-bottom: 0.7rem;
  border-bottom: 1px solid var(--idg-border);
}
.idg-toolbar strong,
.idg-group-title {
  font-size: 0.82rem;
  font-weight: 700;
}
.idg-reset,
.idg-demo {
  padding: 0.35rem 0.6rem;
  color: inherit;
  background: var(--bs-body-bg, #fff);
  border: 1px solid var(--idg-border);
  border-radius: 5px;
  font: inherit;
  font-size: 0.78rem;
  line-height: 1.2;
  cursor: pointer;
}
.idg-reset:hover,
.idg-demo:hover {
  border-color: currentColor;
}
.idg-reset:focus-visible,
.idg-demo:focus-visible,
.interdep-dim-grid input:focus-visible {
  outline: 3px solid rgba(23, 105, 170, 0.35);
  outline-offset: 2px;
}
.idg-group-title {
  justify-content: flex-start;
  padding-bottom: 0.35rem;
  border-bottom: 1px solid var(--idg-border);
}
.idg-swatches {
  display: inline-flex;
  gap: 0.25rem;
}
.idg-swatch {
  width: 0.65rem;
  height: 0.65rem;
  border-radius: 50%;
}
.idg-control {
  display: grid;
  gap: 0.25rem;
}
.idg-control label {
  min-width: 0;
  font-size: 0.84rem;
  font-weight: 600;
}
.idg-control output {
  flex: 0 0 5.5ch;
  text-align: right;
  font-variant-numeric: tabular-nums;
  font-weight: 700;
}
.idg-control input[type="range"] {
  width: 100%;
  min-height: 1.75rem;
  margin: 0;
  cursor: pointer;
}
.idg-actor input { accent-color: var(--idg-actor); }
.idg-partner input { accent-color: var(--idg-partner); }
.idg-mean input { accent-color: var(--idg-mean); }
.idg-within input { accent-color: var(--idg-within); }
.idg-relations {
  padding-top: 0.7rem;
  border-top: 1px solid var(--idg-border);
  font-size: 0.8rem;
  line-height: 1.6;
}
.idg-equations {
  display: grid;
  gap: 0.4rem;
  min-width: 0;
  width: 100%;
  max-width: 100%;
  padding: 0.75rem;
  background: var(--bs-body-bg, #fff);
  border: 1px solid var(--idg-border);
  border-radius: 6px;
  overflow-x: auto;
  font-family: Georgia, "Times New Roman", serif;
  font-size: 0.94rem;
  font-variant-numeric: tabular-nums;
  white-space: nowrap;
}
.idg-equations > div:last-child {
  white-space: normal;
}
.idg-equation-label {
  display: inline-block;
  width: 3.3rem;
  font-family: var(--bs-body-font-family, sans-serif);
  font-size: 0.7rem;
  font-weight: 700;
  text-transform: uppercase;
}
.idg-equation-number {
  display: inline-block;
  text-align: right;
}
.idg-equation-number--apim {
  width: 4ch;
}
.idg-equation-number--dim {
  width: 5ch;
}
.idg-equation-number--result {
  width: 6ch;
}
.idg-stage {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(min(100%, 16rem), 1fr));
  gap: 1rem;
  align-items: start;
}
.idg-plot {
  display: block;
  width: 100%;
  max-width: 29rem;
  aspect-ratio: 1;
  margin: 0 auto;
  background: var(--bs-body-bg, #fff);
  border: 1px solid var(--idg-border);
  border-radius: 6px;
  cursor: crosshair;
  touch-action: none;
  user-select: none;
}
.idg-explanation {
  display: grid;
  gap: 1rem;
  font-size: 0.82rem;
  line-height: 1.45;
}
.idg-explanation p {
  margin: 0.3rem 0 0;
}
.idg-demo-row {
  align-items: flex-start;
}
.idg-demo-row strong {
  display: block;
}
.idg-mini-equation {
  margin-top: 0.4rem;
  font-family: Georgia, "Times New Roman", serif;
  font-size: 0.86rem;
}
.idg-result {
  font-weight: 700;
}
.idg-halo {
  fill: var(--idg-actor);
  fill-opacity: 0.14;
}
.idg-point {
  fill: var(--idg-actor);
  stroke: var(--bs-body-bg, #fff);
  stroke-width: 2;
}
.idg-sr-only {
  position: absolute;
  width: 1px;
  height: 1px;
  padding: 0;
  margin: -1px;
  overflow: hidden;
  clip: rect(0, 0, 0, 0);
  white-space: nowrap;
  border: 0;
}
@media (max-width: 860px) {
  .interdep-dim-grid {
    grid-template-columns: 1fr;
  }
  .idg-controls {
    grid-template-columns: repeat(2, minmax(0, 1fr));
  }
  .idg-toolbar,
  .idg-relations {
    grid-column: 1 / -1;
  }
}
@media (max-width: 620px) {
  .interdep-dim-grid {
    margin-right: -0.35rem;
    margin-left: -0.35rem;
    padding: 0.75rem;
  }
  .idg-controls,
  .idg-stage {
    grid-template-columns: 1fr;
  }
  .idg-toolbar,
  .idg-relations {
    grid-column: auto;
  }
  .idg-equations {
    font-size: 0.84rem;
  }
}
@media (prefers-reduced-motion: reduce) {
  .interdep-dim-grid * {
    scroll-behavior: auto !important;
  }
}
[data-bs-theme="dark"] .interdep-dim-grid {
  --idg-actor: #75b7ff;
  --idg-partner: #d0a6ff;
  --idg-mean: #ff8c82;
  --idg-within: #f0c75e;
}
</style>

<div class="interdep-dim-grid" id="interdep-dim-grid">
  <div class="idg-controls">
    <div class="idg-toolbar">
      <strong>Predictor coordinates</strong>
      <button class="idg-reset" type="button" data-idg-reset>Reset</button>
    </div>

    <div class="idg-control-group">
      <div class="idg-group-title">
        <span class="idg-swatches" aria-hidden="true">
          <span class="idg-swatch" style="background: var(--idg-actor)"></span>
          <span class="idg-swatch" style="background: var(--idg-partner)"></span>
        </span>
        APIM coordinates
      </div>
      <div class="idg-control idg-actor">
        <div class="idg-control-head">
          <label for="idg-actor">Grand-mean-centered actor, <i>x</i><sub>actor</sub></label>
          <output for="idg-actor" data-idg-output="actor">0.0</output>
        </div>
        <input id="idg-actor" data-idg-input="actor" type="range"
               min="-5" max="5" step="0.1" value="0">
      </div>
      <div class="idg-control idg-partner">
        <div class="idg-control-head">
          <label for="idg-partner">Grand-mean-centered partner, <i>x</i><sub>partner</sub></label>
          <output for="idg-partner" data-idg-output="partner">0.0</output>
        </div>
        <input id="idg-partner" data-idg-input="partner" type="range"
               min="-5" max="5" step="0.1" value="0">
      </div>
    </div>

    <div class="idg-control-group">
      <div class="idg-group-title">
        <span class="idg-swatches" aria-hidden="true">
          <span class="idg-swatch" style="background: var(--idg-mean)"></span>
          <span class="idg-swatch" style="background: var(--idg-within)"></span>
        </span>
        DIM coordinates
      </div>
      <div class="idg-control idg-mean">
        <div class="idg-control-head">
          <label for="idg-mean">Dyad mean, <i>x</i><sub>mean</sub></label>
          <output for="idg-mean" data-idg-output="mean">0.0</output>
        </div>
        <input id="idg-mean" data-idg-input="mean" type="range"
               min="-5" max="5" step="0.05" value="0">
      </div>
      <div class="idg-control idg-within">
        <div class="idg-control-head">
          <label for="idg-within">Member deviation, <i>x</i><sub>dev</sub></label>
          <output for="idg-within" data-idg-output="within">0.0</output>
        </div>
        <input id="idg-within" data-idg-input="within" type="range"
               min="-5" max="5" step="0.05" value="0">
      </div>
    </div>

    <div class="idg-relations">
      <div><i>x</i><sub>mean</sub> = (<i>x</i><sub>actor</sub> + <i>x</i><sub>partner</sub>) / 2</div>
      <div><i>x</i><sub>dev</sub> = (<i>x</i><sub>actor</sub> - <i>x</i><sub>partner</sub>) / 2</div>
    </div>
  </div>

  <div class="idg-visual">
    <div class="idg-equations">
      <div><span class="idg-equation-label">APIM</span><span data-idg-equation="apim">0.00 = 0.00</span></div>
      <div><span class="idg-equation-label">DIM</span><span data-idg-equation="dim">0.00 = 0.00</span></div>
      <div><span class="idg-equation-label">Slopes</span><span data-idg-slopes>Loading fitted slopes...</span></div>
    </div>

    <div class="idg-stage">
      <svg class="idg-plot" data-idg-plot viewBox="0 0 320 320"
           role="img" aria-labelledby="idg-plot-title idg-plot-description">
        <title id="idg-plot-title">APIM and DIM coordinate grid</title>
        <desc id="idg-plot-description">The selected point has actor, partner, dyad-mean, and within-dyad values of zero.</desc>
        <defs>
          <clipPath id="idg-plot-clip">
            <rect x="30" y="30" width="260" height="260" rx="4"></rect>
          </clipPath>
        </defs>
        <rect x="30" y="30" width="260" height="260" rx="4"
              fill="none" stroke="currentColor" stroke-opacity="0.35"></rect>
        <g data-idg-grid-lines clip-path="url(#idg-plot-clip)"></g>
        <g data-idg-axis-labels></g>
        <circle class="idg-halo" data-idg-halo cx="160" cy="160" r="13"></circle>
        <circle class="idg-point" data-idg-point cx="160" cy="160" r="6"></circle>
      </svg>

      <div class="idg-explanation">
        <div>
          <div class="idg-demo-row">
            <strong>Shared dyad level</strong>
            <button class="idg-demo" type="button" data-idg-demo="mean">Show</button>
          </div>
          <p>When both members are one point higher, the dyad mean is one point higher and the deviation is unchanged.</p>
          <div class="idg-mini-equation"><i>b</i><sub>mean</sub> = <i>b</i><sub>actor</sub> + <i>b</i><sub>partner</sub></div>
        </div>
        <div>
          <div class="idg-demo-row">
            <strong>Member deviation</strong>
            <button class="idg-demo" type="button" data-idg-demo="within">Show</button>
          </div>
          <p>When one member is one point higher and the other one point lower, the dyad mean is unchanged and the deviation is one point.</p>
          <div class="idg-mini-equation"><i>b</i><sub>dev</sub> = <i>b</i><sub>actor</sub> - <i>b</i><sub>partner</sub></div>
        </div>
        <p>The dot uses APIM coordinates. The diagonal lines show the same point in DIM coordinates.</p>
      </div>
    </div>
  </div>
  <noscript><p>Enable JavaScript to manipulate this figure. The equations and discussion above provide the same transformation.</p></noscript>
</div>

<script>
(function () {
  "use strict";

  var root = document.getElementById("interdep-dim-grid");
  if (!root || root.dataset.initialized === "true") return;
  root.dataset.initialized = "true";

  var limit = 5;
  var plotScale = 130 / limit;
  var slopes = {
    actor: `r dim_grid_js_number("actor")`,
    partner: `r dim_grid_js_number("partner")`,
    mean: `r dim_grid_js_number("mean")`,
    within: `r dim_grid_js_number("within")`
  };
  var colors = {
    actor: "var(--idg-actor)",
    partner: "var(--idg-partner)",
    mean: "var(--idg-mean)",
    within: "var(--idg-within)"
  };
  var inputs = {};
  var outputs = {};
  ["actor", "partner", "mean", "within"].forEach(function (name) {
    inputs[name] = root.querySelector('[data-idg-input="' + name + '"]');
    outputs[name] = root.querySelector('[data-idg-output="' + name + '"]');
  });

  var plot = root.querySelector("[data-idg-plot]");
  var point = root.querySelector("[data-idg-point]");
  var halo = root.querySelector("[data-idg-halo]");
  var description = root.querySelector("#idg-plot-description");
  var equationAPIM = root.querySelector('[data-idg-equation="apim"]');
  var equationDIM = root.querySelector('[data-idg-equation="dim"]');
  var slopeSummary = root.querySelector("[data-idg-slopes]");

  var hasFittedSlopes = [
    slopes.actor,
    slopes.partner,
    slopes.mean,
    slopes.within
  ].every(Number.isFinite);

  if (!hasFittedSlopes) {
    equationAPIM.textContent = "Fitted slopes unavailable";
    equationDIM.textContent = "Fitted slopes unavailable";
    slopeSummary.textContent =
      "Install glmmTMB and refit the models to activate this figure.";
    root.querySelectorAll("input, button").forEach(function (control) {
      control.disabled = true;
    });
    description.textContent =
      "The coordinate grid is inactive because fitted slopes are unavailable.";
    return;
  }

  function roundAPIM(value) {
    return Math.round(Number(value) * 10) / 10;
  }

  function roundDIM(value) {
    return Math.round(Number(value) * 20) / 20;
  }

  function clamp(value, min, max) {
    return Math.min(max, Math.max(min, Number(value)));
  }

  function format(value, digits) {
    var normalized = Math.abs(value) < 1e-10 ? 0 : value;
    return normalized.toFixed(digits);
  }

  function formatSigned(value, digits) {
    var normalized = Math.abs(value) < 1e-10 ? 0 : value;
    var sign = normalized < 0 ? "\u2212" : "\u2007";
    return sign + Math.abs(normalized).toFixed(digits);
  }

  function formatEquationNumber(value, digits, widthClass) {
    return '<span class="idg-equation-number idg-equation-number--' +
      widthClass + '">' + formatSigned(value, digits) + '</span>';
  }

  function formatCoefficient(value, following) {
    var normalized = Math.abs(value) < 1e-10 ? 0 : value;
    var magnitude = Math.abs(normalized).toFixed(2);
    if (following) {
      return normalized < 0 ? " \u2212 " + magnitude : " + " + magnitude;
    }
    return normalized < 0 ? "\u2212" + magnitude : magnitude;
  }

  function xPixel(value) {
    return 160 + value * plotScale;
  }

  function yPixel(value) {
    return 160 - value * plotScale;
  }

  function feasibleMean(within) {
    var bound = roundDIM(limit - Math.abs(within));
    return [-bound, bound];
  }

  function feasibleWithin(mean) {
    var bound = roundDIM(limit - Math.abs(mean));
    return [-bound, bound];
  }

  function render() {
    var actor = Number(inputs.actor.value);
    var partner = Number(inputs.partner.value);
    var mean = Number(inputs.mean.value);
    var within = Number(inputs.within.value);
    var apimContribution = slopes.actor * actor + slopes.partner * partner;
    var dimContribution = slopes.mean * mean + slopes.within * within;

    outputs.actor.value = formatSigned(actor, 1);
    outputs.partner.value = formatSigned(partner, 1);
    outputs.mean.value = formatSigned(mean, 2);
    outputs.within.value = formatSigned(within, 2);
    equationAPIM.innerHTML = formatCoefficient(slopes.actor, false) + " \u00d7 " +
      formatEquationNumber(actor, 1, "apim") +
      formatCoefficient(slopes.partner, true) + " \u00d7 " +
      formatEquationNumber(partner, 1, "apim") + " = " +
      formatEquationNumber(apimContribution, 2, "result");
    equationDIM.innerHTML = formatCoefficient(slopes.mean, false) + " \u00d7 " +
      formatEquationNumber(mean, 2, "dim") +
      formatCoefficient(slopes.within, true) + " \u00d7 " +
      formatEquationNumber(within, 2, "dim") + " = " +
      formatEquationNumber(dimContribution, 2, "result");
    slopeSummary.innerHTML =
      "<i>b</i><sub>mean</sub> = <i>b</i><sub>actor</sub> + " +
      "<i>b</i><sub>partner</sub> = " + formatCoefficient(slopes.mean, false) +
      "; <i>b</i><sub>dev</sub> = <i>b</i><sub>actor</sub> \u2212 " +
      "<i>b</i><sub>partner</sub> = " + formatCoefficient(slopes.within, false);

    point.setAttribute("cx", xPixel(actor));
    point.setAttribute("cy", yPixel(partner));
    halo.setAttribute("cx", xPixel(actor));
    halo.setAttribute("cy", yPixel(partner));
    description.textContent = "The selected point has actor " + format(actor, 1) +
      ", partner " + format(partner, 1) + ", dyad mean " + format(mean, 2) +
      ", and member deviation " + format(within, 2) + ".";
  }

  function setFromAPIM(actor, partner) {
    actor = roundAPIM(clamp(actor, -limit, limit));
    partner = roundAPIM(clamp(partner, -limit, limit));
    var mean = roundDIM((actor + partner) / 2);
    var within = roundDIM((actor - partner) / 2);
    inputs.actor.value = actor;
    inputs.partner.value = partner;
    inputs.mean.value = mean;
    inputs.within.value = within;
    render();
  }

  function setFromDIM(mean, within) {
    var meanBounds = feasibleMean(within);
    mean = roundDIM(clamp(mean, meanBounds[0], meanBounds[1]));
    var withinBounds = feasibleWithin(mean);
    within = roundDIM(clamp(within, withinBounds[0], withinBounds[1]));
    inputs.mean.value = mean;
    inputs.within.value = within;
    inputs.actor.value = roundAPIM(mean + within);
    inputs.partner.value = roundAPIM(mean - within);
    render();
  }

  inputs.actor.addEventListener("input", function () {
    setFromAPIM(inputs.actor.value, inputs.partner.value);
  });
  inputs.partner.addEventListener("input", function () {
    setFromAPIM(inputs.actor.value, inputs.partner.value);
  });
  inputs.mean.addEventListener("input", function () {
    var within = roundDIM(inputs.within.value);
    var bounds = feasibleMean(within);
    var mean = roundDIM(clamp(inputs.mean.value, bounds[0], bounds[1]));
    setFromDIM(mean, within);
  });
  inputs.within.addEventListener("input", function () {
    var mean = roundDIM(inputs.mean.value);
    var bounds = feasibleWithin(mean);
    var within = roundDIM(clamp(inputs.within.value, bounds[0], bounds[1]));
    setFromDIM(mean, within);
  });

  root.querySelector("[data-idg-reset]").addEventListener("click", function () {
    setFromAPIM(0, 0);
  });
  root.querySelector('[data-idg-demo="mean"]').addEventListener("click", function () {
    setFromAPIM(1, 1);
  });
  root.querySelector('[data-idg-demo="within"]').addEventListener("click", function () {
    setFromAPIM(1, -1);
  });

  function setFromPointer(event) {
    var bounds = plot.getBoundingClientRect();
    var svgX = (event.clientX - bounds.left) / bounds.width * 320;
    var svgY = (event.clientY - bounds.top) / bounds.height * 320;
    setFromAPIM((svgX - 160) / plotScale, (160 - svgY) / plotScale);
  }

  var dragging = false;
  plot.addEventListener("pointerdown", function (event) {
    dragging = true;
    plot.setPointerCapture(event.pointerId);
    setFromPointer(event);
  });
  plot.addEventListener("pointermove", function (event) {
    if (dragging) setFromPointer(event);
  });
  plot.addEventListener("pointerup", function (event) {
    dragging = false;
    if (plot.hasPointerCapture(event.pointerId)) plot.releasePointerCapture(event.pointerId);
  });
  plot.addEventListener("pointercancel", function () {
    dragging = false;
  });
  function drawGrid() {
    var namespace = "http://www.w3.org/2000/svg";
    var lines = root.querySelector("[data-idg-grid-lines]");
    var labels = root.querySelector("[data-idg-axis-labels]");

    function addLine(x1, y1, x2, y2, color, opacity, dash) {
      var line = document.createElementNS(namespace, "line");
      line.setAttribute("x1", x1);
      line.setAttribute("y1", y1);
      line.setAttribute("x2", x2);
      line.setAttribute("y2", y2);
      line.setAttribute("stroke", color);
      line.setAttribute("stroke-opacity", opacity);
      line.setAttribute("stroke-width", dash ? "1" : "1.5");
      if (dash) line.setAttribute("stroke-dasharray", dash);
      lines.appendChild(line);
    }

    [-4, -2, 0, 2, 4].forEach(function (tick) {
      var secondary = tick === 0 ? null : "3 3";
      var opacity = tick === 0 ? "0.65" : "0.24";
      addLine(xPixel(tick), 30, xPixel(tick), 290, colors.actor, opacity, secondary);
      addLine(30, yPixel(tick), 290, yPixel(tick), colors.partner, opacity, secondary);
      addLine(xPixel(-limit), yPixel(2 * tick + limit), xPixel(limit), yPixel(2 * tick - limit), colors.mean, opacity, secondary);
      addLine(xPixel(-limit), yPixel(-limit - 2 * tick), xPixel(limit), yPixel(limit - 2 * tick), colors.within, opacity, secondary);
    });

    function addLabel(text, x, y, color, anchor) {
      var label = document.createElementNS(namespace, "text");
      label.setAttribute("x", x);
      label.setAttribute("y", y);
      label.setAttribute("fill", color);
      label.setAttribute("font-size", "11");
      label.setAttribute("font-weight", "700");
      label.setAttribute("text-anchor", anchor || "middle");
      label.textContent = text;
      labels.appendChild(label);
    }

    addLabel("actor", 286, 151, colors.actor, "end");
    addLabel("partner", 160, 22, colors.partner);
    addLabel("dyad mean", 247, 66, colors.mean);
    addLabel("deviation", 248, 261, colors.within);
  }

  drawGrid();
  setFromAPIM(0, 0);
})();
</script>
```

### Random-effect transformation

The DIM and APIM models above already use the same sum-and-difference
random-effects parameterization [@delrosarioPracticalGuideSpecifying2025]. See
the [exchangeable residual-structure section of the APIM vignette](apim.html#exchangeable-residual-structure)
for the derivation and back-transformation to the member-level covariance matrix.

## Intensive Longitudinal DIM

For longitudinal DIM, predictors are decomposed into within-person and
between-person components before the dyadic decomposition
[@bolgerIntensiveLongitudinalMethods2013;
@gistelinckMultilevelAutoregressiveDyadic2020]. The default `"auto"` selects
`"time_2l"` when both `time` and `predictors` are supplied. It also
retains raw dyad-occasion means and member deviations. The decomposed
columns used below are:

1. The `cwp` dyad mean captures a shared occasion-specific shift from the two
   members' usual levels (shared occasion-level variation).
2. The `cwp` member deviation captures which member is further above or
   below their own usual level on that occasion.
3. The `cbp` dyad mean captures the dyad's shared usual level relative to the
   sample's grand mean (stable between-dyad differences).
4. The `cbp` member deviation captures each member's stable difference
   from the dyad's usual level.

The `cbp` terms use each member's mean across the observed occasions to
estimate that member's longer-run usual level. With few occasions (small
\(T\)), especially when the predictor has low stability over time, these person
means can be unreliable. The associated between-person estimates can therefore
be biased or imprecise, so they should be interpreted cautiously
[@gottfredsonStraightforwardApproachCoping2019].

Use `temporal_predictor_decomposition = "none"` to construct only the raw
dyad-occasion mean and member deviation.

```{r prepare-ild-exchangeable}
ild_exchangeable_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  time = diaryday,
  predictors = provided_support,
  model_type = c("apim", "dim"),
  seed = 123
)

print(ild_exchangeable_data)
```

The example below estimates same-day associations between support and closeness
and includes `diaryday` to adjust for a linear trend across the study.

```{r fit-ild-dim, eval = has_glmmTMB}

dim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person DIM
    .i_provided_support_cwp_dyad_mean +
    .i_provided_support_cwp_within_dyad_dev +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_dev +

    # Stable exchangeable dyad-level covariance
    us(1 | coupleID) +
    us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID) +

    # Residual (same-day) exchangeable dyad-level covariance
    us(1 | coupleID:diaryday) +
    us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

summary(dim_ILD)

```

### Interpretation of concurrent ILD DIM coefficients

The same interpretation as before applies longitudinally: dyad-mean effects describe
both expected outcomes for individual members and expected couple-average
outcomes, whereas within-dyad effects describe both member deviations and
expected differences between partners.

- The `cbp` dyad-mean estimate (about 1.51) means that, comparing couples whose
  average usual support differs by one point while holding the stable difference
  between partners constant, expected couple-average closeness is 1.51 points
  higher for the higher-support couple. Equivalently, each member is expected to
  report 1.51 points higher closeness.

- The `cwp` dyad-mean estimate (about 0.49) means that when both members are one
  point above their respective usual support levels, each member's expected
  closeness is 0.49 points higher than when both are at their usual levels,
  holding the difference between their momentary deviations constant.
  Equivalently, expected couple-average closeness is 0.49 points higher on that
  occasion.

- The `cbp` member-deviation estimate (about 0.78) means that if partners differ by
  one point in their usual support levels, they are expected to differ by 0.78
  points in closeness, holding the couple's average usual support and the other
  predictors constant. In member terms, suppose one member is 0.5 points above
  the couple's average usual support and the other is 0.5 points below it. Their
  expected closeness is then 0.39 points above and below the couple's predicted
  mean, respectively.

- The `cwp` member-deviation estimate (about 0.06) means that if one partner's
  momentary deviation from usual support is one point higher than the other
  partner's deviation, they are expected to differ by 0.06 points in closeness,
  holding the occasion-specific dyad mean and the other predictors constant.
  In member terms, suppose their momentary deviations are 0.5 points above and
  below the occasion-specific dyad mean. Their expected closeness is then 0.03
  points above and below the couple's predicted mean, respectively.

### Equivalence of APIM and DIM in ILD

The equivalent APIM uses actor and partner effects on both levels:

```{r fit-ild-apim, eval = has_glmmTMB}

apim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person APIM
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # Stable exchangeable dyad-level covariance
    us(1 | coupleID)  + us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID) +

    # Residual (same-day) exchangeable dyad-level covariance
    us(1 | coupleID:diaryday) +
    us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

The two ILD models again have identical fit statistics:

```{r compare-ild-fit, eval = has_glmmTMB}
data.frame(
  model = c("DIM", "APIM"),
  AIC = c(AIC(dim_ILD), AIC(apim_ILD)),
  BIC = c(BIC(dim_ILD), BIC(apim_ILD)),
  logLik = c(as.numeric(logLik(dim_ILD)), as.numeric(logLik(apim_ILD)))
)
```

The equivalence holds separately for the within-person (`cwp`) and
between-person (`cbp`) predictor components. For the within-person component:

$$
b_{\mathrm{cwp,mean}}
= b_{\mathrm{cwp,actor}} + b_{\mathrm{cwp,partner}}
$$

$$
b_{\mathrm{cwp,diff}}
= b_{\mathrm{cwp,actor}} - b_{\mathrm{cwp,partner}}
$$

For the between-person component:

$$
b_{\mathrm{cbp,mean}}
= b_{\mathrm{cbp,actor}} + b_{\mathrm{cbp,partner}}
$$

$$
b_{\mathrm{cbp,diff}}
= b_{\mathrm{cbp,actor}} - b_{\mathrm{cbp,partner}}
$$

This also means that an APIM parameterization can be used on one level and a
DIM parameterization on the other. For example:

```{r fit-ild-apim-within-dim-between, eval = has_glmmTMB}

apim_dim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_dev +

    # Stable exchangeable dyad-level covariance
    us(1 | coupleID)  + us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID) +

    # Same-day exchangeable dyad-level covariance
    us(1 | coupleID:diaryday) +
    us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

This mixed parameterization still estimates the same model:

```{r compare-ild-all-fit, eval = has_glmmTMB}
data.frame(
  model = c("DIM within / DIM between", "APIM within / APIM between", "APIM within / DIM between"),
  AIC = c(AIC(dim_ILD), AIC(apim_ILD), AIC(apim_dim_ILD)),
  BIC = c(BIC(dim_ILD), BIC(apim_ILD), BIC(apim_dim_ILD)),
  logLik = c(
    as.numeric(logLik(dim_ILD)),
    as.numeric(logLik(apim_ILD)),
    as.numeric(logLik(apim_dim_ILD))
  )
)
```

### Including Random Slopes

Random slopes can be included in the DIM by adding the corresponding
within-person effects to the stable dyad-level random-effect blocks:

The following syntax illustrates the full random-slope specification. It is not
fitted here because the example data do not support this complex random-effects
structure.

```{r fit-ild-dim-random, eval = FALSE}

dim_ILD_random <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person DIM
    .i_provided_support_cwp_dyad_mean +
    .i_provided_support_cwp_within_dyad_dev +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_dev +

    # Stable dyad-level covariance with within-person random slopes
    us(1 +
       .i_provided_support_cwp_dyad_mean +
       .i_provided_support_cwp_within_dyad_dev
     | coupleID)  +
    us(0 +
       .i_diff_assumed_exchangeable_arbitrary +
       .i_diff_assumed_exchangeable_arbitrary:.i_provided_support_cwp_dyad_mean +
       .i_diff_assumed_exchangeable_arbitrary:.i_provided_support_cwp_within_dyad_dev
     | coupleID) +

    # Same-day exchangeable dyad-level covariance
    us(1 | coupleID:diaryday) +
    us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

The first stable dyad-level random effects block contains the shared DIM intercept,
dyad-mean slope, and member-deviation slope. The second block contains
their member-difference counterparts, included through the `.i_diff_*` interactions.
These uncorrelated blocks allow the two members to have different random slopes while preserving
exchangeability.

#### Transforming DIM random slopes to APIM slopes

Applying the same transformation to the random-slope coefficients proceeds in
two steps. First, transform the DIM dyad-mean and member-deviation random
slopes into APIM actor and partner random slopes. Random effects use $u$, with
subscripts that write out `actor`, `partner`, `mean`, and `dev`. For the shared
block,

$$
u_{\mathrm{actor},j}
= \frac{u_{\mathrm{mean},j} + u_{\mathrm{dev},j}}{2},
\qquad
u_{\mathrm{partner},j}
= \frac{u_{\mathrm{mean},j} - u_{\mathrm{dev},j}}{2},
$$

and for the `.i_diff_*` block, marked by a tilde,

$$
\widetilde{u}_{\mathrm{actor},j}
= \frac{\widetilde{u}_{\mathrm{mean},j} + \widetilde{u}_{\mathrm{dev},j}}{2},
\qquad
\widetilde{u}_{\mathrm{partner},j}
= \frac{\widetilde{u}_{\mathrm{mean},j} - \widetilde{u}_{\mathrm{dev},j}}{2}.
$$

The shared and `.i_diff_*` random intercepts remain unchanged.

We now have the shared and `.i_diff_*` actor and partner effects which are then
back-transformed into the complete and more readily interpretable member-specific
actor-partner covariance matrix. This is described in the
[exchangeable random-slope back-transformation in the APIM vignette](apim.html#exchangeable-random-slope-back-transformation).

### Dynamic ILD DIM example

The ILD models above do not model residual serial dependence. One way to model
dynamics or to account for temporal dependency is to include lagged outcomes as predictors.

**Note:** Dynamic models, especially with small time series, are subject to bias.
This, and the choice between raw and within-person-centered outcome lags, are
addressed in the [APIM vignette's discussion of dynamic models](apim.html#dynamic-models).

Brief example to obtain lagged raw and within-person centered versions of the outcome:

```{r prepare-ild-dim-dynamic}
ild_exchangeable_data_dynamic <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  time = diaryday,
  predictors = c(provided_support, closeness),
  lag_predictors = closeness,
  model_type = "dim",
  seed = 123
)

print(ild_exchangeable_data_dynamic)
```

Lags are matched at exactly `diaryday - 1`, so omitted diary days are not
bridged.

The raw lagged outcome scores can be included in the model as follows:

```{r fit-ild-dim-dynamic, eval = has_glmmTMB}
dim_ILD_lag_raw <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    # Raw lagged outcomes
    .i_closeness_dyad_mean_gmc_lag1 + .i_closeness_within_dyad_dev_lag1 +

    diaryday +

    # Within-person DIM
    .i_provided_support_cwp_dyad_mean +
    .i_provided_support_cwp_within_dyad_dev +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_dev +

    # Stable exchangeable dyad-level covariance
    us(1 | coupleID)  +
    us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID) +

    # Same-day exchangeable dyad-level covariance
    us(1 | coupleID:diaryday) +
    us(0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data_dynamic

  # The model did not converge with the default optimizer
  , control = glmmTMB::glmmTMBControl(
      optimizer = stats::optim,
      optArgs = list(method = "BFGS")
    )
)

summary(dim_ILD_lag_raw)
```

The lagged dyad-mean coefficient describes how a one-point higher shared prior
outcome level relates to both members' current outcomes. The lagged
member-deviation coefficient describes how a one-point prior difference
between partners relates to their current expected difference.

All predictor effects in this model are conditional on both members' prior
outcomes.

---

Return to the [Actor-Partner Interdependence Model vignette](apim.html), see
the [Mixed-Composition APIM vignette](mixed-apim.html) or the
[Dyadic Score Model vignette](dsm.html) for related model specifications, or
return to the [Overview](index.html).

## References

::: {#refs}
:::
