---
title: "Dyad-Individual Model"
description: >
  Construction and interpretation of DIM predictors, including their
  relationship to APIM effects in dyadic multilevel models for exchangeable
  dyads.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dyad-Individual Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

This vignette focuses on the Dyad-Individual Model (DIM) for dyadic multilevel
models and its relationship to the exchangeable Actor-Partner Interdependence 
Model (APIM). The DIM separates a predictor into the dyad's shared level and each member's deviation from that level. While the APIM expresses effects in terms of two
interdependent individuals, the DIM expresses them in terms of the dyad's
shared level and the contrast between partners.

For the broader data-preparation workflow of the `interdep` package, see the
[Getting Started vignette](getting-started.html). For distinguishable,
exchangeable, generalized, and intensive longitudinal APIMs, see the
[Actor-Partner Interdependence Model vignette](apim.html). For APIMs that
combine distinguishable and exchangeable dyad compositions, see the
[Mixed-Composition APIM vignette](mixed-apim.html). For the dyadic score
model, see the [Dyadic Score Model vignette](dsm.html). Whereas the DIM is a
different parameterization of the exchangeable APIM, the Dyadic Score Model
(DSM) is a different parameterization of the distinguishable APIM.

## Cross-Sectional Gaussian DIM

The current DIM implementation prepares predictors for one exchangeable dyad
composition. Exchangeability means that swapping the two member labels does not
change the model. One way to make this assumption at the data-preparation layer
is to omit `role` from `prepare_interdep_data()`. This treats all dyads as the
same exchangeable composition.

```{r prepare-cross-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  predictors = communication,
  # Create both APIM and DIM columns for comparison.
  model_type = c("apim", "dim"),
  seed = 123
)

# Print the first two dyads.
print(cross_exchangeable_data, n = 4)
```

The function creates arbitrary member labels encoding a member-difference
contrast `.i_diff_*` as 1 for one partner and -1 for the other. This is needed
for the correct random effects specification. The fixed `seed` makes these
arbitrary member labels reproducible.

Passing a `role` is also possible when it leads to exactly one
exchangeable composition (e.g., only female-female dyads).

In a dataset with mixed dyad types, the right preparation depends on the analysis goal.

For instance, we can filter dyads and keep only male-male and female-female
dyads. Because DIM currently supports a single type of exchangeable dyad, we
might pool those like this, if theoretically justified:

```{r prepare-cross-filter-pool-same-sex}
cross_same_sex_pooled_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = satisfaction,
  model_type = c("apim", "dim"),
  # Remove male-female dyads.
  include_compositions = c("male-male", "female-female"),
  pool_compositions = list(
    "same-sex-couples" = c("male-male", "female-female")
  ),
  seed = 123
)

print(cross_same_sex_pooled_data, n = 4)
```

If we want to include only male-female couples and treat those as exchangeable,
we can do:

```{r prepare-cross-filter-set-male-female-exchangeable}
cross_male_female_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = satisfaction,
  model_type = c("apim", "dim"),
  include_compositions = "male-female",
  set_exchangeable_compositions = "male-female",
  seed = 123
)

print(cross_male_female_exchangeable_data, n = 4)
```

Pooling compositions constrains them to share DIM effects. These are substantive
modeling decisions, not only data-processing choices, and should follow the
research question.

### Example DIM Model

For members \(j \in \{1, 2\}\) of dyad \(i\), the DIM decomposition is computed
as:

$$
\bar{x}_i = \frac{x_{i1} + x_{i2}}{2}, \qquad
d_{ij} = x_{ij} - \bar{x}_i.
$$

The two deviations of the two partners have equal magnitude and opposite signs:
\(d_{i1} = -d_{i2}\). The variables that enter the DIM fixed effects then represent:

1. A dyad-mean variable that is grand-mean centered. This describes a couple's
shared level compared to all other couples.
2. A within-dyad variable describing each partner's deviation from the couple
mean, equivalently half the signed difference between partners.

The resulting estimated fixed effects are a reparameterization of the APIM actor
and partner effects. The same random-effects structure can therefore be used for both
fixed-effect parameterizations: a dyad-level intercept and a dyad-level
difference contrast indexed by `.i_diff_assumed_exchangeable_arbitrary`.
In `glmmTMB`, with `dispformula = ~ 0`, these random effects represent the two members' Gaussian residual variance and covariance.

The intercept and difference contrast are specified as separate random-effects
terms, which constrains their correlation to zero. This preserves
exchangeability: swapping the arbitrary member labels leaves the intercept term
unchanged but reverses the difference contrast. A nonzero correlation between
them would therefore make the random-effects distribution depend on the
arbitrary labeling.

The full model can be estimated as:

```{r fit-cross-dim, eval=has_glmmTMB}

dim_1 <- glmmTMB::glmmTMB(
  satisfaction ~

    # Pooled fixed intercept
    1 +

    # Between-couple effect
    .i_communication_dyad_mean_gmc +

    # Within-couple effect
    .i_communication_within_dyad_dev +

    # Residual Gaussian variance-covariance
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID)
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

summary(dim_1)
```

```{r glmmTMB-missing, echo = FALSE, eval = !has_glmmTMB}
message("Install glmmTMB to run the fitted-model examples in this vignette.")
```

### Model interpretation

Because this Gaussian model uses an identity link, fixed coefficients are
interpreted in units of the outcome, e.g., "satisfaction":

The exchangeable Gaussian DIM is algebraically equivalent to the
reduced, label-invariant Dyadic Score Model (DSM). Therefore, each coefficient has
both an individual-member interpretation and an equivalent couple mean/difference
interpretation.

- The intercept (about 5.04) is the expected satisfaction of either member,
  and therefore the expected couple-average satisfaction, when both members'
  communication equals the sample grand mean.

- The dyad-mean estimate (about 2.00) means that, comparing couples with the
  same communication difference between partners, a one-point higher
  couple-average communication level is associated with a 2.00-point higher
  expected couple-average satisfaction. Equivalently, each member's expected
  satisfaction is 2.00 points higher.

- The within-dyad-deviation estimate (about 1.52) means that a one-point
  difference in communication between partners is associated with a 1.52-point
  difference in their expected satisfaction, holding their average communication
  constant. Equivalently, if one member is 1 point above the dyad mean, holding
  the dyad mean constant implies that the other is 1 point below it. The members
  are then 2 points apart in communication. The first member's expected
  satisfaction is then 1.52 points above the couple's predicted mean, the second
  member's is 1.52 points below it, and they are expected to differ by 3.04
  points in satisfaction.

### Demonstrating model equivalence to APIM

The same model can be written in APIM form. Since we have requested both sets of
variables from `prepare_interdep_data`, we can fit one directly. For more
guidance on APIM specifications and different models, see the
[Actor-Partner Interdependence Model vignette](apim.html).

```{r fit-cross-apim, eval = has_glmmTMB}

apim_1 <- glmmTMB::glmmTMB(
  satisfaction ~ 1 +
    .i_communication_actor + .i_communication_partner +
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID)
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

```

The two models have identical fit statistics:

```{r compare-cross-fit, eval=has_glmmTMB}
data.frame(
  model = c("DIM", "APIM"),
  AIC = c(AIC(dim_1), AIC(apim_1)),
  BIC = c(BIC(dim_1), BIC(apim_1)),
  logLik = c(as.numeric(logLik(dim_1)), as.numeric(logLik(apim_1)))
)
```

This demonstrates that the same statistical model is being estimated with different
parameterizations and coefficient interpretations.

Once APIM estimates are present, one can easily obtain DIM estimates, and
the other way around. The slope coefficients relate as follows:

$$
\beta_{\text{dyad mean}} =
\beta_{\text{actor}} + \beta_{\text{partner}}
$$

and

$$
\beta_{\text{within-dyad deviation}} =
\beta_{\text{actor}} - \beta_{\text{partner}}
$$

Conversely:

$$
\beta_{\text{actor}} =
\frac{\beta_{\text{dyad mean}} + \beta_{\text{within-dyad deviation}}}{2}
$$

and

$$
\beta_{\text{partner}} =
\frac{\beta_{\text{dyad mean}} - \beta_{\text{within-dyad deviation}}}{2}
$$

### Why Are These Models Equivalent? Exploring the Reparameterization

An intuitive way to think about this is:

- When the dyad mean goes up by 1 unit while the difference between partners
  remains stable, both partners' values must go up by 1. Both the actor and
  partner effects therefore contribute, which is why the dyad-mean effect is
  the actor effect + the partner effect.

- When a person's deviation from the dyad mean goes up by 1 unit while the dyad
  mean remains constant, the other partner's value must go **down** by 1 unit.
  The actor value therefore changes by +1 and the partner value by -1, which is
  why the within-dyad-deviation effect is the actor effect - the partner effect.

The grid below shows the same predictor values in both coordinate systems. The
horizontal and vertical axes are actor and partner values centered at the
sample grand mean. The diagonal axes are their dyad mean and within-dyad
deviation.

The displayed slope values approximate the fitted example above and illustrate
that the two forms make the same change in the linear predictor relative to the
grand-mean reference. The intercept is omitted from both displayed equations.

```{=html}
<style>
.interdep-dim-grid {
  --idg-actor: #1769aa;
  --idg-partner: #7b3fa1;
  --idg-mean: #c43c35;
  --idg-within: #8a6500;
  --idg-border: var(--bs-border-color, #c9ced6);
  --idg-surface: var(--bs-tertiary-bg, #f5f6f8);
  --idg-ink: var(--bs-body-color, #202428);
  box-sizing: border-box;
  display: grid;
  grid-template-columns: minmax(15rem, 19rem) minmax(0, 1fr);
  gap: 1.25rem;
  margin: 1.25rem 0 2rem;
  padding: 1rem;
  color: var(--idg-ink);
  background: var(--idg-surface);
  border: 1px solid var(--idg-border);
  border-radius: 8px;
}
.interdep-dim-grid *,
.interdep-dim-grid *::before,
.interdep-dim-grid *::after {
  box-sizing: border-box;
}
.idg-controls,
.idg-visual,
.idg-control-group {
  display: grid;
  align-content: start;
  gap: 0.9rem;
  min-width: 0;
}
.idg-toolbar,
.idg-group-title,
.idg-control-head,
.idg-demo-row {
  display: flex;
  align-items: center;
  justify-content: space-between;
  gap: 0.75rem;
}
.idg-toolbar {
  padding-bottom: 0.7rem;
  border-bottom: 1px solid var(--idg-border);
}
.idg-toolbar strong,
.idg-group-title {
  font-size: 0.82rem;
  font-weight: 700;
}
.idg-reset,
.idg-demo {
  padding: 0.35rem 0.6rem;
  color: inherit;
  background: var(--bs-body-bg, #fff);
  border: 1px solid var(--idg-border);
  border-radius: 5px;
  font: inherit;
  font-size: 0.78rem;
  line-height: 1.2;
  cursor: pointer;
}
.idg-reset:hover,
.idg-demo:hover {
  border-color: currentColor;
}
.idg-reset:focus-visible,
.idg-demo:focus-visible,
.interdep-dim-grid input:focus-visible {
  outline: 3px solid rgba(23, 105, 170, 0.35);
  outline-offset: 2px;
}
.idg-group-title {
  justify-content: flex-start;
  padding-bottom: 0.35rem;
  border-bottom: 1px solid var(--idg-border);
}
.idg-swatches {
  display: inline-flex;
  gap: 0.25rem;
}
.idg-swatch {
  width: 0.65rem;
  height: 0.65rem;
  border-radius: 50%;
}
.idg-control {
  display: grid;
  gap: 0.25rem;
}
.idg-control label {
  min-width: 0;
  font-size: 0.84rem;
  font-weight: 600;
}
.idg-control output {
  flex: 0 0 5.5ch;
  text-align: right;
  font-variant-numeric: tabular-nums;
  font-weight: 700;
}
.idg-control input[type="range"] {
  width: 100%;
  min-height: 1.75rem;
  margin: 0;
  cursor: pointer;
}
.idg-actor input { accent-color: var(--idg-actor); }
.idg-partner input { accent-color: var(--idg-partner); }
.idg-mean input { accent-color: var(--idg-mean); }
.idg-within input { accent-color: var(--idg-within); }
.idg-relations {
  padding-top: 0.7rem;
  border-top: 1px solid var(--idg-border);
  font-size: 0.8rem;
  line-height: 1.6;
}
.idg-equations {
  display: grid;
  gap: 0.4rem;
  min-width: 0;
  width: 100%;
  max-width: 100%;
  padding: 0.75rem;
  background: var(--bs-body-bg, #fff);
  border: 1px solid var(--idg-border);
  border-radius: 6px;
  overflow-x: auto;
  font-family: Georgia, "Times New Roman", serif;
  font-size: 0.94rem;
  font-variant-numeric: tabular-nums;
  white-space: nowrap;
}
.idg-equations > div:last-child {
  white-space: normal;
}
.idg-equation-label {
  display: inline-block;
  width: 3.3rem;
  font-family: var(--bs-body-font-family, sans-serif);
  font-size: 0.7rem;
  font-weight: 700;
  text-transform: uppercase;
}
.idg-equation-number {
  display: inline-block;
  text-align: right;
}
.idg-equation-number--apim {
  width: 4ch;
}
.idg-equation-number--dim {
  width: 5ch;
}
.idg-equation-number--result {
  width: 6ch;
}
.idg-stage {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(min(100%, 16rem), 1fr));
  gap: 1rem;
  align-items: start;
}
.idg-plot {
  display: block;
  width: 100%;
  max-width: 29rem;
  aspect-ratio: 1;
  margin: 0 auto;
  background: var(--bs-body-bg, #fff);
  border: 1px solid var(--idg-border);
  border-radius: 6px;
  cursor: crosshair;
  touch-action: none;
  user-select: none;
}
.idg-explanation {
  display: grid;
  gap: 1rem;
  font-size: 0.82rem;
  line-height: 1.45;
}
.idg-explanation p {
  margin: 0.3rem 0 0;
}
.idg-demo-row {
  align-items: flex-start;
}
.idg-demo-row strong {
  display: block;
}
.idg-mini-equation {
  margin-top: 0.4rem;
  font-family: Georgia, "Times New Roman", serif;
  font-size: 0.86rem;
}
.idg-result {
  font-weight: 700;
}
.idg-halo {
  fill: var(--idg-actor);
  fill-opacity: 0.14;
}
.idg-point {
  fill: var(--idg-actor);
  stroke: var(--bs-body-bg, #fff);
  stroke-width: 2;
}
.idg-sr-only {
  position: absolute;
  width: 1px;
  height: 1px;
  padding: 0;
  margin: -1px;
  overflow: hidden;
  clip: rect(0, 0, 0, 0);
  white-space: nowrap;
  border: 0;
}
@media (max-width: 860px) {
  .interdep-dim-grid {
    grid-template-columns: 1fr;
  }
  .idg-controls {
    grid-template-columns: repeat(2, minmax(0, 1fr));
  }
  .idg-toolbar,
  .idg-relations {
    grid-column: 1 / -1;
  }
}
@media (max-width: 620px) {
  .interdep-dim-grid {
    margin-right: -0.35rem;
    margin-left: -0.35rem;
    padding: 0.75rem;
  }
  .idg-controls,
  .idg-stage {
    grid-template-columns: 1fr;
  }
  .idg-toolbar,
  .idg-relations {
    grid-column: auto;
  }
  .idg-equations {
    font-size: 0.84rem;
  }
}
@media (prefers-reduced-motion: reduce) {
  .interdep-dim-grid * {
    scroll-behavior: auto !important;
  }
}
[data-bs-theme="dark"] .interdep-dim-grid {
  --idg-actor: #75b7ff;
  --idg-partner: #d0a6ff;
  --idg-mean: #ff8c82;
  --idg-within: #f0c75e;
}
</style>

<div class="interdep-dim-grid" id="interdep-dim-grid">
  <div class="idg-controls">
    <div class="idg-toolbar">
      <strong>Predictor coordinates</strong>
      <button class="idg-reset" type="button" data-idg-reset>Reset</button>
    </div>

    <div class="idg-control-group">
      <div class="idg-group-title">
        <span class="idg-swatches" aria-hidden="true">
          <span class="idg-swatch" style="background: var(--idg-actor)"></span>
          <span class="idg-swatch" style="background: var(--idg-partner)"></span>
        </span>
        APIM coordinates
      </div>
      <div class="idg-control idg-actor">
        <div class="idg-control-head">
          <label for="idg-actor">Grand-mean-centered actor, <i>x</i><sub>actor</sub></label>
          <output for="idg-actor" data-idg-output="actor">0.0</output>
        </div>
        <input id="idg-actor" data-idg-input="actor" type="range"
               min="-5" max="5" step="0.1" value="0">
      </div>
      <div class="idg-control idg-partner">
        <div class="idg-control-head">
          <label for="idg-partner">Grand-mean-centered partner, <i>x</i><sub>partner</sub></label>
          <output for="idg-partner" data-idg-output="partner">0.0</output>
        </div>
        <input id="idg-partner" data-idg-input="partner" type="range"
               min="-5" max="5" step="0.1" value="0">
      </div>
    </div>

    <div class="idg-control-group">
      <div class="idg-group-title">
        <span class="idg-swatches" aria-hidden="true">
          <span class="idg-swatch" style="background: var(--idg-mean)"></span>
          <span class="idg-swatch" style="background: var(--idg-within)"></span>
        </span>
        DIM coordinates
      </div>
      <div class="idg-control idg-mean">
        <div class="idg-control-head">
          <label for="idg-mean">Dyad mean, <i>x</i><sub>mean</sub></label>
          <output for="idg-mean" data-idg-output="mean">0.0</output>
        </div>
        <input id="idg-mean" data-idg-input="mean" type="range"
               min="-5" max="5" step="0.1" value="0">
      </div>
      <div class="idg-control idg-within">
        <div class="idg-control-head">
          <label for="idg-within">Within-dyad deviation, <i>x</i><sub>within</sub></label>
          <output for="idg-within" data-idg-output="within">0.0</output>
        </div>
        <input id="idg-within" data-idg-input="within" type="range"
               min="-5" max="5" step="0.1" value="0">
      </div>
    </div>

    <div class="idg-relations">
      <div><i>x</i><sub>mean</sub> = (<i>x</i><sub>actor</sub> + <i>x</i><sub>partner</sub>) / 2</div>
      <div><i>x</i><sub>within</sub> = (<i>x</i><sub>actor</sub> - <i>x</i><sub>partner</sub>) / 2</div>
    </div>
  </div>

  <div class="idg-visual">
    <div class="idg-equations">
      <div><span class="idg-equation-label">APIM</span><span data-idg-equation="apim">0.00 = 0.00</span></div>
      <div><span class="idg-equation-label">DIM</span><span data-idg-equation="dim">0.00 = 0.00</span></div>
      <div><span class="idg-equation-label">Slopes</span><span><i>b</i><sub>mean</sub> = 1.76 + 0.24 = 2.00; <i>b</i><sub>within</sub> = 1.76 - 0.24 = 1.52</span></div>
    </div>

    <div class="idg-stage">
      <svg class="idg-plot" data-idg-plot viewBox="0 0 320 320"
           role="img" aria-labelledby="idg-plot-title idg-plot-description">
        <title id="idg-plot-title">APIM and DIM coordinate grid</title>
        <desc id="idg-plot-description">The selected point has actor, partner, dyad-mean, and within-dyad values of zero.</desc>
        <defs>
          <clipPath id="idg-plot-clip">
            <rect x="30" y="30" width="260" height="260" rx="4"></rect>
          </clipPath>
        </defs>
        <rect x="30" y="30" width="260" height="260" rx="4"
              fill="none" stroke="currentColor" stroke-opacity="0.35"></rect>
        <g data-idg-grid-lines clip-path="url(#idg-plot-clip)"></g>
        <g data-idg-axis-labels></g>
        <circle class="idg-halo" data-idg-halo cx="160" cy="160" r="13"></circle>
        <circle class="idg-point" data-idg-point cx="160" cy="160" r="6"></circle>
      </svg>

      <div class="idg-explanation">
        <div>
          <div class="idg-demo-row">
            <strong>Shared dyad level</strong>
            <button class="idg-demo" type="button" data-idg-demo="mean">Show</button>
          </div>
          <p>When both members are one point higher, the dyad mean is one point higher and the deviation is unchanged.</p>
          <div class="idg-mini-equation"><i>b</i><sub>mean</sub> = <i>b</i><sub>actor</sub> + <i>b</i><sub>partner</sub></div>
        </div>
        <div>
          <div class="idg-demo-row">
            <strong>Within-dyad difference</strong>
            <button class="idg-demo" type="button" data-idg-demo="within">Show</button>
          </div>
          <p>When one member is one point higher and the other one point lower, the dyad mean is unchanged and the deviation is one point.</p>
          <div class="idg-mini-equation"><i>b</i><sub>within</sub> = <i>b</i><sub>actor</sub> - <i>b</i><sub>partner</sub></div>
        </div>
        <p>The dot uses APIM coordinates. The diagonal lines show the same point in DIM coordinates.</p>
      </div>
    </div>
  </div>
  <noscript><p>Enable JavaScript to manipulate this figure. The equations and discussion above provide the same transformation.</p></noscript>
</div>

<script>
(function () {
  "use strict";

  var root = document.getElementById("interdep-dim-grid");
  if (!root || root.dataset.initialized === "true") return;
  root.dataset.initialized = "true";

  var limit = 5;
  var plotScale = 130 / limit;
  var slopes = { actor: 1.76, partner: 0.24, mean: 2.00, within: 1.52 };
  var colors = {
    actor: "var(--idg-actor)",
    partner: "var(--idg-partner)",
    mean: "var(--idg-mean)",
    within: "var(--idg-within)"
  };
  var inputs = {};
  var outputs = {};
  ["actor", "partner", "mean", "within"].forEach(function (name) {
    inputs[name] = root.querySelector('[data-idg-input="' + name + '"]');
    outputs[name] = root.querySelector('[data-idg-output="' + name + '"]');
  });

  var plot = root.querySelector("[data-idg-plot]");
  var point = root.querySelector("[data-idg-point]");
  var halo = root.querySelector("[data-idg-halo]");
  var description = root.querySelector("#idg-plot-description");
  var equationAPIM = root.querySelector('[data-idg-equation="apim"]');
  var equationDIM = root.querySelector('[data-idg-equation="dim"]');

  function roundAPIM(value) {
    return Math.round(Number(value) * 10) / 10;
  }

  function roundDIM(value) {
    return Math.round(Number(value) * 20) / 20;
  }

  function clamp(value, min, max) {
    return Math.min(max, Math.max(min, Number(value)));
  }

  function format(value, digits) {
    var normalized = Math.abs(value) < 1e-10 ? 0 : value;
    return normalized.toFixed(digits);
  }

  function formatSigned(value, digits) {
    var normalized = Math.abs(value) < 1e-10 ? 0 : value;
    var sign = normalized < 0 ? "\u2212" : "\u2007";
    return sign + Math.abs(normalized).toFixed(digits);
  }

  function formatEquationNumber(value, digits, widthClass) {
    return '<span class="idg-equation-number idg-equation-number--' +
      widthClass + '">' + formatSigned(value, digits) + '</span>';
  }

  function xPixel(value) {
    return 160 + value * plotScale;
  }

  function yPixel(value) {
    return 160 - value * plotScale;
  }

  function feasibleMean(within) {
    var bound = roundDIM(limit - Math.abs(within));
    return [-bound, bound];
  }

  function feasibleWithin(mean) {
    var bound = roundDIM(limit - Math.abs(mean));
    return [-bound, bound];
  }

  function render() {
    var actor = Number(inputs.actor.value);
    var partner = Number(inputs.partner.value);
    var mean = Number(inputs.mean.value);
    var within = Number(inputs.within.value);
    var apimContribution = slopes.actor * actor + slopes.partner * partner;
    var dimContribution = slopes.mean * mean + slopes.within * within;

    outputs.actor.value = formatSigned(actor, 1);
    outputs.partner.value = formatSigned(partner, 1);
    outputs.mean.value = formatSigned(mean, 2);
    outputs.within.value = formatSigned(within, 2);
    equationAPIM.innerHTML = "1.76 \u00d7 " +
      formatEquationNumber(actor, 1, "apim") + " + 0.24 \u00d7 " +
      formatEquationNumber(partner, 1, "apim") + " = " +
      formatEquationNumber(apimContribution, 2, "result");
    equationDIM.innerHTML = "2.00 \u00d7 " +
      formatEquationNumber(mean, 2, "dim") + " + 1.52 \u00d7 " +
      formatEquationNumber(within, 2, "dim") + " = " +
      formatEquationNumber(dimContribution, 2, "result");

    point.setAttribute("cx", xPixel(actor));
    point.setAttribute("cy", yPixel(partner));
    halo.setAttribute("cx", xPixel(actor));
    halo.setAttribute("cy", yPixel(partner));
    description.textContent = "The selected point has actor " + format(actor, 1) +
      ", partner " + format(partner, 1) + ", dyad mean " + format(mean, 2) +
      ", and within-dyad deviation " + format(within, 2) + ".";
  }

  function setFromAPIM(actor, partner) {
    actor = roundAPIM(clamp(actor, -limit, limit));
    partner = roundAPIM(clamp(partner, -limit, limit));
    var mean = roundDIM((actor + partner) / 2);
    var within = roundDIM((actor - partner) / 2);
    inputs.actor.value = actor;
    inputs.partner.value = partner;
    inputs.mean.value = mean;
    inputs.within.value = within;
    render();
  }

  function setFromDIM(mean, within) {
    var meanBounds = feasibleMean(within);
    mean = roundDIM(clamp(mean, meanBounds[0], meanBounds[1]));
    var withinBounds = feasibleWithin(mean);
    within = roundDIM(clamp(within, withinBounds[0], withinBounds[1]));
    inputs.mean.value = mean;
    inputs.within.value = within;
    inputs.actor.value = roundAPIM(mean + within);
    inputs.partner.value = roundAPIM(mean - within);
    render();
  }

  inputs.actor.addEventListener("input", function () {
    setFromAPIM(inputs.actor.value, inputs.partner.value);
  });
  inputs.partner.addEventListener("input", function () {
    setFromAPIM(inputs.actor.value, inputs.partner.value);
  });
  inputs.mean.addEventListener("input", function () {
    var within = roundDIM(inputs.within.value);
    var bounds = feasibleMean(within);
    var mean = roundDIM(clamp(inputs.mean.value, bounds[0], bounds[1]));
    setFromDIM(mean, within);
  });
  inputs.within.addEventListener("input", function () {
    var mean = roundDIM(inputs.mean.value);
    var bounds = feasibleWithin(mean);
    var within = roundDIM(clamp(inputs.within.value, bounds[0], bounds[1]));
    setFromDIM(mean, within);
  });

  root.querySelector("[data-idg-reset]").addEventListener("click", function () {
    setFromAPIM(0, 0);
  });
  root.querySelector('[data-idg-demo="mean"]').addEventListener("click", function () {
    setFromAPIM(1, 1);
  });
  root.querySelector('[data-idg-demo="within"]').addEventListener("click", function () {
    setFromAPIM(1, -1);
  });

  function setFromPointer(event) {
    var bounds = plot.getBoundingClientRect();
    var svgX = (event.clientX - bounds.left) / bounds.width * 320;
    var svgY = (event.clientY - bounds.top) / bounds.height * 320;
    setFromAPIM((svgX - 160) / plotScale, (160 - svgY) / plotScale);
  }

  var dragging = false;
  plot.addEventListener("pointerdown", function (event) {
    dragging = true;
    plot.setPointerCapture(event.pointerId);
    setFromPointer(event);
  });
  plot.addEventListener("pointermove", function (event) {
    if (dragging) setFromPointer(event);
  });
  plot.addEventListener("pointerup", function (event) {
    dragging = false;
    if (plot.hasPointerCapture(event.pointerId)) plot.releasePointerCapture(event.pointerId);
  });
  plot.addEventListener("pointercancel", function () {
    dragging = false;
  });
  function drawGrid() {
    var namespace = "http://www.w3.org/2000/svg";
    var lines = root.querySelector("[data-idg-grid-lines]");
    var labels = root.querySelector("[data-idg-axis-labels]");

    function addLine(x1, y1, x2, y2, color, opacity, dash) {
      var line = document.createElementNS(namespace, "line");
      line.setAttribute("x1", x1);
      line.setAttribute("y1", y1);
      line.setAttribute("x2", x2);
      line.setAttribute("y2", y2);
      line.setAttribute("stroke", color);
      line.setAttribute("stroke-opacity", opacity);
      line.setAttribute("stroke-width", dash ? "1" : "1.5");
      if (dash) line.setAttribute("stroke-dasharray", dash);
      lines.appendChild(line);
    }

    [-4, -2, 0, 2, 4].forEach(function (tick) {
      var secondary = tick === 0 ? null : "3 3";
      var opacity = tick === 0 ? "0.65" : "0.24";
      addLine(xPixel(tick), 30, xPixel(tick), 290, colors.actor, opacity, secondary);
      addLine(30, yPixel(tick), 290, yPixel(tick), colors.partner, opacity, secondary);
      addLine(xPixel(-limit), yPixel(2 * tick + limit), xPixel(limit), yPixel(2 * tick - limit), colors.mean, opacity, secondary);
      addLine(xPixel(-limit), yPixel(-limit - 2 * tick), xPixel(limit), yPixel(limit - 2 * tick), colors.within, opacity, secondary);
    });

    function addLabel(text, x, y, color, anchor) {
      var label = document.createElementNS(namespace, "text");
      label.setAttribute("x", x);
      label.setAttribute("y", y);
      label.setAttribute("fill", color);
      label.setAttribute("font-size", "11");
      label.setAttribute("font-weight", "700");
      label.setAttribute("text-anchor", anchor || "middle");
      label.textContent = text;
      labels.appendChild(label);
    }

    addLabel("actor", 286, 151, colors.actor, "end");
    addLabel("partner", 160, 22, colors.partner);
    addLabel("dyad mean", 247, 66, colors.mean);
    addLabel("within", 248, 261, colors.within);
  }

  drawGrid();
  setFromAPIM(0, 0);
})();
</script>
```

The DIM and APIM intercepts are not expected to be equal because the DIM dyad
mean is grand-mean centered, whereas the APIM predictors retain their original
scale.

In this example:

```{r compare-cross-coefficients, echo = TRUE, eval = has_glmmTMB}
apim_coef <- glmmTMB::fixef(apim_1)$cond
dim_coef <- glmmTMB::fixef(dim_1)$cond

b_actor <- apim_coef[[".i_communication_actor"]]
b_partner <- apim_coef[[".i_communication_partner"]]

b_dyad_mean <- dim_coef[[".i_communication_dyad_mean_gmc"]]
b_within_dyad <- dim_coef[[".i_communication_within_dyad_dev"]]


cat("From APIM model:\n",
     "  actor effect:                  ", round(b_actor, 3), "\n",
     "  partner effect:                ", round(b_partner, 3), "\n\n",

     "DIM transformation:\n",
     "  actor effect + partner effect: ", round(b_actor + b_partner, 3), "\n",
     "  actor effect - partner effect: ", round(b_actor - b_partner, 3), "\n\n",

     "From DIM model:\n",
     "  dyad-mean effect:              ", round(b_dyad_mean, 3), "\n",
     "  within-dyad-deviation effect:  ", round(b_within_dyad, 3), "\n"
)

```

## Intensive Longitudinal DIM

For longitudinal DIM, predictors are decomposed into within-person and
between-person components before the dyadic decomposition. The default `"auto"`
strategy selects `"time_2l"` when both `time` and `predictors` are supplied:

1. The `cwp` dyad mean captures a shared occasion-specific shift from the two
   members' usual levels (shared occasion-level variation).
2. The `cwp` within-dyad deviation captures which member is further above or
   below their own usual level on that occasion.
3. The `cbp` dyad mean captures the dyad's shared usual level relative to the
   sample's grand mean (stable between-dyad differences).
4. The `cbp` within-dyad deviation captures each member's stable difference
   from the dyad's usual level.

`temporal_predictor_decomposition = "none"` is not available for longitudinal
DIM predictor construction.

```{r prepare-ild-exchangeable}
ild_exchangeable_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  time = diaryday,
  predictors = provided_support,
  model_type = c("apim", "dim"),
  seed = 123
)

print(ild_exchangeable_data)
```

```{r fit-ild-dim, eval = has_glmmTMB}

dim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person DIM
    .i_provided_support_cwp_dyad_mean +
    .i_provided_support_cwp_within_dyad_dev +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_dev +

    # Stable exchangeable dyad-level covariance
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID) +

    # Residual (same-day) exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

summary(dim_ILD)

```

### Interpretation of ILD DIM coefficients

The same interpretation as before applies longitudinally: dyad-mean effects describe
both expected outcomes for individual members and expected couple-average
outcomes, whereas within-dyad effects describe both member deviations and
expected differences between partners.

- The `cbp` dyad-mean estimate (about 1.51) means that, comparing couples whose
  average usual support differs by one point while holding the stable difference
  between partners constant, expected couple-average closeness is 1.51 points
  higher for the higher-support couple. Equivalently, each member is expected to
  report 1.51 points higher closeness.

- The `cwp` dyad-mean estimate (about 0.49) means that when both members are one
  point above their respective usual support levels, each member's expected
  closeness is 0.49 points higher than when both are at their usual levels,
  holding the difference between their momentary deviations constant.
  Equivalently, expected couple-average closeness is 0.49 points higher on that
  occasion.

- The `cbp` within-dyad estimate (about 0.78) means that if partners differ by
  one point in their usual support levels, they are expected to differ by 0.78
  points in closeness, holding the couple's average usual support and the other
  predictors constant. In member terms, a one-point increase in one member's
  stable deviation from the dyad's usual support level, with the other member's
  deviation decreasing by one point, is associated with a 0.78-point increase
  for the first member and a 0.78-point decrease for the second.

- The `cwp` within-dyad estimate (about 0.06) means that if one partner's
  momentary deviation from usual support is one point higher than the other
  partner's deviation, they are expected to differ by 0.06 points in closeness,
  holding the occasion-specific dyad mean and the other predictors constant.
  Equivalently, increasing one member's momentary deviation by one point while
  decreasing the other member's deviation by one point is associated with a
  0.06-point increase for the first member and a 0.06-point decrease for the
  second.

### Equivalence of APIM and DIM in ILD

The equivalent APIM uses actor and partner effects on both levels:

```{r fit-ild-apim, eval = has_glmmTMB}

apim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person APIM
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # Stable exchangeable dyad-level covariance
    (1 | coupleID)  + (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID) +

    # Residual (same-day) exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

The two ILD models again have identical fit statistics:

```{r compare-ild-fit, eval = has_glmmTMB}
data.frame(
  model = c("DIM", "APIM"),
  AIC = c(AIC(dim_ILD), AIC(apim_ILD)),
  BIC = c(BIC(dim_ILD), BIC(apim_ILD)),
  logLik = c(as.numeric(logLik(dim_ILD)), as.numeric(logLik(apim_ILD)))
)
```

The equivalence holds separately for the within-person (`cwp`) and
between-person (`cbp`) predictor components. For the within-person component:

$$
\beta_{\text{cwp dyad mean}} =
\beta_{\text{cwp actor}} + \beta_{\text{cwp partner}}
$$

$$
\beta_{\text{cwp within-dyad deviation}} =
\beta_{\text{cwp actor}} - \beta_{\text{cwp partner}}
$$

For the between-person component:

$$
\beta_{\text{cbp dyad mean}} =
\beta_{\text{cbp actor}} + \beta_{\text{cbp partner}}
$$

$$
\beta_{\text{cbp within-dyad deviation}} =
\beta_{\text{cbp actor}} - \beta_{\text{cbp partner}}
$$

This also means that an APIM parameterization can be used on one level and a
DIM parameterization on the other. For example:

```{r fit-ild-apim-within-dim-between, eval = has_glmmTMB}

apim_dim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_dev +

    # Stable exchangeable dyad-level covariance
    (1 | coupleID)  + (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID) +

    # Same-day exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

This mixed parameterization still estimates the same model:

```{r compare-ild-all-fit, eval = has_glmmTMB}
data.frame(
  model = c("DIM within / DIM between", "APIM within / APIM between", "APIM within / DIM between"),
  AIC = c(AIC(dim_ILD), AIC(apim_ILD), AIC(apim_dim_ILD)),
  BIC = c(BIC(dim_ILD), BIC(apim_ILD), BIC(apim_dim_ILD)),
  logLik = c(
    as.numeric(logLik(dim_ILD)),
    as.numeric(logLik(apim_ILD)),
    as.numeric(logLik(apim_dim_ILD))
  )
)
```

## Including Random Slopes

Random-slope DIM and APIM parameterizations can be written analogously by adding
the corresponding within-person effects to the stable dyad-level random-effect
blocks. These larger models do not converge cleanly with the example data, so
the chunks are not evaluated. Such models can be fit when the study design,
data, and convergence diagnostics support the added complexity.

A random-slope standard deviation describes how much the corresponding
association varies across dyads around its fixed effect. Correlations within a
random-effects block describe how its latent dyad-specific effects co-vary; they
are not correlations between observed predictors or partners. Separate
random-effects terms define separate covariance blocks, so correlations across
those blocks are constrained to zero, preserving the exchangeability condition
described above.

For an exchangeable APIM, the sum-difference random-effects representation
involving `.i_diff_*` can be rotated back to a constrained actor-partner
representation. This can make actor and partner random effects easier to
interpret, while offering no analogous interpretive advantage for the DIM. The
[APIM vignette](apim.html) describes this rotation and the constraints required
to preserve exchangeability.

In the APIM:

```{r fit-ild-apim-random, eval = FALSE}

apim_ILD_random <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person APIM
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # Stable dyad-level covariance with within-person random slopes
    (1 +
         .i_provided_support_cwp_actor +
         .i_provided_support_cwp_partner
       | coupleID)  +
    (0 +
         .i_diff_assumed_exchangeable_arbitrary +
         .i_diff_assumed_exchangeable_arbitrary:.i_provided_support_cwp_actor +
         .i_diff_assumed_exchangeable_arbitrary:.i_provided_support_cwp_partner
       | coupleID) +

    # Same-day exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

In the DIM:

```{r fit-ild-dim-random, eval = FALSE}

dim_ILD_random <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person DIM
    .i_provided_support_cwp_dyad_mean +
    .i_provided_support_cwp_within_dyad_dev +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_dev +

    # Stable dyad-level covariance with within-person random slopes
    (1 +
       .i_provided_support_cwp_dyad_mean +
       .i_provided_support_cwp_within_dyad_dev
     | coupleID)  +
    (0 +
       .i_diff_assumed_exchangeable_arbitrary +
       .i_diff_assumed_exchangeable_arbitrary:.i_provided_support_cwp_dyad_mean +
       .i_diff_assumed_exchangeable_arbitrary:.i_provided_support_cwp_within_dyad_dev
     | coupleID) +

    # Same-day exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

---

**Continue** with the [Dyadic Score Model vignette](dsm.html),

refer to the: 

- [Actor-Partner Interdependence Model (APIM) vignette](apim.html),
- [Mixed-Composition APIM vignette](mixed-apim.html),

or return to the [Overview](index.html).
