---
title: "Dyad-Individual Model"
description: >
  Construction and interpretation of DIM predictors, including their
  relationship to APIM effects in dyadic multilevel models for exchangeable
  dyads.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dyad-Individual Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

This vignette focuses on the Dyad-Individual Model (DIM) for dyadic multilevel
models and its relationship to the Actor-Partner Interdependence Model (APIM).
The DIM separates a predictor into the dyad's shared level and each member's
deviation from that level.

For broader guidance on using this package for data preparation for various
dyadic model types, especially APIMs with multiple dyad types, generalized
outcomes, intensive longitudinal models, and optimizer choices, see the
[Getting Started vignette](getting-started.html) and the
[Actor-Partner Interdependence Model vignette](apim.html). For undirected
dyadic score outcomes, see the
[Undirected Dyadic Score Model vignette](undirected-dsm.html).

## Cross-Sectional Gaussian DIM

The current DIM implementation prepares predictors for one exchangeable dyad
composition. Exchangeability means that swapping the two member labels does not
change the model. One way to make this assumption is to omit `role` from
`prepare_interdep_data()`, which treats all dyads as the same exchangeable
composition.

```{r prepare-cross-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  predictors = communication,
  # Create both APIM and DIM columns for comparison.
  model_type = c("apim", "dim"),
  seed = 123
)

# Print the first two dyads.
print(cross_exchangeable_data, n = 4)
```

The fixed `seed` makes the arbitrary member labels used to construct the
exchangeable-dyad difference contrast `.i_diff_*` reproducible.

Passing a `role` is also possible when it leads to exactly one
exchangeable composition (e.g., only female-female dyads).

In a dataset with mixed dyad types, the right preparation depends on the analysis goal.

For instance, we can filter dyads and keep only male-male and female-female
dyads. Because DIM currently supports a single type of exchangeable dyad, we
might pool those like this:

```{r prepare-cross-filter-pool-same-sex}
cross_same_sex_pooled_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = satisfaction,
  model_type = c("apim", "dim"),
  # Remove male-female dyads.
  include_compositions = c("male-male", "female-female"),
  pool_compositions = list(
    "same-sex-couples" = c("male-male", "female-female")
  ),
  seed = 123
)

print(cross_same_sex_pooled_data, n = 4)
```

If we want to include only male-female couples and treat those as exchangeable,
we can do:

```{r prepare-cross-filter-set-male-female-exchangeable}
cross_male_female_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = satisfaction,
  model_type = c("apim", "dim"),
  include_compositions = "male-female",
  set_exchangeable_compositions = "male-female",
  seed = 123
)

print(cross_male_female_exchangeable_data, n = 4)
```

Pooling compositions constrains them to share DIM effects. These are substantive
modeling decisions, not only data-processing choices, and should follow the
research question.

### Example DIM Model

For members \(j \in \{1, 2\}\) of dyad \(i\), the DIM decomposition is

$$
\bar{x}_i = \frac{x_{i1} + x_{i2}}{2}, \qquad
d_{ij} = x_{ij} - \bar{x}_i.
$$

The two deviations have equal magnitude and opposite signs:
\(d_{i1} = -d_{i2}\). The variables that enter the DIM fixed effects are:

1. A dyad-mean variable that is grand-mean centered. This describes a couple's
shared level compared to all other couples.
2. A within-dyad variable describing each partner's deviation from the couple
mean, equivalently half the signed difference between partners.

The fixed effects are a reparameterization of the APIM actor and partner
effects. The same random-effects structure can therefore be used for both
fixed-effect parameterizations: a dyad-level intercept and a dyad-level
difference contrast indexed by `.i_diff_assumed_exchangeable`. In `glmmTMB`,
with `dispformula = ~ 0`, these random effects represent the two members'
Gaussian residual variance and covariance.

The intercept and difference contrast are specified as separate random-effects
terms, which constrains their correlation to zero. This preserves
exchangeability: swapping the arbitrary member labels leaves the intercept term
unchanged but reverses the difference contrast. A nonzero correlation between
them would therefore make the random-effects distribution depend on the
arbitrary labeling.

```{r fit-cross-dim, eval=has_glmmTMB}

dim_1 <- glmmTMB::glmmTMB(
  satisfaction ~

    # Pooled fixed intercept
    1 +

    # Between-couple effect
    .i_communication_raw_dyad_mean_gmc +

    # Within-couple effect
    .i_communication_raw_within_dyad_deviation +

    # Residual Gaussian variance-covariance
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable | coupleID)
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

summary(dim_1)
```

```{r glmmTMB-missing, echo = FALSE, eval = !has_glmmTMB}
message("Install glmmTMB to run the fitted-model examples in this vignette.")
```

Because this Gaussian model uses an identity link, fixed coefficients are
interpreted in satisfaction units:

- The intercept (about 5.04) is the expected satisfaction
  for a member whose communication equals the mean of a dyad with average
  communication.

- The dyad-mean estimate (about 2.00) means that a shared one-point
  difference in both members' communication is associated with a two-point
  difference in expected satisfaction, holding their difference constant.

- The within-dyad-deviation estimate (about 1.52) means that the member whose
  communication is one point higher than their partner's is expected to report
  satisfaction 1.52 points higher, holding the other model terms constant. This
  is a within-dyad association, not the expected change from increasing only one
  member's communication, which would also change the dyad mean.

The same model can be written in APIM form. Since we have requested both sets of
variables from `prepare_interdep_data`, we can fit one directly. For more
guidance on APIM specifications and different models, see the
[Actor-Partner Interdependence Model vignette](apim.html).

```{r fit-cross-apim, eval = has_glmmTMB}

apim_1 <- glmmTMB::glmmTMB(
  satisfaction ~ 1 +
    .i_communication_raw_actor + .i_communication_raw_partner +
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable | coupleID)
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

```

The two models have identical fit statistics:

```{r compare-cross-fit, eval=has_glmmTMB}
data.frame(
  model = c("DIM", "APIM"),
  AIC = c(AIC(dim_1), AIC(apim_1)),
  BIC = c(BIC(dim_1), BIC(apim_1)),
  logLik = c(as.numeric(logLik(dim_1)), as.numeric(logLik(apim_1)))
)
```

This demonstrates that the same statistical model is being estimated with different
parameterizations and coefficient interpretations.

The slope coefficients relate as follows:

$$
\beta_{\text{dyad mean}} =
\beta_{\text{actor}} + \beta_{\text{partner}}
$$

and

$$
\beta_{\text{within-dyad deviation}} =
\beta_{\text{actor}} - \beta_{\text{partner}}
$$

Conversely, the actor effect is half the sum of the two DIM effects, and the
partner effect is half their difference. The DIM and APIM intercepts are not
expected to be equal because the DIM dyad mean is grand-mean centered, whereas
the APIM predictors retain their original scale.

In this example:

```{r compare-cross-coefficients, eval=has_glmmTMB}
apim_coef <- glmmTMB::fixef(apim_1)$cond
dim_coef <- glmmTMB::fixef(dim_1)$cond

b_actor <- apim_coef[[".i_communication_raw_actor"]]
b_partner <- apim_coef[[".i_communication_raw_partner"]]

b_dyad_mean <- dim_coef[[".i_communication_raw_dyad_mean_gmc"]]
b_within_dyad <- dim_coef[[".i_communication_raw_within_dyad_deviation"]]

cat("From APIM model:\n")
cat("  actor effect:                  ", round(b_actor, 3), "\n")
cat("  partner effect:                ", round(b_partner, 3), "\n\n")

cat("  DIM transformation:\n")
cat("  actor effect + partner effect: ", round(b_actor + b_partner, 3), "\n")
cat("  actor effect - partner effect: ", round(b_actor - b_partner, 3), "\n\n")

cat("From DIM model:\n")
cat("  dyad-mean effect:              ", round(b_dyad_mean, 3), "\n")
cat("  within-dyad-deviation effect:  ", round(b_within_dyad, 3), "\n")
```

## Intensive Longitudinal DIM

For longitudinal DIM, predictors are decomposed into within-person and
between-person components before the dyadic decomposition. The default `"auto"`
strategy selects `"time_2l"` when both `time` and `predictors` are supplied:

1. The `cwp` dyad mean captures a shared occasion-specific shift from the two
   members' usual levels (shared occasion-level variation).
2. The `cwp` within-dyad deviation captures which member is further above or
   below their own usual level on that occasion.
3. The `cbp` dyad mean captures the dyad's shared usual level relative to the
   sample's grand mean (stable between-dyad differences).
4. The `cbp` within-dyad deviation captures each member's stable difference
   from the dyad's usual level.

`temporal_predictor_decomposition = "none"` is not available for longitudinal
DIM or undirected DSM predictor construction.

```{r prepare-ild-exchangeable}
ild_exchangeable_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  time = diaryday,
  predictors = provided_support,
  model_type = c("apim", "dim"),
  seed = 123
)

print(ild_exchangeable_data)
```

```{r fit-ild-dim, eval = has_glmmTMB}

dim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person DIM
    .i_provided_support_cwp_dyad_mean +
    .i_provided_support_cwp_within_dyad_deviation +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_deviation +

    # Stable exchangeable dyad-level covariance
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable | coupleID) +

    # Residual (same-day) exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

summary(dim_ILD)

```

The fitted coefficients can be interpreted as:

- The `cwp` dyad-mean estimate (about 0.49) describes the expected change in
  closeness when both members are one point above their respective usual support
  levels.

- The `cbp` dyad-mean estimate (about 1.51) instead compares dyads whose members'
  average usual support levels differ by one point.

- The `cwp` within-dyad estimate (about 0.06) describes the expected closeness
  difference associated with a one-point difference between members in their
  momentary deviations from usual support.

- The `cbp` within-dyad estimate (about 0.78) describes the expected closeness
  difference associated with a one-point difference between members in their
  usual support levels.

The equivalent APIM uses actor and partner effects on both levels:

```{r fit-ild-apim, eval = has_glmmTMB}

apim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person APIM
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # Stable exchangeable dyad-level covariance
    (1 | coupleID)  + (0 + .i_diff_assumed_exchangeable | coupleID) +

    # Residual (same-day) exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

The two ILD models again have identical fit statistics:

```{r compare-ild-fit, eval = has_glmmTMB}
data.frame(
  model = c("DIM", "APIM"),
  AIC = c(AIC(dim_ILD), AIC(apim_ILD)),
  BIC = c(BIC(dim_ILD), BIC(apim_ILD)),
  logLik = c(as.numeric(logLik(dim_ILD)), as.numeric(logLik(apim_ILD)))
)
```

The equivalence holds separately for the within-person (`cwp`) and
between-person (`cbp`) predictor components. For the within-person component:

$$
\beta_{\text{cwp dyad mean}} =
\beta_{\text{cwp actor}} + \beta_{\text{cwp partner}}
$$

$$
\beta_{\text{cwp within-dyad deviation}} =
\beta_{\text{cwp actor}} - \beta_{\text{cwp partner}}
$$

For the between-person component:

$$
\beta_{\text{cbp dyad mean}} =
\beta_{\text{cbp actor}} + \beta_{\text{cbp partner}}
$$

$$
\beta_{\text{cbp within-dyad deviation}} =
\beta_{\text{cbp actor}} - \beta_{\text{cbp partner}}
$$

This also means that an APIM parameterization can be used on one level and a
DIM parameterization on the other. For example:

```{r fit-ild-apim-within-dim-between, eval = has_glmmTMB}

apim_dim_ILD <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_deviation +

    # Stable exchangeable dyad-level covariance
    (1 | coupleID)  + (0 + .i_diff_assumed_exchangeable | coupleID) +

    # Same-day exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

This mixed parameterization still estimates the same model:

```{r compare-ild-all-fit, eval = has_glmmTMB}
data.frame(
  model = c("DIM within / DIM between", "APIM within / APIM between", "APIM within / DIM between"),
  AIC = c(AIC(dim_ILD), AIC(apim_ILD), AIC(apim_dim_ILD)),
  BIC = c(BIC(dim_ILD), BIC(apim_ILD), BIC(apim_dim_ILD)),
  logLik = c(
    as.numeric(logLik(dim_ILD)),
    as.numeric(logLik(apim_ILD)),
    as.numeric(logLik(apim_dim_ILD))
  )
)
```

## Including Random Slopes

Random-slope DIM and APIM parameterizations can be written analogously by adding
the corresponding within-person effects to the stable dyad-level random-effect
blocks. These larger models do not converge cleanly with the example data, so
the chunks are not evaluated. Such models can be fit when the study design,
data, and convergence diagnostics support the added complexity.

A random-slope standard deviation describes how much the corresponding
association varies across dyads around its fixed effect. Correlations within a
random-effects block describe how its latent dyad-specific effects co-vary; they
are not correlations between observed predictors or partners. Separate
random-effects terms define separate covariance blocks, so correlations across
those blocks are constrained to zero, preserving the exchangeability condition
described above.

For an exchangeable APIM, the sum-difference random-effects representation
involving `.i_diff_*` can be rotated back to a constrained actor-partner
representation. This can make actor and partner random effects easier to
interpret, while offering no analogous interpretive advantage for the DIM. The
[APIM vignette](apim.html) describes this rotation and the constraints required
to preserve exchangeability.

In the APIM:

```{r fit-ild-apim-random, eval = FALSE}

apim_ILD_random <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person APIM
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # Stable dyad-level covariance with within-person random slopes
    (1 +
         .i_provided_support_cwp_actor +
         .i_provided_support_cwp_partner
       | coupleID)  +
    (0 +
         .i_diff_assumed_exchangeable +
         .i_diff_assumed_exchangeable:.i_provided_support_cwp_actor +
         .i_diff_assumed_exchangeable:.i_provided_support_cwp_partner
       | coupleID) +

    # Same-day exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

In the DIM:

```{r fit-ild-dim-random, eval = FALSE}

dim_ILD_random <- glmmTMB::glmmTMB(
  closeness ~
    1 +

    diaryday +

    # Within-person DIM
    .i_provided_support_cwp_dyad_mean +
    .i_provided_support_cwp_within_dyad_deviation +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_deviation +

    # Stable dyad-level covariance with within-person random slopes
    (1 +
       .i_provided_support_cwp_dyad_mean +
       .i_provided_support_cwp_within_dyad_deviation
     | coupleID)  +
    (0 +
       .i_diff_assumed_exchangeable +
       .i_diff_assumed_exchangeable:.i_provided_support_cwp_dyad_mean +
       .i_diff_assumed_exchangeable:.i_provided_support_cwp_within_dyad_deviation
     | coupleID) +

    # Same-day exchangeable dyad-level covariance
    (1 | coupleID:diaryday) +
    (0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

For APIM formulas with distinguishable, exchangeable, generalized, and
mixed-composition dyads, see the
[Actor-Partner Interdependence Model vignette](apim.html). For undirected dyadic
score outcomes, see the
[Undirected Dyadic Score Model vignette](undirected-dsm.html). For the broader
data-preparation workflow, return to the
[Getting Started vignette](getting-started.html).
