---
title: "Dyadic Score Model (DSM)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dyadic Score Model (DSM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```

This vignette focuses on undirected Dyadic Score Model (DSM) preparation. 

For the broader data-preparation workflow of the `interdep` package, see the
[Getting Started vignette](getting-started.html). For distinguishable,
exchangeable, generalized, and intensive longitudinal APIMs, see the
[Actor-Partner Interdependence Model vignette](apim.html). For APIMs that
combine distinguishable and exchangeable dyad compositions, see the
[Mixed-Composition APIM vignette](mixed-apim.html). For the dyad-individual
model (DIM), see the [Dyad-Individual Model vignette](dim.html). Whereas the DSM 
is a different parametrization of the distinguishable APIM, the Dyadic Score 
Model (DSM) is a different parametrisation of the exchangeable APIM. 


> This vignette is under construction. A directional DSM implementation is planned.

---

Refer to the: 

- [Actor-Partner Interdependence Model (APIM) vignette](apim.html),
- [Mixed-Composition APIM vignette](mixed-apim.html), or
- [Dyad-Individual Model vignette](dim.html),

or return to the [Overview](index.html).
