---
title: "Dyadic Score Model (DSM)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dyadic Score Model (DSM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```

This vignette focuses on the Dyadic Score Model (DSM) for distinguishable
dyads and its relationship to the distinguishable Actor-Partner
Interdependence Model (APIM). The DSM expresses associations in terms of the
dyad's shared level and the directional difference between partners.

For the broader data-preparation workflow, see the
[Getting Started vignette](getting-started.html). For distinguishable,
exchangeable, generalized, and intensive longitudinal APIMs, see the
[Actor-Partner Interdependence Model vignette](apim.html). For APIMs combining
distinguishable and exchangeable compositions, see the
[Mixed-Composition APIM vignette](mixed-apim.html). For the exchangeable
Dyad-Individual Model (DIM), see the
[Dyad-Individual Model vignette](dim.html).

## Preparing DSM Data

A DSM requires a substantively meaningful direction. We therefore supply the
two roles in `dsm_role_order`. Here, `c("female", "male")` defines every
difference as female minus male, independently of row order.

```{r prepare-cross-dsm, eval=FALSE}
cross_dsm_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  model_type = "dsm",
  dsm_role_order = c("female", "male")
)

print(cross_dsm_data, n = 4)
```

For predictor values `X_female` and `X_male`, `interdep` creates:

```text
.i_communication_dyad_mean_gmc
  = (X_female + X_male) / 2, grand-mean centered

.i_communication_dyad_difference
  = X_female - X_male

.i_dsm_role_contrast
  = +0.5 for female and -0.5 for male
```

The dyad mean and signed difference are repeated on both member rows. The
original member-level outcome remains unchanged and is selected in the model
formula. No outcome argument or outcome transformation is needed.

The `+0.5/-0.5` contrast is important: coefficients involving the contrast
equal the full female-minus-male outcome difference. Reversing
`dsm_role_order` reverses all directional differences and the contrast, but
does not change fitted member outcomes.

## Cross-Sectional Gaussian DSM

Let `XLevel` denote the centered predictor dyad mean, `XDiff` the signed
predictor difference, and `C` the DSM role contrast. Iida et al.'s full DSM can
be fitted to the unchanged member-level outcome using:

```text
outcome ~ XLevel + XDiff + C + XLevel:C + XDiff:C
```

For a cross-sectional Gaussian DSM, a correlated dyad random intercept and
role-contrast slope represent unexplained outcome-level and outcome-difference
variation. The ordinary member-level residual variance is suppressed because
the three random-effect covariance parameters already span the three unique
elements of the two-member residual covariance matrix.

```{r fit-cross-dsm, eval=FALSE}
dsm_model <- glmmTMB::glmmTMB(
  satisfaction ~
    .i_communication_dyad_mean_gmc +
    .i_communication_dyad_difference +
    .i_dsm_role_contrast +
    .i_communication_dyad_mean_gmc:.i_dsm_role_contrast +
    .i_communication_dyad_difference:.i_dsm_role_contrast +
    (1 + .i_dsm_role_contrast | coupleID),
  dispformula = ~ 0,
  family = gaussian(),
  data = cross_dsm_data
)

summary(dsm_model)
```

In this example, the fixed effects have the following interpretations:

1. **Intercept:** the expected dyad-average satisfaction when the couple has
   the sample-average communication level and no female-male communication
   difference.
2. **Communication dyad mean:** the difference in dyad-average satisfaction
   associated with a one-unit higher shared communication level, holding the
   communication difference constant.
3. **Communication difference:** the difference in dyad-average satisfaction
   associated with a one-unit larger female-minus-male communication
   difference, holding the shared level constant.
4. **DSM role contrast:** the expected female-minus-male satisfaction
   difference when the shared communication level is at its reference value
   and the communication difference is zero.
5. **Dyad mean by role contrast:** the change in the female-minus-male
   satisfaction difference associated with a one-unit higher shared
   communication level.
6. **Communication difference by role contrast:** the change in the
   female-minus-male satisfaction difference associated with a one-unit larger
   female-minus-male communication difference.

The third and fifth coefficients are the DSM cross-paths: predictor difference
to outcome level and predictor level to outcome difference. They are omitted
from the reduced DSM but are needed for the full model.

The random intercept variance is unexplained variation in outcome level, and
the random role-contrast slope variance is unexplained variation in the full
directional outcome difference. Their covariance indicates whether unexplained
outcome level and unexplained outcome difference are associated.

## Relationship to the APIM and DIM

For distinguishable dyads, the full DSM and an unconstrained distinguishable
APIM are alternative parameterizations of the same fixed associations. The
APIM estimates separate actor and partner effects for each role. The DSM
re-expresses those four slopes as:

- predictor level to outcome level;
- predictor difference to outcome level;
- predictor level to outcome difference; and
- predictor difference to outcome difference.

Neither parameterization is generally more correct. The APIM is often clearer
for questions about actor and partner effects, whereas the DSM directly
addresses questions about shared dyadic levels and directional discrepancies.

For exchangeable dyads, the direction of a member difference is arbitrary.
The directional intercept, both cross-paths, and the covariance between outcome
level and outcome difference must then be zero. The remaining reduced,
label-invariant DSM is algebraically the Gaussian DIM. In `interdep`, use
`model_type = "dim"` for this exchangeable model and reserve
`model_type = "dsm"` for distinguishable dyads.

---

Refer to the:

- [Actor-Partner Interdependence Model (APIM) vignette](apim.html),
- [Mixed-Composition APIM vignette](mixed-apim.html), or
- [Dyad-Individual Model vignette](dim.html),

or return to the [Overview](index.html).
