---
title: "Dyadic Score Model (DSM)"
author: "Pascal Küng"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dyadic Score Model (DSM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

This vignette focuses on the Dyadic Score Model (DSM) for distinguishable
dyads and its relationship to the distinguishable Actor-Partner
Interdependence Model (APIM). The DSM expresses associations in terms of the
dyad's shared level and the directional difference between partners
[@iidaModelsInterdependentIndividuals2018].

For the broader data-preparation workflow, see the
[Getting Started vignette](getting-started.html). For distinguishable and
exchangeable APIMs in cross-sectional and intensive longitudinal data, see the
[Actor-Partner Interdependence Model vignette](apim.html). For APIMs combining
distinguishable and exchangeable compositions, see the
[Mixed-Composition APIM vignette](mixed-apim.html). For the exchangeable
Dyad-Individual Model (DIM), see the
[Dyad-Individual Model vignette](dim.html).

## Preparing DSM Data

A DSM requires a substantively meaningful direction. We therefore supply the
two roles in `dsm_role_order`. In this example, `c("female", "male")` defines every
difference as female minus male.

```{r prepare-cross-dsm}
cross_dsm_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  model_type = "dsm",
  dsm_role_order = c("female", "male")
)

print(cross_dsm_data, n = 4)
```

For predictor values $X_{female}$ and $X_{male}$, `interdep` then creates:

- `.i_communication_dyad_mean_gmc` $=
  \frac{X_{female} + X_{male}}{2} - \mu_X$ (with $\mu_X$ representing the sample grand
  mean of the dyad-level predictor means)
  
- `.i_communication_within_dyad_diff` $= X_{female} - X_{male}$

- `.i_dsm_role_contrast` $= +0.5$ for female and $-0.5$ for male. 

The dyad mean and signed difference are repeated on both member rows. The
outcome remains unchanged and no transformation is needed.

## Cross-Sectional Gaussian DSM

For a cross-sectional Gaussian DSM, a correlated dyad random intercept and
role-contrast slope represent unexplained outcome-level and outcome-difference
variation. 

```{r fit-cross-dsm, eval=has_glmmTMB}
dsm_model <- glmmTMB::glmmTMB(
  satisfaction ~

    # Outcome-level intercept (also estimated in the DIM)
    1 +
    .i_communication_dyad_mean_gmc +
    .i_communication_within_dyad_diff +
    .i_dsm_role_contrast +
    .i_communication_dyad_mean_gmc:.i_dsm_role_contrast +
    .i_communication_within_dyad_diff:.i_dsm_role_contrast +

    # Outcome-level and outcome-difference residual variances and their covariance
    (1 + .i_dsm_role_contrast | coupleID),
  dispformula = ~ 0,
  family = gaussian(),
  data = cross_dsm_data
)

summary(dsm_model)
```

### Interpreting the DSM paths

For the outcomes given the predictors, the long-format model estimates the same
paths as the conventional score-based DSM
[@iidaModelsInterdependentIndividuals2018].

In the conventional score-based representation, the **predictors** are
decomposed in the same way:

$$
X_{mean} = \frac{X_{female} + X_{male}}{2} - \mu_X, \qquad
X_{diff} = X_{female} - X_{male},
$$

where $\mu_X$ is the sample grand mean of the dyad-level predictor means.

The **outcomes** are also decomposed:

$$
Y_{mean} = \frac{Y_{female} + Y_{male}}{2}, \qquad
Y_{diff} = Y_{female} - Y_{male}.
$$

The long-format model fitted here does not create $Y_{mean}$ and $Y_{diff}$ as
observed variables. Instead, it uses the member-level outcome directly. With
complete outcome pairs, this is an equivalent parameterization of the same
conditional outcome regressions.

Consider the conceptual SEM formulas:

$$
\widehat{Y_{mean}} = a_{10} + a_{11}X_{mean} + a_{12}X_{diff},
$$

$$
\widehat{Y_{diff}} = a_{20} + a_{21}X_{mean} + a_{22}X_{diff}.
$$

The fixed effects from our MLM model map directly to these paths as such:

| Long-format fixed effect | DSM SEM path and interpretation |
|---|---|
| Intercept | $a_{10}$: expected dyad-average satisfaction at the sample-average communication level and no female-male communication difference |
| Communication dyad mean | $a_{11}$: predictor level $\rightarrow$ outcome level |
| Communication difference | $a_{12}$: predictor difference $\rightarrow$ outcome level |
| DSM role contrast | $a_{20}$: expected female-minus-male outcome difference at the predictor reference values |
| Dyad mean $\times$ role contrast | $a_{21}$: predictor level $\rightarrow$ outcome difference |
| Communication difference $\times$ role contrast | $a_{22}$: predictor difference $\rightarrow$ outcome difference |

Thus, for example, $a_{12}$ is the change in dyad-average satisfaction
associated with a one-unit larger female-minus-male communication difference,
holding communication level constant. In contrast, $a_{22}$ is the change in
the female-minus-male satisfaction difference associated with that same
one-unit larger communication difference, holding communication level
constant. The $a_{12}$ and $a_{21}$ coefficients are the DSM cross-paths. They
are omitted from the reduced DSM but are needed for the full model.

The random intercept variance is unexplained variation in outcome level, and
the random role-contrast slope variance is unexplained variation in the full
directional outcome difference. Their covariance indicates whether unexplained
outcome level and unexplained outcome difference are associated.

### Reversing the coding

Instead of computing $X_{female} - X_{male}$, we can reverse the direction and
compute $X_{male} - X_{female}$. This changes the direction of the differences,
but not the substantive model.

```{r prepare-cross-dsm-inverted}
cross_dsm_data_inverted <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  # Request APIM columns too for comparison below.
  model_type = c("dsm", "apim"),
  dsm_role_order = c("male", "female")
)
```

```{r fit-cross-dsm-inverted, eval=has_glmmTMB}
dsm_model_inverted <- glmmTMB::glmmTMB(
  satisfaction ~
    .i_communication_dyad_mean_gmc +
    .i_communication_within_dyad_diff +
    .i_dsm_role_contrast +
    .i_communication_dyad_mean_gmc:.i_dsm_role_contrast +
    .i_communication_within_dyad_diff:.i_dsm_role_contrast +
    (1 + .i_dsm_role_contrast | coupleID),
  dispformula = ~ 0,
  family = gaussian(),
  data = cross_dsm_data_inverted
)

female_minus_male <- glmmTMB::fixef(dsm_model)$cond
male_minus_female <- glmmTMB::fixef(dsm_model_inverted)$cond

knitr::kable(
  data.frame(
    `model term` = names(female_minus_male),
    `female - male` = unname(female_minus_male),
    `male - female` = unname(male_minus_female),
    check.names = FALSE
  ),
  digits = 3,
  align = c("l", "r", "r")
)
```

The two models have identical fitted values and model fit: 

- `.i_communication_within_dyad_diff` reverses because the predictor difference
  reverses.

- `.i_dsm_role_contrast` reverses because the represented outcome difference
  reverses.

- `.i_communication_dyad_mean_gmc:.i_dsm_role_contrast` reverses because only
  the outcome difference reverses.

- `.i_communication_within_dyad_diff:.i_dsm_role_contrast` remains unchanged
  because both differences reverse.

The intercept and `.i_communication_dyad_mean_gmc` also remain unchanged. For
the random effects, the variances of `(Intercept)` and
`.i_dsm_role_contrast` remain unchanged, whereas their covariance reverses
sign.

## Relationship to the APIM and DIM

For distinguishable dyads, the full DSM and an unconstrained distinguishable
APIM are alternative parameterizations of the same fixed associations
[@iidaModelsInterdependentIndividuals2018].

Let's fit the equivalent distinguishable APIM:

```{r fit-cross-apim, eval=has_glmmTMB}
apim_model <- glmmTMB::glmmTMB(
  satisfaction ~
    # Role-specific intercepts
    0 +
    .i_is_female_x_male_female +
    .i_is_female_x_male_male +

    # Role-specific actor effects
    .i_is_female_x_male_female:.i_communication_actor +
    .i_is_female_x_male_male:.i_communication_actor +

    # Role-specific partner effects
    .i_is_female_x_male_female:.i_communication_partner +
    .i_is_female_x_male_male:.i_communication_partner +

    # Role-specific Gaussian residual variance-covariance
    us(0 +
         .i_is_female_x_male_female +
         .i_is_female_x_male_male
       | coupleID),
  dispformula = ~ 0,
  family = gaussian(),
  data = cross_dsm_data_inverted
)
```

The two DSM directions and the APIM have identical fit statistics:

```{r compare-cross-fit, eval=has_glmmTMB}
data.frame(
  model = c("DSM: female - male", "DSM: male - female", "APIM"),
  AIC = round(c(AIC(dsm_model), AIC(dsm_model_inverted), AIC(apim_model)), 3),
  BIC = round(c(BIC(dsm_model), BIC(dsm_model_inverted), BIC(apim_model)), 3),
  logLik = round(c(
    as.numeric(logLik(dsm_model)),
    as.numeric(logLik(dsm_model_inverted)),
    as.numeric(logLik(apim_model))
  ), 3)
)
```

### Fixed-effect transformation

Let $A_f$ and $A_m$ denote the female and male actor effects, and let
$P_f$ and $P_m$ denote the partner effects on the female and male outcomes,
respectively. Let $\alpha_f$ and $\alpha_m$ be the corresponding APIM
intercepts. The APIM slopes transform into the female-minus-male DSM paths as

$$
\begin{aligned}
a_{11} &= \frac{A_f + P_f + A_m + P_m}{2}, &
a_{12} &= \frac{A_f - P_f + P_m - A_m}{4}, \\
a_{21} &= A_f + P_f - A_m - P_m, &
a_{22} &= \frac{A_f + A_m - P_f - P_m}{2}.
\end{aligned}
$$

The APIM predictors retain their original scale, whereas the DSM predictor
level is grand-mean centered. If $\mu_X$ is the grand mean subtracted from the
DSM predictor level, the intercepts therefore transform as

$$
a_{10} = \frac{\alpha_f + \alpha_m}{2} + \mu_X a_{11}, \qquad
a_{20} = \alpha_f - \alpha_m + \mu_X a_{21}.
$$

The reverse slope transformation is

$$
\begin{aligned}
A_f &= \frac{a_{11}}{2} + a_{12} + \frac{a_{21}}{4} + \frac{a_{22}}{2}, &
P_f &= \frac{a_{11}}{2} - a_{12} + \frac{a_{21}}{4} - \frac{a_{22}}{2}, \\
A_m &= \frac{a_{11}}{2} - a_{12} - \frac{a_{21}}{4} + \frac{a_{22}}{2}, &
P_m &= \frac{a_{11}}{2} + a_{12} - \frac{a_{21}}{4} - \frac{a_{22}}{2}.
\end{aligned}
$$

Similarly,

$$
\alpha_f = a_{10} + \frac{a_{20}}{2}
- \mu_X\left(a_{11} + \frac{a_{21}}{2}\right),
$$

$$
\alpha_m = a_{10} - \frac{a_{20}}{2}
- \mu_X\left(a_{11} - \frac{a_{21}}{2}\right).
$$

The following comparison applies the APIM-to-DSM transformation to all six
fixed effects:

```{r compare-cross-coefficients, eval=has_glmmTMB, echo=FALSE}
apim_coef <- glmmTMB::fixef(apim_model)$cond
dsm_coef <- glmmTMB::fixef(dsm_model)$cond

alpha_f <- apim_coef[[".i_is_female_x_male_female"]]
alpha_m <- apim_coef[[".i_is_female_x_male_male"]]
A_f <- apim_coef[[
  ".i_is_female_x_male_female:.i_communication_actor"
]]
A_m <- apim_coef[[
  ".i_is_female_x_male_male:.i_communication_actor"
]]
P_f <- apim_coef[[
  ".i_is_female_x_male_female:.i_communication_partner"
]]
P_m <- apim_coef[[
  ".i_is_female_x_male_male:.i_communication_partner"
]]

mu_X <- mean(
  (cross_dsm_data_inverted$.i_communication_actor +
     cross_dsm_data_inverted$.i_communication_partner) / 2 -
    cross_dsm_data_inverted$.i_communication_dyad_mean_gmc,
  na.rm = TRUE
)

a11_from_apim <- (A_f + P_f + A_m + P_m) / 2
a12_from_apim <- (A_f - P_f + P_m - A_m) / 4
a21_from_apim <- A_f + P_f - A_m - P_m
a22_from_apim <- (A_f + A_m - P_f - P_m) / 2

apim_to_dsm <- c(
  a10 = (alpha_f + alpha_m) / 2 + mu_X * a11_from_apim,
  a11 = a11_from_apim,
  a12 = a12_from_apim,
  a20 = alpha_f - alpha_m + mu_X * a21_from_apim,
  a21 = a21_from_apim,
  a22 = a22_from_apim
)

a10_dsm <- dsm_coef[["(Intercept)"]]
a11_dsm <- dsm_coef[[".i_communication_dyad_mean_gmc"]]
a12_dsm <- dsm_coef[[".i_communication_within_dyad_diff"]]
a20_dsm <- dsm_coef[[".i_dsm_role_contrast"]]
a21_dsm <- dsm_coef[[
  ".i_communication_dyad_mean_gmc:.i_dsm_role_contrast"
]]
a22_dsm <- dsm_coef[[
  ".i_communication_within_dyad_diff:.i_dsm_role_contrast"
]]

cat(
  "From APIM model:\n",
  "  female intercept:       ", round(alpha_f, 3), "\n",
  "  female actor effect:    ", round(A_f, 3), "\n",
  "  female partner effect:  ", round(P_f, 3), "\n",
  "  male intercept:         ", round(alpha_m, 3), "\n",
  "  male actor effect:      ", round(A_m, 3), "\n",
  "  male partner effect:    ", round(P_m, 3), "\n",
  "  centering constant:     ", round(mu_X, 3), "\n\n",

  "DSM transformation:\n",
  "  a10 = (female intercept + male intercept) / 2 +\n",
  "         mu_X * a11:                                      ",
  round(apim_to_dsm[["a10"]], 3), "\n",
  "  a11 = (female actor + female partner + male actor +\n",
  "         male partner) / 2:                               ",
  round(apim_to_dsm[["a11"]], 3), "\n",
  "  a12 = (female actor - female partner + male partner -\n",
  "         male actor) / 4:                                 ",
  round(apim_to_dsm[["a12"]], 3), "\n",
  "  a20 = female intercept - male intercept + mu_X * a21:   ",
  round(apim_to_dsm[["a20"]], 3), "\n",
  "  a21 = female actor + female partner - male actor -\n",
  "         male partner:                                    ",
  round(apim_to_dsm[["a21"]], 3), "\n",
  "  a22 = (female actor + male actor - female partner -\n",
  "         male partner) / 2:                               ",
  round(apim_to_dsm[["a22"]], 3), "\n\n",

  "From DSM model:\n",
  "  outcome-level intercept (a10):       ", round(a10_dsm, 3), "\n",
  "  level -> level (a11):                ", round(a11_dsm, 3), "\n",
  "  difference -> level (a12):           ", round(a12_dsm, 3), "\n",
  "  outcome-difference intercept (a20):  ", round(a20_dsm, 3), "\n",
  "  level -> difference (a21):           ", round(a21_dsm, 3), "\n",
  "  difference -> difference (a22):      ", round(a22_dsm, 3), "\n"
)
```

### Random-effect transformation

Let $u_f$ and $u_m$ be the APIM random effects for the female and male
outcomes. The DSM outcome-level and outcome-difference residuals are the same
random effects expressed in different coordinates:

$$
\begin{pmatrix} r_{YL} \\ r_{YD} \end{pmatrix}
=
\begin{pmatrix} 1/2 & 1/2 \\ 1 & -1 \end{pmatrix}
\begin{pmatrix} u_f \\ u_m \end{pmatrix}.
$$

Applying this rotation to the APIM covariance matrix reproduces the DSM random
intercept variance, intercept-slope covariance, and role-slope variance:

```{r compare-cross-random-effects, eval=has_glmmTMB}
apim_vcov <- as.matrix(glmmTMB::VarCorr(apim_model)$cond$coupleID)
dsm_vcov <- as.matrix(glmmTMB::VarCorr(dsm_model)$cond$coupleID)

rotation <- rbind(
  outcome_level = c(0.5, 0.5),
  outcome_difference = c(1, -1)
)
apim_to_dsm_vcov <- rotation %*% apim_vcov %*% t(rotation)

data.frame(
  parameter = c("Var(r_YL)", "Cov(r_YL, r_YD)", "Var(r_YD)"),
  from_DSM = round(c(
    dsm_vcov[1, 1],
    dsm_vcov[1, 2],
    dsm_vcov[2, 2]
  ), 3),
  from_APIM_transformation = round(c(
    apim_to_dsm_vcov[1, 1],
    apim_to_dsm_vcov[1, 2],
    apim_to_dsm_vcov[2, 2]
  ), 3)
)
```

For exchangeable dyads, the direction of a member difference is arbitrary.
The directional intercept, both cross-paths, and the covariance between outcome
level and outcome difference must then be zero. The remaining reduced,
label-invariant DSM is algebraically the Gaussian Dyad-Individual Model (DIM).
In `interdep`, use `model_type = "dim"` for this exchangeable model and reserve
`model_type = "dsm"` for distinguishable dyads.

Because the Gaussian DIM is the exchangeability-constrained version of the full
DSM, exchangeability can also be tested by comparing these nested models. This
is equivalent to the comparison shown in [Testing distinguishability in the
APIM vignette](apim.html#testing-distinguishability).

## Intensive longitudinal DSM

### Concurrent ILD Gaussian DSM 

Observed person means used to construct the between-person (`cbp`) predictors
can be unreliable when each member contributes few occasions, which can bias
between-person estimates [@gottfredsonStraightforwardApproachCoping2019].

### Current limitations of dyadic ILD designs in R

Short, link to APIM!

#### Dynamic models

Short, link to APIM! Just provide one simple example for how the dynamic 
version can be implemented.


---

Return to the [Actor-Partner Interdependence Model vignette](apim.html), see
the [Mixed-Composition APIM vignette](mixed-apim.html) or the
[Dyad-Individual Model vignette](dim.html) for related model specifications,
or return to the [Overview](index.html).

## References

::: {#refs}
:::
