---
title: "Dyadic Score Model (DSM)"
author: "Pascal Küng"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dyadic Score Model (DSM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

diagram_helper <- if (file.exists("diagram-helpers.Rinc")) {
  "diagram-helpers.Rinc"
} else {
  file.path("vignettes", "diagram-helpers.Rinc")
}
sys.source(diagram_helper, envir = knitr::knit_global())
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

This vignette focuses on the Dyadic Score Model (DSM) for distinguishable
dyads and its relationship to the distinguishable Actor-Partner
Interdependence Model (APIM). The DSM expresses associations in terms of the
dyad's shared level and the directional difference between partners
[@iidaModelsInterdependentIndividuals2018].

For the broader data-preparation workflow, see the
[Getting Started vignette](getting-started.html). For distinguishable and
exchangeable APIMs in cross-sectional and intensive longitudinal data, see the
[Actor-Partner Interdependence Model vignette](apim.html). For APIMs combining
distinguishable and exchangeable compositions, see the
[Mixed-Composition APIM vignette](mixed-apim.html). For the exchangeable
Dyad-Individual Model (DIM), see the
[Dyad-Individual Model vignette](dim.html).

## Preparing DSM Data

A DSM requires a substantively meaningful direction. We therefore supply the
two roles in `dsm_role_order`. In this example, `c("female", "male")` defines every
difference as female minus male.

```{r prepare-cross-dsm}
cross_dsm_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  model_type = "dsm",
  dsm_role_order = c("female", "male")
)

print(cross_dsm_data, n = 4)
```

For predictor values $X_{\mathrm{female}}$ and $X_{\mathrm{male}}$, `interdep`
then creates:

- `.i_communication_dyad_mean_gmc` $=
  \frac{X_{\mathrm{female}} + X_{\mathrm{male}}}{2} - \mu_X$ (with $\mu_X$ representing the sample grand
  mean of the dyad-level predictor means)
  
- `.i_communication_within_dyad_diff` $= X_{\mathrm{female}} - X_{\mathrm{male}}$

- `.i_dsm_role_contrast` $= +0.5$ for female and $-0.5$ for male. 

The dyad mean and signed difference are repeated on both member rows. The
outcome remains unchanged and no transformation is needed.

## Cross-Sectional Gaussian DSM

For a cross-sectional Gaussian DSM, a correlated dyad random intercept and
role-contrast slope represent unexplained outcome-level and outcome-difference
variation. 

```{r conceptual-dsm-diagram, echo=FALSE, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Conceptual cross-sectional DSM. Predictor level and predictor difference each predict both outcome level and outcome difference. The intercepts are a10 and a20; the four regression paths are a11, a12, a21, and a22. The outcome-level and outcome-difference residuals may covary.", fig.alt="Path diagram for a cross-sectional dyadic score model. Communication level and female-minus-male communication difference each predict satisfaction level and female-minus-male satisfaction difference. Paths are labelled a11, a12, a21, and a22, and outcome intercepts are labelled a10 and a20."}
draw_dsm_diagram()
```

The path labels correspond directly to the terms in the model below:

```{r fit-cross-dsm, eval=has_glmmTMB}
dsm_model <- glmmTMB::glmmTMB(
  satisfaction ~

    # Outcome-level intercept
    1 +

    # Predictor level -> outcome level (a11)
    .i_communication_dyad_mean_gmc +

    # Predictor difference -> outcome level (a12)
    .i_communication_within_dyad_diff +

    # Outcome-difference intercept (a20)
    .i_dsm_role_contrast +

    # Predictor level -> outcome difference (a21)
    .i_communication_dyad_mean_gmc:.i_dsm_role_contrast +

    # Predictor difference -> outcome difference (a22)
    .i_communication_within_dyad_diff:.i_dsm_role_contrast +

    # Outcome-level and outcome-difference residual variances and their covariance
    us(1 + .i_dsm_role_contrast | coupleID),
  dispformula = ~ 0,
  family = gaussian(),
  data = cross_dsm_data
)

summary(dsm_model)
```

### Interpreting the DSM paths

For the outcomes given the predictors, the long-format model estimates the same
paths as the conventional score-based DSM
[@iidaModelsInterdependentIndividuals2018].

In the conventional score-based representation, the **predictors** are
decomposed in the same way:

$$
X_{\mathrm{mean}}
= \frac{X_{\mathrm{female}} + X_{\mathrm{male}}}{2} - \mu_X,
\qquad
X_{\mathrm{diff}} = X_{\mathrm{female}} - X_{\mathrm{male}},
$$

where $\mu_X$ is the sample grand mean of the dyad-level predictor means.

The **outcomes** are also decomposed:

$$
Y_{\mathrm{mean}} = \frac{Y_{\mathrm{female}} + Y_{\mathrm{male}}}{2},
\qquad
Y_{\mathrm{diff}} = Y_{\mathrm{female}} - Y_{\mathrm{male}}.
$$

The long-format model fitted here does not create $Y_{\mathrm{mean}}$ and
$Y_{\mathrm{diff}}$ as observed variables. Instead, it uses the member-level
outcome directly. With complete outcome pairs, this is an equivalent
parameterization of the same conditional outcome regressions.

Consider the conceptual SEM formulas:

$$
\widehat{Y_{\mathrm{mean}}}
= a_{10} + a_{11}X_{\mathrm{mean}} + a_{12}X_{\mathrm{diff}},
$$

$$
\widehat{Y_{\mathrm{diff}}}
= a_{20} + a_{21}X_{\mathrm{mean}} + a_{22}X_{\mathrm{diff}}.
$$

The fitted paths for this example are:

```{r fitted-dsm-diagram, echo=FALSE, eval=has_glmmTMB, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Fitted cross-sectional DSM for the example data. Edge and intercept labels show the estimated DSM coefficients.", fig.alt="Fitted path diagram for the example dyadic score model. Communication level and female-minus-male communication difference each predict satisfaction level and female-minus-male satisfaction difference. The six paths and intercepts are labelled with their estimated coefficients."}
dsm_coef_for_diagram <- glmmTMB::fixef(dsm_model)$cond

draw_dsm_diagram(
  estimates = c(
    a10 = dsm_coef_for_diagram[["(Intercept)"]],
    a11 = dsm_coef_for_diagram[[".i_communication_dyad_mean_gmc"]],
    a12 = dsm_coef_for_diagram[[".i_communication_within_dyad_diff"]],
    a20 = dsm_coef_for_diagram[[".i_dsm_role_contrast"]],
    a21 = dsm_coef_for_diagram[[
      ".i_communication_dyad_mean_gmc:.i_dsm_role_contrast"
    ]],
    a22 = dsm_coef_for_diagram[[
      ".i_communication_within_dyad_diff:.i_dsm_role_contrast"
    ]]
  )
)
```

The fixed effects from our MLM model map directly to these paths as such:

| Long-format fixed effect | DSM SEM path and interpretation |
|---|---|
| Intercept | $a_{10}$: expected dyad-average satisfaction at the sample-average communication level and no female-male communication difference |
| Communication dyad mean | $a_{11}$: predictor level $\rightarrow$ outcome level |
| Communication difference | $a_{12}$: predictor difference $\rightarrow$ outcome level |
| DSM role contrast | $a_{20}$: expected female-minus-male outcome difference at the predictor reference values |
| Dyad mean $\times$ role contrast | $a_{21}$: predictor level $\rightarrow$ outcome difference |
| Communication difference $\times$ role contrast | $a_{22}$: predictor difference $\rightarrow$ outcome difference |

Thus, for example, $a_{12}$ is the change in dyad-average satisfaction
associated with a one-unit larger female-minus-male communication difference,
holding communication level constant. In contrast, $a_{22}$ is the change in
the female-minus-male satisfaction difference associated with that same
one-unit larger communication difference, holding communication level
constant. The $a_{12}$ and $a_{21}$ coefficients are the DSM cross-paths. They
are omitted from the reduced DSM but are needed for the full model.

The random intercept variance is unexplained variation in outcome level, and
the random role-contrast slope variance is unexplained variation in the full
directional outcome difference. Their covariance indicates whether unexplained
outcome level and unexplained outcome difference are associated.

### Reversing the coding

Instead of computing $X_{\mathrm{female}} - X_{\mathrm{male}}$, we can reverse
the direction and compute $X_{\mathrm{male}} - X_{\mathrm{female}}$. This
changes the direction of the differences, but not the substantive model.

```{r prepare-cross-dsm-inverted}
cross_dsm_data_inverted <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  # Request APIM columns too for comparison below.
  model_type = c("dsm", "apim"),
  dsm_role_order = c("male", "female")
)
```

```{r fit-cross-dsm-inverted, eval=has_glmmTMB}
dsm_model_inverted <- glmmTMB::glmmTMB(
  satisfaction ~
    .i_communication_dyad_mean_gmc +
    .i_communication_within_dyad_diff +
    .i_dsm_role_contrast +
    .i_communication_dyad_mean_gmc:.i_dsm_role_contrast +
    .i_communication_within_dyad_diff:.i_dsm_role_contrast +
    us(1 + .i_dsm_role_contrast | coupleID),
  dispformula = ~ 0,
  family = gaussian(),
  data = cross_dsm_data_inverted
)

female_minus_male <- glmmTMB::fixef(dsm_model)$cond
male_minus_female <- glmmTMB::fixef(dsm_model_inverted)$cond

knitr::kable(
  data.frame(
    `model term` = names(female_minus_male),
    `female - male` = unname(female_minus_male),
    `male - female` = unname(male_minus_female),
    check.names = FALSE
  ),
  digits = 3,
  align = c("l", "r", "r")
)
```

The two models have identical fitted values and model fit: 

- `.i_communication_within_dyad_diff` reverses because the predictor difference
  reverses.

- `.i_dsm_role_contrast` reverses because the represented outcome difference
  reverses.

- `.i_communication_dyad_mean_gmc:.i_dsm_role_contrast` reverses because only
  the outcome difference reverses.

- `.i_communication_within_dyad_diff:.i_dsm_role_contrast` remains unchanged
  because both differences reverse.

The intercept and `.i_communication_dyad_mean_gmc` also remain unchanged. For
the random effects, the variances of `(Intercept)` and
`.i_dsm_role_contrast` remain unchanged, whereas their covariance reverses
sign.

## Relationship to the APIM and DIM

For distinguishable dyads, the full DSM and an unconstrained distinguishable
APIM are alternative parameterizations of the same fixed associations
[@iidaModelsInterdependentIndividuals2018].

Let's fit the equivalent distinguishable APIM:

```{r fit-cross-apim, eval=has_glmmTMB}
apim_model <- glmmTMB::glmmTMB(
  satisfaction ~
    # Role-specific intercepts
    0 +
    .i_is_female_x_male_female +
    .i_is_female_x_male_male +

    # Role-specific actor effects
    .i_is_female_x_male_female:.i_communication_actor +
    .i_is_female_x_male_male:.i_communication_actor +

    # Role-specific partner effects
    .i_is_female_x_male_female:.i_communication_partner +
    .i_is_female_x_male_male:.i_communication_partner +

    # Role-specific Gaussian residual covariance structure
    us(0 +
         .i_is_female_x_male_female +
         .i_is_female_x_male_male
       | coupleID),
  dispformula = ~ 0,
  family = gaussian(),
  data = cross_dsm_data_inverted
)
```

The two DSM directions and the APIM have identical fit statistics:

```{r compare-cross-fit, eval=has_glmmTMB}
data.frame(
  model = c("DSM: female - male", "DSM: male - female", "APIM"),
  AIC = round(c(AIC(dsm_model), AIC(dsm_model_inverted), AIC(apim_model)), 3),
  BIC = round(c(BIC(dsm_model), BIC(dsm_model_inverted), BIC(apim_model)), 3),
  logLik = round(c(
    as.numeric(logLik(dsm_model)),
    as.numeric(logLik(dsm_model_inverted)),
    as.numeric(logLik(apim_model))
  ), 3)
)
```

### Fixed-effect transformation

Let $b_{\mathrm{actor},\mathrm{female}}$ and
$b_{\mathrm{actor},\mathrm{male}}$ denote the actor effects on the female and
male outcomes. The corresponding partner effects are
$b_{\mathrm{partner},\mathrm{female}}$ and
$b_{\mathrm{partner},\mathrm{male}}$, and the APIM intercepts are
$b_{0,\mathrm{female}}$ and $b_{0,\mathrm{male}}$. Fixed APIM coefficients use
$b$ and write out their effect and outcome role. The numbered paths $a_{10}$
through $a_{22}$ retain the published DSM notation.

The slope transformation can be understood in two steps. First, for each
outcome role $r \in \{\mathrm{female},\mathrm{male}\}$, form the actor-plus-partner
and actor-minus-partner combinations:

$$
\begin{aligned}
b_{\mathrm{sum},r}
&= b_{\mathrm{actor},r} + b_{\mathrm{partner},r}, \\
b_{\mathrm{difference},r}
&= b_{\mathrm{actor},r} - b_{\mathrm{partner},r}.
\end{aligned}
$$

The DSM slopes then follow by combining the female- and male-outcome effects:

$$
\begin{aligned}
a_{11}
&= \frac{b_{\mathrm{sum},\mathrm{female}}
        + b_{\mathrm{sum},\mathrm{male}}}{2},
&
a_{21}
&= b_{\mathrm{sum},\mathrm{female}}
   - b_{\mathrm{sum},\mathrm{male}}, \\
a_{12}
&= \frac{b_{\mathrm{difference},\mathrm{female}}
        - b_{\mathrm{difference},\mathrm{male}}}{4},
&
a_{22}
&= \frac{b_{\mathrm{difference},\mathrm{female}}
        + b_{\mathrm{difference},\mathrm{male}}}{2}.
\end{aligned}
$$

The APIM predictors retain their original scale, whereas the DSM predictor
level is grand-mean centered. Keeping the raw APIM predictors on their original
scale preserves their reference values; the centering difference is handled
explicitly in the intercept transformation. If $\mu_X$ is the grand mean
subtracted from the DSM predictor level, the intercepts transform as

$$
a_{10}
= \frac{b_{0,\mathrm{female}} + b_{0,\mathrm{male}}}{2} + \mu_X a_{11},
\qquad
a_{20} = b_{0,\mathrm{female}} - b_{0,\mathrm{male}} + \mu_X a_{21}.
$$

For the reverse slope transformation, first recover the role-specific
actor-plus-partner and actor-minus-partner combinations:

$$
\begin{aligned}
b_{\mathrm{sum},\mathrm{female}}
&= a_{11} + \frac{a_{21}}{2},
&
b_{\mathrm{sum},\mathrm{male}}
&= a_{11} - \frac{a_{21}}{2}, \\
b_{\mathrm{difference},\mathrm{female}}
&= a_{22} + 2a_{12},
&
b_{\mathrm{difference},\mathrm{male}}
&= a_{22} - 2a_{12}.
\end{aligned}
$$

Then, for each outcome role $r$,

$$
b_{\mathrm{actor},r}
= \frac{b_{\mathrm{sum},r} + b_{\mathrm{difference},r}}{2},
\qquad
b_{\mathrm{partner},r}
= \frac{b_{\mathrm{sum},r} - b_{\mathrm{difference},r}}{2}.
$$

The intercepts transform back as

$$
b_{0,\mathrm{female}} = a_{10} + \frac{a_{20}}{2}
- \mu_X\left(a_{11} + \frac{a_{21}}{2}\right),
$$

$$
b_{0,\mathrm{male}} = a_{10} - \frac{a_{20}}{2}
- \mu_X\left(a_{11} - \frac{a_{21}}{2}\right).
$$

The following comparison applies the APIM-to-DSM transformation to all six
fixed effects:

```{r compare-cross-coefficients, eval=has_glmmTMB, echo=FALSE}
apim_coef <- glmmTMB::fixef(apim_model)$cond
dsm_coef <- glmmTMB::fixef(dsm_model)$cond

b0_female <- apim_coef[[".i_is_female_x_male_female"]]
b0_male <- apim_coef[[".i_is_female_x_male_male"]]
b_actor_female <- apim_coef[[
  ".i_is_female_x_male_female:.i_communication_actor"
]]
b_actor_male <- apim_coef[[
  ".i_is_female_x_male_male:.i_communication_actor"
]]
b_partner_female <- apim_coef[[
  ".i_is_female_x_male_female:.i_communication_partner"
]]
b_partner_male <- apim_coef[[
  ".i_is_female_x_male_male:.i_communication_partner"
]]

mu_X <- mean(
  (cross_dsm_data_inverted$.i_communication_actor +
     cross_dsm_data_inverted$.i_communication_partner) / 2 -
    cross_dsm_data_inverted$.i_communication_dyad_mean_gmc,
  na.rm = TRUE
)

b_sum_female <- b_actor_female + b_partner_female
b_sum_male <- b_actor_male + b_partner_male
b_difference_female <- b_actor_female - b_partner_female
b_difference_male <- b_actor_male - b_partner_male

a11_from_apim <- (b_sum_female + b_sum_male) / 2
a21_from_apim <- b_sum_female - b_sum_male
a12_from_apim <- (b_difference_female - b_difference_male) / 4
a22_from_apim <- (b_difference_female + b_difference_male) / 2

apim_to_dsm <- c(
  a10 = (b0_female + b0_male) / 2 + mu_X * a11_from_apim,
  a11 = a11_from_apim,
  a12 = a12_from_apim,
  a20 = b0_female - b0_male + mu_X * a21_from_apim,
  a21 = a21_from_apim,
  a22 = a22_from_apim
)

a10_dsm <- dsm_coef[["(Intercept)"]]
a11_dsm <- dsm_coef[[".i_communication_dyad_mean_gmc"]]
a12_dsm <- dsm_coef[[".i_communication_within_dyad_diff"]]
a20_dsm <- dsm_coef[[".i_dsm_role_contrast"]]
a21_dsm <- dsm_coef[[
  ".i_communication_dyad_mean_gmc:.i_dsm_role_contrast"
]]
a22_dsm <- dsm_coef[[
  ".i_communication_within_dyad_diff:.i_dsm_role_contrast"
]]

knitr::kable(
  data.frame(
    `DSM path` = names(apim_to_dsm),
    `From APIM transformation` = round(unname(apim_to_dsm), 3),
    `From DSM model` = round(c(
      a10_dsm,
      a11_dsm,
      a12_dsm,
      a20_dsm,
      a21_dsm,
      a22_dsm
    ), 3),
    check.names = FALSE
  ),
  align = c("l", "r", "r"),
  caption = paste0(
    "APIM-to-DSM fixed-effect transformation (centering constant = ",
    round(mu_X, 3),
    ")."
  )
)
```

### Random-effect transformation

Let $u_{\mathrm{female}}$ and $u_{\mathrm{male}}$ be the APIM random effects
for the female and male outcomes. The DSM outcome-level and outcome-difference
residuals are the same random effects expressed in different coordinates:

$$
\begin{pmatrix} r_{Y,\mathrm{mean}} \\ r_{Y,\mathrm{diff}} \end{pmatrix}
=
\begin{pmatrix} 1/2 & 1/2 \\ 1 & -1 \end{pmatrix}
\begin{pmatrix} u_{\mathrm{female}} \\ u_{\mathrm{male}} \end{pmatrix}.
$$

Applying this rotation to the APIM covariance matrix reproduces the DSM random
intercept variance, intercept-slope covariance, and role-slope variance:

```{r compare-cross-random-effects, eval=has_glmmTMB}
apim_vcov <- as.matrix(glmmTMB::VarCorr(apim_model)$cond$coupleID)
dsm_vcov <- as.matrix(glmmTMB::VarCorr(dsm_model)$cond$coupleID)

rotation <- rbind(
  outcome_level = c(0.5, 0.5),
  outcome_difference = c(1, -1)
)
apim_to_dsm_vcov <- rotation %*% apim_vcov %*% t(rotation)

data.frame(
  parameter = c(
    "Var(outcome mean)",
    "Cov(outcome mean, outcome diff)",
    "Var(outcome diff)"
  ),
  from_DSM = round(c(
    dsm_vcov[1, 1],
    dsm_vcov[1, 2],
    dsm_vcov[2, 2]
  ), 3),
  from_APIM_transformation = round(c(
    apim_to_dsm_vcov[1, 1],
    apim_to_dsm_vcov[1, 2],
    apim_to_dsm_vcov[2, 2]
  ), 3)
)
```

For exchangeable dyads, the direction of a member difference is arbitrary.
The directional intercept, both cross-paths, and the covariance between outcome
level and outcome difference must then be zero. The remaining reduced,
label-invariant DSM is algebraically the Gaussian Dyad-Individual Model (DIM).
In `interdep`, use `model_type = "dim"` for this exchangeable model and reserve
`model_type = "dsm"` for distinguishable dyads.

Because the Gaussian DIM is the exchangeability-constrained version of the full
DSM, exchangeability can also be tested by comparing these nested models. This
is equivalent to the comparison shown in [Testing distinguishability in the APIM vignette](apim.html#testing-distinguishability).

## Intensive longitudinal DSM

> The following sections are under construction and are coming soon! This extension
is equivalent to the extension from the cross-sectional DIM to the ILD DIM. Please
refer to that section.

### Interpretation of concurrent ILD DSM coefficients

### Equivalence of APIM and DSM in ILD

### Including Random Slopes

#### Transforming DSM random slopes to APIM slopes

### Dynamic ILD DSM example

The ILD models above do not model residual serial dependence. One way to model
dynamics or to account for temporal dependency is to include lagged outcomes as predictors.

**Note:** Dynamic models, especially with small time series, are subject to bias.
This, and the choice between raw and within-person-centered outcome lags, are
addressed in the [APIM vignette's discussion of dynamic models](apim.html#dynamic-models).

---

Return to the [Actor-Partner Interdependence Model vignette](apim.html), see
the [Mixed-Composition APIM vignette](mixed-apim.html) or the
[Dyad-Individual Model vignette](dim.html) for related model specifications,
or return to the [Overview](index.html).

## References

::: {#refs}
:::
