---
title: "Dyad-Individual Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dyad-Individual Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

This vignette focuses on Dyad-Individual Model (DIM) predictor construction and
its relationship to APIM parameterizations. For broader guidance on fitting
dyadic models with `interdep`-prepared data, especially APIMs with multiple dyad
types, generalized outcomes, intensive longitudinal models, and optimizer
choices, see `vignette("getting-started")`.

## Cross-Sectional Gaussian DIM

The current DIM implementation prepares undirected DIM predictors. This means
that the dyad members are treated as exchangeable for DIM construction. Here we
therefore omit `role`, which makes `prepare_interdep_data()` create a valid
`.i_diff` contrast with values `-1` and `1` within each dyad.

A supplied `role` column is also allowed when it yields exactly one
exchangeable composition, such as all female-female dyads. DIM construction is
currently rejected for distinguishable compositions and for multiple
exchangeable compositions.

```{r prepare-cross-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  predictors = communication,
  # returning both apim and dim columns for comparison
  model_type = c("apim", "dim"),
  seed = 123
)

print(cross_exchangeable_data, n = 20)
```

### Example DIM Model

The random-effects structure contains a dyad-level random intercept and a
within-dyad random deviation indexed by `.i_diff`.

```{r fit-cross-dim, eval=has_glmmTMB}

dim_1 <- glmmTMB::glmmTMB(
  satisfaction ~ 1 +
    .i_communication_raw_dyad_mean_gmc + .i_communication_raw_within_dyad_deviation +
    (1 | coupleID) + (0 + .i_diff | coupleID),
  data = cross_exchangeable_data
)

summary(dim_1)
```

The same model can be written in APIM form:

```{r fit-cross-apim, eval=has_glmmTMB}

apim_1 <- glmmTMB::glmmTMB(
  satisfaction ~ 1 +
    .i_communication_raw_actor + .i_communication_raw_partner +
    (1 | coupleID) + (0 + .i_diff | coupleID),
  data = cross_exchangeable_data
)

summary(apim_1)
```

```{r glmmTMB-missing, eval=!has_glmmTMB}
message("Install glmmTMB to run the fitted-model examples in this vignette.")
```

The two models have identical fit statistics:

```{r compare-cross-fit, eval=has_glmmTMB}
data.frame(
  model = c("DIM", "APIM"),
  AIC = c(AIC(dim_1), AIC(apim_1)),
  BIC = c(BIC(dim_1), BIC(apim_1)),
  logLik = c(as.numeric(logLik(dim_1)), as.numeric(logLik(apim_1)))
)
```

This shows that the same statistical model is being estimated with different
parameterizations and coefficient interpretations.

The coefficients relate as follows:

$$ 
b_{dyad\ mean} = b_{actor} + b_{partner}
$$

and 

$$ 
b_{within\ dyad\ deviation} = b_{actor} - b_{partner}
$$

For example:

```{r compare-cross-coefficients, eval=has_glmmTMB}
apim_coef <- glmmTMB::fixef(apim_1)$cond
dim_coef <- glmmTMB::fixef(dim_1)$cond

b_actor <- round(apim_coef[[".i_communication_raw_actor"]], 3)
b_partner <- round(apim_coef[[".i_communication_raw_partner"]], 3)

b_dyad_mean <- round(dim_coef[[".i_communication_raw_dyad_mean_gmc"]], 3)
b_within_dyad <- round(dim_coef[[".i_communication_raw_within_dyad_deviation"]], 3)

cat("From APIM model:\n")
cat("  actor effect:                  ",  b_actor, "\n")
cat("  partner effect:                ", b_partner, "\n\n")

cat("  Computes to: \n")
cat("  actor effect + partner effect: ", b_actor + b_partner, "\n")
cat("  actor effect - partner effect: ", b_actor - b_partner, "\n\n")

cat("From DIM model:\n")
cat("  dyad-mean effect:              ", b_dyad_mean, "\n")
cat("  within-dyad-deviation effect:  ", b_within_dyad, "\n")
```

## Intensive longitudinal DIM


```{r prepare-ild-exchangeable}
ild_exchangeable_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  # role = gender,
  member = personID,
  time = diaryday,
  model_type = c("apim", "dim"),
  predictors = provided_support,
  seed = 123
)

print(ild_exchangeable_data)
```

```{r fit-ild-dim, eval = has_glmmTMB}

dim_ILD <- glmmTMB::glmmTMB(
  closeness ~ 
    1 + 
    
    diaryday +

    # Within-person DIM
    .i_provided_support_cwp_dyad_mean +
    .i_provided_support_cwp_within_dyad_deviation +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_deviation +

    # random effects for stable non-independence (means)
    us(1 | coupleID)  + us(0 + .i_diff | coupleID) +

    # Same-day residual covariance
    us(1 | coupleID:diaryday) + us(0 + .i_diff | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

summary(dim_ILD)

```

Compare to the APIM on both levels: 

```{r fit-ild-apim, eval = has_glmmTMB}

apim_ILD <- glmmTMB::glmmTMB(
  closeness ~ 
    1 + 
    
    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person APIM
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # random effects for stable non-independence (means)
    us(1 | coupleID)  + us(0 + .i_diff | coupleID) +

    # Same-day residual covariance
    us(1 | coupleID:diaryday) + us(0 + .i_diff | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

summary(apim_ILD)

```

The two ILD models again have identical fit statistics:

```{r compare-ild-fit, eval = has_glmmTMB}
data.frame(
  model = c("DIM", "APIM"),
  AIC = c(AIC(dim_ILD), AIC(apim_ILD)),
  BIC = c(BIC(dim_ILD), BIC(apim_ILD)),
  logLik = c(as.numeric(logLik(dim_ILD)), as.numeric(logLik(apim_ILD)))
)
```

The equivalence holds separately for the within-person and between-person
predictor components. For example:

`b_.i_provided_support_cwp_actor + b_.i_provided_support_cwp_partner`
corresponds to `b_.i_provided_support_cwp_dyad_mean`.

Analogous sum and difference transformations apply to the other within-person
and between-person components.

This also means that an APIM parameterization can be used on one level and a
DIM parameterization on the other.

```{r fit-ild-apim-within-dim-between, eval = has_glmmTMB}

apim_dim_ILD <- glmmTMB::glmmTMB(
  closeness ~ 
    1 + 
    
    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_deviation +

    # random effects for stable non-independence (means)
    us(1 | coupleID)  + us(0 + .i_diff | coupleID) +

    # Same-day residual covariance
    us(1 | coupleID:diaryday) + us(0 + .i_diff | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

summary(apim_dim_ILD)

```

Which is still the same model:

```{r compare-ild-all-fit, eval = has_glmmTMB}
data.frame(
  model = c("DIM", "APIM", "APIM within / DIM between"),
  AIC = c(AIC(dim_ILD), AIC(apim_ILD), AIC(apim_dim_ILD)),
  BIC = c(BIC(dim_ILD), BIC(apim_ILD), BIC(apim_dim_ILD)),
  logLik = c(
    as.numeric(logLik(dim_ILD)),
    as.numeric(logLik(apim_ILD)),
    as.numeric(logLik(apim_dim_ILD))
  )
)
```

## Advanced: Role-Moderated DIM Fixed Effects

A possible extension is to let a distinguishable variable, such as gender,
moderate the DIM fixed effects. Conceptually, this creates role-specific
dyad-mean and within-dyad-deviation effects. Those role-specific DIM effects
can then be translated back to actor and partner effects using the same sum and
difference logic shown above.

This is close to the fixed-effect interpretation of a distinguishable APIM, but
it is not yet full distinguishable DIM support. In particular, the random-effect
and residual structures would also need role-specific treatment before this
could be treated as a complete distinguishable dyadic model.

## Advanced: Random slopes

To include equivalent random slopes between the two models, the within-person 
model coefficients can be entered as slopes as such: 

WARNING: careful about model convergence, the following models do not
converge properly and should not be interpreted. They only show the conceptual 
way of incorporating random slopes. 

In the APIM:

```{r fit-ild-apim-random, eval = FALSE}

apim_ILD_random <- glmmTMB::glmmTMB(
  closeness ~ 
    1 + 
    
    diaryday +

    # Within-person APIM
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Between-person APIM
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # random effects for stable non-independence (means)
    us(1 + .i_provided_support_cwp_actor + .i_provided_support_cwp_partner | coupleID)  + 
    us(0 + .i_diff + .i_diff:.i_provided_support_cwp_actor + .i_diff:.i_provided_support_cwp_partner | coupleID) +

    # Same-day residual covariance
    us(1 | coupleID:diaryday) + us(0 + .i_diff | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```

In the DIM: 

```{r fit-ild-dim-random, eval = FALSE}

dim_ILD_random <- glmmTMB::glmmTMB(
  closeness ~ 
    1 + 
    
    diaryday +

    # Within-person DIM
    .i_provided_support_cwp_dyad_mean +
    .i_provided_support_cwp_within_dyad_deviation +

    # Between-person DIM
    .i_provided_support_cbp_dyad_mean +
    .i_provided_support_cbp_within_dyad_deviation +

    # random effects for stable non-independence (means)
    us(1 + .i_provided_support_cwp_dyad_mean + .i_provided_support_cwp_within_dyad_deviation | coupleID)  + 
    us(0 + .i_diff + .i_diff:.i_provided_support_cwp_dyad_mean + .i_diff:.i_provided_support_cwp_within_dyad_deviation | coupleID) +

    # Same-day residual covariance
    us(1 | coupleID:diaryday) + us(0 + .i_diff | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

```
