---
title: "Dyad-Individual-Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dyad-Individual-Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```


## Cross-Sectional Gaussian DIM

We prepare the data with the `interdep::prepare_interdep_data()` function:

```{r prepare-cross-distinguishable}
cross_distinguishable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  # role = gender, 
  # currently, role cannot be supplied, as we need
  # valid .i_diff columns that are -1 and 1 randomly per couple
  predictors = communication,
  
  # we request both apim and dim variables so we can compare the models
  model_type = "apim_dim",
  seed = 123
)

print(cross_distinguishable_data, n = 20)

```

### Example DIM Model

```{r}

dim_1 <- glmmTMB::glmmTMB(
  satisfaction ~ 1 + 
    .i_communication_raw_dyad_mean + .i_communication_raw_within_dyad_deviation +
    
    (1 | coupleID) + (0 + .i_diff | coupleID)
  
  , data = cross_distinguishable_data
)

summary(dim_1)
```

This model is identical to an APIM as such: 

```{r}

apim_1 <- glmmTMB::glmmTMB(
  satisfaction ~ 1 + 
    .i_communication_raw_actor + .i_communication_raw_partner +
    
    (1 | coupleID) + (0 + .i_diff | coupleID)
  
  , data = cross_distinguishable_data
)

summary(apim_1)
```


Note the identical AIC, BIC, logLIK, etc. (p-value is not interpretable here): 

```{r}
anova(dim_1, apim_1)
```

This is evidence, that we are estimating the exact same model twice, just with different
parametrizations and interpretations of the coefficients!

The coefficients relate as such: 

$$ 
b_{BetweenCouple} = b_{actor} + b_{partner}
$$
and 

$$ 
b_{WithinCoupleDifference} = b_{actor} - b_{partner}
$$

E.g.,:

```{r}
b_actor <- round(apim_1$fit$par[[2]], 3)
b_partner <- round(apim_1$fit$par[[3]], 3)

b_between_couple <- round(dim_1$fit$par[[2]], 3)
b_within_couple <- round(dim_1$fit$par[[3]], 3)

# with: 
cat("From APIM model: \n")
cat(" actor effect: ", b_actor, "\n")
cat(" partner effect: ", b_partner, "\n\n")

cat(" actor effect + partner effect: ", b_actor + b_partner, "\n")
cat(" actor effect - partner effect: ", b_actor - b_partner, "\n\n")


# and: 
cat("From DIM model: \n")
cat(" between-couple effect:", b_between_couple, "\n")
cat(" within-couple effect:", b_within_couple, "\n\n")

```

