---
title: "Dyad-Individual Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dyad-Individual Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

## Cross-Sectional Gaussian DIM

The current DIM implementation prepares non-directed DIM predictors. This means
that the dyad members are treated as exchangeable for DIM construction. Here we
therefore omit `role`, which makes `prepare_interdep_data()` create a valid
`.i_diff` contrast with values `-1` and `1` within each dyad.

```{r prepare-cross-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  predictors = communication,
  # returning both apim and dim columns for comparison
  model_type = "apim_dim",
  seed = 123
)

print(cross_exchangeable_data, n = 20)
```

### Example DIM Model

The random-effects structure contains a dyad-level random intercept and a
within-dyad random deviation indexed by `.i_diff`.

```{r fit-cross-dim, eval=has_glmmTMB}

dim_1 <- glmmTMB::glmmTMB(
  satisfaction ~ 1 +
    .i_communication_raw_dyad_mean + .i_communication_raw_within_dyad_deviation +
    (1 | coupleID) + (0 + .i_diff | coupleID),
  data = cross_exchangeable_data
)

summary(dim_1)
```

The same model can be written in APIM form:

```{r fit-cross-apim, eval=has_glmmTMB}

apim_1 <- glmmTMB::glmmTMB(
  satisfaction ~ 1 +
    .i_communication_raw_actor + .i_communication_raw_partner +
    (1 | coupleID) + (0 + .i_diff | coupleID),
  data = cross_exchangeable_data
)

summary(apim_1)
```

```{r glmmTMB-missing, eval=!has_glmmTMB}
message("Install glmmTMB to run the fitted-model examples in this vignette.")
```

The two models have identical fit statistics:

```{r compare-cross-fit, eval=has_glmmTMB}
data.frame(
  model = c("DIM", "APIM"),
  AIC = c(AIC(dim_1), AIC(apim_1)),
  BIC = c(BIC(dim_1), BIC(apim_1)),
  logLik = c(as.numeric(logLik(dim_1)), as.numeric(logLik(apim_1)))
)
```

This shows that the same statistical model is being estimated with different
parameterizations and coefficient interpretations.

The coefficients relate as follows:

$$ 
b_{dyad\ mean} = b_{actor} + b_{partner}
$$
and 

$$ 
b_{within\ dyad\ deviation} = b_{actor} - b_{partner}
$$

For example:

```{r compare-cross-coefficients, eval=has_glmmTMB}
apim_coef <- glmmTMB::fixef(apim_1)$cond
dim_coef <- glmmTMB::fixef(dim_1)$cond

b_actor <- round(apim_coef[[".i_communication_raw_actor"]], 3)
b_partner <- round(apim_coef[[".i_communication_raw_partner"]], 3)

b_dyad_mean <- round(dim_coef[[".i_communication_raw_dyad_mean"]], 3)
b_within_dyad <- round(dim_coef[[".i_communication_raw_within_dyad_deviation"]], 3)

cat("From APIM model:\n")
cat("  actor effect:                  ",  b_actor, "\n")
cat("  partner effect:                ", b_partner, "\n\n")

cat("  Computes to: \n")
cat("  actor effect + partner effect: ", b_actor + b_partner, "\n")
cat("  actor effect - partner effect: ", b_actor - b_partner, "\n\n")

cat("From DIM model:\n")
cat("  dyad-mean effect:              ", b_dyad_mean, "\n")
cat("  within-dyad-deviation effect:  ", b_within_dyad, "\n")
```
