---
title: "Generalized Non-Gaussian Actor-Partner Interdependence Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalized Non-Gaussian Actor-Partner Interdependence Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```


## Dichotomous outcomes

## Ordinal outcomes

## Categorical outcomes

## Poisson and Negative Binomial Outcomes

##

## Example and excursion: semi-continuous cross-sectional data

!NOTE: potentially we should separate non-normal models into their own vignette, since they are actually
quite complex and models depend on 

`example_dyadic_crosssectional_tweedie` has the same dyadic structure as before, but the
outcome has exact zeros and positive skewed values, as in
some physical activity outcomes, in contrast to the Gaussian distribution from the 
previous example. 

```{r tweedie-raw}
print(head(example_dyadic_crosssectional_tweedie))
hist(example_dyadic_crosssectional_tweedie$physical_activity, breaks = 20)
```

Validation and modeling work the same way because the dyadic structure is the same.
The random-effect interpretation is different from the Gaussian case: Tweedie
random effects are latent effects on the log-mean scale, while the
observation-level Tweedie variance remains part of the model.

### Distinguishable Tweedie APIM

```{r prepare-tweedie-distinguishable}
tweedie_distinguishable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_tweedie,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = motivation,
  # when no model_type is specified, apim is the default.
  seed = 123
)

print(tweedie_distinguishable_data)
summary(tweedie_distinguishable_data)
```


```{r fit-tweedie-distinguishable, eval = FALSE}

tweedie_distinguishable_model <- glmmTMB(
  physical_activity ~ 
    # remove standard intercept and model separate intercepts for male and female
    0 + .i_is_female_x_male_female + .i_is_female_x_male_male + 
    
    # gender-specific slopes for motivation actor effect
    .i_is_female_x_male_female:.i_motivation_actor + .i_is_female_x_male_male:.i_motivation_actor +
    
    # gender-specific slopes for motivation partner effect
    .i_is_female_x_male_female:.i_motivation_partner + .i_is_female_x_male_male:.i_motivation_partner +
    
    # keep a simple couple-level latent effect for stable non-independence
    # important limitation: this can only induce positive partner dependence
    (1 | coupleID) 
  
  # allow role-specific Tweedie dispersion
  , dispformula = ~ 0 + .i_is_female_x_male_female + .i_is_female_x_male_male
  , family = tweedie()
  , data = tweedie_distinguishable_data
)

summary(tweedie_distinguishable_model)

```

### Exchangeable Tweedie APIM

```{r prepare-tweedie-exchangeable}
tweedie_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_tweedie,
  group = coupleID,
  member = personID,
  # role = gender,
  predictors = motivation,
  seed = 123
)

print(tweedie_exchangeable_data)
summary(tweedie_exchangeable_data)
```

```{r fit-tweedie-exchangeable, eval = FALSE}

tweedie_exchangeable_model <- glmmTMB(
  physical_activity ~ 
    # pooled intercept 
    1 + 
    
    # pooled actor slope for motivation
    .i_motivation_actor +
    
    # pooled partner slope for motivation
    .i_motivation_partner +
    
    # exchangeable latent dyad block on the log-mean scale
    (1 | coupleID) + (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID)
    
  # estimate a single pooled Tweedie dispersion parameter
  , dispformula = ~ 1
  , family = tweedie()
  , data = tweedie_exchangeable_data
)

summary(tweedie_exchangeable_model)

```
