---
title: "Getting Started"
author: "Pascal Küng"
description: >
  Core workflow for validating dyadic data, identifying dyad compositions, and
  preparing model-ready columns for dyadic multilevel models.
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```

## Installation

You can install the development version of `interdep` from GitHub with:

```{r installation, eval=FALSE}
# install.packages("pak")
pak::pak("Pascal-Kueng/interdep")
```

## About this Vignette

`interdep` helps researchers prepare cross-sectional and intensive longitudinal
dyadic data for (generalized) multilevel models. It automatically creates
model-ready columns for dyadic multilevel model parameterizations such as
Actor-Partner Interdependence Models (APIM), Dyad-Individual Models (DIM), and
Dyadic Score Models (DSM). DIM currently supports one exchangeable dyad
composition, whereas DSM supports one distinguishable dyad composition.

**This vignette focuses on the automatic data preparation step.**

For guidance and examples on how to use the prepared data to estimate
cross-sectional, generalized, and intensive longitudinal APIMs, see the
[Actor-Partner Interdependence Model vignette](apim.html).
For models that combine distinguishable and exchangeable dyad compositions,
see the [Mixed-Composition APIM vignette](mixed-apim.html).

For guidance on how to use the Dyad-Individual Model (DIM) parameterization,
including dyad-mean and within-dyad-deviation predictors and their equivalence
to APIM effects (via an interactive example) in exchangeable dyads, see the
[Dyad-Individual Model vignette](dim.html).

For the dyadic score parameterization and its relationship to the APIM for
distinguishable dyads, see the [Dyadic Score Model vignette](dsm.html).

For an in-depth tutorial covering data preparation and model fitting, but also
additional steps like diagnostics and assumption checks, see
[Distinguishable and Exchangeable Dyads: Bayesian Multilevel Modelling](https://pascal-kueng.github.io/05DyadicDataAnalysis/).

## Prerequisites

The basic data structure needed for `interdep` is a long data frame where dyads
are stacked on top of each other and both members of a dyad appear as separate
rows.

If your raw data are currently in wide format (for time or dyads or both),
reshape them to this long structure before using `prepare_interdep_data()`. See the
[tidyr pivoting vignette](https://tidyr.tidyverse.org/articles/pivot.html) or
the [`pivot_longer()` reference](https://tidyr.tidyverse.org/reference/pivot_longer.html).

Roughly, the expected structure for `interdep` is:

- For cross-sectional data: one row per `dyad x member`

| dyad | member | x | y |
|---:|---:|---:|---:|
| 1 | 1 | 4.2 | 7.1 |
| 1 | 2 | 5.0 | 6.4 |
| 2 | 1 | 3.8 | 5.9 |
| 2 | 2 | 4.5 | 6.8 |

- For intensive longitudinal data: at most one row per `dyad x time x member`

| dyad | time | member | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 2 | 5.0 | 6.4 |
| 1 | 2 | 1 | 4.0 | 6.9 |
| 1 | 2 | 2 | 5.3 | 6.6 |


Measured variables may contain missing values. The structural `group`,
`member`, and optional `time` variables must not contain missing values.

In intensive longitudinal data, missing measurement occasions can be represented
by absent rows, as long as the time variable preserves the observed measurement
occasions. For example:

| dyad | personID | time | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 3 | 4.0 | 6.9 |
| 1 | 2 | 1 | 5.3 | 6.6 |
| 1 | 2 | 2 | 4.7 | 6.1 |
| 1 | 2 | 3 | 5.1 | 6.4 |

In this example, the row for person 1 at time 2 is absent. The time variable
preserves the observed measurement occasions and skips from time 1 to time 3
for that person. `interdep` accepts this structure without requiring a
placeholder row for the missing occasion.


## Data preparation for distinguishable dyads

`example_dyadic_crosssectional` is a simulated cross-sectional dataset for
distinguishable dyads. Each dyad has two rows: one for each member.

```{r cross-sectional-raw}
print(head(example_dyadic_crosssectional))
```

We validate and prepare the data with the function `prepare_interdep_data()`

```{r prepare-cross-distinguishable}
cross_distinguishable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,

  # In this example, we optionally specify a predictor variable
  # and a model type to generate the columns needed for that model type.
  predictors = communication,
  model_type = "apim"
)

print(cross_distinguishable_data, n = 4)

```

The function automatically recognized that in this dataset there are 95
female-male dyads and created APIM-relevant variables. These generated `.i_*`
columns can be used directly in model formulas.

Here is a simple example:

```{r fit-cross-distinguishable-gaussian, eval = requireNamespace("glmmTMB", quietly = TRUE), results = "hide"}
library(glmmTMB)

cross_distinguishable_model <- glmmTMB(
  satisfaction ~

    # Gender-specific intercepts
    0 +
    .i_is_female_x_male_female +
    .i_is_female_x_male_male +

    # Gender-specific actor effects
    .i_is_female_x_male_female:.i_communication_actor +
    .i_is_female_x_male_male:.i_communication_actor +

    # Gender-specific partner effects
    .i_is_female_x_male_female:.i_communication_partner +
    .i_is_female_x_male_male:.i_communication_partner +

    # Dyad-level unstructured random effects represent the two partner
    # residual variances and their covariance when dispformula = ~ 0.
    # This is glmmTMB-specific syntax! lme4 and brms use different syntax.
    us(0 +
         .i_is_female_x_male_female +
         .i_is_female_x_male_male
       | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_distinguishable_data
)
```

The model is fitted when `glmmTMB` is installed, but its output is omitted here
to keep the focus on data preparation. For fitted APIM examples, see the
[Actor-Partner Interdependence Model vignette](apim.html).

## Data preparation for exchangeable dyads

For a dataset with only one type of dyad that should be treated as exchangeable,
omit the `role` argument:

```{r prepare-cross-exchangeable-basic}
cross_exchangeable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  seed = 123
)

print(cross_exchangeable_data, n = 4)

```

The generated `.i_diff_assumed_exchangeable_arbitrary` contrast assigns `-1`
and `1` to the two members of each exchangeable dyad. Its direction is
arbitrary, and `seed` makes the assignment reproducible. When role compositions
are available, each exchangeable composition receives its own contrast, such as
`.i_diff_female_x_female_arbitrary`, which is `0` for all other compositions.
We use a fixed seed in the examples below for consistent results.

Alternatively, for more control, we can explicitly set dyad types to exchangeable:

```{r prepare-cross-set-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = "male-female",
  seed = 123
)

print(cross_exchangeable_data, n = 4)

```

*Note* that whenever you need to refer to a dyad type, the order of members does not matter
(e.g., `male-female` and `female-male` will both work), and you can use different separators
like `male_female`, `male_x_female`, or `male female`.

For exchangeable dyads, we can request DIM predictor columns. This works here
because omitting `role` treats all dyads as a single exchangeable composition.

```{r prepare-cross-dim}
cross_dim_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  predictors = communication,
  model_type = "dim",
  seed = 123
)

print(cross_dim_data, n = 4)
```

For distinguishable dyads, DSM preparation additionally requires a stable role
variable and an explicit role order. The role order defines the direction of
all DSM predictor differences and the DSM role contrast. Outcomes remain
unchanged and are selected later in the fitted-model formula.

```{r prepare-cross-dsm}
cross_dsm_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  predictors = communication,
  model_type = "dsm",
  dsm_role_order = c("female", "male")
)

print(cross_dsm_data, n = 4)
```

## Incomplete dyads and missing roles

By default, `prepare_interdep_data()` stops when a dyad has only one observed
member or when a member's role cannot be resolved from the observed rows. These
cases can also be dropped before validation continues.

```{r incomplete-drop}
incomplete_data <- data.frame(
  coupleID = c(1, 2, 2, 3, 3, 4, 4, 5),
  personID = c(1, 3, 4, 5, 6, 7, 8, 10),
  gender = c("female", "female", NA, "female", "female", "female", "male", NA),
  satisfaction = c(5.2, 4.8, 4.9, 5.1, 5.0, 4.7, 4.6, 3.0)
)

incomplete_dropped_data <- prepare_interdep_data(
  incomplete_data,
  group = coupleID,
  member = personID,
  role = gender,
  incomplete_dyads = "drop",
  missing_role = "drop",
  seed = 123
)

print(incomplete_dropped_data)
```


## Intensive longitudinal dyadic data

`example_dyadic_ILD` is a simulated intensive longitudinal dyadic dataset. Each
dyad has repeated observations over `diaryday`, with one row per person-day.

The first rows look like this:

```{r ild-raw}
print(example_dyadic_ILD, n = 26)
```

To prepare intensive longitudinal data, pass the `time` variable to
`prepare_interdep_data()`.

### ILD APIM preparation

```{r prepare-ild-apim}
ild_apim_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  model_type = "apim",
  seed = 123
)

print(ild_apim_data)

```

By default, numeric predictors in longitudinal APIM preparation are
decomposed into within-person and between-person components. This temporal
predictor decomposition is controlled by `temporal_predictor_decomposition`.
The default `"auto"` setting selects `"time_2l"` for this longitudinal setup.

## Data with multiple and mixed-composition dyads

`example_dyadic_crosssectional_mixed` contains three dyad compositions in the
same data object: distinguishable female-male dyads and exchangeable
female-female and male-male dyads.

Let's have `interdep` infer the compositions automatically:

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data, n = 4)
```

We can use this data to model these dyad types as separate or in the same model.
The [APIMs with Mixed Dyad Compositions vignette](mixed-apim.html) shows both
mixed-composition formulas and practical convergence notes.

### Keeping only selected dyad compositions (filtering)

Sometimes a mixed dataset contains dyad compositions that should not be part of
a given analysis. Use `include_compositions` to keep only dyads whose *observed*
composition matches the requested labels. The filtering happens before exchangeability constraints and pooling, so `set_exchangeable_compositions` and
`pool_compositions` arguments can only refer to retained types of dyads.

```{r prepare-mixed-cross-sectional-included}
mixed_cross_data_included <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  include_compositions = c("female-female", "male-male"),
  seed = 123
)

print(mixed_cross_data_included, n = 4)
```

### Setting distinguishable dyads to be treated as exchangeable

As mentioned earlier, omitting the `role` argument treated all
dyads as one exchangeable composition, effectively pooling them.

For more control, a distinguishable dyad composition in a mixed dataset can be
treated as exchangeable. This specification
keeps the differentiation between the kinds of dyads
(e.g., `male-male`, `female-female`, and `male-female`).


```{r prepare-mixed-cross-sectional-exchangeable}
mixed_cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = c("male-female"),
  seed = 123
)

print(mixed_cross_exchangeable_data, n = 4)
```


### Pooling different dyad compositions

Sometimes for theoretical or practical reasons, we may want to
pool different exchangeable dyad compositions and analyze them as if they were one.
This also allows testing various constraints via model comparisons.

For instance, let's pool `male-male` and `female-female` dyads
and name them `same-sex` dyads:

```{r prepare-mixed-cross-sectional_pooled}
mixed_cross_data_pooled <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  pool_compositions = list(
    "same-sex" = c("male-male", "female_female")
  ),
  seed = 123
)

print(mixed_cross_data_pooled)
```

Note that you cannot pool distinguishable dyads. To pool `female-male` with
`male-male`, we first have to treat
`female-male` as exchangeable:


```{r prepare-mixed-cross-sectional_pooled_constrained}
mixed_cross_data_pooled_constrained <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = "male female",
  pool_compositions = list(
    "pooled_exchangeable" = c("male-male", "male_female")
  ),
  seed = 123
)

print(mixed_cross_data_pooled_constrained)
```

---

**Continue** with the 
[Actor-Partner Interdependence Model (APIM) vignette](apim.html),

refer to the: 

- [Mixed-Composition APIM vignette](mixed-apim.html),
- [Dyad-Individual Model vignette](dim.html),
- [Dyadic Score Model vignette](dsm.html),

or return to the [Overview](index.html).
