---
title: "Getting Started"
description: >
  Core workflow for validating dyadic data, identifying dyad compositions, and
  preparing model-ready columns for dyadic multilevel models.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```

## About this Vignette

`interdep` helps researchers prepare cross-sectional and intensive longitudinal 
dyadic data for (generalized) multilevel models. It supports common dyadic 
designs with one type of exchangeable or distinguishable dyad, and it also 
handles studies where different kinds of dyads appear in the same dataset, 
such as female-male, female-female, and male-male couples. 

It automatically creates model-ready columns for dyadic multilevel model 
parameterizations such as Actor-Partner Interdependence Models (APIM), 
Dyad-Individual Models (DIM), and undirected Dyadic Score Models (DSM). 
Current DIM and undirected DSM helpers require a single type of exchangeable 
dyad composition.

This vignette focuses on the automatic data preparation step.

For guidance and examples on how to use the prepared data to estimate various 
APIM models from simple Gaussian cross-sectional to generalized intensive 
longitudinal (ILD) models with multiple dyad types, see the 
[Actor-Partner Interdependence Model vignette](apim.html). 

For guidance on how to use the Dyad-Individual Model (DIM) parameterization, 
including dyad-mean and within-dyad-deviation predictors and their equivalence 
to APIM effects in exchangeable dyads, see the [Dyad-Individual Model vignette](dim.html). 

For undirected dyadic score outcomes, see the [Undirected Dyadic Score Model vignette](undirected-dsm.html).

## Prerequisites

The basic data structure needed for `interdep` is a long data frame where dyads 
are stacked on top of each other and both members of a dyad appear as separate 
rows. Roughly, the expected structure is:

- For cross-sectional data: one row per `dyad x member`

| dyad | member | x | y |
|---:|---:|---:|---:|
| 1 | 1 | 4.2 | 7.1 |
| 1 | 2 | 5.0 | 6.4 |
| 2 | 1 | 3.8 | 5.9 |
| 2 | 2 | 4.5 | 6.8 |

- For intensive longitudinal data: at most one row per `dyad x time x member`

| dyad | time | member | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 2 | 5.0 | 6.4 |
| 1 | 2 | 1 | 4.0 | 6.9 |
| 1 | 2 | 2 | 5.3 | 6.6 |


Measured variables may contain missing values. The structural `group`,
`member`, and optional `time` variables must be complete when observed (no
missing values are allowed).

In intensive longitudinal data, missing measurement occasions can be represented
by absent rows, as long as the time variable preserves the observed measurement
occasions. For example:

| dyad | personID | time | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 3 | 4.0 | 6.9 |
| 1 | 2 | 1 | 5.3 | 6.6 |
| 1 | 2 | 2 | 4.7 | 6.1 |
| 1 | 2 | 3 | 5.1 | 6.4 |

In this example, note that the row for person 1 at time 2 is absent. The time 
variable still uses the observed measurement occasions and skips from time 1 to 
time 3 for that person, which is fine and gives the models all the information 
they need. 


## Data preparation for distinguishable dyads

`example_dyadic_crosssectional` is a simulated cross-sectional dataset for
distinguishable dyads. Each dyad has two rows: one for each member.

```{r cross-sectional-raw}
print(head(example_dyadic_crosssectional))
```

We validate and prepare the data with the function `prepare_interdep_data()`

```{r prepare-cross-distinguishable}
cross_distinguishable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  
  # In this example, we optionally specify a predictor variable 
  # and a model type to generate the columns needed for that model type.
  predictors = communication,
  model_type = "apim"
)

print(cross_distinguishable_data, n = 4)

```

The function automatically recognized that in this dataset there are 95 
female-male dyads and created APIM-relevant variables. These generated `.i_*` 
columns can be used directly in model formulas. 

Here is a simple example:

```{r fit-cross-distinguishable-gaussian, eval = FALSE}
library(glmmTMB)

cross_distinguishable_model <- glmmTMB(
  satisfaction ~ 

    # Gender-specific intercepts
    0 + 
    .i_is_female_x_male_female + 
    .i_is_female_x_male_male + 
    
    # Gender-specific actor effects
    .i_is_female_x_male_female:.i_communication_raw_actor + 
    .i_is_female_x_male_male:.i_communication_raw_actor +
    
    # Gender-specific partner effects
    .i_is_female_x_male_female:.i_communication_raw_partner + 
    .i_is_female_x_male_male:.i_communication_raw_partner +
    
    # Dyad-level unstructured random effects represent the two partner
    # residual variances and their covariance when dispformula = ~ 0.
    # This is glmmTMB-specific syntax; lme4 and brms use different syntax.
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID)
  
  # Fix the observation-level Gaussian dispersion near zero so the dyad-level
  # random effects carry the partner residual covariance.
  , dispformula = ~ 0 
  , family = gaussian()
  , data = cross_distinguishable_data
)

summary(cross_distinguishable_model)

```

Models are not run and summarized in this vignette. For fitted APIM examples,
see the [Actor-Partner Interdependence Model vignette](apim.html).

## Data preparation for exchangeable dyads

For a dataset with only one type of dyad that should be treated as exchangeable, 
we can simply omit the `role` argument: 

```{r prepare-cross-exchangeable-basic}
cross_exchangeable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  seed = 123
)

print(cross_exchangeable_data, n = 4)

```

Alternatively, for more control, we can explicitly set dyad types to exchangeable:

```{r prepare-cross-set-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = "male-female",
  seed = 123
)

print(cross_exchangeable_data, n = 4)

```

*Note* that whenever you need to refer to a dyad type, the order of members does not matter
(e.g., 'male-female' and 'female-male' will both work), and you can use different separators 
like 'male_female', 'male_x_female' or even 'male female'. 

For exchangeable dyads, we can also request DIM and undirected DSM columns.
This works here because omitting `role` treats all dyads as a single type of
exchangeable dyads, and DIM and undirected DSM currently require exchangeable
dyads. Undirected DSM preparation additionally needs `outcomes`, because
it creates dyad-mean and within-dyad-deviation columns for both the predictor
and the outcome.

```{r prepare-cross-dim-dsm}
cross_dim_dsm_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  predictors = communication,
  outcomes = satisfaction,
  model_type = c("dim", "undirected_dsm"),
  seed = 123
)

print(cross_dim_dsm_data, n = 4)
```

## Incomplete dyads and missing roles

By default, `prepare_interdep_data()` stops when a dyad has only one observed
member or when a member's role cannot be resolved from the observed rows. These
cases can also be dropped before validation continues.

```{r incomplete-drop}
incomplete_data <- data.frame(
  coupleID = c(1, 2, 2, 3, 3, 4, 4, 5),
  personID = c(1, 3, 4, 5, 6, 7, 8, 10),
  gender = c("female", "female", NA, "female", "female", "female", "male", NA),
  satisfaction = c(5.2, 4.8, 4.9, 5.1, 5.0, 4.7, 4.6, 3.0)
)

incomplete_dropped_data <- prepare_interdep_data(
  incomplete_data,
  group = coupleID,
  member = personID,
  role = gender,
  incomplete_dyads = "drop",
  missing_role = "drop",
  seed = 123
)

print(incomplete_dropped_data)
```


## Intensive longitudinal dyadic data

`example_dyadic_ILD` is a simulated intensive longitudinal dyadic dataset. Each
dyad has repeated observations over `diaryday`, with one row per person-day.

The first rows look like this: 

```{r ild-raw}
print(example_dyadic_ILD, n = 26)
```

To prepare intensive longitudinal data, simply pass the `time` variable to 
`prepare_interdep_data`. 

### ILD APIM preparation

```{r prepare-ild-apim}
ild_apim_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  model_type = "apim",
  seed = 123
)

print(ild_apim_data)

```

By default, numeric predictors in longitudinal APIM preparation are
decomposed into within-person and between-person components. This temporal
predictor decomposition can be controlled via the
`temporal_predictor_decomposition` argument. Usually for ILD, the `time_2l`
option is the correct one. 

## Data with multiple and mixed-composition dyads

`example_dyadic_crosssectional_mixed` contains three dyad compositions in the
same data object: distinguishable female-male dyads and exchangeable
female-female and male-male dyads.

Letting the `interdep` package just infer the compositions automatically: 

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data, n = 4)
```

We can use this data to model these dyad types as separate or in the same model;
the [Actor-Partner Interdependence Model vignette](apim.html) shows both
mixed-composition formulas and practical convergence notes.

### Setting distinguishable dyads to be treated as exchangeable

As mentioned earlier, omitting the `role` argument treated all
dyads as one exchangeable composition, effectively pooling them. 

For more control, it is possible to separately set each distinguishable dyad 
composition in a mixed dataset to be treated as exchangeable. This specification 
keeps the differentiation between the kinds of dyads 
(e.g., `male-male`, `female-female`, and `male-female`).


```{r prepare-mixed-cross-sectional-exchangeable}
mixed_cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = c("male-female"),
  seed = 123
)

print(mixed_cross_exchangeable_data, n = 4)
```


### Pooling different dyad compositions

Sometimes for theoretical or practical reasons, we may want to 
**pool different exchangeable dyads** and analyze them as if they were one. 
This also allows testing various constraints via model comparisons later. 

For instance, let's pool `male-male` and `female-female` dyads
and name them `same-sex` dyads: 

```{r prepare-mixed-cross-sectional_pooled}
mixed_cross_data_pooled <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  pool_compositions = list(
    "same-sex" = c("male-male", "female_female")
  ),
  seed = 123
)

print(mixed_cross_data_pooled)
```

Note that you cannot pool distinguishable dyads. If we want to pool
`female-male` with `male-male` for some reason, we first have to treat
`female-male` as exchangeable:


```{r prepare-mixed-cross-sectional_pooled_constrained}
mixed_cross_data_pooled_constrained <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = "male_x_female",
  pool_compositions = list(
    "same-sex" = c("male-male", "male_female")
  ),
  seed = 123
)

print(mixed_cross_data_pooled_constrained)
```


If you want to exclude some dyad compositions completely, filter them before
preparation. A dedicated composition-inclusion argument is planned.

For model formulas and interpretation, continue with the
[Actor-Partner Interdependence Model vignette](apim.html),
[Dyad-Individual Model vignette](dim.html), or
[Undirected Dyadic Score Model vignette](undirected-dsm.html).
