---
title: "Getting Started"
author: "Pascal Küng"
description: >
  Core workflow for validating dyadic data, identifying dyad compositions, and
  preparing model-ready columns for dyadic multilevel models.
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dyadMLM)
```

## Installation

You can install the development version with:

```{r installation, eval=FALSE}
install.packages("dyadMLM", repos = c(
  "https://pascal-kueng.r-universe.dev",
  "https://cloud.r-project.org"
  )
)
```

## About this vignette

`dyadMLM` helps researchers prepare cross-sectional and intensive longitudinal
dyadic data for (generalized) multilevel models.

**This vignette focuses on automatic data preparation for multilevel models
(MLMs).** For a comparison of MLM and structural equation modeling (SEM)
approaches to dyadic data, see @ledermannAnalyzingDyadicData2017.

The model-fitting examples in the model-specific vignettes use the
`glmmTMB` package.

Post-estimation functions in `dyadMLM`, including model comparison and
back-transformation of exchangeable random-effect covariance structures, are
described in the [Actor-Partner Interdependence Model vignette](https://pascal-kueng.github.io/dyadMLM/articles/apim.html).

Other vignettes cover the [Dyad-Individual Model](dim.html) and its equivalence
to the *exchangeable APIM*, and the [Dyadic Score Model](dsm.html) and its
equivalence to the *distinguishable APIM*. They also describe the corresponding
coefficient and covariance transformations.

The [online package overview](https://pascal-kueng.github.io/dyadMLM/)
provides the current online versions of these vignettes and the complete
function reference.

## Prerequisites

The basic data structure needed for `dyadMLM` is a long data frame where dyads
are stacked on top of each other and both members of a dyad appear as separate
rows.

Roughly, the expected structure for `dyadMLM` is:

- For cross-sectional data: one row per `dyad x member`

| dyad | member | x | y |
|---:|---:|---:|---:|
| 1 | 1 | 4.2 | 7.1 |
| 1 | 2 | 5.0 | 6.4 |
| 2 | 1 | 3.8 | 5.9 |
| 2 | 2 | 4.5 | 6.8 |

- For intensive longitudinal data: at most one row per `dyad x time x member`

| dyad | time | member | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 2 | 5.0 | 6.4 |
| 1 | 2 | 1 | 4.0 | 6.9 |
| 1 | 2 | 2 | 5.3 | 6.6 |

Measured variables may contain missing values, but `dyad`, `member`, and
optional `time` must be complete; missing measurement occasions may instead be
represented by absent rows.

If your raw data are currently in wide format (for time or dyads or both),
reshape them to this long structure first. See the
[tidyr pivoting vignette](https://tidyr.tidyverse.org/articles/pivot.html) or
the [`pivot_longer()` reference](https://tidyr.tidyverse.org/reference/pivot_longer.html).

## Data preparation for distinguishable dyads

`dyads_cross` is a simulated cross-sectional dataset that contains three dyad
compositions.

Because these example data contain multiple compositions, `keep_compositions`
selects the composition modeled below. It can be omitted when the supplied data
already contain only the intended composition.

```{r cross-sectional-raw, echo = FALSE}
head(dyadMLM::dyads_cross)
```

We validate and prepare the data with the function `dyadMLM::prepare_dyad_data()`.

```{r prepare-cross-distinguishable}
cross_distinguishable_data <- dyadMLM::prepare_dyad_data(
  data = dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  predictors = provided_support,
  model_types = "apim",
  add_apim_gmc_predictors = TRUE, # Optional grand-mean centering

  # All three observed compositions in `dyads_cross` are detected and retained by
  # default. This example focuses on `female-male` dyads, so we restrict the
  # analysis here.
  keep_compositions = "female-male"
)

print(cross_distinguishable_data, n = 4)

```

The function retained 120 female-male dyads and created raw and grand-mean
centered APIM variables [@kennyPartnerEffectsRelationship1999].

For fitted APIM examples using these columns, see the
[Actor-Partner Interdependence Model vignette](apim.html).

## Data preparation for exchangeable dyads

For this example, we pretend that we have a dataset with no distinguishable
variable (e.g., same-sex friend dyads). Then, we simply omit the `role` argument:

```{r prepare-cross-exchangeable-basic}
cross_exchangeable_data <- dyadMLM::prepare_dyad_data(
  data = dyads_cross,
  dyad = coupleID,
  member = personID,
  predictors = provided_support,
  model_types = "apim",
  add_apim_gmc_predictors = TRUE,
  seed = 123
)

print(cross_exchangeable_data, n = 4)

```

The generated `.member_contrast_arbitrary` contrast assigns `-1`
and `1` to the two members of each exchangeable dyad
[@delrosarioPracticalGuideSpecifying2025]. Its direction is arbitrary, and
`seed` makes the assignment reproducible.

Refer to the APIM vignette's
[exchangeable APIM section](apim.html#exchangeable-residual-structure)
for how to use these columns to specify an exchangeable dyadic APIM and recover
the constrained actor-partner variance-covariance structure with
[`dyadMLM::recover_exchangeable_covariance()`](https://pascal-kueng.github.io/dyadMLM/reference/recover_exchangeable_covariance.html).


## Generating DIM and DSM columns

For exchangeable dyads, we can request DIM predictor columns.
**DIM preparation requires exactly one exchangeable composition**. This can be
achieved by omitting `role`. For more control over compositions, see
[Working with multiple dyad compositions](#working-with-multiple-dyad-compositions).

```{r prepare-cross-dim}
cross_dim_data <- dyadMLM::prepare_dyad_data(
  data = dyads_cross,
  dyad = coupleID,
  member = personID,
  predictors = provided_support,
  model_types = "dim",
  seed = 123
)

print(cross_dim_data, n = 4)
```

For distinguishable dyads, we can request DSM columns.
**DSM preparation currently requires exactly one distinguishable composition**.
To compute these columns, an explicit role order is required.
The role order defines the direction of all DSM predictor differences and the
DSM role contrast [@iidaModelsInterdependentIndividuals2018].

```{r prepare-cross-dsm}
cross_dsm_data <- dyadMLM::prepare_dyad_data(
  data = dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  predictors = provided_support,
  model_types = "dsm",
  dsm_role_order = c("female", "male"),
  keep_compositions = "female-male"
)

print(cross_dsm_data, n = 4)
```

APIM GMC uses all retained non-missing values. DIM and DSM center complete-pair
dyad means. The constants may differ with one-sided missingness.

In DIM and DSM, mean predictors are grand-mean centered by convention; thus,
no additional argument is required to request them.

## Intensive longitudinal dyadic data

`dyads_ild` is a simulated intensive longitudinal dyadic dataset. Each dyad
has repeated observations over `diaryday`, with one row per person-day.

```{r ild-raw, echo=FALSE}
head(dyads_ild)
```

To prepare intensive longitudinal data, pass the `time` variable to
`dyadMLM::prepare_dyad_data()`.

```{r prepare-ild-apim}
ild_apim_data <- dyadMLM::prepare_dyad_data(
  dyads_ild,
  dyad = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  model_types = "apim",
  keep_compositions = "female-male",
  seed = 123
)

print(ild_apim_data, n = 6)

```

By default, numeric predictors in longitudinal APIM preparation are
decomposed into within-person and between-person components
[@bolgerIntensiveLongitudinalMethods2013]. This temporal predictor decomposition
is controlled by `temporal_decomposition`. The default `"auto"`
setting selects `"2l"` (2-level temporal decomposition).

`add_apim_gmc_predictors = TRUE` requires resolved
`temporal_decomposition = "none"`, since the between-person component is
always already grand-mean centered by convention.

Note that observed person means used to construct the between-person (`cbp`)
predictors can be unreliable when each member contributes few occasions, which
can bias between-person estimates [@gottfredsonStraightforwardApproachCoping2019].

For fitted concurrent examples and AR(1) specifications for distinguishable
and exchangeable dyads, refer to the
[intensive longitudinal APIM section](apim.html#intensive-longitudinal-apims).

### Preparing lagged predictors

Lagged versions of variables, including an outcome that is also passed to
`predictors` for dynamic models, can be obtained through the
`lag1_predictors` argument.

Lagging respects the dyad and member structure, matches observations at exactly
`time - 1`, and does not bridge missing occasions when rows are missing
in the dataset.

```{r prepare-ild-apim-dynamic}
ild_apim_data_dynamic <- dyadMLM::prepare_dyad_data(
  dyads_ild,
  dyad = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = closeness,
  lag1_predictors = closeness,
  model_types = "apim",
  keep_compositions = "female-female",
  seed = 123
)

print(ild_apim_data_dynamic, n = 6)

```

**Note:** Whether to use the prepared raw or within-person-centered lagged
outcome depends on the research question and the data. See the
[dynamic ILD APIM example](apim.html#dynamic-models) for a more detailed
discussion and guidance.

## Working with multiple dyad compositions

`dyads_cross` contains three dyad compositions:
distinguishable female-male dyads and exchangeable female-female and male-male
dyads [@Bolger2025Unified].

Let's have `dyadMLM` infer the compositions automatically:

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- dyadMLM::prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data, n = 4)
```

Note that when role compositions are available, each *exchangeable* composition
receives its own difference contrast, such as
`.member_contrast_female_x_female_arbitrary`, which is `0` for all other
compositions.

We can use this data to model these dyad types as separate or in the same model.

### Keeping only selected dyad compositions (filtering)

Sometimes a mixed dataset contains dyad compositions that should not be part of
a given analysis. Use `keep_compositions` to keep only dyads whose *observed*
composition matches the requested labels. The filtering happens before
exchangeability constraints and pooling, so `set_exchangeable_compositions`
and `pool_compositions` (described later) can only refer to retained dyad
compositions.

```{r prepare-mixed-cross-sectional-included}
mixed_cross_data_included <- dyadMLM::prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  keep_compositions = c("female-female", "male-male"),
  seed = 123
)

print(mixed_cross_data_included, n = 4)
```

*Note* that whenever you need to refer to a dyad type, the order of members does
not matter (e.g., `male-female` and `female-male` will both work), and you can
use different separators like `male_female`, `male_x_female`, or `male female`.

### Setting distinguishable dyads to be treated as exchangeable

`set_exchangeable_compositions` changes how selected role-defined compositions
are modeled while preserving their composition labels. By contrast, omitting
`role` indicates that no distinguishing variable is available, so all dyads are
treated as one exchangeable composition with a single global arbitrary-member
contrast.

```{r prepare-mixed-cross-sectional-exchangeable}
mixed_cross_exchangeable_data <- dyadMLM::prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = c("male-female"),
  seed = 123
)

print(mixed_cross_exchangeable_data, n = 4)
```


### Pooling different dyad compositions

Sometimes, we may want to pool selected *exchangeable* dyad compositions and
analyze them as if they were one. Pooling can impose equality constraints among
compositions. After fitting nested pooled and unpooled models to the same
observations, these constraints can be tested with
`dyadMLM::compare_nested_models()`; see
[testing distinguishability in the APIM vignette](apim.html#testing-distinguishability)
for the model-comparison workflow.

For instance, let's pool `male-male` and `female-female` dyads
and name them `same-sex` dyads:

```{r prepare-mixed-cross-sectional_pooled}
mixed_cross_data_pooled <- dyadMLM::prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  pool_compositions = list(
    "same-sex" = c("male-male", "female_female")
  ),
  seed = 123
)

print(mixed_cross_data_pooled)
```

Note that you cannot pool distinguishable dyads. If we wanted to pool
`female-male` with `male-male`, we would first have to treat `female-male` as
exchangeable:

```{r prepare-mixed-cross-sectional_pooled_constrained}
mixed_cross_data_pooled_constrained <- dyadMLM::prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = "male female",
  pool_compositions = list(
    "pooled_exchangeable" = c("male-male", "male_female")
  ),
  seed = 123
)

print(mixed_cross_data_pooled_constrained)
```

## Citation

If you use `dyadMLM`, please cite the version of the package you use. Obtain the
citation via:

```{r eval = TRUE}
citation("dyadMLM")
```

---

**Continue** with the
[Actor-Partner Interdependence Model (APIM) vignette](apim.html).

Related model-specific vignettes:

- [Dyad-Individual Model vignette](dim.html),
- [Dyadic Score Model vignette](dsm.html),

or return to [About this vignette](#about-this-vignette).

## References

::: {#refs}
:::
