---
title: "Getting Started"
description: >
  Core workflow for validating dyadic data, identifying dyad compositions, and
  preparing model-ready columns for dyadic multilevel models.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

## About this Vignette

`interdep` helps researchers prepare cross-sectional and intensive longitudinal 
dyadic data for (generalized) multilevel models. It supports common dyadic 
designs with one type of exchangeable or distinguishable dyad, and it also 
handles studies where different kinds of dyads appear in the same dataset, 
such as female-male, female-female, and male-male couples. 

It automatically creates model-ready columns for dyadic multilevel model 
parameterizations such as Actor-Partner Interdependence Models (APIM), 
Dyad-Individual Modles (DIM), and undirected Dyadic Score Models (DSM). 
Current DIM and undirected DSM helpers require a single type of exchangeable 
dyad composition.

This vignette focuses on the automatic data preparation step.

For guidance and examples on how to use the prepared data to estimate various 
APIM models from simple Gaussian cross-sectional to generalized intensive 
longitudinal (ILD) models with mutliple dyad types, see the 
[Actor-Partner Interdependence Model vignette](apim.html). 

For guidance on how to use the Dyad-Individual Model (DIM) parameterization, 
including dyad-mean and within-dyad-deviation predictors and their equivalence 
to APIM effects in exchangeable dyads, see the [Dyad-Individual Model vignette](dim.html). 

For undirected dyadic score outcomes, see the [Undirected Dyadic Score Model vignette](undirected-dsm.html).

## Prerequisites

The basic data structure needed for `interdep` is a long data frame where dyads 
are stacked on top of each other and both members of a dyad appear as separate 
rows. Roughtly, the expected structure is:

- For cross-sectional data: one row per `dyad x member`

| dyad | member | x | y |
|---:|---:|---:|---:|
| 1 | 1 | 4.2 | 7.1 |
| 1 | 2 | 5.0 | 6.4 |
| 2 | 1 | 3.8 | 5.9 |
| 2 | 2 | 4.5 | 6.8 |

- For intensive longitudinal data: at most one row per `dyad x time x member`

| dyad | time | member | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 2 | 5.0 | 6.4 |
| 1 | 2 | 1 | 4.0 | 6.9 |
| 1 | 2 | 2 | 5.3 | 6.6 |


Measured variables may contain missing values. The structural `group`,
`member`, and optional `time` variables must be complete, or not observed (e.g.,
no NA values are allowed).

In intensive longitudinal data, missing measurement occasions can be represented
by absent rows, as long as the time variable preserves the observed measurement
occasions. For example:

| dyad | personID | time | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 3 | 4.0 | 6.9 |
| 1 | 2 | 1 | 5.3 | 6.6 |
| 1 | 2 | 2 | 4.7 | 6.1 |
| 1 | 2 | 3 | 5.1 | 6.4 |

In this example, note that the row for person 1 at time 2 is absent. The time 
variable still uses the observed measurement occasions and skips from time 1 to 
time 3 for that person, which is fine and gives the models all the information 
they need. 


## Data preparation for cross-sectional dyadic data

`example_dyadic_crosssectional` is a simulated cross-sectional dataset for
distinguishable dyads. Each dyad has two rows: one for each member.

```{r cross-sectional-raw}
print(head(example_dyadic_crosssectional))
```

We validate and prepare the data with the function `prepare_interdep_data()`

This is the minimal call: 

```{r prepare-cross-exchangeable-basic}
cross_exchangeable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  seed = 123
)

print(cross_exchangeable_data, n = 20)

```

Note that by not providing a `role` variable, we get a single type of
exchangeable dyads. If we want to distinguish by gender, we do: 

```{r prepare-cross-distinguishable}
cross_distinguishable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  
  # In this example, we also add predictors and a model-type to get 
  # model-ready columns for this model-type
  predictors = communication,
  model_type = "apim",
  seed = 123
)

print(cross_distinguishable_data, n = 20)

```
The function automatically recognized that in this dataset there are 95 
female-male dyads and created APIM-relevant variables. These generated `.i_*` 
columns can be used directly in model formulas. 

Here is a simple example:

```{r fit-cross-distinguishable-gaussian, eval = FALSE}
library(glmmTMB)

cross_distinguishable_model <- glmmTMB(
  satisfaction ~ 
    # remove standard intercept and model separate intercepts for male and female
    0 + 
    .i_is_female_x_male_female + 
    .i_is_female_x_male_male + 
    
    # gender-specific slopes for communication actor effect (no main effect)
    .i_is_female_x_male_female:.i_communication_raw_actor + 
    .i_is_female_x_male_male:.i_communication_raw_actor +
    
    # gender-specific slopes for communication partner effect (no main effect)
    .i_is_female_x_male_female:.i_communication_raw_partner + 
    .i_is_female_x_male_male:.i_communication_raw_partner +
    
    # residual covariance in glmmTMB can be modeled with random intercepts
    # when dispformula ~ 0. We model gender-specific residual variances and 
    # a correlation between the partners' residuals. 
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID)
  
  , dispformula = ~ 0 
  , family = gaussian()
  , data = cross_distinguishable_data
)

summary(cross_distinguishable_model)

```

Models are not run and summarized in this vignette. For fitted APIM examples,
see the [Actor-Partner Interdependence Model vignette](apim.html).


## Incomplete dyads and missing roles

By default, `prepare_interdep_data()` stops when a dyad has only one observed
member or when a member's role cannot be resolved from the observed rows. These
cases can also be dropped before validation continues.

```{r incomplete-drop}
incomplete_data <- data.frame(
  coupleID = c(1, 2, 2, 3, 3, 4, 4, 5),
  personID = c(1, 3, 4, 5, 6, 7, 8, 10),
  gender = c("female", "female", NA, "female", "female", "female", "male", NA),
  satisfaction = c(5.2, 4.8, 4.9, 5.1, 5.0, 4.7, 4.6, 3.0)
)

incomplete_dropped_data <- prepare_interdep_data(
  incomplete_data,
  group = coupleID,
  member = personID,
  role = gender,
  incomplete_dyads = "drop",
  missing_role = "drop",
  seed = 123
)

print(incomplete_dropped_data)
```


## Intensive longitudinal dyadic data

`example_dyadic_ILD` is a simulated intensive longitudinal dyadic dataset. Each
dyad has repeated observations over `diaryday`, with one row per person-day.

```{r ild-raw}
print(head(example_dyadic_ILD, n = 26), n = 26)
```

For longitudinal data, simply pass the `time` variable as well.

### Exchangeable ILD preparation

```{r prepare-ild-exchangeable-dim}
ild_exchangeable_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  time = diaryday,
  predictors = provided_support,
  
  # This time we request DIM-style variables. This only works
  # for exchangeable dyads, which is why we omit the `role` variable here. 
  model_type = 'dim',
  seed = 123
)

print(ild_exchangeable_data)

```

By default, numeric predictors in longitudinal model-ready preparation are
decomposed into within-person and between-person components. This temporal
predictor decomposition can be controlled via the
`temporal_predictor_decomposition` argument. Usually for ILD, the `time_2l`
option is the correct one. 

## Data with multiple and mixed-composition dyads

`example_dyadic_crosssectional_mixed` contains three dyad compositions in the
same data object: distinguishable female-male dyads and exchangeable
female-female and male-male dyads.


```{r mixed-cross-sectional-raw}
print(example_dyadic_crosssectional_mixed)
```

Letting the `interdep` package just infer the compositions automatically: 

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data)
```

We can use this data to model these dyad types as separate or in the same model;
the [Actor-Partner Interdependence Model vignette](apim.html) shows both
mixed-composition formulas and practical convergence notes.

### Setting distinguishable dyads to be treated as exchangeable

Previously we treated all dyads in the same dataset as the same type of indistinguishable
by omitting the `role` argument. If you need more control, you can separately 
set each distinguishable dyad composition in the dataset to be treated as 
exchangeable.  

Note that whenever you need to refer to a dyad-type, the order of members does not matter
(e.g., 'male-female' and 'female-male' will both work), and you can use different separators 
like 'male_female', 'male_x_female' or even 'male female'. 


```{r prepare-mixed-cross-sectional-exchangeable}
mixed_cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = c('male-female'),
  seed = 123
)

print(mixed_cross_exchangeable_data)
```

This specification keeps the differentiation between the kinds of dyads
(`male-male`, `female-female`, and `male-female`), while treating the
`male-female` composition as exchangeable.


### Pooling different dyad-compositions

Sometimes for theoretical or practical reasons, we may want to 
**pool different exchangeable dyads** and analyze them as if they were one. 
This also allows to test various constraints via model comparisons. 

For instance, let's pool `male-male` and `female-female` dyads
and name them `same-sex` dyads: 

```{r prepare-mixed-cross-sectional_pooled}
mixed_cross_data_pooled <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  pool_compositions = list(
    'same-sex' = c('male-male', 'female_female')
  ),
  seed = 123
)

print(mixed_cross_data_pooled)
```

Note that you cannot pool distinguishable dyads. If you would want to pool female-male with male-male you could treat female-male as exchangeable: 



```{r prepare-mixed-cross-sectional_pooled_constrained}
mixed_cross_data_pooled_constrained <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = 'male_x_female',
  pool_compositions = list(
    'same-sex' = c('male-male', 'male_female')
  ),
  seed = 123
)

print(mixed_cross_data_pooled_constrained)
```

If you want to exclude some dyad compositions completely, filter them before
preparation. A dedicated composition-inclusion argument is planned.
