---
title: "Getting Started"
description: >
  Core workflow for validating dyadic data, identifying dyad compositions, and
  preparing model-ready columns for dyadic multilevel models.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```

## Installation

You can install the development version of `interdep` from GitHub with:

```{r installation, eval=FALSE}
# install.packages("pak")
pak::pak("Pascal-Kueng/interdep")
```

## About this Vignette

`interdep` helps researchers prepare cross-sectional and intensive longitudinal
dyadic data for (generalized) multilevel models. It supports common dyadic
designs with one type of exchangeable or distinguishable dyad, and it also
handles studies where different kinds of dyads appear in the same dataset,
such as female-male, female-female, and male-male couples.

It automatically creates model-ready columns for dyadic multilevel model
parameterizations such as Actor-Partner Interdependence Models (APIM),
Dyad-Individual Models (DIM), and undirected Dyadic Score Models (DSM).
Current DIM and undirected DSM helpers require a single type of exchangeable
dyad composition.

This vignette focuses on the automatic data preparation step.

For guidance and examples on how to use the prepared data to estimate
cross-sectional, generalized, and intensive longitudinal APIMs, see the
[Actor-Partner Interdependence Model vignette](apim.html).
For models that combine distinguishable and exchangeable dyad compositions,
see the [Mixed-Composition APIM vignette](mixed-apim.html).

For guidance on how to use the Dyad-Individual Model (DIM) parameterization,
including dyad-mean and within-dyad-deviation predictors and their equivalence
to APIM effects in exchangeable dyads, see the [Dyad-Individual Model vignette](dim.html).

For undirected dyadic score outcomes, see the [Undirected Dyadic Score Model vignette](undirected-dsm.html).

## Prerequisites

The basic data structure needed for `interdep` is a long data frame where dyads
are stacked on top of each other and both members of a dyad appear as separate
rows. Roughly, the expected structure is:

If your raw data are currently in wide format (for time or dyads or both),
reshape them to this long structure before using `prepare_interdep_data()`. See the
[tidyr pivoting vignette](https://tidyr.tidyverse.org/articles/pivot.html) or
the [`pivot_longer()` reference](https://tidyr.tidyverse.org/reference/pivot_longer.html).

- For cross-sectional data: one row per `dyad x member`

| dyad | member | x | y |
|---:|---:|---:|---:|
| 1 | 1 | 4.2 | 7.1 |
| 1 | 2 | 5.0 | 6.4 |
| 2 | 1 | 3.8 | 5.9 |
| 2 | 2 | 4.5 | 6.8 |

- For intensive longitudinal data: at most one row per `dyad x time x member`

| dyad | time | member | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 2 | 5.0 | 6.4 |
| 1 | 2 | 1 | 4.0 | 6.9 |
| 1 | 2 | 2 | 5.3 | 6.6 |


Measured variables may contain missing values. The structural `group`,
`member`, and optional `time` variables must not contain missing values.

In intensive longitudinal data, missing measurement occasions can be represented
by absent rows, as long as the time variable preserves the observed measurement
occasions. For example:

| dyad | personID | time | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 3 | 4.0 | 6.9 |
| 1 | 2 | 1 | 5.3 | 6.6 |
| 1 | 2 | 2 | 4.7 | 6.1 |
| 1 | 2 | 3 | 5.1 | 6.4 |

In this example, the row for person 1 at time 2 is absent. The time variable
preserves the observed measurement occasions and skips from time 1 to time 3
for that person. `interdep` accepts this structure without requiring a
placeholder row for the missing occasion.


## Data preparation for distinguishable dyads

`example_dyadic_crosssectional` is a simulated cross-sectional dataset for
distinguishable dyads. Each dyad has two rows: one for each member.

```{r cross-sectional-raw}
knitr::kable(
  head(example_dyadic_crosssectional),
  digits = 2,
  caption = "First six observations."
)
```

We validate and prepare the data with the function `prepare_interdep_data()`

```{r prepare-cross-distinguishable}
cross_distinguishable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,

  # In this example, we optionally specify a predictor variable
  # and a model type to generate the columns needed for that model type.
  predictors = communication,
  model_type = "apim"
)

print(cross_distinguishable_data, n = 4)

```

The function automatically recognized that in this dataset there are 95
female-male dyads and created APIM-relevant variables. These generated `.i_*`
columns can be used directly in model formulas.

Here is a simple example:

```{r fit-cross-distinguishable-gaussian, eval = requireNamespace("glmmTMB", quietly = TRUE), results = "hide"}
library(glmmTMB)

cross_distinguishable_model <- glmmTMB(
  satisfaction ~

    # Gender-specific intercepts
    0 +
    .i_is_female_x_male_female +
    .i_is_female_x_male_male +

    # Gender-specific actor effects
    .i_is_female_x_male_female:.i_communication_raw_actor +
    .i_is_female_x_male_male:.i_communication_raw_actor +

    # Gender-specific partner effects
    .i_is_female_x_male_female:.i_communication_raw_partner +
    .i_is_female_x_male_male:.i_communication_raw_partner +

    # Dyad-level unstructured random effects represent the two partner
    # residual variances and their covariance when dispformula = ~ 0.
    # This is glmmTMB-specific syntax! lme4 and brms use different syntax.
    us(0 +
         .i_is_female_x_male_female +
         .i_is_female_x_male_male
       | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_distinguishable_data
)
```

The model is fitted when `glmmTMB` is installed, but its output is omitted here
to keep the focus on data preparation. For fitted APIM examples, see the
[Actor-Partner Interdependence Model vignette](apim.html).

## Data preparation for exchangeable dyads

For a dataset with only one type of dyad that should be treated as exchangeable,
omit the `role` argument:

```{r prepare-cross-exchangeable-basic}
cross_exchangeable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  seed = 123
)

print(cross_exchangeable_data, n = 4)

```

The generated `.i_diff_assumed_exchangeable_arbitrary` contrast assigns `-1`
and `1` to the two members of each exchangeable dyad. Its direction is
arbitrary, and `seed` makes the assignment reproducible. When role compositions
are available, each exchangeable composition receives its own contrast, such as
`.i_diff_female_x_female_arbitrary`, which is `0` for all other compositions.
We use a fixed seed in the examples below for consistent results.

Alternatively, for more control, we can explicitly set dyad types to exchangeable:

```{r prepare-cross-set-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = "male-female",
  seed = 123
)

print(cross_exchangeable_data, n = 4)

```

*Note* that whenever you need to refer to a dyad type, the order of members does not matter
(e.g., `male-female` and `female-male` will both work), and you can use different separators
like `male_female`, `male_x_female`, or `male female`.

For exchangeable dyads, we can also request DIM and undirected DSM columns.
This works here because omitting `role` treats all dyads as a single type of
exchangeable dyads, and DIM and undirected DSM currently require exchangeable
dyads. Undirected DSM preparation additionally needs `outcomes`, because
it creates dyad-mean and within-dyad-deviation columns for both the predictor
and the outcome.

```{r prepare-cross-dim-dsm}
cross_dim_dsm_data <- prepare_interdep_data(
  data = example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  predictors = communication,
  outcomes = satisfaction,
  model_type = c("dim", "undirected_dsm"),
  seed = 123
)

print(cross_dim_dsm_data, n = 4)
```

## Incomplete dyads and missing roles

By default, `prepare_interdep_data()` stops when a dyad has only one observed
member or when a member's role cannot be resolved from the observed rows. These
cases can also be dropped before validation continues.

```{r incomplete-drop}
incomplete_data <- data.frame(
  coupleID = c(1, 2, 2, 3, 3, 4, 4, 5),
  personID = c(1, 3, 4, 5, 6, 7, 8, 10),
  gender = c("female", "female", NA, "female", "female", "female", "male", NA),
  satisfaction = c(5.2, 4.8, 4.9, 5.1, 5.0, 4.7, 4.6, 3.0)
)

incomplete_dropped_data <- prepare_interdep_data(
  incomplete_data,
  group = coupleID,
  member = personID,
  role = gender,
  incomplete_dyads = "drop",
  missing_role = "drop",
  seed = 123
)

print(incomplete_dropped_data)
```


## Intensive longitudinal dyadic data

`example_dyadic_ILD` is a simulated intensive longitudinal dyadic dataset. Each
dyad has repeated observations over `diaryday`, with one row per person-day.

The first rows look like this:

```{r ild-raw}
knitr::kable(
  head(example_dyadic_ILD, n = 8),
  digits = 2,
  caption = "First eight person-day observations."
)
```

To prepare intensive longitudinal data, pass the `time` variable to
`prepare_interdep_data()`.

### ILD APIM preparation

```{r prepare-ild-apim}
ild_apim_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  model_type = "apim",
  seed = 123
)

print(ild_apim_data)

```

By default, numeric predictors in longitudinal APIM preparation are
decomposed into within-person and between-person components. This temporal
predictor decomposition can be controlled via the
`temporal_predictor_decomposition` argument. Use `time_2l` when within-person
and between-person associations should be estimated separately.

## Data with multiple and mixed-composition dyads

`example_dyadic_crosssectional_mixed` contains three dyad compositions in the
same data object: distinguishable female-male dyads and exchangeable
female-female and male-male dyads.

Let `interdep` infer the compositions automatically:

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data, n = 4)
```

We can use this data to model these dyad types as separate or in the same model.
The [Actor-Partner Interdependence Model vignette](apim.html) shows both
mixed-composition formulas and practical convergence notes.

### Keeping only selected dyad compositions (filtering)

Sometimes a mixed dataset contains dyad compositions that should not be part of
a given analysis. Use `include_compositions` to keep only dyads whose *observed*
composition matches the requested labels. The filtering happens before exchangeability
constraints and pooling, so `set_exchangeable_compositions` and
`pool_compositions` arguments can only refer to retained types of dyads.

```{r prepare-mixed-cross-sectional-included}
mixed_cross_data_included <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  include_compositions = c("female-female", "male-male"),
  seed = 123
)

print(mixed_cross_data_included, n = 4)
```

### Setting distinguishable dyads to be treated as exchangeable

As mentioned earlier, omitting the `role` argument treated all
dyads as one exchangeable composition, effectively pooling them.

For more control, a distinguishable dyad composition in a mixed dataset can be
treated as exchangeable. This specification
keeps the differentiation between the kinds of dyads
(e.g., `male-male`, `female-female`, and `male-female`).


```{r prepare-mixed-cross-sectional-exchangeable}
mixed_cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = c("male-female"),
  seed = 123
)

print(mixed_cross_exchangeable_data, n = 4)
```


### Pooling different dyad compositions

Sometimes for theoretical or practical reasons, we may want to
**pool different exchangeable dyads** and analyze them as if they were one.
This also allows testing various constraints via model comparisons later.

For instance, let's pool `male-male` and `female-female` dyads
and name them `same-sex` dyads:

```{r prepare-mixed-cross-sectional_pooled}
mixed_cross_data_pooled <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  pool_compositions = list(
    "same-sex" = c("male-male", "female_female")
  ),
  seed = 123
)

print(mixed_cross_data_pooled)
```

Note that you cannot pool distinguishable dyads. To pool `female-male` with
`male-male`, we first have to treat
`female-male` as exchangeable:


```{r prepare-mixed-cross-sectional_pooled_constrained}
mixed_cross_data_pooled_constrained <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = "male_x_female",
  pool_compositions = list(
    "pooled_exchangeable" = c("male-male", "male_female")
  ),
  seed = 123
)

print(mixed_cross_data_pooled_constrained)
```


For model formulas and interpretation, continue with the
[Actor-Partner Interdependence Model vignette](apim.html),
[Mixed-Composition APIM vignette](mixed-apim.html),
[Dyad-Individual Model vignette](dim.html), or
[Undirected Dyadic Score Model vignette](undirected-dsm.html).

For an in-depth tutorial covering data preparation, model fitting, diagnostics,
and assumption checks, see
[Distinguishable and Exchangeable Dyads: Bayesian Multilevel Modelling](https://pascal-kueng.github.io/05DyadicDataAnalysis/).
It uses `interdep` for cross-sectional and intensive longitudinal APIM and DIM
workflows, with models fitted primarily using `brms`
([source](https://github.com/Pascal-Kueng/05DyadicDataAnalysis),
[DOI](https://doi.org/10.5281/zenodo.17400655)).
