---
title: "Getting Started"
description: >
  Core workflow for validating dyadic data, identifying dyad compositions, and
  preparing model-ready columns for dyadic multilevel models.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

`interdep` helps researchers prepare cross-sectional and intensive longitudinal
dyadic data for multilevel models, including generalized multilevel models. It
supports common dyadic studies with one kind of dyad, and it also handles
studies where different kinds of dyads appear in the same dataset, such as
female-male, female-female, and male-male couples. It creates
composition-aware, model-ready columns for dyadic multilevel model
parameterizations such as APIM, DIM, and undirected DSM. Current DIM and
undirected DSM helpers require one exchangeable dyad composition.

This vignette focuses mainly on APIM-style actor and partner effects in
multilevel model formulas, especially models with multiple dyad types,
intensive longitudinal data, and generalized outcomes. For the Dyad-Individual
Model (DIM) parameterization, including
dyad-mean and within-dyad-deviation predictors and their equivalence to APIM
effects in exchangeable dyads, see the
[Dyad-Individual Model vignette](dim.html).

The basic data structure is a long data frame where dyads are stacked on top of
each other and both members of a dyad appear as separate rows. For
cross-sectional data, each complete dyad contributes one row per member. For
intensive longitudinal data, each observed member-occasion has at most one row.
Dyads must contain two unique members overall, but member-occasion rows may be
absent.

The expected structure is:

- cross-sectional: one row per `dyad x member`

| dyad | member | x | y |
|---:|---:|---:|---:|
| 1 | 1 | 4.2 | 7.1 |
| 1 | 2 | 5.0 | 6.4 |
| 2 | 1 | 3.8 | 5.9 |
| 2 | 2 | 4.5 | 6.8 |

- intensive longitudinal: at most one row per `dyad x time x member`

| dyad | time | member | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 2 | 5.0 | 6.4 |
| 1 | 2 | 1 | 4.0 | 6.9 |
| 1 | 2 | 2 | 5.3 | 6.6 |


Measured variables may contain missing values. The structural `group`,
`member`, and optional `time` variables must be complete. Missing or incomplete
role and dyad information can be handled with the function arguments `missing_role` and
`incomplete_dyads`. 

In intensive longitudinal data, missing measurement occasions can be represented
by absent rows, as long as the time variable preserves the observed measurement
occasions. For example:

| dyad | personID | time | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 3 | 4.0 | 6.9 |
| 1 | 2 | 1 | 5.3 | 6.6 |
| 1 | 2 | 2 | 4.7 | 6.1 |
| 1 | 2 | 3 | 5.1 | 6.4 |

*Note* that the row for person 1 at time 2 is absent. The time variable still uses
the observed measurement occasions and skips from time 1 to time 3 for that
person, which is fine. 


`prepare_interdep_data()` validates the expected structure, returns a tibble
with class `interdep_data`, and adds the dyad-composition labels and
model-ready columns needed for dyadic multilevel model formulas.

Omit `role` only when all partners should be treated as exchangeable.

The examples build up from simple dyadic structures to composition-aware models:

- distinguishable dyads, where member roles are modeled separately;
- exchangeable dyads, where arbitrary member labels are handled with sum-diff
  columns;
- semi-continuous and intensive longitudinal outcomes;
- mixed-composition dyadic MLMs that combine distinguishable and exchangeable
  dyads in one model, illustrated in the
  [mixed-composition cross-sectional](#mixed-cross-sectional-gaussian-model) and
  [mixed-composition intensive longitudinal](#mixed-intensive-longitudinal-gaussian-model)
  examples.

## Cross-sectional dyadic data

`example_dyadic_crosssectional` is a simulated cross-sectional dataset for
distinguishable dyads. Each dyad has two rows: one for each member.

```{r cross-sectional-raw}
print(example_dyadic_crosssectional)
```

We validate and prepare the data with the function `prepare_interdep_data()`

```{r prepare-cross-distinguishable}
cross_distinguishable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  role = gender,
  
  # if we specify a predictor and a model type, we get transformed variables. 
  # here, .i_communication_raw_actor and .i_communication_raw_partner
  predictors = communication,
  model_type = "apim",
  seed = 123
)

print(cross_distinguishable_data, n = 20)

```

The generated `.i_*` columns can be used directly in model formulas.

### Distinguishable Gaussian APIM

```{r fit-cross-distinguishable-gaussian, eval = FALSE}
library(glmmTMB)

cross_distinguishable_model <- glmmTMB(
  satisfaction ~ 
    # remove standard intercept and model separate intercepts for male and female
    0 + 
    .i_is_female_x_male_female + 
    .i_is_female_x_male_male + 
    
    # gender-specific slopes for communication actor effect (no main effect)
    .i_is_female_x_male_female:.i_communication_raw_actor + 
    .i_is_female_x_male_male:.i_communication_raw_actor +
    
    # gender-specific slopes for communication partner effect (no main effect)
    .i_is_female_x_male_female:.i_communication_raw_partner + 
    .i_is_female_x_male_male:.i_communication_raw_partner +
    
    # residual covariance in glmmTMB can be modeled with random intercepts
    # when dispformula ~ 0. We model gender-specific residual variances and 
    # a correlation between the partners' residuals. 
    us(0 + 
         .i_is_female_x_male_female + 
         .i_is_female_x_male_male 
       | coupleID)
  
  , dispformula = ~ 0 
  , family = gaussian()
  , data = cross_distinguishable_data
)

summary(cross_distinguishable_model)

```

Note that models are not run and summarized in this vignette. For more infos
See the vignette !TODO insert link to APIM vignette.

### Exchangeable Gaussian APIM

We can treat **all dyads** as the same type of exchangeable dyads by simply 
omitting the `role` argument. The prepared data contains a composition indicator 
and a sum-diff contrast for the assumed exchangeable dyads.

```{r prepare-cross-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  predictors = communication,
  model_type = "apim",
  seed = 123
)

print(cross_exchangeable_data)
```


```{r fit-cross-exchangeable-gaussian-diff, eval = FALSE}

cross_exchangeable_diff_model <- glmmTMB(
  satisfaction ~ 
    # pooled intercept 
    1 + 
    
    # pooled actor slope for communication
    .i_communication_raw_actor +
    
    # pooled partner slope for communication
    .i_communication_raw_partner +
    
    # exchangeable residual covariance via sum-diff
    us(1 | coupleID) + 
    us(0 + .i_diff_assumed_exchangeable | coupleID)
    
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

summary(cross_exchangeable_diff_model)

```

Helper functions to rotate these structures back to partner-level
interpretations are planned for v.0.0.1.500


## Incomplete dyads and missing roles

By default, `prepare_interdep_data()` stops when a dyad has only one observed
member or when a member's role cannot be resolved from the observed rows. These
cases can also be dropped before validation continues.

The package example datasets are kept structurally complete, so this section
uses a small artificial example instead. With `"drop"`, dyads with incomplete
structure or unresolved role information are removed as whole dyads. The printed header records which dyads were removed.

```{r incomplete-drop}
incomplete_data <- data.frame(
  coupleID = c(1, 2, 2, 3, 3, 4, 4, 5),
  personID = c(1, 3, 4, 5, 6, 7, 8, 10),
  gender = c("female", "female", NA, "female", "female", "female", "male", NA),
  satisfaction = c(5.2, 4.8, 4.9, 5.1, 5.0, 4.7, 4.6, 3.0)
)

incomplete_dropped_data <- prepare_interdep_data(
  incomplete_data,
  group = coupleID,
  member = personID,
  role = gender,
  incomplete_dyads = "drop",
  missing_role = "drop",
  seed = 123
)

print(incomplete_dropped_data)
```



## Intensive longitudinal dyadic data

`example_dyadic_ILD` is a simulated intensive longitudinal dyadic dataset. Each
dyad has repeated observations over `diaryday`, with one row per person-day.

```{r ild-raw}
print(head(example_dyadic_ILD, n = 26), n = 26)
```

For longitudinal data, simply pass the `time` variable as well.

### Distinguishable ILD APIM

```{r prepare-ild-distinguishable}
ild_distinguishable_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(ild_distinguishable_data)

```

By default, numeric predictors in longitudinal APIM preparation are decomposed
into within-person and between-person components. This temporal predictor
decomposition can be controlled via the `temporal_predictor_decomposition`
argument. Usually for ILD, the `time_2l` option is the correct one. 

Example of an exchangeable ILD APIM. Note that *instead of omitting the role variable here*, 
we use a different method and set the dyad-type explicitly to exchangeable. 

```{r prepare-ild-exchangeable}
ild_exchangeable_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = c('male-female'), 
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(ild_exchangeable_data)
```

Note that whenever you need to refer to a dyad-type, the order of members does not matter
(e.g., 'male-female' and 'female-male' will both work), and you can use different separators 
like 'male_female', 'male_x_female' or even 'male female'. 

```{r fit-ild-exchangeable, eval = FALSE}

ild_exchangeable_model <- glmmTMB(
  closeness ~ 
    1 + 
    
    diaryday +

    # Pooled within-person actor and partner effects
    .i_provided_support_cwp_actor +
    .i_provided_support_cwp_partner +

    # Pooled between-person actor and partner effects
    .i_provided_support_cbp_actor +
    .i_provided_support_cbp_partner +

    # random effects for stable non-independence (means)
    us(1 | coupleID)  + us(0 + .i_diff_female_x_male | coupleID) +

    # Same-day residual covariance
    us(1 | coupleID:diaryday) + us(0 + .i_diff_female_x_male  | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

summary(ild_exchangeable_model)

```


## Mixed-composition dyads

`example_dyadic_crosssectional_mixed` contains three dyad compositions in the
same data object: distinguishable female-male dyads and exchangeable
female-female and male-male dyads.

Here, we focus on the preparation steps that are possible with `interdep`. For example 
models for multilevel models with multiple different types of distinguishable 
and indistinguishable dyads and a discussion, check the vignette here: !TODO add link.


```{r mixed-cross-sectional-raw}
print(example_dyadic_crosssectional_mixed)
```

Letting the `interdep` package just infer the compositions automatically: 

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data)
```

We can use this data to model these dyad types as separate (see Vignette X). ! TODO

However, sometimes for theoretical or practical reasons, we may want to pool 
different exchangeable dyads and analyze them as if they were one. 
This also allows to test various constraints via model comparisons. 

For instance, let's pool `male-male` and `female-female` dyads
and name them `same-sex` dyads: 

```{r prepare-mixed-cross-sectional}
mixed_cross_data_pooled <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  pool_compositions = list(
    'same-sex' = c('male-male', 'female-female')
  ),
  seed = 123
)

print(mixed_cross_data)
```

Note that you way use 'male-male', 'male_male', 'male_x_male' (what the package uses), or 'male male'. 


