---
title: "Preparing dyadic data for APIMs with interdep"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing dyadic data for APIMs with interdep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

`interdep` is being developed for dyadic data analysis with multilevel models (MLM). 
The basic data structure is a long data frame where dyads are stacked on top of each other and
both members of a dyad appear as separate rows. For cross-sectional data, each
complete dyad contributes one row per member. For intensive longitudinal data,
each complete dyad contributes one row per member and measurement occasion.

The expected structure is:

- cross-sectional: one row per `dyad x member`

| dyad | member | x | y |
|---:|---:|---:|---:|
| 1 | 1 | 4.2 | 7.1 |
| 1 | 2 | 5.0 | 6.4 |
| 2 | 1 | 3.8 | 5.9 |
| 2 | 2 | 4.5 | 6.8 |

- intensive longitudinal: one row per `dyad x time x member`

| dyad | time | member | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 2 | 5.0 | 6.4 |
| 1 | 2 | 1 | 4.0 | 6.9 |
| 1 | 2 | 2 | 5.3 | 6.6 |


Measured variables may contain missing values. The structural `group`,
`member`, and optional `time` variables must be complete. Missing or incomplete
role and dyad information can be handled with the function arguments `missing_role` and
`incomplete_dyads`. 

In intensive longitudinal data, missing measurement occasions can be represented
by absent rows, as long as the time variable preserves the observed measurement
occasions. For example:

| dyad | personID | time | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 3 | 4.0 | 6.9 |
| 1 | 2 | 1 | 5.3 | 6.6 |
| 1 | 2 | 2 | 4.7 | 6.1 |
| 1 | 2 | 3 | 5.1 | 6.4 |

*Note* that the row for person 1 at time 2 is absent. The time variable still uses
the observed measurement occasions and skips from time 1 to time 3 for that
person, which is fine. 


`prepare_interdep_data()` validates the expected structure,
returns a tibble with class `interdep_data`, derives and adds dyad-composition 
labels that are needed for modeling. 

Omit `role` only when all partners should be treated as exchangeable.

## Cross-sectional dyadic data

`example_dyadic_crosssectional` is a simulated cross-sectional dataset for
distinguishable dyads. Each dyad has two rows: one for each member.

```{r cross-sectional-raw}
print(head(example_dyadic_crosssectional))
```

We validate and prepare the data with the function `prepare_interdep_data()`

```{r prepare-cross-distinguishable}
cross_distinguishable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  role = gender,
  member = personID,
  seed = 123
)

print(cross_distinguishable_data)
summary(cross_distinguishable_data)
```

The generated `.i_*` columns can be used directly in model formulas.

```{r cross-distinguishable-columns}
names(cross_distinguishable_data)[startsWith(names(cross_distinguishable_data), ".i_")]
```

### Distinguishable Gaussian APIM


```{r fit-cross-distinguishable-gaussian, eval = has_glmmTMB}
library(glmmTMB)

cross_distinguishable_model <- glmmTMB(
  satisfaction ~ 
    # remove standard intercept and model separate intercepts for male and female
    0 + .i_is_female_x_male_female + .i_is_female_x_male_male + 
    
    # gender-specific slopes for communication (no main effect)
    .i_is_female_x_male_female:communication + .i_is_female_x_male_male:communication +
    
    # residual covariance in glmmTMB can be modeled with random intercepts
    # when dispformula ~ 0. We model gender-specific residual variances and 
    # a correlation between the partners' residuals. 
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID)
  
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_distinguishable_data
)

summary(cross_distinguishable_model)

```

### Exchangeable Gaussian APIM

We can treat all dyads as exchangeable by omitting the `role` argument.
The prepared data contains a composition indicator and a sum-diff contrast for
the assumed exchangeable dyads.

```{r prepare-cross-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  seed = 123
)

print(cross_exchangeable_data)
summary(cross_exchangeable_data)
```


```{r fit-cross-exchangeable-gaussian-diff, eval = has_glmmTMB}

cross_exchangeable_diff_model <- glmmTMB(
  satisfaction ~ 
    # pooled intercept 
    1 + 
    
    # pooled (actor) slope for communication
    communication +
    
    # exchangeable residual covariance via sum-diff
    us(1 | coupleID) + us(0 + .i_diff_assumed_exchangeable | coupleID)
    
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

summary(cross_exchangeable_diff_model)

```

Helper functions to rotate these structures back to partner-level
interpretations are planned.

### Test distinguishability

Aside from using a Wald test on the first model, nested model comparisons
require models fit to the same prepared data object. Helpers for creating those
constrained columns are planned.


## Incomplete dyads and missing roles

By default, `prepare_interdep_data()` stops when a dyad has only one observed
member or when a member's role cannot be resolved from the observed rows. These
cases can also be dropped before validation continues.

With `"drop"`, dyads with incomplete structure or unresolved role information
are removed as whole dyads.

```{r incomplete-drop}
incomplete_data <- data.frame(
  coupleID = c(1, 2, 2, 3, 3, 4, 4),
  personID = c(1, 1, 2, 1, 2, 1, 2),
  gender = c("female", "female", NA, "female", "female", "female", "male"),
  satisfaction = c(5.2, 4.8, 4.9, 5.1, 5.0, 4.7, 4.6)
)

incomplete_dropped_data <- prepare_interdep_data(
  incomplete_data,
  group = coupleID,
  member = personID,
  role = gender,
  incomplete_dyads = "drop",
  missing_role = "drop",
  seed = 123
)

incomplete_dropped_data
```



## Semi-continuous cross-sectional data

`example_dyadic_crosssectional_tweedie` has the same dyadic structure as before, but the
outcome has exact zeros and positive skewed values, as in
some physical activity outcomes, in contrast to the Gaussian distribution from the 
previous example. 

```{r tweedie-raw}
print(head(example_dyadic_crosssectional_tweedie))
hist(example_dyadic_crosssectional_tweedie$physical_activity, breaks = 20)
```

Validation and modeling work the same way because the dyadic structure is the same.
The random-effect interpretation is different from the Gaussian case: Tweedie
random effects are latent effects on the log-mean scale, while the
observation-level Tweedie variance remains part of the model.

### Distinguishable Tweedie APIM

```{r prepare-tweedie-distinguishable}
tweedie_distinguishable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_tweedie,
  group = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(tweedie_distinguishable_data)
summary(tweedie_distinguishable_data)
```


```{r fit-tweedie-distinguishable, eval = has_glmmTMB}

tweedie_distinguishable_model <- glmmTMB(
  physical_activity ~ 
    # remove standard intercept and model separate intercepts for male and female
    0 + .i_is_female_x_male_female + .i_is_female_x_male_male + 
    
    # gender-specific slopes for motivation (no main effect)
    .i_is_female_x_male_female:motivation + .i_is_female_x_male_male:motivation +
    
    # keep a simple couple-level latent effect for stable non-independence
    
    # important limitation: this can only induce positive partner dependence
    (1 | coupleID) 
  
  # allow role-specific Tweedie dispersion
  , dispformula = ~ 0 + .i_is_female_x_male_female + .i_is_female_x_male_male
  , family = tweedie()
  , data = tweedie_distinguishable_data
)

summary(tweedie_distinguishable_model)

```

### Exchangeable Tweedie APIM

```{r prepare-tweedie-exchangeable}
tweedie_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional_tweedie,
  group = coupleID,
  member = personID,
  # role = gender,
  seed = 123
)

print(tweedie_exchangeable_data)
summary(tweedie_exchangeable_data)
```

```{r fit-tweedie-exchangeable, eval = has_glmmTMB}

tweedie_exchangeable_model <- glmmTMB(
  physical_activity ~ 
    # pooled intercept 
    1 + 
    
    # pooled (actor) slope for motivation
    motivation +
    
    # exchangeable latent dyad block on the log-mean scale
    (1 | coupleID) + (0 + .i_diff_assumed_exchangeable | coupleID)
    
  # estimate a single pooled Tweedie dispersion parameter
  , dispformula = ~ 1
  , family = tweedie()
  , data = tweedie_exchangeable_data
)

summary(tweedie_exchangeable_model)

```

## Intensive longitudinal dyadic data

`example_dyadic_ILD` is a simulated intensive longitudinal dyadic dataset. Each
dyad has repeated observations over `diaryday`, with one row per person-day.

```{r ild-raw}
print(head(example_dyadic_ILD, n = 26), n = 26)
```

For longitudinal data, pass the time variable as well.

### Distinguishable ILD APIM

```{r prepare-ild-distinguishable}
ild_distinguishable_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  role = gender,
  member = personID,
  time = diaryday,
  seed = 123
)

print(ild_distinguishable_data)
summary(ild_distinguishable_data)
```

For now, not within- and between-person centred. 

```{r fit-ild-distinguishable, eval = has_glmmTMB}

ild_distinguishable_model <- glmmTMB(
  closeness ~ 0 + 
    
    .i_is_female_x_male_female + .i_is_female_x_male_male + 
    
    .i_is_female_x_male_female:diaryday + .i_is_female_x_male_male:diaryday +
    
    # random effects for stable non-independence (means)
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID)  +

    # Same-day residual covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID:diaryday) 

  , dispformula = ~ 0  
  , family = gaussian()
  , data = ild_distinguishable_data
)

summary(ild_distinguishable_model)

```

### Exchangeable ILD APIM

```{r prepare-ild-exchangeable}
ild_exchangeable_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  # role = gender,
  member = personID,
  time = diaryday,
  seed = 123
)

print(ild_exchangeable_data)
summary(ild_exchangeable_data)
```

For now, not within- and between-person centred. 

```{r fit-ild-exchangeable, eval = has_glmmTMB}

ild_exchangeable_model <- glmmTMB(
  closeness ~ 
    1 + 
    
    diaryday +

    # random effects for stable non-independence (means)
    us(1 | coupleID)  + us(0 + .i_diff_assumed_exchangeable | coupleID) +

    # Same-day residual covariance
    us(1 | coupleID:diaryday) + us(0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = ild_exchangeable_data
)

summary(ild_exchangeable_model)

```

## Semi-continuous intensive longitudinal dyadic data

`example_dyadic_ILD_tweedie` has the same intensive longitudinal dyadic
structure as `example_dyadic_ILD`, but the outcome is semi-continuous.

```{r ild-tweedie-raw}
print(head(example_dyadic_ILD_tweedie, n = 26), n = 26)
hist(example_dyadic_ILD_tweedie$physical_activity, breaks = 20)
```

### Distinguishable Tweedie ILD APIM

```{r prepare-ild-tweedie-distinguishable}
ild_tweedie_distinguishable_data <- prepare_interdep_data(
  example_dyadic_ILD_tweedie,
  group = coupleID,
  role = gender,
  member = personID,
  time = diaryday,
  seed = 123
)

print(ild_tweedie_distinguishable_data)
summary(ild_tweedie_distinguishable_data)
```

For now, not within- and between-person centred.

For Tweedie models, the random-effect blocks are latent effects on the log-mean
scale. They can induce partner dependence, but they are not residual covariance
structures in the same direct sense as in Gaussian models, because the Tweedie
observation-level variance remains part of the model.

The first model uses role-specific stable couple effects and a simple same-day
shared latent shock. This is the easier teaching model: it keeps role-specific
Tweedie dispersion, but the same-day latent shock can only induce positive
partner dependence.

```{r fit-ild-tweedie-distinguishable-shared-day, eval = has_glmmTMB}

ild_tweedie_distinguishable_shared_day_model <- glmmTMB(
  physical_activity ~ 0 +

    .i_is_female_x_male_female + .i_is_female_x_male_male +

    .i_is_female_x_male_female:diaryday + .i_is_female_x_male_male:diaryday +

    # random effects for stable non-independence (means)
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # same-day shared latent shock; positive dependence only
    (1 | coupleID:diaryday)

  , dispformula = ~ 0 + .i_is_female_x_male_female + .i_is_female_x_male_male
  , family = tweedie()
  , data = ild_tweedie_distinguishable_data
)

summary(ild_tweedie_distinguishable_shared_day_model)

```

The second model uses a role-specific same-day latent covariance block. This is
closer to the Gaussian ILD covariance structure and can represent positive or
negative same-day partner dependence. In this ILD example, the repeated paired
occasions provide enough information to also estimate role-specific Tweedie
dispersion.

```{r fit-ild-tweedie-distinguishable-latent-day-cov, eval = has_glmmTMB}

ild_tweedie_distinguishable_latent_day_cov_model <- glmmTMB(
  physical_activity ~ 0 +

    .i_is_female_x_male_female + .i_is_female_x_male_male +

    .i_is_female_x_male_female:diaryday + .i_is_female_x_male_male:diaryday +

    # random effects for stable non-independence (means)
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # same-day role-specific latent covariance; positive or negative dependence
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID:diaryday)

  , dispformula = ~ 0 + .i_is_female_x_male_female + .i_is_female_x_male_male
  # in case of non-convergence, a first simplification could be to remove the
  # role-specific dispersion formula
  , family = tweedie()
  , data = ild_tweedie_distinguishable_data
)

summary(ild_tweedie_distinguishable_latent_day_cov_model)

```

The fuller same-day latent covariance structure is much more fragile in the
cross-sectional case because each dyad contributes only one paired occasion. The
same pair of observations must inform the fixed effects, Tweedie dispersion and
power, and the dyadic dependence structure. In ILD data, each dyad contributes
many paired occasions, so the stable couple-level block and the same-day
occasion-level block are informed by repeated within-dyad patterns over time.

### Exchangeable Tweedie ILD APIM

```{r prepare-ild-tweedie-exchangeable}
ild_tweedie_exchangeable_data <- prepare_interdep_data(
  example_dyadic_ILD_tweedie,
  group = coupleID,
  member = personID,
  time = diaryday,
  seed = 123
)

print(ild_tweedie_exchangeable_data)
summary(ild_tweedie_exchangeable_data)
```

For now, not within- and between-person centred.

This is the exchangeable analogue of the fuller Tweedie ILD model above. The
sum-diff random-effect blocks provide stable and same-day exchangeable latent
covariance structures. The Tweedie dispersion stays pooled because the
sum-diff signs identify exchangeable positions, not substantive roles.

```{r fit-ild-tweedie-exchangeable, eval = has_glmmTMB}

ild_tweedie_exchangeable_model <- glmmTMB(
  physical_activity ~
    1 +

    diaryday +

    # stable exchangeable latent covariance
    (1 | coupleID) + (0 + .i_diff_assumed_exchangeable | coupleID) +
  
    # same-day exchangeable latent covariance
    (1 | coupleID:diaryday) + (0 + .i_diff_assumed_exchangeable | coupleID:diaryday)

  # pooled dispersion; exchangeable positions should not define dispersion differences
  , dispformula = ~ 1
  , family = tweedie()
  , data = ild_tweedie_exchangeable_data
)

summary(ild_tweedie_exchangeable_model)

```

## Unified cross-sectional Gaussian model

`example_dyadic_crosssectional_unified` contains three dyad compositions in the
same data object: distinguishable female-male dyads and exchangeable
female-female and male-male dyads.

```{r unified-cross-sectional-raw}
print(head(example_dyadic_crosssectional_unified))
```

```{r prepare-unified-cross-sectional}
unified_cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional_unified,
  group = coupleID,
  role = gender,
  member = personID,
  seed = 123
)

print(unified_cross_data)
summary(unified_cross_data)
```

The data were simulated with the following fixed effects and residual covariance
parameters. For exchangeable dyads, `sum_variance` and `diff_variance` imply the
partner correlation.

```{r unified-cross-sectional-targets, echo=FALSE}
unified_cross_targets <- data.frame(
  block = c(
    "female_x_male",
    "female_x_male",
    "female_x_male",
    "female_x_female",
    "female_x_female",
    "female_x_female",
    "male_x_male",
    "male_x_male",
    "male_x_male"
  ),
  parameter = c(
    "female_mean",
    "male_mean",
    "correlation",
    "mean",
    "sum_variance",
    "diff_variance",
    "mean",
    "sum_variance",
    "diff_variance"
  ),
  value = c(
    5.50,
    4.50,
    -0.30,
    5.80,
    0.675,
    0.325,
    4.20,
    0.63375,
    1.05625
  )
)

unified_cross_targets
```

```{r fit-unified-cross-sectional-gaussian, eval = has_glmmTMB}

unified_cross_gaussian_model <- glmmTMB(
  satisfaction ~ 0 +

    .i_is_female_x_male_female + .i_is_female_x_male_male +
    .i_is_female_x_female + .i_is_male_x_male +

    # distinguishable female-male residual covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # exchangeable female-female residual covariance via sum-diff
    (0 + .i_is_female_x_female | coupleID) +
    (0 + .i_diff_female_x_female | coupleID) +

    # exchangeable male-male residual covariance via sum-diff
    (0 + .i_is_male_x_male | coupleID) +
    (0 + .i_diff_male_x_male | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = unified_cross_data
)

summary(unified_cross_gaussian_model)

```
