---
title: "Validating dyadic data with interdep"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating dyadic data with interdep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```

`interdep` is being developed for dyadic data analysis. The basic data
structure is a long data frame where dyads are stacked on top of each other and
both members of a dyad appear as separate rows. For cross-sectional data, each
dyad should contribute one row per member. For intensive longitudinal data, each
dyad should contribute one row per member and measurement occasion.

The expected structure is:

- cross-sectional: one row per `dyad x member`

| dyad | member | x | y |
|---:|---:|---:|---:|
| 1 | 1 | 4.2 | 7.1 |
| 1 | 2 | 5.0 | 6.4 |
| 2 | 1 | 3.8 | 5.9 |
| 2 | 2 | 4.5 | 6.8 |

- intensive longitudinal: one row per `dyad x time x member`

| dyad | time | member | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 2 | 5.0 | 6.4 |
| 1 | 2 | 1 | 4.0 | 6.9 |
| 1 | 2 | 2 | 5.3 | 6.6 |


Measured variables may contain missing values. Structural variables used for
validation should not. `prepare_interdep_data()` validates the expected
structure, returns a tibble with class `interdep_data`, and stores the validated
dyad, member, optional role, and optional time columns as metadata for later
package functions. It also derives dyad-composition labels used by later
functions. Omit `role` only when partners should be treated as
exchangeable.

In intensive longitudinal data, missing measurement occasions can be represented
by absent rows, as long as the time variable preserves the observed measurement
occasions. For example:

| dyad | personID | time | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 3 | 4.0 | 6.9 |
| 1 | 2 | 1 | 5.3 | 6.6 |
| 1 | 2 | 2 | 4.7 | 6.1 |
| 1 | 2 | 3 | 5.1 | 6.4 |

*Note* that the row for person 1 at time 2 is absent. The time variable still uses
the observed measurement occasions and skips from time 1 to time 3 for that
person, which is fine. 

## Cross-sectional dyadic data

`example_dyadic_crosssectional` is a simulated cross-sectional dataset for
distinguishable dyads. Each dyad has two rows: one for each member.

```{r cross-sectional-data}
print(head(example_dyadic_crosssectional))
```

```{r validate-cross-sectional}
cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  role = gender,
  member = personID
)

class(cross_data)
attr(cross_data, "interdep")
```



## Semi-continuous cross-sectional data

`example_dyadic_crosssectional_tweedie` has the same dyadic structure, but the
outcome has exact zeros and positive skewed values, as in
some physical activity outcomes, in contrast to the Gaussian distribution from the 
previous example. 

```{r tweedie-data}
print(head(example_dyadic_crosssectional_tweedie))
hist(example_dyadic_crosssectional_tweedie$physical_activity, breaks = 20)
```

Validation works the same way because the dyadic structure is the same.

```{r validate-tweedie}
tweedie_data <- prepare_interdep_data(
  example_dyadic_crosssectional_tweedie,
  group = coupleID,
  role = gender,
  member = personID
)

attr(tweedie_data, "interdep")
```

## Intensive longitudinal dyadic data

`example_dyadic_ILD` is a simulated intensive longitudinal dyadic dataset. Each
dyad has repeated observations over `diaryday`, with one row per person-day.

```{r ild-data}
print(head(example_dyadic_ILD, n = 26), n = 26)
```

For longitudinal data, pass the time variable as well.

```{r validate-ild}
ild_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  role = gender,
  member = personID,
  time = diaryday
)

class(ild_data)
attr(ild_data, "interdep")
```
