---
title: "Validating dyadic data with interdep"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating dyadic data with interdep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```

`interdep` is being developed for dyadic data analysis with multilevel models (MLM). 
The basic data structure is a long data frame where dyads are stacked on top of each other and
both members of a dyad appear as separate rows. For cross-sectional data, each
complete dyad contributes one row per member. For intensive longitudinal data,
each complete dyad contributes one row per member and measurement occasion.

The expected structure is:

- cross-sectional: one row per `dyad x member`

| dyad | member | x | y |
|---:|---:|---:|---:|
| 1 | 1 | 4.2 | 7.1 |
| 1 | 2 | 5.0 | 6.4 |
| 2 | 1 | 3.8 | 5.9 |
| 2 | 2 | 4.5 | 6.8 |

- intensive longitudinal: one row per `dyad x time x member`

| dyad | time | member | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 2 | 5.0 | 6.4 |
| 1 | 2 | 1 | 4.0 | 6.9 |
| 1 | 2 | 2 | 5.3 | 6.6 |


Measured variables may contain missing values. The structural `group`,
`member`, and optional `time` variables must be complete; missing or incomplete
role and dyad information can be handled with the function arguments `missing_role` and
`incomplete_dyads`. 

In intensive longitudinal data, missing measurement occasions can be represented
by absent rows, as long as the time variable preserves the observed measurement
occasions. For example:

| dyad | personID | time | x | y |
|---:|---:|---:|---:|---:|
| 1 | 1 | 1 | 4.2 | 7.1 |
| 1 | 1 | 3 | 4.0 | 6.9 |
| 1 | 2 | 1 | 5.3 | 6.6 |
| 1 | 2 | 2 | 4.7 | 6.1 |
| 1 | 2 | 3 | 5.1 | 6.4 |

*Note* that the row for person 1 at time 2 is absent. The time variable still uses
the observed measurement occasions and skips from time 1 to time 3 for that
person, which is fine. 


`prepare_interdep_data()` validates the expected structure,
returns a tibble with class `interdep_data`, derives and adds dyad-composition 
labels that are needed for modeling. 

Omit `role` only when all partners should be treated as exchangeable.

## Cross-sectional dyadic data

`example_dyadic_crosssectional` is a simulated cross-sectional dataset for
distinguishable dyads. Each dyad has two rows: one for each member.

```{r cross-sectional-data}
print(head(example_dyadic_crosssectional))
```

We validate and prepare the data with the function `prepare_interdep_data()`

```{r validate-cross-sectional}
cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  role = gender,
  member = personID
)

print(cross_data)
summary(cross_data)
```

### Example: simple dyadic model for distinguishable dyads

```{r validate-cross-sectional-distinguishable}
cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  role = gender,
  member = personID
)

print(cross_data)
summary(cross_data)
```

```{r estimate-distinguishable}
library(glmmTMB)

model1 <- glmmTMB(
  satisfaction ~ 
    # remove standard intercept and model separate intercepts for male and female
    0 + .i_is_female_x_male_female + .i_is_female_x_male_male + 
    
    # gender-specific slopes for communication (no main effect)
    .i_is_female_x_male_female:communication + .i_is_female_x_male_male:communication +
    
    # residual covariance in glmmTMB can be modeled with random intercepts
    # when dispformula ~ 0. We model gender-specific residual variances and 
    # a correlation between the partners residuals. 
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID)
  
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_data
)

summary(model1)

```

### Example: simple dyadic model for exchangeable dyads

We can constrain ally dyads to be exchangeable by omitting the `role` argument: 

```{r validate-cross-sectional-exchangeable}
cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  # role = gender,
  member = personID
)

print(cross_data)
summary(cross_data)
```


```{r estimate-exchangeable1}

model2 <- glmmTMB(
  satisfaction ~ 
    # pooled intercept 
    1 + 
    
    # pooled (actor) slope for communication
    communication +
    
    # specify a single pooled residual covariance but allow for a correlation
    # between the partners' residuals. 
    homcs(0 + .i_is_arbitrary_role_1 + .i_is_arbitrary_role_2 | coupleID)
  
    # alternative specification with the sum-diff approach that can be rotated back.
    # us(1 | coupleID) + us(0 + Idiff | coupleID)
    
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_data
)

summary(model2)

```

Here is the equivalent model (note identical AIC, BIC, logLik etc.). The random effects
could be rotated back if needed to obtain a correlation etc. 

```{r estimate-exchangeable2}

model2.5 <- glmmTMB(
  satisfaction ~ 
    # pooled intercept 
    1 + 
    
    # pooled (actor) slope for communication
    communication +
    
    # alternative specification.
    us(1 | coupleID) + us(0 + .i_diff | coupleID)
    
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_data
)

summary(model2.5)

```

Demonstrating that the models are exactly identical:
```{r showing-equivalence}

anova(model2, model2.5)

```


### Test distinguishability

Aside from using a Wald-Test on the first model, we can also compare models, since
they are nested. 

```{r model-comparison}

# comparing the model for distinguishable dyads with the model for exchangeable dyads. 
anova(model1, model2)

```


## Incomplete dyads and missing roles

By default, `prepare_interdep_data()` stops when a dyad has only one observed
member or when a member's role is unknown in every row. These cases can also be
kept with unknown composition labels or dropped before validation continues.

With `"keep"`, the original rows are retained. The incomplete dyad and the
missing role are represented with `unknown` in the derived composition labels.

```{r incomplete-keep}
incomplete_data <- data.frame(
  coupleID = c(1, 2, 2, 3, 3, 4, 4),
  personID = c(1, 1, 2, 1, 2, 1, 2),
  gender = c("female", "female", NA, "female", "female", "female", "male"),
  satisfaction = c(5.2, 4.8, 4.9, 5.1, 5.0, 4.7, 4.6)
)

kept_data <- prepare_interdep_data(
  incomplete_data,
  group = coupleID,
  member = personID,
  role = gender,
  incomplete_dyads = "drop",
  missing_role = "keep"
)

kept_data
```



## Semi-continuous cross-sectional data

`example_dyadic_crosssectional_tweedie` has the same dyadic structure as before, but the
outcome has exact zeros and positive skewed values, as in
some physical activity outcomes, in contrast to the Gaussian distribution from the 
previous example. 

```{r tweedie-data}
print(head(example_dyadic_crosssectional_tweedie))
hist(example_dyadic_crosssectional_tweedie$physical_activity, breaks = 20)
```

Validation works the same way because the dyadic structure is the same.

```{r validate-tweedie}
tweedie_data <- prepare_interdep_data(
  example_dyadic_crosssectional_tweedie,
  group = coupleID,
  role = gender,
  member = personID
)

print(tweedie_data)
summary(tweedie_data)
```

## Intensive longitudinal dyadic data

`example_dyadic_ILD` is a simulated intensive longitudinal dyadic dataset. Each
dyad has repeated observations over `diaryday`, with one row per person-day.

```{r ild-data}
print(head(example_dyadic_ILD, n = 26), n = 26)
```

For longitudinal data, pass the time variable as well.

```{r validate-ild}
ild_data <- prepare_interdep_data(
  example_dyadic_ILD,
  group = coupleID,
  role = gender,
  member = personID,
  time = diaryday
)

print(ild_data)
summary(ild_data)
```
