---
title: "APIMs with Mixed Dyad Compositions"
author: "Pascal Küng"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{APIMs with Mixed Dyad Compositions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```


This vignette covers APIMs that combine distinguishable and exchangeable dyad
compositions in one analysis. It assumes familiarity with the data-preparation
workflow in the [Getting Started vignette](getting-started.html) and with the
single-composition models in the
[Actor-Partner Interdependence Model vignette](apim.html). For DIM predictors and their
equivalence to APIM effects in exchangeable dyads, see the
[Dyad-Individual Model vignette](dim.html). For DSM predictor scores and their
relationship to APIM effects in distinguishable dyads, see the
[Dyadic Score Model vignette](dsm.html).



> This vignette is under construction and for now only contains a few preliminary example models. Please check back soon!

# Cross-sectional mixed-dyads APIM

The data were simulated with the following fixed effects and residual covariance
parameters. For exchangeable dyads, `sum_variance` and `diff_variance` imply the
partner correlation.

```{r mixed-cross-sectional-targets, echo=FALSE}
mixed_cross_targets <- data.frame(
  block = c(
    "female_x_male",
    "female_x_male",
    "female_x_male",
    "female_x_female",
    "female_x_female",
    "female_x_female",
    "male_x_male",
    "male_x_male",
    "male_x_male"
  ),
  parameter = c(
    "female_mean",
    "male_mean",
    "correlation",
    "mean",
    "sum_variance",
    "diff_variance",
    "mean",
    "sum_variance",
    "diff_variance"
  ),
  value = c(
    5.50,
    4.50,
    -0.30,
    5.80,
    0.675,
    0.325,
    4.20,
    0.63375,
    1.05625
  )
)

mixed_cross_targets
```

Let's see if we can recover those parameters. 

We first prepare the data with `prepare_interdep_data`. It automatically finds
all dyad compositions. 

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data, n = 4)
```

We basically simply estimate a separate distinguishable or exchangeable APIM 
per dyad composition. The tricky part is the random effects setup. 

.i_diff variables need to be structured as containing -1 and 1 for the corresponding
composition, and 0 for each other composition. `prepare_interdep_data` does this
automatically. Then, the following model can be specified:

```{r fit-mixed-cross-sectional-gaussian, eval = FALSE}

mixed_cross_gaussian_model <- glmmTMB(
  satisfaction ~ 0 +

    .i_is_female_x_male_female + .i_is_female_x_male_male +
    .i_is_female_x_female + .i_is_male_x_male +

    # distinguishable female-male residual covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # exchangeable female-female residual covariance via sum-diff
    (0 + .i_is_female_x_female | coupleID) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID) +

    # exchangeable male-male residual covariance via sum-diff
    (0 + .i_is_male_x_male | coupleID) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_cross_data
)

summary(mixed_cross_gaussian_model)

```

### Single versus separate mixed-composition fits

The mixed-composition models in this vignette fit all dyad compositions in one
model call. That is useful when the goal is to
compare effects across compositions, test equality constraints, keep one
model-based covariance matrix for those comparisons, or intentionally share
parameters such as a common Tweedie power parameter.

A fixed-effect formula with all compositions in one model call does not, by
itself, create partial pooling across dyad compositions. The
composition-specific intercepts and slopes above are ordinary fixed effects, so
estimates for female-female or male-male dyads are not automatically shrunk
toward estimates from the other dyad types. If every mean, variance, and
covariance parameter is composition-specific, the likelihood largely factorizes
by composition. In the cross-sectional Gaussian example, the single
mixed-composition fit and three separate composition-specific fits give the same
log-likelihood and fixed-effect estimates up to numerical tolerance.

Partial pooling requires a different model specification, such as a common effect
plus composition deviations with a hierarchical prior, or a random dyad-type
effect. With only a few dyad types, frequentist random dyad-type variance
components are usually weakly identified. Complete pooling, where a slope is
constrained to be identical across compositions, can be more stable and more
powerful when the constraint is substantively correct, but it is biased when the
dyad compositions truly differ.


## Mixed-composition intensive longitudinal Gaussian model

`example_dyadic_ILD_mixed` contains the same three dyad compositions as the
mixed-composition cross-sectional example, but each dyad contributes repeated paired
observations over `diaryday`.

To prepare the data: 

```{r prepare-mixed-ild}
mixed_ild_data <- prepare_interdep_data(
  example_dyadic_ILD_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(mixed_ild_data)
```


This first mixed-composition ILD model includes composition-specific intercepts, fixed
time slopes, and within-person and between-person actor and partner effects.
Random time slopes are omitted; with separate stable and same-day dyadic
covariance blocks, random time slopes add many weakly identified parameters and
are a likely first source of convergence problems.

This maximal covariance model is a demanding all-in-one likelihood. In this
simulated example, separate composition-specific fits recover the same
fixed-effect solution cleanly, while the all-in-one mixed-composition fit can
report false convergence even when it reaches the same likelihood and has a
positive definite Hessian.
Treat convergence diagnostics, Hessian diagnostics, and comparison with simpler
or separate fits as part of the workflow. If the all-in-one fit is not clean, fit
the composition-specific models separately for estimation, or simplify the
same-day and stable covariance blocks before adding more random effects.

```{r fit-mixed-ild-gaussian, eval = FALSE}

mixed_ild_gaussian_model <- glmmTMB(
  closeness ~ 0 +

    # Composition-specific intercepts
    .i_is_female_x_male_female + .i_is_female_x_male_male +
    
    .i_is_female_x_female + 
    .i_is_male_x_male +

    # Composition-specific time trends
    .i_is_female_x_male_female:diaryday +
    .i_is_female_x_male_male:diaryday +
    
    .i_is_female_x_female:diaryday +
    
    .i_is_male_x_male:diaryday +

    # Composition-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +
    
    .i_is_female_x_female:.i_provided_support_cwp_actor +
    
    .i_is_male_x_male:.i_provided_support_cwp_actor +

    # Composition-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    .i_is_female_x_female:.i_provided_support_cwp_partner +
    
    .i_is_male_x_male:.i_provided_support_cwp_partner +

    # Composition-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +
    
    .i_is_female_x_female:.i_provided_support_cbp_actor +
    
    .i_is_male_x_male:.i_provided_support_cbp_actor +

    # Composition-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    .i_is_female_x_female:.i_provided_support_cbp_partner +
    
    .i_is_male_x_male:.i_provided_support_cbp_partner +

    # stable dyad-level covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +
    
    (0 + .i_is_female_x_female | coupleID) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID) +
    
    (0 + .i_is_male_x_male | coupleID) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID) +

    # same-day covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID:diaryday) +
    
    (0 + .i_is_female_x_female | coupleID:diaryday) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID:diaryday) +
    
    (0 + .i_is_male_x_male | coupleID:diaryday) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_ild_data
)

summary(mixed_ild_gaussian_model)

```


## Mixed-composition intensive longitudinal Tweedie model

`example_dyadic_ILD_mixed_tweedie` has the same mixed-composition intensive
longitudinal dyadic structure, but the outcome is semi-continuous.

```{r mixed-ild-tweedie-raw}
print(head(example_dyadic_ILD_mixed_tweedie, 20))
hist(example_dyadic_ILD_mixed_tweedie$physical_activity, breaks = 20)
```

```{r prepare-mixed-ild-tweedie}
mixed_ild_tweedie_data <- prepare_interdep_data(
  example_dyadic_ILD_mixed_tweedie,
  group = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(mixed_ild_tweedie_data)
summary(mixed_ild_tweedie_data)
```

As in the Gaussian mixed-composition ILD model, the mean model contains
composition-specific intercepts, fixed time slopes, and within-person and
between-person actor and partner effects. The dyadic random-effect blocks are
latent effects on the log-mean scale, not residual covariance parameters in the
Gaussian sense. This is the most fragile example in the vignette because it
combines composition-specific mean effects, composition-specific latent dyadic
dependence, composition-specific dispersion, and a common Tweedie power
parameter. A practical workflow is to fit separate Gaussian or simpler Tweedie
models first, then add shared-parameter constraints only when they answer a
specific comparison question.

```{r fit-mixed-ild-tweedie, eval = FALSE}

mixed_ild_tweedie_model <- glmmTMB(
  physical_activity ~ 0 +

    # Intercepts for each composition_role
    .i_is_female_x_male_female + 
    .i_is_female_x_male_male +
    
    .i_is_female_x_female + 
    
    .i_is_male_x_male +

    # C-specific specific time trends
    .i_is_female_x_male_female:diaryday + 
    .i_is_female_x_male_male:diaryday +
    
    .i_is_female_x_female:diaryday +
    
    .i_is_male_x_male:diaryday +

    # Composition-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +
    
    .i_is_female_x_female:.i_provided_support_cwp_actor +
    
    .i_is_male_x_male:.i_provided_support_cwp_actor +

    # Composition-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    .i_is_female_x_female:.i_provided_support_cwp_partner +
    
    .i_is_male_x_male:.i_provided_support_cwp_partner +

    # Composition-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +
    
    .i_is_female_x_female:.i_provided_support_cbp_actor +
    
    .i_is_male_x_male:.i_provided_support_cbp_actor +

    # Composition-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    .i_is_female_x_female:.i_provided_support_cbp_partner +
    
    .i_is_male_x_male:.i_provided_support_cbp_partner +

    # stable dyad-level latent covariance
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +
    
    (0 + .i_is_female_x_female | coupleID) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID) +
    
    (0 + .i_is_male_x_male | coupleID) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID) +

    # same-day latent covariance
    (0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID:diaryday) +
    
    (0 + .i_is_female_x_female | coupleID:diaryday) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID:diaryday) +
    
    (0 + .i_is_male_x_male | coupleID:diaryday) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID:diaryday)

    # recommended first simplification in convergence issues is to set dispformula = ~ 1
  , dispformula = ~ 0 +
      .i_is_female_x_male_female + .i_is_female_x_male_male +
      .i_is_female_x_female + .i_is_male_x_male
  , family = tweedie()
  , data = mixed_ild_tweedie_data
)

summary(mixed_ild_tweedie_model)

```

---

**Continue** with the [Dyad-Individual Model vignette](dim.html),

refer to the: 

- [Actor-Partner Interdependence Model (APIM) vignette](apim.html),
- [Dyadic Score Model vignette](dsm.html),

or return to the [Overview](index.html).
