---
title: "APIMs with Mixed Dyad Compositions"
author: "Pascal Küng"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{APIMs with Mixed Dyad Compositions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```


This vignette covers APIMs that combine distinguishable and exchangeable dyad
compositions in one analysis. It assumes familiarity with the data-preparation
workflow in the [Getting Started vignette](getting-started.html) and with the
single-composition models in the
[Actor-Partner Interdependence Model vignette](apim.html). For DIM predictors and their
equivalence to APIM effects in exchangeable dyads, see the
[Dyad-Individual Model vignette](dim.html). For DSM predictor scores and their
relationship to APIM effects in distinguishable dyads, see the
[Dyadic Score Model vignette](dsm.html).

## Cross-sectional mixed-composition APIM

The data were simulated with the following fixed effects and residual covariance
parameters. For exchangeable dyads, `sum_variance` and `diff_variance` imply the
partner correlation.

```{r mixed-cross-sectional-targets, echo=FALSE}
mixed_cross_targets <- data.frame(
  block = c(
    "female_x_male",
    "female_x_male",
    "female_x_male",
    "female_x_female",
    "female_x_female",
    "female_x_female",
    "male_x_male",
    "male_x_male",
    "male_x_male"
  ),
  parameter = c(
    "female_mean",
    "male_mean",
    "correlation",
    "mean",
    "sum_variance",
    "diff_variance",
    "mean",
    "sum_variance",
    "diff_variance"
  ),
  value = c(
    5.50,
    4.50,
    -0.30,
    5.80,
    0.675,
    0.325,
    4.20,
    0.63375,
    1.05625
  )
)

mixed_cross_targets
```

We first prepare the data with `prepare_interdep_data()`. It identifies all
observed dyad compositions.

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data, n = 4)
```

For each composition, the model uses either distinguishable or exchangeable
APIM terms. Most of the additional work is in the random-effects structure.

The `.i_diff_*` variables contain `-1` and `1` for the corresponding
exchangeable composition and `0` for all other compositions.
`prepare_interdep_data()` creates this structure automatically. The model can
then be specified as follows:

```{r fit-mixed-cross-sectional-gaussian, eval = FALSE}

mixed_cross_gaussian_model <- glmmTMB::glmmTMB(
  satisfaction ~ 0 +

    .i_is_female_x_male_female + .i_is_female_x_male_male +
    .i_is_female_x_female + .i_is_male_x_male +

    # distinguishable female-male residual covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # exchangeable female-female residual covariance via sum-diff
    (0 + .i_is_female_x_female | coupleID) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID) +

    # exchangeable male-male residual covariance via sum-diff
    (0 + .i_is_male_x_male | coupleID) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_cross_data
)

summary(mixed_cross_gaussian_model)

```

### Single versus separate mixed-composition fits

The mixed-composition models in this vignette fit all dyad compositions in one
model call. This is useful when the goal is to compare effects across
compositions, test equality constraints, or make those comparisons within one
joint fitted model.

A fixed-effect formula with all compositions in one model call does not, by
itself, create partial pooling across dyad compositions. The
composition-specific intercepts and slopes above are ordinary fixed effects, so
estimates for female-female or male-male dyads are not automatically shrunk
toward estimates from the other dyad types. If every mean, variance, and
covariance parameter is composition-specific, the likelihood largely
factorizes by composition. In that case, a combined fit can closely match
separate composition-specific fits. Its main advantage is that comparisons can
be made within one fitted model.

Partial pooling requires a different model specification, such as a common effect
plus composition deviations with a hierarchical prior, or a random dyad-type
effect. With only a few dyad types, frequentist random dyad-type variance
components are usually weakly identified. Complete pooling, where a slope is
constrained to be identical across compositions, can improve precision and
stability when the equality constraint is substantively reasonable. However,
it can hide meaningful differences between dyad compositions.


## Mixed-composition intensive longitudinal Gaussian model

`example_dyadic_ILD_mixed` contains the same three dyad compositions as the
mixed-composition cross-sectional example, but each dyad contributes repeated
paired observations over `diaryday`.

We prepare the data in the same way, now adding `time` and a predictor:

```{r prepare-mixed-ild}
mixed_ild_data <- prepare_interdep_data(
  example_dyadic_ILD_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(mixed_ild_data)
```


This mixed-composition ILD model includes composition-specific intercepts,
fixed time slopes, and within-person and between-person actor and partner
effects. Random time slopes are omitted because the separate stable and
same-day covariance blocks already introduce many variance parameters. They
should be added only when the research question and data support the additional
complexity.

The all-in-one covariance model is demanding and may produce optimizer
convergence warnings. A positive definite Hessian and estimates that are close
to simpler or separate fits are useful checks, but neither should be treated as
automatic proof that a warning can be ignored. If the combined fit is not
clean, fit the composition-specific models separately or simplify the stable
and same-day covariance blocks before adding more random effects.

```{r fit-mixed-ild-gaussian, eval = FALSE}

mixed_ild_gaussian_model <- glmmTMB::glmmTMB(
  closeness ~ 0 +

    # Composition-specific intercepts
    .i_is_female_x_male_female + .i_is_female_x_male_male +
    
    .i_is_female_x_female + 
    .i_is_male_x_male +

    # Composition-specific time trends
    .i_is_female_x_male_female:diaryday +
    .i_is_female_x_male_male:diaryday +
    
    .i_is_female_x_female:diaryday +
    
    .i_is_male_x_male:diaryday +

    # Composition-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +
    
    .i_is_female_x_female:.i_provided_support_cwp_actor +
    
    .i_is_male_x_male:.i_provided_support_cwp_actor +

    # Composition-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    .i_is_female_x_female:.i_provided_support_cwp_partner +
    
    .i_is_male_x_male:.i_provided_support_cwp_partner +

    # Composition-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +
    
    .i_is_female_x_female:.i_provided_support_cbp_actor +
    
    .i_is_male_x_male:.i_provided_support_cbp_actor +

    # Composition-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    .i_is_female_x_female:.i_provided_support_cbp_partner +
    
    .i_is_male_x_male:.i_provided_support_cbp_partner +

    # stable dyad-level covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +
    
    (0 + .i_is_female_x_female | coupleID) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID) +
    
    (0 + .i_is_male_x_male | coupleID) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID) +

    # same-day covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID:diaryday) +
    
    (0 + .i_is_female_x_female | coupleID:diaryday) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID:diaryday) +
    
    (0 + .i_is_male_x_male | coupleID:diaryday) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_ild_data
)

summary(mixed_ild_gaussian_model)

```


## Practical takeaway

Mixed dyad compositions do not require separate data-preparation calls.
`interdep` creates the composition-specific indicator and difference columns,
and the fitted model can place the corresponding fixed effects and covariance
terms in one formula.

It is often useful to begin with separate composition-specific models or
simpler covariance structures. Combine them when a comparison or equality
constraint is part of the research question, and add further random effects
only when the data support the additional complexity.

---

**Continue** with the [Dyad-Individual Model vignette](dim.html), refer to the
[Actor-Partner Interdependence Model vignette](apim.html) or the
[Dyadic Score Model vignette](dsm.html), or return to the
[Overview](index.html).
