---
title: "APIMs with Mixed Dyad Compositions"
author: "Pascal Küng"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{APIMs with Mixed Dyad Compositions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

diagram_helper <- if (file.exists("diagram-helpers.Rinc")) {
  "diagram-helpers.Rinc"
} else {
  file.path("vignettes", "diagram-helpers.Rinc")
}
sys.source(diagram_helper, envir = knitr::knit_global())
```

```{r setup}
library(interdep)
```

This vignette covers APIMs that combine distinguishable and exchangeable dyad
compositions in one analysis. The unified SEM presentation by Bolger,
Laurenceau, and DiGiovanni is the basis and inspiration for this vignette
[@Bolger2025Unified]. Here, we implement the same general idea in one multilevel
model.

For the general data-preparation workflow, go to the
[Getting Started vignette](getting-started.html). For regular APIMs, refer to the
[Actor-Partner Interdependence Model vignette](apim.html). For DIM predictors and their
equivalence to APIM effects in exchangeable dyads, see the
[Dyad-Individual Model vignette](dim.html). For DSM predictor scores and their
relationship to APIM effects in distinguishable dyads, see the
[Dyadic Score Model vignette](dsm.html).

## Cross-sectional mixed-composition APIM

The data were simulated with the following fixed effects and covariance
parameters. For exchangeable dyads, `sum_variance` and `diff_variance` imply the
within-dyad correlation.

```{r mixed-cross-sectional-targets, echo=FALSE}
mixed_cross_targets <- data.frame(
  block = c(
    "female_x_male",
    "female_x_male",
    "female_x_male",
    "female_x_female",
    "female_x_female",
    "female_x_female",
    "male_x_male",
    "male_x_male",
    "male_x_male"
  ),
  parameter = c(
    "female_mean",
    "male_mean",
    "correlation",
    "mean",
    "sum_variance",
    "diff_variance",
    "mean",
    "sum_variance",
    "diff_variance"
  ),
  value = c(
    5.50,
    4.50,
    -0.30,
    5.80,
    0.675,
    0.325,
    4.20,
    0.63375,
    1.05625
  )
)

mixed_cross_targets
```

We first prepare the data with `prepare_interdep_data()`, which identifies all
observed dyad compositions.

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data, n = 4)
```

For each composition, the model uses either distinguishable or exchangeable
APIM terms. For exchangeable compositions, it uses a sum-and-difference
parameterization. This gives the two members equal variances while allowing
their within-dyad covariance to be estimated
[@delrosarioPracticalGuideSpecifying2025].

The returned `.i_diff_*` variables contain `-1` and `1` for the corresponding
exchangeable composition and `0` for all other compositions.
`prepare_interdep_data()` creates this structure automatically.
The [APIM vignette](apim.html#exchangeable-residual-structure) derives
the resulting covariance matrix and shows how to back-transform the sum and
difference variances to the usual equal-variance member-level covariance
matrix.

The three composition-specific blocks are combined within one joint model:

```{r mixed-composition-diagram, echo=FALSE, fig.width=10, fig.height=5.5, out.width="100%", fig.cap="Structure of the cross-sectional mixed-composition APIM. Female-male dyads have distinguishable member intercepts and an unrestricted residual covariance block. Each same-gender composition has a pooled intercept and an exchangeable residual covariance block. The three blocks are estimated jointly but are not connected by covariance parameters.", fig.alt="Diagram of one joint mixed-composition model containing three separate blocks. The female-male block has distinct female and male intercepts, separate residual variances, and their covariance. Female-female and male-male blocks each have one shared intercept, equal member residual variances, and a covariance. No covariance parameters connect the three composition blocks."}
draw_mixed_apim_diagram()
```

The model can then be specified as follows (example for an intercept-only model):

```{r fit-mixed-cross-sectional-gaussian}

mixed_cross_gaussian_model <- glmmTMB::glmmTMB(
  satisfaction ~ 0 +

    ##### INTERCEPTS ######
    # fixed intercepts for individuals from distinguishable female-male couples
    .i_is_female_x_male_female + .i_is_female_x_male_male +

    # fixed pooled intercept for female-female couples
    .i_is_female_x_female +

    # fixed pooled intercept for male-male couples
    .i_is_male_x_male +

    ##### RESIDUAL STRUCTURE ######
    # distinguishable female-male residual covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # exchangeable female-female residual covariance via sum-diff
    us(0 + .i_is_female_x_female | coupleID) +
    us(0 + .i_diff_female_x_female_arbitrary | coupleID) +

    # exchangeable male-male residual covariance via sum-diff
    us(0 + .i_is_male_x_male | coupleID) +
    us(0 + .i_diff_male_x_male_arbitrary | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_cross_data
)

summary(mixed_cross_gaussian_model)

```

The four fixed-effect indicators are mutually exclusive. With $I$ denoting the
generated `.i_is_{composition_role}` indicator and $b_0$ a fixed intercept, the
fixed part of the full model can be written as:

$$
\mu_{ij}^{\mathrm{fixed}} =
b_{0,\mathrm{FM,F}} I_{\mathrm{FM,F},ij} +
b_{0,\mathrm{FM,M}} I_{\mathrm{FM,M},ij} +
b_{0,\mathrm{FF}} I_{\mathrm{FF},ij} +
b_{0,\mathrm{MM}} I_{\mathrm{MM},ij}.
$$

In `FM,F`, the first part identifies the female-male composition and the final
`F` identifies the female outcome member. The same composition-first convention
is used throughout this vignette.

The active indicator for each person is shown below:

| Row | $I_{\mathrm{FM,F}}$ | $I_{\mathrm{FM,M}}$ | $I_{\mathrm{FF}}$ | $I_{\mathrm{MM}}$ | Fixed intercept |
|:--|--:|--:|--:|--:|:--|
| Female in a female-male dyad | **1** | 0 | 0 | 0 | $b_{0,\mathrm{FM,F}}$ |
| Male in a female-male dyad | 0 | **1** | 0 | 0 | $b_{0,\mathrm{FM,M}}$ |
| Member of a female-female dyad | 0 | 0 | **1** | 0 | $b_{0,\mathrm{FF}}$ |
| Member of a male-male dyad | 0 | 0 | 0 | **1** | $b_{0,\mathrm{MM}}$ |

Therefore, for a member of a male-male dyad, for example, the fixed part reduces
to just $b_{0,\mathrm{MM}}$ via:

$$
0 \cdot b_{0,\mathrm{FM,F}} + 0 \cdot b_{0,\mathrm{FM,M}} +
0 \cdot b_{0,\mathrm{FF}} + 1 \cdot b_{0,\mathrm{MM}}
= b_{0,\mathrm{MM}}.
$$

The random-effects terms reduce in the same way because their indicator columns
are also `0` outside the relevant composition. The model combines an
unrestricted distinguishable covariance block for female-male dyads with a
separate equal-variance covariance block for each exchangeable composition.
The main APIM vignette introduces the
[distinguishable](apim.html#distinguishable-residual-structure) and
[exchangeable](apim.html#exchangeable-residual-structure) structures.

The random-effects terms form composition-specific blocks, so no covariance
parameters are estimated across compositions. Such parameters would not be
identified because each dyad belongs to only one composition.

This illustration only covered the fixed-intercept model. The ILD example below
shows how composition-specific slopes can be added.

### Single versus separate mixed-composition fits

The mixed-composition models described in this vignette fit all dyad compositions in one
model call. This is useful when the goal is to compare effects across
compositions, test equality constraints, or make those comparisons within one
joint fitted model.

`marginaleffects::hypotheses()` allows for testing linear
combinations of `glmmTMB` fixed effects. In these equations, `conditional_`
identifies coefficients from the conditional model. A positive estimate means
that the estimate on the left side is higher than the estimate on the right
side.

```{r mixed-cross-sectional-contrasts, eval = FALSE}
# Test whether male-male and female-female pooled intercepts are different
marginaleffects::hypotheses(
  mixed_cross_gaussian_model,
  hypothesis = "conditional_.i_is_male_x_male = conditional_.i_is_female_x_female",
  re.form = NA
)

# Test whether the average female-male intercept is different from the pooled
# male-male intercept
marginaleffects::hypotheses(
  mixed_cross_gaussian_model,
  hypothesis = "(conditional_.i_is_female_x_male_female + conditional_.i_is_female_x_male_male) / 2 = conditional_.i_is_male_x_male",
  re.form = NA
)

# Test whether the male intercept from a female-male couple is different from
# the pooled male intercept from a male-male couple
marginaleffects::hypotheses(
  mixed_cross_gaussian_model,
  hypothesis = "conditional_.i_is_female_x_male_male = conditional_.i_is_male_x_male",
  re.form = NA
)

```

Since we only want to compare fixed-effects here, we set `re.form = NA`.

It is important to note that such a fixed-effect formula with all compositions
in one model call does not create partial pooling across dyad
compositions. Estimates for female-female, for instance, are not
automatically shrunk toward estimates from the other dyad types. In this model,
almost every parameter is composition-specific. This means that the likelihood largely
factorizes by composition. Therefore, separate composition-specific fits would
largely recover the same parameters.

The main advantage of estimating everything in a single model is that formal
parameter comparisons can be made within one fitted model, and that models with
different versions of pooling can be compared. For example, it can be tested
whether pooling male-male and female-female couples as same-sex substantially
worsens model fit, considering both the fixed effects and random-effects
structure. Such model comparisons require the restricted and unrestricted
models to use the same observations and require the restricted model to be
nested within the unrestricted model.

If partial pooling is desired, for instance with few dyads for one particular type,
a different model specification would be required, such as a common effect
plus composition deviations.

### Model comparison with a restricted model

The [regular APIM vignette](apim.html#testing-distinguishability) introduces
`compare_interdep_models()` by comparing a distinguishable APIM with its
exchangeable restriction. Here, we use the same approach for a
mixed-composition constraint. We treat female-male dyads as exchangeable and
pool them with male-male dyads, while modeling female-female dyads separately.
In an applied analysis, these constraints would require a theoretical or empirical
justification.

```{r prepare-mixed-cross-sectional-constrained}
mixed_cross_data_constrained <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = "male-female",
  pool_compositions = list(
    "non_female_x_female" = c("male-male", "female-male")
  ),
  seed = 123
)

mixed_cross_gaussian_model_constrained <- glmmTMB::glmmTMB(
  satisfaction ~ 0 +
    .i_is_female_x_female +
    .i_is_non_female_x_female +

    # exchangeable female-female residual covariance via sum-diff
    us(0 + .i_is_female_x_female | coupleID) +
    us(0 + .i_diff_female_x_female_arbitrary | coupleID) +

    # pooled female-male and male-male residual covariance via sum-diff
    us(0 + .i_is_non_female_x_female | coupleID) +
    us(0 + .i_diff_non_female_x_female_arbitrary | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_cross_data_constrained
)

compare_interdep_models(
  restricted = mixed_cross_gaussian_model_constrained,
  full = mixed_cross_gaussian_model
)
```

Here, the constrained model fits substantially worse
than the full model. Note that the test evaluates all imposed constraints
jointly, including the pooled fixed effects and residual covariance structure.
It does not show which restriction accounts for the difference.

## Mixed-composition intensive longitudinal (ILD) Gaussian model

`example_dyadic_ILD_mixed` contains the same three dyad compositions as the
mixed-composition cross-sectional example, but each dyad contributes repeated
paired observations over `diaryday`.

To keep the example lighter, we pool female-female and male-male dyads into one
same-sex composition. This is a simplifying modeling assumption and does not
reflect all differences used to simulate these example data. In an applied
analysis, pooling should be theoretically justified or evaluated by comparing
pooled and unpooled models fitted to the same observations.

Aside from the pooling, we prepare the data in the same way, now adding `time`
and a predictor:

```{r prepare-mixed-ild}
mixed_ild_data <- prepare_interdep_data(
  example_dyadic_ILD_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  pool_compositions = list(
    "same-sex" = c("male-male", "female-female")
  ),
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(mixed_ild_data)
```

Note that observed person means used to construct the between-person (`cbp`) predictors
can be unreliable when each member contributes few occasions, which can bias
between-person estimates [@gottfredsonStraightforwardApproachCoping2019].

This mixed-composition ILD model includes composition-specific fixed
intercepts, time slopes, and within-person and between-person actor and partner
effects. It omits random slopes, as they would add several variances and
covariances to an already large model.

This model is complex and may produce optimizer convergence warnings,
especially when the random-effects structure is too
complex for the available data or some components are close to zero
[@delrosarioPracticalGuideSpecifying2025]. When near-zero covariance components
make a frequentist fit unstable, regularizing priors in a Bayesian model can
sometimes help, but they do not replace careful model checks
[@delrosarioPracticalGuideSpecifying2025]. A Bayesian implementation of this
mixed-composition model is beyond the scope of this vignette. For these example
data, we use the BFGS optimizer with a higher iteration limit.

```{r fit-mixed-ild-gaussian, eval = FALSE}

mixed_ild_gaussian_model <- glmmTMB::glmmTMB(
  closeness ~ 0 +

    # Composition-specific intercepts
    .i_is_female_x_male_female + .i_is_female_x_male_male +

    .i_is_same_sex +

    # Composition-specific time trends
    .i_is_female_x_male_female:diaryday +
    .i_is_female_x_male_male:diaryday +

    .i_is_same_sex:diaryday +

    # Composition-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +

    .i_is_same_sex:.i_provided_support_cwp_actor +

    # Composition-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +

    .i_is_same_sex:.i_provided_support_cwp_partner +

    # Composition-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +

    .i_is_same_sex:.i_provided_support_cbp_actor +

    # Composition-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +

    .i_is_same_sex:.i_provided_support_cbp_partner +

    # stable dyad-level covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    us(0 + .i_is_same_sex | coupleID) +
    us(0 + .i_diff_same_sex_arbitrary | coupleID) +

    # same-day covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID:diaryday) +

    us(0 + .i_is_same_sex | coupleID:diaryday) +
    us(0 + .i_diff_same_sex_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_ild_data
  , control = glmmTMB::glmmTMBControl(
      optimizer = stats::optim,
      optArgs = list(method = "BFGS"),
      optCtrl = list(maxit = 10000)
    )
)

summary(mixed_ild_gaussian_model)

```

This example omits random slopes, as they would add further complexity. To
extend the model with random slopes, the following blocks would replace the
stable dyad-level covariance terms in the model above:

```{r mixed-ild-random-slopes, eval = FALSE}
# distinguishable female-male random intercepts and slopes
us(
  0 +
    .i_is_female_x_male_female +
    .i_is_female_x_male_male +
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner |
    coupleID
) +

# exchangeable same-sex shared block
us(
  0 +
    .i_is_same_sex +
    .i_is_same_sex:.i_provided_support_cwp_actor +
    .i_is_same_sex:.i_provided_support_cwp_partner |
    coupleID
) +

# exchangeable same-sex difference block
us(
  0 +
    .i_diff_same_sex_arbitrary +
    .i_diff_same_sex_arbitrary:.i_provided_support_cwp_actor +
    .i_diff_same_sex_arbitrary:.i_provided_support_cwp_partner |
    coupleID
)
```

This extension estimates a very large covariance structure and may not be
supported by many datasets. Regularizing priors in a Bayesian framework may help
with estimation, but they do not necessarily make an unsupported random-effects
structure appropriate.

## Takeaway

Mixed dyad compositions do not always require separate models.
`interdep` creates the necessary columns to fit a joint model and apply various
pooling and exchangeability constraints.

It is often useful to begin with separate composition-specific models or
simpler covariance structures. Combine them when an equality constraint or a
comparison of effects across compositions is part of the research question,
and add further random effects only when the data support the additional
complexity.

---

Return to the [Actor-Partner Interdependence Model vignette](apim.html), see
the [Dyad-Individual Model vignette](dim.html) or the
[Dyadic Score Model vignette](dsm.html) for alternative parameterizations, or
return to the [Overview](index.html).

## References

::: {#refs}
:::
