---
title: "APIMs with Mixed Dyad Compositions"
author: "Pascal Küng"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{APIMs with Mixed Dyad Compositions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```


This vignette covers APIMs that combine distinguishable and exchangeable dyad
compositions in one analysis. The unified model presented by Bolger, Laurenceau,
and DiGiovanni is the basis and inspiration for this vignette. Their presentation
includes a hybrid model for these dyad types in SEM [@Bolger2025Unified]. Here,
we implement the same general idea in one multilevel model. The vignette assumes familiarity
with the data-preparation workflow in the
[Getting Started vignette](getting-started.html) and with the single-composition models in the
[Actor-Partner Interdependence Model vignette](apim.html). For DIM predictors and their
equivalence to APIM effects in exchangeable dyads, see the
[Dyad-Individual Model vignette](dim.html). For DSM predictor scores and their
relationship to APIM effects in distinguishable dyads, see the
[Dyadic Score Model vignette](dsm.html).

## Cross-sectional mixed-composition APIM

The data were simulated with the following fixed effects and covariance
parameters. For exchangeable dyads, `sum_variance` and `diff_variance` imply the
partner correlation.

```{r mixed-cross-sectional-targets, echo=FALSE}
mixed_cross_targets <- data.frame(
  block = c(
    "female_x_male",
    "female_x_male",
    "female_x_male",
    "female_x_female",
    "female_x_female",
    "female_x_female",
    "male_x_male",
    "male_x_male",
    "male_x_male"
  ),
  parameter = c(
    "female_mean",
    "male_mean",
    "correlation",
    "mean",
    "sum_variance",
    "diff_variance",
    "mean",
    "sum_variance",
    "diff_variance"
  ),
  value = c(
    5.50,
    4.50,
    -0.30,
    5.80,
    0.675,
    0.325,
    4.20,
    0.63375,
    1.05625
  )
)

mixed_cross_targets
```

We first prepare the data with `prepare_interdep_data()`, which identifies all
observed dyad compositions.

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- prepare_interdep_data(
  example_dyadic_crosssectional_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data, n = 4)
```

For each composition, the model uses either distinguishable or exchangeable
APIM terms. For exchangeable compositions, it uses a sum-and-difference
parameterization. This gives the two members equal variances while allowing
their within-dyad covariance to be estimated
[@delrosarioPracticalGuideSpecifying2025].

The returned `.i_diff_*` variables contain `-1` and `1` for the corresponding
exchangeable composition and `0` for all other compositions.
`prepare_interdep_data()` creates this structure automatically.

The model can then be specified as follows (example for an intercept-only model):

```{r fit-mixed-cross-sectional-gaussian, eval = FALSE}

mixed_cross_gaussian_model <- glmmTMB::glmmTMB(
  satisfaction ~ 0 +

    ##### INTERCEPTS ######
    # fixed intercepts for individuals from distinguishable female-male couples
    .i_is_female_x_male_female + .i_is_female_x_male_male +

    # fixed pooled intercept for female-female couples
    .i_is_female_x_female +

    # fixed pooled intercept for male-male couples
    .i_is_male_x_male +

    ##### RESIDUAL STRUCTURE ######
    # distinguishable female-male residual covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +

    # exchangeable female-female residual covariance via sum-diff
    (0 + .i_is_female_x_female | coupleID) +
    (0 + .i_diff_female_x_female_arbitrary | coupleID) +

    # exchangeable male-male residual covariance via sum-diff
    (0 + .i_is_male_x_male | coupleID) +
    (0 + .i_diff_male_x_male_arbitrary | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_cross_data
)

summary(mixed_cross_gaussian_model)

```

The four fixed-effect indicators are mutually exclusive. With $I$ denoting a
shorter label for each generated composition-and-role indicator, the fixed part
of the model can be written as:

$$
\mu_{ij}^{\mathrm{fixed}} =
\beta_{FM,F} I_{FM,F,ij} +
\beta_{FM,M} I_{FM,M,ij} +
\beta_{FF} I_{FF,ij} +
\beta_{MM} I_{MM,ij}.
$$

The active indicator for each person is shown below:

| Row | $I_{FM,F}$ | $I_{FM,M}$ | $I_{FF}$ | $I_{MM}$ | Fixed intercept |
|:--|--:|--:|--:|--:|:--|
| Female in a female-male dyad | **1** | 0 | 0 | 0 | $\beta_{FM,F}$ |
| Male in a female-male dyad | 0 | **1** | 0 | 0 | $\beta_{FM,M}$ |
| Member of a female-female dyad | 0 | 0 | **1** | 0 | $\beta_{FF}$ |
| Member of a male-male dyad | 0 | 0 | 0 | **1** | $\beta_{MM}$ |

Therefore, for a member of a male-male dyad, for example, the fixed part reduces
to

$$
0\beta_{FM,F} + 0\beta_{FM,M} + 0\beta_{FF} + 1\beta_{MM}
= \beta_{MM}.
$$

The random-effects terms reduce in the same way because their indicator columns
are also `0` outside the relevant composition. For a female-male dyad, the
`us()` term gives the female and male members separate random intercepts and
estimates their covariance:

$$
\mu_{Fj} = \beta_{FM,F} + u_{Fj}, \qquad
\mu_{Mj} = \beta_{FM,M} + u_{Mj}.
$$

For an exchangeable composition $C$ (female-female or male-male), the two
random-effects terms give a shared dyad effect $u_{Cj}$ and an opposite member
effect $v_{Cj}$:

$$
\mu_{Cij} = \beta_C + u_{Cj} + d_{Cij}v_{Cj},
\qquad d_{Cij} \in \{-1, 1\}.
$$

Thus, the shared effect moves both members together, while the difference effect
moves them in opposite directions. Its `-1`/`1` direction is arbitrary.

The five random-effects terms in the model form separate blocks: one correlated
female-male block, two female-female sum-and-difference blocks, and two
male-male sum-and-difference blocks. No random-effect covariance is estimated
between compositions. The two blocks within each exchangeable composition are
also separate, so the shared and difference effects are uncorrelated. Because
this is a cross-sectional model with one observation per member and
`dispformula = ~ 0`, these dyad-level random effects represent the member
variances and within-dyad covariance without an additional observation-level
residual variance.

This illustration is limited to the fixed-intercept model. The ILD example below
shows how composition-specific slopes can be added.

### Single versus separate mixed-composition fits

The mixed-composition models described in this vignette fit all dyad compositions in one
model call. This is useful when the goal is to compare effects across
compositions, test equality constraints, or make those comparisons within one
joint fitted model.

`marginaleffects::hypotheses()` provides a string interface for testing linear
combinations of `glmmTMB` fixed effects. In these equations, `conditional_`
identifies coefficients from the conditional model. A positive estimate means
that the quantity on the left side is higher than the quantity on the right
side.

```{r mixed-cross-sectional-contrasts, eval = FALSE}
# Test whether male-male and female-female pooled intercepts are different
marginaleffects::hypotheses(
  mixed_cross_gaussian_model,
  hypothesis = "conditional_.i_is_male_x_male = conditional_.i_is_female_x_female",
  re.form = NA
)

# Test whether the average female-male intercept is different from the pooled
# male-male intercept
marginaleffects::hypotheses(
  mixed_cross_gaussian_model,
  hypothesis = "(conditional_.i_is_female_x_male_female + conditional_.i_is_female_x_male_male) / 2 = conditional_.i_is_male_x_male",
  re.form = NA
)

# Test whether the male intercept from a female-male couple is different from
# the pooled male intercept from a male-male couple
marginaleffects::hypotheses(
  mixed_cross_gaussian_model,
  hypothesis = "conditional_.i_is_female_x_male_male = conditional_.i_is_male_x_male",
  re.form = NA
)

```

These are population-level fixed-effect comparisons. Setting `re.form = NA`
excludes dyad-specific random effects from the quantities being compared; it
does not remove the random-effects structure from the fitted model. Uncertainty
is calculated for the fixed-effect comparison, which is the intended target
here.

It is important to note that such a fixed-effect formula with all compositions
in one model call does not, by itself, create partial pooling across dyad
compositions. The composition-specific intercepts and slopes above are ordinary
fixed effects, so estimates for female-female or male-male dyads are not
automatically shrunk toward estimates from the other dyad types. If every mean,
variance, and covariance parameter is composition-specific, the likelihood largely
factorizes by composition. In that case, a combined fit can closely match
separate composition-specific fits. Its main advantage is that formal
parameter comparisons can be made within one fitted model, and that models with
different versions of pooling can be compared. For example, it can be tested
whether pooling male-male and female-female couples as same-sex substantially
worsens model fit, considering both the fixed effects and random-effects
structure. Such model comparisons require the restricted and unrestricted
models to use the same observations; separate fits to different compositions
should not be compared this way.

If partial pooling is desired, for instance with few dyads for one particular type,
a different model specification is required, such as a common effect
plus composition deviations with a hierarchical prior, or a random dyad-type
effect. However, with only a few dyad types, frequentist random dyad-type variance
components are usually weakly identified.

## Mixed-composition intensive longitudinal (ILD) Gaussian model

`example_dyadic_ILD_mixed` contains the same three dyad compositions as the
mixed-composition cross-sectional example, but each dyad contributes repeated
paired observations over `diaryday`.

To keep the example lighter, we pool female-female and male-male dyads into one
same-sex composition. This is a simplifying modeling assumption and does not
reflect all differences used to simulate these example data. In an applied
analysis, pooling should be theoretically justified or evaluated by comparing
pooled and unpooled models fitted to the same observations.

Aside from the pooling, we prepare the data in the same way, now adding `time`
and a predictor:

```{r prepare-mixed-ild}
mixed_ild_data <- prepare_interdep_data(
  example_dyadic_ILD_mixed,
  group = coupleID,
  member = personID,
  role = gender,
  pool_compositions = list(
    "same-sex" = c("male-male", "female-female")
  ),
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(mixed_ild_data)
```

Note that observed person means used to construct the between-person (`cbp`) predictors
can be unreliable when each member contributes few occasions, which can bias
between-person estimates [@gottfredsonStraightforwardApproachCoping2019].

This mixed-composition ILD model includes composition-specific fixed
intercepts, time slopes, and within-person and between-person actor and partner
effects. It omits random slopes, as they would add several variances and
covariances to an already large model.

This model is complex and may produce optimizer convergence warnings,
especially when the random-effects structure is too
complex for the available data or some components are close to zero
[@delrosarioPracticalGuideSpecifying2025]. When near-zero covariance components
make a frequentist fit unstable, regularizing priors in a Bayesian model can
sometimes help, but they do not replace careful model checks
[@delrosarioPracticalGuideSpecifying2025]. A Bayesian implementation of this
mixed-composition model is beyond the scope of this vignette. For these example
data, we use the BFGS optimizer with a higher iteration limit.

```{r fit-mixed-ild-gaussian, eval = FALSE}

mixed_ild_gaussian_model <- glmmTMB::glmmTMB(
  closeness ~ 0 +

    # Composition-specific intercepts
    .i_is_female_x_male_female + .i_is_female_x_male_male +
    
    .i_is_same_sex +

    # Composition-specific time trends
    .i_is_female_x_male_female:diaryday +
    .i_is_female_x_male_male:diaryday +
    
    .i_is_same_sex:diaryday +

    # Composition-specific within-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +
    
    .i_is_same_sex:.i_provided_support_cwp_actor +

    # Composition-specific within-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner +
    
    .i_is_same_sex:.i_provided_support_cwp_partner +

    # Composition-specific between-person actor effects
    .i_is_female_x_male_female:.i_provided_support_cbp_actor +
    .i_is_female_x_male_male:.i_provided_support_cbp_actor +
    
    .i_is_same_sex:.i_provided_support_cbp_actor +

    # Composition-specific between-person partner effects
    .i_is_female_x_male_female:.i_provided_support_cbp_partner +
    .i_is_female_x_male_male:.i_provided_support_cbp_partner +
    
    .i_is_same_sex:.i_provided_support_cbp_partner +

    # stable dyad-level covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID) +
    
    (0 + .i_is_same_sex | coupleID) +
    (0 + .i_diff_same_sex_arbitrary | coupleID) +

    # same-day covariance
    us(0 + .i_is_female_x_male_female + .i_is_female_x_male_male | coupleID:diaryday) +
    
    (0 + .i_is_same_sex | coupleID:diaryday) +
    (0 + .i_diff_same_sex_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_ild_data
  , control = glmmTMB::glmmTMBControl(
      optimizer = stats::optim,
      optArgs = list(method = "BFGS"),
      optCtrl = list(maxit = 10000)
    )
)

summary(mixed_ild_gaussian_model)

```

This worked example omits random slopes, as they would add further complexity.
One possible extension is to let the within-person actor and partner effects
vary across dyads. The following blocks would replace the stable dyad-level
covariance terms in the model above. Each `us()` block estimates all variances
and covariances among the terms inside it.

```{r mixed-ild-random-slopes, eval = FALSE}
# distinguishable female-male random intercepts and slopes
us(
  0 +
    .i_is_female_x_male_female +
    .i_is_female_x_male_male +
    .i_is_female_x_male_female:.i_provided_support_cwp_actor +
    .i_is_female_x_male_male:.i_provided_support_cwp_actor +
    .i_is_female_x_male_female:.i_provided_support_cwp_partner +
    .i_is_female_x_male_male:.i_provided_support_cwp_partner |
    coupleID
) +

# exchangeable same-sex shared block
us(
  0 +
    .i_is_same_sex +
    .i_is_same_sex:.i_provided_support_cwp_actor +
    .i_is_same_sex:.i_provided_support_cwp_partner |
    coupleID
) +

# exchangeable same-sex difference block
us(
  0 +
    .i_diff_same_sex_arbitrary +
    .i_diff_same_sex_arbitrary:.i_provided_support_cwp_actor +
    .i_diff_same_sex_arbitrary:.i_provided_support_cwp_partner |
    coupleID
)
```

This extension estimates a very large covariance structure and may not be
supported by many datasets. Regularizing priors may help with estimation, but 
they do not necessarily make an unsupported random-effects structure appropriate.

## Practical takeaway

Mixed dyad compositions do not require separate data-preparation calls.
`interdep` creates the composition-specific indicator and difference columns,
and the fitted model can place the corresponding fixed effects and covariance
terms in one formula.

It is often useful to begin with separate composition-specific models or
simpler covariance structures. Combine them when an equality constraint or a
comparison of effects across compositions is part of the research question,
and add further random effects only when the data support the additional
complexity.

---

Return to the [Actor-Partner Interdependence Model vignette](apim.html), see
the [Dyad-Individual Model vignette](dim.html) or the
[Dyadic Score Model vignette](dsm.html) for alternative parameterizations, or
return to the [Overview](index.html).
