---
title: "People Observed Without Their Partner"
author: "Pascal Küng"
description: >
  Keep singles, people whose partner did not take part, and partners lost or
  gained during a study in dyadic models, with two-part partner predictors.
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{People Observed Without Their Partner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dyadMLM)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

Dyadic samples often include people observed without their partner. Some
have no partner (singles, widowed people). Others have a partner who did not
take part. And in longitudinal studies, people can lose or gain a partner.
This vignette shows how `prepare_dyad_data()` keeps all of them with
`partner_exists`, and how to model them together with complete dyads.

It builds on the
[Actor-Partner Interdependence Model vignette](apim.html), which introduces
the exchangeable residual structure with shared and contrast terms
([Modeling the residual random-effects structure](apim.html#exchangeable-residual-structure)).

> **Software.** These models use `glmmTMB`. Standard `brms` models with one
> outcome variable do not currently support the exact Gaussian residual
> covariance structures shown here.

## Two kinds of people observed alone

People observed alone need different handling, depending on whether a partner
exists:

- **No partner** (e.g., single or widowed): there is no partner value. We call
  these people *singletons*. Their partner predictors are set to 0, and the
  model gets a status term (two-part coding, below).
- **Partner not in the data**: a partner exists but did not take part. Their
  partner predictors are missing.

`prepare_dyad_data()` tells them apart through `partner_exists`. In
cross-sectional data, it gives each kind its own composition (`.composition`)
and indicator column (`.is_*`):

| Situation | `partner_exists` | With `role` | Without `role` |
|---|---|---|---|
| Both members observed | `TRUE` | e.g., `female_x_male` | `assumed_exchangeable` |
| No partner (singleton) | `FALSE` | `singleton_<role>` | `singleton` |
| Partner not in the data | `TRUE` | `<role>_x_missing` | `missing_partner` |

In longitudinal data, labels stay the same across occasions. For example, a
couple keeps its composition after a separation or death. `partner_exists`
then marks the occasions without a partner.

Base `partner_exists` on recorded status, such as relationship status or
widowhood, not on whether both people supplied data. A rule such as
`dplyr::n() == 2` mixes the two kinds of people observed alone.

## Partners not in the data

This is the simpler case. The partner exists, so their values are simply
missing, as for any other missing data. People whose partner did not take
part (`<role>_x_missing`) keep their rows, so models can use their available
outcomes [@knaflCrossSectionalFamilyDyads2009]. Models with partner predictors
drop these rows (listwise deletion), while actor-only models keep them. Check
`nobs()`, and compare results with and without these people, for example in an
actor-only model. An appropriate dyadic imputation model can retain them
under a plausible missing-at-random assumption.

When reporting, count the rows left out only because the partner predictor is
missing (here with placeholder column names):

```{r report-missing-partners, eval = FALSE}
left_out <- is.na(data$.x_gmc_partner) &
  !is.na(data$outcome) & !is.na(data$.x_gmc_actor)
sum(left_out)                              # n rows
dplyr::n_distinct(data$personID[left_out]) # from m people
```

For example: "*n* rows from *m* people had missing partner predictors. They
were retained for centering and excluded from models including those
predictors."

## Singletons and complete dyads in one model

Complete dyads and singletons can share one model. This lets us estimate actor
effects while accounting for dependence between partners. Estimation can be
less reliable with few complete dyads and many people observed alone
[@duNumberDyadsEstimation2016].

For this illustration, we simulate new closeness scores using the support
scores in `dyads_cross`, with **one observation per person**. Each original
composition (female–male, female–female, and male–male) provides 80 complete
dyads and 40 singletons (`has_partner = 0`).

```{r create-dyad-solo-example, include = FALSE}
set.seed(20261001)
solo_source <- dyads_cross |>
  dplyr::group_by(coupleID) |>
  dplyr::mutate(
    source_composition = paste(sort(as.character(gender)), collapse = "_x_"),
    partner_support = rev(provided_support),
    shared_error = rnorm(1, sd = 0.6)
  ) |>
  dplyr::ungroup()

singleton_clusters <- solo_source |>
  dplyr::distinct(coupleID, source_composition) |>
  dplyr::group_by(source_composition) |>
  dplyr::slice_sample(prop = 1 / 3) |>
  dplyr::ungroup()

removed_members <- solo_source |>
  dplyr::semi_join(singleton_clusters, by = "coupleID") |>
  dplyr::group_by(coupleID) |>
  dplyr::slice_sample(n = 1) |>
  dplyr::ungroup()

dyad_solo_example <- solo_source |>
  dplyr::anti_join(removed_members, by = "personID") |>
  dplyr::mutate(
    has_partner = as.integer(!coupleID %in% singleton_clusters$coupleID),
    closeness = 3 + 0.4 * (gender == "female") + 0.3 * has_partner +
      0.6 * (provided_support - 5) +
      0.3 * ifelse(has_partner == 1, partner_support - 5, 0) +
      shared_error + rnorm(dplyr::n(), sd = ifelse(gender == "female", 0.8, 1))
  ) |>
  dplyr::select(
    personID, coupleID, gender, source_composition,
    has_partner, closeness, provided_support
  )
```

### Prepare the data

`coupleID` identifies a complete dyad or one singleton. `personID` identifies
a person. Give unrelated singletons different `coupleID` values, never a
shared ID or `NA`.

We use the male-male sample for exchangeable dyads and the female-male sample
for distinguishable dyads. The female-female sample works the same way. We
keep each singleton's original simulation group (`source_composition`) only to
construct these examples. A singleton has no dyad composition.

```{r prepare-dyads-and-singletons, message = FALSE}
library(dplyr)

prepare_sample <- function(composition) {
  dyad_solo_example |>
    filter(source_composition == composition) |>
    prepare_dyad_data(
      dyad = coupleID, member = personID, role = gender,
      predictors = provided_support, add_apim_gmc_predictors = TRUE,
      partner_exists = has_partner == 1, seed = 123
    )
}

mm_data <- prepare_sample("male_x_male")
fm_data <- prepare_sample("female_x_male")
mm_data
```

The printed summary lists 80 male-male dyads and 40 singleton men. Singletons
get their own indicator (`.is_singleton_male`) and no member contrast. The
summary also suggests the two-part terms used below.

### Two-part partner predictors

A partner score is undefined when no partner exists. To retain these rows,
`prepare_dyad_data()` uses the two-part predictor of
@dziakTwoPartPredictors2017 (Method 1). After centering, partner predictors
are set to 0 where no partner existed, and `.partner_exists` (1 or 0) marks
these rows. Include `.partner_exists` as a fixed effect, so that rows without a
partner get their own mean, unless fixed intercepts already separate these
rows. Otherwise, the zeros are treated as real partner values.

The `gmc` columns are centered on the mean of everyone's support scores
before the zeros are set. With `family = gaussian()` and `dispformula = ~ 0`,
the random-effects terms represent the residual covariance (see the
[APIM vignette](apim.html#exchangeable-residual-structure)). Here, singletons
get their own residual variance, and dyad members get the exchangeable
structure:

```{r dyad-solo-partner-effects, eval = has_glmmTMB}
mm_separate <- glmmTMB::glmmTMB(
  closeness ~ 1 + .partner_exists +
    .provided_support_gmc_actor + .provided_support_gmc_partner +
    (0 + .is_singleton_male | coupleID) +
    (0 + .is_exchangeable | coupleID) +
    (0 + .member_contrast_arbitrary | coupleID),
  family = gaussian(), dispformula = ~ 0,
  data = mm_data
)

nobs(mm_separate)
```

All 200 people remain in the fit: 160 dyad members and 40 singletons. Here,
`1 + .partner_exists` is the same as `0 + .is_exchangeable + .is_singleton_male`,
because all singletons have no partner.

With $x_A$ and $x_P$ denoting centered scores, the fixed means are:

$$
\mu =
\begin{cases}
\beta_0 + \beta_A x_A, & \text{without a partner}, \\
\beta_0 + \beta_H + \beta_A x_A + \beta_P x_P,
  & \text{with a partner}.
\end{cases}
$$

The status coefficient $\beta_H$ compares people with and without a partner
at equal actor support, when partner support equals the centering reference.
It is not "the effect of
having a partner". The partner slope applies to people with a partner. The
actor slope is shared.

### Residual variances: separate or shared

The models in this section differ in whether singletons and dyad members have
equal **residual variances**: the variation left after accounting for the
fixed effects. Separate random-effects terms are independent. Matrix diagonals
are variances, and their square roots are SDs. We omit the negligible residual
variance retained by `glmmTMB` from the matrices.

#### Exchangeable dyads: separate singleton variance

The model above pools the two members' variances but estimates a separate
singleton variance. Let $a$ and $b$ be the **variances**, not SDs, of the
shared and contrast terms:

$$
R_{\mathrm{dyad}} =
\begin{pmatrix}a+b & a-b \\ a-b & a+b\end{pmatrix},
\qquad R_{\mathrm{singleton}} = [\tau^2].
$$

Dyad-member SDs are $\sqrt{a+b}$ and their correlation is $(a-b)/(a+b)$. The
singleton SD is $\tau$. The partner covariance can be positive or negative.
Use this when singletons may be more or less variable than dyad members.

#### Exchangeable dyads: pool all member variances

To share the variance, add the singleton indicator to **both** exchangeable
terms, so both variance components contribute to everyone:

```{r dyad-solo-exchangeable-pooled, eval = has_glmmTMB}
mm_pooled <- glmmTMB::glmmTMB(
  closeness ~ 1 + .partner_exists +
    .provided_support_gmc_actor + .provided_support_gmc_partner +
    (0 + I(.is_exchangeable + .is_singleton_male) | coupleID) +
    (0 + I(.member_contrast_arbitrary + .is_singleton_male) | coupleID),
  family = gaussian(), dispformula = ~ 0,
  data = mm_data
)

recover_exchangeable_covariance(mm_pooled)
```

$$
R_{\mathrm{dyad}} =
\begin{pmatrix}a+b & a-b \\ a-b & a+b\end{pmatrix},
\qquad R_{\mathrm{singleton}} = [a+b].
$$

Everyone now has residual SD $\sqrt{a+b}$, so the recovered diagonal is also
the singleton variance. Singletons inform this pooled variance. Only complete
dyads provide paired observations to identify $a-b$. Pooling can still change
the estimated partner covariance and correlation through the shared
parameters.

**Do not add the singleton indicator to the shared term only.** This gives
singletons variance $a$, while dyad members retain $a+b$, which is a
different constraint.

#### Distinguishable dyads: separate singleton variances

For female-male dyads, allow singleton men and women to have different
residual variances from their counterparts in dyads. Fixed intercepts for
each group already separate rows with and without a partner here, so
`.partner_exists` is not needed:

```{r dyad-solo-distinguishable-separate, eval = has_glmmTMB}
fm_separate <- glmmTMB::glmmTMB(
  closeness ~ 0 + .is_male + .is_female +
    .is_singleton_male + .is_singleton_female +
    .provided_support_gmc_actor + .provided_support_gmc_partner +
    (0 + .is_singleton_male | coupleID) +
    (0 + .is_singleton_female | coupleID) +
    (0 + .is_male + .is_female | coupleID),
  family = gaussian(), dispformula = ~ 0,
  data = fm_data
)
```

With dyad members ordered male, female:

$$
R_{\mathrm{dyad}} =
\begin{pmatrix}v_M & c \\ c & v_F\end{pmatrix},
\qquad R_{\mathrm{singleton},M} = [\tau_M^2],
\qquad R_{\mathrm{singleton},F} = [\tau_F^2].
$$

All four variances are estimated separately. The partner correlation is
$c / \sqrt{v_M v_F}$.

#### Distinguishable dyads: pool variances by role

If singleton men can share the male dyad-member variance, and likewise for
women, add each singleton indicator to the indicator of their role:

```{r dyad-solo-distinguishable-pooled, eval = has_glmmTMB}
fm_pooled <- glmmTMB::glmmTMB(
  closeness ~ 0 + .is_male + .is_female +
    .is_singleton_male + .is_singleton_female +
    .provided_support_gmc_actor + .provided_support_gmc_partner +
    (0 + I(.is_male + .is_singleton_male) +
       I(.is_female + .is_singleton_female) | coupleID),
  family = gaussian(), dispformula = ~ 0,
  data = fm_data
)
```

$$
R_{\mathrm{dyad}} =
\begin{pmatrix}v_M & c \\ c & v_F\end{pmatrix},
\qquad R_{\mathrm{singleton},M} = [v_M],
\qquad R_{\mathrm{singleton},F} = [v_F].
$$

Each singleton helps estimate the residual variance for their role. Complete
dyads provide the paired observations needed to estimate covariance $c$.
Pooling does not create covariance between unrelated singletons.

### Testing whether to pool

Each pooled model is a restricted version of the separate one, so a
likelihood-ratio test compares them. Test with the final fixed effects:
predictors left out add unexplained variance to some groups only (here,
partner support to dyad members).

```{r dyad-solo-pooling-test, eval = has_glmmTMB}
anova(mm_pooled, mm_separate)
anova(fm_pooled, fm_separate)
```

The singletons here were simulated with the same residual variance as dyad
members, so equal variances are plausible, and neither test rejects pooling.
The next model builds on `mm_pooled`.
The same works for pooling only some variances, for example only men's. A
non-significant test does not show that the variances are equal. Use a pooled
structure when equal residual variances are plausible on substantive grounds.
With separate singleton variances, only shared fixed effects can
still link singleton and dyad estimates.

### Sharing or separating means and slopes

With one indicator per group, each group gets its own intercept. Singletons
keep their own mean, through `.partner_exists` or their own intercept
(two-part coding). In actor-only models, people whose partner did not take
part can share a mean with dyad members, e.g. `I(.is_male + .is_male_x_missing)`.
Slopes are shared by adding the indicators, also with singletons, e.g.
`I(.is_male + .is_singleton_male):x`.

In the other direction, an interaction with `.partner_exists` allows the actor
slope of singletons to differ from that of dyad members:

```{r dyad-solo-separate-actor-effects, eval = has_glmmTMB}
mm_by_status <- update(
  mm_pooled, . ~ . + .partner_exists:.provided_support_gmc_actor
)

estimates <- summary(mm_by_status)$coefficients$cond
estimates[
  c(".provided_support_gmc_actor", ".partner_exists:.provided_support_gmc_actor"),
  c("Estimate", "Std. Error")
]
```

The actor coefficient is the slope for singletons. Add the interaction
coefficient to obtain the slope for dyad members. The interaction is about
`r if (has_glmmTMB) round(abs(estimates[".partner_exists:.provided_support_gmc_actor", "z value"]), 1) else 2`
standard errors from 0, although both generating actor slopes are 0.6. Such
chance results are a reason to test only differences you planned to test.

## Scenarios

With `partner_exists`, `prepare_dyad_data()` prepares these cases directly.
Two-part predictors also work with repeated observations
[@henryTwoPartFatherChild2022]. Model within-person changes separately from
between-person differences [@curranDisaggregationWithinBetween2011].

Each scenario below shows the situation, the preparation, and a model. In
the timelines, ● is a row with data, ○ is an occasion without a row while
the partner exists, an empty cell has neither a row nor a partner, and the
last line is `partner_exists`. Random effects are kept minimal here (see the
sections above for exchangeable structures).

### Singletons and partners who did not take part (cross-sectional)

| Person | Partner's row | `partner_exists` |
|---|---|---|
| Couple member | ● | `TRUE` |
| Singleton (no partner) | | `FALSE` |
| Person whose partner did not answer | ○ | `TRUE` |

```{r scenario-cross-sectional, eval = FALSE}
prepared <- prepare_dyad_data(
  survey, dyad = coupleID, member = personID, predictors = support,
  add_apim_gmc_predictors = TRUE, partner_exists = partnered, seed = 1
)

glmmTMB::glmmTMB(
  closeness ~ 0 + .is_exchangeable + .is_singleton +
    .support_gmc_actor + .support_gmc_partner + (1 | coupleID),
  data = prepared
)
```

Singletons get 0 as partner predictor. The fixed intercepts (`.is_exchangeable`,
`.is_singleton`) already represent partner status here, so `.partner_exists`
is not needed, although the printed suggestion includes it. `(1 | coupleID)`
pools all residual variances (see above). People whose
partner did not answer have a missing partner predictor and drop out of this
model.

### Partners who never took part (diary)

| day | 1 | 2 | 3 | 4 |
|---|---|---|---|---|
| Anna | ● | ● | ● | ● |
| Ben (never took part) | ○ | ○ | ○ | ○ |
| `partner_exists` | `TRUE` | `TRUE` | `TRUE` | `TRUE` |

```{r scenario-partner-not-in-data, eval = FALSE}
prepared <- prepare_dyad_data(
  diary, dyad = coupleID, member = personID, time = day,
  predictors = support, partner_exists = TRUE, seed = 1
)
```

Nothing is set to 0. Anna is labeled `missing_partner` and keeps her rows for
actor-only models. Models with partner predictors leave her out.

### A partner dies (panel)

| wave | 1 | 2 | 3 | 4 | 5 |
|---|---|---|---|---|---|
| Anna | ● | ● | ● | ● | ● |
| Ben | ● | ● | ● | | |
| `partner_exists` | `TRUE` | `TRUE` | `TRUE` | `FALSE` | `FALSE` |

```{r scenario-death, eval = FALSE}
prepared <- prepare_dyad_data(
  panel, dyad = coupleID, member = personID, time = wave,
  predictors = health, lag1_predictors = health,
  partner_exists = partnered, seed = 1
)

glmmTMB::glmmTMB(
  wellbeing ~ 1 + .partner_exists_cwp + .partner_exists_cbp +
    .partner_exists_lag1 +
    .health_cwp_actor + .health_cbp_actor +
    .health_cwp_partner + .health_cwp_partner_lag1 +
    .health_cbp_partner_when_exists + .health_cbp_partner_after_exists +
    (1 | coupleID / personID),
  data = prepared
)
```

The fixed effects follow the terms that `prepare_dyad_data()` suggests for
these data. From wave 4, Ben's momentary partner values in Anna's rows are 0.
His usual level moves to `.health_cbp_partner_after_exists`, so its
association with Anna's well-being after his death is estimated separately.
At wave 4, the lagged value still holds his wave-3 value, paired with
`.partner_exists_lag1`.

### A couple separates, both keep taking part

| wave | 1 | 2 | 3 | 4 | 5 |
|---|---|---|---|---|---|
| Anna | ● | ● | ● | ● | ● |
| Ben | ● | ● | ● | ● | ● |
| `partner_exists` | `TRUE` | `TRUE` | `TRUE` | `FALSE` | `FALSE` |

The preparation and model are the same as for a death. Both keep their rows.
After the separation, the contemporaneous momentary partner predictors are 0.
The former partner's usual level stays available in `_after_exists` and
includes their occasions after the separation. At the first occasion after
the separation, the lagged values still hold the former partner's values.

### A relationship forms during the study

| wave | 1 | 2 | 3 | 4 | 5 |
|---|---|---|---|---|---|
| Anna | ● | ● | ● | ● | ● |
| Ben (joins with the relationship) | | | ● | ● | ● |
| `partner_exists` | `FALSE` | `FALSE` | `TRUE` | `TRUE` | `TRUE` |

```{r scenario-formation, eval = FALSE}
prepared <- prepare_dyad_data(
  panel, dyad = coupleID, member = personID, time = wave,
  predictors = health, partner_exists = partnered, seed = 1
)

glmmTMB::glmmTMB(
  wellbeing ~ 1 + .partner_exists_cwp + .partner_exists_cbp +
    .health_cwp_actor + .health_cbp_actor + .health_cwp_partner +
    .health_cbp_partner_when_exists + .health_cbp_partner_before_exists +
    (1 | coupleID / personID),
  data = prepared
)
```

Before the relationship, Ben's usual level is in
`.health_cbp_partner_before_exists`. It can relate to Anna's outcome through
selection (who ends up together). If the data also contain losses, add
`.partner_before_exists` and `.health_cbp_partner_after_exists`, so that
occasions before a relationship and after a loss get their own mean and
slope.

### An on-off relationship

| wave | 1 | 2 | 3 | 4 | 5 |
|---|---|---|---|---|---|
| Anna | ● | ● | ● | ● | ● |
| Ben | ● | ● | ● | ● | ● |
| `partner_exists` | `TRUE` | `FALSE` | `FALSE` | `TRUE` | `TRUE` |

`prepare_dyad_data()` treats Ben as the same partner returning and says so.
The gap counts as "after a partnership" (`_after_exists`). A different, new
partner is not supported: end the person's data before the new partnership.

### Advanced: actor variables that only exist with a partner

Some of a person's own variables also only exist when a partner exists, for
example ratings of the partner or of the relationship. They are usually
missing where no partner exists, so rows without a partner drop out of models
that use them. `prepare_dyad_data()` does not yet apply two-part coding to
such actor variables ([#86](https://github.com/Pascal-Kueng/dyadMLM/issues/86)), but you can do it yourself after preparing the
data.

Here, `satisfaction` is a daily relationship rating in a diary study,
prepared with `partner_exists`. The person means behind its `cwp` and `cbp`
columns come only from days with a partner, because the rating is missing on
the other days. So only the zeros are missing:

- The momentary part (`cwp`) is set to 0 where no partner exists.
- The usual level (`cbp`) is split by partner status, as dyadMLM does for
  partner usual levels when the status changes within a dyad. Each column
  holds the usual level in its status and 0 elsewhere.

`dplyr::if_else()` is needed, because multiplying by `.partner_exists` keeps
missing values (`0 * NA` is `NA`).

```{r actor-two-part-recipe, eval = FALSE}
prepared <- prepare_dyad_data(
  diary, dyad = coupleID, member = personID, time = day,
  predictors = c(support, satisfaction), partner_exists = partnered, seed = 1
)

prepared <- prepared |>
  mutate(
    # First day on which the couple had a partner (Inf if never).
    first_partnered = if (any(.partner_exists == 1)) {
      min(day[.partner_exists == 1])
    } else {
      Inf
    },
    .by = coupleID
  ) |>
  mutate(
    # Momentary part: 0 where no partner exists.
    .satisfaction_cwp_actor = if_else(
      .partner_exists == 0, 0, .satisfaction_cwp_actor
    ),
    # Usual level, split by partner status.
    .satisfaction_cbp_actor_when_exists = if_else(
      .partner_exists == 1, .satisfaction_cbp_actor, 0
    ),
    .satisfaction_cbp_actor_before_exists = if_else(
      .partner_exists == 0 & day < first_partnered & is.finite(first_partnered),
      .satisfaction_cbp_actor, 0
    ),
    .satisfaction_cbp_actor_after_exists = if_else(
      .partner_exists == 0 & day > first_partnered,
      .satisfaction_cbp_actor, 0
    )
  )
```

Include these columns in place of `.satisfaction_cwp_actor` and
`.satisfaction_cbp_actor`, together with the status terms that
`prepare_dyad_data()` suggests. Drop a split column that is 0 everywhere (for
example, `_before_exists` when no relationship begins during the study). In
cross-sectional data, or when the status never changes within a dyad, only
the zeros are needed: set the actor columns to 0 where `.partner_exists` is 0.
If both
occasions before a relationship and after a loss occur, also include
`.partner_before_exists`.

With lags (here `lag1_predictors = satisfaction` in `prepare_dyad_data()`),
set lagged actor columns to 0 where no partner existed at the previous
occasion. Use `.partner_exists_lag1` if it was created. Otherwise, use the
couple's status at exactly the previous occasion (from either member), and
keep missing values where neither member was observed then:

```{r actor-two-part-lags, eval = FALSE}
prepared <- prepared |>
  mutate(
    # The couple's status at the previous day, NA if nobody was observed then.
    partnered_lag1 = partnered[match(day - 1, day)],
    .by = coupleID
  ) |>
  mutate(
    .satisfaction_cwp_actor_lag1 = if_else(
      partnered_lag1 %in% FALSE, 0, .satisfaction_cwp_actor_lag1
    )
  )
```

## References

::: {#refs}
:::
