---
title: "Undirected Dyadic Score Model (DSM)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Undirected Dyadic Score Model (DSM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
```

This vignette focuses on undirected Dyadic Score Model (DSM) preparation. For
the main data requirements and validation workflow, start with the
[Getting Started vignette](getting-started.html). For actor-partner
predictor preparation, see the
[Actor-Partner Interdependence Model vignette](apim.html). For APIMs that
combine distinguishable and exchangeable dyad compositions, see the
[Mixed-Composition APIM vignette](mixed-apim.html). For dyad-mean and
within-dyad-deviation predictors, see the
[Dyad-Individual Model vignette](dim.html).

> This vignette is under construction. Please check back soon!

For the main data-preparation workflow, return to the
[Getting Started vignette](getting-started.html). For actor-partner models, see
the [Actor-Partner Interdependence Model vignette](apim.html) and the advanced
[Mixed-Composition APIM vignette](mixed-apim.html). For the DIM
parameterization, see the [Dyad-Individual Model vignette](dim.html).

For an in-depth tutorial covering data preparation, model fitting, diagnostics,
and assumption checks, see
[Distinguishable and Exchangeable Dyads: Bayesian Multilevel Modelling](https://pascal-kueng.github.io/05DyadicDataAnalysis/).
It uses `interdep` for cross-sectional and intensive longitudinal APIM and DIM
workflows, with models fitted primarily using `brms`
([source](https://github.com/Pascal-Kueng/05DyadicDataAnalysis),
[DOI](https://doi.org/10.5281/zenodo.17400655)).
