---
title: "Undirected Dyadic Score Model (DSM)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Undirected Dyadic Score Model (DSM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interdep)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

This vignette focuses on undirected Dyadic Score Model (DSM) preparation. For
the main data requirements and validation workflow, start with the
[Getting Started vignette](getting-started.html). For actor-partner
predictor preparation, see the
[Actor-Partner Interdependence Model vignette](apim.html). For APIMs that
combine distinguishable and exchangeable dyad compositions, see the
[Mixed-Composition APIM vignette](mixed-apim.html). For dyad-mean and
within-dyad-deviation predictors, see the
[Dyad-Individual Model vignette](dim.html).

> This vignette is under construction and for now only contains a few preliminary example models. Please check back soon!


The model is an expantion of the DIM, but we now also decompose the outcome. 
In preparaten, we therefore need to specify which variable is the outcome. 

```{r prepare-cross-exchangeable}
cross_exchangeable_data <- prepare_interdep_data(
  example_dyadic_crosssectional,
  group = coupleID,
  member = personID,
  predictors = communication,
  outcomes = satisfaction,
  # Create both APIM and DSM columns for comparison.
  model_type = c("apim", "undirected_dsm"),
  seed = 123
)

# Print the first two dyads.
print(cross_exchangeable_data, n = 4)
```


## The separate models approach

Since `glmmTMB` is a univariate framwork, we can't model both mean and difference
decomposed parts of the outcome in the same model. An approximation that does
not estimate the residual correlation between the mean and difference is to 
estimate two seaparate models. 

The dyad-mean model is a DIM (see DIM Vignette for interpretation and explanations),
with the outcome being the dyad-mean of the outcome. (Note: could and should we control
for the dyad difference?? What does Iida say?)

```{r fit-cross-dsm, eval=has_glmmTMB}

dim_dyad_mean <- glmmTMB::glmmTMB(
  .i_satisfaction_dyad_mean ~

    # Pooled fixed intercept
    1 +

    # Between-couple effect
    .i_communication_dyad_mean_gmc +

    # Within-couple effect
    .i_communication_within_dyad_deviation +

    # Residual Gaussian variance-covariance
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID)
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

summary(dim_dyad_mean)
```

```{r fit-cross-dsm, eval=has_glmmTMB}

dim_within_dyad_deviation <- glmmTMB::glmmTMB(
  .i_satisfaction_within_dyad_deviation ~

    # Pooled fixed intercept
    1 +

    # Between-couple effect
    .i_communication_dyad_mean_gmc +

    # Within-couple effect
    .i_communication_within_dyad_deviation +

    # Residual Gaussian variance-covariance
    (1 | coupleID) +
    (0 + .i_diff_assumed_exchangeable_arbitrary | coupleID)
  , dispformula = ~ 0
  , family = gaussian()
  , data = cross_exchangeable_data
)

summary(dim_within_dyad_deviation)
```


For the main data-preparation workflow, return to the
[Getting Started vignette](getting-started.html). For actor-partner models, see
the [Actor-Partner Interdependence Model vignette](apim.html) and the advanced
[Mixed-Composition APIM vignette](mixed-apim.html). For the DIM
parameterization, see the [Dyad-Individual Model vignette](dim.html).

For an in-depth tutorial covering data preparation, model fitting, diagnostics,
and assumption checks, see
[Distinguishable and Exchangeable Dyads: Bayesian Multilevel Modelling](https://pascal-kueng.github.io/05DyadicDataAnalysis/).
It uses `interdep` for cross-sectional and intensive longitudinal APIM and DIM
workflows, with models fitted primarily using `brms`
([source](https://github.com/Pascal-Kueng/05DyadicDataAnalysis),
[DOI](https://doi.org/10.5281/zenodo.17400655)).
