---
title: "Dyadic Data Analysis with R"
subtitle: "Intensive Longitudinal Workshop Exercises"
author:
  - "Pascal Küng"
  - "Turu Stadler"
format:
  html:
    embed-resources: true
    toc: true
    toc-title: "ILD Exercises"
editor: source
---

# Exercises

These exercises extend the cross-sectional APIM workflow to intensive
longitudinal dyadic data.

Download the complete workshop materials from the
[workshop page](https://pascal-kueng.github.io/dyadMLM/workshop/). Package
documentation and vignettes are available on the
[`dyadMLM` homepage](https://pascal-kueng.github.io/dyadMLM/).


# Getting started

Keep this file, `00_setup.R`, `dyadic-data.rds`, and `hints.R` in the same
folder.

If you have not already run the workshop setup, run the next chunk once, in a
fresh R session. It installs or updates the workshop packages and any
dependencies that need updating, but does not update unrelated packages. This
may take several minutes.

```{r eval=FALSE}
source("00_setup.R")
```

Load the required packages and optional hints:

```{r}
library(glmmTMB)
library(dyadMLM)
library(wbCorr)
library(easystats)
library(DHARMa)
library(dplyr)
library(ggplot2)

source("hints.R")
```

Optional hints appear as commented `hint()` calls. Uncomment and run them one
at a time.

Load and inspect the intensive longitudinal dataset:

```{r}
raw_ild_data <- readRDS("dyadic-data.rds")

head(raw_ild_data)
names(raw_ild_data)
```


# The ILD workflow

The same six-step workflow still applies:

1. Define dyads, distinguishability, and research questions
2. Prepare the data and validate the dyadic and temporal structure
3. Describe measures, trajectories, and within- and between-person relationships
4. Specify and estimate the model
5. Verify model adequacy, temporal dependence, and robustness
6. Visualize, interpret, and report results

ILD additionally requires decisions about matched occasions, time gaps,
within- and between-person variation, trends, lagging, and residual temporal
dependence.


# 1. Define the research question and temporal target

Choose one outcome and one predictor. State the actor and partner questions.

> [Insert your research question]

Also decide whether your primary question is:

- **Concurrent:** are partners' predictor and outcome values associated on the
  same occasion?
- **Lagged:** do prior values predict either partner's later outcome?

> [Insert your temporal target and reasoning]

```{r}
# hint("ild_research_question", "concept")
```


# 2. Prepare and validate the data

Use `dyadMLM::prepare_dyad_data()` and save the result as `my_ild_data`.

```{r}
?prepare_dyad_data()

# hint("ild_prepare_data", "concept")
# hint("ild_prepare_data", "structure")
```

Your preparation call should identify the dyad, member, role, time variable,
and focal predictor. Use a two-level temporal decomposition for a concurrent
ILD APIM. If your research question is lagged, also create the required lagged
predictors without bridging missing occasions.

```{r}

```


## Validate the dyadic and temporal structure

Check:

- exactly two members per retained dyad
- stable member roles
- at most one row per dyad, member, and occasion
- the scheduled and observed time range
- missing observations, compliance, and matched partner occasions
- actual time gaps

```{r}

```

Summarize what you found:

> [Insert your data-structure and missingness summary]


# 3. Describe trajectories and relationships

```{r}
# hint("ild_describe_data", "concept")
```


## Plot paired trajectories

Plot the outcome and focal predictor over time. Show both members of each dyad
and inspect a manageable selection of dyads rather than drawing every
trajectory in one panel.

```{r}

```


## Separate within- and between-person information

At minimum:

- describe the focal variables by role
- calculate ICCs
- calculate within- and between-person correlations
- note whether multilevel reliability would be needed for multi-item measures
  (item-level data are not included here)

```{r}

```

Describe the main patterns:

> [Insert your descriptive interpretation]


# 4. Specify and estimate the model

Begin with the scientific mean structure. Include plausible time trends before
adding a residual temporal process.

```{r eval=FALSE}
# hint("ild_model", "concept")
# hint("ild_model", "structure")

my_ild_model <- glmmTMB(
  formula = YOUR_ILD_MODEL_FORMULA,
  data = my_ild_data,
  family = gaussian()
)

summary(my_ild_model)
```


## Choose the temporal specification

Decide whether the remaining temporal dependence is best represented by:

- no residual temporal process
- discrete-time AR(1) for equally spaced scheduled occasions
- continuous-time decay for unequal time gaps
- lagged actor and partner outcomes when carryover is the scientific target

Document the time scale, lag, trend, covariance structure, and any constraints.

```{r}
# hint("ild_time_structure", "concept")
```

If required, update your model with the selected temporal structure:

```{r eval=FALSE}

```


# 5. Verify adequacy and robustness

Check estimation first: convergence, the Hessian, and whether all parameter
estimates are plausible.


## Residual diagnostics

Use the ILD DHARMa block from the applied tutorial. Inspect overall calibration,
patterns against fitted values and time, and remaining temporal or
cross-partner dependence.

```{r eval=FALSE}
# hint("ild_diagnostics", "concept")
```


## Robustness

Consider whether conclusions depend on:

- the chosen time trend
- the temporal covariance structure
- treatment of incomplete or unmatched occasions
- substantively questionable dyads or measurement periods

Do not select a specification solely because it changes a significance label.

> [Insert your robustness conclusion]


# 6. Visualize, interpret, and report

Report the actor and partner effects with uncertainty. Clearly distinguish
within-person from between-person effects and describe the temporal and
cross-partner covariance structure.

```{r}
# hint("ild_report_results", "concept")
# hint("ild_report_results", "structure")
```

Plot the focal fitted effects or model-implied trajectories:

```{r}

```

Interpret the results in practical terms:

> [Insert your interpretation]


# Optional: exchangeability as a model constraint

If scientifically defensible, fit the exchangeability-constrained version as a
change to the completed distinguishable ILD workflow. Compare the models and
document every imposed constraint.

```{r eval=FALSE}

```

> [Insert your distinguishability conclusion]
