---
title: "Prototype simulation study: Gaussian ILD partner-dependence checks"
output:
  html_document:
    toc: true
    number_sections: true
    code_folding: hide
params:
  profile: "smoke"
  outer_repetitions: 0
  inner_simulations: 0
  n_dyads: 0
  n_times: 0
---

This development study exercises the Gaussian intensive-longitudinal (ILD)
partner-dependence prototype. It repeats the fitted workflow for stable,
concurrent, and own-member lag dependence, then uses a known-parameter control
to exercise the joint finite-series consequences of shared/difference process
persistence. Final deterministic sections check exact gaps and the distinction
between equal-dyad and equal-edge weighting.

The predictive reference is plug-in and conditional on the fitted-row design.
Observed positions and middle intervals are descriptive; they are not
calibrated tests, p-values, or estimates of power or type-I error.

```{r setup, include=FALSE}
package_root <- normalizePath(file.path("..", "..", ".."))
devtools::load_all(package_root, quiet = TRUE)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)

profiles <- list(
  smoke = list(
    outer_repetitions = 3L,
    inner_simulations = 49L,
    n_dyads = 40L,
    n_times = 12L
  ),
  pilot = list(
    outer_repetitions = 20L,
    inner_simulations = 199L,
    n_dyads = 60L,
    n_times = 16L
  )
)

profile_name <- as.character(params$profile)
if (!profile_name %in% names(profiles)) {
  stop("profile must be one of: smoke or pilot.", call. = FALSE)
}
settings <- profiles[[profile_name]]

apply_positive_override <- function(value, current, name) {
  if (
    length(value) != 1L ||
      !is.numeric(value) ||
      is.na(value) ||
      !is.finite(value) ||
      value < 0 ||
      value != floor(value) ||
      value > .Machine$integer.max
  ) {
    stop(name, " must be zero or a positive whole number.", call. = FALSE)
  }
  value <- as.integer(value)
  if (value == 0L) current else value
}

n_outer <- apply_positive_override(
  params$outer_repetitions,
  settings$outer_repetitions,
  "outer_repetitions"
)
n_inner <- apply_positive_override(
  params$inner_simulations,
  settings$inner_simulations,
  "inner_simulations"
)
n_dyads <- apply_positive_override(
  params$n_dyads,
  settings$n_dyads,
  "n_dyads"
)
n_times <- apply_positive_override(
  params$n_times,
  settings$n_times,
  "n_times"
)

stopifnot(
  n_outer >= 1L,
  n_inner >= 20L,
  n_dyads >= 12L,
  n_times >= 6L
)

base_seed <- 20260827L
demonstration_lags <- 1:min(3L, n_times - 1L)
```

# Study profiles

The selected profile is `r profile_name`: `r n_outer` outer repetitions,
`r n_inner` complete response simulations per fitted model, `r n_dyads` dyads,
and `r n_times` scheduled occasions. The intended profiles are:

```{r profile-table, echo=FALSE}
profile_table <- do.call(
  rbind,
  lapply(names(profiles), function(profile) {
    data.frame(profile = profile, profiles[[profile]], row.names = NULL)
  })
)
knitr::kable(profile_table)
```

The smoke profile checks execution only. The pilot profile checks runtime,
convergence, and qualitative scenario behavior against exact finite-series
targets. Neither profile estimates power or error rates. A larger calibration
study should be designed only if a formal calibration estimand is first
specified.

# Distinguishable core population

The fitted study uses three targeted distinguishable populations rather than
one large omnibus covariance model. This keeps the restrictions transparent
and avoids making successful estimation of ten covariance parameters a
prerequisite for validating the diagnostic calculations.

The stable population generates

\[
y_{dmt}=\mu_{dmt}+b_{dm}+\epsilon_{dmt},
\]

where \(b_d\) is a stable bivariate dyad effect with

\[
B=\begin{pmatrix}.64&.48\\.48&1.00\end{pmatrix},
\]

and \(\epsilon_{dmt}\) is independent Gaussian noise with variance .25.
The concurrent population instead generates

\[
y_{dmt}=\mu_{dmt}+e_{dmt},
\]

where the bivariate same-occasion shock has

\[
W=\begin{pmatrix}.49&.294\\.294&.49\end{pmatrix}.
\]

The separate temporal population generates

\[
y_{dmt}=\mu_{dmt}+a_{dmt},
\]

where each \(a_{dm,t}\) is a role-specific stationary AR(1) process with
variance .64 and \((\phi_1,\phi_2)=(.80,.20)\). The fixed mean in all three
populations contains role-specific intercepts and trends plus a time-varying
predictor. It is held fixed across outer repetitions, matching the
diagnostic's conditional-on-design target.

```{r distinguishable-dgp}
draw_multivariate_normal <- function(n, covariance) {
  standard_normal <- matrix(
    stats::rnorm(n * ncol(covariance)),
    nrow = n,
    ncol = ncol(covariance)
  )
  standard_normal %*% chol(covariance)
}

draw_stationary_ar <- function(n_series, n_times, variance, phi) {
  values <- matrix(NA_real_, nrow = n_series, ncol = n_times)
  values[, 1L] <- stats::rnorm(n_series, sd = sqrt(variance))
  innovation_sd <- sqrt(variance * (1 - phi^2))
  if (n_times >= 2L) {
    for (time_index in 2:n_times) {
      values[, time_index] <-
        phi * values[, time_index - 1L] +
        stats::rnorm(n_series, sd = innovation_sd)
    }
  }
  values
}

make_distinguishable_design <- function(n_dyads, n_times, seed) {
  design <- expand.grid(
    dyad_number = seq_len(n_dyads),
    time_number = seq_len(n_times),
    role_number = 1:2,
    KEEP.OUT.ATTRS = FALSE
  )
  design <- design[
    order(design$dyad_number, design$time_number, design$role_number),
  ]
  rownames(design) <- NULL

  design$dyad <- factor(design$dyad_number)
  design$member <- factor(
    paste(design$dyad_number, design$role_number, sep = "_")
  )
  design$role <- factor(
    paste0("member_", design$role_number),
    levels = c("member_1", "member_2")
  )
  design$role_1 <- as.numeric(design$role_number == 1L)
  design$role_2 <- as.numeric(design$role_number == 2L)
  design$member_sign <- ifelse(design$role_number == 1L, -1, 1)
  design$time_f <- factor(
    design$time_number,
    levels = seq_len(n_times)
  )
  design$time_c <-
    (design$time_number - (n_times + 1) / 2) / (n_times - 1)
  design$member_series <- interaction(
    design$dyad,
    design$member,
    drop = TRUE
  )

  set.seed(seed)
  shared_x <- stats::rnorm(n_dyads * n_times, sd = 0.5)
  shared_index <-
    (design$dyad_number - 1L) * n_times + design$time_number
  design$x <- shared_x[shared_index] + stats::rnorm(nrow(design), sd = 0.5)
  design$expected_mean <-
    -0.20 * design$role_1 +
    0.20 * design$role_2 +
    0.30 * design$role_1 * design$time_c -
    0.15 * design$role_2 * design$time_c +
    0.40 * design$x
  design
}

draw_stable_distinguishable_data <- function(design, seed) {
  set.seed(seed)
  n_dyads <- nlevels(design$dyad)

  stable_covariance <- matrix(
    c(0.64, 0.48, 0.48, 1.00),
    nrow = 2L,
    byrow = TRUE
  )
  stable_effects <- draw_multivariate_normal(
    n_dyads,
    stable_covariance
  )
  dyad_index <- design$dyad_number
  role_index <- design$role_number
  stable_value <- stable_effects[cbind(dyad_index, role_index)]

  data <- design
  data$y <-
    data$expected_mean + stable_value + stats::rnorm(nrow(data), sd = 0.5)
  data
}

draw_concurrent_distinguishable_data <- function(design, seed) {
  set.seed(seed)
  n_dyads <- nlevels(design$dyad)
  n_times <- nlevels(design$time_f)
  concurrent_covariance <- matrix(
    c(0.49, 0.294, 0.294, 0.49),
    nrow = 2L,
    byrow = TRUE
  )
  concurrent_effects <- draw_multivariate_normal(
    n_dyads * n_times,
    concurrent_covariance
  )

  dyad_index <- design$dyad_number
  time_index <- design$time_number
  role_index <- design$role_number
  occasion_index <- (dyad_index - 1L) * n_times + time_index
  concurrent_value <- concurrent_effects[cbind(occasion_index, role_index)]

  data <- design
  data$y <- data$expected_mean + concurrent_value
  data
}

draw_temporal_distinguishable_data <- function(design, seed) {
  set.seed(seed)
  n_dyads <- nlevels(design$dyad)
  n_times <- nlevels(design$time_f)
  ar_role_1 <- draw_stationary_ar(n_dyads, n_times, 0.64, 0.80)
  ar_role_2 <- draw_stationary_ar(n_dyads, n_times, 0.64, 0.20)
  dyad_index <- design$dyad_number
  time_index <- design$time_number
  role_index <- design$role_number
  ar_value <- ifelse(
    role_index == 1L,
    ar_role_1[cbind(dyad_index, time_index)],
    ar_role_2[cbind(dyad_index, time_index)]
  )

  data <- design
  data$y <- data$expected_mean + ar_value
  data
}

distinguishable_design <- make_distinguishable_design(
  n_dyads = n_dyads,
  n_times = n_times,
  seed = base_seed + 1000L
)
```

# Fitted structures and targeted restrictions

The stable population is fitted with or without its stable role covariance.
The concurrent population is fitted with or without its same-occasion partner
covariance. The temporal population is fitted with either separate role-
specific AR processes or one pooled member-series AR process. In each
comparison, the correct and restricted model use the same generated response
data.

```{r fitted-model-functions}
fit_distinguishable_model <- function(data, model_name) {
  fixed_formula <-
    y ~ 0 + role_1 + role_2 + role_1:time_c + role_2:time_c + x

  formula <- switch(
    model_name,
    stable_correct = stats::update.formula(
      fixed_formula,
      . ~ . +
        us(0 + role_1 + role_2 | dyad)
    ),
    stable_omitted = stats::update.formula(
      fixed_formula,
      . ~ .
    ),
    concurrent_correct = stats::update.formula(
      fixed_formula,
      . ~ . +
        (1 | dyad:time_f) +
        (0 + member_sign | dyad:time_f)
    ),
    concurrent_independent = stats::update.formula(
      fixed_formula,
      . ~ .
    ),
    ar_correct = stats::update.formula(
      fixed_formula,
      . ~ . +
        ar1(0 + role_1:time_f | dyad) +
        ar1(0 + role_2:time_f | dyad)
    ),
    pooled_ar = stats::update.formula(
      fixed_formula,
      . ~ . +
        ar1(0 + time_f | member_series)
    ),
    stop("Unknown fitted model: ", model_name, call. = FALSE)
  )
  dispersion_formula <- if (model_name %in% c(
    "stable_correct",
    "stable_omitted",
    "concurrent_independent"
  )) {
    ~1
  } else {
    ~0
  }
  fit_control <- glmmTMB::glmmTMBControl(profile = FALSE)

  glmmTMB::glmmTMB(
    formula,
    dispformula = dispersion_formula,
    family = stats::gaussian(),
    control = fit_control,
    data = data
  )
}

extract_ild_results <- function(
  check,
  scenario,
  outer_rep,
  method,
  n_dyads,
  n_times
) {
  check$statistics_table |>
    dplyr::filter(.data$parameterization %in% c("member", "profile")) |>
    dplyr::mutate(
      outside_95 = dplyr::case_when(
        !is.finite(.data$observed_value) |
          !is.finite(.data$replicated_lower) |
          !is.finite(.data$replicated_upper) ~ NA,
        .data$observed_value < .data$replicated_lower |
          .data$observed_value > .data$replicated_upper ~ TRUE,
        TRUE ~ FALSE
      ),
      direction = dplyr::case_when(
        !is.finite(.data$observed_value) |
          !is.finite(.data$replicated_lower) |
          !is.finite(.data$replicated_upper) ~ "undefined",
        .data$observed_value < .data$replicated_lower ~ "below",
        .data$observed_value > .data$replicated_upper ~ "above",
        TRUE ~ "inside"
      ),
      scenario = scenario,
      outer_rep = outer_rep,
      method = method,
      design_n_dyads = n_dyads,
      design_n_times = n_times
    ) |>
    dplyr::select(
      .data$scenario,
      .data$outer_rep,
      .data$method,
      .data$design_n_dyads,
      .data$design_n_times,
      .data$statistic_id,
      .data$component,
      .data$curve,
      .data$lag,
      .data$statistic,
      .data$label,
      .data$observed_value,
      .data$replicated_median,
      .data$replicated_lower,
      .data$replicated_upper,
      .data$observed_quantile,
      .data$n_defined_simulations,
      .data$n_dyads,
      .data$n_edges,
      .data$outside_95,
      .data$direction
    )
}

run_fitted_check <- function(
  data,
  scenario,
  outer_rep,
  nsim,
  seed
) {
  fit_warnings <- character()
  fit_start <- proc.time()[["elapsed"]]
  fit <- tryCatch(
    withCallingHandlers(
      fit_distinguishable_model(data, scenario),
      warning = function(warning) {
        fit_warnings <<- c(fit_warnings, conditionMessage(warning))
        invokeRestart("muffleWarning")
      }
    ),
    error = function(error) error
  )
  fit_seconds <- proc.time()[["elapsed"]] - fit_start

  status <- tibble::tibble(
    scenario = scenario,
    outer_rep = outer_rep,
    status = NA_character_,
    convergence = NA_integer_,
    pd_hess = NA,
    fit_seconds = fit_seconds,
    check_seconds = NA_real_,
    warnings = paste(unique(fit_warnings), collapse = " | "),
    error = NA_character_
  )

  if (inherits(fit, "error")) {
    status$status <- "fit_error"
    status$error <- conditionMessage(fit)
    return(list(status = status, statistics = NULL))
  }

  status$convergence <- fit$fit$convergence
  status$pd_hess <- fit$sdr$pdHess
  if (!isTRUE(status$convergence == 0L) || !isTRUE(status$pd_hess)) {
    status$status <- "fit_problem"
    return(list(status = status, statistics = NULL))
  }

  check_warnings <- character()
  check_start <- proc.time()[["elapsed"]]
  check <- tryCatch(
    withCallingHandlers({
      simulations <- simulate_dyad_responses(
        fit,
        nsim = nsim,
        seed = seed
      )
      check_partner_dependence(
        simulations,
        dyad = data$dyad,
        role = data$role,
        plot = FALSE,
        response = "model-centred",
        member = data$member,
        time = data$time_f,
        lags = 1L,
        weighting = "dyad"
      )
    }, warning = function(warning) {
      check_warnings <<- c(check_warnings, conditionMessage(warning))
      invokeRestart("muffleWarning")
    }),
    error = function(error) error
  )
  status$check_seconds <- proc.time()[["elapsed"]] - check_start
  status$warnings <- paste(
    unique(c(fit_warnings, check_warnings)),
    collapse = " | "
  )

  if (inherits(check, "error")) {
    status$status <- "check_error"
    status$error <- conditionMessage(check)
    return(list(status = status, statistics = NULL))
  }

  status$status <- "success"
  list(
    status = status,
    statistics = extract_ild_results(
      check = check,
      scenario = scenario,
      outer_rep = outer_rep,
      method = "fitted plug-in",
      n_dyads = n_dyads,
      n_times = n_times
    )
  )
}
```

The prespecified expected movements are:

```{r target-table}
primary_targets <- tibble::tribble(
  ~scenario, ~component, ~curve, ~lag, ~statistic, ~expected,
  "stable_omitted", "stable", "partner", NA_integer_,
    "partner_correlation", "above",
  "concurrent_independent", "concurrent", "partner", NA_integer_,
    "partner_correlation", "above",
  "pooled_ar", "own_lag", "member_1", 1L,
    "correlation", "above",
  "pooled_ar", "own_lag", "member_2", 1L,
    "correlation", "below"
)

knitr::kable(primary_targets)
```

# Outer loop

Every outer repetition generates one response vector from each targeted
population. It fits the correct and restricted model to each corresponding
response vector. Fit errors, convergence failures, non-positive-definite
Hessians, warnings, and check errors are recorded without retries.

```{r run-fitted-study}
statuses <- list()
statistics <- list()
status_index <- 0L
statistics_index <- 0L
model_names <- c(
  "stable_correct",
  "stable_omitted",
  "concurrent_correct",
  "concurrent_independent",
  "ar_correct",
  "pooled_ar"
)

for (outer_rep in seq_len(n_outer)) {
  stable_data <- draw_stable_distinguishable_data(
    distinguishable_design,
    seed = base_seed + 100000L + outer_rep
  )
  concurrent_data <- draw_concurrent_distinguishable_data(
    distinguishable_design,
    seed = base_seed + 200000L + outer_rep
  )
  temporal_data <- draw_temporal_distinguishable_data(
    distinguishable_design,
    seed = base_seed + 300000L + outer_rep
  )

  for (model_index in seq_along(model_names)) {
    scenario <- model_names[[model_index]]
    generated_data <- if (scenario %in% c(
      "stable_correct",
      "stable_omitted"
    )) {
      stable_data
    } else if (scenario %in% c(
      "concurrent_correct",
      "concurrent_independent"
    )) {
      concurrent_data
    } else {
      temporal_data
    }
    check_seed <-
      base_seed + 10000000L + 10000L * model_index + outer_rep
    result <- run_fitted_check(
      data = generated_data,
      scenario = scenario,
      outer_rep = outer_rep,
      nsim = n_inner,
      seed = check_seed
    )

    status_index <- status_index + 1L
    statuses[[status_index]] <- result$status
    if (!is.null(result$statistics)) {
      statistics_index <- statistics_index + 1L
      statistics[[statistics_index]] <- result$statistics
    }
  }
}

run_status <- dplyr::bind_rows(statuses)
outer_results <- dplyr::bind_rows(statistics)
```

# Fitted-study results

## Completion

```{r completion-table, echo=FALSE}
completion <- run_status |>
  dplyr::group_by(.data$scenario) |>
  dplyr::summarise(
    attempted = dplyr::n(),
    successful = sum(.data$status == "success"),
    success_rate = mean(.data$status == "success"),
    warning_runs = sum(nzchar(.data$warnings)),
    fit_problems = sum(.data$status == "fit_problem"),
    fit_errors = sum(.data$status == "fit_error"),
    check_errors = sum(.data$status == "check_error"),
    median_fit_seconds = stats::median(.data$fit_seconds),
    median_check_seconds = stats::median(
      .data$check_seconds,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

knitr::kable(completion, digits = 3)
```

All summaries below are conditional on a successful fit and check. The status
table above remains part of the evidence; failed runs are not silently removed
or replaced.

```{r failure-details, echo=FALSE}
failure_details <- run_status |>
  dplyr::filter(.data$status != "success" | nzchar(.data$warnings)) |>
  dplyr::select(
    .data$scenario,
    .data$outer_rep,
    .data$status,
    .data$convergence,
    .data$pd_hess,
    .data$warnings,
    .data$error
  )

if (nrow(failure_details) > 0L) {
  knitr::kable(failure_details)
} else {
  cat("No fit/check failures or warnings were recorded.\n")
}
```

## Correct fitted model

Flexible fitted-model positions may cluster near the middle because the model
estimated the same dependence features from the observed dataset. This section
is a compatibility check, not a calibration analysis.

```{r correct-summary, echo=FALSE}
if (nrow(outer_results) > 0L) {
  correct_summary <- outer_results |>
    dplyr::filter(
      .data$scenario %in% c(
        "stable_correct",
        "concurrent_correct",
        "ar_correct"
      ),
      .data$statistic %in% c("partner_correlation", "correlation"),
      is.na(.data$lag) | .data$lag == 1L
    ) |>
    dplyr::group_by(
      .data$scenario,
      .data$component,
      .data$curve,
      .data$lag,
      .data$statistic
    ) |>
    dplyr::summarise(
      repetitions = dplyr::n(),
      median_position = stats::median(.data$observed_quantile, na.rm = TRUE),
      outside_rate = mean(.data$outside_95, na.rm = TRUE),
      minimum_defined = min(.data$n_defined_simulations),
      .groups = "drop"
    )
  knitr::kable(correct_summary, digits = 3)
} else {
  cat("No fitted checks completed successfully.\n")
}
```

## Targeted restrictions

```{r targeted-summary, echo=FALSE}
if (nrow(outer_results) > 0L) {
  targeted_results <- primary_targets |>
    dplyr::inner_join(
      outer_results,
      by = c("scenario", "component", "curve", "lag", "statistic")
    )

  targeted_summary <- targeted_results |>
    dplyr::group_by(
      .data$scenario,
      .data$component,
      .data$curve,
      .data$lag,
      .data$expected
    ) |>
    dplyr::summarise(
      repetitions = dplyr::n(),
      median_observed = stats::median(.data$observed_value, na.rm = TRUE),
      median_reference = stats::median(
        .data$replicated_median,
        na.rm = TRUE
      ),
      median_position = stats::median(
        .data$observed_quantile,
        na.rm = TRUE
      ),
      expected_direction_rate = mean(
        .data$direction == .data$expected,
        na.rm = TRUE
      ),
      .groups = "drop"
    )
  knitr::kable(targeted_summary, digits = 3)
} else {
  cat("No targeted fitted checks completed successfully.\n")
}
```

# Known-parameter shared/difference persistence contrast

The fitted study above targets member-mean, same-occasion, and role-specific
own-member persistence. This known-parameter comparison exercises the
exchangeable cross-lag calculation without requiring a directional VAR
implementation in `glmmTMB`.

Let \(c_{dt}\) be a shared dyad-mean AR process and \(q_{dt}\) a member-
difference AR process:

\[
W_{d1t}=c_{dt}-q_{dt},\qquad W_{d2t}=c_{dt}+q_{dt}.
\]

Both modes have variance .50, but \(\phi_+=.80\) and \(\phi_-=.20\). Before
member demeaning, their lag-zero partner covariance is zero, their lag-1 own-
member correlation is .50, and their lag-1 cross-member correlation is .30.
Those are not the diagnostic-scale targets: finite-series member demeaning
changes all three summaries. A restricted reference with both persistence
parameters fixed to .50 therefore provides a joint shared/difference process
contrast, not an isolated cross-lag perturbation.

The exact diagnostic-scale targets follow from applying the same centring
matrix used conceptually by the diagnostic to the finite-series covariance
matrices. This oracle is independent of the diagnostic implementation.

```{r finite-series-oracle}
ar_covariance <- function(phi, n_times, variance = 1) {
  variance * phi^abs(outer(seq_len(n_times), seq_len(n_times), `-`))
}

centre_covariance <- function(covariance) {
  n_times <- nrow(covariance)
  centre <- diag(n_times) - matrix(1 / n_times, n_times, n_times)
  centre %*% covariance %*% centre
}

edge_moments <- function(
  cross_covariance,
  source_covariance,
  target_covariance,
  lag
) {
  source <- seq_len(nrow(source_covariance) - lag)
  target <- source + lag
  c(
    covariance = mean(cross_covariance[cbind(source, target)]),
    variance_x = mean(diag(source_covariance)[source]),
    variance_y = mean(diag(target_covariance)[target]),
    n_edges = length(source)
  )
}

moment_correlation <- function(moments) {
  moments[["covariance"]] /
    sqrt(moments[["variance_x"]] * moments[["variance_y"]])
}

exchangeable_targets <- function(phi_plus, phi_minus, reference) {
  plus <- centre_covariance(ar_covariance(phi_plus, n_times, 0.50))
  minus <- centre_covariance(ar_covariance(phi_minus, n_times, 0.50))
  own_covariance <- plus + minus
  cross_covariance <- plus - minus

  targets <- list(data.frame(
    reference = reference,
    component = "concurrent",
    lag = NA_integer_,
    exact_target = moment_correlation(edge_moments(
      cross_covariance,
      own_covariance,
      own_covariance,
      lag = 0L
    ))
  ))
  for (lag in demonstration_lags) {
    targets[[length(targets) + 1L]] <- data.frame(
      reference = reference,
      component = c("own_lag", "cross_lag"),
      lag = lag,
      exact_target = c(
        moment_correlation(edge_moments(
          own_covariance,
          own_covariance,
          own_covariance,
          lag
        )),
        moment_correlation(edge_moments(
          cross_covariance,
          own_covariance,
          own_covariance,
          lag
        ))
      )
    )
  }
  dplyr::bind_rows(targets)
}

finite_cross_lag_targets <- dplyr::bind_rows(
  exchangeable_targets(0.80, 0.20, "correct"),
  exchangeable_targets(0.50, 0.50, "restricted")
)
```

The same oracle gives known-mean finite-schedule DGP benchmarks for the fitted
study. They are comparison targets, not exact expectations after estimating
the fixed-effect centre and calculating a finite-dyad sample statistic.

```{r core-target-table, echo=FALSE}
finite_ar_target <- function(phi) {
  covariance <- centre_covariance(ar_covariance(phi, n_times))
  moment_correlation(edge_moments(
    covariance,
    covariance,
    covariance,
    lag = 1L
  ))
}

core_target_table <- data.frame(
  summary = c(
    "Stable-population member means",
    "Concurrent-population within-member responses",
    "Temporal member 1, lag 1",
    "Temporal member 2, lag 1",
    "Restricted pooled temporal reference, lag 1"
  ),
  analytic_dgp_target = c(
    0.48 / sqrt(
      (0.64 + 0.25 / n_times) * (1.00 + 0.25 / n_times)
    ),
    0.60,
    finite_ar_target(0.80),
    finite_ar_target(0.20),
    finite_ar_target(0.50)
  )
)
knitr::kable(core_target_table, digits = 3)
```

```{r cross-lag-functions}
draw_exchangeable_response <- function(
  design,
  seed,
  phi_plus,
  phi_minus
) {
  set.seed(seed)
  n_dyads <- nlevels(design$dyad)
  n_times <- nlevels(design$time_f)
  stable_plus <- stats::rnorm(n_dyads, sd = sqrt(0.45))
  stable_minus <- stats::rnorm(n_dyads, sd = sqrt(0.15))
  dynamic_plus <- draw_stationary_ar(n_dyads, n_times, 0.50, phi_plus)
  dynamic_minus <- draw_stationary_ar(n_dyads, n_times, 0.50, phi_minus)

  dyad_index <- design$dyad_number
  time_index <- design$time_number
  sign <- design$member_sign
  design$expected_mean +
    stable_plus[dyad_index] +
    sign * stable_minus[dyad_index] +
    dynamic_plus[cbind(dyad_index, time_index)] +
    sign * dynamic_minus[cbind(dyad_index, time_index)]
}

make_manual_simulations <- function(
  observed,
  simulated,
  response_center,
  reference,
  seed
) {
  result <- list(
    observed_response = as.numeric(observed),
    simulated_responses = simulated,
    response_center = as.numeric(response_center),
    model_frame = data.frame(y = as.numeric(observed)),
    backend = "glmmTMB",
    family = "gaussian",
    link = "identity",
    reference = reference,
    random_effects = "known population draws",
    parameter_uncertainty = "not applicable",
    center = "known fixed expected response",
    center_target = paste0(
      "population response mean over zero-mean random effects ",
      "(Gaussian identity)"
    ),
    target = paste0(
      "unconditional known-parameter replication under the fixed design, ",
      "with all random effects newly generated"
    ),
    nsim = nrow(simulated),
    seed = as.integer(seed),
    call = match.call()
  )
  class(result) <- c("dyadMLM_response_simulations", "list")
  result
}

simulate_exchangeable_reference <- function(
  design,
  nsim,
  seed,
  phi_plus,
  phi_minus
) {
  simulated <- vapply(
    seq_len(nsim),
    function(simulation_index) {
      draw_exchangeable_response(
        design = design,
        seed = seed + simulation_index,
        phi_plus = phi_plus,
        phi_minus = phi_minus
      )
    },
    numeric(nrow(design))
  )
  t(simulated)
}
```

```{r run-cross-lag-demonstration}
cross_lag_observed <- draw_exchangeable_response(
  distinguishable_design,
  seed = base_seed + 30000000L,
  phi_plus = 0.80,
  phi_minus = 0.20
)
cross_lag_correct_matrix <- simulate_exchangeable_reference(
  distinguishable_design,
  nsim = n_inner,
  seed = base_seed + 31000000L,
  phi_plus = 0.80,
  phi_minus = 0.20
)
cross_lag_restricted_matrix <- simulate_exchangeable_reference(
  distinguishable_design,
  nsim = n_inner,
  seed = base_seed + 32000000L,
  phi_plus = 0.50,
  phi_minus = 0.50
)

cross_lag_correct_simulations <- make_manual_simulations(
  observed = cross_lag_observed,
  simulated = cross_lag_correct_matrix,
  response_center = distinguishable_design$expected_mean,
  reference = "known-parameter correct reference",
  seed = base_seed + 31000000L
)
cross_lag_restricted_simulations <- make_manual_simulations(
  observed = cross_lag_observed,
  simulated = cross_lag_restricted_matrix,
  response_center = distinguishable_design$expected_mean,
  reference = "known-parameter restricted reference",
  seed = base_seed + 32000000L
)

run_exchangeable_check <- function(simulations) {
  check_partner_dependence(
    simulations,
    dyad = distinguishable_design$dyad,
    plot = FALSE,
    response = "model-centred",
    member = distinguishable_design$member,
    time = distinguishable_design$time_f,
    lags = demonstration_lags,
    weighting = "dyad"
  )
}

cross_lag_correct_check <- run_exchangeable_check(
  cross_lag_correct_simulations
)
cross_lag_restricted_check <- run_exchangeable_check(
  cross_lag_restricted_simulations
)
```

```{r cross-lag-table, echo=FALSE}
extract_cross_lag_demo <- function(check, reference) {
  check$statistics_table |>
    dplyr::filter(
      .data$component %in% c("concurrent", "own_lag", "cross_lag"),
      .data$statistic %in% c(
        "exchangeable_partner_correlation",
        "correlation"
      ),
      is.na(.data$lag) | .data$lag <= 3L
    ) |>
    dplyr::mutate(reference = reference) |>
    dplyr::select(
      .data$reference,
      .data$component,
      .data$curve,
      .data$lag,
      .data$observed_value,
      .data$replicated_median,
      .data$replicated_lower,
      .data$replicated_upper,
      .data$observed_quantile
    )
}

cross_lag_table <- dplyr::bind_rows(
  extract_cross_lag_demo(cross_lag_correct_check, "correct"),
  extract_cross_lag_demo(cross_lag_restricted_check, "restricted")
) |>
  dplyr::left_join(
    finite_cross_lag_targets,
    by = c("reference", "component", "lag")
  )
knitr::kable(cross_lag_table, digits = 3)
```

The restricted comparison is a known-parameter diagnostic control, not a
claim that a fitted directional VAR is currently supported. Compare each
replicated median with its exact finite-series target. Because member demeaning
couples the summaries, the concurrent, own-lag, and cross-lag differences must
be interpreted jointly.

# Exact gaps

The diagnostic defines lags by factor-level distance, never by adjacency of
the remaining rows. In this fixture, member 1 lacks scheduled occasion 2 but
retains occasions 1 and 3. The pair from 1 to 3 is therefore a lag-2 edge, not
a lag-1 edge.

```{r gap-illustration}
gap_data <- expand.grid(
  dyad_number = 1:4,
  time_number = 1:4,
  member_number = 1:2,
  KEEP.OUT.ATTRS = FALSE
)
gap_data <- gap_data[
  order(
    gap_data$dyad_number,
    gap_data$time_number,
    gap_data$member_number
  ),
]
gap_data <- gap_data[
  !(gap_data$member_number == 1L & gap_data$time_number == 2L),
]
rownames(gap_data) <- NULL
gap_data$dyad <- factor(gap_data$dyad_number)
gap_data$member <- factor(
  paste(gap_data$dyad_number, gap_data$member_number, sep = "_")
)
gap_data$role <- factor(
  paste0("member_", gap_data$member_number),
  levels = c("member_1", "member_2")
)
gap_data$time_f <- factor(gap_data$time_number, levels = 1:4)
gap_data$y <-
  sin(gap_data$time_number) +
  0.2 * gap_data$member_number +
  0.1 * gap_data$dyad_number

set.seed(base_seed + 40000000L)
gap_simulated <- t(vapply(
  seq_len(20L),
  function(index) gap_data$y + stats::rnorm(nrow(gap_data), sd = 0.2),
  numeric(nrow(gap_data))
))
gap_simulations <- make_manual_simulations(
  observed = gap_data$y,
  simulated = gap_simulated,
  response_center = rep(0, nrow(gap_data)),
  reference = "deterministic gap illustration",
  seed = base_seed + 40000000L
)
gap_check <- check_partner_dependence(
  gap_simulations,
  dyad = gap_data$dyad,
  role = gap_data$role,
  plot = FALSE,
  response = "raw",
  member = gap_data$member,
  time = gap_data$time_f,
  lags = 1:2,
  weighting = "dyad"
)

gap_support <- gap_check$statistics_table |>
  dplyr::filter(.data$component %in% c("own_lag", "cross_lag")) |>
  dplyr::select(
    .data$component,
    .data$curve,
    .data$lag,
    .data$weighting,
    .data$n_dyads,
    .data$n_edges,
    .data$structural_reason
  )
knitr::kable(gap_support)
```

# Dyad versus edge weighting

For a deliberately heterogeneous schedule, half the dyads contribute two
12-occasion member series with \(\phi=.80\); the other half contribute two
5-occasion series with \(\phi=.20\). Equal-dyad weighting gives each dyad the
same total influence. Equal-edge weighting emphasizes the longer, more
persistent series. The exact covariance oracle above makes this contrast
visible without another 99--199 Monte Carlo datasets. The deterministic
package tests separately verify that the implementation reproduces the
prespecified weighted moments.

```{r weighting-oracle, echo=FALSE}
ar_pair_moments <- function(phi, n_times) {
  covariance <- centre_covariance(ar_covariance(phi, n_times))
  edge_moments(covariance, covariance, covariance, lag = 1L)
}

combine_pair_moments <- function(first, second, first_weight) {
  combined <- first_weight * first[c(
    "covariance", "variance_x", "variance_y"
  )] + (1 - first_weight) * second[c(
    "covariance", "variance_x", "variance_y"
  )]
  moment_correlation(combined)
}

long_moments <- ar_pair_moments(0.80, 12L)
short_moments <- ar_pair_moments(0.20, 5L)
long_edge_weight <- long_moments[["n_edges"]] /
  (long_moments[["n_edges"]] + short_moments[["n_edges"]])

weighting_table <- data.frame(
  weighting = c("dyad", "edge"),
  long_series_weight = c(0.50, long_edge_weight),
  exact_lag_1_target = c(
    combine_pair_moments(long_moments, short_moments, 0.50),
    combine_pair_moments(
      long_moments,
      short_moments,
      long_edge_weight
    )
  ),
  n_dyads = 80L,
  n_edges = 1200L
)
knitr::kable(weighting_table, digits = 3)
```

After member demeaning, the exact long- and short-series lag-1 correlations
are `r round(moment_correlation(long_moments), 3)` and
`r round(moment_correlation(short_moments), 3)`. The resulting dyad- and
edge-weighted targets differ because correlation is calculated from the pooled
weighted moments, not by averaging the two series-specific correlations.

# Interpretation and promotion boundary

The prototype is behaving in the intended qualitative direction when:

1. the correct fitted model completes reliably without systematic tail
   movement;
2. removing stable or same-occasion covariance moves its prespecified summary
   in the expected direction;
3. pooling the .80/.20 AR processes moves the two role-specific own-lag curves
   in opposite directions;
4. the shared/difference persistence control reproduces the exact joint
   finite-series concurrent, own-lag, and cross-lag targets;
5. exact gaps retain the intended factor-level lag; and
6. the two weighting rules agree for balanced schedules but target visibly
   different summaries when series length and persistence are associated.

Promotion should additionally require deterministic package tests, explicit
ILD simulation-backend tests, at least 95% fit/check completion in the core
cells, and expert review of the finite-series decomposition and weighting
estimand. The fitted restrictions are deliberately strong engineering checks;
their observed positions must not be described as power or specificity
estimates. The member-mean, same-occasion, and lagged summaries are correlated,
so movement outside the prespecified target is expected and should remain
visible during review.

```{r run-conclusion, echo=FALSE, results='asis'}
minimum_success <- min(completion$success_rate)
cat(sprintf(
  paste0(
    "This is a **%s run**. It recorded a minimum cell-specific completion ",
    "rate of %.1f%%. Interpret the fitted tables as integration evidence and ",
    "the covariance-oracle tables as the statistical targets.\n"
  ),
  profile_name,
  100 * minimum_success
))
```

# Reproducibility

All data and simulation seeds are deterministic functions of `r base_seed`.
The fixed design is generated once, while every outer response dataset and
every fitted-model reference uses its own explicit seed. Failures are never
retried.

A very small execution check can override the selected profile without editing
this file:

```r
rmarkdown::render(
  "dev/diagnostic_checks/gaussian-ild/outer-simulation-study.Rmd",
  params = list(
    profile = "smoke",
    outer_repetitions = 1,
    inner_simulations = 20,
    n_dyads = 24,
    n_times = 8
  )
)
```

```{r session-info}
sessionInfo()
```
