---
title: "Checking covariance pooling with generalized outcomes"
description: "Composition checks and model comparisons when latent covariance is pooled."
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
params:
  results_directory: "../report-data/generalized-covariance-pooling"
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 12, fig.height = 6, dpi = 160, out.width = "100%")
read_results <- function(filename) {
  read.csv(file.path(params$results_directory, filename))
}
conditions <- read_results("conditions.csv")
summary <- read_results("summary.csv") |>
  left_join(conditions, by = c("family", "condition"))
fit_counts <- read_results("fits.csv")
screening <- read_results("screening.csv")
settings <- read_results("settings.csv")
family_labels <- c(poisson = "Poisson", nbinom1 = "Negative binomial (NB1)",
  nbinom2 = "Negative binomial (NB2)", nbinom12 = "Negative binomial (NB12)", Gamma = "Gamma",
  beta = "Beta", beta_family = "Beta", lognormal = "Lognormal",
  tweedie = "Tweedie", binomial = "Binomial", betabinomial = "Beta-binomial",
  ordinal = "Ordinal", t = "Student t", skewnormal = "Skew-normal", bell = "Bell",
  genpois = "Generalized Poisson", compois = "Conway-Maxwell-Poisson",
  truncated_poisson = "Truncated Poisson", truncated_nbinom1 = "Truncated NB1",
  truncated_nbinom2 = "Truncated NB2", truncated_compois = "Truncated COM-Poisson",
  truncated_genpois = "Truncated generalized Poisson", zi_poisson = "Zero-inflated Poisson",
  hurdle_nbinom2 = "Hurdle NB2")
family_label <- function(family) {
  dplyr::recode(family, !!!family_labels)
}
scenario_labels <- c(null = "Correct pooling",
  variance_small = "Latent SD ratio\n1.25", variance_large = "Latent SD ratio\n1.50",
  correlation_small = "Latent correlations\n0.10 / 0.30 / 0.50",
  correlation_large = "Latent correlations\n-0.10 / 0.30 / 0.70")
summary <- summary |> mutate(family = family_label(family),
  scenario = factor(scenario, levels = names(scenario_labels), labels = scenario_labels))
false_alarm_results <- summary |> filter(scenario == "Correct pooling")
plot_rates <- function(results, methods, y_label, y_maximum = 1) {
  results <- results |> filter(method %in% names(methods)) |>
    mutate(method = factor(method, levels = names(methods), labels = methods))
  ggplot(results, aes(n_dyads, rate, colour = method, linetype = method)) +
    geom_line() +
    geom_errorbar(aes(ymin = lower, ymax = upper), width = 5, show.legend = FALSE) +
    geom_point(size = 2, show.legend = FALSE) +
    scale_x_continuous(breaks = sort(unique(conditions$n_dyads))) +
    scale_y_continuous(labels = scales::label_percent(), limits = c(0, y_maximum)) +
    labs(x = "Total number of dyads", y = y_label, colour = NULL, linetype = NULL) +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom", legend.key.width = grid::unit(1.5, "cm"),
      panel.grid.minor = element_blank())
}
```

```{r preliminary, results='asis'}
if (any(conditions$repetitions < 500)) {
  cat("> **Preliminary run.** These few datasets check that the study works. ",
    "The planned study uses 500 datasets per condition.\n\n")
}
```

## What the study asks

Can composition checks reveal inappropriate covariance pooling for non-Gaussian
outcomes? This extends the [Gaussian study](covariance-pooling.html) to models with
a Gaussian latent effect for each partner, on the model's link scale.

Both models correctly allow different means and actor and partner effects for
female-female, female-male, and male-male dyads. The fuller model estimates a
separate latent covariance for each composition. The pooled model estimates one
exchangeable covariance across all dyads. **Both retain partner dependence and
the family's observation-level variation.**

Screening retained
`r paste(family_label(screening$family[screening$selected]), collapse = ", ")`
under our fitting and recovery criteria. The other `r sum(!screening$selected)`
candidates did not qualify in these settings (see the
[screening code](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/generalized-covariance-pooling/screen-families.R)).

The full model provides a check on fitting: it uses the structure that generated
the data. Few flags do not by themselves establish accurate parameter estimates.
Each condition uses `r paste(unique(conditions$repetitions), collapse = ", ")`
generated datasets and `r paste(unique(conditions$reference_draws), collapse = ", ")`
reference simulations per checked fit. Dyads are divided roughly equally among
the three compositions, so 100 total dyads means about 33 per composition.

## False alarms when pooling is correct

A flag means at least one of the 14 composition summaries falls outside its middle
95% simulated range. We check these same summaries for the pooled and full models.
Flags can occur by chance, especially when checking several summaries. They invite
investigation, not formal rejection of the model. The model comparison uses a
likelihood-ratio test at 5%.

```{r false-alarms, fig.height=max(4, 3 * ceiling(n_distinct(summary$family) / 2)), fig.alt="False-alarm rates for composition checks of the pooled and full models and for model comparisons, by family and total number of dyads."}
false_alarm_methods <- c(composition_any = "Pooled-model checks", full_any = "Full-model checks",
  likelihood_ratio = "Model comparison")
false_alarm_axis_maximum <- false_alarm_results |>
  filter(method %in% names(false_alarm_methods)) |>
  summarise(upper = max(upper, na.rm = TRUE)) |>
  pull(upper)
false_alarm_axis_maximum <- max(0.1, ceiling(false_alarm_axis_maximum * 10) / 10)
plot_rates(false_alarm_results, false_alarm_methods, "False-alarm rate", false_alarm_axis_maximum) +
  facet_wrap(~family, ncol = 2) +
  geom_hline(yintercept = 0.05, colour = "grey60", linetype = "dashed") +
  scale_colour_manual(values = rep("black", 3)) +
  scale_linetype_manual(values = c("solid", "dotted", "longdash"))
```

The grey line marks 5% as a reference, not a guaranteed rate. Points use only
datasets with a usable fit and check. Bars show 95% intervals for uncertainty from
the finite number of study datasets. Missing results do not count as agreement.

## When latent covariance is pooled incorrectly

We vary either the female/male latent SD ratio within female-male dyads or the
latent partner correlations across compositions. Correlation labels list
female-female, female-male, and male-male dyads, in that order.

The checks compare **outcome summaries**, not latent parameters. A latent
correlation of 0.3 need not give an outcome correlation of 0.3. Unlike in the
Gaussian study, preserving pooled latent covariance need not preserve pooled
outcome covariance.

These plots count **any composition flag**, without claiming that it identifies
the right summary or direction. The link function and changes in other fitted
parameters complicate that attribution. The full model's flags remain a reference
for what happens with the correct structure.

Higher flag rates alone do not establish a better formal test. Read them alongside
the false-alarm rates, which need not be equal across methods.

```{r alternatives, fig.height=max(4, 2.6 * n_distinct(summary$family)), fig.alt="Any composition flag for the pooled and full models and rejection by the model comparison, across families, sample sizes, and four latent covariance conditions."}
summary |> filter(scenario != "Correct pooling") |>
  plot_rates(c(composition_any = "Pooled-model checks", full_any = "Full-model checks",
    likelihood_ratio = "Model comparison"), "Flag or rejection rate") +
  facet_grid(family ~ scenario) +
  scale_colour_manual(values = c("#D55E00", "grey50", "#009E73")) +
  scale_linetype_manual(values = c("solid", "dotted", "longdash"))
```

## Study details

<details>
<summary>Data and models</summary>

Every dataset is prepared with `prepare_dyad_data()`. The covariance terms follow
the package's dyadic structures. The pooled model has a shared dyad effect and an
arbitrary member-contrast effect. The fuller model has separate shared and contrast
effects for each same-role composition and an unstructured female-male block.
They estimate two and seven latent covariance parameters, respectively. The
likelihood-ratio comparison uses five degrees of freedom. Family-specific
dispersion, where applicable, has the same specification in both models.

```{r family-settings}
settings |> mutate(family = family_label(family),
  response_settings = gsub("; ", ", ", response_settings, fixed = TRUE)) |>
  select(family, link, response_settings) |>
  knitr::kable(col.names = c("Family", "Link", "Response distribution"))
```

```{r zero-components, results='asis'}
if (any(settings$family %in% c("zi_poisson", "hurdle_nbinom2"))) {
  cat("For zero-inflated and hurdle models, the Gaussian latent effects are in the ",
    "response component. Extra zeros occur independently with probability 0.25. ",
    "Both models estimate a zero-component intercept, without random effects.\n\n")
}
```

```{r tweedie-sampler, results='asis'}
if ("tweedie" %in% settings$family) {
  cat("Tweedie simulations use the [equivalent sampler repair](",
    "https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/README.md#tweedie-simulation-repair)",
    " from the earlier study to avoid slow draws when fitted power approaches 2. ",
    "Fitting is unchanged.\n\n", sep = "")
}
```

Under correct pooling, latent SDs are 0.6 and latent correlations are 0.3.
The SD conditions preserve the sum of the two female-male variances and their
covariance. The correlation conditions change the two same-role correlations by
equal amounts in opposite directions. These changes preserve pooled latent
covariance, but the response distribution and link affect their consequences for
outcome summaries.

Predictors are standard normal, with partner correlation 0.3, and are independent
of the latent effects. Role- and composition-specific coefficients match the
[Gaussian study](covariance-pooling.html), with `log(3)` added to each intercept
for families with a log link.
Conditional on the predictors and latent effects, the partners' outcomes are
drawn independently from the response family.

Checks subtract fitted fixed-effect predictions. Pooled and composition checks
use the same reference simulations, holding fitted parameters and predictors
fixed without refitting. These are descriptive checks of selected summaries,
not calibrated significance tests. The
[Stan guide](https://mc-stan.org/docs/stan-users-guide/posterior-predictive-checks.html#which-statistics-to-test)
discusses why summaries already estimated freely by a model can be less informative
for checking it.

The mean model is correct throughout. In practice, omitted predictors can also
cause mismatches. Agreement does not establish that fixed effects can be pooled
or that the model fits well in every respect.

</details>

<details>
<summary>Checks across all compositions</summary>

For comparison, the table checks four summaries across all dyads, using the pooled
model at the largest sample size. These overall summaries can hide differences
between compositions.

```{r pooled-rates}
summary |> filter(n_dyads == max(n_dyads), method == "pooled_any") |>
  mutate(scenario = gsub("\n", " ", scenario),
    percentage = ifelse(checked > 0, sprintf("%.1f%%", 100 * rate), "—")) |>
  select(family, scenario, checked, percentage) |>
  knitr::kable(col.names = c("Family", "Condition", "Available datasets", "Flagged"))
```

</details>

<details>
<summary>Fitting and saved results</summary>

```{r fitting}
fit_counts |> mutate(family = family_label(family),
  model = recode(model, pooled = "Pooled", composition = "Full")) |> group_by(family, model) |>
  summarise(across(c(attempted, usable, with_warnings, check_failures,
    comparison_failures), sum), .groups = "drop") |>
  knitr::kable(col.names = c("Family", "Model", "Attempted", "Usable", "Warnings",
    "Check failures", "Comparison failures"))
```

Rates exclude failed fits, convergence or Hessian problems, and unavailable
checks. Warnings are counted separately and do not necessarily invalidate a fit.
Model comparisons require both fits, so methods can use different subsets of
datasets. The screening does not establish that
excluded families are generally unsuitable, or that fitting problems were caused
by non-identifiability.

The [study script](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/generalized-covariance-pooling/run.R)
records per-dataset results and exports the tables used here. Re-rendering this
report changes no fitted models or results.

</details>
