---
title: "Common Fate Model"
bibliography: ../../vignettes/references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Common Fate Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

diagram_candidates <- c(
  "diagram-helpers.Rinc",
  "../../vignettes/diagram-helpers.Rinc",
  "vignettes/diagram-helpers.Rinc"
)
diagram_helper <- diagram_candidates[file.exists(diagram_candidates)][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(diagram_helper, envir = knitr::knit_global())
```

```{r setup}
library(interdep)
```

> This vignette is a draft and is under construction.

The Common Fate Model (CFM) represents the two members' reports as indicators
of a shared latent dyad-level construct [@ledermannAnalyzingDyadicData2017;
@iidaModelsInterdependentIndividuals2018]. In the predictor-to-outcome CFM
below, the shared predictor level predicts the shared outcome level. The
unit-loading paths define each latent variable from the two member reports.

```{r conceptual-cfm-diagram, echo=FALSE, fig.width=10, fig.height=5.8, out.width="100%", fig.cap="SEM representation of a predictor-to-outcome Common Fate Model. Each pair of member reports loads with a fixed unit loading on a shared latent dyad-level construct. The latent predictor level predicts the latent outcome level. Same-member predictor and outcome residuals may covary, and the latent outcome level has a residual disturbance.", fig.alt="Structural equation path diagram for a Common Fate Model. X 1 and X 2 are unit-loading indicators of latent X level, and Y 1 and Y 2 are unit-loading indicators of latent Y level. Latent X level predicts latent Y level with coefficient b level. The residuals of X 1 and Y 1 covary, as do the residuals of X 2 and Y 2. Latent Y level has a residual disturbance."}
draw_cfm_diagram()
```

The orange curved paths retain the predictor--outcome association specific to
each member rather than assigning it to the shared dyad-level effect. Only the
SEM representation is shown here; there is no individual-level multilevel-model
diagram for this CFM.
