---
title: "Common Fate Model"
bibliography: ../../vignettes/references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Common Fate Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

diagram_candidates <- c(
  "diagram-helpers.Rinc",
  "../../vignettes/diagram-helpers.Rinc",
  "vignettes/diagram-helpers.Rinc"
)
diagram_helper <- diagram_candidates[file.exists(diagram_candidates)][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(diagram_helper, envir = knitr::knit_global())
```

```{r setup, message=FALSE}
library(interdep)
library(lavaan)
```

> This vignette is a draft and is under construction.

The Common Fate Model (CFM) represents the two members' reports as indicators
of a shared latent dyad-level construct [@ledermannAnalyzingDyadicData2017;
@iidaModelsInterdependentIndividuals2018]. In the predictor-to-outcome CFM
below, the shared predictor level predicts the shared outcome level. The
unit-loading paths define each latent variable from the two member reports.

```{r conceptual-cfm-diagram, echo=FALSE, fig.width=10, fig.height=5.8, out.width="100%", fig.cap="SEM representation of a predictor-to-outcome Common Fate Model. Each pair of member reports loads with a fixed unit loading on a shared latent dyad-level construct. The latent predictor level predicts the latent outcome level. Same-member predictor and outcome residuals may covary, and the latent outcome level has a residual disturbance.", fig.alt="Structural equation path diagram for a Common Fate Model. X 1 and X 2 are unit-loading indicators of latent X level, and Y 1 and Y 2 are unit-loading indicators of latent Y level. Latent X level predicts latent Y level with coefficient b level. The residuals of X 1 and Y 1 covary, as do the residuals of X 2 and Y 2. Latent Y level has a residual disturbance."}
draw_cfm_diagram()
```

The curved paths, labelled $c_{\mathrm{res},1}$ and
$c_{\mathrm{res},2}$, retain the predictor--outcome association specific to
each member rather than assigning it to the shared dyad-level effect. To match
the DIM and DSM diagrams, the shared-level path uses the predictor-mean/between-
dyad field color, whereas the member-specific residual paths use the
within-dyad field color. These colors describe fields of variation, not actor
and partner effects. Only the SEM representation is shown here; there is no
individual-level multilevel-model diagram for this CFM.

## Model specification

Each shared level is represented by a two-indicator confirmatory factor model.
The regression between those two latent levels makes the complete CFM a
structural equation model.

For dyad $j$ and member $i \in \{1, 2\}$, the standard CFM can be written as

$$
X_{ij} = X_{\mathrm{level},j} + \varepsilon_{Xij}, \qquad
Y_{ij} = Y_{\mathrm{level},j} + \varepsilon_{Yij},
$$

with the dyad-level regression

$$
Y_{\mathrm{level},j}
= b_{\mathrm{level}}X_{\mathrm{level},j} + r_{Yj}.
$$

Following Model 4 of Iida et al. [-@iidaModelsInterdependentIndividuals2018],
**both** loadings for each latent variable are fixed to 1. This differs from
`lavaan`'s usual marker-variable default, which fixes only one loading. The unit
loadings treat the two reports symmetrically and keep the latent variables on the
original measurement scales; they do not define the latent scores as literal
arithmetic averages.

The model freely estimates:

- the variance of the latent predictor level and the residual variance of the
  latent outcome level;
- all four indicator residual variances;
- the dyad-level regression $b_{\mathrm{level}}$; and
- the two same-member predictor--outcome residual covariances.

All other indicator residual covariances are fixed to zero by omission. Iida et
al.'s standard CFM does not constrain the two members' intercepts, residual
variances, or same-member residual covariances to be equal. With a mean
structure, the four observed intercepts are free and the two latent means are
fixed to zero. The mean structure is therefore saturated and does not change the
model's one degree of freedom. At the covariance level, the model estimates nine
parameters from the ten unique observed covariance elements, leaving one degree
of freedom.

## Data

To compare the CFM directly with the other vignettes, we use the same
`example_dyadic_crosssectional` data, with `communication` as the predictor and
`satisfaction` as the outcome. `lavaan` represents the two members as separate
observed variables, so the long-format data first need one row per dyad.

```{r prepare-cfm-data}
cfm_data <- reshape(
  example_dyadic_crosssectional[
    c("coupleID", "gender", "communication", "satisfaction")
  ],
  idvar = "coupleID",
  timevar = "gender",
  direction = "wide"
)

names(cfm_data) <- gsub(".", "_", names(cfm_data), fixed = TRUE)
cfm_data <- cfm_data[
  c(
    "coupleID",
    "communication_female", "communication_male",
    "satisfaction_female", "satisfaction_male"
  )
]
row.names(cfm_data) <- NULL

head(cfm_data)
```

The `lavaan` syntax directly follows Iida et al.'s Mplus specification, now
using the variables from the package example:

```{r cfm-model-syntax}
cfm_model <- '
  # Shared dyad levels: fix all four loadings to 1
  communication_level =~
    1 * communication_female + 1 * communication_male
  satisfaction_level =~
    1 * satisfaction_female + 1 * satisfaction_male

  # Dyad-level association
  satisfaction_level ~ b_level * communication_level

  # Same-member residual covariances
  communication_female ~~ c_female * satisfaction_female
  communication_male ~~ c_male * satisfaction_male
'
```

Observed residual variances, the latent communication variance, and the latent
satisfaction disturbance variance are free by default. The four observed
intercepts are also freely estimated. Full-information maximum likelihood uses
all dyads, including those with a partially missing member report.

```{r fit-cfm}
cfm_fit <- suppressWarnings(
  sem(
    cfm_model,
    data = cfm_data,
    estimator = "ML",
    missing = "fiml",
    meanstructure = TRUE
  )
)

loading_constraints <- parameterTable(cfm_fit)
loading_constraints <- loading_constraints[
  loading_constraints$op == "=~",
]

cfm_admissible <- suppressWarnings(lavInspect(cfm_fit, "post.check"))

stopifnot(
  lavInspect(cfm_fit, "converged"),
  fitMeasures(cfm_fit, "df") == 1,
  nrow(loading_constraints) == 4,
  all(loading_constraints$free == 0),
  all(loading_constraints$ustart == 1),
  all(loading_constraints$est == 1)
)
```

The checks above verify convergence, the expected one degree of freedom, and
that all four unstandardized loadings remain fixed at exactly 1. Admissibility
is checked separately because convergence alone does not rule out impossible
variance estimates.

```{r extract-cfm-estimates, include=FALSE}
cfm_all_estimates <- parameterEstimates(cfm_fit, standardized = TRUE)

key_labels <- c("b_level", "c_female", "c_male")
cfm_estimates <- cfm_all_estimates[
  match(key_labels, cfm_all_estimates$label),
]

variance_rows <- with(
  cfm_all_estimates,
  op == "~~" & lhs == rhs
)
cfm_variances <- cfm_all_estimates[variance_rows, ]

outcome_variance <- cfm_variances[
  cfm_variances$lhs == "satisfaction_level",
]
outcome_variance_negative <- is.finite(outcome_variance$est) &&
  outcome_variance$est < 0
cfm_r_squared <- if (cfm_admissible) {
  lavInspect(cfm_fit, "r2")[["satisfaction_level"]]
} else {
  NA_real_
}

cfm_b <- cfm_estimates[cfm_estimates$label == "b_level", ]
cfm_residual_correlations <- cfm_estimates$std.all[
  cfm_estimates$label %in% c("c_female", "c_male")
]

cfm_fitted_caption <- paste(
  "Fitted standard CFM for the package's cross-sectional example.",
  "All four unstandardized loadings are fixed at 1.",
  "The shared-level path and same-member residual correlations are read",
  "directly from the fitted model. The label below the latent outcome reports",
  "R-squared for an admissible solution or flags an inadmissible variance."
)
cfm_fitted_alt <- if (cfm_admissible) {
  sprintf(
    paste(
      "Fitted CFM. Dyad-level effect %.2f, standardized effect %.2f, female",
      "and male residual correlations %.2f and %.2f, and outcome R-squared %.2f."
    ),
    cfm_b$est, cfm_b$std.all,
    cfm_residual_correlations[1], cfm_residual_correlations[2],
    cfm_r_squared
  )
} else if (outcome_variance_negative) {
  sprintf(
    paste(
      "Fitted CFM. Dyad-level effect %.2f, female and male residual correlations",
      "%.2f and %.2f, and outcome disturbance variance %.2f (inadmissible)."
    ),
    cfm_b$est,
    cfm_residual_correlations[1], cfm_residual_correlations[2],
    outcome_variance$est
  )
} else {
  paste(
    "Fitted Common Fate Model. The solution is inadmissible; inspect the",
    "reported variance estimates for the failed diagnostic."
  )
}
```

```{r cfm-diagnostic, echo=FALSE, results='asis'}
if (cfm_admissible) {
  cat(
    "> **Model diagnostic:** The model converged and all estimated variances ",
    "are admissible.\n\n",
    sep = ""
  )
} else if (outcome_variance_negative) {
  cat(
    sprintf(
      paste0(
        "> **Model diagnostic:** The model converged, but the latent ",
        "satisfaction disturbance variance is %.2f. This negative variance ",
        "makes the solution inadmissible, so its paths and fit indices should ",
        "not be interpreted substantively.\n\n"
      ),
      outcome_variance$est
    )
  )
} else {
  cat(
    "> **Model diagnostic:** The model converged, but at least one ",
    "post-estimation admissibility check failed. Inspect the variance table ",
    "before interpreting the model.\n\n",
    sep = ""
  )
}
```

### Fitted model

The fitted diagram is generated from `cfm_fit`: none of its displayed numbers
are typed into the figure. For an admissible solution, the shared-level path
shows both the unstandardized effect and standardized $\beta$, and the latent
outcome shows $R^2$. For an inadmissible solution, the diagram instead displays
the offending variance. The curved-path labels are residual **correlations**,
which are easier to compare than scale-dependent covariances.

```{r fitted-cfm-diagram, echo=FALSE, fig.width=10, fig.height=5.8, out.width="100%", fig.cap=cfm_fitted_caption, fig.alt=cfm_fitted_alt}
draw_cfm_diagram(
  b_level = cfm_b$est,
  beta_level = cfm_b$std.all,
  residual_correlations = cfm_residual_correlations,
  r_squared = cfm_r_squared,
  outcome_residual_variance = outcome_variance$est,
  admissible = cfm_admissible,
  member_ids = c("F", "M"),
  member_names = c("Female", "Male"),
  predictor_name = "communication",
  outcome_name = "satisfaction"
)
```

```{r cfm-interpretation, echo=FALSE, results='asis'}
if (cfm_admissible) {
  cat(
    sprintf(
      paste0(
        "A one-unit higher latent shared communication level is associated ",
        "with a %.2f-unit higher latent shared satisfaction level. The ",
        "standardized association is %.2f.\n\n"
      ),
      cfm_b$est, cfm_b$std.all
    )
  )
} else if (outcome_variance_negative) {
  cat(
    sprintf(
      paste0(
        "The model estimates a shared dyad-level coefficient of ",
        "$b_{\\mathrm{level}} = %.2f$. If the CFM solution were admissible, ",
        "this would mean that a one-unit higher latent shared communication ",
        "level is associated with a %.2f-unit higher latent shared ",
        "satisfaction level. However, the model estimates a negative latent ",
        "satisfaction disturbance variance. The solution is therefore ",
        "inadmissible, and this coefficient should not be interpreted as a ",
        "reliable substantive estimate.\n\n"
      ),
      cfm_b$est, cfm_b$est
    )
  )
} else {
  cat(
    "The fitted solution is inadmissible and should not be interpreted ",
    "substantively. Inspect the reported variance estimates to identify the ",
    "failed diagnostic.\n\n",
    sep = ""
  )
}
```

The indicator residual variances and latent outcome disturbance variance are
reported separately below. The freely estimated observed intercepts reproduce
the four member-specific means and are omitted from the diagram.

```{r cfm-fit-table, echo=FALSE}
fit_indices <- fitMeasures(
  cfm_fit,
  c("chisq", "df", "pvalue", "cfi", "tli", "rmsea", "srmr")
)

knitr::kable(
  data.frame(
    Measure = c("Chi-square", "df", "p", "CFI", "TLI", "RMSEA", "SRMR"),
    Estimate = as.numeric(fit_indices)
  ),
  digits = 3,
  caption = "Model fit for the package's cross-sectional example."
)
```

```{r cfm-parameter-table, echo=FALSE}
term_labels <- c(
  b_level = "Dyad-level effect",
  c_female = "Female residual covariance",
  c_male = "Male residual covariance"
)

knitr::kable(
  data.frame(
    Parameter = unname(term_labels[cfm_estimates$label]),
    Estimate = cfm_estimates$est,
    SE = cfm_estimates$se,
    p = cfm_estimates$pvalue,
    `Standardized estimate` = cfm_estimates$std.all,
    check.names = FALSE
  ),
  digits = 3,
  caption = "Key estimates from the standard CFM."
)
```

```{r cfm-variance-table, echo=FALSE}
residual_names <- c(
  "communication_female", "communication_male",
  "satisfaction_female", "satisfaction_male",
  "satisfaction_level"
)
residual_variances <- cfm_variances[
  match(residual_names, cfm_variances$lhs),
]

residual_labels <- c(
  "Female communication residual",
  "Male communication residual",
  "Female satisfaction residual",
  "Male satisfaction residual",
  "Latent satisfaction disturbance"
)

knitr::kable(
  data.frame(
    Parameter = residual_labels,
    `Unstandardized variance` = residual_variances$est,
    `Standardized residual variance` = residual_variances$std.all,
    check.names = FALSE
  ),
  digits = 3,
  caption = "Residual variance estimates. A negative value identifies an inadmissible solution."
)
```

We do not force the negative variance to zero. Such a bound would change the
estimation problem, place the estimate on the boundary, and require additional
care for uncertainty and model-fit tests. If the example data are changed later,
the diagnostic text and fitted diagram will update from the new model fit.
