---
title: "Beyond Zero Correlation: An Exchangeable Location–Scale DIM"
author: "Pascal Küng"
description: >
  A worked introduction to an exchangeable DIM and APIM extension in which
  the dyad-mean residual is associated with within-dyad residual variability.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beyond Zero Correlation: An Exchangeable Location–Scale DIM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  dpi = 144
)

diagram_candidates <- c(
  "../../vignettes/diagram-helpers.Rinc",
  "vignettes/diagram-helpers.Rinc",
  "diagram-helpers.Rinc"
)
diagram_helper <- diagram_candidates[file.exists(diagram_candidates)][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.", call. = FALSE)
}
sys.source(diagram_helper, envir = knitr::knit_global())
diagram_device <- .vignette_diagram_device()
knitr::opts_chunk$set(dev = diagram_device, fig.ext = diagram_device)

seesaw_candidates <- c(
  "../workshop/R/draw-residual-seesaw.R",
  "dev/workshop/R/draw-residual-seesaw.R",
  "../../dev/workshop/R/draw-residual-seesaw.R"
)
seesaw_helper <- seesaw_candidates[file.exists(seesaw_candidates)][1]

if (is.na(seesaw_helper)) {
  stop("Could not find dev/workshop/R/draw-residual-seesaw.R.", call. = FALSE)
}
sys.source(seesaw_helper, envir = knitr::knit_global())

fmt <- function(x, digits = 2) {
  formatC(x, format = "f", digits = digits)
}

fmt_p <- function(x) {
  format.pval(x, digits = 2, eps = 0.001)
}

plot_colours <- list(
  ink = "#263238",
  muted = "#64748B",
  grid = "#E2E8F0",
  surface = "#F8FAFC",
  residual_1 = "#004D40",
  residual_2 = "#00A6B2",
  residual_2_text = "#007486",
  residual_scale = "#546E7A",
  standard = "#94A3B8",
  truth = "#334155"
)

transparent <- function(colour, alpha = 0.25) {
  grDevices::adjustcolor(colour, alpha.f = alpha)
}
```

In an exchangeable dyad, the member labels are arbitrary. Swapping them leaves
the dyad mean unchanged but reverses the signed member difference. The standard
Dyad-Individual Model (DIM) respects this symmetry by setting the residual
covariance between the mean and the signed difference to zero.

But zero signed covariance does **not** imply constant discrepancy. Partners
may be farther apart at some dyad-mean residual levels while either arbitrary
member remains equally likely to be higher. This vignette simulates that
pattern, shows what the standard DIM misses, fits an exchangeable
location–scale extension, and translates the result back to Actor-Partner
Interdependence Model (APIM) language.

By the end, the distinction should be clear:

- signed residual correlation summarizes whether elevator position is
  **linearly** associated with a preferred signed seat difference;
- conditional sign symmetry—not zero correlation alone—is what makes the
  residual model exchangeable;
- residual scale asks how far apart the two seats tend to be;
- exchangeability forbids a preferred arbitrary direction but does not forbid
  changing residual scale.

Everything used to fit the extension below lives inside this development
vignette. It is a mathematical prototype, not an exported `interdep` function.

## Elevator and seesaw coordinates

For dyad $j$, define the dyad mean and half-difference:

$$
M_{Xj}=\frac{X_{1j}+X_{2j}}{2},
\qquad
D_{Xj}=\frac{X_{1j}-X_{2j}}{2},
$$

and analogously for $M_{Yj}$ and $D_{Yj}$. $D_X$ is member 1's deviation from
the dyad mean; member 2's deviation is $-D_X$. We can return to the member
scores using

$$
X_{1j}=M_{Xj}+D_{Xj},
\qquad
X_{2j}=M_{Xj}-D_{Xj}.
$$

The two coordinates have a useful physical interpretation:

- the **elevator** $M$ moves both partners together;
- the **seesaw** $D$ moves one partner up and the other down.

```{r elevator-seesaw, echo=FALSE, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="The elevator residual moves both members together. The seesaw residual moves them apart. Swapping arbitrary seats reverses the seesaw but not the elevator.", fig.alt="Two-panel illustration. An elevator moves both members of a dyad together and represents the dyad-mean residual. A seesaw moves the members in opposite directions and represents the within-dyad residual."}
draw_residual_seesaw("dim")
```

For outcome residuals,

$$
e_{1j}=r_{Mj}+r_{Dj},
\qquad
e_{2j}=r_{Mj}-r_{Dj}.
$$

Swapping the two arbitrary member labels produces

$$
(r_M,r_D)\longmapsto(r_M,-r_D).
$$

> **Central idea.** Exchangeability says that the seesaw has no preferred
> side. It does not say that every dyad's seesaw must swing equally far.

The package DSM uses a full difference rather than the half-difference used
here. If $\Delta=2D$, the difference residual standard deviation doubles and
its log-variance intercept increases by $\log(4)$. The substantive model and
the scale slope introduced below do not change.

## From the DSM to the standard DIM

In these score coordinates, a full linear Dyadic Score Model (DSM) is

$$
M_Y=a_{10}+a_{11}M_X+a_{12}D_X+r_M,
$$

$$
D_Y=a_{20}+a_{21}M_X+a_{22}D_X+r_D.
$$

After a member-label swap, $M_X$ and $M_Y$ stay unchanged, whereas $D_X$ and
$D_Y$ reverse sign. For the usual linear Gaussian model with one constant
residual covariance matrix, exchangeability therefore requires

$$
a_{12}=a_{20}=a_{21}=0,
\qquad
\operatorname{Cov}(r_M,r_D)=0.
$$

What remains is the standard DIM:

$$
M_Y=\beta_0+\beta_BM_X+r_M,
$$

$$
D_Y=\beta_WD_X+r_D.
$$

$\beta_B$ is the between-dyad effect and $\beta_W$ is the within-dyad effect.
The missing extension is not an ordinary mean–difference correlation. It is a
model for how the **variance** of the signed difference changes.

To see why, write

$$
r_D=SA,
\qquad
S\in\{-1,+1\},
\qquad
A\geq 0.
$$

$A$ records how far the seesaw tilts, and $S$ records its arbitrary direction.
Under exchangeability, both signs are equally likely conditional on $r_M$ and
$A$. Thus $E(r_D\mid r_M,A)=0$ and consequently

$$
\begin{aligned}
\operatorname{Cov}(r_M,r_D)
&=E(r_Mr_D) \\
&=E\!\left[r_ME(r_D\mid r_M)\right] \\
&=0.
\end{aligned}
$$

The signed covariance cancels even if $A$ changes strongly with $r_M$. The
usual Gaussian DIM goes further: zero covariance plus joint normality makes
$r_M$ and $r_D$ independent, implying one constant seesaw variance.
Exchangeability itself does not require that constant variance.

## The location–scale extension

Standardize the elevator residual:

$$
z_{Mj}=\frac{r_{Mj}}{\sigma_M},
\qquad
r_{Mj}\sim\mathcal{N}(0,\sigma_M^2).
$$

Keep the seesaw residual centered and symmetric, but allow its variance to
depend on elevator position:

$$
r_{Dj}\mid r_{Mj}
\sim
\mathcal{N}\!\left(0,v_D(z_{Mj})\right),
$$

$$
\log v_D(z_{Mj})
=2\log(\sigma_{D0})+\lambda z_{Mj}.
$$

Here $\sigma_{D0}$ is the seesaw residual standard deviation at $z_M=0$—that
is, for dyads whose outcome mean is exactly as predicted.
The scale slope $\lambda$ has a direct interpretation:

- $\lambda=0$ gives the standard Gaussian DIM;
- $\lambda>0$ means that the seesaw spreads out as the residual elevator rises;
- $\lambda<0$ means that the seesaw tightens as the residual elevator rises.

A one-standard-deviation increase in $r_M$ multiplies the seesaw variance by
$\exp(\lambda)$ and its standard deviation by $\exp(\lambda/2)$. At every
elevator position, however,

$$
E(r_D\mid r_M)=0.
$$

The model changes gap **magnitude**, not gap **direction**, and therefore
remains exchangeable.

## A worked simulation

```{r simulation-values}
simulation_truth <- list(
  J = 1500L,
  beta_0 = 5,
  beta_between = 0.60,
  beta_within = 0.30,
  sigma_mean = 0.80,
  sigma_dev0 = 0.55,
  lambda = 0.90,
  seed = 123L
)
```

We simulate `r format(simulation_truth$J, big.mark = ",")` complete dyads. The
true between-dyad effect is `r fmt(simulation_truth$beta_between)`, the true
within-dyad effect is `r fmt(simulation_truth$beta_within)`, and the true scale
slope is `r fmt(simulation_truth$lambda)`. Both predictor coordinates are
centered, so the intercept is the expected outcome mean for a dyad at the
average predictor level.

<details>
<summary><strong>Show the complete simulation helper</strong></summary>

```{r simulation-helper}
simulate_location_scale_dim <- function(
    J,
    beta_0,
    beta_between,
    beta_within,
    sigma_mean,
    sigma_dev0,
    lambda,
    seed
) {
  set.seed(seed)

  # Predictor elevator and seesaw coordinates
  m_x <- rnorm(J)
  m_x <- m_x - mean(m_x)
  d_x <- rnorm(J, sd = 0.75)
  d_x <- d_x - mean(d_x)

  # Exchangeable residual elevator and seesaw
  z_mean <- rnorm(J)
  z_dev <- rnorm(J)
  r_mean <- sigma_mean * z_mean
  sd_dev <- sigma_dev0 * exp(lambda * z_mean / 2)
  r_dev <- sd_dev * z_dev

  # Outcome coordinates and member scores
  m_y <- beta_0 + beta_between * m_x + r_mean
  d_y <- beta_within * d_x + r_dev

  data.frame(
    dyad = seq_len(J),
    x1 = m_x + d_x,
    x2 = m_x - d_x,
    y1 = m_y + d_y,
    y2 = m_y - d_y,
    m_x, d_x, m_y, d_y,
    r_mean, r_dev, z_mean, z_dev, sd_dev
  )
}
```

</details>

```{r simulate-data}
dat <- do.call(simulate_location_scale_dim, simulation_truth)

with(dat, c(
  signed_correlation = cor(r_mean, r_dev),
  magnitude_association = cor(r_mean, abs(r_dev))
))
```

The signed correlation is close to zero, as expected from the conditionally
symmetric construction: there is essentially no **linear** association between
elevator position and the signed tilt. The magnitude association is positive:
the seesaw tends to tilt farther at higher elevator residuals. Conditional sign
symmetry—not the sample correlation by itself—is what guarantees that neither
arbitrary seat is preferred.

The simulation core is only four lines:

```{r simulation-core, eval=FALSE}
r_mean <- sigma_mean * z_mean
sd_dev <- sigma_dev0 * exp(lambda * z_mean / 2)
r_dev <- rnorm(J, mean = 0, sd = sd_dev)
d_y <- beta_within * d_x + r_dev
```

Positive and negative seesaw tilts remain equally likely. Only their spread
changes.

## Fit the standard DIM first

For one complete cross-sectional pair per dyad, the two score-coordinate mean
equations can be fitted as

```{r standard-dim-equations, eval=FALSE}
lm(m_y ~ m_x, data = dat)
lm(d_y ~ 0 + d_x, data = dat)
```

These calls show the two conditional-mean equations. The joint fit below
estimates the same mean effects together with the residual model needed to
compare the standard and extended DIM on exactly the same likelihood. That
likelihood factorizes as

$$
p(r_M,r_D)=p(r_M)\,p(r_D\mid r_M).
$$

The standard model fixes $\lambda=0$; the extended model estimates it.

<details>
<summary><strong>Show the internal joint maximum-likelihood implementation</strong></summary>

The implementation includes the constant Jacobian for transforming each raw
outcome pair $(Y_1,Y_2)$ to $(M_Y,D_Y)$. The reported log likelihood and AIC are
therefore on the raw member-outcome scale.

```{r joint-likelihood}
dim_ls_nll <- function(par, dat, extended = TRUE) {
  b0 <- par[["b0"]]
  b_between <- par[["b_between"]]
  b_within <- par[["b_within"]]
  log_sd_mean <- par[["log_sd_mean"]]
  log_sd_dev0 <- par[["log_sd_dev0"]]
  lambda <- if (extended) par[["lambda"]] else 0

  r_mean <- dat$m_y - b0 - b_between * dat$m_x
  r_dev <- dat$d_y - b_within * dat$d_x
  z_mean <- r_mean * exp(-log_sd_mean)
  log_var_dev <- 2 * log_sd_dev0 + lambda * z_mean

  if (
    any(!is.finite(c(z_mean, log_var_dev))) ||
      any(log_var_dev < -700 | log_var_dev > 700)
  ) {
    return(1e100)
  }

  log_f_mean <- dnorm(r_mean, 0, exp(log_sd_mean), log = TRUE)
  log_f_dev <- -0.5 * (
    log(2 * pi) + log_var_dev + r_dev^2 * exp(-log_var_dev)
  )

  # |d(M_Y, D_Y) / d(Y_1, Y_2)| = 1/2 for every dyad.
  nll <- -sum(log_f_mean + log_f_dev) + nrow(dat) * log(2)
  if (is.finite(nll)) nll else 1e100
}

fit_dim_ls <- function(dat, extended = TRUE) {
  needed <- c("m_x", "d_x", "m_y", "d_y")
  if (!all(needed %in% names(dat))) {
    stop("`dat` must contain m_x, d_x, m_y, and d_y.", call. = FALSE)
  }

  complete_numeric <- vapply(
    dat[needed],
    function(x) is.numeric(x) && all(is.finite(x)),
    logical(1)
  )
  if (!all(complete_numeric) || nrow(dat) <= 6L) {
    stop("Complete finite numeric dyad scores are required.", call. = FALSE)
  }

  mean_start <- lm.fit(cbind(1, dat$m_x), dat$m_y)
  dev_start <- lm.fit(matrix(dat$d_x, ncol = 1), dat$d_y)
  sd_mean_start <- sqrt(mean(mean_start$residuals^2))
  sd_dev_start <- sqrt(mean(dev_start$residuals^2))

  if (
    anyNA(c(mean_start$coefficients, dev_start$coefficients)) ||
      min(sd_mean_start, sd_dev_start) < 1e-8
  ) {
    stop("Rank deficiency or near-zero residual scale.", call. = FALSE)
  }

  start_common <- c(
    b0 = unname(mean_start$coefficients[1]),
    b_between = unname(mean_start$coefficients[2]),
    b_within = unname(dev_start$coefficients[1]),
    log_sd_mean = log(sd_mean_start),
    log_sd_dev0 = log(sd_dev_start)
  )

  starts <- if (extended) {
    lapply(c(-0.75, 0, 0.75), function(x) {
      c(start_common, lambda = x)
    })
  } else {
    list(start_common)
  }

  lower <- if (extended) {
    c(rep(-Inf, 3), -10, -10, -6)
  } else {
    c(rep(-Inf, 3), -10, -10)
  }
  upper <- if (extended) {
    c(rep(Inf, 3), 10, 10, 6)
  } else {
    c(rep(Inf, 3), 10, 10)
  }

  fits <- lapply(starts, function(start) {
    nlminb(
      start,
      dim_ls_nll,
      dat = dat,
      extended = extended,
      lower = lower,
      upper = upper,
      control = list(
        eval.max = 3000,
        iter.max = 2000,
        rel.tol = 1e-10,
        x.tol = 1e-8
      )
    )
  })

  converged <- vapply(
    fits,
    function(x) x$convergence == 0 && is.finite(x$objective),
    logical(1)
  )
  if (!any(converged)) {
    stop("No multistart optimization converged.", call. = FALSE)
  }
  candidates <- fits[converged]
  best <- candidates[[which.min(vapply(
    candidates,
    `[[`,
    numeric(1),
    "objective"
  ))]]

  at_bound <- (
    is.finite(lower) & abs(best$par - lower) < 1e-6
  ) | (
    is.finite(upper) & abs(best$par - upper) < 1e-6
  )
  if (any(at_bound)) {
    stop("The optimum reached an imposed parameter bound.", call. = FALSE)
  }

  hessian <- tryCatch(
    optimHess(
      best$par,
      dim_ls_nll,
      dat = dat,
      extended = extended
    ),
    error = function(error) {
      stop("Could not compute the observed Hessian.", call. = FALSE)
    }
  )

  if (any(!is.finite(hessian))) {
    stop("The observed Hessian contains non-finite values.", call. = FALSE)
  }
  hessian <- (hessian + t(hessian)) / 2
  eigenvalues <- tryCatch(
    eigen(hessian, symmetric = TRUE, only.values = TRUE)$values,
    error = function(error) {
      stop("Could not decompose the observed Hessian.", call. = FALSE)
    }
  )
  if (
    min(eigenvalues) <= sqrt(.Machine$double.eps) * max(eigenvalues)
  ) {
    stop("The observed Hessian is not positive definite.", call. = FALSE)
  }
  vcov <- tryCatch(
    chol2inv(chol(hessian)),
    error = function(error) {
      stop("Could not invert the observed Hessian.", call. = FALSE)
    }
  )
  dimnames(vcov) <- list(names(best$par), names(best$par))

  list(
    par = best$par,
    vcov = vcov,
    nll = best$objective,
    logLik = -best$objective,
    AIC = 2 * best$objective + 2 * length(best$par),
    convergence = best$convergence,
    message = best$message,
    hessian_eigenvalues = eigenvalues,
    multistart_objectives = vapply(fits, `[[`, numeric(1), "objective")
  )
}
```

</details>

The actual model calls are deliberately short:

```{r fit-models}
fit_standard <- fit_dim_ls(dat, extended = FALSE)
fit_extended <- fit_dim_ls(dat, extended = TRUE)
```

In this run, both models converged and the extended optimum passed the local
curvature check.

```{r prepare-model-results, include=FALSE}
se_standard <- sqrt(diag(fit_standard$vcov))
se_extended <- sqrt(diag(fit_extended$vcov))

estimate_row <- function(
    parameter, truth, standard, extended, standard_se, extended_se) {
  data.frame(
    Parameter = parameter,
    Truth = truth,
    `Standard DIM` = standard,
    `Standard SE` = standard_se,
    `Extended DIM` = extended,
    `Extended SE` = extended_se,
    check.names = FALSE
  )
}

truth_marginal_sd_dev <- simulation_truth$sigma_dev0 * exp(
  simulation_truth$lambda^2 / 4
)
extended_baseline_sd_dev <- exp(
  fit_extended$par[["log_sd_dev0"]]
)
extended_marginal_sd_dev <- extended_baseline_sd_dev * exp(
  fit_extended$par[["lambda"]]^2 / 4
)
extended_marginal_gradient <- c(
  log_sd_dev0 = extended_marginal_sd_dev,
  lambda = extended_marginal_sd_dev * fit_extended$par[["lambda"]] / 2
)
extended_marginal_vcov <- fit_extended$vcov[
  names(extended_marginal_gradient),
  names(extended_marginal_gradient),
  drop = FALSE
]
extended_marginal_se_dev <- sqrt(drop(
  t(extended_marginal_gradient) %*%
    extended_marginal_vcov %*%
    extended_marginal_gradient
))

model_results <- rbind(
  estimate_row(
    "Between-dyad effect (beta_B)", simulation_truth$beta_between,
    fit_standard$par[["b_between"]], fit_extended$par[["b_between"]],
    se_standard[["b_between"]], se_extended[["b_between"]]
  ),
  estimate_row(
    "Within-dyad effect (beta_W)", simulation_truth$beta_within,
    fit_standard$par[["b_within"]], fit_extended$par[["b_within"]],
    se_standard[["b_within"]], se_extended[["b_within"]]
  ),
  estimate_row(
    "Elevator residual SD (sigma_M)", simulation_truth$sigma_mean,
    exp(fit_standard$par[["log_sd_mean"]]),
    exp(fit_extended$par[["log_sd_mean"]]),
    exp(fit_standard$par[["log_sd_mean"]]) * se_standard[["log_sd_mean"]],
    exp(fit_extended$par[["log_sd_mean"]]) * se_extended[["log_sd_mean"]]
  ),
  estimate_row(
    "Baseline seesaw residual SD at z_M = 0 (sigma_D0)",
    simulation_truth$sigma_dev0,
    NA_real_,
    extended_baseline_sd_dev,
    NA_real_,
    exp(fit_extended$par[["log_sd_dev0"]]) * se_extended[["log_sd_dev0"]]
  ),
  estimate_row(
    "Marginal seesaw residual SD (sqrt(E[v_D]))",
    truth_marginal_sd_dev,
    exp(fit_standard$par[["log_sd_dev0"]]),
    extended_marginal_sd_dev,
    exp(fit_standard$par[["log_sd_dev0"]]) * se_standard[["log_sd_dev0"]],
    extended_marginal_se_dev
  ),
  estimate_row(
    "Log-variance slope (lambda)", simulation_truth$lambda,
    NA_real_, fit_extended$par[["lambda"]],
    NA_real_, se_extended[["lambda"]]
  )
)

comparison <- data.frame(
  Model = c("Standard DIM", "Extended DIM"),
  Parameters = c(5, 6),
  `Log likelihood` = c(fit_standard$logLik, fit_extended$logLik),
  AIC = c(fit_standard$AIC, fit_extended$AIC),
  check.names = FALSE
)

lambda_hat <- fit_extended$par[["lambda"]]
lambda_se <- se_extended[["lambda"]]
lambda_ci <- lambda_hat + c(-1, 1) * qnorm(0.975) * lambda_se
variance_multiplier <- exp(lambda_hat)
sd_multiplier <- exp(lambda_hat / 2)

lr <- 2 * (fit_extended$logLik - fit_standard$logLik)
lr_p <- pchisq(lr, df = 1, lower.tail = FALSE)

residual_components <- function(fit, dat, extended = TRUE) {
  par <- fit$par
  r_mean <- dat$m_y - par[["b0"]] - par[["b_between"]] * dat$m_x
  r_dev <- dat$d_y - par[["b_within"]] * dat$d_x
  sd_mean <- exp(par[["log_sd_mean"]])
  z_mean <- r_mean / sd_mean
  lambda <- if (extended) par[["lambda"]] else 0
  sd_dev <- exp(par[["log_sd_dev0"]] + lambda * z_mean / 2)

  data.frame(
    r_mean = r_mean,
    r_dev = r_dev,
    z_mean = z_mean,
    sd_dev = sd_dev,
    z_dev = r_dev / sd_dev
  )
}

res_standard <- residual_components(fit_standard, dat, extended = FALSE)
res_extended <- residual_components(fit_extended, dat, extended = TRUE)
```

```{r parameter-table, echo=FALSE}
format_estimate_se <- function(estimate, se) {
  if (!is.finite(estimate)) {
    return("—")
  }
  if (!is.finite(se)) {
    return(fmt(estimate, 3))
  }
  paste0(fmt(estimate, 3), " (", fmt(se, 3), ")")
}

display_results <- data.frame(
  Parameter = model_results$Parameter,
  Truth = vapply(model_results$Truth, fmt, character(1), digits = 3),
  `Standard DIM estimate (SE)` = mapply(
    format_estimate_se,
    model_results[["Standard DIM"]],
    model_results[["Standard SE"]]
  ),
  `Extended DIM estimate (SE)` = mapply(
    format_estimate_se,
    model_results[["Extended DIM"]],
    model_results[["Extended SE"]]
  ),
  check.names = FALSE
)
display_results[
  display_results$Parameter == "Log-variance slope (lambda)",
  "Standard DIM estimate (SE)"
] <- "0 (fixed)"

knitr::kable(
  display_results,
  align = c("l", "r", "r", "r"),
  caption = paste(
    "True and fitted parameters (model-based SEs).",
    "The standard DIM is misspecified by design and fixes lambda to zero."
  )
)
```

The extended estimate $\hat\beta_B=`r fmt(fit_extended$par[["b_between"]])`$
means that a one-unit higher dyad-average $X$ is associated with a
`r fmt(fit_extended$par[["b_between"]])`-unit higher expected dyad-average
$Y$. The estimate
$\hat\beta_W=`r fmt(fit_extended$par[["b_within"]])`$ means that a one-unit
larger member difference in $X$ is associated with a
`r fmt(fit_extended$par[["b_within"]])`-unit larger expected member difference
in $Y$ in the same direction, holding the dyad-average $X$ constant.

The mean effects are recovered well by both models. This illustrates an
important limit on the claim: a missed scale relationship does **not**
automatically bias the between- or within-dyad slopes when the conditional mean
is correct and the predictors are exogenous.

The standard DIM instead compresses the changing seesaw spread into one
marginal root-mean-square standard deviation,
$\sqrt{E\{v_D(z_M)\}}$. It has no separate baseline $\sigma_{D0}$ at
$z_M=0$. The extended model estimates both the baseline and the implied
marginal spread, so the table keeps those quantities on separate rows.

```{r model-comparison, echo=FALSE}
comparison_display <- comparison
comparison_display[c("Log likelihood", "AIC")] <- lapply(
  comparison_display[c("Log likelihood", "AIC")],
  round,
  digits = 1
)
knitr::kable(comparison_display, align = c("l", "r", "r", "r"))
```

Here the extended model lowers AIC by
`r fmt(fit_standard$AIC - fit_extended$AIC, 1)`. The nested likelihood-ratio
statistic is `r fmt(lr, 1)` on one degree of freedom
($p$ `r if (lr_p < 0.001) "< .001" else paste0("= ", fmt_p(lr_p))`). The fit
comparison is useful in this controlled simulation, but a substantive analysis
should use profile likelihood or a parametric bootstrap rather than relying on
one Wald or asymptotic likelihood-ratio test.

## What the standard DIM misses

The next figure uses **fitted standard-DIM residuals**, not the known simulation
errors. This is how the problem could first appear in a real analysis.

```{r residual-plot-helpers, include=FALSE}
equal_count_sd <- function(x, residual, groups = 10) {
  breaks <- unique(quantile(
    x,
    probs = seq(0, 1, length.out = groups + 1),
    na.rm = TRUE
  ))
  group <- cut(x, breaks = breaks, include.lowest = TRUE, labels = FALSE)

  data.frame(
    x = as.numeric(tapply(x, group, mean)),
    sd = as.numeric(tapply(residual, group, stats::sd)),
    n = as.numeric(table(group))
  )
}

draw_plot_frame <- function(xlim, ylim, xlab, ylab, main) {
  plot(
    NA,
    xlim = xlim,
    ylim = ylim,
    xlab = xlab,
    ylab = ylab,
    main = main,
    bty = "n",
    col.axis = plot_colours$muted,
    col.lab = plot_colours$ink,
    col.main = plot_colours$ink,
    las = 1
  )
  abline(
    v = pretty(xlim),
    h = pretty(ylim),
    col = plot_colours$grid,
    lwd = 0.8
  )
  box(col = plot_colours$grid)
}
```

```{r standard-misses, echo=FALSE, fig.width=9.5, fig.height=4.3, out.width="100%", fig.cap="The signed residual correlation is near zero, yet the seesaw spread increases with the elevator residual. The standard DIM captures the center but misses the changing residual spread.", fig.alt="Two residual diagnostic panels. The left panel shows a fan-shaped cloud of signed seesaw residuals around zero as the standardized elevator residual increases, despite near-zero signed correlation. The right panel shows binned seesaw residual standard deviations rising while the standard DIM predicts one horizontal constant."}
old_par <- par(
  mfrow = c(1, 2),
  mar = c(4.5, 4.7, 3.2, 1),
  oma = c(0, 0, 0, 0),
  cex = 0.92,
  family = "sans"
)

x_limit <- as.numeric(quantile(res_standard$z_mean, c(0.005, 0.995)))
y_limit <- max(abs(quantile(res_standard$r_dev, c(0.003, 0.997)))) * c(-1, 1)
signed_residual_correlation <- cor(
  res_standard$r_mean,
  res_standard$r_dev
)
if (abs(signed_residual_correlation) < 0.005) {
  signed_residual_correlation <- 0
}

draw_plot_frame(
  x_limit,
  y_limit,
  expression(hat(z)[M]),
  expression(hat(r)[D]),
  "Signed seesaw residual"
)
abline(h = 0, col = plot_colours$muted, lwd = 1.2)
abline(
  h = c(-1, 1) * 1.96 * exp(fit_standard$par[["log_sd_dev0"]]),
  col = plot_colours$standard,
  lwd = 2,
  lty = 2
)
points(
  res_standard$z_mean,
  res_standard$r_dev,
  pch = 16,
  cex = 0.55,
  col = transparent(plot_colours$residual_scale, 0.26)
)
legend(
  "topleft",
  legend = c(
    paste0(
      "signed r = ",
      fmt(signed_residual_correlation, 2)
    ),
    "standard 95% limits"
  ),
  lty = c(NA, 2),
  pch = c(NA, NA),
  col = c(plot_colours$ink, plot_colours$standard),
  bty = "n",
  text.col = plot_colours$ink,
  lwd = c(NA, 2),
  cex = 0.88
)

binned_standard <- equal_count_sd(
  res_standard$z_mean,
  res_standard$r_dev
)
y_sd_limit <- c(0, max(binned_standard$sd) * 1.18)
draw_plot_frame(
  x_limit,
  y_sd_limit,
  expression(hat(z)[M]),
  "Seesaw residual SD",
  "Spread within equal-count bins"
)
abline(
  h = exp(fit_standard$par[["log_sd_dev0"]]),
  col = plot_colours$standard,
  lwd = 2.5,
  lty = 2
)
lines(
  binned_standard$x,
  binned_standard$sd,
  col = transparent(plot_colours$residual_scale, 0.55),
  lwd = 1.5
)
points(
  binned_standard$x,
  binned_standard$sd,
  pch = 21,
  bg = plot_colours$residual_scale,
  col = "white",
  cex = 1.25,
  lwd = 1.2
)
legend(
  "topleft",
  legend = c("empirical binned SD", "standard DIM"),
  pch = c(21, NA),
  pt.bg = c(plot_colours$residual_scale, NA),
  col = c(plot_colours$residual_scale, plot_colours$standard),
  lty = c(NA, 2),
  lwd = c(NA, 2.5),
  bty = "n",
  cex = 0.88
)

par(old_par)
```

> Signed mean–difference correlation asks “Is elevator position linearly
> associated with a preferred seat difference?” Conditional sign symmetry
> rules out a preferred arbitrary seat. The scale pattern asks “How far apart
> are the seats?” These are different questions.

The binned residual plot is a useful diagnostic, not a valid two-stage
estimator. Regressing $|\hat r_D|$ on $\hat r_M$ would treat estimated residuals
as observed without error. The joint model instead estimates the mean and scale
parts together.

## Interpret the extended model

The fitted scale slope is
$\hat\lambda=`r fmt(lambda_hat)`$ (Wald 95% CI
[`r fmt(lambda_ci[1])`, `r fmt(lambda_ci[2])`]). For each one-SD higher
elevator residual, the fitted seesaw variance is multiplied by

$$
\exp(\hat\lambda)=`r fmt(variance_multiplier)`,
$$

and its standard deviation is multiplied by

$$
\exp(\left.\hat\lambda\middle/2\right.)=`r fmt(sd_multiplier)`.
$$

This is not evidence that member 1 becomes higher than member 2. It means that
the **gap** becomes more variable, with either sign still equally likely.

```{r prepare-scale-table, include=FALSE}
scale_positions <- c(-1, 0, 1)
scale_table <- data.frame(
  `Elevator residual z_M` = scale_positions,
  `Standard constant SD` = rep(
    exp(fit_standard$par[["log_sd_dev0"]]),
    length(scale_positions)
  ),
  `Extended seesaw SD` = exp(
    fit_extended$par[["log_sd_dev0"]] +
      fit_extended$par[["lambda"]] * scale_positions / 2
  ),
  `Full member-gap SD` = 2 * exp(
    fit_extended$par[["log_sd_dev0"]] +
      fit_extended$par[["lambda"]] * scale_positions / 2
  ),
  check.names = FALSE
)
```

```{r scale-table, echo=FALSE}
scale_display <- scale_table
scale_display[-1] <- lapply(scale_display[-1], round, digits = 2)
knitr::kable(
  scale_display,
  align = c("r", "r", "r", "r"),
  caption = paste(
    "Fitted residual spread at three elevator positions.",
    "The full member gap is e1 - e2 = 2 r_D."
  )
)
```

The standard model must use the same seesaw SD everywhere. The extended model
estimates a much narrower residual member gap at $z_M=-1$ than at $z_M=1$.

```{r extended-scale-figure, echo=FALSE, fig.width=9.5, fig.height=4.3, out.width="100%", fig.cap="Left: the fitted scale function tracks the changing seesaw spread. Right: after division by the fitted conditional SD, the standardized residual spread is approximately constant.", fig.alt="Two model-checking panels. The left shows empirical binned seesaw spread increasing with elevator position, tracked by the fitted extended DIM and the simulation truth while the standard DIM remains horizontal. The right shows that binned standardized residual spread is approximately one after applying the extended model."}
old_par <- par(
  mfrow = c(1, 2),
  mar = c(4.5, 4.7, 3.2, 1),
  oma = c(0, 0, 0, 0),
  cex = 0.92,
  family = "sans"
)

x_grid <- seq(-3, 3, length.out = 300)
binned_extended <- equal_count_sd(
  res_extended$z_mean,
  res_extended$r_dev
)
extended_curve <- exp(
  fit_extended$par[["log_sd_dev0"]] + lambda_hat * x_grid / 2
)
truth_curve <- simulation_truth$sigma_dev0 * exp(
  simulation_truth$lambda * x_grid / 2
)
standard_curve <- exp(fit_standard$par[["log_sd_dev0"]])
y_max <- max(c(binned_extended$sd, extended_curve, truth_curve)) * 1.08

draw_plot_frame(
  c(-3, 3),
  c(0, y_max),
  expression(hat(z)[M]),
  "Seesaw residual SD",
  "Estimated scale function"
)
abline(
  h = standard_curve,
  col = plot_colours$standard,
  lwd = 2.4,
  lty = 2
)
lines(
  x_grid,
  truth_curve,
  col = plot_colours$truth,
  lwd = 1.7,
  lty = 3
)
lines(
  x_grid,
  extended_curve,
  col = plot_colours$residual_2_text,
  lwd = 3
)
points(
  binned_extended$x,
  binned_extended$sd,
  pch = 21,
  bg = plot_colours$residual_scale,
  col = "white",
  cex = 1.2,
  lwd = 1.2
)
legend(
  "topleft",
  legend = c(
    "empirical binned SD",
    "extended DIM",
    "simulation truth",
    "standard DIM"
  ),
  pch = c(21, NA, NA, NA),
  pt.bg = c(plot_colours$residual_scale, NA, NA, NA),
  col = c(
    plot_colours$residual_scale,
    plot_colours$residual_2_text,
    plot_colours$truth,
    plot_colours$standard
  ),
  lty = c(NA, 1, 3, 2),
  lwd = c(NA, 3, 1.7, 2.4),
  bty = "n",
  cex = 0.84
)

binned_z_dev <- equal_count_sd(
  res_extended$z_mean,
  res_extended$z_dev
)
draw_plot_frame(
  c(-3, 3),
  c(0, max(1.35, max(binned_z_dev$sd) * 1.12)),
  expression(hat(z)[M]),
  expression("SD of fitted " * hat(z)[D]),
  "After modeling the spread"
)
abline(h = 1, col = plot_colours$residual_2_text, lwd = 2.5)
lines(
  binned_z_dev$x,
  binned_z_dev$sd,
  col = transparent(plot_colours$residual_scale, 0.55),
  lwd = 1.5
)
points(
  binned_z_dev$x,
  binned_z_dev$sd,
  pch = 21,
  bg = plot_colours$residual_scale,
  col = "white",
  cex = 1.2,
  lwd = 1.2
)
legend(
  "topleft",
  legend = c("binned SD", "expected value"),
  pch = c(21, NA),
  pt.bg = c(plot_colours$residual_scale, NA),
  col = c(plot_colours$residual_scale, plot_colours$residual_2_text),
  lty = c(NA, 1),
  lwd = c(NA, 2.5),
  bty = "n",
  cex = 0.88
)

par(old_par)
```

The right panel is a direct model check: after dividing each seesaw residual by
its fitted conditional standard deviation, the residual spread should no
longer change systematically with elevator position. A real analysis should
also inspect Q–Q plots and influential dyads; the scale slope is particularly
sensitive to unusual squared residuals.

### The label-swap check

An exchangeable model must give the same likelihood after independently
swapping the two arbitrary seats in any dyad. A swap reverses both $D_X$ and
$D_Y$ for that dyad.

```{r label-swap-check}
set.seed(2468)
swap_sign <- sample(c(-1, 1), nrow(dat), replace = TRUE)
dat_swapped <- dat[c("m_x", "d_x", "m_y", "d_y")]
dat_swapped$d_x <- swap_sign * dat$d_x
dat_swapped$d_y <- swap_sign * dat$d_y

unname(
  dim_ls_nll(fit_extended$par, dat_swapped, extended = TRUE) -
    dim_ls_nll(fit_extended$par, dat, extended = TRUE)
)
```

The difference is exactly zero apart from possible machine rounding. The
extension changes the size of $r_D$, never its preferred sign.

## Translate the fitted result to APIM language

The DIM does not contain information absent from the APIM. It rotates the same
two-member data into coordinates that make exchangeability easier to see.

### Fixed effects

For member $i$ with partner $k$,

$$
E(Y_i\mid X_i,X_k)
=\beta_0
+\frac{\beta_B+\beta_W}{2}X_i
+\frac{\beta_B-\beta_W}{2}X_k.
$$

Thus the exchangeable actor and partner effects are

$$
a=\frac{\beta_B+\beta_W}{2},
\qquad
p=\frac{\beta_B-\beta_W}{2}.
$$

```{r apim-fixed-table, echo=FALSE}
actor_hat <- (
  fit_extended$par[["b_between"]] +
    fit_extended$par[["b_within"]]
) / 2
partner_hat <- (
  fit_extended$par[["b_between"]] -
    fit_extended$par[["b_within"]]
) / 2

apim_fixed <- data.frame(
  Effect = c("Actor effect", "Partner effect"),
  Formula = c("(beta_B + beta_W) / 2", "(beta_B - beta_W) / 2"),
  Truth = round(c(
    (simulation_truth$beta_between + simulation_truth$beta_within) / 2,
    (simulation_truth$beta_between - simulation_truth$beta_within) / 2
  ), 3),
  Estimate = round(c(actor_hat, partner_hat), 3)
)
knitr::kable(apim_fixed, align = c("l", "l", "r", "r"))
```

The location–scale extension does not change this mapping. It changes the
residual distribution around the member-level predictions. In member-level
language, the fitted actor effect is `r fmt(actor_hat)` and the fitted partner
effect is `r fmt(partner_hat)`.

### Residual covariance

Recall that

$$
e_1=r_M+r_D,
\qquad
e_2=r_M-r_D.
$$

With constant component variances $v_M$ and $v_D$ and
$\operatorname{Cov}(r_M,r_D)=0$, the expanded exchangeable APIM residual
covariance matrix is

$$
\boldsymbol{\Sigma}_e
=
\begin{pmatrix}
v_M+v_D & v_M-v_D\\
v_M-v_D & v_M+v_D
\end{pmatrix}.
$$

The partner residual correlation is therefore

$$
\rho_e=\frac{v_M-v_D}{v_M+v_D}.
$$

In the extended model, let

$$
\bar v_D=E\!\left[v_D(z_M)\right]
$$

denote the average seesaw variance. Because $z_M\sim\mathcal{N}(0,1)$ here,

$$
\bar v_D
=\sigma_{D0}^2\exp\!\left(\frac{\lambda^2}{2}\right).
$$

A conventional Gaussian exchangeable APIM with one constant covariance matrix
can match the extended model's **marginal** covariance matrix,

$$
\boldsymbol{\Sigma}_e
=
\begin{pmatrix}
\sigma_M^2+\bar v_D & \sigma_M^2-\bar v_D\\
\sigma_M^2-\bar v_D & \sigma_M^2+\bar v_D
\end{pmatrix}.
$$

But that covariance matrix alone cannot identify $\lambda$. The extended model
is identified by the association between elevator position and the squared or
absolute seesaw residual—that is, by the **changing shape** of the residual
cloud.

In APIM residuals, the extension can be written directly as

$$
\log
\operatorname{Var}
\left(
\frac{e_1-e_2}{2}
\ \middle|\
\frac{e_1+e_2}{2}=m
\right)
=2\log(\sigma_{D0})+\lambda\frac{m}{\sigma_M}.
$$

In plain language: the residual **average** is associated with the residual
**gap magnitude**.

```{r prepare-apim-cloud, include=FALSE}
sd_mean_hat <- exp(fit_extended$par[["log_sd_mean"]])
sd_dev0_hat <- exp(fit_extended$par[["log_sd_dev0"]])
v_dev_average <- sd_dev0_hat^2 * exp(lambda_hat^2 / 2)
member_variance <- sd_mean_hat^2 + v_dev_average
member_covariance <- sd_mean_hat^2 - v_dev_average
member_rho <- member_covariance / member_variance

# Match the conventional panel to the extended panel's marginal seesaw
# variance. Reusing the same latent standard-normal draws isolates shape.
r_mean_cloud <- sd_mean_hat * dat$z_mean
r_dev_standard_cloud <- sqrt(v_dev_average) * dat$z_dev
r_dev_extended_cloud <- sd_dev0_hat * exp(
  lambda_hat * dat$z_mean / 2
) * dat$z_dev

make_member_residuals <- function(r_mean, r_dev) {
  data.frame(
    e1 = r_mean + r_dev,
    e2 = r_mean - r_dev
  )
}

cloud_standard <- make_member_residuals(
  r_mean_cloud,
  r_dev_standard_cloud
)
cloud_extended <- make_member_residuals(
  r_mean_cloud,
  r_dev_extended_cloud
)
```

For this fitted model, the marginal residual variance of either member is
`r fmt(member_variance)`, the partner residual covariance is
`r fmt(member_covariance)`, and the resulting marginal residual correlation is
`r fmt(member_rho)`. These are average summaries; they do not describe how the
cloud's width changes internally.

```{r apim-cloud, echo=FALSE, fig.width=9.5, fig.height=4.7, out.width="100%", fig.cap="Both APIM residual clouds have the same model-implied marginal variance and correlation. Only the extended model captures how the anti-diagonal seesaw width changes along the diagonal elevator.", fig.alt="Two scatterplots of member 1 against member 2 residuals with identical axes. The conventional Gaussian exchangeable APIM forms one roughly elliptical cloud. The extended APIM has the same average covariance but its anti-diagonal width fans out along the diagonal elevator direction."}
old_par <- par(
  mfrow = c(1, 2),
  mar = c(4.5, 4.6, 3.2, 1),
  oma = c(0, 0, 0, 0),
  cex = 0.92,
  family = "sans",
  pty = "s"
)

all_cloud_values <- c(
  cloud_standard$e1,
  cloud_standard$e2,
  cloud_extended$e1,
  cloud_extended$e2
)
cloud_limit <- max(abs(quantile(all_cloud_values, c(0.002, 0.998))))
cloud_limit <- c(-cloud_limit, cloud_limit)

draw_cloud <- function(cloud, title) {
  draw_plot_frame(
    cloud_limit,
    cloud_limit,
    expression(e[1]),
    expression(e[2]),
    title
  )
  abline(a = 0, b = 1, col = plot_colours$residual_2, lwd = 1.8)
  abline(a = 0, b = -1, col = plot_colours$residual_scale, lwd = 1.8, lty = 2)
  points(
    cloud$e1,
    cloud$e2,
    pch = 16,
    cex = 0.5,
    col = transparent(plot_colours$ink, 0.19)
  )
  text(
    cloud_limit[2] * 0.62,
    cloud_limit[2] * 0.82,
    "elevator",
    col = plot_colours$residual_2_text,
    cex = 0.86,
    font = 2
  )
  text(
    cloud_limit[2] * 0.62,
    cloud_limit[1] * 0.82,
    "seesaw",
    col = plot_colours$residual_scale,
    cex = 0.86,
    font = 2
  )
  legend(
    "topleft",
    legend = paste0("model-implied marginal rho = ", fmt(member_rho, 2)),
    bty = "n",
    text.col = plot_colours$ink,
    cex = 0.85
  )
}

draw_cloud(cloud_standard, "Conventional Gaussian APIM")
draw_cloud(cloud_extended, "Location-scale APIM")

par(old_par)
```

> A conventional Gaussian exchangeable APIM with one constant covariance
> matrix can match the overall covariance ellipse while missing the funnel
> inside it. The DIM rotation makes that hidden structure visible.

The two plots use the same $\sigma_M^2$ and the same marginal $\bar v_D$, so
their model-implied member residual variance, covariance, and correlation are
the same. The standard panel has one constant anti-diagonal width. The extended
panel redistributes that same average seesaw variation across the elevator.

## Optional extensions

The residual-driven model is only one symmetry-compatible extension. Two
related ideas answer different questions.

### Let an observed predictor mean explain seesaw variability

An easier distributional model uses an observed, standardized dyad-level
predictor $z_X$:

$$
\log v_D(z_X)=2\log(\sigma_{D0})+\lambda_Xz_X.
$$

$\lambda_X$ asks whether dyads with a higher average $X$ show larger or smaller
unexplained outcome differences, without predicting which member is higher.
Suppose $r_M\mid z_X\sim\mathcal{N}(0,v_M)$ with constant $v_M$,
$r_D\mid z_X\sim\mathcal{N}(0,v_D(z_X))$, and the two components are
conditionally independent. The member-level conditional covariance matrix is

$$
\boldsymbol{\Sigma}_e(z_X)
=
\begin{pmatrix}
v_M+v_D(z_X) & v_M-v_D(z_X)\\
v_M-v_D(z_X) & v_M+v_D(z_X)
\end{pmatrix},
$$

with conditional residual correlation

$$
\rho_e(z_X)
=\frac{v_M-v_D(z_X)}{v_M+v_D(z_X)}.
$$

This varying-correlation expression belongs to the **observed-predictor**
version. In the residual-driven model, conditioning on $r_M$ fixes the
elevator coordinate, so an ordinary conditional member correlation is not the
right summary.

### Directionless nonlinear cross-paths

Exchangeability also permits nonlinear conditional-mean relationships that
have no preferred member direction:

$$
M_Y=\beta_0+\beta_BM_X+qD_X^2+r_M,
$$

$$
D_Y=(\beta_W+hM_X)D_X+r_D.
$$

| Exchangeable term | Substantive question |
|---|---|
| $D_X^2\rightarrow M_Y$ | Do dyads who differ more on $X$ have a different average $Y$? |
| $M_XD_X\rightarrow D_Y$ | Does the within-dyad $X$–$Y$ association depend on average $X$? |

$D_X^2$ is unchanged by a member swap, whereas $M_XD_X$ reverses sign together
with $D_Y$. Both terms are therefore exchangeable. They extend the conditional
mean rather than the residual variance and are not needed for the fitted
location–scale model above.

<details>
<summary><strong>Show the nonlinear APIM back-transformation</strong></summary>

In APIM coordinates, their full expansion is

$$
\begin{aligned}
E(Y_i\mid X_i,X_k)
&=\beta_0
+\frac{\beta_B+\beta_W}{2}X_i
+\frac{\beta_B-\beta_W}{2}X_k \\
&\quad
+\frac{q+h}{4}X_i^2
+\frac{q-h}{4}X_k^2
-\frac{q}{2}X_iX_k.
\end{aligned}
$$

Thus a discrepancy term and a mean-by-deviation moderation in DIM coordinates
become actor-squared, partner-squared, and actor-by-partner terms in APIM
coordinates. The model remains an exchangeable APIM, but it is no longer the
standard **linear** APIM.

</details>

## What to report—and what not to claim

A substantive application should report at least:

- the between- and within-dyad effects, $\beta_B$ and $\beta_W$;
- the baseline seesaw SD $\sigma_{D0}$ and where $z_M=0$ lies;
- $\lambda$ and the variance multiplier $\exp(\lambda)$;
- predicted seesaw or full member-gap SDs at meaningful elevator positions;
- diagnostics for the elevator residual and the standardized seesaw residual;
- for a new implementation, verification that reversing arbitrary member
  labels leaves the likelihood unchanged.

Four cautions matter:

1. **Association is not temporal or causal.** The elevator and seesaw residuals
   are two coordinates computed from the same cross-sectional outcomes.
2. **A scale relationship does not by itself imply biased mean slopes.** Bias
   becomes a clearer concern when an omitted term also changes the conditional
   mean—for example, if $D_X^2$ predicts $M_Y$ and is associated with $M_X$.
3. **Bounded outcomes can create mechanical fanning.** Floor and ceiling effects
   can link a dyad's mean to its possible member gap even without a substantive
   scale process.
4. **One gap per dyad is data-hungry.** $\lambda$ is learned from the association
   between $r_M$ and $r_D^2$, making it sensitive to unusual dyads. A paper-ready
   implementation should use TMB or Stan, regularization where appropriate,
   profile or bootstrap inference, and simulation-based calibration.

Complete outcome pairs are required for the score-space likelihood used here.
Repeated observations require a different multilevel location–scale model.

## Take-away

The standard linear Gaussian DIM is the exchangeability-constrained DSM. Its
zero signed mean–difference residual covariance is a genuine consequence of
arbitrary member labels. One constant **conditional** difference variance,
however, is an additional homoscedastic Gaussian restriction—not a requirement
of exchangeability.

The location–scale DIM relaxes that restriction while preserving symmetry. In
APIM coordinates, the same model says that the residual average is associated
with the residual gap magnitude. A conventional Gaussian exchangeable APIM
with one constant covariance matrix can reproduce its average covariance but
not the changing residual-cloud shape that identifies $\lambda$.

> Exchangeability removes the direction of an arbitrary member difference. It
> does not require the magnitude of that difference to be constant.

For the more technical derivations and the developing paper plan, see
[`paper-idea-exchangeable-location-scale-dim-apim.md`](../paper-idea-exchangeable-location-scale-dim-apim.md).
