---
title: "Generalized Non-Gaussian Actor-Partner Interdependence Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalized Non-Gaussian Actor-Partner Interdependence Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dyadMLM)
```


## Dichotomous outcomes

## Ordinal outcomes

## Categorical outcomes

## Poisson and negative-binomial outcomes

`dyads_nbinom_cross` has the same dyadic structure as `dyads_cross`, but its
`conflict_count` outcome is a non-negative count with greater variance than a
Poisson model allows. The data were generated from a negative-binomial model
with a log link. Dyad compositions are selected below through
`keep_compositions`, so that `prepare_dyad_data()` performs the filtering at
the dyad level.

```{r nbinom-raw}
print(head(dyads_nbinom_cross))
hist(dyads_nbinom_cross$conflict_count, breaks = 20)
```

Validation and preparation work in the same way as for a Gaussian outcome
because the dyadic structure is unchanged. The interpretation of the fitted
model differs: fixed and random effects act on the log-mean scale, while the
observation-level negative-binomial variance remains part of the model.

### Distinguishable negative-binomial APIM

```{r prepare-nbinom-distinguishable}
nbinom_distinguishable_data <- dyadMLM::prepare_dyad_data(
  dyads_nbinom_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  predictors = stress,
  # dyads_nbinom_cross contains three compositions; retain `female-male` here.
  keep_compositions = "female-male",
  seed = 123
)

print(nbinom_distinguishable_data)
summary(nbinom_distinguishable_data)
```


```{r fit-nbinom-distinguishable, eval = FALSE}
nbinom_distinguishable_model <- glmmTMB::glmmTMB(
  conflict_count ~ 0 +
    # Role-specific intercepts
    .dy_is_female_x_male_female +
    .dy_is_female_x_male_male +

    # Role-specific actor effects of stress
    .dy_is_female_x_male_female:.dy_stress_actor +
    .dy_is_female_x_male_male:.dy_stress_actor +

    # Role-specific partner effects of stress
    .dy_is_female_x_male_female:.dy_stress_partner +
    .dy_is_female_x_male_male:.dy_stress_partner +

    # Shared latent dyad effect for stable non-independence
    (1 | coupleID),
  family = glmmTMB::nbinom2(link = "log"),
  data = nbinom_distinguishable_data
)

summary(nbinom_distinguishable_model)
```

This shared random intercept is deliberately simple. It can induce only
positive partner dependence on the latent log-mean scale; more flexible
role-specific covariance structures require enough repeated information to
support their additional parameters.

### Exchangeable negative-binomial APIM

Here we retain male-male dyads, which are exchangeable based on their
observed roles.

```{r prepare-nbinom-exchangeable}
nbinom_exchangeable_data <- dyadMLM::prepare_dyad_data(
  dyads_nbinom_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  predictors = stress,
  keep_compositions = "male-male",
  seed = 123
)

print(nbinom_exchangeable_data)
summary(nbinom_exchangeable_data)
```

```{r fit-nbinom-exchangeable, eval = FALSE}
nbinom_exchangeable_model <- glmmTMB::glmmTMB(
  conflict_count ~
    # Pooled intercept and actor and partner effects
    1 +
    .dy_stress_actor +
    .dy_stress_partner +

    # Exchangeable latent dyad covariance on the log-mean scale
    (1 | coupleID) +
    (0 + .dy_member_contrast_male_x_male_arbitrary | coupleID),
  family = glmmTMB::nbinom2(link = "log"),
  data = nbinom_exchangeable_data
)

summary(nbinom_exchangeable_model)
```
