---
title: "Generalized Non-Gaussian Actor-Partner Interdependence Models"
output: rmarkdown::html_vignette
bibliography: ../../vignettes/references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Generalized Non-Gaussian Actor-Partner Interdependence Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(dyadMLM)
```

## Scope

This development vignette covers two non-Gaussian APIMs fitted with
`glmmTMB`: a cross-sectional negative-binomial count model and a longitudinal
Tweedie model for an outcome with exact zeros and positive continuous values
[@brooksGlmmTMBBalancesSpeed2017]. Dichotomous, ordinal, and unordered
categorical outcomes require family-specific guidance and are deferred.

For a non-Gaussian response, a random effect is not an observed-scale residual.
Fixed and random effects act on the link scale, and the conditional response
distribution still has its own variance. This distinction is especially
important in the longitudinal Tweedie example below.


## Cross-sectional negative-binomial APIM

`dyads_nbinom_cross` has the same dyadic structure as `dyads_cross`, but its
`conflict_count` outcome is a non-negative count with greater variance than a
Poisson model allows. The data were generated from a negative-binomial model
with a log link.

```{r nbinom-raw}
print(head(dyads_nbinom_cross))
hist(dyads_nbinom_cross$conflict_count, breaks = 20)
```

The examples use `add_apim_gmc_predictors = TRUE`, which retains the raw APIM
columns and adds stress variables centered over all retained non-missing
values. Using the raw columns instead changes the intercept reference.

### Distinguishable negative-binomial APIM

```{r prepare-nbinom-distinguishable}
nbinom_distinguishable_data <- dyadMLM::prepare_dyad_data(
  dyads_nbinom_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  predictors = stress,
  add_apim_gmc_predictors = TRUE,
  keep_compositions = "female-male",
  seed = 123
)

print(nbinom_distinguishable_data)
```

```{r fit-nbinom-distinguishable, eval = FALSE}
nbinom_distinguishable_model <- glmmTMB::glmmTMB(
  conflict_count ~ 0 +
    # Role-specific intercepts
    .is_female +
    .is_male +

    # Role-specific actor effects of stress
    .is_female:.stress_gmc_actor +
    .is_male:.stress_gmc_actor +

    # Role-specific partner effects of stress
    .is_female:.stress_gmc_partner +
    .is_male:.stress_gmc_partner +

    # Shared latent dyad effect for stable non-independence
    (1 | coupleID),
  family = glmmTMB::nbinom2(link = "log"),
  data = nbinom_distinguishable_data
)

summary(nbinom_distinguishable_model)
```

The fixed coefficients are conditional log-mean effects. Exponentiating a
coefficient gives the corresponding conditional mean ratio. The shared random
intercept is deliberately simple: it can induce only positive partner
dependence on the latent log-mean scale.

### Exchangeable negative-binomial APIM

Here we retain male-male dyads, which are exchangeable based on their observed
roles.

```{r prepare-nbinom-exchangeable}
nbinom_exchangeable_data <- dyadMLM::prepare_dyad_data(
  dyads_nbinom_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  predictors = stress,
  add_apim_gmc_predictors = TRUE,
  keep_compositions = "male-male",
  seed = 123
)

print(nbinom_exchangeable_data)
```

```{r fit-nbinom-exchangeable, eval = FALSE}
nbinom_exchangeable_model <- glmmTMB::glmmTMB(
  conflict_count ~
    # Pooled intercept and actor and partner effects
    1 +
    .stress_gmc_actor +
    .stress_gmc_partner +

    # Exchangeable latent dyad covariance on the log-mean scale
    (1 | coupleID) +
    (0 + .member_contrast_arbitrary | coupleID),
  family = glmmTMB::nbinom2(link = "log"),
  data = nbinom_exchangeable_data
)

summary(nbinom_exchangeable_model)
```


## Longitudinal Tweedie APIM

For \(1 < p < 2\), the Tweedie family is a compound Poisson-gamma distribution:
it has a point mass at zero and a continuous, right-skewed positive part. We
write the conditional model as

\[
Y_{ijt} \mid b_{ij}, u_{ijt}
  \sim \operatorname{Tweedie}(\mu_{ijt}, \phi_{r(i)}, p),
\qquad
\operatorname{Var}(Y_{ijt}\mid b_{ij},u_{ijt})
  = \phi_{r(i)}\mu_{ijt}^{p},
\]

\[
\log(\mu_{ijt}) = x_{ijt}^{\mathsf T}\beta + b_{ij} + u_{ijt}.
\]

Here, \(b_{ij}\) represents stable member-specific heterogeneity within dyad
\(j\). The pair \((u_{1jt},u_{2jt})\) is a Gaussian dyad-occasion effect. It
represents same-occasion partner dependence on the log-mean scale; it is not an
observed-scale residual. The Tweedie dispersion \(\phi\) remains in the model.

Consequently, `dispformula = ~ 0` is not the Gaussian trick used elsewhere in
this package. For Tweedie models:

- `dispformula = ~ 1` estimates one common conditional dispersion;
- `dispformula = ~ 0 + .is_female + .is_male` estimates two positive,
  role-specific dispersions on `glmmTMB`'s log-dispersion scale. The `0` removes
  the formula intercept; it does **not** set dispersion to zero.

### What identifies latent occasion covariance and dispersion?

After integrating over one Gaussian latent layer with variance \(q\), let
\(m = E(Y)\) denote the marginal mean. Under a log link,

\[
\operatorname{Var}(Y)
  = \phi m^p \exp\{p(p-1)q/2\}
    + m^2\{\exp(q)-1\}.
\]

The two variance sources therefore have different mean-variance scaling. If
the two partners' latent occasion effects have covariance \(c\), then

\[
\operatorname{Cov}(Y_{1jt},Y_{2jt})
  = m_{1jt}m_{2jt}\{\exp(c)-1\}.
\]

Conditional Tweedie dispersion cannot by itself produce this cross-partner
covariance. Repeated complete partner occasions, variation in the conditional
mean, the positive outcome distribution, and the zero mass all help separate
the parameters.

Thus, both of the following are *generically* identified in an adequately
repeated design:

- distinguishable dyads with a two-member occasion covariance, two
  role-specific dispersions, and an estimated Tweedie power;
- exchangeable dyads with an exchangeable occasion covariance, one common
  dispersion, and an estimated Tweedie power.

This is not guaranteed by parameter counting. At the occasion layer alone, the
first model adds six nuisance parameters (three covariance, two dispersion, one
power), whereas the second adds four (two exchangeable covariance, one
dispersion, one power). Stable intercepts and random slopes add further
parameters. Counts describe burden, not whether distinct parameters enter the
likelihood differently.

An important exact exception occurs when every dyad has only one occasion: a
stable member-intercept covariance block and a dyad-occasion intercept
covariance block then have the same design, so only their covariance sum is
identified. With repeated occasions, the stable block also induces
across-occasion covariance and the occasion block does not.


## Reproducible Tweedie teaching data

No Tweedie outcome is shipped with the package. The next hidden chunk uses the
structure and stress predictor in `dyads_nbinom_ild` as a shell and generates a
compound Poisson-gamma outcome. The simulation includes a focal within-person
actor random slope, stable partner covariance, and same-occasion partner
covariance. This is a teaching simulation, not evidence that every design will
identify the same model.

```{r simulate-tweedie-data, include = FALSE}
simulate_tweedie <- function(mean, dispersion, power) {
  poisson_mean <-
    mean^(2 - power) / (dispersion * (2 - power))
  gamma_shape <- (2 - power) / (power - 1)
  gamma_scale <-
    dispersion * (power - 1) * mean^(power - 1)

  component_count <- stats::rpois(length(mean), poisson_mean)
  response <- numeric(length(mean))
  positive <- component_count > 0
  response[positive] <- stats::rgamma(
    sum(positive),
    shape = component_count[positive] * gamma_shape,
    scale = gamma_scale[positive]
  )
  response
}

set.seed(20260811)
tweedie_distinguishable_data <- dyadMLM::prepare_dyad_data(
  dyads_nbinom_ild,
  dyad = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = stress,
  model_types = "apim",
  keep_compositions = "female-male",
  include_arbitrary_member_contrast = TRUE,
  seed = 123
)

distinguishable_dyad_ids <-
  unique(tweedie_distinguishable_data$coupleID)
distinguishable_stable_sds <- c(0.35, 0.50, 0.12, 0.15)
distinguishable_stable_correlation <- matrix(
  c(
    1.00, 0.35, 0.20, 0.05,
    0.35, 1.00, 0.05, 0.20,
    0.20, 0.05, 1.00, 0.20,
    0.05, 0.20, 0.20, 1.00
  ),
  nrow = 4,
  byrow = TRUE
)
distinguishable_stable_covariance <-
  diag(distinguishable_stable_sds) %*%
  distinguishable_stable_correlation %*%
  diag(distinguishable_stable_sds)
distinguishable_stable_effects <- matrix(
  stats::rnorm(length(distinguishable_dyad_ids) * 4),
  ncol = 4
) %*% chol(distinguishable_stable_covariance)
distinguishable_stable_index <- match(
  tweedie_distinguishable_data$coupleID,
  distinguishable_dyad_ids
)

distinguishable_occasion_keys <- unique(
  tweedie_distinguishable_data[c("coupleID", "diaryday")]
)
distinguishable_occasion_sds <- c(0.25, 0.35)
distinguishable_occasion_correlation <- 0.35
distinguishable_occasion_covariance <-
  diag(distinguishable_occasion_sds) %*%
  matrix(
    c(
      1, distinguishable_occasion_correlation,
      distinguishable_occasion_correlation, 1
    ),
    nrow = 2
  ) %*%
  diag(distinguishable_occasion_sds)
distinguishable_occasion_effects <- matrix(
  stats::rnorm(nrow(distinguishable_occasion_keys) * 2),
  ncol = 2
) %*% chol(distinguishable_occasion_covariance)
distinguishable_occasion_index <- match(
  paste(
    tweedie_distinguishable_data$coupleID,
    tweedie_distinguishable_data$diaryday,
    sep = "-"
  ),
  paste(
    distinguishable_occasion_keys$coupleID,
    distinguishable_occasion_keys$diaryday,
    sep = "-"
  )
)

female <- tweedie_distinguishable_data$.is_female
male <- tweedie_distinguishable_data$.is_male
distinguishable_time_centered <-
  tweedie_distinguishable_data$diaryday -
  mean(tweedie_distinguishable_data$diaryday)

distinguishable_linear_predictor <-
  female * 1.00 + male * 1.20 +
  female * -0.010 * distinguishable_time_centered +
  male * -0.005 * distinguishable_time_centered +
  female * 0.25 * tweedie_distinguishable_data$.stress_cwp_actor +
  male * 0.20 * tweedie_distinguishable_data$.stress_cwp_actor +
  female * 0.10 * tweedie_distinguishable_data$.stress_cwp_partner +
  male * 0.08 * tweedie_distinguishable_data$.stress_cwp_partner +
  female * 0.18 * tweedie_distinguishable_data$.stress_cbp_actor +
  male * 0.15 * tweedie_distinguishable_data$.stress_cbp_actor +
  female * 0.08 * tweedie_distinguishable_data$.stress_cbp_partner +
  male * 0.06 * tweedie_distinguishable_data$.stress_cbp_partner +
  female * distinguishable_stable_effects[
    distinguishable_stable_index, 1
  ] +
  male * distinguishable_stable_effects[
    distinguishable_stable_index, 2
  ] +
  female * distinguishable_stable_effects[
    distinguishable_stable_index, 3
  ] * tweedie_distinguishable_data$.stress_cwp_actor +
  male * distinguishable_stable_effects[
    distinguishable_stable_index, 4
  ] * tweedie_distinguishable_data$.stress_cwp_actor +
  female * distinguishable_occasion_effects[
    distinguishable_occasion_index, 1
  ] +
  male * distinguishable_occasion_effects[
    distinguishable_occasion_index, 2
  ]

tweedie_distinguishable_data$time_centered <-
  distinguishable_time_centered
tweedie_distinguishable_data$tweedie_amount <- simulate_tweedie(
  mean = exp(distinguishable_linear_predictor),
  dispersion = female * 0.55 + male * 0.85,
  power = 1.50
)

set.seed(20260812)
tweedie_exchangeable_data <- dyadMLM::prepare_dyad_data(
  dyads_nbinom_ild,
  dyad = coupleID,
  member = personID,
  role = gender,
  time = diaryday,
  predictors = stress,
  model_types = "apim",
  keep_compositions = "female-female",
  seed = 123
)

exchangeable_dyad_ids <- unique(tweedie_exchangeable_data$coupleID)
exchangeable_shared_covariance <- matrix(
  c(
    0.38^2, 0.20 * 0.38 * 0.12,
    0.20 * 0.38 * 0.12, 0.12^2
  ),
  nrow = 2
)
exchangeable_difference_covariance <- matrix(
  c(
    0.22^2, -0.10 * 0.22 * 0.08,
    -0.10 * 0.22 * 0.08, 0.08^2
  ),
  nrow = 2
)
exchangeable_shared_effects <- matrix(
  stats::rnorm(length(exchangeable_dyad_ids) * 2),
  ncol = 2
) %*% chol(exchangeable_shared_covariance)
exchangeable_difference_effects <- matrix(
  stats::rnorm(length(exchangeable_dyad_ids) * 2),
  ncol = 2
) %*% chol(exchangeable_difference_covariance)
exchangeable_stable_index <- match(
  tweedie_exchangeable_data$coupleID,
  exchangeable_dyad_ids
)

exchangeable_occasion_keys <- unique(
  tweedie_exchangeable_data[c("coupleID", "diaryday")]
)
exchangeable_occasion_shared <- stats::rnorm(
  nrow(exchangeable_occasion_keys),
  0,
  0.22
)
exchangeable_occasion_difference <- stats::rnorm(
  nrow(exchangeable_occasion_keys),
  0,
  0.14
)
exchangeable_occasion_index <- match(
  paste(
    tweedie_exchangeable_data$coupleID,
    tweedie_exchangeable_data$diaryday,
    sep = "-"
  ),
  paste(
    exchangeable_occasion_keys$coupleID,
    exchangeable_occasion_keys$diaryday,
    sep = "-"
  )
)

exchangeable_contrast <-
  tweedie_exchangeable_data$.member_contrast_arbitrary
exchangeable_time_centered <-
  tweedie_exchangeable_data$diaryday -
  mean(tweedie_exchangeable_data$diaryday)
exchangeable_linear_predictor <-
  1.05 - 0.008 * exchangeable_time_centered +
  0.22 * tweedie_exchangeable_data$.stress_cwp_actor +
  0.09 * tweedie_exchangeable_data$.stress_cwp_partner +
  0.16 * tweedie_exchangeable_data$.stress_cbp_actor +
  0.07 * tweedie_exchangeable_data$.stress_cbp_partner +
  exchangeable_shared_effects[exchangeable_stable_index, 1] +
  exchangeable_contrast *
    exchangeable_difference_effects[exchangeable_stable_index, 1] +
  exchangeable_shared_effects[exchangeable_stable_index, 2] *
    tweedie_exchangeable_data$.stress_cwp_actor +
  exchangeable_contrast *
    exchangeable_difference_effects[exchangeable_stable_index, 2] *
    tweedie_exchangeable_data$.stress_cwp_actor +
  exchangeable_occasion_shared[exchangeable_occasion_index] +
  exchangeable_contrast *
    exchangeable_occasion_difference[exchangeable_occasion_index]

tweedie_exchangeable_data$time_centered <- exchangeable_time_centered
tweedie_exchangeable_data$tweedie_amount <- simulate_tweedie(
  mean = exp(exchangeable_linear_predictor),
  dispersion = rep(0.65, nrow(tweedie_exchangeable_data)),
  power = 1.50
)
```

```{r inspect-tweedie-data}
c(
  distinguishable_zero_proportion = mean(
    tweedie_distinguishable_data$tweedie_amount == 0
  ),
  exchangeable_zero_proportion = mean(
    tweedie_exchangeable_data$tweedie_amount == 0
  )
)

hist(
  tweedie_distinguishable_data$tweedie_amount,
  breaks = 30,
  main = "Simulated Tweedie outcome",
  xlab = "Amount"
)
```


### Distinguishable Tweedie APIM

The focused candidate below allows the stable intercepts and the focal
within-person actor slopes to vary and covary across both roles. It is maximal
only for those selected stable coefficients. If heterogeneity in time trends or
partner effects is scientifically relevant, their slope variances also belong
in the starting candidate. This follows the general random-effects workflow of
@delrosarioPracticalGuideSpecifying2025. Their fitted examples are Gaussian,
however; the same-occasion Tweedie block below is this vignette's link-scale
extension, not a structure evaluated in their paper.

The primary model starts with a common dispersion because the research question
concerns fixed actor and partner effects, not role differences in conditional
variability. For illustration, the within-person actor effect is designated as
the focal heterogeneous slope. The partner fixed effect remains in the mean
model; add its random slope to the starting candidate if heterogeneity in that
effect is also scientifically important.

```{r fit-tweedie-distinguishable, eval = FALSE}
tweedie_distinguishable_model <- glmmTMB::glmmTMB(
  tweedie_amount ~ 0 +
    # Role-specific fixed means and time trends
    .is_female +
    .is_male +
    .is_female:time_centered +
    .is_male:time_centered +

    # Role-specific within-person actor and partner effects
    .is_female:.stress_cwp_actor +
    .is_male:.stress_cwp_actor +
    .is_female:.stress_cwp_partner +
    .is_male:.stress_cwp_partner +

    # Role-specific between-person actor and partner effects
    .is_female:.stress_cbp_actor +
    .is_male:.stress_cbp_actor +
    .is_female:.stress_cbp_partner +
    .is_male:.stress_cbp_partner +

    # Stable role-specific intercepts and focal actor slopes
    us(
      0 +
        .is_female +
        .is_male +
        .is_female:.stress_cwp_actor +
        .is_male:.stress_cwp_actor |
        coupleID
    ) +

    # Same-occasion latent Gaussian covariance on the log scale
    us(
      0 + .is_female + .is_male |
        coupleID:diaryday
    ),
  dispformula = ~ 1,
  family = glmmTMB::tweedie(link = "log"),
  data = tweedie_distinguishable_data,
  control = glmmTMB::glmmTMBControl(profile = TRUE)
)

summary(tweedie_distinguishable_model)
```

If role-specific conditional dispersion is scientifically motivated, fit it as
a prespecified extension while leaving the mean and random-effects structure
unchanged:

```{r fit-tweedie-distinguishable-role-dispersion, eval = FALSE}
tweedie_distinguishable_role_dispersion_model <- update(
  tweedie_distinguishable_model,
  dispformula = ~ 0 + .is_female + .is_male
)
```

This extension is generically identified, but it asks the likelihood to
separate two role-specific latent occasion variances, two role-specific
Tweedie dispersions, and the Tweedie power. It may therefore be much more
weakly identified than the common-dispersion model.

#### Pooling the random covariance while retaining role fixed effects

For distinguishable dyads, pooling random effects does not require pooling the
fixed effects. When partner-swap symmetry is substantively defensible for the
random covariance - including equal corresponding role variances and mirrored
cross-coefficient covariances - the shared/difference specification below
imposes that exchangeable covariance while retaining every role-specific
coefficient in the mean model [@delrosarioPracticalGuideSpecifying2025].

```{r fit-tweedie-distinguishable-pooled-covariance, eval = FALSE}
tweedie_distinguishable_pooled_covariance_model <- glmmTMB::glmmTMB(
  tweedie_amount ~ 0 +
    .is_female +
    .is_male +
    .is_female:time_centered +
    .is_male:time_centered +
    .is_female:.stress_cwp_actor +
    .is_male:.stress_cwp_actor +
    .is_female:.stress_cwp_partner +
    .is_male:.stress_cwp_partner +
    .is_female:.stress_cbp_actor +
    .is_male:.stress_cbp_actor +
    .is_female:.stress_cbp_partner +
    .is_male:.stress_cbp_partner +

    # Exchangeable stable covariance; fixed effects remain distinguishable
    us(1 + .stress_cwp_actor | coupleID) +
    us(
      0 +
        .member_contrast_arbitrary +
        .member_contrast_arbitrary:.stress_cwp_actor |
        coupleID
    ) +

    # Exchangeable same-occasion covariance
    us(1 | coupleID:diaryday) +
    us(
      0 + .member_contrast_arbitrary |
        coupleID:diaryday
    ),
  dispformula = ~ 1,
  family = glmmTMB::tweedie(link = "log"),
  data = tweedie_distinguishable_data,
  control = glmmTMB::glmmTMBControl(profile = TRUE)
)
```

This is one possible simplification rung, not an automatic destination. If the
roles are fundamentally different in their heterogeneity, a defensible
alternative is to preserve role-specific variances while removing only the
higher-order random-coefficient correlations:

```{r distinguishable-diagonal-random-slopes, eval = FALSE}
# Replace the stable four-coefficient `us()` block by these two blocks:
us(0 + .is_female + .is_male | coupleID) +
diag(
  0 +
    .is_female:.stress_cwp_actor +
    .is_male:.stress_cwp_actor |
    coupleID
)
```

The focal slope variances remain. The `diag()` block removes intercept-slope,
slope-slope, and cross-role slope correlations; it does not remove the stable
partner-intercept covariance or the dyad-occasion covariance.

If equal role-specific slope variances are also defensible, pool those two
variances without removing the slopes:

```{r distinguishable-pooled-random-slopes, eval = FALSE}
homdiag(
  0 +
    .is_female:.stress_cwp_actor +
    .is_male:.stress_cwp_actor |
    coupleID
)
```


### Exchangeable Tweedie APIM

For exchangeable dyads, role-specific mean distinctions are absent and the
random covariance is pooled from the outset. Separate shared and
arbitrary-difference blocks impose those exchangeable covariance constraints
[@delrosarioPracticalGuideSpecifying2025]. The common Tweedie dispersion used
here is an additional family-specific constraint; it is not discussed in their
Gaussian examples.

```{r fit-tweedie-exchangeable, eval = FALSE}
tweedie_exchangeable_model <- glmmTMB::glmmTMB(
  tweedie_amount ~
    1 +
    time_centered +
    .stress_cwp_actor +
    .stress_cwp_partner +
    .stress_cbp_actor +
    .stress_cbp_partner +

    # Stable shared and difference intercepts and focal actor slopes
    us(1 + .stress_cwp_actor | coupleID) +
    us(
      0 +
        .member_contrast_arbitrary +
        .member_contrast_arbitrary:.stress_cwp_actor |
        coupleID
    ) +

    # Same-occasion shared and difference effects
    us(1 | coupleID:diaryday) +
    us(
      0 + .member_contrast_arbitrary |
        coupleID:diaryday
    ),
  dispformula = ~ 1,
  family = glmmTMB::tweedie(link = "log"),
  data = tweedie_exchangeable_data,
  control = glmmTMB::glmmTMBControl(profile = TRUE)
)

summary(tweedie_exchangeable_model)

dyadMLM::recover_exchangeable_covariance(
  tweedie_exchangeable_model
)
```

The shared and difference blocks are independent. This is part of the
exchangeability constraint: freely correlating shared and difference effects
would generally reintroduce unequal member variances. If the intercept-slope
correlations are unsupported, split each two-coefficient stable block into
separate one-coefficient blocks while retaining both intercept and focal slope
variances.

### Formula validation

The complete formulas were run separately because the full fits take roughly
two to three minutes each on the development machine. In one simulation with
120 dyads and 14 occasions, both models had convergence code zero, a
positive-definite Hessian, and no correlation at a boundary:

| Model | Tweedie power: truth, estimate | Dispersion: truth, estimate | Informative weak-identification signal |
|:--|:--|:--|:--|
| Distinguishable, role-specific dispersion | 1.50, 1.511 | female: 0.55, 0.548; male: 0.85, 0.829 | occasion correlation was 0.579 rather than 0.35 |
| Exchangeable, common dispersion | 1.50, 1.499 | 0.65, 0.663 | occasion difference SD was 0.080 rather than 0.14 |

For the exchangeable fit, the underestimated difference variance produced a
recovered member-level occasion correlation of 0.777 rather than the simulated
0.424. The covariance recovery algebra was correct; the fitted variance
decomposition was imprecise. This is exactly why convergence and a
positive-definite Hessian are not enough to establish strong identification.
One favorable simulation also does not establish repeated-sampling bias or
coverage.

A second simulation made the simplification decision more concrete. With 300
distinguishable dyads and eight occasions, the model with a correlated
intercept block and a separate `diag()` focal-slope block converged with a
positive-definite Hessian; its same-occasion correlation was 0.383 rather than
the simulated 0.40. Replacing those blocks with one four-coefficient `us()`
block failed to converge and estimated a slope correlation of 0.982 even though
the simulated correlation was 0.20. The near-boundary correlation was therefore
a symptom of the unsupported covariance decomposition, not evidence of a
nearly perfect population relationship.

In the same design, common- and role-specific-dispersion core models both
converged. Their four focal actor and partner estimates differed by less than
0.001, as did their standard errors, even though the role-specific dispersion
model detected the simulated dispersion difference. This supports common
dispersion as a practical first model when the research question concerns fixed
mean effects, followed by the role-specific model as a planned extension or
sensitivity analysis.


## A fixed-effect-first specification strategy

@delrosarioPracticalGuideSpecifying2025 recommend starting with all relevant
random variances and covariances, holding the fixed-effects model constant,
changing one random component at a time, and pooling equality-related
parameters before setting them to zero. Their worked examples are Gaussian;
the following adapts that workflow to a Tweedie APIM.

1. **Fix the scientific mean model first.** Prespecify centering, time, actor
   and partner effects, role interactions, family, and link. Do not select the
   covariance structure according to whether a focal fixed effect becomes
   significant.

2. **Start with the scientifically relevant heterogeneity.** Include stable
   intercepts, the focal within-person random slope when its effect could vary
   across dyads, and a minimal same-occasion partner covariance. More elaborate
   serial structure is a separate question.

3. **Pool before trimming, where equality is defensible.** For distinguishable
   dyads, first consider a common Tweedie dispersion and equality constraints
   on specific role-dependent nuisance parameters, such as the two focal slope
   variances. This need not mean imposing exchangeability on the complete
   covariance block. Pooling part or all of the random covariance does not
   require pooling role-specific fixed effects. Do not pool fundamentally
   different roles merely to obtain convergence.

4. **Trim covariance parameters before scientifically important variances, in
   a theory-based order.** Do not assume that every intercept-slope covariance
   is small: the paper treats within-person covariances as potentially among
   the largest. In its physiological example, cross-person,
   cross-variable covariances were expected to be smallest and were simplified
   before within-person covariances. For this APIM, simplify unsupported
   higher-order coefficient covariances before deleting their associated
   variances. Removing one of those covariances is not the same as deleting the
   same-occasion partner covariance, which represents the dyadic pairing at
   Level 1 [@delrosarioPracticalGuideSpecifying2025].

5. **Remove nonfocal random slopes before the focal slope.** This is the
   fixed-effect-priority recommendation adopted in this vignette, not an order
   tested by the paper. A random slope for a within-person predictor protects
   fixed-effect inference against genuine effect heterogeneity. Remove the
   focal slope only when its variance is consistently negligible and profiles,
   multistarts, and fixed-effect sensitivity checks are reassuring.

6. **Do not mechanically delete intercepts last.** Stable random intercepts and
   same-occasion covariance represent different clustering features. Remove a
   particular component only if the design and diagnostics support a value near
   zero. The most reduced model in the paper's worked sequence removes the
   G-matrix covariances but retains pooled intercept, stability-slope, and
   influence-slope variances as well as its Gaussian repeated partner structure.
   It therefore does not motivate automatically deleting slopes and then
   intercepts. If a minimally defensible model remains unstable, the design may
   not support the intended decomposition
   [@delrosarioPracticalGuideSpecifying2025].

An alternative build-up strategy reaches the same decisions from the other
direction: begin with all relevant variances, add the pooled covariances
expected to be largest or most important first, and, for distinguishable dyads,
relax those equality constraints one at a time while support remains adequate.
Whichever direction is used should be chosen in advance and applied one
component at a time [@delrosarioPracticalGuideSpecifying2025].


## Diagnosing weak identification

A zero convergence code and a positive-definite Hessian are necessary checks,
not proof of strong identification.

```{r tweedie-fit-checks, eval = FALSE}
tweedie_distinguishable_model$fit$convergence
tweedie_distinguishable_model$sdr$pdHess
glmmTMB::VarCorr(tweedie_distinguishable_model)
glmmTMB::fixef(tweedie_distinguishable_model)$disp
glmmTMB::diagnose(tweedie_distinguishable_model)
```

Weak identification often creates a flat likelihood ridge. For example, the
data may estimate a total marginal variance well while allowing latent
occasion variance and Tweedie dispersion to trade off. The inverse Hessian is
then large along that ridge, producing large standard errors and parameter
estimator correlations near ±1. A fitted random-effect correlation near ±1
can likewise indicate a nearly rank-deficient covariance matrix rather than a
truly perfect population relationship.

There is also a computational reason to calibrate these models: each member
coordinate of a dyad-occasion random effect is informed by one response, and
`glmmTMB` integrates over the many latent effects using a Laplace approximation.
Generic likelihood identification therefore does not guarantee accurate
finite-sample approximation or stable optimization.

Useful checks include:

- inspect standard deviations, covariances, correlations, and covariance-matrix
  eigenvalues rather than correlations alone;
- inspect the strongest correlations among all estimated parameters;
- profile important variance and dispersion parameters;
- refit from materially different starting values and check whether the same
  likelihood is reached with different variance decompositions;
- refit prespecified pooled, diagonal, and full structures and compare the
  focal fixed estimate, standard error, and interval;
- use repeated design-matched simulation to assess bias, interval coverage,
  boundary frequency, and convergence.

```{r tweedie-estimator-correlations, eval = FALSE}
full_vcov <- vcov(tweedie_distinguishable_model, full = TRUE)
estimator_correlations <- stats::cov2cor(as.matrix(full_vcov))

sort(
  abs(estimator_correlations[upper.tri(estimator_correlations)]),
  decreasing = TRUE
)[1:10]

conditional_covariance <-
  glmmTMB::VarCorr(tweedie_distinguishable_model)$cond
occasion_covariance <- as.matrix(
  conditional_covariance[["coupleID:diaryday"]]
)
eigen(occasion_covariance, symmetric = TRUE)$values
```

Large estimates that persist across parameterizations, have bounded profiles,
recover under repeated simulation, and become more precise with more dyads and
complete paired occasions are more credible as real heterogeneity. Estimates
that move along a ridge, hit boundaries, or change drastically across nearly
equivalent parameterizations are more consistent with weak identification.

Simulation should match the intended number of dyads and occasions, mean range,
zero proportion, Tweedie power and dispersion, role imbalance, missingness, and
number of complete partner occasions. It should vary those features rather than
validate only one favorable parameter setting. Using simulation during design
and preregistering the specification path are also central recommendations of
@delrosarioPracticalGuideSpecifying2025.


## Takeaway

An occasion-level Gaussian random-effect pair is a valid way to model
same-occasion partner dependence in a non-Gaussian APIM, but it is latent
covariance on the link scale. Tweedie dispersion remains conditional response
variation. Both role-specific distinguishable and pooled exchangeable
dispersion/covariance models can be identified with adequate repeated data;
the distinguishable model is simply more demanding and more likely to be
weakly identified.

For fixed-effect research questions, retain the focal mean structure, focal
random-slope variance, stable clustering, and minimal same-occasion partner
covariance as priorities. Simplify equality-related nuisance parameters and
higher-order random-coefficient correlations first, and treat boundary
estimates as prompts for profiles and sensitivity analyses rather than as an
automatic deletion rule.

---

Return to the
[Actor-Partner Interdependence Model vignette](https://pascal-kueng.github.io/dyadMLM/articles/apim.html),
see the
[Dyad-Individual Model vignette](https://pascal-kueng.github.io/dyadMLM/articles/dim.html)
or the
[Dyadic Score Model vignette](https://pascal-kueng.github.io/dyadMLM/articles/dsm.html)
for alternative parameterizations, or return to the
[online package overview](https://pascal-kueng.github.io/dyadMLM/).

## References

::: {#refs}
:::
