---
title: "APIMs with Mixed Dyad Compositions"
author: "Pascal Küng"
bibliography: ../../vignettes/references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{APIMs with Mixed Dyad Compositions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

diagram_candidates <- c(
  "diagram-helpers.Rinc",
  "../../vignettes/diagram-helpers.Rinc",
  "vignettes/diagram-helpers.Rinc"
)
diagram_helper <- diagram_candidates[file.exists(diagram_candidates)][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.", call. = FALSE)
}
sys.source(diagram_helper, envir = knitr::knit_global())

diagram_device <- .vignette_diagram_device()
knitr::opts_chunk$set(
  dev = diagram_device,
  fig.ext = diagram_device,
  dev.args = list(bg = "transparent")
)
```

```{r setup}
library(dyadMLM)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
mixed_apim_fitted_alt <-
  "Fitted mixed-composition APIM diagram unavailable."
```

Suppose you study how support relates to closeness in female–male,
female–female and male–male couples. Some people have no partner, or take part
without their partner. You want to compare groups while accounting for the
fact that partners' scores are related.

This vignette shows how to combine dyad groups in one APIM and how to include
people observed alone. You can choose which effects, variances and partner
correlations differ between groups and which are shared.

Distinguishable models allow parameters to differ by member role; exchangeable
models use the same parameters for both members. We adapt the structural
equation modeling approach of @Bolger2025Unified to a multilevel model.

> **Software.** These models use `glmmTMB`. Standard `brms` models with one
> outcome variable do not currently support the exact Gaussian residual
> covariance structures shown here.

This vignette builds on the
[Actor-Partner Interdependence Model vignette](apim.html). For the broader
package workflow and an overview of the other model-specific vignettes,
including the [Dyad-Individual Model](dim.html) and
[Dyadic Score Model](dsm.html), see the
[online package overview](https://pascal-kueng.github.io/dyadMLM/).

## Cross-sectional mixed-composition APIM

`dyads_cross` was simulated with actor and partner effects. We begin with an
intercept-only model to focus on residual covariance within each composition.

We first prepare the data with `dyadMLM::prepare_dyad_data()`, which identifies all
observed dyad compositions.

```{r prepare-mixed-cross-sectional}
mixed_cross_data <- dyadMLM::prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  seed = 123
)

print(mixed_cross_data, n = 4)
```

Each composition uses the
[distinguishable](apim.html#distinguishable-residual-structure) or
[exchangeable](apim.html#exchangeable-residual-structure) covariance structure
described in the APIM vignette.

The returned `.member_contrast_*` variables contain `-1` and `1` for the corresponding
exchangeable composition and `0` for all other compositions.
`dyadMLM::prepare_dyad_data()` creates this structure automatically.

The three composition-specific blocks are combined within one joint model:

```{r mixed-composition-diagram, echo=FALSE, fig.width=10, fig.height=5.5, out.width="100%", fig.cap="Structure of the cross-sectional mixed-composition APIM (Intercept only example). Female-male dyads have distinguishable member intercepts and an unrestricted residual covariance block. Each same-gender composition has a pooled intercept and an exchangeable residual covariance block. The three blocks are estimated jointly but are not connected by covariance parameters.", fig.alt="Diagram of one joint mixed-composition model containing three separate blocks. The female-male block has distinct female and male intercepts, separate residual variances, and their covariance. Female-female and male-male blocks each have one shared intercept, equal member residual variances, and a covariance. No covariance parameters connect the three composition blocks."}
draw_mixed_apim_diagram()
```

With one observation per person, the random-effects terms below represent
within-dyad residual covariance. `dispformula = ~ 0` fixes the remaining
Gaussian residual variance near zero.

We fit the intercept-only model as follows:

```{r fit-mixed-cross-sectional-gaussian, eval = has_glmmTMB}

mixed_cross_gaussian_model <- glmmTMB::glmmTMB(
  closeness ~ 0 +

    ##### INTERCEPTS ######
    # fixed intercepts for individuals from distinguishable female-male couples
    .is_female_x_male_female + .is_female_x_male_male +

    # fixed pooled intercept for female-female couples
    .is_female_x_female +

    # fixed pooled intercept for male-male couples
    .is_male_x_male +

    ##### RESIDUAL STRUCTURE ######
    # distinguishable female-male residual covariance
    us(0 + .is_female_x_male_female + .is_female_x_male_male | coupleID) +

    # exchangeable female-female residual covariance via sum-diff
    us(0 + .is_female_x_female | coupleID) +
    us(0 + .member_contrast_female_x_female_arbitrary | coupleID) +

    # exchangeable male-male residual covariance via sum-diff
    us(0 + .is_male_x_male | coupleID) +
    us(0 + .member_contrast_male_x_male_arbitrary | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_cross_data
)

summary(mixed_cross_gaussian_model)

```

The next diagram shows the fitted intercepts, residual standard deviations,
and within-dyad correlations. For exchangeable dyads, member residual SDs and
correlations are calculated from the fitted shared and contrast components.

```{r prepare-fitted-mixed-apim-diagram, include=FALSE, eval=has_glmmTMB}
mixed_apim_diagram_values <- .extract_mixed_apim_diagram_values(
  mixed_cross_gaussian_model
)
mixed_apim_diagram_estimates <- mixed_apim_diagram_values$estimates
mixed_apim_diagram_residuals <- mixed_apim_diagram_values$residuals

mixed_apim_fitted_alt <- sprintf(
  paste(
    "Fitted mixed-composition APIM. Female-male intercepts %.2f and %.2f,",
    "residual SDs %.2f and %.2f, and residual correlation %.2f.",
    "Female-female intercept %.2f, residual SD %.2f, and correlation %.2f.",
    "Male-male intercept %.2f, residual SD %.2f, and correlation %.2f."
  ),
  mixed_apim_diagram_estimates[["intercept_fm_female"]],
  mixed_apim_diagram_estimates[["intercept_fm_male"]],
  mixed_apim_diagram_residuals$fm[["sd_female"]],
  mixed_apim_diagram_residuals$fm[["sd_male"]],
  mixed_apim_diagram_residuals$fm[["correlation"]],
  mixed_apim_diagram_estimates[["intercept_ff"]],
  mixed_apim_diagram_residuals$ff[["sd"]],
  mixed_apim_diagram_residuals$ff[["correlation"]],
  mixed_apim_diagram_estimates[["intercept_mm"]],
  mixed_apim_diagram_residuals$mm[["sd"]],
  mixed_apim_diagram_residuals$mm[["correlation"]]
)
```

```{r fitted-mixed-apim-diagram, echo=FALSE, eval=has_glmmTMB, fig.width=10, fig.height=5.5, out.width="100%", fig.cap="Fitted intercept-only mixed-composition APIM. Intercepts and composition-specific residual parameters are extracted from the fitted joint model; the two exchangeable blocks are shown on the member scale.", fig.alt=mixed_apim_fitted_alt}
draw_mixed_apim_diagram(model = mixed_cross_gaussian_model)
```

Exactly one of the four fixed-effect indicators is 1 for each person; the
others are 0. With $I$ denoting a generated `.is_{composition_role}` indicator
and $b_0$ a fixed intercept, the model's fixed mean is:

$$
\mu_{ij}^{\mathrm{fixed}} =
b_{0,\mathrm{FM,F}} I_{\mathrm{FM,F},ij} +
b_{0,\mathrm{FM,M}} I_{\mathrm{FM,M},ij} +
b_{0,\mathrm{FF}} I_{\mathrm{FF},ij} +
b_{0,\mathrm{MM}} I_{\mathrm{MM},ij}.
$$

In `FM,F`, the first part identifies the female-male composition and the final
`F` identifies the female member. The same composition-first convention
is used throughout this vignette.

The active indicator for each person is shown below:

| Row | $I_{\mathrm{FM,F}}$ | $I_{\mathrm{FM,M}}$ | $I_{\mathrm{FF}}$ | $I_{\mathrm{MM}}$ | Fixed intercept |
|:--|--:|--:|--:|--:|:--|
| Female in a female-male dyad | **1** | 0 | 0 | 0 | $b_{0,\mathrm{FM,F}}$ |
| Male in a female-male dyad | 0 | **1** | 0 | 0 | $b_{0,\mathrm{FM,M}}$ |
| Member of a female-female dyad | 0 | 0 | **1** | 0 | $b_{0,\mathrm{FF}}$ |
| Member of a male-male dyad | 0 | 0 | 0 | **1** | $b_{0,\mathrm{MM}}$ |

For a member of a male-male dyad, this becomes:

$$
0 \cdot b_{0,\mathrm{FM,F}} + 0 \cdot b_{0,\mathrm{FM,M}} +
0 \cdot b_{0,\mathrm{FF}} + 1 \cdot b_{0,\mathrm{MM}}
= b_{0,\mathrm{MM}}.
$$

The random-effects terms reduce in the same way because their indicator columns
are also `0` outside the relevant composition. The model combines an
unrestricted distinguishable covariance block for female-male dyads with a
separate equal-variance covariance block for each exchangeable composition.

The random-effects terms form composition-specific blocks, so no covariance
parameters are estimated across compositions. Such parameters would not be
identified because each dyad belongs to only one composition.

The longitudinal example below adds slopes for each composition.

### Single versus separate mixed-composition fits

These models fit all dyad compositions in one call. This is useful for
comparing effects across compositions or testing whether parameters can be
shared.

`marginaleffects::hypotheses()` tests linear combinations of `glmmTMB` fixed
effects. In these equations, `conditional_` identifies coefficients for the
modeled mean of `closeness`. A positive estimated difference means the
coefficient or average on the left is higher than the one on the right.

```{r mixed-cross-sectional-contrasts, eval = FALSE}
# Test whether male-male and female-female pooled intercepts are different
marginaleffects::hypotheses(
  mixed_cross_gaussian_model,
  hypothesis = "conditional_.is_male_x_male = conditional_.is_female_x_female",
  re.form = NA
)

# Test whether the average female-male intercept is different from the pooled
# male-male intercept
marginaleffects::hypotheses(
  mixed_cross_gaussian_model,
  hypothesis = "(conditional_.is_female_x_male_female + conditional_.is_female_x_male_male) / 2 = conditional_.is_male_x_male",
  re.form = NA
)

# Test whether the male intercept from a female-male couple is different from
# the pooled male intercept from a male-male couple
marginaleffects::hypotheses(
  mixed_cross_gaussian_model,
  hypothesis = "conditional_.is_female_x_male_male = conditional_.is_male_x_male",
  re.form = NA
)

```

We set `re.form = NA` to compare fixed effects only.

Fitting all compositions together does not itself create partial pooling:
estimates are not automatically pulled toward those of other compositions.
Here, each estimated mean and covariance parameter belongs to one composition.
Separate fits therefore target the same parameters, apart from numerical
fitting differences.

We can also compare models that require parameters to be equal across
compositions. For example, we can test whether sharing fixed effects and
covariance parameters between male-male and female-female dyads worsens fit.
Use `dyadMLM::compare_nested_models()`, as shown in
[Testing distinguishability in the APIM vignette](apim.html#testing-distinguishability).
Both models must use the same observations. The restricted model must be
obtained by constraining parameters of the full model.

Partial pooling requires a different model, such as a common effect plus
random composition deviations.

### Model comparison with a restricted model

The [APIM vignette](apim.html#testing-distinguishability) compares a
distinguishable APIM with its exchangeable restriction. Here, we require
female-male and male-male dyads to share an intercept and an exchangeable
residual covariance matrix. Female-female dyads keep their own parameters.
These constraints need theoretical or empirical justification.

```{r prepare-mixed-cross-sectional-constrained, eval = has_glmmTMB}
mixed_cross_data_constrained <- dyadMLM::prepare_dyad_data(
  dyads_cross,
  dyad = coupleID,
  member = personID,
  role = gender,
  set_exchangeable_compositions = "male-female",
  pool_compositions = list(
    "non_female_x_female" = c("male-male", "female-male")
  ),
  seed = 123
)

mixed_cross_gaussian_model_constrained <- glmmTMB::glmmTMB(
  closeness ~ 0 +
    .is_female_x_female +
    .is_non_female_x_female +

    # exchangeable female-female residual covariance via sum-diff
    us(0 + .is_female_x_female | coupleID) +
    us(0 + .member_contrast_female_x_female_arbitrary | coupleID) +

    # pooled female-male and male-male residual covariance via sum-diff
    us(0 + .is_non_female_x_female | coupleID) +
    us(0 + .member_contrast_non_female_x_female_arbitrary | coupleID)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_cross_data_constrained
)

dyadMLM::compare_nested_models(
  mixed_cross_gaussian_model_constrained,
  mixed_cross_gaussian_model
)
```

Here, the constrained model fits worse than the full model. The test evaluates
all imposed constraints together, including the shared fixed effects and
residual covariance structure. It does not show which constraint accounts
for the difference.

## Combining dyads and individuals

Complete dyads and individuals observed alone can share one model. This lets
us estimate actor effects while accounting for dependence between partners.
A *singleton* is a person observed alone: either they have no partner or their
partner did not participate.

For this illustration, we simulate new closeness scores using the support
scores in `dyads_cross`, with **one observation per person**. Each original
composition provides 80 complete dyads and 40 singletons. The singletons
represent unpartnered people. Deleting a partner's row alone would instead
leave a partnered person with missing partner data.

```{r create-dyad-solo-example, include = FALSE}
set.seed(20261001)
solo_source <- dyads_cross |>
  dplyr::group_by(coupleID) |>
  dplyr::mutate(
    source_composition = paste(sort(as.character(gender)), collapse = "_x_"),
    partner_support = rev(provided_support),
    shared_error = rnorm(1, sd = 0.6)
  ) |>
  dplyr::ungroup()

singleton_clusters <- solo_source |>
  dplyr::distinct(coupleID, source_composition) |>
  dplyr::group_by(source_composition) |>
  dplyr::slice_sample(prop = 1 / 3) |>
  dplyr::ungroup()

removed_members <- solo_source |>
  dplyr::semi_join(singleton_clusters, by = "coupleID") |>
  dplyr::group_by(coupleID) |>
  dplyr::slice_sample(n = 1) |>
  dplyr::ungroup()

dyad_solo_example <- solo_source |>
  dplyr::anti_join(removed_members, by = "personID") |>
  dplyr::mutate(
    has_partner = as.integer(!coupleID %in% singleton_clusters$coupleID),
    closeness = 3 + 0.4 * (gender == "female") + 0.3 * has_partner +
      0.6 * (provided_support - 5) +
      0.3 * ifelse(has_partner == 1, partner_support - 5, 0) +
      shared_error + rnorm(dplyr::n(), sd = ifelse(gender == "female", 0.8, 1))
  ) |>
  dplyr::select(
    personID, coupleID, gender, source_composition,
    has_partner, closeness, provided_support
  )
```

### Prepare the data

`coupleID` identifies a complete dyad or one singleton; `personID` identifies
a person. Give unrelated singletons different `coupleID` values, never a
shared ID or `NA`.

First, let `dyadMLM` prepare the complete dyads. `incomplete_dyads = "drop"`
excludes singletons from this step. We keep the actor and partner scores and
combine the two same-gender member contrasts into `q`. Within an exchangeable
dyad, these contrasts assign one member `+1` and the other `-1`.

```{r prepare-complete-dyads-for-solo-example, message = FALSE}
library(dplyr)

complete_dyads <- dyad_solo_example |>
  prepare_dyad_data(
    dyad = coupleID, member = personID, role = gender,
    predictors = provided_support, incomplete_dyads = "drop", seed = 123
  ) |>
  as_tibble() |>
  transmute(
    coupleID, personID, gender, closeness, has_partner,
    composition = as.character(.composition),
    is_solo = 0L,
    x_actor = .provided_support_actor,
    x_partner = .provided_support_partner,
    q = .member_contrast_female_x_female_arbitrary +
      .member_contrast_male_x_male_arbitrary
  )
```

Next, prepare the singletons. Their actor score is their own support score.
Keep the undefined partner score as `NA` for now, and set `q = 1` so both
exchangeable variance components can contribute to singleton variance.
We keep each singleton's original simulation group only to construct these
examples. An unpartnered person has no dyad composition.

```{r prepare-singletons}
singletons <- dyad_solo_example |>
  add_count(coupleID) |>
  filter(n == 1) |>
  transmute(
    coupleID, personID, gender, closeness, has_partner,
    composition = source_composition,
    is_solo = 1L,
    x_actor = provided_support,
    x_partner = NA_real_,
    q = 1
  )
```

Stack the rows, then add numeric 0/1 indicators. A product such as
`is_solo * is_male` is 1 only for singleton men. We use the female-male sample
for distinguishable models and the male-male sample for exchangeable models;
the female-female sample works the same way.

```{r combine-dyads-and-singletons}
combined_data <- bind_rows(complete_dyads, singletons) |>
  mutate(
    is_dyad = 1L - is_solo,
    is_male = as.integer(gender == "male"),
    is_female = as.integer(gender == "female"),
    is_solo_male = is_solo * is_male,
    is_solo_female = is_solo * is_female,
    is_dyad_male = is_dyad * is_male,
    is_dyad_female = is_dyad * is_female
  )

fm_data <- combined_data |> filter(composition == "female_x_male")
mm_data <- combined_data |> filter(composition == "male_x_male")
```

### Start with intercepts

The models below allow different intercepts for singletons and dyad members.
The distinguishable models add the same gender difference in both groups.
Actor and partner predictors are added later; here, their unexplained effects
contribute to the residual covariance.

The four models differ in whether singletons and dyad members have equal
**residual variances**: the variation left after accounting for the modeled
mean differences. With `family = gaussian()` and `dispformula = ~ 0`,
the random-effects terms represent residual covariance. Separate terms are
independent. Matrix diagonals are variances; their square roots are SDs. We
omit the negligible residual variance retained by `glmmTMB` from the matrices.

### Distinguishable dyads: separate singleton variances

For female-male dyads, allow singleton men and women to have different residual
variances from their counterparts in dyads:

```{r dyad-solo-distinguishable-separate, eval = has_glmmTMB}
fm_separate <- glmmTMB::glmmTMB(
  closeness ~ is_solo + gender +
    (0 + is_solo_male | personID) +
    (0 + is_solo_female | personID) +
    (0 + is_dyad_male + is_dyad_female | coupleID),
  family = gaussian(), dispformula = ~ 0,
  data = fm_data
)
```

With dyad members ordered male, female:

$$
R_{\mathrm{dyad}} =
\begin{pmatrix}v_M & c \\ c & v_F\end{pmatrix},
\qquad R_{\mathrm{solo},M} = [\tau_M^2],
\qquad R_{\mathrm{solo},F} = [\tau_F^2].
$$

All four variances are estimated separately. The partner correlation is
$c / \sqrt{v_M v_F}$. Use this when equal singleton and dyad-member variances
are not plausible.

### Distinguishable dyads: pool variances by role

If singleton men can share the male dyad-member variance, and likewise for
women, use the role indicators for everyone:

```{r dyad-solo-distinguishable-pooled, eval = has_glmmTMB}
fm_pooled <- glmmTMB::glmmTMB(
  closeness ~ is_solo + gender +
    (0 + is_male + is_female | coupleID),
  family = gaussian(), dispformula = ~ 0,
  data = fm_data
)
```

$$
R_{\mathrm{dyad}} =
\begin{pmatrix}v_M & c \\ c & v_F\end{pmatrix},
\qquad R_{\mathrm{solo},M} = [v_M],
\qquad R_{\mathrm{solo},F} = [v_F].
$$

Each singleton helps estimate the residual variance for their role. Complete
dyads provide the paired observations needed to estimate covariance $c$.
Pooling does not create covariance between unrelated singletons. The partner
correlation remains freely estimated.

### Exchangeable dyads: separate singleton variance

For exchangeable dyads, pool the two members' variances but estimate a separate
singleton variance:

```{r dyad-solo-exchangeable-separate, eval = has_glmmTMB}
mm_separate <- glmmTMB::glmmTMB(
  closeness ~ is_solo +
    (0 + is_solo | personID) +
    (0 + is_dyad | coupleID) +
    (0 + is_dyad:q | coupleID),
  family = gaussian(), dispformula = ~ 0,
  data = mm_data
)
```

Let $a$ and $b$ be the **variances**, not SDs, of the shared and contrast terms:

$$
R_{\mathrm{dyad}} =
\begin{pmatrix}a+b & a-b \\ a-b & a+b\end{pmatrix},
\qquad R_{\mathrm{solo}} = [\tau^2].
$$

Dyad-member SDs are $\sqrt{a+b}$ and their correlation is $(a-b)/(a+b)$;
the singleton SD is $\tau$. The partner covariance can be positive or negative.
Use this when singletons may be more or less variable than dyad members.

### Exchangeable dyads: pool all member variances

With `q = +1` for singletons, both variance components contribute to everyone:

```{r dyad-solo-exchangeable-pooled, eval = has_glmmTMB}
mm_pooled <- glmmTMB::glmmTMB(
  closeness ~ is_solo +
    (1 | coupleID) + (0 + q | coupleID),
  family = gaussian(), dispformula = ~ 0,
  data = mm_data
)
```

$$
R_{\mathrm{dyad}} =
\begin{pmatrix}a+b & a-b \\ a-b & a+b\end{pmatrix},
\qquad R_{\mathrm{solo}} = [a+b].
$$

Everyone now has residual SD $\sqrt{a+b}$. Singletons inform this pooled
variance; only complete dyads provide paired observations to identify $a-b$.
Pooling can still change the estimated partner covariance and
correlation through the shared parameters.

Use either pooled structure when equal residual variances are plausible,
whether the singletons are incomplete dyads or genuinely unpartnered people.
Missing a partner's data does not by itself justify pooling: people whose
partners do not participate may differ from those observed together. With
separate singleton variances, only shared mean parameters can still link
singleton and dyad estimates.

**Do not use `q = 0` for singletons to impose equal variances:** it gives them
variance $a$, while dyad members retain $a+b$. This is a different constraint.
Arbitrary singleton signs
(`+1` or `-1`) give the same pooled model because the two terms are independent.

### Adding actor and partner effects

A partner score is undefined when no partner exists. To retain these rows,
use the two-part predictor of @dziakTwoPartPredictors2017 (Method 1): include
status and the score, coding the score zero for unpartnered people.

At each measurement occasion:

- If the participant **has a partner**, set `has_partner = 1`. Use the partner's
  predictor value if known; otherwise leave it as `NA`.
- If the participant **has no partner**, set `has_partner = 0` and the recoded
  predictor `partner_value = 0`, never `NA`.

`has_partner` records relationship status, not whether both people supplied
data. Here, the singleton rows are unpartnered by design. In your own data,
use the recorded status; do not infer it from the number of rows per couple.

Subtract the mean actor score from both actor and partner scores. Then set
`partner_value` to zero for unpartnered people. The original `x_partner`
column keeps its missing values.

```{r prepare-two-part-partner-predictor}
mm_data <- mm_data |>
  mutate(
    x_actor_c = x_actor - mean(x_actor),
    x_partner_c = x_partner - mean(x_actor),
    partner_value = if_else(has_partner == 1, x_partner_c, 0)
  )
```

Now add partner status and the recoded score. We keep the pooled exchangeable
covariance here; any appropriate structure above can be used.

```{r dyad-solo-partner-effects, eval = has_glmmTMB}
mm_apim <- glmmTMB::glmmTMB(
  closeness ~ has_partner + x_actor_c + partner_value +
    (1 | coupleID) + (0 + q | coupleID),
  family = gaussian(), dispformula = ~ 0,
  data = mm_data
)

nobs(mm_apim)
```

All 200 people remain in the fit: 160 dyad members and 40 singletons.

With $x_A$ and $x_P$ denoting centered scores, the fixed means are:

$$
\mu =
\begin{cases}
\beta_0 + \beta_A x_A, & \text{unpartnered}, \\
\beta_0 + \beta_H + \beta_A x_A + \beta_P x_P,
  & \text{partnered}.
\end{cases}
$$

The status coefficient compares partnered and unpartnered people when partner
support equals the centering reference. The partner slope applies among
partnered people; the actor slope is shared.

An existing partner's unknown predictor stays `NA`. Such rows are normally
excluded from the fit unless you address the missing data, for example through
multiple imputation. A missing partner outcome alone does not imply that the
partner predictor is missing.

With repeated observations, status and partner scores may vary by occasion.
Because `partner_value` is zero for unpartnered observations, multiplying it
by `has_partner` does not change it. Account for dependence within people
and dyads, keeping person IDs stable if relationships change. Model
within-person changes separately from between-person differences
[@curranDisaggregationWithinBetween2011]; a random intercept alone does not do
this. If centering or decomposing partner scores, preserve structural zeros
in the terms that apply only during partnered occasions.

### Separate actor slopes

Here, all singletons are unpartnered. Add an interaction with `has_partner`
to allow their actor slope to differ from that of complete-dyad members,
keeping the partner predictor and covariance structure unchanged.

```{r dyad-solo-separate-actor-effects, eval = has_glmmTMB}
mm_apim_by_status <- update(mm_apim, . ~ . + has_partner:x_actor_c)
b <- glmmTMB::fixef(mm_apim_by_status)$cond

tibble(
  Effect = c("Actor: unpartnered", "Actor: partnered", "Partner: partnered"),
  `Generating value` = c(0.6, 0.6, 0.3),
  Estimate = c(
    b["x_actor_c"],
    b["x_actor_c"] + b["has_partner:x_actor_c"],
    b["partner_value"]
  )
) |> knitr::kable(digits = 3)
```

The actor coefficient is the slope for unpartnered people. Add the interaction
coefficient to obtain the slope for partnered people. Both generating actor
slopes are 0.6 here, so their fitted difference reflects sampling variation.
If singletons include partnered people whose partners did not participate,
use `is_solo * x_actor_c` for different actor slopes among people observed
alone versus with their partner, retaining `has_partner` and `partner_value`.

## Mixed-composition intensive longitudinal (ILD) Gaussian model

`dyads_ild` contains the same three dyad compositions used in the
cross-sectional example, with repeated paired observations over `diaryday`.

To keep the example lighter, we pool female-female and male-male dyads into one
same-sex composition. This is a simplifying modeling assumption and does not
reflect all differences used to simulate these example data. In an applied
analysis, pooling should be theoretically justified or evaluated by comparing
pooled and unpooled models fitted to the same observations.

Aside from the pooling, we prepare the data in the same way, now adding `time`
and a predictor:

```{r prepare-mixed-ild}
mixed_ild_data <- dyadMLM::prepare_dyad_data(
  dyads_ild,
  dyad = coupleID,
  member = personID,
  role = gender,
  pool_compositions = list(
    "same-sex" = c("male-male", "female-female")
  ),
  time = diaryday,
  predictors = provided_support,
  seed = 123
)

print(mixed_ild_data)
```

The within-person (`cwp`) and between-person (`cbp`) predictors follow the
[longitudinal APIM preparation](apim.html#intensive-longitudinal-apims), which
explains their interpretation and the limitations of observed person means.

This mixed-composition ILD model includes composition-specific fixed
intercepts, time slopes, and within-person and between-person actor and partner
effects. It omits random slopes, as they would add several variances and
covariances to an already large model.

The dyad-level random terms describe differences that persist across days.
The dyad-by-day terms describe residual variation and partner covariance on
the same day.

This model can be difficult to fit when the data provide too little information
about its variance and covariance parameters. Variances close to zero can also
cause estimation problems [@delrosarioPracticalGuideSpecifying2025].
For these example data, we use the BFGS optimizer with a higher iteration limit.

```{r fit-mixed-ild-gaussian, eval = FALSE}

mixed_ild_gaussian_model <- glmmTMB::glmmTMB(
  closeness ~ 0 +

    # Composition-specific intercepts
    .is_female_x_male_female + .is_female_x_male_male +

    .is_same_sex +

    # Composition-specific time trends
    .is_female_x_male_female:diaryday +
    .is_female_x_male_male:diaryday +

    .is_same_sex:diaryday +

    # Composition-specific within-person actor effects
    .is_female_x_male_female:.provided_support_cwp_actor +
    .is_female_x_male_male:.provided_support_cwp_actor +

    .is_same_sex:.provided_support_cwp_actor +

    # Composition-specific within-person partner effects
    .is_female_x_male_female:.provided_support_cwp_partner +
    .is_female_x_male_male:.provided_support_cwp_partner +

    .is_same_sex:.provided_support_cwp_partner +

    # Composition-specific between-person actor effects
    .is_female_x_male_female:.provided_support_cbp_actor +
    .is_female_x_male_male:.provided_support_cbp_actor +

    .is_same_sex:.provided_support_cbp_actor +

    # Composition-specific between-person partner effects
    .is_female_x_male_female:.provided_support_cbp_partner +
    .is_female_x_male_male:.provided_support_cbp_partner +

    .is_same_sex:.provided_support_cbp_partner +

    # stable dyad-level covariance
    us(0 + .is_female_x_male_female + .is_female_x_male_male | coupleID) +

    us(0 + .is_same_sex | coupleID) +
    us(0 + .member_contrast_same_sex_arbitrary | coupleID) +

    # same-day covariance
    us(0 + .is_female_x_male_female + .is_female_x_male_male | coupleID:diaryday) +

    us(0 + .is_same_sex | coupleID:diaryday) +
    us(0 + .member_contrast_same_sex_arbitrary | coupleID:diaryday)

  , dispformula = ~ 0
  , family = gaussian()
  , data = mixed_ild_data
  # Non-default settings help this example converge.
  , control = glmmTMB::glmmTMBControl(
      optimizer = stats::optim,
      optArgs = list(method = "BFGS"),
      optCtrl = list(maxit = 10000)
    )
)

summary(mixed_ild_gaussian_model)

```

To add random actor and partner slopes, replace the stable dyad-level
covariance terms with the following blocks:

```{r mixed-ild-random-slopes, eval = FALSE}
# distinguishable female-male random intercepts and slopes
us(
  0 +
    .is_female_x_male_female +
    .is_female_x_male_male +
    .is_female_x_male_female:.provided_support_cwp_actor +
    .is_female_x_male_male:.provided_support_cwp_actor +
    .is_female_x_male_female:.provided_support_cwp_partner +
    .is_female_x_male_male:.provided_support_cwp_partner |
    coupleID
) +

# exchangeable same-sex shared block
us(
  0 +
    .is_same_sex +
    .is_same_sex:.provided_support_cwp_actor +
    .is_same_sex:.provided_support_cwp_partner |
    coupleID
) +

# exchangeable same-sex difference block
us(
  0 +
    .member_contrast_same_sex_arbitrary +
    .member_contrast_same_sex_arbitrary:.provided_support_cwp_actor +
    .member_contrast_same_sex_arbitrary:.provided_support_cwp_partner |
    coupleID
)
```

This extension estimates a very large covariance structure and may not be
supported by many datasets.

## Takeaway

These examples bring established methods into one R workflow: compare dyad
groups, choose which parameters to share, and include people observed alone.
`dyadMLM` creates the columns for the mixed-composition models.

It is often useful to begin with separate composition-specific models or
simpler covariance structures. Combine them when an equality constraint or a
comparison of effects across compositions is part of the research question,
and add further random effects only when the data support the additional
complexity.

[Simulations](mixed-apim-recovery.html) supported parameter recovery for the
cross-sectional and random-intercept Gaussian models combining distinguishable
and exchangeable dyads. Some 95% Wald confidence intervals included the true value
less often than expected, and many random-slope fits failed the fit checks.

Singleton models were not included. The proposed extension would check both
variance choices and shared or separate actor slopes, including partnered
people observed alone.

---

Return to the [Actor-Partner Interdependence Model vignette](apim.html), see
the [Dyad-Individual Model vignette](dim.html) or the
[Dyadic Score Model vignette](dsm.html) for alternative parameterizations, or
return to the
[online package overview](https://pascal-kueng.github.io/dyadMLM/).

## References

::: {#refs}
:::
