---
title: "Dyadic Data Analysis with R"
subtitle: "Workshop Exercises"
author:
  - "Pascal Küng"
  - "Turu Stadler"
format:
  html:
    embed-resources: true
    toc: true
    toc-title: "Exercises"
editor: source
---

# Exercises

These exercises follow the six-step cross-sectional APIM workflow from the
applied tutorial.

Download the complete workshop materials from the 
[workshop page](https://pascal-kueng.github.io/dyadMLM/workshop/). 
Package documentation and vignettes describing model fitting for dyadic MLMs
are available through the [`dyadMLM` homepage](https://pascal-kueng.github.io/dyadMLM/).

# Getting started

An R Markdown (`.Rmd`) file combines *markdown* text -- such as this paragraph --
with runnable R code chunks.

Run code chunks individually via the green little play button on the 
top-right of each chunk. Try it and run the next chunk. It installs/updates
all workshop packages and their dependencies. This may take several minutes.

```{r eval=FALSE}
source("00_setup.R")
```



In RStudio, insert a new R chunk with
Ctrl+Alt+I on Windows or Linux, or Cmd+Option+I on macOS.

Try it and insert a new chunk between the green lines: 

---



---


Load all required packages:

```{r}
library(glmmTMB)
library(dyadMLM)
library(easystats, quietly = TRUE) # This package likes to be especially noisy
library(DHARMa)
library(dplyr)

source("hints.R")

```


While going through the exercises you can sometimes spot optional hints: 

```{r}
# hint('example_hint', 'run_chunk_without_comment') 
```

To get the hints, delete the `#` of the hint you want to read. `#` marks 
the whole line a comment that is not evaluated by R. 

After deleting the `#`, run the chunk (again). You can try it above.


# Loading, inspecting, and preparing the data

Follow a simplified workflow:

1. Define dyads, distinguishability, and research questions
2. Prepare the data and validate the dyadic structure
3. Describe and visualize measures and observed relationships
4. Estimate the model
5. Verify model adequacy, distinguishability, and robustness
6. Visualize, interpret, and report results

# 1. Define dyads, distinguishability, and research questions

You may use any of the variables in the dataset as outcome or predictor.

Place `dyadic-person-means.rds` in the same folder as this exercise file, then
load and inspect the dataset:

```{r}
raw_dyad_data <- readRDS("dyadic-person-means.rds")

head(raw_dyad_data)
names(raw_dyad_data)
```


Choose your research question:

> [Insert your RQ]

```{r}
# hint("research_question", "concept")
```

# 2. Prepare the data and validate the dyadic structure

You may use `dyadMLM::prepare_dyad_data()`. Load the package:

```{r}
library(dyadMLM)
```

For help with the arguments and how to use the function, run:

```{r}
?prepare_dyad_data()
```

Adding a `?` before any package function will produce a helpful help-page.

---

Prepare and validate the data for a **distinguishable** model now.

Hint: distinguish partners using `role = gender`. Set
`include_arbitrary_member_contrast = TRUE` to use the same prepared data for
distinguishable and exchangeable models later. Use
`add_apim_gmc_predictors = TRUE` if you want the grand-mean-centered APIM
predictors used in the tutorial.

Save the prepared data as `my_data`; later exercise code uses this name.

```{r}
# hint("prepare_data", "concept")
# hint("prepare_data", "structure")
```

Inspect the prepared data and verify the dyad IDs, member IDs, roles, row
counts, missingness, and generated columns:

```{r}

```


# 3. Describe and visualize measures and observed relationships

Adapt the summary and plotting code from the tutorial to your focal variables.
At minimum, summarize the variables by role, inspect their distributions,
calculate observed dyadic correlations, and visualize partner similarity.

```{r}
# hint("describe_data", "concept")
```

Describe what you see and how you interpret the summaries and plots:

[Insert your brief summary here]


# 4. Estimate the model

First, estimate a model for distinguishable dyads:

```{r eval=FALSE}
# hint("distinguishable_model", "concept")
# hint("distinguishable_model", "structure")
# hint("distinguishable_model", "example_code")

my_model <- glmmTMB(
  formula = YOUR_MODEL_FORMULA,
  data = my_data,
  family = gaussian()
)

summary(my_model)

```



# 5. Verify model adequacy, distinguishability, and robustness

### Adequacy

Check if the model converged and if all estimates look plausible.

Then, use DHARMa for residual diagnostics via simulation:

```{r eval=FALSE}
my_model_dharma_residuals <- DHARMa::simulateResiduals(
  fittedModel = my_model,
  n = 10000,

  # Re-simulate the random-effects block that represents the residuals.
  simulateREs = "unconditional",

  # Account for the correlation between partners' residuals.
  rotation = "estimated",
  seed = 123
)

plot(my_model_dharma_residuals)
```

Next, inspect each role separately:

```{r eval=FALSE}
female_dharma_residuals <- DHARMa::recalculateResiduals(
  my_model_dharma_residuals,
  sel = ~ .is_female == 1,
  rotation = NULL,
  seed = 123
)

male_dharma_residuals <- DHARMa::recalculateResiduals(
  my_model_dharma_residuals,
  sel = ~ .is_male == 1,
  rotation = NULL,
  seed = 123
)

old_par <- par(mfrow = c(1, 2))
DHARMa::plotResiduals(female_dharma_residuals, main = "Women")
DHARMa::plotResiduals(male_dharma_residuals, main = "Men")
par(old_par)
```

Interpreting these residual diagnostics takes practice. Consult the DHARMa
vignette section on
[interpreting residuals and recognizing misspecification](https://stat.ethz.ch/CRAN/web/packages/DHARMa/vignettes/DHARMa.html#interpreting-residuals-and-recognizing-misspecification-problems).
For now, briefly assess whether the diagnostics appear acceptable overall and
for each role.


### Robustness

Optionally, use the supplied helper to inspect unusual predictor–outcome
combinations separately by role:

```{r eval=FALSE}
# The function is defined in the separate helper file.
source("helpers.R")

women_mahalanobis_plot <- plot_mahalanobis(
  data = dplyr::filter(my_data, gender == "female"),
  x = "YOUR_ACTOR_PREDICTOR_COLUMN",
  y = "YOUR_OUTCOME_COLUMN",
  title = "Women"
)

men_mahalanobis_plot <- plot_mahalanobis(
  data = dplyr::filter(my_data, gender == "male"),
  x = "YOUR_ACTOR_PREDICTOR_COLUMN",
  y = "YOUR_OUTCOME_COLUMN",
  title = "Men"
)

see::plots(women_mahalanobis_plot, men_mahalanobis_plot)
```

Mahalanobis distance identifies unusual bivariate combinations, not formal
model influence, and the displayed regression line is not the adjusted APIM
effect. Use the labels to guide data inspection rather than to automatically
exclude dyads.

If a dyad was flagged for a substantive or procedural reason, optionally refit
the model without it:

```{r eval=FALSE}
data_without_selected_dyads <- dplyr::filter(
  my_data,
  !couple_id %in% DYAD_IDS_TO_EXCLUDE
)

model_without_selected_dyads <- update(
  my_model,
  data = data_without_selected_dyads
)
```

Do not exclude dyads merely because their removal changes a significance label.


### Distinguishability

Fit the equivalent model under exchangeability constraints. Hint: you need an
`.arbitrary_member_contrast` column in the data. You may already have included
one earlier, or you may need to prepare the data again. Check your column names
with `names(my_data)`.


```{r}
# hint("distinguishability", "concept")
# hint("distinguishability", "structure")

```

Compare the distinguishable and exchangeable models using
`dyadMLM::compare_nested_models()`.

Insert your brief conclusion, and choose whether you want to continue with
the distinguishable or exchangeable model for the rest of the exercise:

[Insert conclusion]

[Insert decision and reasoning]


# 6. Visualize, interpret, and report results

Print a fixed-effects table using `parameters::model_parameters()` (remember
to use `?` if you need help with arguments):

```{r}
# hint("report_results", "concept")
# hint("report_results", "structure")
```

Print a random-effects table:

```{r}

```

Plot the fixed effects:

```{r}

```


Interpret your results in practical terms. Remember that the interpretation
of the intercepts depends on whether you used grand-mean-centered variables.

Your interpretation:

[insert interpretation]


You have now completed the APIM workflow!


## Compute DIM estimates from the exchangeable APIM

The transformation from exchangeable APIM coefficients to DIM estimates is
particularly simple.
Check the output from your **exchangeable** model, and enter your coefficients
here:

```{r}

# hint("dim_transformation", "concept")

b_actor <- YOUR_ACTOR_COEFFICIENT
b_partner <- YOUR_PARTNER_COEFFICIENT

```

Recover the DIM estimates:

```{r}

b_mean <- YOUR_FORMULA
b_dev <- YOUR_FORMULA

cat("The between-dyad effect is: ", round(b_mean, 2), "\n", # newline
    "The within-dyad effect is: ", round(b_dev, 2)
)

```


Interpret the effects:

[Insert interpretation]


Optionally, fit a DIM or DSM directly using the package vignettes:

- [DIM vignette](https://pascal-kueng.github.io/dyadMLM/articles/dim.html)
- [DSM vignette](https://pascal-kueng.github.io/dyadMLM/articles/dsm.html)


# Intensive longitudinal dyads

The intensive longitudinal exercises will be added with the ILD tutorial.
