---
title: "Dyadic Data Analysis with R"
subtitle: "Intensive Longitudinal Workshop Exercises"
author:
  - "Pascal Küng"
  - "Turu Stadler"
format:
  html:
    embed-resources: true
    toc: true
    toc-title: "ILD Exercises"
editor: source
---

# Exercises

These exercises are a fast-finisher or take-home extension of the
cross-sectional APIM workflow. Complete the cross-sectional exercise first;
faster participants can continue here during the workshop.

Download the complete workshop materials from the
[workshop page](https://pascal-kueng.github.io/dyadMLM/workshop/). Package
documentation and vignettes are available on the
[`dyadMLM` homepage](https://pascal-kueng.github.io/dyadMLM/).


# Getting started

Keep `04_exercises_ild.Rmd`, `dyadic-ild-exercise.rds`, `00_setup.R`,
`hints.R`, and `helpers.R` in the same folder.

If you have not already run the workshop setup, run the next chunk once, in a
fresh R session. It installs or updates the workshop packages and any
dependencies that need updating, but does not update unrelated packages. This
may take several minutes.

```{r eval=FALSE}
source("00_setup.R")
```

Load the required packages, optional hints, and ILD diagnostic helpers:

```{r}
library(glmmTMB)
library(dyadMLM)
library(wbCorr)
library(easystats, quietly = TRUE)
library(DHARMa)
library(dplyr)
library(ggplot2)

source("hints.R")
source("helpers.R")
```

Optional hints appear as commented `hint()` calls. Uncomment and run them one
at a time.


# The ILD workflow

The same six-step workflow still applies:

1. Define dyads, roles, and research questions
2. Prepare the data and validate the dyadic structure
3. Describe and visualize measures and observed relationships
4. Estimate the models
5. Verify model adequacy, role differences, and robustness
6. Visualize, interpret, and report results

ILD additionally requires attention to matched occasions, time gaps,
within- and between-person variation, time trends, and residual temporal
dependence.


# 1. Define dyads, roles, and research questions

The cross-sectional exercise let you choose the predictor and outcome. Here,
everyone uses one prescribed question and a focused simulated dataset so that
the focus remains on the new ILD decisions and results can be compared across
groups. This is a teaching choice, not a claim that there is only one sensible
question or causal direction.

Load and inspect the focused workshop dataset:

```{r}
raw_ild_data <- readRDS("dyadic-ild-exercise.rds")

raw_ild_data
names(raw_ild_data)
```

The data contain newly simulated daily scores for female--male dyads only.
Gender is the distinguishing variable. Higher scores indicate more provided
support and greater closeness.

Our concurrent research question is:

> **How are daily deviations and stable between-person differences in provided
> support associated with one's own and one's partner's closeness on the same
> day?**

Unlike the prospective workshop example, this question concerns the same day.
Therefore, do not request lagged predictors; use the current-day within-person
columns generated by `prepare_dyad_data()`. In particular,
`lag1_predictors` is omitted and these columns do not end in `_lag1`.

Translate this into four questions:

> [Insert the within-person actor question]

> [Insert the within-person partner question]

> [Insert the between-person actor question]

> [Insert the between-person partner question]

Briefly state what the concurrent design can and cannot establish:

> [Insert your answer]


# 2. Prepare the data and validate the dyadic structure

Prepare a **distinguishable** APIM dataset with
`dyadMLM::prepare_dyad_data()`. Reuse the preparation workflow from the
cross-sectional exercise, now adding the time variable and a two-level temporal
decomposition.

Your call should:

- identify `coupleID`, `personID`, `gender`, and `diaryday`
- use `provided_support` as the predictor
- treat gender as the distinguishing role
- create APIM predictors at the within-person and between-person levels

Save the result as `my_ild_data`; all later models should use this object.

```{r eval=FALSE}
my_ild_data <- dyadMLM::prepare_dyad_data(
  data = raw_ild_data,
  # Complete the required arguments.
)

# hint("ild_preparation", "solution")
```

Add two time variables:

- `diaryday_c`: centered at the study midpoint and scaled in full-study units
- `diaryday_f`: a factor containing every scheduled day from 0 through 13

These names are used below.

```{r eval=FALSE}
my_ild_data <- my_ild_data |>
  dplyr::mutate(
    # Create diaryday_c and diaryday_f.
  )

# hint("ild_preparation", "time_variables")
```

Inspect the prepared data and its generated columns:

```{r eval=FALSE}
my_ild_data
names(my_ild_data)
```


## Validate the dyadic and temporal structure

This teaching dataset is deliberately complete and balanced. Run the checks
anyway: the same checks reveal unmatched occasions and time gaps in real ILD.

Check:

- the number of retained dyads and members
- whether every dyad contains exactly two members with stable roles
- whether dyad--member--day combinations are unique
- the scheduled and observed day ranges
- missing observations and matched partner occasions
- whether consecutive observations are always one day apart

```{r}
# hint("ild_validation", "functions")
# hint("ild_validation", "solution")

```

Summarize what you found:

> [Insert your data-structure and missingness summary]


# 3. Describe and visualize measures and observed relationships


## Describe the focal variables

Reuse the simple tools from the cross-sectional exercise: summarize
`provided_support` and `closeness` overall and by role, then inspect
role-specific histograms. Means and standard deviations are sufficient here.

```{r}
# Overall and role-specific summaries
# hint("ild_descriptives", "summary")

# Role-specific histograms
# hint("ild_descriptives", "histograms")

```


## Plot paired trajectories

Plot both variables over time for a manageable selection of dyads. Make the two
members identifiable and avoid drawing every trajectory in one panel.

```{r}
trajectory_dyads <- sort(unique(my_ild_data$coupleID))[1:4]

trajectory_data <- my_ild_data |>
  dplyr::filter(coupleID %in% trajectory_dyads) |>
  dplyr::select(
    coupleID, personID, diaryday, gender,
    provided_support, closeness
  ) |>
  tidyr::pivot_longer(
    cols = c(provided_support, closeness),
    names_to = "measure",
    values_to = "score"
  )

# Complete the mappings and facet specification, then run the plot:
# ggplot2::ggplot(
#   trajectory_data,
#   ggplot2::aes(
#     x = ...,
#     y = ...,
#     color = ...,
#     group = ...
#   )
# ) +
#   ggplot2::geom_line(na.rm = TRUE) +
#   ggplot2::geom_point(na.rm = TRUE) +
#   ggplot2::facet_grid(... ~ ..., scales = "free_y") +
#   ggplot2::labs(x = "Diary day", y = NULL, color = "Gender") +
#   ggplot2::theme_minimal()

# hint("ild_trajectories", "structure")
# hint("ild_trajectories", "solution")

```


## Optional extension: separate within- and between-person information

If time permits, adapt the wide-data correlation workflow from the
cross-sectional exercise. Create one row per couple and day, with separate
female and male columns for each focal variable. Then use `wbCorr()` with
`coupleID` as the cluster to inspect:

- ICCs for each role-specific series
- correlations among daily deviations within couples
- correlations among role-specific person means between couples

Map the relevant entries to the actor and partner questions. Treat these
correlations as descriptive rather than as replacements for the APIM.
Multilevel reliability would also matter for multi-item measures, but the
item-level data needed to estimate it are not included here.

```{r}
# hint("ild_level_associations", "structure")
# hint("ild_level_associations", "solution")

```

Describe the main patterns:

> [Insert your descriptive interpretation]


# 4. Estimate the initial distinguishable model

Begin with the scientific mean structure and no residual temporal process. Fit
a Gaussian distinguishable APIM with:

- role-specific intercepts and linear time trends
- role-specific within-person actor and partner effects
- role-specific between-person actor and partner effects
- stable role-specific covariance at the dyad level
- same-occasion role-specific covariance at the dyad--day level

Use `dispformula = ~ 0` so the modeled dyad--day block represents the Gaussian
observation-level residual structure. Adapt the distinguishable ILD model from
the applied tutorial and inspect `names(my_ild_data)` when selecting columns.

To keep this exercise focused and fitting times manageable, use random
intercepts but no random slopes. In substantive analyses, consider
theoretically plausible within-person random slopes when the data can support
them; their omission here is a teaching and computational simplification.

The covariance model has several components. Use this tested optimizer control
for every candidate so that the comparison does not also change the estimation
strategy:

```{r}
ild_control <- glmmTMB::glmmTMBControl(
  profile = TRUE,
  optimizer = stats::optim,
  optArgs = list(method = "BFGS"),
  optCtrl = list(maxit = 2000, reltol = 1e-10)
)
```

```{r eval=FALSE}
# hint("ild_models", "distinguishable_no_ar")

distinguishable_no_ar_model <- glmmTMB::glmmTMB(
  formula = YOUR_DISTINGUISHABLE_FORMULA,
  dispformula = ~ 0,
  family = gaussian(),
  data = my_ild_data,
  control = ild_control
)

summary(distinguishable_no_ar_model)
```

Inspect the model output for convergence or Hessian warnings, boundary
estimates, and the plausibility of all estimates before continuing. A zero
convergence code is not enough when the Hessian is non-positive-definite or a
covariance estimate is on its boundary.


# 5. Verify model adequacy, role differences, and robustness


## Does the model reproduce the temporal dependence?

Use the workshop helper to simulate block-whitened DHARMa residuals. The helper
accounts for fitted dependence within each dyad rather than treating every
person-day as independent. Pass `role` and request `by_role = TRUE` so the
female and male lag-1 checks are not averaged together.

```{r eval=FALSE}
no_ar_dharma_residuals <- simulate_ild_dharma(
  model = distinguishable_no_ar_model,
  dyad = my_ild_data$coupleID,
  member = my_ild_data$personID,
  time = my_ild_data$diaryday,
  role = my_ild_data$gender,
  n = 1000,
  seed = 123
)

plot(
  no_ar_dharma_residuals,
  title = "Block-whitened DHARMa residuals"
)

test_ild_lag1(no_ar_dharma_residuals, by_role = TRUE)
```

Consider:

- overall calibration and visible residual patterns
- the magnitude of the observed lag-1 correlation for each role
- whether each correlation is typical under simulations from the fitted model

Do not base the decision on a significance label alone.

> [Insert your diagnostic interpretation]

> [State whether you retain residual independence or add AR(1), and why]


## If selected: add role-specific AR(1)

If you decided that residual independence is adequate, skip this section.

Otherwise, add separate female and male AR(1) processes using `diaryday_f`.
Retain the dyad--day block because AR(1) dependence and same-occasion partner
covariance represent different parts of the residual structure.

```{r eval=FALSE}
# hint("ild_ar1", "role_specific_syntax")
# hint("ild_models", "distinguishable_ar1")

ild_ar1_start <- make_ild_ar1_start(distinguishable_no_ar_model)

distinguishable_ar1_model <- update(
  distinguishable_no_ar_model,
  formula = YOUR_AR1_FORMULA_UPDATE,
  start = ild_ar1_start
)
```

The start helper reuses the fitted no-AR coefficients and covariance estimates,
then inserts moderate interior AR values. It does not use the simulation truth.

Check estimation again. Then rerun the DHARMa and lag-1 checks using the
**AR(1) model object**, and decide whether the revised specification is
adequate.

```{r eval=FALSE}
ar1_dharma_residuals <- simulate_ild_dharma(
  model = distinguishable_ar1_model,
  dyad = my_ild_data$coupleID,
  member = my_ild_data$personID,
  time = my_ild_data$diaryday,
  role = my_ild_data$gender,
  n = 1000,
  seed = 123
)

plot(
  ar1_dharma_residuals,
  title = "AR(1) block-whitened DHARMa residuals"
)

test_ild_lag1(ar1_dharma_residuals, by_role = TRUE)
```

> [Insert your updated diagnostic interpretation]


## Retain the distinguishable temporal specification

Save the selected distinguishable model and its matching residual object under
the names below. Do not return to the no-AR objects if you selected AR(1).

```{r eval=FALSE}
distinguishable_model <- YOUR_SELECTED_DISTINGUISHABLE_MODEL
distinguishable_dharma_residuals <- YOUR_MATCHING_DHARMA_OBJECT
```


## Do the focal associations differ by role?

The cross-sectional exercise covered a global exchangeability test. Here, use a
more focused comparison: pool only the four within- and between-person
actor/partner associations. Keep role-specific intercepts, time trends, stable
covariance, same-occasion covariance, and both AR(1) processes unchanged.

This creates a regular four-degree-of-freedom test of one clear question:

> **Do the four focal APIM associations differ between women and men?**

Because the temporal and covariance terms are identical in both candidates,
the comparison does not mix that question with a change in residual structure.
The same fixed-effect update works whether you selected the no-AR or AR(1)
distinguishable model.

```{r eval=FALSE}
# hint("ild_models", "pooled_apim_effects")

pooled_apim_effects_start <- make_pooled_apim_effects_start(
  distinguishable_model
)

pooled_apim_effects_model <- update(
  distinguishable_model,
  formula = YOUR_POOLED_EFFECTS_FORMULA_UPDATE,
  start = pooled_apim_effects_start
)

summary(pooled_apim_effects_model)
```

Check estimation, then compare the matched models:

```{r eval=FALSE}
dyadMLM::compare_nested_models(
  distinguishable_model,
  pooled_apim_effects_model
)
```

Choose the model you will retain. Consider statistical evidence, parsimony, and
whether common actor and partner associations are substantively defensible.
Do not call the restricted model exchangeable: its baseline, covariance, and
AR(1) parameters remain role-specific.

> [Insert your role-difference conclusion and decision]

Use one stable name for all further processing and reporting:

```{r eval=FALSE}
retained_model <- YOUR_RETAINED_MODEL
```


## If pooled: recheck adequacy

If you retained `pooled_apim_effects_model`, repeat the DHARMa and lag-1 checks
with `retained_model`. If you retained the already-checked distinguishable
model, skip this step.

```{r eval=FALSE}
retained_model_dharma_residuals <- simulate_ild_dharma(
  model = retained_model,
  dyad = my_ild_data$coupleID,
  member = my_ild_data$personID,
  time = my_ild_data$diaryday,
  role = my_ild_data$gender,
  n = 1000,
  seed = 123
)

plot(
  retained_model_dharma_residuals,
  title = "Retained-model block-whitened residuals"
)

test_ild_lag1(retained_model_dharma_residuals, by_role = TRUE)
```

Briefly assess whether the pooled-effects model remains adequate:

> [Insert your conclusion, or state that you skipped this step]


## Robustness

If you fitted both no-AR and AR(1) candidates, compare their focal estimates and
state whether modeling temporal dependence materially changes the substantive
conclusions. If time permits, assess a different plausible time trend. Do not
select a specification solely because it changes a significance label.

> [Insert your robustness conclusion]

For a future nonnegative, right-skewed outcome, `sqrt()` or `log1p()` may
improve the scale; inspect its histogram, refit, and rerun all diagnostics. An
appropriate non-Gaussian family may instead be preferable, but generalized
families have different assumptions and require case-specific choices.


# 6. Visualize, interpret, and report results

Use `retained_model` for every table, plot, and interpretation below.

Print the fixed and random effects, and inspect the fitted covariance
parameters:

```{r}
# hint("ild_reporting", "solution")

```

Interpret each AR(1) SD as the stationary SD of that role's persistent latent
residual process. The dyad--day correlation describes only the remaining
nonpersistent same-occasion component; it is not the correlation of the
complete residual after the AR(1) components are added.

Plot the focal fixed effects or model-implied associations:

```{r}
# hint("ild_reporting", "plot")

```

Write a brief results paragraph that:

- describes the analysis sample and temporal structure
- explains the selected temporal specification
- explains the role-difference decision
- reports within-person and between-person actor and partner effects with
  estimates and 95% confidence intervals
- summarizes stable dyad covariance, same-occasion partner covariance, and any
  fitted AR(1) parameters
- reports the adequacy and robustness checks

Interpret the estimates as associations, not causal effects. If effects are
role-specific, a significant estimate for one role and a nonsignificant
estimate for the other do not by themselves establish that the roles differ.

```{r}
# hint("ild_reporting", "paragraph_template")
```

> [Insert your brief results paragraph]


You have now completed the ILD APIM workflow!
