---
title: "Exploratory APIM by signup position"
subtitle: "Lower-userID initiator versus higher-userID partner"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

library(dplyr)
library(ggplot2)
library(glmmTMB)
library(tidyr)
```

# Purpose

The workshop's distinguishable APIM uses gender to distinguish partners. This
exploratory analysis instead distinguishes partners by signup position:

- **Person 1 / initiator:** the member with the lower `userID`;
- **Person 2 / invited partner:** the member with the higher `userID`.

In these study data, the lower ID is also labelled `_A` and the higher ID `_B`.
The interpretation of `_A` as the signup initiator comes from knowledge of the
study's signup process; it is not a measured motivation variable. Consequently,
the analysis can test whether APIM parameters vary by signup position, but it
cannot establish that any difference is caused by motivation.

# Data and role construction

```{r read-data}
data_candidates <- c(
  "long.rds",
  file.path("dev", "workshop", "long.rds")
)
data_file <- data_candidates[file.exists(data_candidates)][1]

if (is.na(data_file)) {
  stop("Could not find dev/workshop/long.rds.", call. = FALSE)
}

workshop_raw <- readRDS(data_file)

members <- workshop_raw |>
  distinct(coupleID, userID, gender, which_partner) |>
  group_by(coupleID) |>
  arrange(userID, .by_group = TRUE) |>
  mutate(
    member_count = n(),
    signup_position = row_number(),
    signup_role = factor(
      if_else(signup_position == 1L, "initiator", "partner"),
      levels = c("initiator", "partner")
    )
  ) |>
  ungroup()

if (anyNA(members$userID) || any(members$member_count != 2L)) {
  stop("Each dyad must contain exactly two non-missing userIDs.", call. = FALSE)
}

role_matches_system_label <-
  (members$signup_role == "initiator" & members$which_partner == "_A") |
  (members$signup_role == "partner" & members$which_partner == "_B")

if (!all(role_matches_system_label)) {
  stop("Lower/higher userID does not consistently match _A/_B.", call. = FALSE)
}

role_validation <- members |>
  count(signup_role, which_partner, name = "members")

knitr::kable(
  role_validation,
  caption = "Validation of userID order against the study-system role label."
)
```

All `r n_distinct(members$coupleID)` dyads have one lower-ID `_A` member and one
higher-ID `_B` member. The role distinction therefore retains all dyads,
including one female--female and one male--male couple. By contrast, the
workshop's gender-distinguishable example retains only the 36 female--male
couples.

We construct daily MVPA exactly as in the conceptual workshop deck and then
aggregate the diary observations to person-level means.

```{r aggregate-data}
mean_or_na <- function(x) {
  if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
}

pilot_aggregated <- workshop_raw |>
  select(
    coupleID, userID, ss_pa, ss_pa_min_solo, ss_pa_min_collab,
    ss_emo_pleasure
  ) |>
  inner_join(
    members |>
      select(coupleID, userID, signup_role, gender),
    by = c("coupleID", "userID")
  ) |>
  transmute(
    couple_id = coupleID,
    person_id = userID,
    signup_role,
    gender = factor(gender, levels = c(1, 2), labels = c("female", "male")),
    provided_emotional_support = ss_emo_pleasure,
    mvpa_solo = if_else(ss_pa == 0, 0, ss_pa_min_solo),
    mvpa_joint = if_else(ss_pa == 0, 0, ss_pa_min_collab),
    mvpa_total = mvpa_solo + mvpa_joint
  ) |>
  group_by(couple_id, person_id, signup_role, gender) |>
  summarise(
    across(
      c(provided_emotional_support, mvpa_total),
      mean_or_na
    ),
    .groups = "drop"
  )

if (any(!is.finite(pilot_aggregated$provided_emotional_support)) ||
    any(!is.finite(pilot_aggregated$mvpa_total))) {
  stop("The aggregated APIM variables must be finite.", call. = FALSE)
}
```

## Descriptives and an important overlap

```{r descriptives}
role_descriptives <- pilot_aggregated |>
  group_by(signup_role) |>
  summarise(
    people = n(),
    support_mean = mean(provided_emotional_support),
    support_sd = sd(provided_emotional_support),
    mvpa_mean = mean(mvpa_total),
    mvpa_sd = sd(mvpa_total),
    .groups = "drop"
  )

role_by_gender <- pilot_aggregated |>
  count(signup_role, gender) |>
  pivot_wider(names_from = gender, values_from = n, values_fill = 0)

support_wide <- pilot_aggregated |>
  select(couple_id, signup_role, provided_emotional_support) |>
  pivot_wider(
    names_from = signup_role,
    values_from = provided_emotional_support
  )

partner_support_correlation <- with(
  support_wide,
  cor(initiator, partner, use = "complete.obs")
)

signup_gender_correlation <- with(
  pilot_aggregated,
  cor(as.integer(gender), as.integer(signup_role))
)

knitr::kable(
  role_descriptives,
  digits = 2,
  caption = "Person-level descriptives by signup position."
)

knitr::kable(
  role_by_gender,
  caption = "Signup position by gender."
)
```

Signup position is associated with gender in this small sample: Person 1 is
male in 24 of 38 dyads, whereas Person 2 is female in 24 of 38 dyads. The two
partners' average support scores are also correlated
($r = `r sprintf("%.2f", partner_support_correlation)`$). Role-specific APIM
coefficients should therefore be treated as exploratory conditional
associations, not as clean effects of signup position or motivation.

# Prepare the role-distinguishable APIM

`prepare_dyad_data()` now receives `signup_role` rather than gender. Both actor
and partner support scores are then centered using one pooled grand mean, just
as in the workshop's gender-distinguishable APIM.

```{r prepare-distinguishable-apim}
signup_apim_data <- dyadMLM::prepare_dyad_data(
  pilot_aggregated,
  dyad = couple_id,
  member = person_id,
  role = signup_role,
  predictors = provided_emotional_support,
  model_types = "apim",
  keep_compositions = "initiator-partner"
)

actor_term <- ".dy_provided_emotional_support_actor"
partner_term <- ".dy_provided_emotional_support_partner"
complete_predictor_pair <-
  is.finite(signup_apim_data[[actor_term]]) &
  is.finite(signup_apim_data[[partner_term]])

support_grand_mean <- mean(
  (
    signup_apim_data[[actor_term]][complete_predictor_pair] +
      signup_apim_data[[partner_term]][complete_predictor_pair]
  ) / 2
)

signup_apim_data[[actor_term]] <-
  signup_apim_data[[actor_term]] - support_grand_mean
signup_apim_data[[partner_term]] <-
  signup_apim_data[[partner_term]] - support_grand_mean

support_grand_mean
```

The common centering value is
`r sprintf("%.2f", support_grand_mean)` on the original support scale. Thus,
each role-specific intercept is predicted MVPA when both partners' support is
at this pooled sample mean.

# Fit the distinguishable APIM

The fixed effects and the complete two-person residual covariance structure
are allowed to differ by signup position.

```{r fit-distinguishable-apim}
person1_indicator <- ".dy_is_initiator_x_partner_initiator"
person2_indicator <- ".dy_is_initiator_x_partner_partner"

signup_distinguishable_apim <- glmmTMB(
  mvpa_total ~ 0 +
    .dy_is_initiator_x_partner_initiator +
    .dy_is_initiator_x_partner_partner +
    .dy_is_initiator_x_partner_initiator:
      .dy_provided_emotional_support_actor +
    .dy_is_initiator_x_partner_partner:
      .dy_provided_emotional_support_actor +
    .dy_is_initiator_x_partner_initiator:
      .dy_provided_emotional_support_partner +
    .dy_is_initiator_x_partner_partner:
      .dy_provided_emotional_support_partner +
    us(
      0 + .dy_is_initiator_x_partner_initiator +
        .dy_is_initiator_x_partner_partner | couple_id
    ),
  dispformula = ~0,
  family = gaussian(),
  data = signup_apim_data
)

summary(signup_distinguishable_apim)
```

# Extract results for the diagram

The workshop's shared distinguishable-APIM diagram extractor deliberately uses
female/male labels. To avoid changing that helper or mislabelling this model,
the extraction and signup-position diagram are defined locally in this file.

```{r extract-diagram-results}
signup_apim_terms <- c(
  intercept_person1 =
    ".dy_is_initiator_x_partner_initiator",
  intercept_person2 =
    ".dy_is_initiator_x_partner_partner",
  actor_person1 = paste0(person1_indicator, ":", actor_term),
  actor_person2 = paste0(person2_indicator, ":", actor_term),
  partner_effect_on_person1 =
    paste0(person1_indicator, ":", partner_term),
  partner_effect_on_person2 =
    paste0(person2_indicator, ":", partner_term)
)

extract_signup_apim_diagram_values <- function(
    model, terms, dyad_group, person1_random, person2_random) {
  coefficient_table <- summary(model)$coefficients$cond
  missing_terms <- setdiff(unname(terms), rownames(coefficient_table))

  if (length(missing_terms)) {
    stop(
      "Could not find model term(s): ", paste(missing_terms, collapse = ", "),
      call. = FALSE
    )
  }

  random_covariance <- glmmTMB::VarCorr(model)$cond[[dyad_group]]
  random_sds <- attr(random_covariance, "stddev")
  random_correlations <- attr(random_covariance, "correlation")

  random_terms <- c(person1_random, person2_random)
  if (!all(random_terms %in% names(random_sds))) {
    stop("Could not find both role-specific residual terms.", call. = FALSE)
  }

  list(
    estimates = setNames(
      coefficient_table[unname(terms), "Estimate"],
      names(terms)
    ),
    standard_errors = setNames(
      coefficient_table[unname(terms), "Std. Error"],
      names(terms)
    ),
    p_values = setNames(
      coefficient_table[unname(terms), "Pr(>|z|)"],
      names(terms)
    ),
    residuals = c(
      sd_person1 = unname(random_sds[[person1_random]]),
      sd_person2 = unname(random_sds[[person2_random]]),
      correlation = unname(
        random_correlations[person1_random, person2_random]
      )
    ),
    terms = terms
  )
}

signup_diagram_values <- extract_signup_apim_diagram_values(
  signup_distinguishable_apim,
  terms = signup_apim_terms,
  dyad_group = "couple_id",
  person1_random = person1_indicator,
  person2_random = person2_indicator
)

format_p_value <- function(p_value) {
  format.pval(p_value, digits = 3, eps = .001)
}

fixed_effect_table <- tibble(
  parameter = c(
    "Intercept: Person 1",
    "Intercept: Person 2",
    "Actor effect: Person 1",
    "Actor effect: Person 2",
    "Partner effect on Person 1 outcome",
    "Partner effect on Person 2 outcome"
  ),
  estimate = unname(signup_diagram_values$estimates),
  standard_error = unname(signup_diagram_values$standard_errors),
  p_value = format_p_value(unname(signup_diagram_values$p_values))
)

residual_table <- tibble(
  parameter = c(
    "Residual SD: Person 1",
    "Residual SD: Person 2",
    "Residual correlation"
  ),
  estimate = unname(signup_diagram_values$residuals)
)

knitr::kable(
  fixed_effect_table,
  digits = 3,
  caption = "Fixed effects extracted for the signup-position APIM diagram."
)

knitr::kable(
  residual_table,
  digits = 3,
  caption = "Residual quantities extracted for the diagram."
)
```

## Fitted signup-position diagram

```{r fitted-signup-position-diagram, echo=FALSE, fig.width=11, fig.height=6}
significance_stars <- function(p_value) {
  if (!is.finite(p_value)) return("")
  if (p_value < .001) return("***")
  if (p_value < .01) return("**")
  if (p_value < .05) return("*")
  ""
}

draw_signup_apim_diagram <- function(values) {
  estimates <- values$estimates
  p_values <- values$p_values
  residuals <- values$residuals

  actor_colour <- "#1769AA"
  partner_colour <- "#7B3FA1"
  residual_colour <- "#64748B"
  ink_colour <- "#263238"

  old_parameters <- par(no.readonly = TRUE)
  on.exit(par(old_parameters), add = TRUE)
  par(mar = c(0.3, 0.3, 2.1, 0.3), xpd = NA)

  plot.new()
  plot.window(xlim = c(0, 10.7), ylim = c(0, 10))

  node <- function(x, y, label, width = 2.3, height = 1.15) {
    rect(
      x - width / 2, y - height / 2,
      x + width / 2, y + height / 2,
      col = "#F8FAFC", border = ink_colour, lwd = 1.5
    )
    text(x, y, label, col = ink_colour, cex = 0.9)
  }

  edge_label <- function(x, y, label, colour) {
    label_width <- strwidth(label, cex = 0.82, font = 2)
    label_height <- strheight(label, cex = 0.82, font = 2)
    rect(
      x - label_width / 2 - 0.06, y - label_height / 2 - 0.07,
      x + label_width / 2 + 0.06, y + label_height / 2 + 0.07,
      col = "white", border = NA
    )
    text(x, y, label, col = colour, cex = 0.82, font = 2)
  }

  formatted_effect <- function(label, parameter) {
    paste0(
      label, " = ", sprintf("%.2f", estimates[[parameter]]),
      significance_stars(p_values[[parameter]])
    )
  }

  x_predictor <- 1.55
  x_outcome <- 6.45
  x_residual <- 9.05
  y_person1 <- 7.45
  y_person2 <- 2.55

  node(
    x_predictor, y_person1,
    "Provided support\nPerson 1 (GM-centered)", width = 2.8
  )
  node(
    x_predictor, y_person2,
    "Provided support\nPerson 2 (GM-centered)", width = 2.8
  )
  node(x_outcome, y_person1, "MVPA\nPerson 1")
  node(x_outcome, y_person2, "MVPA\nPerson 2")
  node(
    x_residual, y_person1,
    paste0("Residual 1\nSD = ", sprintf("%.2f", residuals[["sd_person1"]])),
    width = 1.75
  )
  node(
    x_residual, y_person2,
    paste0("Residual 2\nSD = ", sprintf("%.2f", residuals[["sd_person2"]])),
    width = 1.75
  )

  arrows(
    2.98, y_person1, 5.24, y_person1,
    length = 0.10, lwd = 2.2, col = actor_colour
  )
  arrows(
    2.98, y_person2, 5.24, y_person2,
    length = 0.10, lwd = 2.2, col = actor_colour
  )
  arrows(
    2.88, y_person2 + 0.35, 5.30, y_person1 - 0.35,
    length = 0.10, lwd = 2.2, col = partner_colour
  )
  arrows(
    2.88, y_person1 - 0.35, 5.30, y_person2 + 0.35,
    length = 0.10, lwd = 2.2, col = partner_colour
  )

  arrows(
    8.15, y_person1, 7.65, y_person1,
    length = 0.09, lwd = 1.7, col = residual_colour
  )
  arrows(
    8.15, y_person2, 7.65, y_person2,
    length = 0.09, lwd = 1.7, col = residual_colour
  )
  arrows(
    x_residual, y_person2 + 0.62,
    x_residual, y_person1 - 0.62,
    code = 3, length = 0.09, lwd = 1.7, col = residual_colour
  )

  edge_label(
    4.10, 7.88,
    formatted_effect("Actor P1", "actor_person1"), actor_colour
  )
  edge_label(
    4.10, 2.12,
    formatted_effect("Actor P2", "actor_person2"), actor_colour
  )
  edge_label(
    4.12, 6.15,
    formatted_effect("Partner effect on P1", "partner_effect_on_person1"),
    partner_colour
  )
  edge_label(
    4.12, 3.85,
    formatted_effect("Partner effect on P2", "partner_effect_on_person2"),
    partner_colour
  )

  text(
    x_outcome, 8.30,
    paste0("Intercept P1 = ", sprintf("%.2f", estimates[["intercept_person1"]])),
    col = ink_colour, cex = 0.85
  )
  text(
    x_outcome, 1.70,
    paste0("Intercept P2 = ", sprintf("%.2f", estimates[["intercept_person2"]])),
    col = ink_colour, cex = 0.85
  )
  text(
    9.85, 5,
    paste0("Residual r = ", sprintf("%.2f", residuals[["correlation"]])),
    col = residual_colour, cex = 0.85, srt = 90
  )

  title(
    "Distinguishable APIM by signup position",
    col.main = ink_colour, font.main = 2
  )
}

draw_signup_apim_diagram(signup_diagram_values)
```

Person 1 and Person 2 refer to lower and higher `userID`, respectively. Actor
effects link each member's support to their own MVPA. Each partner effect is
named by the **outcome member**: for example, "partner effect on P1" links
Person 2's support to Person 1's MVPA. Stars use the model's asymptotic Wald
tests.

# Which role-specific coefficients differ?

The following Wald contrasts isolate the three fixed-effect equality
restrictions. They are more informative than inspecting whether one
role-specific coefficient is significant and the other is not.

```{r role-contrasts}
fixed_estimates <- glmmTMB::fixef(signup_distinguishable_apim)$cond
fixed_covariance <- vcov(signup_distinguishable_apim)$cond

contrast_matrix <- rbind(
  "Intercept: Person 1 - Person 2" = c(1, -1, 0, 0, 0, 0),
  "Actor effect: Person 1 - Person 2" = c(0, 0, 1, -1, 0, 0),
  "Partner effect: on Person 1 - on Person 2" = c(0, 0, 0, 0, 1, -1)
)
colnames(contrast_matrix) <- unname(signup_apim_terms)

ordered_estimates <- fixed_estimates[unname(signup_apim_terms)]
ordered_covariance <- fixed_covariance[
  unname(signup_apim_terms), unname(signup_apim_terms), drop = FALSE
]

contrast_estimates <- drop(contrast_matrix %*% ordered_estimates)
contrast_ses <- sqrt(diag(
  contrast_matrix %*% ordered_covariance %*% t(contrast_matrix)
))
contrast_z <- contrast_estimates / contrast_ses

role_contrast_table <- tibble(
  contrast = rownames(contrast_matrix),
  estimate = contrast_estimates,
  standard_error = contrast_ses,
  z = contrast_z,
  p_value = format_p_value(
    2 * pnorm(abs(contrast_z), lower.tail = FALSE)
  )
)

knitr::kable(
  role_contrast_table,
  digits = 3,
  caption = "Wald contrasts between signup-position-specific fixed effects."
)
```

# Compare with an exchangeable APIM

To ask whether signup position is jointly informative, we compare the model
above with an exchangeable APIM fitted to the **same 38 dyads**. The reduced
model constrains the intercept, actor effect, partner effect, and residual
variance to equality across positions while retaining within-couple residual
correlation.

```{r fit-exchangeable-apim}
signup_exchangeable_data <- dyadMLM::prepare_dyad_data(
  pilot_aggregated,
  dyad = couple_id,
  member = person_id,
  role = signup_role,
  predictors = provided_emotional_support,
  model_types = "apim",
  keep_compositions = "initiator-partner",
  set_exchangeable_compositions = "initiator-partner",
  seed = 123
)

signup_exchangeable_data[[actor_term]] <-
  signup_exchangeable_data[[actor_term]] - support_grand_mean
signup_exchangeable_data[[partner_term]] <-
  signup_exchangeable_data[[partner_term]] - support_grand_mean

signup_exchangeable_apim <- glmmTMB(
  mvpa_total ~ 0 +
    .dy_is_initiator_x_partner +
    .dy_provided_emotional_support_actor +
    .dy_provided_emotional_support_partner +
    us(0 + .dy_is_initiator_x_partner | couple_id) +
    us(
      0 + .dy_member_contrast_initiator_x_partner_arbitrary | couple_id
    ),
  dispformula = ~0,
  family = gaussian(),
  data = signup_exchangeable_data
)

model_comparison <- dyadMLM::compare_nested_glmmTMB_models(
  signup_exchangeable_apim,
  signup_distinguishable_apim
)

model_comparison
```

```{r information-criteria}
fit_table <- tibble(
  model = c("Exchangeable", "Distinguishable by signup position"),
  parameters = c(
    attr(logLik(signup_exchangeable_apim), "df"),
    attr(logLik(signup_distinguishable_apim), "df")
  ),
  log_likelihood = c(
    as.numeric(logLik(signup_exchangeable_apim)),
    as.numeric(logLik(signup_distinguishable_apim))
  ),
  AIC = c(AIC(signup_exchangeable_apim), AIC(signup_distinguishable_apim)),
  BIC = c(BIC(signup_exchangeable_apim), BIC(signup_distinguishable_apim))
)

knitr::kable(fit_table, digits = 2, caption = "Model-fit comparison.")
```

```{r comparison-values, include=FALSE}
comparison_loglik <- c(
  exchangeable = as.numeric(logLik(signup_exchangeable_apim)),
  distinguishable = as.numeric(logLik(signup_distinguishable_apim))
)
comparison_df <- c(
  exchangeable = attr(logLik(signup_exchangeable_apim), "df"),
  distinguishable = attr(logLik(signup_distinguishable_apim), "df")
)
lrt_chisq <- 2 * diff(comparison_loglik)
lrt_df <- diff(comparison_df)
lrt_p <- pchisq(lrt_chisq, df = lrt_df, lower.tail = FALSE)
delta_aic <- AIC(signup_exchangeable_apim) - AIC(signup_distinguishable_apim)
delta_bic <- BIC(signup_distinguishable_apim) - BIC(signup_exchangeable_apim)
```

# Separate gender from signup-position moderation

The position-distinguishable model alone cannot show whether its slope
differences merely reflect the association between signup position and gender.
We can address this more directly by including both moderators in one model:

```{r combined-model-formula, eval=FALSE}
mvpa_total ~
  actor * (gender + signup_role) +
  partner * (gender + signup_role) +
  residual_covariance
```

This expands to the actor and partner main effects, the gender and signup-role
main effects, and four two-way interactions. Thus, each gender interaction is
adjusted for signup-position moderation and each signup-position interaction
is adjusted for gender moderation. Here gender is the **outcome member's**
gender, matching the convention in the workshop's distinguishable APIM.

We use the same exchangeable residual covariance in every model below. This
avoids privileging either gender or signup position in the residual structure,
and ensures that the comparisons isolate fixed-effect slope moderation.

```{r fit-combined-moderation-apims}
combined_full_formula <-
  mvpa_total ~
  .dy_provided_emotional_support_actor * (gender + signup_role) +
  .dy_provided_emotional_support_partner * (gender + signup_role) +
  us(0 + .dy_is_initiator_x_partner | couple_id) +
  us(
    0 + .dy_member_contrast_initiator_x_partner_arbitrary | couple_id
  )

# Retain the gender main effect but omit both gender slope interactions.
combined_signup_only_formula <-
  mvpa_total ~
  .dy_provided_emotional_support_actor * signup_role +
  .dy_provided_emotional_support_partner * signup_role +
  gender +
  us(0 + .dy_is_initiator_x_partner | couple_id) +
  us(
    0 + .dy_member_contrast_initiator_x_partner_arbitrary | couple_id
  )

# Retain the signup-position main effect but omit both position interactions.
combined_gender_only_formula <-
  mvpa_total ~
  .dy_provided_emotional_support_actor * gender +
  .dy_provided_emotional_support_partner * gender +
  signup_role +
  us(0 + .dy_is_initiator_x_partner | couple_id) +
  us(
    0 + .dy_member_contrast_initiator_x_partner_arbitrary | couple_id
  )

# Retain both main effects but omit all four slope interactions.
combined_no_moderation_formula <-
  mvpa_total ~
  .dy_provided_emotional_support_actor +
  .dy_provided_emotional_support_partner +
  gender + signup_role +
  us(0 + .dy_is_initiator_x_partner | couple_id) +
  us(
    0 + .dy_member_contrast_initiator_x_partner_arbitrary | couple_id
  )

fit_combined_apim <- function(formula, data) {
  glmmTMB(
    formula,
    dispformula = ~0,
    family = gaussian(),
    data = data
  )
}

combined_full_apim <- fit_combined_apim(
  combined_full_formula, signup_exchangeable_data
)
combined_signup_only_apim <- fit_combined_apim(
  combined_signup_only_formula, signup_exchangeable_data
)
combined_gender_only_apim <- fit_combined_apim(
  combined_gender_only_formula, signup_exchangeable_data
)
combined_no_moderation_apim <- fit_combined_apim(
  combined_no_moderation_formula, signup_exchangeable_data
)

combined_models <- list(
  combined_full_apim,
  combined_signup_only_apim,
  combined_gender_only_apim,
  combined_no_moderation_apim
)
combined_models_converged <- vapply(
  combined_models,
  function(model) model$fit$convergence == 0L && isTRUE(model$sdr$pdHess),
  logical(1)
)

if (!all(combined_models_converged)) {
  warning("At least one combined APIM has a convergence or Hessian problem.")
}
```

## Unique interaction coefficients

Female Person 1 members are the reference group under R's default treatment
coding. The main effects are included to respect model hierarchy. The four
interaction coefficients are the unique slope-moderation estimates of primary
interest.

```{r combined-fixed-effects}
combined_coefficients <- summary(combined_full_apim)$coefficients$cond

combined_fixed_table <- tibble(
  term = rownames(combined_coefficients),
  estimate = combined_coefficients[, "Estimate"],
  standard_error = combined_coefficients[, "Std. Error"],
  p_value = format_p_value(combined_coefficients[, "Pr(>|z|)"])
) |>
  mutate(
    parameter = case_when(
      term == "(Intercept)" ~ "Intercept: female Person 1",
      term == actor_term ~ "Actor effect: female Person 1",
      term == partner_term ~ "Partner effect: female Person 1 outcome",
      term == "gendermale" ~ "Main effect: male versus female",
      term == "signup_rolepartner" ~ "Main effect: Person 2 versus Person 1",
      grepl(actor_term, term, fixed = TRUE) &
        grepl("gendermale", term, fixed = TRUE) ~
          "Actor x gender (male - female)",
      grepl(actor_term, term, fixed = TRUE) &
        grepl("signup_rolepartner", term, fixed = TRUE) ~
          "Actor x signup position (Person 2 - Person 1)",
      grepl(partner_term, term, fixed = TRUE) &
        grepl("gendermale", term, fixed = TRUE) ~
          "Partner x gender (male - female outcome)",
      grepl(partner_term, term, fixed = TRUE) &
        grepl("signup_rolepartner", term, fixed = TRUE) ~
          "Partner x signup position (Person 2 - Person 1 outcome)",
      TRUE ~ term
    )
  ) |>
  select(parameter, estimate, standard_error, p_value)

knitr::kable(
  combined_fixed_table,
  digits = 3,
  caption = paste(
    "Combined gender and signup-position APIM.",
    "Female Person 1 members are the reference group."
  )
)
```

Because this model assumes additive moderation, the gender difference in each
slope is held constant across signup positions, and the signup-position
difference is held constant across genders. A three-way interaction would relax
that assumption, but would be ambitious with only 38 dyads.

The fitted slopes for the four observed gender-by-position combinations are:

```{r combined-simple-slopes}
combined_estimates <- glmmTMB::fixef(combined_full_apim)$cond
combined_names <- names(combined_estimates)

find_interaction_term <- function(first, second, available_terms) {
  candidates <- c(
    paste(first, second, sep = ":"),
    paste(second, first, sep = ":")
  )
  match <- intersect(candidates, available_terms)
  if (length(match) != 1L) {
    stop(
      "Could not uniquely resolve interaction between ", first,
      " and ", second, ".",
      call. = FALSE
    )
  }
  match
}

actor_gender_term <- find_interaction_term(
  actor_term, "gendermale", combined_names
)
actor_position_term <- find_interaction_term(
  actor_term, "signup_rolepartner", combined_names
)
partner_gender_term <- find_interaction_term(
  partner_term, "gendermale", combined_names
)
partner_position_term <- find_interaction_term(
  partner_term, "signup_rolepartner", combined_names
)

combined_simple_slopes <- tidyr::expand_grid(
  gender = c("female", "male"),
  signup_position = c("Person 1", "Person 2")
) |>
  mutate(
    predicted_mvpa_at_mean_support =
      unname(combined_estimates[["(Intercept)"]]) +
      (gender == "male") * unname(combined_estimates[["gendermale"]]) +
      (signup_position == "Person 2") *
        unname(combined_estimates[["signup_rolepartner"]]),
    actor_effect =
      unname(combined_estimates[[actor_term]]) +
      (gender == "male") * unname(combined_estimates[[actor_gender_term]]) +
      (signup_position == "Person 2") *
        unname(combined_estimates[[actor_position_term]]),
    partner_effect_on_outcome =
      unname(combined_estimates[[partner_term]]) +
      (gender == "male") * unname(combined_estimates[[partner_gender_term]]) +
      (signup_position == "Person 2") *
        unname(combined_estimates[[partner_position_term]])
  )

knitr::kable(
  combined_simple_slopes,
  digits = 2,
  caption = "Point estimates implied by the additive combined model."
)
```

## Joint tests: gender or signup position?

Seeing one interaction reach significance and another not does not itself test
whether the moderators differ. Each moderator also contributes two interaction
terms--one for the actor slope and one for the partner slope. We therefore
remove both terms for one moderator at a time and compare that reduced model
with the full model. Both main effects remain in every comparison.

```{r conditional-moderation-tests}
likelihood_ratio_row <- function(reduced, full, test, sample) {
  reduced_loglik <- logLik(reduced)
  full_loglik <- logLik(full)
  chi_square <- 2 * (
    as.numeric(full_loglik) - as.numeric(reduced_loglik)
  )
  degrees_freedom <-
    attr(full_loglik, "df") - attr(reduced_loglik, "df")

  tibble(
    sample,
    test,
    chi_square,
    df = degrees_freedom,
    p_value = pchisq(
      chi_square, df = degrees_freedom, lower.tail = FALSE
    )
  )
}

all_dyad_moderation_tests <- bind_rows(
  likelihood_ratio_row(
    combined_signup_only_apim,
    combined_full_apim,
    "Gender interactions beyond signup position",
    "All 38 dyads"
  ),
  likelihood_ratio_row(
    combined_gender_only_apim,
    combined_full_apim,
    "Signup-position interactions beyond gender",
    "All 38 dyads"
  ),
  likelihood_ratio_row(
    combined_no_moderation_apim,
    combined_full_apim,
    "All four slope interactions jointly",
    "All 38 dyads"
  )
)

combined_fit_table <- tibble(
  model = c(
    "No slope moderation",
    "Signup-position moderation only",
    "Gender moderation only",
    "Both moderators"
  ),
  fitted_model = list(
    combined_no_moderation_apim,
    combined_signup_only_apim,
    combined_gender_only_apim,
    combined_full_apim
  )
) |>
  mutate(
    parameters = vapply(
      fitted_model, function(model) attr(logLik(model), "df"), numeric(1)
    ),
    log_likelihood = vapply(
      fitted_model, function(model) as.numeric(logLik(model)), numeric(1)
    ),
    AIC = vapply(fitted_model, AIC, numeric(1)),
    BIC = vapply(fitted_model, BIC, numeric(1))
  ) |>
  select(-fitted_model)

knitr::kable(
  all_dyad_moderation_tests |>
    mutate(p_value = format_p_value(p_value)),
  digits = 3,
  caption = "Conditional likelihood-ratio tests of slope moderation."
)

knitr::kable(
  combined_fit_table,
  digits = 2,
  caption = "Fit of the combined moderation models for all 38 dyads."
)
```

The full model converged with a positive-definite Hessian. Conditional on the
signup-position interactions, the two gender interactions do not jointly
improve fit. Conditional on the gender interactions, the two signup-position
interactions do improve fit.

## Sensitivity using the workshop's 36 female--male dyads

The combined analysis above can validly include the two same-gender couples
because gender is a person-level moderator. Nevertheless, the original
gender-distinguishable workshop APIM uses only female--male dyads. We therefore
repeat the two conditional tests on those same 36 dyads.

```{r mixed-gender-sensitivity}
mixed_gender_couples <- pilot_aggregated |>
  group_by(couple_id) |>
  summarise(genders = n_distinct(gender), .groups = "drop") |>
  filter(genders == 2L) |>
  pull(couple_id)

combined_mixed_data <- signup_exchangeable_data |>
  filter(couple_id %in% mixed_gender_couples)

# Recenter both APIM predictors on the pooled mean of this sensitivity sample.
mixed_centering_offset <- mean(
  (
    combined_mixed_data[[actor_term]] +
      combined_mixed_data[[partner_term]]
  ) / 2
)
combined_mixed_data[[actor_term]] <-
  combined_mixed_data[[actor_term]] - mixed_centering_offset
combined_mixed_data[[partner_term]] <-
  combined_mixed_data[[partner_term]] - mixed_centering_offset

combined_mixed_full_apim <- fit_combined_apim(
  combined_full_formula, combined_mixed_data
)
combined_mixed_signup_only_apim <- fit_combined_apim(
  combined_signup_only_formula, combined_mixed_data
)
combined_mixed_gender_only_apim <- fit_combined_apim(
  combined_gender_only_formula, combined_mixed_data
)

mixed_gender_moderation_tests <- bind_rows(
  likelihood_ratio_row(
    combined_mixed_signup_only_apim,
    combined_mixed_full_apim,
    "Gender interactions beyond signup position",
    "36 female-male dyads"
  ),
  likelihood_ratio_row(
    combined_mixed_gender_only_apim,
    combined_mixed_full_apim,
    "Signup-position interactions beyond gender",
    "36 female-male dyads"
  )
)

all_moderation_tests <- bind_rows(
  all_dyad_moderation_tests,
  mixed_gender_moderation_tests
)

knitr::kable(
  mixed_gender_moderation_tests |>
    mutate(p_value = format_p_value(p_value)),
  digits = 3,
  caption = "Sensitivity analysis using the 36 female--male dyads."
)
```

The sensitivity analysis gives the same qualitative answer, so the result is
not driven by the two same-gender couples.

```{r combined-inline-values, include=FALSE}
all_gender_test <- all_moderation_tests |>
  filter(
    sample == "All 38 dyads",
    test == "Gender interactions beyond signup position"
  )
all_signup_test <- all_moderation_tests |>
  filter(
    sample == "All 38 dyads",
    test == "Signup-position interactions beyond gender"
  )
mixed_gender_test <- all_moderation_tests |>
  filter(
    sample == "36 female-male dyads",
    test == "Gender interactions beyond signup position"
  )
mixed_signup_test <- all_moderation_tests |>
  filter(
    sample == "36 female-male dyads",
    test == "Signup-position interactions beyond gender"
  )

combined_actor_gender_estimate <-
  unname(combined_estimates[[actor_gender_term]])
combined_actor_gender_p <-
  combined_coefficients[actor_gender_term, "Pr(>|z|)"]
combined_partner_gender_estimate <-
  unname(combined_estimates[[partner_gender_term]])
combined_partner_gender_p <-
  combined_coefficients[partner_gender_term, "Pr(>|z|)"]
combined_actor_position_estimate <-
  unname(combined_estimates[[actor_position_term]])
combined_actor_position_p <-
  combined_coefficients[actor_position_term, "Pr(>|z|)"]
combined_partner_position_estimate <-
  unname(combined_estimates[[partner_position_term]])
combined_partner_position_p <-
  combined_coefficients[partner_position_term, "Pr(>|z|)"]
```

# Exploratory conclusion

The joint comparison favours allowing parameters to vary by signup position,
$\chi^2(`r lrt_df`) = `r sprintf("%.2f", lrt_chisq)`$,
$p = `r sprintf("%.3f", lrt_p)`$. AIC also favours the role-distinguishable
model by `r sprintf("%.2f", delta_aic)` points, whereas BIC favours the simpler
exchangeable model by `r sprintf("%.2f", delta_bic)` points. Thus the evidence
is interesting but not unequivocal. The pronounced role differences are in the
actor and partner slopes rather than in predicted MVPA levels or residual
standard deviations:

- higher support from Person 1 is positively associated with both Person 1's
  own MVPA and Person 2's MVPA;
- higher support from Person 2 is negatively associated with both Person 2's
  own MVPA and Person 1's MVPA;
- the two intercepts and residual SDs are very similar.

This is intriguing enough to discuss as an alternative distinguishability
scheme. It is not, however, evidence that signup motivation produces these
associations. Signup position has a modest association with gender: Person 1
includes 14 women and 24 men, whereas Person 2 includes 24 women and 14 men.
Coding female = 1, male = 2, Person 1 = 1, and Person 2 = 2 gives
$r = `r sprintf("%.3f", signup_gender_correlation)`$ (equivalently,
$\phi = V = `r sprintf("%.3f", abs(signup_gender_correlation))`$). This is not
near-collinearity, so the two moderator sets can be estimated together.

In the full combined model, both signup-position interactions remain clearly
different from zero after accounting for gender (actor:
$b = `r sprintf("%.2f", combined_actor_position_estimate)`$,
$p = `r sprintf("%.3f", combined_actor_position_p)`$; partner:
$b = `r sprintf("%.2f", combined_partner_position_estimate)`$,
$p = `r sprintf("%.3f", combined_partner_position_p)`$). Neither gender
interaction remains clearly different from zero after accounting for signup
position (actor: $b = `r sprintf("%.2f", combined_actor_gender_estimate)`$,
$p = `r sprintf("%.3f", combined_actor_gender_p)`$; partner:
$b = `r sprintf("%.2f", combined_partner_gender_estimate)`$,
$p = `r sprintf("%.3f", combined_partner_gender_p)`$).

The joint tests give the same answer. Signup-position moderation remains after
accounting for gender, $\chi^2(`r all_signup_test$df`) =
`r sprintf("%.2f", all_signup_test$chi_square)`$,
$p = `r sprintf("%.3f", all_signup_test$p_value)`$, whereas gender moderation
does not remain after accounting for signup position,
$\chi^2(`r all_gender_test$df`) =
`r sprintf("%.2f", all_gender_test$chi_square)`$,
$p = `r sprintf("%.3f", all_gender_test$p_value)`$. Restricting the analysis to
the same 36 female--male dyads leads to the same conclusion (signup position:
$p = `r sprintf("%.3f", mixed_signup_test$p_value)`$; gender:
$p = `r sprintf("%.3f", mixed_gender_test$p_value)`$). Statistically, the slope
differences therefore align more clearly with signup position than with gender
in this sample.

This identifies unique conditional associations, not causal responsibility.
The sample contains only 38 dyads, the partners' support scores are strongly
correlated, and signup position is only a proxy for the hypothesized difference
in motivation. The combined model also assumes additive two-way moderation;
there is not enough information here for a convincing gender-by-position
three-way moderation analysis.
