---
title: "Dyadic Data Analysis with R"
subtitle: "Part II: Applied Tutorial"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
csl: apa.csl
resources:
  - uzh-logo.svg
  - charite-logo.svg
format:
  revealjs:
    embed-resources: true
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: false
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup-applied-tutorial
#| include: false

library(dplyr)
library(ggplot2)

slide_table <- function(x, digits = 2, ...) {
  knitr::kable(
    x,
    format = "html",
    digits = digits,
    format.args = list(
      nsmall = digits,
      scientific = FALSE
    ),
    row.names = FALSE,
    ...
  )
}

diagram_helper_candidates <- c(
  "diagram-helpers.Rinc",
  file.path("vignettes", "diagram-helpers.Rinc"),
  file.path("..", "..", "vignettes", "diagram-helpers.Rinc")
)
diagram_helper <- diagram_helper_candidates[
  file.exists(diagram_helper_candidates)
][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(
  normalizePath(diagram_helper),
  envir = knitr::knit_global()
)

```

## Today's aims and program

::: {.fancy-table}

| Time | Focus |
|---|---|
| [11:00–12:45]{.agenda-completed} | [Introduction and types of dyadic models]{.agenda-completed} |
| [12:45–14:00]{.agenda-completed} | [*Lunch break*]{.agenda-completed} |
| 14:00–15:30 | Estimating cross-sectional dyadic models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

:::

<!-- ca. 2 Minutes -->

# Software and frameworks

## SEM vs. MLM {.smaller}

- Many simple dyadic models can be represented in either framework.
- The best framework is the one that represents your conceptual model most naturally [@ledermannAnalyzingDyadicData2017].

:::: {.columns .fragment}

::: {.column width="50%"}

**SEM / DSEM is often more natural for**

- Latent dyadic constructs (e.g., a CFM)
- Item-level measurement models for reflective measures
- Linked outcomes or equations [e.g., mediation; @ledermannMediationDyadicData2009]
- Short dynamic panels in wide-format [e.g., LD-APIM; @gistelinckMultilevelAutoregressiveSmall2021]
- Intensive lagged or coupled dynamics via DSEM/RDSEM [e.g., @asparouhovDynamicStructuralEquation2018]

:::

::: {.column width="50%"}

**MLM is often more natural for**

- Parsimonious observed-variable models with modest numbers of dyads
- Nested grouping and flexible random-effect structures
- Observed-outcome growth or concurrent ILD models with flexible time trends
- Complex observed-variable regressions—interactions, polynomials, splines,
  and dyadic RSA—especially with repeated measures or additional nesting
- Specialized generalized mixed models in R, including zero-inflated, hurdle,
  and Tweedie models

:::

::::

::: {.subtle}
[@asparouhovDynamicStructuralEquation2018;
@asparouhovComparisonModelsAnalysis2020;
@driverContinuousTimeStructural2017;
@gistelinckMultilevelAutoregressiveSmall2021;
@delrosarioPracticalGuideSpecifying2025;
@schonbrodtTestingSimilarityEffects2018;
@nestlerResponseSurfaceAnalysis2019;
@brooksGlmmTMBBalancesSpeed2017]
:::

## SEM vs. MLM: Missing data {.smaller}

:::{.fragment}
- **MLM: Complete-case**
  - **Outcome missing:** Valid under MAR, MLM uses remaining observations
  - **Predictor missing:** Valid under MAR, if missingness does **not depend** on the outcome after accounting for predictors
  
:::

:::{.fragment}
- **SEM with FIML and joint likelihood**
  - Valid under MAR: Models predictors and outcomes jointly

:::

:::{.fragment}
- **MI: Imputation**
  - Handles missing outcomes, predictors, and auxiliary information
  - Must match the analysis and preserve person, dyad, role, and time structure

:::

::: {.fragment}
Include important predictors of missingness and missing values. FIML and MI
cover more MAR situations than row omission; none automatically solves MNAR.

:::

[@ledermannAnalyzingDyadicData2017; @hughesAccountingMissingData2019; @grundMultipleImputationMissing2018]{.subtle}

<!-- ca. 2.5 Minutes -->

## Why R {.smaller}

:::: {.columns}

::: {.column width="46%"}

### Why we teach R

:::{.fragment}
- **Free to use:** no license fees or institutional access barriers
- **Open source:** statistical implementations can be inspected, audited, and adapted
- **Reproducible:** workflows can be rerun, reviewed, shared, and version-controlled
- **Extensible:** specialized methods integrate with data, graphics, and reporting
:::

:::

::: {.column width="54%"}

### Current trade-offs

:::{.fragment}

- **SEM/DSEM:** R's capabilities are spread across specialized packages
  - (Multilevel) SEM: `lavaan` (e.g., CLPM/RI-CLPM), but has no native DSEM or     residual AR(1) structure. Niche packages for DSEM or continuous time SEM       `ctsem` exist, but no complete package can rival Mplus. 
- **MLM/ILD:** `glmmTMB` and `brms` both model serial dependence
  - Neither directly offers a separable partner-by-time residual structure such
    as $\mathrm{UN}\otimes\mathrm{AR}(1)$; SAS `PROC MIXED` and SPSS `MIXED` do
    for Gaussian repeated-measures models

<!--TODO: cite the ctsem package -->
:::

:::

::::

::: {.subtle}
[@lavaanMultilevelSEM;
@mulderThreeExtensions2021;
@blavaanTwoLevelSEM;
@koslowskiMlts2025;
@asparouhovDynamicStructuralEquation2018;
@asparouhovComparisonModelsAnalysis2020;
@kristensenCovarianceStructuresGlmmTMB2023;
@buerknerAutocorrelationStructuresBrms;
@sasInstituteMixedProcedure;
@ibmCovarianceStructureList]
:::

<!-- ca. 2.5 Minutes -->

# Estimating cross-sectional dyadic models in R

## One possible workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. **Verify model adequacy, distinguishability (if applicable), and robustness**
6. **Visualize, interpret, and report results**

::: {.subtle}
Synthesized from @kenny2006dyadic, @ledermannToolboxProgramsRestructure2015, and @stasGivingDyadicData2018.
:::

:::{.notes}

- In this example we will use the same dataset that we already used in Part I

:::

## One possible workflow

1. [**Define dyads, distinguishability, and research questions**]{.text-highlight}
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. **Verify model adequacy, distinguishability (if applicable), and robustness**
6. **Visualize, interpret, and report results**

::: {.subtle}
Synthesized from @kenny2006dyadic, @ledermannToolboxProgramsRestructure2015, and @stasGivingDyadicData2018.
:::


## The Dataset

**Same as before: **

- 36 physically inactive romantic couples
- Both partners intended to become more physically active
- Smartphone-based planning interventions and JITAIs
- Daily diaries over 55 days

**Aggregated for cross-sectional analyses**


## Research Question {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ:]{.rq-label} [In romantic couples, is a person's **self-efficacy** linked to physical activity in both partners?]{.rq-text}
:::
:::

## Measurement - predictor

### Self-efficacy

> **Rough English translation of the original German item**<br>
> “I am confident that I can be physically active tomorrow, even if it becomes
> difficult.”
>
> [Adapted from @sniehottaBridgingIntentionBehaviour2005]{.subtle}

0 = *not at all true today* <br>
5 = *completely true today*<br>

<!-- ca. 1 Minute -->

## Measurement - outcome

### Moderate to vigorous physical activity (MVPA)

> **Rough English translation of the original German item**<br>
> “How many minutes did you spend engaging in moderate-to-vigorous physical
> activity today?”
>
> [Adapted from @amireaultGodinShephardQuestionnaire2015]{.subtle}

- **Alone:** _____ minutes
- **Together with your partner:** _____ minutes

**Total MVPA = solo MVPA + joint MVPA.**

<!-- ca. 1 Minute -->

# Distinguishable dyads
Distinguishing by gender

:::{.notes}

Almost done with first step, we still have to decide on distinguishability. 
In our data, probably makes sense to distinguish by gender. Not always!

-> Gender, but TEST!

:::

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. [**Prepare the data and validate the dyadic structure**]{.text-highlight}
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Load and inspect the raw data

```{r}
library(dplyr) # for easier data-wrangling, exports functions like `select()`

raw_dyad_data <- readRDS("dyadic-person-means.rds") |>
  dplyr::select(
    couple_id, person_id, gender, efficacy, total_mvpa
  )

# usually use head(raw_dyad_data), slide_table is a custom function for html tables on these slides
slide_table(raw_dyad_data[1:4,])
```

:::{.subtle}

<br>

This dataset is stored as an R object. Other packages provide functions to import various formats, such as: `readxl::read_excel()`, `utils::read.csv()`, `haven::read_sav()`, `haven::read_sas()`

:::

:::{.notes}

Manual inspection at each step is important! 

Example that Lilly told me: weird associations, until discovered that 
missing values were coded as 99, in a software that didn't recognize this 
as a missing value.

:::

## Create the required format and validate the data

**For the distinguishable APIM we need to add:**

- `efficacy_actor` and `efficacy_partner` columns
- Dummy coded numeric variables for the two roles

:::{.fragment}
**Validate things like: **

- no dyad with more than two members,
- each dyad-member combination occurs only once (per occasion),
- gender labels are present and consistent
- you only have the type of dyad you expect
:::

## Create the required format and validate the data

We can use the `dyadMLM` package to help with data preparation and validation.

```{r}
#| output-location: slide

library(dyadMLM)

apim_distinguishable_data <- dyadMLM::prepare_dyad_data( # exported function from dyadMLM
  data = raw_dyad_data,
  dyad = couple_id,
  member = person_id,
  role = gender,
  predictors = efficacy, 
  model_types = 'apim',
  # if we want to fit an exchangeable model from the same data we can:
  include_arbitrary_member_contrast = TRUE,
  add_apim_gmc_predictors = TRUE, 
  seed = 123
) 

print(apim_distinguishable_data)

```


## Create the required format and validate the data [Manual code →](#manual-apim-preparation){.results-forward-link} {#prepared-apim-data}

Looking at the columns immediately relevant to us:

```{r}

head(apim_distinguishable_data) |>
  dplyr::select(
    couple_id, person_id, gender, .is_female, .is_male, .efficacy_gmc, .efficacy_gmc_actor, .efficacy_gmc_partner, total_mvpa
  ) |>
  slide_table()

```

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. [**Describe and visualize measures and observed relationships**]{.text-highlight}
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Create a wide view for role-specific exploration

The models will **NOT** use this dataset

```{r}
apim_distinguishable_wide_data <- apim_distinguishable_data |>
  dplyr::select(couple_id, gender, efficacy, total_mvpa) |>
  tidyr::pivot_wider(
    names_from = gender,
    values_from = c(efficacy, total_mvpa)
  )

slide_table(head(apim_distinguishable_wide_data))
```


## Describe focal variables per role

```{r}
library(report)

apim_distinguishable_wide_data |>
  dplyr::select(-couple_id) |>
  report::report_table() |>
  slide_table()
```

## Visualize focal variable by partner role

```{r}

library(ggplot2)
library(see)

plot_role_histograms <- function(data, group, var) {
  var_label <- rlang::as_label(rlang::enquo(var))

  ggplot(
    data,
    aes(x = {{ var }}, y = after_stat(density))
  ) +
    geom_histogram(
      bins = 15, boundary = 0,
      fill = "steelblue", color = "white", alpha = 0.75
    ) +
    geom_density(
      bounds = c(0, Inf), color = "navy", linewidth = 1.1
    ) +
    facet_wrap(vars({{ group }}), nrow = 1) +
    labs(x = var_label, y = "Density") +
    see::theme_modern(base_size = 16)
}

```

## Visualize focal variable by partner role - Efficacy

```{r}
#| dev: svglite

plot_role_histograms(apim_distinguishable_data, gender, efficacy)
```


## Visualize focal variable by partner role - MVPA

```{r}
#| dev: svglite

plot_role_histograms(apim_distinguishable_data, gender, total_mvpa)
```

## Observed dyadic correlations

The wide data ensure that each couple contributes once.

```{r}
library(correlation)
library(insight)

apim_distinguishable_wide_data |>
  correlation::correlation(
    select = c("efficacy_female", "total_mvpa_female"),
    select2 = c("efficacy_male", "total_mvpa_male")
  ) |> 
  summary() |>
  insight::print_md(footer = "")
```


## Visualize partner similarity

```{r}
plot_partner_similarity <- function(data, x, y, id, title, n_mahalanobis = 3) {
  x_name <- rlang::as_name(rlang::enquo(x))
  y_name <- rlang::as_name(rlang::enquo(y))
  xy <- cbind(data[[x_name]], data[[y_name]])
  labels <- data |>
    dplyr::mutate(
      .distance = mahalanobis(xy, colMeans(xy), cov(xy))
    ) |>
    dplyr::slice_max(.distance, n = n_mahalanobis)

  ggplot(data, aes(x = {{ x }}, y = {{ y }})) +
    geom_point(color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "navy") +
    geom_text(
      data = labels,
      aes(label = {{ id }}),
      vjust = -0.6, check_overlap = TRUE, size = 4
    ) +
    labs(title = title, x = "Female", y = "Male") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
    see::theme_modern(base_size = 16)
}
```

## Visualize partner similarity

```{r}
#| output-location: slide
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.5
#| out-width: 100%
#| fig-cap: "Labels: four largest Mahalanobis distances in each plot."

n_mahalanobis <- 4

see::plots(
  plot_partner_similarity(
    apim_distinguishable_wide_data,
    efficacy_female, efficacy_male, couple_id, "Efficacy",
    n_mahalanobis = n_mahalanobis
  ),
  plot_partner_similarity(
    apim_distinguishable_wide_data,
    total_mvpa_female, total_mvpa_male, couple_id, "Daily MVPA",
    n_mahalanobis = n_mahalanobis
  )
)
```

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. [**Estimate the models**]{.text-highlight}
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Distinguishable APIM

```{r}
#| label: applied-conceptual-distinguishable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with role-specific actor and partner paths and correlated outcome residuals."

draw_apim_diagram("distinguishable", predictors_centered = TRUE,
                  labels = c(predictor = "Efficacy", outcome = "MVPA"))
```

:::{.notes}

This is the same conceptual model introduced in Part I.

[Sources]

- @kenny2006dyadic

:::

## The APIM starts with two member-specific equations

For each dyad $i$, each member with role
$r \in \{\mathrm{M},\mathrm{F}\}$ has their own predictor
$X^{(\mathrm{A})}_{ri}$ and their partner's predictor
$X^{(\mathrm{P})}_{ri}$:

$$
  \begin{gathered}
  Y_{\mathrm{M},i}
  = b_{0,\mathrm{M}}
  + \color{#1769AA}{a_{\mathrm{M}} \times X^{(\mathrm{A})}_{\mathrm{M},i}}
  + \color{#7B3FA1}{p_{\mathrm{M}} \times X^{(\mathrm{P})}_{\mathrm{M},i}}
  + \epsilon_{\mathrm{M},i}
  \\
  Y_{\mathrm{F},i}
  = b_{0,\mathrm{F}}
  + \color{#1769AA}{a_{\mathrm{F}} \times X^{(\mathrm{A})}_{\mathrm{F},i}}
  + \color{#7B3FA1}{p_{\mathrm{F}} \times X^{(\mathrm{P})}_{\mathrm{F},i}}
  + \epsilon_{\mathrm{F},i}
  \end{gathered}
$$

[**- Actor effects**]{style="color: #1769AA"} [**- Partner effects**]{style="color: #7B3FA1"}

In long format, each equation applies to one member row of the dyad.

How can we combine these two equations into one model?


## Role indicators combine the equations

The indicators $D_{\mathrm{M},ri}$ and $D_{\mathrm{F},ri}$ identify the male
or female row:

$$
\begin{aligned}
Y_{ri} ={}&
\underbrace{
  b_{0,\mathrm{M}} \times D_{\mathrm{M},ri}
  + b_{0,\mathrm{F}} \times D_{\mathrm{F},ri}
}_{\text{role-specific intercepts}}
\\[0.35em]
&+
\color{#1769AA}{
\underbrace{
  a_{\mathrm{M}} \times D_{\mathrm{M},ri} \times X^{(\mathrm{A})}_{ri}
  + a_{\mathrm{F}} \times D_{\mathrm{F},ri} \times X^{(\mathrm{A})}_{ri}
}_{\text{role-specific actor effects}}
}
\\[0.35em]
&+
\color{#7B3FA1}{
\underbrace{
  p_{\mathrm{M}} \times D_{\mathrm{M},ri} \times X^{(\mathrm{P})}_{ri}
  + p_{\mathrm{F}} \times D_{\mathrm{F},ri} \times X^{(\mathrm{P})}_{ri}
}_{\text{role-specific partner effects}}
}
+ \epsilon_{ri}.
\end{aligned}
$$

:::{.text-smaller}

For a male row: $D_\mathrm{M}=1,\ D_\mathrm{F}=0$. For a female row: the reverse.

:::

:::{.notes}

The role indicators correspond to `.is_male` and `.is_female`. The actor and
partner values correspond to `.efficacy_gmc_actor` and `.efficacy_gmc_partner`.
The role index is written first, followed by the dyad index, as in
$Y_{\mathrm{male},i}$ and $Y_{\mathrm{female},i}$.

[Sources]

- @kenny2006dyadic

:::

## Residuals remain correlated within dyads

The two member rows from the same dyad retain a joint residual structure:

$$
\operatorname{Cov}
\begin{pmatrix}
\epsilon_{\mathrm{F}i} \\
\epsilon_{\mathrm{M}i}
\end{pmatrix}
= \boldsymbol{\Sigma}_{\epsilon}
=
\begin{bmatrix}
\sigma_{\epsilon_\mathrm{F}}^{2}
&
\rho_{\epsilon_\mathrm{F}\epsilon_\mathrm{M}}
\sigma_{\epsilon_\mathrm{F}}\sigma_{\epsilon_\mathrm{M}} \\
\rho_{\epsilon_\mathrm{F}\epsilon_\mathrm{M}}
\sigma_{\epsilon_\mathrm{F}}\sigma_{\epsilon_\mathrm{M}}
&
\sigma_{\epsilon_\mathrm{M}}^{2}
\end{bmatrix}
$$


## Fitting the distinguishable APIM in glmmTMB

```{r}
#| code-line-numbers: "1-23|4|5-6|8-10|12-14|16-18|19-20|23|1-23"
#| output-location: slide

library(glmmTMB)

apim_distinguishable_model <- glmmTMB(
  total_mvpa ~
    # Role-specific intercepts
    0 + .is_female + .is_male +

    # Role-specific actor effects
    .is_female:.efficacy_gmc_actor +
    .is_male:.efficacy_gmc_actor +

    # Role-specific partner effects
    .is_female:.efficacy_gmc_partner +
    .is_male:.efficacy_gmc_partner +

    # Within-dyad residual covariance
    (0 + .is_female + .is_male | couple_id),
  dispformula = ~ 0,
  family = gaussian(),
  data = apim_distinguishable_data
)

summary(apim_distinguishable_model)

  
```

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. [**Verify model adequacy, distinguishability, and robustness**]{.text-highlight}
6. **Visualize, interpret, and report results**

:::{.notes}

Before interpreting results, verify estimation, model adequacy,
distinguishability, and robustness.

:::

## Verify model adequacy, distinguishability, and robustness

1. Did estimation succeed?
2. Does the model reproduce the important data patterns?
3. Is distinguishability supported?
4. Are conclusions robust or driven by few dyads or consequential
   modeling choices? (sensitivity analyses)

## Did estimation succeed / converge?

The model converged without warnings... a good first sign!

:::{.fragment}
**Other things to look out for in the model output:**

- Estimated variances near $0$ or correlations near $\pm 1$ $\rightarrow$ possible boundary fit
- `NaN` or infinite estimates, standard errors, or model fit indices

:::

## Residual diagnostics / reproduction of data patterns

Residual diagnostics are not as straightforward anymore

- Linear models: straightforward checks
- Mixed models: residuals arise at multiple levels
- Generalized models: expectations depend on the family and link
- Dyadic models: partners' residuals are correlated

**Solution:** Compare the data with simulations from the fitted model.


## DHARMa: one diagnostic scale across models

The `DHARMa` package [@hartigDHARMaResidualDiagnostics2026]:

- Simulates outcomes from the fitted model
- Checks whether the observed residuals behave as expected under simulations from the fitted model.
- Adequate fit $\rightarrow$ approximately uniform, pattern-free residuals

One function works across many model families!

## Simulate the full dyadic residual structure

```{r}
#| label: simulate-dharma-distinguishable
#| cache: true
#| cache.extra: !expr digest::digest(list(stats::logLik(apim_distinguishable_model), stats::model.frame(apim_distinguishable_model)))

library(DHARMa)

apim_distinguishable_dharma_residuals <- DHARMa::simulateResiduals(
  fittedModel = apim_distinguishable_model,
  n = 10000,
  
  # Since our model uses dispformula = ~ 0, we cannot use DHARMa's default
  # conditional simulations. With unconditional, the random-effects block representing
  # our residuals is re-simulated.
  simulateREs = "unconditional",
  
  # Male and female residuals are correlated. DHARMa estimates this correlation
  # from the simulations and rotates the residuals to "remove" it.
  rotation = 'estimated',
  seed = 123
)
```

## Inspect DHARMa residuals {.smaller}

:::: {.columns}

::: {.column width="75%"}

```{r}
#| fig-format: svg
#| echo: false

plot(apim_distinguishable_dharma_residuals)

```

:::

::: {.column width="25%"}

**In this model:**

- Uniformity and dispersion are plausible
- No simulation outliers
- A fitted-value pattern remains

**Next:** inspect roles without rotation.

:::

::::

More targeted checks and guidance on interpretation of the plots:
[DHARMa vignette](https://stat.ethz.ch/CRAN/web/packages/DHARMa/vignettes/DHARMa.html).

## Inspect DHARMa residuals by role

```{r}

apim_distinguishable_female_dharma_residuals <- DHARMa::recalculateResiduals(
  apim_distinguishable_dharma_residuals,
  
  # Recalculate from the same simulations for one role.
  # One observation per couple: no rotation needed.
  sel = ~ .is_female == 1,
  rotation = NULL,
  seed = 123
)

apim_distinguishable_male_dharma_residuals <- DHARMa::recalculateResiduals(
  apim_distinguishable_dharma_residuals,
  sel = ~ .is_male == 1,
  rotation = NULL,
  seed = 123
)
```

## The fitted-value pattern appears among women {.smaller}


:::: {.columns}

::: {.column width="75%"}

```{r}
#| echo: false
#| fig-format: svg
#| fig-width: 10
#| fig-height: 4.6
#| out-width: 100%

old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 4.5, 1))
DHARMa::plotResiduals(
  apim_distinguishable_female_dharma_residuals,
  main = "Women"
)
DHARMa::plotResiduals(
  apim_distinguishable_male_dharma_residuals,
  main = "Men"
)
par(old_par)
```


:::

::: {.column width="25%"}

**In this model:**

- Pattern detected among women
- No comparable pattern among men

**Next:** try a defensible outcome scale.


:::

::::

More targeted checks and guidance on interpretation of the plots:
[DHARMa vignette](https://stat.ethz.ch/CRAN/web/packages/DHARMa/vignettes/DHARMa.html).

## Square-root MVPA reduces the fitted-value pattern {.smaller}

```{r}
#| label: simulate-dharma-distinguishable-sqrt
#| include: false
#| cache: true
#| cache.extra: !expr digest::digest(list(stats::logLik(apim_distinguishable_model), stats::model.frame(apim_distinguishable_model)))

apim_distinguishable_sqrt_data <- apim_distinguishable_data |>
  dplyr::mutate(sqrt_total_mvpa = sqrt(total_mvpa))

apim_distinguishable_sqrt_model <- update(
  apim_distinguishable_model,
  sqrt_total_mvpa ~ .,
  data = apim_distinguishable_sqrt_data
)

apim_distinguishable_sqrt_dharma_residuals <- DHARMa::simulateResiduals(
  fittedModel = apim_distinguishable_sqrt_model,
  n = 10000,
  simulateREs = "unconditional",
  rotation = 'estimated',
  seed = 123
)
```

:::: {.columns}

::: {.column width="70%"}

```{r}
#| echo: false
#| fig-format: svg

plot(apim_distinguishable_sqrt_dharma_residuals)

```
:::

::: {.column width="30%"}

**Compared with raw MVPA:**

- Uniformity and dispersion are plausible
- No simulation outliers
- No pooled fitted-value pattern

**Substantive result unchanged:** actor effects remain positive and supported;
partner effects remain unsupported.

:::

::::

:::{.subtle}

Treat this as a prespecified or transparently reported sensitivity analysis—not
a scale selected by diagnostic p-values.

:::

## Test distinguishability

- Fit model with exchangeability constraints

- Compare models

- Pool only if theoretically defensible, aligned with RQ and theory, and 
compatible with the data
  - Depending on goals you can also report the distinguishable model and 
  report results from this analysis

However, non-significant comparisons are not proof that roles are identical.

[@gistelinckIndistinguishabilityTests2018]{.subtle}

## Estimate and compare: 

```{r include=FALSE}

apim_exchangeable_comparison_model <- glmmTMB(
  total_mvpa ~
    # Pooled intercept
    1 +

    # Pooled actor effect
    .efficacy_gmc_actor +

    # Pooled partner effect
    .efficacy_gmc_partner +

    # Within-dyad residual covariance
    (1 | couple_id) +
    (0 + .member_contrast_arbitrary | couple_id),
  dispformula = ~ 0,
  family = gaussian(),
  data = apim_distinguishable_data
)

summary(apim_exchangeable_comparison_model)

dyadMLM::recover_exchangeable_covariance(apim_exchangeable_comparison_model)

```

```{r}

dyadMLM::compare_nested_models(
  apim_distinguishable_model, apim_exchangeable_comparison_model
)

```

## Test robustness - are conclusions driven by few dyads?

```{r}
plot_actor_association <- function(data, model, role, title,
                                   n_mahalanobis = 3) {
  xy <- cbind(data$.efficacy_gmc_actor, data$total_mvpa)
  labels <- data |>
    dplyr::mutate(.distance = mahalanobis(xy, colMeans(xy), cov(xy))) |>
    dplyr::slice_max(.distance, n = n_mahalanobis)
  coefficients <- glmmTMB::fixef(model)$cond

  ggplot(data, aes(.efficacy_gmc_actor, total_mvpa)) +
    geom_point(color = "steelblue") +
    geom_abline(intercept = coefficients[role],
                slope = coefficients[paste0(role, ":.efficacy_gmc_actor")],
                color = "navy", linewidth = 16 / 11) +
    geom_text(
      data = labels, aes(label = couple_id),
      vjust = -0.6, check_overlap = TRUE, size = 4
    ) +
    labs(
      title = title,
      x = "Actor efficacy (grand-mean centered)",
      y = "Daily MVPA (minutes)"
    ) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
    see::theme_modern(base_size = 16) +
    theme(plot.title.position = "panel",
          plot.title = element_text(hjust = 0.5))
}
```

## Test robustness - are conclusions driven by few dyads?

```{r}
#| output-location: slide
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.2
#| out-width: 100%
#| fig-cap: "Labels: four largest Mahalanobis distances in each plot."

n_mahalanobis <- 4

see::plots(
  plot_actor_association(
    dplyr::filter(apim_distinguishable_data, gender == "female"),
    apim_distinguishable_model, ".is_female", "Women",
    n_mahalanobis = n_mahalanobis
  ),
  plot_actor_association(
    dplyr::filter(apim_distinguishable_data, gender == "male"),
    apim_distinguishable_model, ".is_male", "Men",
    n_mahalanobis = n_mahalanobis
  )
) +
  patchwork::plot_layout(
    axes = "collect",
    axis_titles = "collect"
  )
```


## Omitting Couple 28 changes the associations


```{r}

apim_without_couple_28 <- update(
  apim_distinguishable_model,
  data = filter(apim_distinguishable_data, couple_id != 28)
)

actor_sensitivity <- parameters::compare_parameters(
  `Full data` = apim_distinguishable_model,
  `Without Couple 28` = apim_without_couple_28,
  keep = "\\.efficacy_gmc_actor"
)

```



```{r}
#| echo: false

actor_sensitivity_table <- format(
  actor_sensitivity,
  select = "{estimate}{stars}",
  format = "text"
) |>
  filter(
    !is.na(Parameter),
    Parameter != "Observations"
  ) |>
  mutate(
    Parameter = c("Women", "Men")
  ) |>
  rename(
    `Actor slope` = Parameter
  )

slide_table(
  actor_sensitivity_table,
  align = c("l", "r", "r")
)
```

:::{.subtle}
\* $p < .05$; \*\* $p < .01$. Couple 28 was flagged in both visual screens.
Its omission weakens both slopes; only the men's result crosses $p=.05$.
:::

:::{.notes}

Couple 28 had the largest Mahalanobis distance in both role-specific visual
screens. Mahalanobis distance identifies an unusual value combination, not
formal coefficient influence. This targeted refit asks the narrower question:
do the actor-effect conclusions materially change without that dyad?

Both actor slopes remain positive. The women's slope decreases from 11.87 to
8.44 and remains significant. The men's slope decreases from 7.97 to 5.23,
with its two-sided Wald $p$-value moving from .027 to .078.

The full-data analysis remains primary. Do not remove Couple 28 merely because
its omission changes a significance label. If several cases were
substantively or procedurally flagged before modeling, repeat the same
sensitivity check for those cases.

:::

## Other sensitivity analyses you could do

Fit a small set of (prespecified) models to test: 

- coding decisions
- different plausible outcome families or transformations
- missing-data handling
- covariates with a substantive confounding or precision rationale

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. [**Visualize, interpret, and report results**]{.text-highlight}

## Reporting results and answering the research question

```{r}
library(parameters)

parameters::model_parameters(apim_distinguishable_model, effects = 'fixed') |>
  insight::print_md(footer = "")

```

::: {.fragment}
In this sample, higher self-efficacy was associated with more of one’s own MVPA
for both women and men, but not clearly with the partner’s MVPA.
:::

## Reporting results and answering the research question

Random effects in our model represent the residual structure

```{r}

parameters::model_parameters(apim_distinguishable_model, effects = 'random') |>
  insight::print_md(footer = "")

```

## Visualize results

```{r}
#| fig-format: svg
#| out-width: 80%


parameters::model_parameters(apim_distinguishable_model, effects = "fixed") |>
  plot()

```

## What should be reported?

- **Analysis sample:** dyads and observations, exclusions, aggregation, and missing-data handling
- **Model specification:** outcome family and link, fixed effects, dyadic covariance structure, and equality constraints
- **Estimation and model decision:** estimator, software and version, convergence, boundary estimates, distinguishability comparison, and retained model
- **Results:** focal estimates with 95% confidence intervals in substantive units and the fitted variance-covariance parameters
- **Checks and interpretation:** residual diagnostics, influential dyads, prespecified sensitivity analyses, material changes, and causal limits



# Exchangeable dyads
The same model but constrained

## Workflow

1. [**Define dyads, distinguishability, and research questions**]{.text-highlight}
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Define dyads, distinguishability, and research questions

> Same data and RQ, now with exchangeability constraints for illustration.

:::{.notes}

Since we now know that a distinguishable model is not better than an exchangeable one,
we can use the same RQ and data to demonstrate the exchangeable workflow.

:::

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. [**Prepare the data and validate the dyadic structure**]{.text-highlight}
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Create the required format and validate the data

You can use `dyadMLM::prepare_dyad_data()` to prepare and validate the data. 

For exchangeable dyads, we can just omit the `role` argument. 
If we want to estimate both distinguishable and exchangeable versions, 
we would pass `role` and set `include_arbitrary_member_contrast = TRUE`.

```{r}
#| output-results: slide

apim_exchangeable_data <- dyadMLM::prepare_dyad_data(
  data = raw_dyad_data,
  dyad = couple_id,
  member = person_id,
  predictors = efficacy,
  model_types = 'apim',
  add_apim_gmc_predictors = TRUE,
  seed = 123
)

print(apim_exchangeable_data)

```

## Create the required format and validate the data

Looking at the relevant columns: 

```{r}
head(apim_exchangeable_data) |>
  dplyr::select(
    couple_id, person_id, .member_contrast_arbitrary, .efficacy_gmc, .efficacy_gmc_actor, .efficacy_gmc_partner, total_mvpa
  ) |>
  slide_table()
```


## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. [**Describe and visualize measures and observed relationships**]{.text-highlight}
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Describe focal variables (pooled)

No wide-view of the data is needed for exchangeable dyads. 
We can describe the data directly to obtain pooled estimates: 

```{r}
apim_exchangeable_data |>
  dplyr::select(efficacy, total_mvpa) |>
  report::report_table() |>
  slide_table()
```

## Visualize focal variables

```{r}

plot_pooled_histogram <- function(data, var) {
  var_label <- rlang::as_label(rlang::enquo(var))

  ggplot(
    data,
    aes(x = {{ var }}, y = after_stat(density))
  ) +
    geom_histogram(
      bins = 15, boundary = 0,
      fill = "steelblue", color = "white", alpha = 0.75
    ) +
    geom_density(
      bounds = c(0, Inf), color = "navy", linewidth = 1.1
    ) +
    see::theme_modern(base_size = 16)
}

```

## Visualize focal variables

```{r}
#| fig-format: svg
#| out-width: 100%

see::plots(
  plot_pooled_histogram(apim_exchangeable_data, efficacy),
  plot_pooled_histogram(apim_exchangeable_data, total_mvpa)
)
```

## Observed correlations

For exchangeable dyads, ICCs summarize partner similarity:

```{r}
library(wbCorr)

apim_exchangeable_correlations <- wbCorr(
  apim_exchangeable_data[,c('efficacy','total_mvpa')],
  cluster = apim_exchangeable_data$couple_id
)

summary(apim_exchangeable_correlations, 'wb')
```

**Diagonal ICCs:** expected correlation between partners on each variable

[**Off-diagonals:** top right = within-couple correlation; bottom left = between-couple correlation]{.subtle}

## Visualize partner similarity

```{r}
#| output-location: slide
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.5
#| out-width: 100%
#| fig-cap: "Each couple is shown twice—once with each person as the focal member."

partner_similarity_data <- apim_exchangeable_data |>
  dplyr::group_by(couple_id) |>
  dplyr::mutate(
    efficacy_partner = rev(efficacy),
    total_mvpa_partner = rev(total_mvpa)
  ) |>
  dplyr::ungroup()

see::plots(
  plot_partner_similarity(
    partner_similarity_data,
    efficacy, efficacy_partner, couple_id, "Efficacy",
    n_mahalanobis = 0
  ) +
    labs(x = "Focal member", y = "Partner"),
  plot_partner_similarity(
    partner_similarity_data,
    total_mvpa, total_mvpa_partner, couple_id, "Daily MVPA",
    n_mahalanobis = 0
  ) +
    labs(x = "Focal member", y = "Partner")
)
```

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. [**Estimate the models**]{.text-highlight}
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Exchangeable APIM

```{r}
#| label: applied-conceptual-exchangeable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual exchangeable APIM with pooled actor and partner paths and correlated outcome residuals."

draw_apim_diagram("exchangeable", predictors_centered = TRUE,
                  labels = c(predictor = "Efficacy", outcome = "MVPA"))
```

## In MLM we constrain / pool by omitting terms

Both partners' formulas are identical, so no need to combine them. We can just write: 

For each dyad $i$, each member $j$ has their own predictor 
$X^{(\mathrm{A})}_{j,i}$ and their partner's predictor $X^{(\mathrm{P})}_{j,i}$:

$$
  Y_{j,i}
  = b_{0}
  + \color{#1769AA}{a \times X^{(\mathrm{A})}_{j,i}}
  + \color{#7B3FA1}{p \times X^{(\mathrm{P})}_{j,i}}
  + \epsilon_{j,i}
$$

[**- Actor effects**]{style="color: #1769AA"} [**- Partner effects**]{style="color: #7B3FA1"}

In long format, each equation applies to one member row of the dyad.

The model pools the effects, but actor and partner values are exchanged between
member rows, so their predictions may still differ.

## The tricky part: residuals need to remain correlated within dyads

- The two member rows from the same dyad retain a joint residual structure

- Both members now have the same residual variance (pooled)

- Position 1 and 2 within the dyad are arbitrary


$$
\operatorname{Cov}
\begin{pmatrix}
\epsilon_{1i} \\
\epsilon_{2i}
\end{pmatrix}
= \boldsymbol{\Sigma}_{\epsilon}
=
\begin{bmatrix}
\sigma_{\epsilon}^{2}
&
\rho_{\epsilon_1\epsilon_2}\sigma_{\epsilon}^{2} \\
\rho_{\epsilon_1\epsilon_2}\sigma_{\epsilon}^{2}
&
\sigma_{\epsilon}^{2}
\end{bmatrix}
$$

## Fitting the exchangeable APIM in glmmTMB

```{r results='hide'}
#| code-line-numbers: "1-25|2|3-4|6-7|9-10|12-20|14-15|17-18|20|1-23"
apim_exchangeable_model <- glmmTMB(
  total_mvpa ~
    # Pooled intercept -------------------------------------
    1 + 

    # Pooled actor effect ----------------------------------
    .efficacy_gmc_actor +

    # Pooled partner effect --------------------------------
    .efficacy_gmc_partner +

    # Within-dyad residual covariance (Sum/Deviation) ------
  
    # Residuals of partners might be similar (positively correlated)
    (1 | couple_id) +
    
    # Residuals of partners might move opposite (negatively correlated)
    (0 + .member_contrast_arbitrary | couple_id),
  
  dispformula = ~ 0,
  family = gaussian(),
  data = apim_exchangeable_data
)

summary(apim_exchangeable_model)
```

## The mean-deviation parameterization

- The DIM represents member residuals using two uncorrelated components:

  - Mean block `(1 | couple_id)`:
    $r_{\mathrm{M}i}=(\epsilon_{1i}+\epsilon_{2i})/2$; both members move together

  - Deviation block `(0 + .member_contrast_arbitrary | couple_id)`:
    $r_{\mathrm{D}i}=(\epsilon_{1i}-\epsilon_{2i})/2$; members move in opposite directions

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. [**Verify model adequacy, distinguishability, and robustness**]{.text-highlight}
6. **Visualize, interpret, and report results**

## Verify model adequacy, distinguishability, and robustness

Essentially the same steps and code can be used to check:

1. Did estimation succeed?
2. Does the model reproduce the important data patterns?
3. Does a distinguishable model fit clearly better?
4. Are conclusions robust or driven by few dyads or consequential
   modeling choices? (sensitivity analyses)

:::{.notes}

Regarding distinguishability: Sensitivity analyses where different ways of
distinguishing are tested are possible (if there are feasible ways)

:::


## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. [**Visualize, interpret, and report results**]{.text-highlight}


## Reporting results and answering the research questions

```{r}
parameters::model_parameters(apim_exchangeable_model, effects = 'fixed') |>
  insight::print_md(footer = "")
```

::: {.fragment}
In the parsimonious exchangeable model, higher self-efficacy was associated
with more of one’s own MVPA, but not clearly with the partner’s MVPA.
:::


## Reporting results and answering the research questions

```{r}
parameters::model_parameters(apim_exchangeable_model, effects = 'random') |>
  insight::print_md(footer = "")
```

:::{.fragment}
Often it is preferable to rotate the covariance back to the familiar member-level covariance matrix: 

- Member residual variance:
  $\operatorname{Var}(\epsilon_{1i})=\operatorname{Var}(\epsilon_{2i})
  =\operatorname{Var}(r_{\mathrm{M}i})+\operatorname{Var}(r_{\mathrm{D}i})$

- Partner residual covariance:
  $\operatorname{Cov}(\epsilon_{1i},\epsilon_{2i})
  =\operatorname{Var}(r_{\mathrm{M}i})-\operatorname{Var}(r_{\mathrm{D}i})$
  
:::

## Reporting results and answering the research questions

No need to do so by hand. Rotating back with `dyadMLM`:

```{r}

dyadMLM::recover_exchangeable_covariance(apim_exchangeable_model)

```


## Visualize results


```{r}
#| fig-format: svg
#| out-width: 80%

parameters::model_parameters(apim_exchangeable_model, effects = 'fixed') |>
  plot()

```

# Optional: DIM and DSM

::: {.workshop-next}
Worked examples are available in the package vignettes.

[Open the DIM vignette →](https://pascal-kueng.github.io/dyadMLM/articles/dim.html){.workshop-next-link}

[Open the DSM vignette →](https://pascal-kueng.github.io/dyadMLM/articles/dsm.html){.workshop-next-link}
:::

# Intensive longitudinal dyads

## ILD extends rather than replaces the workflow {.smaller}

:::: {.columns}

::: {.column width="43%"}

### The same core workflow

::: {.subtle}

1. Define dyads, roles, distinguishability, and research questions
2. Check data structure, completeness, and missing-data handling
3. Describe measures and dyadic relationships
4. Specify, document, and estimate the APIM
5. Verify estimation; check residuals and influential dyads
6. Interpret, visualize, and report the results

:::

:::

::: {.column width="57%"}

### ILD additionally requires

- Check compliance, matched occasions, and actual time gaps
  [@revolPreprocessingExperienceSampling2024]
- Plot paired trajectories and inspect raw temporal and cross-partner patterns
  [@bolgerIntensiveLongitudinalMethods2013;
  @revolPreprocessingExperienceSampling2024]
- Separate within- and between-person variation
  - ICCs, level-specific correlations, and multilevel reliability where
    applicable [@bolgerIntensiveLongitudinalMethods2013;
    @hamakerCuriousCaseCrossSectional2024;
    @geldhofReliabilityEstimationMultilevel2014]
- Choose the temporal specification
  - Time scale and lag, trends/stationarity, concurrent versus lagged effects,
    and discrete versus continuous time
    [@driverContinuousTimeStructural2017]
- Check residual temporal and cross-partner dependence

:::

::::

::: {.subtle}
Workflow synthesized from @bolgerIntensiveLongitudinalMethods2013,
@revolPreprocessingExperienceSampling2024, @mcneishPrimerTwoLevel2020, and
@delrosarioPracticalGuideSpecifying2025.
:::

:::{.notes}

For ILD-specific power analysis, see @lafitPowerLAPIMApplication2022. Perry et
al. belongs with interpretation of fitted dynamics: their change-score APIM and
vector-field diagrams show the joint implications of actor and partner
coefficients [@perryGraphicMethodsInterpreting2017].

We will use `wbCorr` for pooled within- and between-person correlations
[@kuengWbCorrBivariateWithin2026]; @hamakerCuriousCaseCrossSectional2024
provides the methodological rationale.

`omegaSEM()` is provided by `multilevelTools`, not `semTools`. It implements
level-specific internal-consistency coefficients based on a two-level
one-factor model [@wileyMultilevelTools2025;
@geldhofReliabilityEstimationMultilevel2014]. These coefficients are not
temporal stability or the precision of observed person means; for the latter,
see @laiCompositeReliabilityMultilevel2021.

:::

## Distinguishable APIM

- Formula
- Random effects and residual covariance explained
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships on both levels (wbCorr)
  - Estimating the model
  - Checking assumptions
  - Visualizing results
  
## Exchangeable APIM

- Formula
- Random effects and residual covariance explained
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships on both levels (wbCorr)
  - Estimating the model
  - Checking assumptions
  - Visualizing results

## AR1, OU, and Dynamic Models

- When to use dynamic models vs. concurrent models
- Current possibilities in R MLM for AR1 or ou models
- Current possibilites in R MLM for dynamic models
- Alternatives (DSEM in MPLUS or lavaan for X)

## Power analysis is the clearest reporting gap

```{r}
#| label: plot-longitudinal-apim-reporting-gaps
#| fig-format: svg
#| echo: false
#| fig-width: 10.5
#| fig-height: 4.5
#| out-width: 92%
#| fig-align: center
#| fig-alt: "Bar chart of reporting practices in 73 longitudinal APIM studies: 17.8 percent reported a power analysis, 64.4 percent documented model constraints, and 75.3 percent reported a missing-data strategy."

reporting_gaps <- tibble(
  item = factor(
    c("Power analysis", "Model constraints", "Missing-data strategy"),
    levels = c("Missing-data strategy", "Model constraints", "Power analysis")
  ),
  reported = c(17.8, 64.4, 75.3)
)

ggplot(reporting_gaps, aes(item, reported, fill = item == "Power analysis")) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = paste0(reported, "%")),
    hjust = -0.15, size = 7
  ) +
  coord_flip(clip = "off") +
  scale_fill_manual(values = c("TRUE" = "#0028a5", "FALSE" = "#9aa7bd")) +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 20),
    labels = paste0(seq(0, 100, 20), "%"),
    expand = c(0, 0)
  ) +
  labs(
    x = NULL,
    y = "Studies reporting the item"
  ) +
  theme_minimal(base_size = 21) +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(5.5, 20, 5.5, 5.5)
  )
```

Systematic review of 73 longitudinal APIM studies published from 2008–2025
[@mohammadzadehYazdBeyondIndividual2026].

:::{.notes}

- The chart shows reporting rates because they are easier to read as adherence
  to good practice. The paper directly reports 17.8% for power analyses; 64.4%
  and 75.3% are the complements of the reported omission rates.
- This was a methodological systematic review, not a meta-analysis of effect
  sizes. Its sample was restricted to longitudinal APIM studies with at least
  70 dyads and follow-up longer than one year.
- The review therefore supports “reporting gaps remain common,” not the broader
  claim that all dyadic analyses are reported inadequately.

[Sources]

- Mohammadzadeh Yazd et al. (2026), abstract:
  <https://doi.org/10.1177/02654075261443593>

:::

## Make every dyadic analysis decision visible

1. **Preregister** actor and partner hypotheses and key analysis decisions.
2. **Plan and report power** using dyad-specific tools.
3. **Explain missing data:** describe missingness and the handling strategy.
4. **Specify the full model:** roles and distinguishability, actor and partner
   paths, constraints, and covariance structure.

The review also recommends multi-method measurement and more diverse
populations

[@mohammadzadehYazdBeyondIndividual2026]{.subtle}


# Excercises in R

Short break - initiate package install before you leave

```{r eval = FALSE}

install.packages(
  c(
    "dyadMLM", "wbCorr", "glmmTMB", "DHARMa", "see",
    "dplyr", "ggplot2", "tibble"
  ),
  repos = c(
    "https://pascal-kueng.r-universe.dev",
    "https://cloud.r-project.org"
  )
)

```


- Go to https://pascal-kueng.github.io/dyadMLM/workshop/ 
- or Teams channel

<!--

- DHARMa plots!?! -> as required package for workshop -> link to materials or do in tutorial? Also check AR1 etc., etc. can dharma do this if resid is set to almost zero?

- glmmTMB -> as required 
- dyadMLM -> as required (via R-universe, see instructions).
- dplyr -> required
- tibble -> required
- lavaan -> required
- ggplot2 -> required

-------------->

<!-- ILD-part: 

- see plots on: https://quantdev.ssri.psu.edu/tutorials/apa-ati-intensive-longitudinal-data-session-m-dyadic-multilevel-model

-->

<!-- Excercises part: separate files! Include on overview page?

Create exercises with data visalization, etc. Panel plots, histograms
correlations (wb corr)?

-->

## References

::: {#refs}
:::

# Appendix

## Create the APIM variables manually [← Data preparation](#prepared-apim-data){.appendix-back-link} {#manual-apim-preparation .smaller}

This shortcut assumes validation has confirmed exactly two rows per dyad.

```{r}
#| eval: false

apim_manual_data <- raw_dyad_data |>
  dplyr::mutate(
    .is_female = as.numeric(gender == "female"),
    .is_male = as.numeric(gender == "male"),
    .efficacy_gmc = efficacy - mean(efficacy, na.rm = TRUE)
  ) |>
  dplyr::group_by(couple_id) |>
  dplyr::mutate(
    .efficacy_gmc_actor = .efficacy_gmc,
    .efficacy_gmc_partner = rev(.efficacy_gmc)
  ) |>
  dplyr::ungroup()
```
