---
title: "Dyadic Data Analysis with R"
subtitle: "Part II: Applied Tutorial"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
csl: apa.csl
resources:
  - uzh-logo.svg
  - charite-logo.svg
format:
  revealjs:
    embed-resources: true
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: false
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup-pilot-data
#| include: false

library(dplyr)
library(ggplot2)

slide_table <- function(x, digits = 2, ...) {
  knitr::kable(
    x,
    format = "html",
    digits = digits,
    format.args = list(
      nsmall = digits,
      scientific = FALSE
    ),
    row.names = FALSE,
    ...
  )
}

diagram_helper_candidates <- c(
  "diagram-helpers.Rinc",
  file.path("vignettes", "diagram-helpers.Rinc"),
  file.path("..", "..", "vignettes", "diagram-helpers.Rinc")
)
diagram_helper <- diagram_helper_candidates[
  file.exists(diagram_helper_candidates)
][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(
  normalizePath(diagram_helper),
  envir = knitr::knit_global()
)

```

## Today's aims and program

::: {.fancy-table}

| Time | Focus |
|---|---|
| 11:00–12:45 | Introduction and types of dyadic models |
| 12:45–14:00 | *Lunch break*|
| 14:00–15:30 | Estimating cross-sectional dyadic models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

:::

<!-- TODO @codex: please grey out the first two points, intro and lunch break). -->

<!-- ca. 2 Minutes -->

# Software and frameworks

## SEM vs. MLM {.smaller}

- Many simple dyadic models can be represented in either framework.
- The best framework is the one that represents your conceptual model most naturally [@ledermannAnalyzingDyadicData2017].

:::: {.columns .fragment}

::: {.column width="50%"}

**SEM / DSEM is often more natural for**

- Latent dyadic constructs (e.g., a CFM)
- Item-level measurement models for reflective measures
- Linked outcomes or equations [e.g., mediation; @ledermannMediationDyadicData2009]
- Short dynamic panels in wide-format [e.g., LD-APIM; @gistelinckMultilevelAutoregressiveSmall2021]
- Intensive lagged or coupled dynamics via DSEM/RDSEM [e.g., @asparouhovDynamicStructuralEquation2018]

:::

::: {.column width="50%"}

**MLM is often more natural for**

- Parsimonious observed-variable models with modest numbers of dyads
- Nested grouping and flexible random-effect structures
- Observed-outcome growth or concurrent ILD models with flexible time trends
- Complex observed-variable regressions—interactions, polynomials, splines,
  and dyadic RSA—especially with repeated measures or additional nesting
- Specialized generalized mixed models in R, including zero-inflated, hurdle,
  and Tweedie models

:::

::::

::: {.subtle}
[@asparouhovDynamicStructuralEquation2018;
@asparouhovComparisonModelsAnalysis2020;
@driverContinuousTimeStructural2017;
@gistelinckMultilevelAutoregressiveSmall2021;
@delrosarioPracticalGuideSpecifying2025;
@schonbrodtTestingSimilarityEffects2018;
@nestlerResponseSurfaceAnalysis2019;
@brooksGlmmTMBBalancesSpeed2017]
:::

## SEM vs. MLM: Missing data {.smaller}

:::{.fragment}
- **MLM: Complete-case**
  - **Outcome missing:** Valid under MAR, ML uses remaining observations
  - **Predictor missing:** Valid under MAR, if missingness does **not depend** on the outcome after accounting for predictors
  
:::

:::{.fragment}
- **SEM with FIML and joint likelihood**
  - Valid under MAR: Models predictors and outcomes jointly

:::

:::{.fragment}
- **MI: Imputation**
  - Handles missing outcomes, predictors, and auxiliary information
  - Must match the analysis and preserve person, dyad, role, and time structure

:::

::: {.fragment}
Include important predictors of missingness and missing values. FIML and MI
cover more MAR situations than row omission; none automatically solves MNAR.

:::

[@ledermannAnalyzingDyadicData2017; @hughesAccountingMissingData2019; @grundMultipleImputationMissing2018]{.subtle}

<!-- ca. 2.5 Minutes -->

## Why R—and where other software helps {.smaller}

:::: {.columns}

::: {.column width="46%"}

### Why we teach R

:::{.fragment}
- **Free to use:** no license fees or institutional access barriers
- **Open source:** statistical implementations can be inspected, audited, and adapted
- **Reproducible:** workflows can be rerun, reviewed, shared, and version-controlled
- **Extensible:** specialized methods integrate with data, graphics, and reporting
:::

:::

::: {.column width="54%"}

### Current trade-offs

:::{.fragment}

- **SEM/DSEM:** R's capabilities are spread across specialized packages
  - `lavaan` handles explicitly specified, fixed-wave lagged SEMs (e.g.,
    CLPM/RI-CLPM), but has no native DSEM or residual AR(1) structure
  - Version 0.7 adds restricted two-level random slopes; `mlts` implements
    Bayesian multilevel time-series models, while Mplus integrates DSEM/RDSEM
- **MLM/ILD:** `glmmTMB` and `brms` both model serial dependence
  - Neither directly offers a separable partner-by-time residual structure such
    as $\mathrm{UN}\otimes\mathrm{AR}(1)$; SAS `PROC MIXED` and SPSS `MIXED` do
    for Gaussian repeated-measures models

:::

:::

::::

::: {.subtle}
[@lavaanMultilevelSEM;
@mulderThreeExtensions2021;
@blavaanTwoLevelSEM;
@koslowskiMlts2025;
@asparouhovDynamicStructuralEquation2018;
@asparouhovComparisonModelsAnalysis2020;
@kristensenCovarianceStructuresGlmmTMB2023;
@buerknerAutocorrelationStructuresBrms;
@sasInstituteMixedProcedure;
@ibmCovarianceStructureList]
:::

<!-- ca. 2.5 Minutes -->

# Estimating cross-sectional dyadic models in R

## One possible workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and simple relationships**
4. **Estimate the models**
5. **Verify model adequacy, (distinguishability, if applicable), and robustness**
6. **Visualize, interpret, and report results**

<!-- konsistent wie hier -->
::: {.subtle}
Synthesized from @kenny2006dyadic, @ledermannToolboxProgramsRestructure2015, and @stasGivingDyadicData2018.
:::

:::{.notes}

- In this example we will use the same dataset that we already used in Part I

:::

## One possible workflow

1. [**Define dyads, distinguishability, and research questions**]{.text-highlight}
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and simple relationships**
4. **Estimate the models**
5. **Verify model adequacy, (distinguishability, if applicable), and robustness**
6. **Visualize, interpret, and report results**

<!-- konsistent wie hier -->
::: {.subtle}
Synthesized from @kenny2006dyadic, @ledermannToolboxProgramsRestructure2015, and @stasGivingDyadicData2018.
:::


## The Dataset

**Same as before: **

- 36 physically inactive romantic couples
- Both partners intended to become more physically active
- Smartphone-based planning interventions and JITAIs
- Daily diaries over 55 days

**Aggregated for cross-sectional analyses**


## Research Question {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ:]{.rq-label} [In romantic couples, is a person's **self-efficacy** linked to physical activity in both partners?]{.rq-text}
:::
:::

## Measurement - predictor

### Self-efficacy

> **Rough English translation of the original German item**<br>
> “I am confident that I can be physically active tomorrow, even if it becomes
> difficult.”
>
> [Adapted from @sniehottaBridgingIntentionBehaviour2005]{.subtle}

0 = *not at all true today* <br>
5 = *completely true today*<br>

<!-- ca. 1 Minute -->

## Measurement - outcome

### Moderate to vigorous physical activity (MVPA)

> **Rough English translation of the original German item**<br>
> “How many minutes did you spend engaging in moderate-to-vigorous physical
> activity today?”
>
> [Adapted from @amireaultGodinShephardQuestionnaire2015]{.subtle}

- **Alone:** _____ minutes
- **Together with your partner:** _____ minutes

**Total MVPA = solo MVPA + joint MVPA.**

<!-- ca. 1 Minute -->

# Distinguishable dyads
Distinguishing by gender

:::{.notes}

Almost done with first step, we still have to decide on distinguishability. 
In our data, probably makes sense to distinguish by gender. Not always!

-> Gender, but TEST!

:::

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. [**Prepare the data and validate the dyadic structure**]{.text-highlight}
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Load and inspect the raw data

```{r}
library(dplyr) # for easier data-wrangling, exports functions like `select()`

df <- readRDS("dyadic-person-means.rds") |>
  dplyr::select(
    couple_id, person_id, gender, efficacy, total_mvpa
  )

# usually use head(df), slide_table is a custom function for html tables on these slides
slide_table(df[1:4,]) 
```

:::{.subtle}

<br>

This dataset is stored as an R object. Other packages provide functions to import various formats, such as: `readxl::read_excel()`, `utils::read.csv()`, `haven::read_sav()`, `haven::read_sas()`

:::

:::{.notes}

Manual inspection at each step is important! 

Example that Lilly told me: weird associations, until discovered that 
missing values were coded as 99, in a software that didn't recognize this 
as a missing value.

:::

## Create the required format and validate the data

**For the distinguishable APIM we need to add:**

- `efficacy_actor` and `efficacy_partner` columns
- Dummy coded numeric variables for the two roles

:::{.fragment}
**Validate things like: **

- no dyad with more than two members,
- each dyad-member combination occurs only once (per occasion),
- gender labels are present and consistent
- you only have the type of dyad you expect
:::

## Create the required format and validate the data

We can use the `dyadMLM` package to help with data preparation and validation.

```{r}
#| output-location: slide

library(dyadMLM)

df_apim <- dyadMLM::prepare_dyad_data( # exported function from dyadMLM
  data = df,
  dyad = couple_id,
  member = person_id,
  role = gender,
  predictors = efficacy, 
  model_types = 'apim',
  # if we want to fit an exchangeable model from the same data we can:
  include_arbitrary_member_contrast = TRUE,
  add_apim_gmc_predictors = TRUE, 
  seed = 123
) 

print(df_apim)

```


## Create the required format and validate the data
<!-- TODO: link to an appendix, just like we did in the dyad-day.qmd file, with code-snippets of how to manually achieve the
is_male, is_female, and _actor and _partner variables. -->


Looking at the columns immediately relevant to us:

```{r}

df_apim[1:6,] |>
  dplyr::select(
    couple_id, person_id, gender, .is_female, .is_male, .efficacy_gmc, .efficacy_gmc_actor, .efficacy_gmc_partner, total_mvpa
  ) |>
  slide_table()

```

<!-- TODO: color code partner values? -->

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. [**Describe and visualize measures and observed relationships**]{.text-highlight}
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Create a wide view for role-specific exploration

The models will **NOT** use this dataset

```{r}
df_wide <- df_apim |>
  dplyr::select(couple_id, gender, efficacy, total_mvpa) |>
  tidyr::pivot_wider(
    names_from = gender,
    values_from = c(efficacy, total_mvpa)
  )

slide_table(head(df_wide))
```


## Describe focal variables per role

```{r}
library(report)

df_wide |>
  dplyr::select(-couple_id) |>
  report::report_table() |>
  slide_table()
```

## Visualize focal variable by partner role

```{r}

library(ggplot2)
library(see)

plot_dyadic_histograms <- function(data, group, var) {
  var_label <- rlang::as_label(rlang::enquo(var))

  ggplot(
    data,
    aes(x = {{ var }}, y = after_stat(density))
  ) +
    geom_histogram(
      bins = 15, boundary = 0,
      fill = "steelblue", color = "white", alpha = 0.75
    ) +
    geom_density(
      bounds = c(0, Inf), color = "navy", linewidth = 1.1
    ) +
    facet_wrap(vars({{ group }}), nrow = 1) +
    labs(x = var_label, y = "Density") +
    see::theme_modern(base_size = 16)
}

```

## Visualize focal variable by partner role - Efficacy

```{r}
#| dev: svglite

plot_dyadic_histograms(df_apim, gender, efficacy)
```


## Visualize focal variable by partner role - MVPA

```{r}
#| dev: svglite

plot_dyadic_histograms(df_apim, gender, total_mvpa)
```

## Observed dyadic correlations

The wide data ensure that each couple contributes once.

```{r}
library(correlation)
library(insight)

df_wide |>
  correlation::correlation(
    select = c("efficacy_female", "total_mvpa_female"),
    select2 = c("efficacy_male", "total_mvpa_male")
  ) |> 
  summary() |>
  insight::print_md(footer = "")
```


## Visualize partner similarity

```{r}
plot_partner_similarity <- function(data, x, y, id, title, n_mahalanobis = 3) {
  x_name <- rlang::as_name(rlang::enquo(x))
  y_name <- rlang::as_name(rlang::enquo(y))
  xy <- cbind(data[[x_name]], data[[y_name]])
  labels <- data |>
    dplyr::mutate(
      .distance = mahalanobis(xy, colMeans(xy), cov(xy))
    ) |>
    dplyr::slice_max(.distance, n = n_mahalanobis)

  ggplot(data, aes(x = {{ x }}, y = {{ y }})) +
    geom_point(color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "navy") +
    geom_text(
      data = labels,
      aes(label = {{ id }}),
      vjust = -0.6, check_overlap = TRUE, size = 4
    ) +
    labs(title = title, x = "Female", y = "Male") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
    see::theme_modern(base_size = 16)
}
```

## Visualize partner similarity

```{r}
#| output-location: slide
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.5
#| out-width: 100%
#| fig-cap: "Labels: four largest Mahalanobis distances in each plot."

n_mahalanobis <- 4

see::plots(
  plot_partner_similarity(
    df_wide, efficacy_female, efficacy_male, couple_id, "Efficacy",
    n_mahalanobis = n_mahalanobis
  ),
  plot_partner_similarity(
    df_wide, total_mvpa_female, total_mvpa_male, couple_id, "Daily MVPA",
    n_mahalanobis = n_mahalanobis
  )
)
```

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. [**Estimate the models**]{.text-highlight}
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Distinguishable APIM

```{r}
#| label: applied-conceptual-distinguishable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with role-specific actor and partner paths and correlated outcome residuals."

draw_apim_diagram("distinguishable", predictors_centered = TRUE,
                  labels = c(predictor = "Efficacy", outcome = "MVPA"))
```

:::{.notes}

This is the same conceptual model introduced in Part I.

[Sources]

- @kenny2006dyadic

:::

## The APIM starts with two member-specific equations

For each dyad $i$, each member with role
$r \in \{\mathrm{M},\mathrm{F}\}$ has their own predictor
$X^{(\mathrm{A})}_{ri}$ and their partner's predictor
$X^{(\mathrm{P})}_{ri}$:

$$
  \begin{gathered}
  Y_{\mathrm{M},i}
  = b_{0,\mathrm{M}}
  + \color{#1769AA}{a_{\mathrm{M}} \times X^{(\mathrm{A})}_{\mathrm{M},i}}
  + \color{#7B3FA1}{p_{\mathrm{M}} \times X^{(\mathrm{P})}_{\mathrm{M},i}}
  + \epsilon_{\mathrm{M},i}
  \\
  Y_{\mathrm{F},i}
  = b_{0,\mathrm{F}}
  + \color{#1769AA}{a_{\mathrm{F}} \times X^{(\mathrm{A})}_{\mathrm{F},i}}
  + \color{#7B3FA1}{p_{\mathrm{F}} \times X^{(\mathrm{P})}_{\mathrm{F},i}}
  + \epsilon_{\mathrm{F},i}
  \end{gathered}
$$

[**- Actor effects**]{style="color: #1769AA"} [**- Partner effects**]{style="color: #7B3FA1"}

In long format, each equation applies to one member row of the dyad.

How can we combine these two equations into one model?


## Role indicators combine the equations

The indicators $D_{\mathrm{M},ri}$ and $D_{\mathrm{F},ri}$ identify the male
or female row:

$$
\begin{aligned}
Y_{ri} ={}&
\underbrace{
  b_{0,\mathrm{M}} \times D_{\mathrm{M},ri}
  + b_{0,\mathrm{F}} \times D_{\mathrm{F},ri}
}_{\text{role-specific intercepts}}
\\[0.35em]
&+
\color{#1769AA}{
\underbrace{
  a_{\mathrm{M}} \times D_{\mathrm{M},ri} \times X^{(\mathrm{A})}_{ri}
  + a_{\mathrm{F}} \times D_{\mathrm{F},ri} \times X^{(\mathrm{A})}_{ri}
}_{\text{role-specific actor effects}}
}
\\[0.35em]
&+
\color{#7B3FA1}{
\underbrace{
  p_{\mathrm{M}} \times D_{\mathrm{M},ri} \times X^{(\mathrm{P})}_{ri}
  + p_{\mathrm{F}} \times D_{\mathrm{F},ri} \times X^{(\mathrm{P})}_{ri}
}_{\text{role-specific partner effects}}
}
+ \epsilon_{ri}.
\end{aligned}
$$

:::{.text-smaller}

For a male row: $D_\mathrm{M}=1,\ D_\mathrm{F}=0$. For a female row: the reverse.

:::

:::{.notes}

The role indicators correspond to `.is_male` and `.is_female`. The actor and
partner values correspond to `.efficacy_actor` and `.efficacy_partner`.
The role index is written first, followed by the dyad index, as in
$Y_{\mathrm{male},i}$ and $Y_{\mathrm{female},i}$.

[Sources]

- @kenny2006dyadic

:::

## Residuals remain correlated within dyads

The two member rows from the same dyad retain a joint residual structure:

$$
\operatorname{Cov}
\begin{pmatrix}
\epsilon_{\mathrm{F}i} \\
\epsilon_{\mathrm{M}i}
\end{pmatrix}
= \boldsymbol{\Sigma}_{\epsilon}
=
\begin{bmatrix}
\sigma_{\epsilon_\mathrm{F}}^{2}
&
\rho_{\epsilon_\mathrm{F}\epsilon_\mathrm{M}}
\sigma_{\epsilon_\mathrm{F}}\sigma_{\epsilon_\mathrm{M}} \\
\rho_{\epsilon_\mathrm{F}\epsilon_\mathrm{M}}
\sigma_{\epsilon_\mathrm{F}}\sigma_{\epsilon_\mathrm{M}}
&
\sigma_{\epsilon_\mathrm{M}}^{2}
\end{bmatrix}
$$


## Fitting the distinguishable APIM in glmmTMB

```{r}
#| code-line-numbers: "1-23|4|5-6|8-10|12-14|16-18|19-20|23|1-23"
#| output-location: slide

library(glmmTMB)

apim_distinguishable_model <- glmmTMB(
  total_mvpa ~
    # Role-specific intercepts
    0 + .is_female + .is_male +

    # Role-specific actor effects
    .is_female:.efficacy_gmc_actor +
    .is_male:.efficacy_gmc_actor +

    # Role-specific partner effects
    .is_female:.efficacy_gmc_partner +
    .is_male:.efficacy_gmc_partner +

    # Within-dyad residual covariance
    us(0 + .is_female + .is_male | couple_id),
  dispformula = ~ 0,
  family = gaussian(),
  data = df_apim
)

summary(apim_distinguishable_model)

  
```

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. [**Verify model adequacy, distinguishability, and robustness**]{.text-highlight}
6. **Visualize, interpret, and report results**

:::{.notes}

BEFORE interpreting and reporting results, we should ...

:::

## Verify model adequacy, distinguishability, and robustness

1. Did estimation succeed?
2. Does the model reproduce the important data patterns?
3. Are exchangeability constraints defensible?
4. Are conclusions driven by one dyad or one consequential modeling choice?

## Did estimation succeed / converge?

The model converged without warnings... a good first sign!

:::{.fragment}
**Other things to look out for in the model output:**

- Estimated variances near $0$ or correlations near $\pm 1$ $\rightarrow$ possible boundary fit
- `NaN` or infinite estimates, standard errors, or model fit indices

:::

## Residual diagnostics / reproduction of data patterns

Residual diagnostics are not as straightforward anymore

:::{.incremental}
- Linear models: straightforward checks
- Mixed models: residuals arise at multiple levels
- Generalized models: expectations depend on the family and link
- Dyadic models: partners' residuals are correlated

:::

:::{.fragment}

**Solution:** Compare the data with simulations from the fitted model.

:::

## DHARMa: one diagnostic scale across models

The `DHARMa` package [@hartigDHARMaResidualDiagnostics2026]:

- Simulates outcomes from the fitted model
- Checks whether the observed residuals behave as expected under simulations from the fitted model.
- Adequate fit $\rightarrow$ approximately uniform, pattern-free residuals
- Offers more tests in its [vignette](https://stat.ethz.ch/CRAN/web/packages/DHARMa/vignettes/DHARMa.html)

One function works across many model families!

## Simulate the full dyadic residual structure

```{r}
library(DHARMa)

dharma_residuals <- DHARMa::simulateResiduals(
  fittedModel = apim_distinguishable_model,
  n = 10000,
  
  # Since our model uses dispformula = ~ 0, we cannot use DHARMa's default
  # conditional simulations. With unconditional, the us() block representing
  # our residuals is re-simulated.
  simulateREs = "unconditional",
  
  # Male and female residuals are correlated. DHARMa estimates this correlation
  # from the simulations and rotates the residuals to "remove" it.
  rotation = 'estimated',
  seed = 123
)
```

## Raw MVPA leaves a fitted-value pattern {.smaller}

:::: {.columns}

::: {.column width="75%"}

```{r}
#| fig-format: svg
#| echo: false

plot(dharma_residuals)

```

:::

::: {.column width="25%"}

**In this model:**

- Uniformity and dispersion are plausible
- No simulation outliers
- A fitted-value pattern remains

**Next:** inspect roles without rotation.

:::

::::

More targeted checks and guidance on interpretation of the plots:
[DHARMa vignette](https://stat.ethz.ch/CRAN/web/packages/DHARMa/vignettes/DHARMa.html).

## Inspect residuals by role

```{r}

female_dharma_residuals <- DHARMa::recalculateResiduals(
  dharma_residuals,
  
  # Recalculate from the same simulations for one role.
  # One observation per couple: no rotation needed.
  sel = ~ .is_female == 1,
  rotation = NULL,
  seed = 123
)

male_dharma_residuals <- DHARMa::recalculateResiduals(
  dharma_residuals,
  sel = ~ .is_male == 1,
  rotation = NULL,
  seed = 123
)
```

## The fitted-value pattern appears among women {.smaller}


:::: {.columns}

::: {.column width="75%"}

```{r}
#| echo: false
#| fig-format: svg
#| fig-width: 10
#| fig-height: 4.6
#| out-width: 100%

old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 4.5, 1))
DHARMa::plotResiduals(
  female_dharma_residuals,
  main = "Women"
)
DHARMa::plotResiduals(
  male_dharma_residuals,
  main = "Men"
)
par(old_par)
```


:::

::: {.column width="25%"}

**In this model:**

- Pattern detected among women
- No comparable pattern among men

**Next:** try a defensible outcome scale.


:::

::::

More targeted checks and guidance on interpretation of the plots:
[DHARMa vignette](https://stat.ethz.ch/CRAN/web/packages/DHARMa/vignettes/DHARMa.html).

## Square-root MVPA reduces the fitted-value pattern {.smaller}

```{r}
#| include: false

df_apim_sqrt <- df_apim |>
  dplyr::mutate(sqrt_total_mvpa = sqrt(total_mvpa))

apim_distinguishable_sqrt_model <- update(
  apim_distinguishable_model,
  sqrt_total_mvpa ~ .,
  data = df_apim_sqrt
)

sqrt_dharma_residuals <- DHARMa::simulateResiduals(
  fittedModel = apim_distinguishable_sqrt_model,
  n = 10000,
  simulateREs = "unconditional",
  rotation = 'estimated',
  seed = 123
)
```

:::: {.columns}

::: {.column width="70%"}

```{r}
#| echo: false
#| fig-format: svg

plot(sqrt_dharma_residuals)

```
:::

::: {.column width="30%"}

**Compared with raw MVPA:**

- Uniformity and dispersion are plausible
- No simulation outliers
- No pooled fitted-value pattern

**Substantive result unchanged:** actor effects remain positive and supported;
partner effects remain unsupported.

:::

::::

:::{.subtle}

Treat this as a prespecified or transparently reported sensitivity analysis—not
a scale selected by diagnostic p-values.

:::

## Test distinguishability

- Fit model with exchangeability constraints

- Compare models

- Pool only if theoretically defensible, aligned with RQ and theory, and 
compatible with the data
  - Depending on goals you can also report the distinguishable model and 
  report results from this analysis

However, non-significant comparisons are not proof that roles are identical.

[@gistelinckIndistinguishabilityTests2018]{.subtle}

## Estimate and compare: 

```{r include=FALSE}

apim_exchangeable_model <- glmmTMB(
  total_mvpa ~
    # Pooled intercept
    1 +

    # Pooled actor effect
    .efficacy_gmc_actor +

    # Pooled partner effect
    .efficacy_gmc_partner +

    # Within-dyad residual covariance
    us(1 | couple_id) +
    us(0 + .member_contrast_arbitrary | couple_id),
  dispformula = ~ 0,
  family = gaussian(),
  data = df_apim
)

summary(apim_exchangeable_model)

dyadMLM::recover_exchangeable_covariance(apim_exchangeable_model)

```

```{r}

dyadMLM::compare_nested_models(
  apim_distinguishable_model, apim_exchangeable_model
)

```

Depending on your analysis goals and preregistration, you may switch to the
more parsimonious exchangeable model, or report the distinguishable model while
reporting no improved fit. In our example, we keep going with the distinguishable
model.

## Visualize role-specific actor associations {.smaller}

```{r}
plot_actor_association <- function(data, model, role, title,
                                   n_mahalanobis = 3) {
  xy <- cbind(data$.efficacy_gmc_actor, data$total_mvpa)
  labels <- data |>
    dplyr::mutate(.distance = mahalanobis(xy, colMeans(xy), cov(xy))) |>
    dplyr::slice_max(.distance, n = n_mahalanobis)
  coefficients <- glmmTMB::fixef(model)$cond

  ggplot(data, aes(.efficacy_gmc_actor, total_mvpa)) +
    geom_point(color = "steelblue") +
    geom_abline(intercept = coefficients[role],
                slope = coefficients[paste0(role, ":.efficacy_gmc_actor")],
                color = "navy", linewidth = 16 / 11) +
    geom_text(
      data = labels, aes(label = couple_id),
      vjust = -0.6, check_overlap = TRUE, size = 4
    ) +
    labs(
      title = title,
      x = "Actor efficacy (grand-mean centered)",
      y = "Daily MVPA (minutes)"
    ) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
    see::theme_modern(base_size = 16) +
    theme(plot.title.position = "panel",
          plot.title = element_text(hjust = 0.5))
}
```

## Couple 28 is still the most unusual {#inspect-unusual-dyad}

```{r}
#| output-location: slide
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.2
#| out-width: 100%
#| fig-cap: "Labels: four largest Mahalanobis distances in each plot."

n_mahalanobis <- 4

see::plots(
  plot_actor_association(
    dplyr::filter(df_apim, gender == "female"),
    apim_distinguishable_model, ".is_female", "Women",
    n_mahalanobis = n_mahalanobis
  ),
  plot_actor_association(
    dplyr::filter(df_apim, gender == "male"),
    apim_distinguishable_model, ".is_male", "Men",
    n_mahalanobis = n_mahalanobis
  )
) +
  patchwork::plot_layout(
    axes = "collect",
    axis_titles = "collect"
  )
```



:::{.notes}

Here Mahalanobis distance is calculated separately for women and men from the
two quantities shown in each panel: actor efficacy and raw MVPA. The four
largest distances are Women: Couples 28, 4, 25, and 20; Men: Couples 28, 2, 25,
and 9.

This is a descriptive screen for unusual value combinations. A distant point
can still lie close to the fitted line, and Mahalanobis distance does not say
how much deleting that dyad changes a coefficient. The next slide answers that
question directly.

The points show raw MVPA. The lines show the full-sample APIM fixed-effect actor
associations with partner efficacy held at its sample mean. Therefore, they are
not simple bivariate regression lines through the raw points.

Because efficacy is grand-mean centered, the fitted APIM intercept and actor
slope give the line for each role when partner efficacy equals its sample mean.
`see` arranges and styles the two plots.

:::

## No leave-one-dyad-out refit reverses either actor effect {#sensitivity-without-unusual-dyad .smaller}

```{r}
#| include: false

apim_cv <- cv::cv(
  apim_distinguishable_model,
  data = df_apim,
  clusterVariables = "couple_id",
  details = TRUE
)

actor_dfbeta <- as.data.frame(
  apim_cv,
  rows = "folds",
  columns = "coefficients"
) |>
  tibble::as_tibble() |>
  dplyr::transmute(
    couple_id = unique(df_apim$couple_id),
    Women =
      apim_cv$coefficients[
        "cond..is_female:.efficacy_gmc_actor"
      ] -
      .data[[
        "coef.cond..is_female:.efficacy_gmc_actor"
      ]],
    Men =
      apim_cv$coefficients[
        "cond..is_male:.efficacy_gmc_actor"
      ] -
      .data[[
        "coef.cond..is_male:.efficacy_gmc_actor"
      ]]
  ) |>
  tidyr::pivot_longer(
    c(Women, Men),
    names_to = "role",
    values_to = "dfbeta"
  ) |>
  dplyr::mutate(
    role = factor(role, levels = c("Women", "Men")),
    marked = dplyr::min_rank(
      dplyr::desc(abs(dfbeta))
    ) <= 4,
    labelled = dplyr::min_rank(
      dplyr::desc(abs(dfbeta))
    ) <= 2,
    .by = role
  )

cv_influence_plot <- ggplot(
  actor_dfbeta,
  aes(x = couple_id, y = dfbeta)
) +
  geom_hline(
    yintercept = 0,
    color = "#1769AA", linewidth = 1
  ) +
  geom_point(color = "#9AA5AD", alpha = 0.8, size = 2.4) +
  geom_point(
    data = dplyr::filter(actor_dfbeta, marked),
    color = "#D55E00", size = 4
  ) +
  geom_text(
    data = dplyr::filter(actor_dfbeta, labelled),
    aes(
      label = couple_id,
      vjust = dplyr::if_else(dfbeta > 0, -0.5, 1.4)
    ),
    color = "#D55E00", fontface = "bold", size = 4
  ) +
  facet_wrap(~role, nrow = 1) +
  scale_x_continuous(breaks = seq(5, 35, by = 5)) +
  coord_cartesian(ylim = c(-4.5, 4.5)) +
  labs(
    x = "Couple omitted",
    y = "Actor-effect DFBETA\n(full − leave-one-dyad-out)"
  ) +
  see::theme_modern(base_size = 16) +
  theme(strip.text = element_text(face = "bold"))
```

:::: {.columns}

::: {.column width="42%"}

```{r}
#| eval: false

# Leave each whole dyad out once
apim_cv <- cv::cv(
  apim_distinguishable_model,
  data = df_apim,
  clusterVariables = "couple_id",
  details = TRUE
)

# Compare the actor coefficients
actor_terms <- grep(
  "actor$",
  names(apim_cv$coefficients),
  value = TRUE
)

deleted_actor <- as.data.frame(
  apim_cv,
  rows = "folds",
  columns = "coefficients"
) |>
  select(contains("actor")) |>
  as.matrix()

actor_dfbeta <- sweep(
  deleted_actor,
  2,
  apim_cv$coefficients[actor_terms],
  \(deleted, full) full - deleted
)
```

:::

::: {.column width="58%"}

```{r}
#| echo: false
#| fig-format: svg
#| fig-width: 8
#| fig-height: 5.2
#| out-width: 100%

cv_influence_plot
```

:::

::::

:::{.subtle}
**Result:** Couple 28 most changes the women's actor effect; Couple 2 most
changes the men's. Across all 36 refits, both actor coefficients remain
positive. [@foxCrossValidatingRegressionModels2026]
:::

:::{.notes}

With `clusterVariables = "couple_id"`, `cv::cv()` omits both partners together.
Cluster-level cross-validation defaults to one fold per cluster. Because 36
folds exceed the package's automatic detail threshold, `details = TRUE` is
needed to retain the coefficient estimates.

The plotted DFBETA is the unstandardized full-sample actor coefficient minus
the corresponding leave-one-dyad-out coefficient. We use the refits to assess
coefficient influence and do not interpret the cross-validation MSE here.

The four largest changes are Women: Couples 28, 4, 20, and 3; Men: Couples 2,
28, 9, and 20. The two largest per role are labelled. All deleted-dyad actor
estimates remain positive.

The fold order follows the first appearance of each cluster in the model data;
here that order is Couples 1 through 36.

[Sources]

- @foxCrossValidatingRegressionModels2026

:::

## Alternative: summarize the same refits in a table {.smaller}

```{r}
#| include: false

full_actor_effects <- c(
  Women = unname(
    glmmTMB::fixef(apim_distinguishable_model)$cond[
      ".is_female:.efficacy_gmc_actor"
    ]
  ),
  Men = unname(
    glmmTMB::fixef(apim_distinguishable_model)$cond[
      ".is_male:.efficacy_gmc_actor"
    ]
  )
)

loo_actor_summary <- actor_dfbeta |>
  dplyr::mutate(
    full_model = unname(
      full_actor_effects[as.character(role)]
    ),
    leave_one_out = full_model - dfbeta
  ) |>
  dplyr::summarise(
    `Full model` = dplyr::first(full_model),
    `Lowest estimate` = sprintf(
      "%.1f (omit Couple %s)",
      min(leave_one_out),
      couple_id[which.min(leave_one_out)]
    ),
    `Highest estimate` = sprintf(
      "%.1f (omit Couple %s)",
      max(leave_one_out),
      couple_id[which.max(leave_one_out)]
    ),
    `Across 36 refits` = dplyr::if_else(
      all(leave_one_out > 0),
      "Always positive",
      "Changed sign"
    ),
    .by = role
  ) |>
  dplyr::rename(`Actor effect` = role)
```

::: {.loo-summary-table}

```{r}
#| echo: false

slide_table(
  loo_actor_summary,
  digits = 1,
  align = c("l", "r", "l", "l", "l")
)
```

:::

:::{.subtle}
**Result:** The complete leave-one-dyad-out range remains positive for both
actor effects. [@foxCrossValidatingRegressionModels2026]
:::

:::{.notes}

This table summarizes the same 36 refits as the preceding plot. It reports the
full-model estimate, the lowest and highest deleted-dyad estimates, and the
dyad omitted at each endpoint.

The table makes the conclusion and numerical range faster to read. The plot
retains more information about the full distribution of influence across
dyads. These are alternative presentations of the same sensitivity analysis,
not two separate analyses.

[Sources]

- @foxCrossValidatingRegressionModels2026

:::

## Test robustness - sensitivity analyses

- Refit a small number of prespecified, defensible alternatives where relevant:
  - coding or centering decisions
  - different plausible outcome families or transformations
  - missing-data handling
  - covariates with a substantive confounding or precision rationale

Report any material changes in the conclusions.

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. [**Visualize, interpret, and report results**]{.text-highlight}

## Your turn: answer the research question {.smaller}

**Write three sentences:** interpret the actor and partner associations,
describe the remaining within-couple correlation, and qualify the result using
the model comparison and sensitivity analyses.

```{r}
apim_model_summary <- summary(apim_distinguishable_model)
apim_model_summary$varcor
stats::printCoefmat(apim_model_summary$coefficients$cond)
```

## Own efficacy showed the clearest associations {.smaller}

```{r}
#| include: false

apim_fixed_ci <- stats::confint(
  apim_distinguishable_model,
  parm = "beta_",
  method = "wald"
)

apim_focal_terms <- c(
  ".is_female:.efficacy_gmc_actor",
  ".is_male:.efficacy_gmc_actor",
  ".is_female:.efficacy_gmc_partner",
  ".is_male:.efficacy_gmc_partner"
)

apim_focal_labels <- c(
  "Own efficacy → women's MVPA",
  "Own efficacy → men's MVPA",
  "Partner efficacy → women's MVPA",
  "Partner efficacy → men's MVPA"
)

apim_focal_results <- tibble::tibble(
  term = apim_focal_terms,
  label = factor(
    apim_focal_labels,
    levels = rev(apim_focal_labels)
  ),
  association = factor(
    c("Actor", "Actor", "Partner", "Partner"),
    levels = c("Actor", "Partner")
  ),
  estimate = apim_fixed_ci[apim_focal_terms, "Estimate"],
  lower = apim_fixed_ci[apim_focal_terms, "2.5 %"],
  upper = apim_fixed_ci[apim_focal_terms, "97.5 %"]
)

apim_within_couple <- VarCorr(
  apim_distinguishable_model
)$cond$couple_id

apim_within_correlation <- attr(
  apim_within_couple,
  "correlation"
)[1, 2]

apim_distinguishability_test <- stats::anova(
  apim_exchangeable_model,
  apim_distinguishable_model
)

apim_distinguishability_p <- apim_distinguishability_test[
  2,
  "Pr(>Chisq)"
]

apim_results_plot <- ggplot(
  apim_focal_results,
  aes(x = estimate, y = label, color = association)
) +
  geom_vline(
    xintercept = 0,
    color = "#6C757D",
    linetype = "dashed"
  ) +
  geom_segment(
    aes(x = lower, xend = upper, yend = label),
    linewidth = 1.8,
    lineend = "round"
  ) +
  geom_point(size = 4.5) +
  scale_color_manual(
    values = c("Actor" = "#1769AA", "Partner" = "#7B3FA1")
  ) +
  scale_x_continuous(
    breaks = seq(-5, 20, by = 5),
    limits = c(-7, 21)
  ) +
  labs(
    x = "Change in daily MVPA (minutes)\nper one-point higher efficacy",
    y = NULL,
    color = NULL
  ) +
  see::theme_modern(base_size = 16) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    axis.text.y = element_text(color = "#273746"),
    panel.grid.major.y = element_blank()
  )
```

:::: {.columns}

::: {.column width="68%"}

```{r}
#| echo: false
#| fig-format: svg
#| fig-width: 8
#| fig-height: 5
#| out-width: 100%

apim_results_plot
```

:::

::: {.column width="32%"}

**One-point higher own efficacy**

- Women: **`r sprintf("%.1f", apim_focal_results$estimate[1])` minutes/day**
- Men: **`r sprintf("%.1f", apim_focal_results$estimate[2])` minutes/day**

**Partner efficacy**

- Both intervals include zero

**Dyadic dependence**

- Within-couple correlation: **`r sprintf("r = %.2f", apim_within_correlation)`**

:::

::::

:::{.subtle}
95% Wald confidence intervals. The distinguishable model did not clearly
improve fit over the exchangeable model
(*p* = `r sprintf("%.3f", apim_distinguishability_p)`). Associations are not causal.
:::

:::{.notes}

The four estimates answer the research question directly. Each is the expected
difference in daily MVPA minutes associated with a one-point difference in
self-efficacy, holding the other predictor constant.

The actor estimates are 11.9 minutes for women (95% CI 3.8 to 20.0) and 8.0
minutes for men (95% CI 0.9 to 15.0). The partner estimates are 3.1 minutes for
women's MVPA (95% CI -5.2 to 11.3) and 4.0 minutes for men's MVPA (95% CI -2.9
to 11.0).

Separate confidence intervals do not test whether the role-specific
coefficients differ. The likelihood-ratio comparison found no clear
improvement from the distinguishable model, so we do not claim gender
differences.

The square-root outcome retained the actor-versus-partner conclusion, and all
leave-one-dyad-out actor coefficients remained positive.

:::

## What should be reported?

- **Analysis sample:** dyads and observations, exclusions, aggregation, and missing-data handling
- **Model specification:** outcome family and link, fixed effects, dyadic covariance structure, and equality constraints
- **Estimation and model decision:** estimator, software and version, convergence, boundary estimates, distinguishability comparison, and retained model
- **Results:** focal estimates with 95% confidence intervals in substantive units and the fitted variance-covariance parameters
- **Checks and interpretation:** residual diagnostics, influential dyads, prespecified sensitivity analyses, material changes, and causal limits



# Exchangeable dyads
The same model but constrained

## Workflow

1. [**Define dyads, distinguishability, and research questions**]{.text-highlight}
2. **Prepare the data and validate the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Define dyads, distinguishability, and research questions

> Same data, RQ, and dyads, but now treated as exchangeable!

:::{.notes}

Since we now know that a distinguishable model is not better than an exchangeable one,
we can use the same RQ and data to demonstrate the exchangeable workflow.

:::

## Prepare the Data



<!-------------------------  Work in progress  -------------------------------->


## Exchangeable APIM

- Row-Level representation
- Formula
- residual matrix?
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships
  - Estimating the model
  - Checking assumptions
  - Visualizing results

## DIM (potentially omit)

- Row-Level representation
- Formula
- residual matrix?
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships
  - Estimating the model
  - Checking assumptions
  - Visualizing results

## Excursion: non-Gaussian generalized models

## Q&A and discussion

# Intensive longitudinal extensions

## ILD extends rather than replaces the workflow {.smaller}

:::: {.columns}

::: {.column width="43%"}

### The same core workflow

::: {.subtle}

1. Define dyads, roles, distinguishability, and research questions
2. Check data structure, completeness, and missing-data handling
3. Describe measures and dyadic relationships
4. Specify, document, and estimate the APIM
5. Verify estimation; check residuals and influential dyads
6. Interpret, visualize, and report the results

:::

:::

::: {.column width="57%"}

### ILD additionally requires

- Check compliance, matched occasions, and actual time gaps
  [@revolPreprocessingExperienceSampling2024]
- Plot paired trajectories and inspect raw temporal and cross-partner patterns
  [@bolgerIntensiveLongitudinalMethods2013;
  @revolPreprocessingExperienceSampling2024]
- Separate within- and between-person variation
  - ICCs, level-specific correlations, and multilevel reliability where
    applicable [@bolgerIntensiveLongitudinalMethods2013;
    @hamakerCuriousCaseCrossSectional2024;
    @geldhofReliabilityEstimationMultilevel2014]
- Choose the temporal specification
  - Time scale and lag, trends/stationarity, concurrent versus lagged effects,
    and discrete versus continuous time
    [@driverContinuousTimeStructural2017]
- Check residual temporal and cross-partner dependence

:::

::::

::: {.subtle}
Workflow synthesized from @bolgerIntensiveLongitudinalMethods2013,
@revolPreprocessingExperienceSampling2024, @mcneishPrimerTwoLevel2020, and
@delrosarioPracticalGuideSpecifying2025.
:::

:::{.notes}

For ILD-specific power analysis, see @lafitPowerLAPIMApplication2022. Perry et
al. belongs with interpretation of fitted dynamics: their change-score APIM and
vector-field diagrams show the joint implications of actor and partner
coefficients [@perryGraphicMethodsInterpreting2017].

We will use `wbCorr` for pooled within- and between-person correlations
[@kuengWbCorrBivariateWithin2026]; @hamakerCuriousCaseCrossSectional2024
provides the methodological rationale.

`omegaSEM()` is provided by `multilevelTools`, not `semTools`. It implements
level-specific internal-consistency coefficients based on a two-level
one-factor model [@wileyMultilevelTools2025;
@geldhofReliabilityEstimationMultilevel2014]. These coefficients are not
temporal stability or the precision of observed person means; for the latter,
see @laiCompositeReliabilityMultilevel2021.

:::

## Distinguishable APIM

- Formula
- Random effects and residual covariance explained
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships on both levels (wbCorr)
  - Estimating the model
  - Checking assumptions
  - Visualizing results
  
## Exchangeable APIM

- Formula
- Random effects and residual covariance explained
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships on both levels (wbCorr)
  - Estimating the model
  - Checking assumptions
  - Visualizing results

## DIM (potentially omit)

- Code
  - Data prep and inspecting and unerstanding data structure
  - Estimating the model show equivalence on both models
  - Checking assumptions
  - Visualizing results

## AR1, OU, and Dynamic Models

- When to use dynamic models vs. concurrent models
- Current possibilities in R MLM for AR1 or ou models
- Current possibilites in R MLM for dynamic models
- Alternatives (DSEM in MPLUS or lavaan for X)

## Power analysis is the clearest reporting gap

```{r}
#| label: plot-longitudinal-apim-reporting-gaps
#| fig-format: svg
#| echo: false
#| fig-width: 10.5
#| fig-height: 4.5
#| out-width: 92%
#| fig-align: center
#| fig-alt: "Bar chart of reporting practices in 73 longitudinal APIM studies: 17.8 percent reported a power analysis, 64.4 percent documented model constraints, and 75.3 percent reported a missing-data strategy."

reporting_gaps <- tibble(
  item = factor(
    c("Power analysis", "Model constraints", "Missing-data strategy"),
    levels = c("Missing-data strategy", "Model constraints", "Power analysis")
  ),
  reported = c(17.8, 64.4, 75.3)
)

ggplot(reporting_gaps, aes(item, reported, fill = item == "Power analysis")) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = paste0(reported, "%")),
    hjust = -0.15, size = 7
  ) +
  coord_flip(clip = "off") +
  scale_fill_manual(values = c("TRUE" = "#0028a5", "FALSE" = "#9aa7bd")) +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 20),
    labels = paste0(seq(0, 100, 20), "%"),
    expand = c(0, 0)
  ) +
  labs(
    x = NULL,
    y = "Studies reporting the item"
  ) +
  theme_minimal(base_size = 21) +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(5.5, 20, 5.5, 5.5)
  )
```

Systematic review of 73 longitudinal APIM studies published from 2008–2025
[@mohammadzadehYazdBeyondIndividual2026].

:::{.notes}

- The chart shows reporting rates because they are easier to read as adherence
  to good practice. The paper directly reports 17.8% for power analyses; 64.4%
  and 75.3% are the complements of the reported omission rates.
- This was a methodological systematic review, not a meta-analysis of effect
  sizes. Its sample was restricted to longitudinal APIM studies with at least
  70 dyads and follow-up longer than one year.
- The review therefore supports “reporting gaps remain common,” not the broader
  claim that all dyadic analyses are reported inadequately.

[Sources]

- Mohammadzadeh Yazd et al. (2026), abstract:
  <https://doi.org/10.1177/02654075261443593>

:::

## Make every dyadic analysis decision visible

1. **Preregister** actor and partner hypotheses and key analysis decisions.
2. **Plan and report power** using dyad-specific tools.
3. **Explain missing data:** describe missingness and the handling strategy.
4. **Specify the full model:** roles and distinguishability, actor and partner
   paths, constraints, and covariance structure.

The review also recommends multi-method measurement and more diverse
populations

[@mohammadzadehYazdBeyondIndividual2026]{.subtle}


# Excercises in R

Short break - initiate package install before you leave

```{r eval = FALSE}

install.packages(
  c(
    "dyadMLM", "wbCorr", "glmmTMB", "DHARMa", "cv",
    "modelbased", "see",
    "dplyr", "ggplot2", "tibble"
  ),
  repos = c(
    "https://pascal-kueng.r-universe.dev",
    "https://cloud.r-project.org"
  )
)

```


- Go to https://pascal-kueng.github.io/dyadMLM/workshop/ 
- or Teams channel

<!--

- DHARMa plots!?! -> as required package for workshop -> link to materials or do in tutorial? Also check AR1 etc., etc. can dharma do this if resid is set to almost zero?

- glmmTMB -> as required 
- dyadMLM -> as required (via R-universe, see instructions).
- dplyr -> required
- tibble -> required
- lavaan -> required
- ggplot2 -> required

-------------->

<!-- ILD-part: 

- see dev note on ar1 processes careful to consider degrees of freedom! relevant here?
  - TODO @codex: are degrees of freedom alreaedy important in the cross-sectional case and should we teach anything regarding those?
- see plots on: https://quantdev.ssri.psu.edu/tutorials/apa-ati-intensive-longitudinal-data-session-m-dyadic-multilevel-model

-->

<!-- Excercises part: separate files! Include on overview page?

Create exercises with data visalization, etc. Panel plots, histograms
correlations (wb corr)?

-->

## References
