---
title: "Dyadic Data Analysis with R"
subtitle: "Part II: Applied Tutorial"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
csl: apa.csl
resources:
  - uzh-logo.svg
  - charite-logo.svg
format:
  revealjs:
    embed-resources: true
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: false
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

## Today's aims and program

| Time | Focus |
|---|---|
| 11:00–12:45 | Introduction and types of dyadic models |
| 12:45–14:00 | *Lunch break*|
| 14:00–15:30 | Estimating cross-sectional dyadic models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

<!-- TODO @codex: please grey out the first two points, intro and lunch break). -->

<!-- ca. 2 Minutes -->

# Software and frameworks

## SEM vs. MLM {.smaller}

- Many simple dyadic models can be represented in either framework.
- The best framework is the one that represents your conceptual model most naturally [@ledermannAnalyzingDyadicData2017].

:::: {.columns .fragment}

::: {.column width="50%"}

**SEM / DSEM is often more natural for**

- Latent dyadic constructs (e.g., a CFM)
- Item-level measurement models for reflective measures
- Linked outcomes or equations [e.g., mediation; @ledermannMediationDyadicData2009]
- Short dynamic panels in wide-format [e.g., LD-APIM; @gistelinckMultilevelAutoregressiveSmall2021]
- Intensive lagged or coupled dynamics via DSEM/RDSEM [e.g., @asparouhovDynamicStructuralEquation2018]

:::

::: {.column width="50%"}

**MLM is often more natural for**

- Parsimonious observed-variable models with modest numbers of dyads
- Nested grouping and flexible random-effect structures
- Observed-outcome growth or concurrent ILD models with flexible time trends
- Complex observed-variable regressions—interactions, polynomials, splines,
  and dyadic RSA—especially with repeated measures or additional nesting
- Specialized generalized mixed models in R, including zero-inflated, hurdle,
  and Tweedie models

:::

::::

::: {.subtle}
[@asparouhovDynamicStructuralEquation2018;
@asparouhovComparisonModelsAnalysis2020;
@driverContinuousTimeStructural2017;
@gistelinckMultilevelAutoregressiveSmall2021;
@delrosarioPracticalGuideSpecifying2025;
@schonbrodtTestingSimilarityEffects2018;
@nestlerResponseSurfaceAnalysis2019;
@brooksGlmmTMBBalancesSpeed2017]
:::

## Missing data: ML, FIML, and MI {.smaller}

- **MLM: Complete-case**
  - **Outcome missing:** Valid under MAR, ML uses remaining observations
  - **Predictor missing:** Valid under MAR, if missingness does **not depend**
    on the outcome after accounting for predictors
- **SEM with FIML and joint likelihood**
  - Valid under MAR: Models predictors and outcomes jointly
- **MI: Imputation**
  - Handles missing outcomes, predictors, and auxiliary information
  - Must match the analysis and preserve person, dyad, role, and time structure

::: {.subtle}
Include important predictors of missingness and missing values. FIML and MI
cover more MAR situations than row omission; none automatically solves MNAR.
[@ledermannAnalyzingDyadicData2017;
@hughesAccountingMissingData2019; @grundMultipleImputationMissing2018]
:::

<!-- TODO @codex, actually insert referecnes to the corresponding points -->
<!-- ca. 2.5 Minutes -->

:::{.notes}

`glmmTMB` omits every incomplete model-frame row, whether the missing value is
an outcome or a predictor. With a missing outcome, that occasion contributes no
outcome information, but the person's or dyad's other occasions remain. This
likelihood analysis can be valid under MAR when the outcome and dependence
models are correctly specified.

Missing predictors are different because an ordinary MLM does not model their
distribution. Row omission can still be valid, but only when being retained
does not depend on the outcome after accounting for the modeled predictors.
MAR alone is not sufficient because missingness in a predictor may depend on an
observed outcome.

FIML uses missing predictors only when they are part of the joint likelihood.
For example, `lavaan` must treat them as stochastic rather than fixed exogenous
covariates. If MLM row omission is valid, FIML's main benefit is efficiency. If
it is not valid, FIML may also reduce bias by using observed outcomes and other
variables to inform the missing predictors.

Multiple imputation is more flexible, but its imputation model must be
congenial, or compatible, with the analysis model. It should include the outcome,
all analysis variables, and useful predictors of missingness and missing values.
For dyadic ILD it must also preserve the person, dyad, role, and temporal
structure, including relevant interactions, nonlinearities, and lags.

Including these variables makes MAR more plausible; it does not prove MAR.
None of these standard approaches automatically handles MNAR.

:::

## Why R—and where other software helps {.smaller}

:::: {.columns}

::: {.column width="46%"}

### Why we teach R

- **Free to use:** no license fees or institutional access barriers
- **Open source:** statistical implementations can be inspected, audited, and adapted
- **Reproducible:** workflows can be rerun, reviewed, shared, and version-controlled
- **Extensible:** specialized methods integrate with data, graphics, and reporting

:::

::: {.column width="54%"}

### Current trade-offs

- **SEM/DSEM:** R's capabilities are spread across specialized packages
  - `lavaan` handles explicitly specified, fixed-wave lagged SEMs (e.g.,
    CLPM/RI-CLPM), but has no native DSEM or residual AR(1) structure
  - Version 0.7 adds restricted two-level random slopes; `mlts` implements
    Bayesian multilevel time-series models, while Mplus integrates DSEM/RDSEM
- **MLM/ILD:** `glmmTMB` and `brms` both model serial dependence
  - Neither directly offers a separable partner-by-time residual structure such
    as $\mathrm{UN}\otimes\mathrm{AR}(1)$; SAS `PROC MIXED` and SPSS `MIXED` do
    for Gaussian repeated-measures models

:::

::::

::: {.subtle}
[@lavaanMultilevelSEM;
@mulderThreeExtensions2021;
@blavaanTwoLevelSEM;
@koslowskiMlts2025;
@asparouhovDynamicStructuralEquation2018;
@asparouhovComparisonModelsAnalysis2020;
@kristensenCovarianceStructuresGlmmTMB2023;
@buerknerAutocorrelationStructuresBrms;
@sasInstituteMixedProcedure;
@ibmCovarianceStructureList]
:::

:::{.notes}

- The vignette's concurrent Gaussian ILD APIM can be reparameterized in
  `lavaan` with one row per dyad-occasion and one outcome per member:
  dyad-level intercepts and `rv()` slopes live at level 2, while the members'
  same-occasion residuals covary at level 1. This equivalence does not work
  with the vignette's stacked member rows because that would require a third
  level.
- When a time-varying predictor appears at both levels, `lavaan` can let an
  `rv()` slope act on its latent within-cluster component. This avoids treating
  observed member means as error-free, targeting the problem described by
  Lüdtke et al. [@lavaanRandomSlopeSyntax; @ludtkeMultilevelLatentCovariate2008].
- This does not provide latent dynamic modeling. In fixed-wave, wide-format
  data, latent within-state variables can be linked with autoregressive and
  cross-lagged paths. A precomputed lag in long data does not make `lavaan`
  time-aware or automatically solve endogeneity and initial-condition problems
  [@gistelinckMultilevelAutoregressiveSmall2021].
- Thus, "AR(1)" is ambiguous: lag-1 structural regressions can be written
  explicitly, but a native time-indexed AR(1) residual covariance for
  long-format ILD is not available.
- `lavaan` 0.7 random slopes are currently limited to two-level, single-group,
  continuous-data models fitted with ML; chi-square-derived fit indices are not
  available. These generic random slopes do not make `lavaan` a DSEM engine.
- `blavaan` currently documents two-level random-intercept models for complete,
  continuous data.
- $\mathrm{UN}\otimes\mathrm{AR}(1)$ is a separable residual covariance across
  partner and time, not a VAR: a VAR models own- and cross-lagged effects in the
  conditional mean.

:::

<!-- ca. 2.5 Minutes -->

# Estimating cross-sectional dyadic models in R

## Cross-sectional APIMs follow six steps {.smaller}

1. **Define dyads, roles, distinguishability, and research questions**
2. **Check data structure, completeness, and missing-data handling**
3. **Describe measures and dyadic relationships**
   - Summary statistics, reliability where applicable, partner correlations,
     and raw relationship plots
4. **Specify, document, and estimate the APIM**
5. **Verify estimation; check residuals and influential dyads**
6. **Interpret, visualize, and report the results**

::: {.subtle}
Before data collection: preregister hypotheses and key model decisions, and
conduct a dyad-specific power analysis
[@stasGivingDyadicData2018; @mohammadzadehYazdBeyondIndividual2026].

Workflow synthesized from @kenny2006dyadic,
@ledermannToolboxProgramsRestructure2015, and @stasGivingDyadicData2018.
:::

:::{.notes}

The systematic review examined 73 longitudinal APIM studies with at least 70
dyads and study periods exceeding 1 year, rather than intensive diary designs.
Only 17.8% reported a power analysis, 24.7% did not report their missing-data
strategy, and 35.6% did not document model constraints. We use its
recommendations as general motivation for preregistration, dyad-specific power
analysis, and transparent model specification—not as prevalence estimates for
ILD studies.

:::

## Distinguishable APIM

- Row-Level representation
- Formula
- residual matrix?
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships
  - Estimating the model
  - Checking assumptions
  - Visualizing results
  
  
```{r}
#| label: setup-distinguishable-apim-example
#| include: false

source(
  "../../vignettes/diagram-helpers.Rinc",
  local = knitr::knit_global()
)

dharma_helper_candidates <- c(
  file.path(
    "tmp-dharma-validation",
    "reference-implementation",
    "dyad_dharma.R"
  ),
  file.path(
    "dev",
    "workshop",
    "tmp-dharma-validation",
    "reference-implementation",
    "dyad_dharma.R"
  )
)
dharma_helper <- dharma_helper_candidates[
  file.exists(dharma_helper_candidates)
][1]

if (is.na(dharma_helper)) {
  stop("Could not find the temporary dyadic DHARMa helper.", call. = FALSE)
}

source(dharma_helper, local = knitr::knit_global())
```

## Example: self-efficacy and MVPA {.smaller}

We average the daily observations for each person. `total_mvpa` is therefore
their mean number of MVPA minutes per day.

```{r}
#| label: aggregate-distinguishable-apim-data

library(dplyr)
library(dyadMLM)
library(glmmTMB)
library(ggplot2)

data_file_candidates <- c(
  "dyadic-data.rds",
  file.path("dev", "workshop", "dyadic-data.rds")
)
data_file <- data_file_candidates[file.exists(data_file_candidates)][1]

if (is.na(data_file)) {
  stop("Could not find dev/workshop/dyadic-data.rds.", call. = FALSE)
}

daily_data <- readRDS(data_file)

person_data <- daily_data |>
  transmute(
    couple_id = coupleID,
    person_id = personID,
    role = factor(
      gender,
      levels = c(1, 2),
      labels = c("female", "male")
    ),
    self_efficacy,
    total_mvpa
  ) |>
  summarise(
    self_efficacy = mean(self_efficacy, na.rm = TRUE),
    total_mvpa = mean(total_mvpa, na.rm = TRUE),
    .by = c(couple_id, person_id, role)
  ) |>
  arrange(couple_id, role)
```

## Prepare the distinguishable APIM data {.smaller}

```{r}
#| label: prepare-distinguishable-apim-data

apim_data <- dyadMLM::prepare_dyad_data(
  person_data,
  dyad = couple_id,
  member = person_id,
  role = role,
  predictors = self_efficacy,
  model_types = "apim",
  keep_compositions = "female-male"
)

efficacy_mean <- mean(apim_data$self_efficacy)

apim_data <- apim_data |>
  mutate(
    .dy_self_efficacy_actor =
      .dy_self_efficacy_actor - efficacy_mean,
    .dy_self_efficacy_partner =
      .dy_self_efficacy_partner - efficacy_mean
  )
```

Both actor and partner scores use the same pooled grand mean. Data preparation
retains 38 female--male couples; the complete-case model uses 36.

## Fit the distinguishable APIM {.smaller}

```{r}
#| label: fit-distinguishable-apim

distinguishable_apim <- glmmTMB::glmmTMB(
  total_mvpa ~
    0 +
    .dy_is_female +
    .dy_is_male +
    .dy_is_female:
      .dy_self_efficacy_actor +
    .dy_is_male:
      .dy_self_efficacy_actor +
    .dy_is_female:
      .dy_self_efficacy_partner +
    .dy_is_male:
      .dy_self_efficacy_partner +
    us(
      0 +
        .dy_is_female +
        .dy_is_male |
        couple_id
    ),
  dispformula = ~0,
  family = gaussian(),
  data = apim_data
)
```

The unstructured dyad random effect estimates the two residual variances and
their covariance.

## Print the fixed effects {.smaller}

```{r}
#| label: print-distinguishable-apim

fixed_effects <- summary(distinguishable_apim)$coefficients$cond

rownames(fixed_effects) <- c(
  "Female intercept", "Male intercept",
  "Female actor effect", "Male actor effect",
  "Female partner effect", "Male partner effect"
)

knitr::kable(fixed_effects, digits = 3)
```

:::{.notes}

One point higher average self-efficacy is associated with about 12 additional
MVPA minutes per day for female actors and 8 for male actors. The corresponding
partner estimates are about 3--4 minutes.

:::

## Plot the fitted APIM

```{r}
#| label: plot-distinguishable-apim
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted distinguishable APIM linking self-efficacy to mean daily MVPA for female and male partners."

draw_apim_diagram(
  "distinguishable",
  model = distinguishable_apim,
  labels = c(
    predictor = "Self-\nefficacy",
    outcome = "Mean daily\nMVPA (minutes)"
  ),
  predictors_centered = TRUE
)
```

## Dyadic residual checks answer two questions {.smaller}

- **Overall shape:** Does a QQ plot adjusted for partner dependence follow the
  diagonal?
- **Patterns by role:** Do residual center and spread stay flat over key
  predictors?

::: {.subtle}
First make sure the model fitted without convergence warnings
[@glmmTMBTroubleshooting2026].
:::

:::{.notes}

The helper uses unconditional simulations, including new couple effects, and
then removes the exact fitted Gaussian response covariance within each couple
by blockwise Cholesky whitening. The global QQ coordinates should therefore not
be plotted against the original predictors. The role plots instead reuse the
original, unwhitened simulations. This adapts DHARMa's documented rotation
principle using a known response-scale covariance
[@hartigDHARMaResidualDiagnostics2026].

:::

## The overall residual shape looks reasonable {.smaller}

:::: {.columns}

::: {.column width="40%"}

```r
apim_checks <- simulate_dyad_dharma(
  distinguishable_apim,
  cluster = "couple_id",
  n = 2000,
  seed = 123
)

DHARMa::plotQQunif(
  apim_checks$joint,
  testUniformity = FALSE,
  testDispersion = FALSE,
  testOutliers = FALSE
)
```

:::

::: {.column width="60%"}

```{r}
#| label: plot-distinguishable-apim-global-dharma
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.5
#| out-width: 94%
#| fig-align: center
#| fig-alt: "Uniform quantile-quantile plot of whitened DHARMa residuals from the distinguishable APIM."

model_frame <- stats::model.frame(distinguishable_apim)
is_female <- model_frame$.dy_is_female == 1
actor_efficacy <- model_frame$.dy_self_efficacy_actor

apim_checks <- simulate_dyad_dharma(
  distinguishable_apim,
  cluster = "couple_id",
  n = 2000,
  seed = 123
)

DHARMa::plotQQunif(
  apim_checks$joint,
  testUniformity = FALSE,
  testDispersion = FALSE,
  testOutliers = FALSE
)
```

:::

::::

- **Here:** points mostly follow the diagonal, so there is no obvious problem
  in the overall residual shape after accounting for partner dependence.
- **A systematic curve or tail gap:** the model may still miss a pattern; use
  the next plots to locate it.

::: {.subtle}
This is a broad screen, not proof that every part of the model is correct.
:::

## Role-specific plots suggest checking sparse extremes {.smaller}

:::: {.columns}

::: {.column width="34%"}

```r
plot_dyad_dharma_roles(
  apim_checks,
  predictor = actor_efficacy,
  roles = list(
    Female = is_female,
    Male = !is_female
  ),
  xlab = "Actor self-efficacy"
)
```

:::

::: {.column width="66%"}

```{r}
#| label: plot-distinguishable-apim-role-patterns
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Four panels showing marginal DHARMa residual center and spread patterns over actor self-efficacy, separately for female and male partners."

plot_dyad_dharma_roles(
  apim_checks,
  predictor = actor_efficacy,
  roles = list(
    Female = is_female,
    Male = !is_female
  ),
  xlab = "Actor self-efficacy"
)
```

:::

::::

- **Expected:** black smooths stay roughly horizontal near 0.5.
- **Meaning:** a center trend suggests a missing mean pattern; a spread trend
  suggests changing residual variance.
- **Here:** the strongest bends are at sparse predictor extremes. Inspect those
  couples before changing the model.

::: {.subtle}
Repeat for partner and other key predictors; focus on sustained patterns rather
than p-values.
:::

:::{.notes}

The spread score is twice the absolute distance of a DHARMa residual from 0.5.
Under the fitted model it is also uniform, so its smooth should stay near 0.5.
In this cross-sectional example each role panel has one row per couple. With
intensive longitudinal data, residuals within a role can remain dependent
across occasions, so these panels should be treated as descriptive.

:::

## Influence belongs to the whole couple {.smaller}

Delete one couple at a time and inspect changes in both fixed effects and the
female--male residual correlation.

```{r}
#| label: check-distinguishable-apim-dyad-influence

source(system.file(
  "other_methods", "influence_mixed.R", package = "glmmTMB"
))

apim_influence <- influence_mixed(
  distinguishable_apim, groups = "couple_id", ncores = 1
)

beta_change <- sweep(
  apim_influence[["fixed.effects[-couple_id]"]],
  2, apim_influence[["fixed.effects"]], "-"
)
beta_change <- sweep(
  beta_change, 2, sqrt(diag(apim_influence[["vcov"]])), "/"
)
max_beta_change <- apply(abs(beta_change), 1, max)

deleted_vcov <- apim_influence[["var.cov.comps[-couple_id]"]]
deleted_rho <- deleted_vcov[, 3] / sqrt(deleted_vcov[, 2] * deleted_vcov[, 4])
full_vcov <- apim_influence[["var.cov.comps"]]
full_rho <- full_vcov[3] / sqrt(full_vcov[2] * full_vcov[4])
deleted_id <- as.numeric(as.character(apim_influence$deleted))
```

The official `glmmTMB` helper performs 36 refits. No couple is removed from the
reported model.

## Two couples drive most of the sensitivity {.smaller}

```{r}
#| label: plot-distinguishable-apim-dyad-influence
#| echo: false
#| fig-width: 11
#| fig-height: 4.5
#| out-width: 92%
#| fig-align: center
#| fig-alt: "Two influence panels showing the largest standardized fixed-effect changes and the residual correlations after deleting each couple."

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
largest <- head(order(max_beta_change, decreasing = TRUE), 8)
barplot(
  rev(max_beta_change[largest]),
  names.arg = rev(apim_influence$deleted[largest]),
  horiz = TRUE, las = 1,
  xlab = "Largest |change| in model-SE units"
)

plot(
  deleted_id, deleted_rho, pch = 1,
  ylim = range(deleted_rho, full_rho),
  xlab = "Deleted couple ID", ylab = "Residual correlation"
)
abline(h = full_rho, lty = 2)
influential_rho <- which(apim_influence$deleted == 113)
points(deleted_id[influential_rho], deleted_rho[influential_rho], pch = 19)
text(deleted_id[influential_rho], deleted_rho[influential_rho], "113", pos = 3)
```

Couples 6 and 113 affect fixed effects most; couple 113 also changes the
estimated covariance most. Investigate and report the sensitivity---do not
delete a couple mechanically [@xiangInfluenceDiagnosticsGeneralized2002].

## Exchangeable APIM

- Row-Level representation
- Formula
- residual matrix?
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships
  - Estimating the model
  - Checking assumptions
  - Visualizing results
  
<!-- TODO: @codex better to have separate data prep and inspecting 
workflows or is it enough with the distinguishable, then just explain and estimate
exchangeable, and compare models? Treat it as a single workflow rather than separate
analyses? Then we could remove data-prep and visualizing variables and bivariate relationships (non-dyadic) 

Same for the DIM, just estimate once, backtransform and show that it is then the same
as exchangeable APIM?
-->

## Testing distinguishability

### Wald test for fixed effects equality

### Model comparison

## DIM (potentially omit)

- Row-Level representation
- Formula
- residual matrix?
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships
  - Estimating the model
  - Checking assumptions
  - Visualizing results

## Excursion: non-Gaussian generalized models

## Q&A and discussion

# Intensive longitudinal extensions

## ILD extends rather than replaces the workflow {.smaller}

:::: {.columns}

::: {.column width="43%"}

::: {.subtle}

### The same core workflow

1. Define dyads, roles, distinguishability, and research questions
2. Check data structure, completeness, and missing-data handling
3. Describe measures and dyadic relationships
4. Specify, document, and estimate the APIM
5. Verify estimation; check residuals and influential dyads
6. Interpret, visualize, and report the results

:::

:::

::: {.column width="57%"}

### ILD additionally requires

- Check compliance, matched occasions, and actual time gaps
  [@revolPreprocessingExperienceSampling2024]
- Plot paired trajectories and inspect raw temporal and cross-partner patterns
  [@bolgerIntensiveLongitudinalMethods2013;
  @revolPreprocessingExperienceSampling2024]
- Separate within- and between-person variation
  - ICCs, level-specific correlations, and multilevel reliability where
    applicable [@bolgerIntensiveLongitudinalMethods2013;
    @hamakerCuriousCaseCrossSectional2024;
    @geldhofReliabilityEstimationMultilevel2014]
- Choose the temporal specification
  - Time scale and lag, trends/stationarity, concurrent versus lagged effects,
    and discrete versus continuous time
    [@driverContinuousTimeStructural2017]
- Check residual temporal and cross-partner dependence

:::

::::

::: {.subtle}
Workflow synthesized from @bolgerIntensiveLongitudinalMethods2013,
@revolPreprocessingExperienceSampling2024, @mcneishPrimerTwoLevel2020, and
@delrosarioPracticalGuideSpecifying2025.
:::

:::{.notes}

For ILD-specific power analysis, see @lafitPowerLAPIMApplication2022. Perry et
al. belongs with interpretation of fitted dynamics: their change-score APIM and
vector-field diagrams show the joint implications of actor and partner
coefficients [@perryGraphicMethodsInterpreting2017].

We will use `wbCorr` for pooled within- and between-person correlations
[@kuengWbCorrBivariateWithin2026]; @hamakerCuriousCaseCrossSectional2024
provides the methodological rationale.

`omegaSEM()` is provided by `multilevelTools`, not `semTools`. It implements
level-specific internal-consistency coefficients based on a two-level
one-factor model [@wileyMultilevelTools2025;
@geldhofReliabilityEstimationMultilevel2014]. These coefficients are not
temporal stability or the precision of observed person means; for the latter,
see @laiCompositeReliabilityMultilevel2021.

:::

## Distinguishable APIM

- Formula
- Random effects and residual covariance explained
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships on both levels (wbCorr)
  - Estimating the model
  - Checking assumptions
  - Visualizing results
  
## Exchangeable APIM

- Formula
- Random effects and residual covariance explained
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships on both levels (wbCorr)
  - Estimating the model
  - Checking assumptions
  - Visualizing results

## DIM (potentially omit)

- Code
  - Data prep and inspecting and unerstanding data structure
  - Estimating the model show equivalence on both models
  - Checking assumptions
  - Visualizing results

## AR1, OU, and Dynamic Models

- When to use dynamic models vs. concurrent models
- Current possibilities in R MLM for AR1 or ou models
- Current possibilites in R MLM for dynamic models
- Alternatives (DSEM in MPLUS or lavaan for X)

# Excercises in R

Short break - initiate package install before you leave

```{r eval = FALSE}

install.packages(
  c('dyadMLM', 'wbCorr', 'glmmTMB', 'DHARMa', 'dplyr', 'ggplot2')
  )

```


- Go to https://pascal-kueng.github.io/dyadMLM/workshop/ 
- or Teams channel

<!--

- DHARMa plots!?! -> as required package for workshop -> link to materials or do in tutorial? Also check AR1 etc., etc. can dharma do this if resid is set to almost zero?

- glmmTMB -> as required 
- dyadMLM -> as required (via pak, see instructions). -> maybe just require pak and we install newest version on workshop-day?
- dplyr -> required
- tibble -> required
- lavaan -> required
- ggplot2 -> required

-------------->

<!-- ILD-part: 

- see dev note on ar1 processes careful to consider degrees of freedom! relevant here?
  - TODO @codex: are degrees of freedom alreaedy important in the cross-sectional case and should we teach anything regarding those?
- see plots on: https://quantdev.ssri.psu.edu/tutorials/apa-ati-intensive-longitudinal-data-session-m-dyadic-multilevel-model

-->

<!-- Excercises part: separate files! Include on overview page?

Create exercises with data visalization, etc. Panel plots, histograms
correlations (wb corr)?

-->
