---
title: "Dyadic Data Analysis with R"
subtitle: "Part II: Applied Tutorial"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
csl: apa.csl
resources:
  - uzh-logo.svg
  - charite-logo.svg
format:
  revealjs:
    embed-resources: true
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: false
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup-pilot-data
#| include: false

library(dplyr)
library(ggplot2)

slide_table <- function(x, digits = 2, ...) {
  knitr::kable(
    x,
    format = "html",
    digits = digits,
    format.args = list(
      nsmall = digits,
      scientific = FALSE
    ),
    row.names = FALSE,
    ...
  )
}

diagram_helper_candidates <- c(
  "diagram-helpers.Rinc",
  file.path("vignettes", "diagram-helpers.Rinc"),
  file.path("..", "..", "vignettes", "diagram-helpers.Rinc")
)
diagram_helper <- diagram_helper_candidates[
  file.exists(diagram_helper_candidates)
][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(
  normalizePath(diagram_helper),
  envir = knitr::knit_global()
)

```

## Today's aims and program

::: {.fancy-table}

| Time | Focus |
|---|---|
| 11:00–12:45 | Introduction and types of dyadic models |
| 12:45–14:00 | *Lunch break*|
| 14:00–15:30 | Estimating cross-sectional dyadic models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

:::

<!-- TODO @codex: please grey out the first two points, intro and lunch break). -->

<!-- ca. 2 Minutes -->

# Software and frameworks

## SEM vs. MLM {.smaller}

- Many simple dyadic models can be represented in either framework.
- The best framework is the one that represents your conceptual model most naturally [@ledermannAnalyzingDyadicData2017].

:::: {.columns .fragment}

::: {.column width="50%"}

**SEM / DSEM is often more natural for**

- Latent dyadic constructs (e.g., a CFM)
- Item-level measurement models for reflective measures
- Linked outcomes or equations [e.g., mediation; @ledermannMediationDyadicData2009]
- Short dynamic panels in wide-format [e.g., LD-APIM; @gistelinckMultilevelAutoregressiveSmall2021]
- Intensive lagged or coupled dynamics via DSEM/RDSEM [e.g., @asparouhovDynamicStructuralEquation2018]

:::

::: {.column width="50%"}

**MLM is often more natural for**

- Parsimonious observed-variable models with modest numbers of dyads
- Nested grouping and flexible random-effect structures
- Observed-outcome growth or concurrent ILD models with flexible time trends
- Complex observed-variable regressions—interactions, polynomials, splines,
  and dyadic RSA—especially with repeated measures or additional nesting
- Specialized generalized mixed models in R, including zero-inflated, hurdle,
  and Tweedie models

:::

::::

::: {.subtle}
[@asparouhovDynamicStructuralEquation2018;
@asparouhovComparisonModelsAnalysis2020;
@driverContinuousTimeStructural2017;
@gistelinckMultilevelAutoregressiveSmall2021;
@delrosarioPracticalGuideSpecifying2025;
@schonbrodtTestingSimilarityEffects2018;
@nestlerResponseSurfaceAnalysis2019;
@brooksGlmmTMBBalancesSpeed2017]
:::

## SEM vs. MLM: Missing data {.smaller}

:::{.fragment}
- **MLM: Complete-case**
  - **Outcome missing:** Valid under MAR, ML uses remaining observations
  - **Predictor missing:** Valid under MAR, if missingness does **not depend** on the outcome after accounting for predictors
  
:::

:::{.fragment}
- **SEM with FIML and joint likelihood**
  - Valid under MAR: Models predictors and outcomes jointly

:::

:::{.fragment}
- **MI: Imputation**
  - Handles missing outcomes, predictors, and auxiliary information
  - Must match the analysis and preserve person, dyad, role, and time structure

:::

::: {.fragment}
Include important predictors of missingness and missing values. FIML and MI
cover more MAR situations than row omission; none automatically solves MNAR.

:::

[@ledermannAnalyzingDyadicData2017; @hughesAccountingMissingData2019; @grundMultipleImputationMissing2018]{.subtle}

<!-- ca. 2.5 Minutes -->

## Why R—and where other software helps {.smaller}

:::: {.columns}

::: {.column width="46%"}

### Why we teach R

:::{.fragment}
- **Free to use:** no license fees or institutional access barriers
- **Open source:** statistical implementations can be inspected, audited, and adapted
- **Reproducible:** workflows can be rerun, reviewed, shared, and version-controlled
- **Extensible:** specialized methods integrate with data, graphics, and reporting
:::

:::

::: {.column width="54%"}

### Current trade-offs

:::{.fragment}

- **SEM/DSEM:** R's capabilities are spread across specialized packages
  - `lavaan` handles explicitly specified, fixed-wave lagged SEMs (e.g.,
    CLPM/RI-CLPM), but has no native DSEM or residual AR(1) structure
  - Version 0.7 adds restricted two-level random slopes; `mlts` implements
    Bayesian multilevel time-series models, while Mplus integrates DSEM/RDSEM
- **MLM/ILD:** `glmmTMB` and `brms` both model serial dependence
  - Neither directly offers a separable partner-by-time residual structure such
    as $\mathrm{UN}\otimes\mathrm{AR}(1)$; SAS `PROC MIXED` and SPSS `MIXED` do
    for Gaussian repeated-measures models

:::

:::

::::

::: {.subtle}
[@lavaanMultilevelSEM;
@mulderThreeExtensions2021;
@blavaanTwoLevelSEM;
@koslowskiMlts2025;
@asparouhovDynamicStructuralEquation2018;
@asparouhovComparisonModelsAnalysis2020;
@kristensenCovarianceStructuresGlmmTMB2023;
@buerknerAutocorrelationStructuresBrms;
@sasInstituteMixedProcedure;
@ibmCovarianceStructureList]
:::

<!-- ca. 2.5 Minutes -->

# Estimating cross-sectional dyadic models in R

## One possible workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the Data**
2. **Check data structure**
3. **Describe and visualize measures and simple relationships**
4. **Estimate the models**
5. **Verify model adequacy, (distinguishability, if applicable), and robustness**
6. **Visualize, interpret, and report results**

<!-- konsistent wie hier -->
::: {.subtle}
Synthesized from @kenny2006dyadic, @ledermannToolboxProgramsRestructure2015, and @stasGivingDyadicData2018.
:::

:::{.notes}

- In this example we will use the same dataset that we already used in Part I

:::

## One possible workflow

1. [**Define dyads, distinguishability, and research questions**]{.text-highlight}
2. **Prepare the Data**
2. **Check data structure**
3. **Describe and visualize measures and simple relationships**
4. **Estimate the models**
5. **Verify model adequacy, (distinguishability, if applicable), and robustness**
6. **Visualize, interpret, and report results**

<!-- konsistent wie hier -->
::: {.subtle}
Synthesized from @kenny2006dyadic, @ledermannToolboxProgramsRestructure2015, and @stasGivingDyadicData2018.
:::


## The Dataset

**Same as before: **

- 36 physically inactive romantic couples
- Both partners intended to become more physically active
- Smartphone-based planning interventions and JITAIs
- Daily diaries over 55 days

**Aggregated for cross-sectional analyses**


## Research Question {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ:]{.rq-label} [In romantic couples, is a person's **self-efficacy** linked to physical activity in both partners?]{.rq-text}
:::
:::

## Measurement - predictor

### Self-efficacy

> **Rough English translation of the original German item**<br>
> “I am confident that I can be physically active tomorrow, even if it becomes
> difficult.”
>
> [Adapted from @sniehottaBridgingIntentionBehaviour2005]{.subtle}

0 = *not at all true today* <br>
5 = *completely true today*<br>

<!-- ca. 1 Minute -->

## Measurement - outcome

### Moderate to vigorous physical activity (MVPA)

> **Rough English translation of the original German item**<br>
> “How many minutes did you spend engaging in moderate-to-vigorous physical
> activity today?”
>
> [Adapted from @amireaultGodinShephardQuestionnaire2015]{.subtle}

- **Alone:** _____ minutes
- **Together with your partner:** _____ minutes

**Total MVPA = solo MVPA + joint MVPA.**

<!-- ca. 1 Minute -->

# Distinguishable dyads
Distinguishing by gender

:::{.notes}

Almost done with first step, we still have to decide on distinguishability. 
In our data, probably makes sense to distinguish by gender. Not always!

-> Gender, but TEST!

:::

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. [**Prepare the data and verify the dyadic structure**]{.text-highlight}
3. **Describe and visualize measures and observed relationships**
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Load and inspect the raw data

```{r}
library(dplyr) # for easier data-wrangling, exports functions like `select()`

df <- readRDS("dyadic-person-means.rds") |>
  dplyr::select(
    couple_id, person_id, gender, efficacy, total_mvpa
  )

# usually use head(df), slide_table is a custom function for html tables on these slides
slide_table(df[1:4,]) 
```

:::{.subtle}

<br>

This dataset is stored as an R object. Other packages provide functions to import various formats, such as: `readxl::read_excel()`, `utils::read.csv()`, `haven::read_sav()`, `haven::read_sas()`

:::

:::{.notes}

Manual inspection at each step is important! 

Example that Lilly told me: weird associations, until discovered that 
missing values were coded as 99, in a software that didn't recognize this 
as a missing value.

:::

## Create the required format and validate the data

**For the distinguishable APIM we need to add:**

- `efficacy_actor` and `efficacy_partner` columns
- Dummy coded numeric variables for the two roles

:::{.fragment}
**Validate things like: **

- no dyad with more than two members,
- each dyad-member combination occurs only once (per occasion),
- gender labels are present and consistent
- you only have the type of dyad you expect
:::

## Create the required format and validate the data

We can use the `dyadMLM` package to help with data preparation and validation.

```{r}
#| output-location: slide

library(dyadMLM)

df_apim <- dyadMLM::prepare_dyad_data( # exported function from dyadMLM
  data = df,
  dyad = couple_id,
  member = person_id,
  role = gender,
  predictors = efficacy, 
  model_types = 'apim'
)


print(df_apim)

```


## Create the required format and validate the data
<!-- TODO: link to an appendix, just like we did in the dyad-day.qmd file, with code-snippets of how to manually achieve the
is_male, is_female, and _actor and _partner variables. -->


Looking at the columns immediately relevant to us:

```{r}

df_apim[1:6,] |>
  dplyr::select(
    couple_id, person_id, gender, .is_female, .is_male, efficacy,
    .efficacy_actor, .efficacy_partner, total_mvpa
  ) |>
  slide_table()

```

<!-- TODO: color code partner values? -->

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and verify the dyadic structure**
3. [**Describe and visualize measures and observed relationships**]{.text-highlight}
4. **Estimate the models**
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Create a wide view for role-specific exploration

The models will **NOT** use this dataset

```{r}
df_wide <- df_apim |>
  dplyr::select(couple_id, gender, efficacy, total_mvpa) |>
  tidyr::pivot_wider(
    names_from = gender,
    values_from = c(efficacy, total_mvpa)
  )

slide_table(head(df_wide))
```


## Describe focal variables per role

```{r}
library(report)

df_wide |>
  dplyr::select(-couple_id) |>
  report::report_table() |>
  slide_table()
```

## Visualize focal variable by partner role

```{r}

library(ggplot2)
library(see)

plot_dyadic_histograms <- function(data, group, var) {
  var_label <- rlang::as_label(rlang::enquo(var))

  ggplot(
    data,
    aes(x = {{ var }}, y = after_stat(density))
  ) +
    geom_histogram(
      bins = 15, boundary = 0,
      fill = "steelblue", color = "white", alpha = 0.75
    ) +
    geom_density(
      bounds = c(0, Inf), color = "navy", linewidth = 1.1
    ) +
    facet_wrap(vars({{ group }}), nrow = 1) +
    labs(x = var_label, y = "Density") +
    see::theme_modern(base_size = 16)
}

```

## Visualize focal variable by partner role - Efficacy

```{r}
#| dev: svglite

plot_dyadic_histograms(df_apim, gender, efficacy)
```


## Visualize focal variable by partner role - MVPA

```{r}
#| dev: svglite

plot_dyadic_histograms(df_apim, gender, total_mvpa)
```

## Observed dyadic correlations

The wide data ensure that each couple contributes once.

```{r}
library(correlation)
library(insight)

df_wide |>
  correlation::correlation(
    select = c("efficacy_female", "total_mvpa_female"),
    select2 = c("efficacy_male", "total_mvpa_male")
  ) |> 
  summary() |>
  insight::print_md(footer = "")
```


## Visualize partner similarity

```{r}
plot_partner_similarity <- function(data, x, y, id, title, n_mahalanobis = 3) {
  x_name <- rlang::as_name(rlang::enquo(x))
  y_name <- rlang::as_name(rlang::enquo(y))
  xy <- cbind(data[[x_name]], data[[y_name]])
  labels <- data |>
    dplyr::mutate(
      .distance = mahalanobis(xy, colMeans(xy), cov(xy))
    ) |>
    dplyr::slice_max(.distance, n = n_mahalanobis)

  ggplot(data, aes(x = {{ x }}, y = {{ y }})) +
    geom_point(color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "navy") +
    geom_text(
      data = labels,
      aes(label = {{ id }}),
      vjust = -0.6, check_overlap = TRUE, size = 4
    ) +
    labs(title = title, x = "Female", y = "Male") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
    see::theme_modern(base_size = 16)
}
```

## Visualize partner similarity

```{r}
#| output-location: slide
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.5
#| out-width: 100%
#| fig-cap: "Labels: four largest Mahalanobis distances in each plot."

n_mahalanobis <- 4

see::plots(
  plot_partner_similarity(
    df_wide, efficacy_female, efficacy_male, couple_id, "Efficacy",
    n_mahalanobis = n_mahalanobis
  ),
  plot_partner_similarity(
    df_wide, total_mvpa_female, total_mvpa_male, couple_id, "Daily MVPA",
    n_mahalanobis = n_mahalanobis
  )
)
```

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the data and verify the dyadic structure**
3. **Describe and visualize measures and observed relationships**
4. [**Estimate the models**]{.text-highlight}
5. **Verify model adequacy, distinguishability, and robustness**
6. **Visualize, interpret, and report results**

## Distinguishable APIM

```{r}
#| label: applied-conceptual-distinguishable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with role-specific actor and partner paths and correlated outcome residuals."

draw_apim_diagram("distinguishable")
```

:::{.notes}

This is the same conceptual model introduced in Part I.

[Sources]

- @kenny2006dyadic

:::

## One response column, two member rows

**Both members' outcomes occupy one response column:** each member contributes
one row.

```{r}
#| label: applied-distinguishable-apim-member-wise
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Member-wise representation of a distinguishable APIM fitted in long format. Each member's actor and partner predictors predict that member's outcome, while the two outcome residuals covary within the dyad."

draw_apim_member_diagram("distinguishable")
```

:::{.notes}

- For the female outcome row, female efficacy is the actor predictor and male
  efficacy is the partner predictor.
- These roles reverse for the male outcome row.
- The two rows are connected through the within-couple residual covariance.

[Sources]

- @kenny2006dyadic

:::

## The APIM starts with two member-specific equations

For each dyad $i$:

$$
  \begin{gathered}
  Y_{\mathrm{male},i}
  = b_{0,\mathrm{M}}
  + \color{#1769AA}{a_{\mathrm{M}} \times X_{\mathrm{male},i}}
  + \color{#7B3FA1}{p_{\mathrm{M}} \times X_{\mathrm{female},i}}
  + \epsilon_{\mathrm{male},i}
  \\
  Y_{\mathrm{female},i}
  = b_{0,\mathrm{F}}
  + \color{#1769AA}{a_{\mathrm{F}} \times X_{\mathrm{female},i}}
  + \color{#7B3FA1}{p_{\mathrm{F}} \times X_{\mathrm{male},i}}
  + \epsilon_{\mathrm{female},i}
  \end{gathered}
$$

[**Actor effects:** own predictor]{style="color: #1769AA"} ·
[**Partner effects:** the other member's predictor]{style="color: #7B3FA1"}

In long format, each equation applies to one member row of the dyad.

:::{.notes}

The signs of the actor and partner associations come from the estimated
coefficients. They are not built into the model equation.

[Sources]

- @kenny2006dyadic

:::

## Role indicators combine the equations

Let $D_{\mathrm{M},ji}$ and $D_{\mathrm{F},ji}$ indicate the role of row
$j$. Let $X^{(\mathrm{A})}_{ji}$ and $X^{(\mathrm{P})}_{ji}$ contain the
actor's and partner's values:

$$
\begin{aligned}
Y_{ji} ={}&
\underbrace{
  b_{0,\mathrm{M}} \times D_{\mathrm{M},ji}
  + b_{0,\mathrm{F}} \times D_{\mathrm{F},ji}
}_{\text{role-specific intercepts}}
\\[0.35em]
&+
\color{#1769AA}{
\underbrace{
  a_{\mathrm{M}} \times D_{\mathrm{M},ji} \times X^{(\mathrm{A})}_{ji}
  + a_{\mathrm{F}} \times D_{\mathrm{F},ji} \times X^{(\mathrm{A})}_{ji}
}_{\text{role-specific actor effects}}
}
\\[0.35em]
&+
\color{#7B3FA1}{
\underbrace{
  p_{\mathrm{M}} \times D_{\mathrm{M},ji} \times X^{(\mathrm{P})}_{ji}
  + p_{\mathrm{F}} \times D_{\mathrm{F},ji} \times X^{(\mathrm{P})}_{ji}
}_{\text{role-specific partner effects}}
}
+ \epsilon_{ji}.
\end{aligned}
$$

:::{.text-smaller}

Male row: $D_\mathrm{M}=1,\ D_\mathrm{F}=0$. Female row: the reverse.

:::

:::{.notes}

The role indicators correspond to `.is_male` and `.is_female`. The actor and
partner values correspond to `.efficacy_actor` and `.efficacy_partner`.

[Sources]

- @kenny2006dyadic

:::

## Residuals remain correlated within dyads

The two member rows from the same dyad retain a joint residual structure:

$$
\operatorname{Cov}
\begin{pmatrix}
\epsilon_{\mathrm{F}i} \\
\epsilon_{\mathrm{M}i}
\end{pmatrix}
= \boldsymbol{\Sigma}_{\epsilon}
=
\begin{bmatrix}
\sigma_{\epsilon_\mathrm{F}}^{2}
&
\rho_{\epsilon_\mathrm{F}\epsilon_\mathrm{M}}
\sigma_{\epsilon_\mathrm{F}}\sigma_{\epsilon_\mathrm{M}} \\
\rho_{\epsilon_\mathrm{F}\epsilon_\mathrm{M}}
\sigma_{\epsilon_\mathrm{F}}\sigma_{\epsilon_\mathrm{M}}
&
\sigma_{\epsilon_\mathrm{M}}^{2}
\end{bmatrix}
$$

In `glmmTMB`, this is represented by:

```r
us(0 + .is_female + .is_male | couple_id),
dispformula = ~ 0
```

:::{.notes}

The unstructured block estimates separate female and male residual variances
and their covariance. Setting `dispformula = ~ 0` avoids adding a second
observation-level residual variance.

[Sources]

- @kenny2006dyadic

:::


<!-------------------------    --------------------------------------------->
  

## Exchangeable APIM

- Row-Level representation
- Formula
- residual matrix?
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships
  - Estimating the model
  - Checking assumptions
  - Visualizing results


## Testing distinguishability

### Wald test for fixed effects equality

### Model comparison

## DIM (potentially omit)

- Row-Level representation
- Formula
- residual matrix?
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships
  - Estimating the model
  - Checking assumptions
  - Visualizing results

## Excursion: non-Gaussian generalized models

## Q&A and discussion

# Intensive longitudinal extensions

## ILD extends rather than replaces the workflow {.smaller}

:::: {.columns}

::: {.column width="43%"}

### The same core workflow

::: {.subtle}

1. Define dyads, roles, distinguishability, and research questions
2. Check data structure, completeness, and missing-data handling
3. Describe measures and dyadic relationships
4. Specify, document, and estimate the APIM
5. Verify estimation; check residuals and influential dyads
6. Interpret, visualize, and report the results

:::

:::

::: {.column width="57%"}

### ILD additionally requires

- Check compliance, matched occasions, and actual time gaps
  [@revolPreprocessingExperienceSampling2024]
- Plot paired trajectories and inspect raw temporal and cross-partner patterns
  [@bolgerIntensiveLongitudinalMethods2013;
  @revolPreprocessingExperienceSampling2024]
- Separate within- and between-person variation
  - ICCs, level-specific correlations, and multilevel reliability where
    applicable [@bolgerIntensiveLongitudinalMethods2013;
    @hamakerCuriousCaseCrossSectional2024;
    @geldhofReliabilityEstimationMultilevel2014]
- Choose the temporal specification
  - Time scale and lag, trends/stationarity, concurrent versus lagged effects,
    and discrete versus continuous time
    [@driverContinuousTimeStructural2017]
- Check residual temporal and cross-partner dependence

:::

::::

::: {.subtle}
Workflow synthesized from @bolgerIntensiveLongitudinalMethods2013,
@revolPreprocessingExperienceSampling2024, @mcneishPrimerTwoLevel2020, and
@delrosarioPracticalGuideSpecifying2025.
:::

:::{.notes}

For ILD-specific power analysis, see @lafitPowerLAPIMApplication2022. Perry et
al. belongs with interpretation of fitted dynamics: their change-score APIM and
vector-field diagrams show the joint implications of actor and partner
coefficients [@perryGraphicMethodsInterpreting2017].

We will use `wbCorr` for pooled within- and between-person correlations
[@kuengWbCorrBivariateWithin2026]; @hamakerCuriousCaseCrossSectional2024
provides the methodological rationale.

`omegaSEM()` is provided by `multilevelTools`, not `semTools`. It implements
level-specific internal-consistency coefficients based on a two-level
one-factor model [@wileyMultilevelTools2025;
@geldhofReliabilityEstimationMultilevel2014]. These coefficients are not
temporal stability or the precision of observed person means; for the latter,
see @laiCompositeReliabilityMultilevel2021.

:::

## Distinguishable APIM

- Formula
- Random effects and residual covariance explained
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships on both levels (wbCorr)
  - Estimating the model
  - Checking assumptions
  - Visualizing results
  
## Exchangeable APIM

- Formula
- Random effects and residual covariance explained
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships on both levels (wbCorr)
  - Estimating the model
  - Checking assumptions
  - Visualizing results

## DIM (potentially omit)

- Code
  - Data prep and inspecting and unerstanding data structure
  - Estimating the model show equivalence on both models
  - Checking assumptions
  - Visualizing results

## AR1, OU, and Dynamic Models

- When to use dynamic models vs. concurrent models
- Current possibilities in R MLM for AR1 or ou models
- Current possibilites in R MLM for dynamic models
- Alternatives (DSEM in MPLUS or lavaan for X)

## Power analysis is the clearest reporting gap

```{r}
#| label: plot-longitudinal-apim-reporting-gaps
#| echo: false
#| fig-width: 10.5
#| fig-height: 4.5
#| out-width: 92%
#| fig-align: center
#| fig-alt: "Bar chart of reporting practices in 73 longitudinal APIM studies: 17.8 percent reported a power analysis, 64.4 percent documented model constraints, and 75.3 percent reported a missing-data strategy."

reporting_gaps <- tibble(
  item = factor(
    c("Power analysis", "Model constraints", "Missing-data strategy"),
    levels = c("Missing-data strategy", "Model constraints", "Power analysis")
  ),
  reported = c(17.8, 64.4, 75.3)
)

ggplot(reporting_gaps, aes(item, reported, fill = item == "Power analysis")) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = paste0(reported, "%")),
    hjust = -0.15, size = 7
  ) +
  coord_flip(clip = "off") +
  scale_fill_manual(values = c("TRUE" = "#0028a5", "FALSE" = "#9aa7bd")) +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 20),
    labels = paste0(seq(0, 100, 20), "%"),
    expand = c(0, 0)
  ) +
  labs(
    x = NULL,
    y = "Studies reporting the item"
  ) +
  theme_minimal(base_size = 21) +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(5.5, 20, 5.5, 5.5)
  )
```

Systematic review of 73 longitudinal APIM studies published from 2008–2025
[@mohammadzadehYazdBeyondIndividual2026].

:::{.notes}

- The chart shows reporting rates because they are easier to read as adherence
  to good practice. The paper directly reports 17.8% for power analyses; 64.4%
  and 75.3% are the complements of the reported omission rates.
- This was a methodological systematic review, not a meta-analysis of effect
  sizes. Its sample was restricted to longitudinal APIM studies with at least
  70 dyads and follow-up longer than one year.
- The review therefore supports “reporting gaps remain common,” not the broader
  claim that all dyadic analyses are reported inadequately.

[Sources]

- Mohammadzadeh Yazd et al. (2026), abstract:
  <https://doi.org/10.1177/02654075261443593>

:::

## Make every dyadic analysis decision visible

1. **Preregister** actor and partner hypotheses and key analysis decisions.
2. **Plan and report power** using dyad-specific tools.
3. **Explain missing data:** describe missingness and the handling strategy.
4. **Specify the full model:** roles and distinguishability, actor and partner
   paths, constraints, and covariance structure.

The review also recommends multi-method measurement and more diverse
populations

[@mohammadzadehYazdBeyondIndividual2026]{.subtle}


# Excercises in R

Short break - initiate package install before you leave

```{r eval = FALSE}

install.packages(
  c("dyadMLM", "wbCorr", "glmmTMB", "DHARMa", "dplyr", "ggplot2", "tibble"),
  repos = c(
    "https://pascal-kueng.r-universe.dev",
    "https://cloud.r-project.org"
  )
)

```


- Go to https://pascal-kueng.github.io/dyadMLM/workshop/ 
- or Teams channel

<!--

- DHARMa plots!?! -> as required package for workshop -> link to materials or do in tutorial? Also check AR1 etc., etc. can dharma do this if resid is set to almost zero?

- glmmTMB -> as required 
- dyadMLM -> as required (via R-universe, see instructions).
- dplyr -> required
- tibble -> required
- lavaan -> required
- ggplot2 -> required

-------------->

<!-- ILD-part: 

- see dev note on ar1 processes careful to consider degrees of freedom! relevant here?
  - TODO @codex: are degrees of freedom alreaedy important in the cross-sectional case and should we teach anything regarding those?
- see plots on: https://quantdev.ssri.psu.edu/tutorials/apa-ati-intensive-longitudinal-data-session-m-dyadic-multilevel-model

-->

<!-- Excercises part: separate files! Include on overview page?

Create exercises with data visalization, etc. Panel plots, histograms
correlations (wb corr)?

-->

## References
