---
title: "Dyadic Data Analysis with R"
subtitle: "Part II: Applied Tutorial"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
csl: apa.csl
resources:
  - uzh-logo.svg
  - charite-logo.svg
format:
  revealjs:
    embed-resources: true
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: false
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup-pilot-data
#| include: false

library(dplyr)
library(ggplot2)

diagram_helper_candidates <- c(
  "diagram-helpers.Rinc",
  file.path("vignettes", "diagram-helpers.Rinc"),
  file.path("..", "..", "vignettes", "diagram-helpers.Rinc")
)
diagram_helper <- diagram_helper_candidates[
  file.exists(diagram_helper_candidates)
][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(
  normalizePath(diagram_helper),
  envir = knitr::knit_global()
)

```

## Today's aims and program

| Time | Focus |
|---|---|
| 11:00–12:45 | Introduction and types of dyadic models |
| 12:45–14:00 | *Lunch break*|
| 14:00–15:30 | Estimating cross-sectional dyadic models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

<!-- TODO @codex: please grey out the first two points, intro and lunch break). -->

<!-- ca. 2 Minutes -->

# Software and frameworks

## SEM vs. MLM {.smaller}

- Many simple dyadic models can be represented in either framework.
- The best framework is the one that represents your conceptual model most naturally [@ledermannAnalyzingDyadicData2017].

:::: {.columns .fragment}

::: {.column width="50%"}

**SEM / DSEM is often more natural for**

- Latent dyadic constructs (e.g., a CFM)
- Item-level measurement models for reflective measures
- Linked outcomes or equations [e.g., mediation; @ledermannMediationDyadicData2009]
- Short dynamic panels in wide-format [e.g., LD-APIM; @gistelinckMultilevelAutoregressiveSmall2021]
- Intensive lagged or coupled dynamics via DSEM/RDSEM [e.g., @asparouhovDynamicStructuralEquation2018]

:::

::: {.column width="50%"}

**MLM is often more natural for**

- Parsimonious observed-variable models with modest numbers of dyads
- Nested grouping and flexible random-effect structures
- Observed-outcome growth or concurrent ILD models with flexible time trends
- Complex observed-variable regressions—interactions, polynomials, splines,
  and dyadic RSA—especially with repeated measures or additional nesting
- Specialized generalized mixed models in R, including zero-inflated, hurdle,
  and Tweedie models

:::

::::

::: {.subtle}
[@asparouhovDynamicStructuralEquation2018;
@asparouhovComparisonModelsAnalysis2020;
@driverContinuousTimeStructural2017;
@gistelinckMultilevelAutoregressiveSmall2021;
@delrosarioPracticalGuideSpecifying2025;
@schonbrodtTestingSimilarityEffects2018;
@nestlerResponseSurfaceAnalysis2019;
@brooksGlmmTMBBalancesSpeed2017]
:::

## SEM vs. MLM: Missing data {.smaller}

- **MLM: Complete-case**
  - **Outcome missing:** Valid under MAR, ML uses remaining observations
  - **Predictor missing:** Valid under MAR, if missingness does **not depend**
    on the outcome after accounting for predictors
- **SEM with FIML and joint likelihood**
  - Valid under MAR: Models predictors and outcomes jointly
- **MI: Imputation**
  - Handles missing outcomes, predictors, and auxiliary information
  - Must match the analysis and preserve person, dyad, role, and time structure

::: {.subtle}
Include important predictors of missingness and missing values. FIML and MI
cover more MAR situations than row omission; none automatically solves MNAR.
[@ledermannAnalyzingDyadicData2017;
@hughesAccountingMissingData2019; @grundMultipleImputationMissing2018]
:::

<!-- ca. 2.5 Minutes -->

## Why R—and where other software helps {.smaller}

:::: {.columns}

::: {.column width="46%"}

### Why we teach R

- **Free to use:** no license fees or institutional access barriers
- **Open source:** statistical implementations can be inspected, audited, and adapted
- **Reproducible:** workflows can be rerun, reviewed, shared, and version-controlled
- **Extensible:** specialized methods integrate with data, graphics, and reporting

:::

::: {.column width="54%"}

### Current trade-offs

- **SEM/DSEM:** R's capabilities are spread across specialized packages
  - `lavaan` handles explicitly specified, fixed-wave lagged SEMs (e.g.,
    CLPM/RI-CLPM), but has no native DSEM or residual AR(1) structure
  - Version 0.7 adds restricted two-level random slopes; `mlts` implements
    Bayesian multilevel time-series models, while Mplus integrates DSEM/RDSEM
- **MLM/ILD:** `glmmTMB` and `brms` both model serial dependence
  - Neither directly offers a separable partner-by-time residual structure such
    as $\mathrm{UN}\otimes\mathrm{AR}(1)$; SAS `PROC MIXED` and SPSS `MIXED` do
    for Gaussian repeated-measures models

:::

::::

::: {.subtle}
[@lavaanMultilevelSEM;
@mulderThreeExtensions2021;
@blavaanTwoLevelSEM;
@koslowskiMlts2025;
@asparouhovDynamicStructuralEquation2018;
@asparouhovComparisonModelsAnalysis2020;
@kristensenCovarianceStructuresGlmmTMB2023;
@buerknerAutocorrelationStructuresBrms;
@sasInstituteMixedProcedure;
@ibmCovarianceStructureList]
:::

<!-- ca. 2.5 Minutes -->

# Estimating cross-sectional dyadic models in R

## One possible workflow

1. **Define dyads, distinguishability, and research questions**
2. **Prepare the Data**
2. **Check data structure**
3. **Describe measures and simple relationships**
4. **Estimate the models**
5. **Verify estimation, check assumptions, identify influential dyads**
6. **Visualize, interpret, and report results**

::: {.subtle}
Synthesized from @kenny2006dyadic, @ledermannToolboxProgramsRestructure2015, and @stasGivingDyadicData2018.
:::

:::{.notes}

- In this example we will use the same dataset that we already used in Part I

:::

## The Dataset

**Same as before: **

- 36 physically inactive romantic couples
- Both partners intended to become more physically active
- Smartphone-based planning interventions and JITAIs
- Daily diaries over 55 days

**Aggregated for cross-sectional analyses**


## Research Question {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ:]{.rq-label} [In romantic couples, is a person's **self-efficacy** linked to physical activity in both partners?]{.rq-text}
:::
:::

## Measurement - predictor

### Self-efficacy

> **Rough English translation of the original German item**<br>
> “I am confident that I can be physically active tomorrow, even if it becomes
> difficult.”
>
> [Adapted from @sniehottaBridgingIntentionBehaviour2005]{.subtle}

0 = *not at all true today* <br>
5 = *completely true today*<br>

<!-- ca. 1 Minute -->

## Measurement - outcome

### Moderate to vigorous physical activity (MVPA)

> **Rough English translation of the original German item**<br>
> “How many minutes did you spend engaging in moderate-to-vigorous physical
> activity today?”
>
> [Adapted from @amireaultGodinShephardQuestionnaire2015]{.subtle}

- **Alone:** _____ minutes
- **Together with your partner:** _____ minutes

**Total MVPA = solo MVPA + joint MVPA.**

<!-- ca. 1 Minute -->

## Workflow

1. [**Define dyads, distinguishability, and research questions**]{.text-highlight}
2. **Prepare the Data**
2. **Check data structure**
3. **Describe measures and simple relationships**
4. **Estimate the models**
5. **Verify estimation, check assumptions, identify influential dyads**
6. **Visualize, interpret, and report results**

:::{.notes}

We have defined a research question, and for this analysis we have
decided to distinguish by gender... although this is often just a standard
and should be considered carefully too. 

:::

## Workflow

1. **Define dyads, distinguishability, and research questions**
2. [**Prepare the Data**]{.text-highlight}
2. **Check data structure**
3. **Describe measures and simple relationships**
4. **Estimate the models**
5. **Verify estimation, check assumptions, identify influential dyads**
6. **Visualize, interpret, and report results**

## Data preparation in R

The dplyr package enables better readable syntax

```{r}
library(dplyr)



```


## Distinguishable APIM

- Row-Level representation
- Formula
- residual matrix?
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships
  - Estimating the model
  - Checking assumptions
  - Visualizing results
  

## Exchangeable APIM

- Row-Level representation
- Formula
- residual matrix?
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships
  - Estimating the model
  - Checking assumptions
  - Visualizing results


## Testing distinguishability

### Wald test for fixed effects equality

### Model comparison

## DIM (potentially omit)

- Row-Level representation
- Formula
- residual matrix?
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships
  - Estimating the model
  - Checking assumptions
  - Visualizing results

## Excursion: non-Gaussian generalized models

## Q&A and discussion

# Intensive longitudinal extensions

## ILD extends rather than replaces the workflow {.smaller}

:::: {.columns}

::: {.column width="43%"}

::: {.subtle}

### The same core workflow

1. Define dyads, roles, distinguishability, and research questions
2. Check data structure, completeness, and missing-data handling
3. Describe measures and dyadic relationships
4. Specify, document, and estimate the APIM
5. Verify estimation; check residuals and influential dyads
6. Interpret, visualize, and report the results

:::

:::

::: {.column width="57%"}

### ILD additionally requires

- Check compliance, matched occasions, and actual time gaps
  [@revolPreprocessingExperienceSampling2024]
- Plot paired trajectories and inspect raw temporal and cross-partner patterns
  [@bolgerIntensiveLongitudinalMethods2013;
  @revolPreprocessingExperienceSampling2024]
- Separate within- and between-person variation
  - ICCs, level-specific correlations, and multilevel reliability where
    applicable [@bolgerIntensiveLongitudinalMethods2013;
    @hamakerCuriousCaseCrossSectional2024;
    @geldhofReliabilityEstimationMultilevel2014]
- Choose the temporal specification
  - Time scale and lag, trends/stationarity, concurrent versus lagged effects,
    and discrete versus continuous time
    [@driverContinuousTimeStructural2017]
- Check residual temporal and cross-partner dependence

:::

::::

::: {.subtle}
Workflow synthesized from @bolgerIntensiveLongitudinalMethods2013,
@revolPreprocessingExperienceSampling2024, @mcneishPrimerTwoLevel2020, and
@delrosarioPracticalGuideSpecifying2025.
:::

:::{.notes}

For ILD-specific power analysis, see @lafitPowerLAPIMApplication2022. Perry et
al. belongs with interpretation of fitted dynamics: their change-score APIM and
vector-field diagrams show the joint implications of actor and partner
coefficients [@perryGraphicMethodsInterpreting2017].

We will use `wbCorr` for pooled within- and between-person correlations
[@kuengWbCorrBivariateWithin2026]; @hamakerCuriousCaseCrossSectional2024
provides the methodological rationale.

`omegaSEM()` is provided by `multilevelTools`, not `semTools`. It implements
level-specific internal-consistency coefficients based on a two-level
one-factor model [@wileyMultilevelTools2025;
@geldhofReliabilityEstimationMultilevel2014]. These coefficients are not
temporal stability or the precision of observed person means; for the latter,
see @laiCompositeReliabilityMultilevel2021.

:::

## Distinguishable APIM

- Formula
- Random effects and residual covariance explained
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships on both levels (wbCorr)
  - Estimating the model
  - Checking assumptions
  - Visualizing results
  
## Exchangeable APIM

- Formula
- Random effects and residual covariance explained
- Code
  - Data prep and inspecting and understanding data structure
  - Visualizing variables and bivariate relationships on both levels (wbCorr)
  - Estimating the model
  - Checking assumptions
  - Visualizing results

## DIM (potentially omit)

- Code
  - Data prep and inspecting and unerstanding data structure
  - Estimating the model show equivalence on both models
  - Checking assumptions
  - Visualizing results

## AR1, OU, and Dynamic Models

- When to use dynamic models vs. concurrent models
- Current possibilities in R MLM for AR1 or ou models
- Current possibilites in R MLM for dynamic models
- Alternatives (DSEM in MPLUS or lavaan for X)

# Excercises in R

Short break - initiate package install before you leave

```{r eval = FALSE}

install.packages(
  c('dyadMLM', 'wbCorr', 'glmmTMB', 'DHARMa', 'dplyr', 'ggplot2')
  )

```


- Go to https://pascal-kueng.github.io/dyadMLM/workshop/ 
- or Teams channel

<!--

- DHARMa plots!?! -> as required package for workshop -> link to materials or do in tutorial? Also check AR1 etc., etc. can dharma do this if resid is set to almost zero?

- glmmTMB -> as required 
- dyadMLM -> as required (via pak, see instructions). -> maybe just require pak and we install newest version on workshop-day?
- dplyr -> required
- tibble -> required
- lavaan -> required
- ggplot2 -> required

-------------->

<!-- ILD-part: 

- see dev note on ar1 processes careful to consider degrees of freedom! relevant here?
  - TODO @codex: are degrees of freedom alreaedy important in the cross-sectional case and should we teach anything regarding those?
- see plots on: https://quantdev.ssri.psu.edu/tutorials/apa-ati-intensive-longitudinal-data-session-m-dyadic-multilevel-model

-->

<!-- Excercises part: separate files! Include on overview page?

Create exercises with data visalization, etc. Panel plots, histograms
correlations (wb corr)?

-->
