---
title: "Dyad-Day"
subtitle: "Dyadic Data Analysis with R"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
resources:
  - uzh-logo.svg
  - charite-logo.svg
format:
  revealjs:
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: true
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup-pilot-data
#| include: false

library(dplyr)
library(ggplot2)
library(tidyr)

data_file <- if (file.exists("long.rds")) "long.rds" else "dev/workshop/long.rds"

pilot <- readRDS(data_file) |>
  select(
    coupleID, userID, ss_pa, ss_pa_min_solo, ss_pa_min_collab,
    ss_emo_pleasure, sc_actionplan
  ) |>
  transmute(
    couple_id = coupleID,
    person_id = userID,
    emotional_support = ss_emo_pleasure,
    collaborative_planning = sc_actionplan,
    mvpa_solo = if_else(ss_pa == 0, 0, ss_pa_min_solo),
    mvpa_joint = if_else(ss_pa == 0, 0, ss_pa_min_collab),
    mvpa_total = mvpa_solo + mvpa_joint
  )

pilot_aggregated <- pilot |>
  group_by(couple_id, person_id) |>
  summarise(
    across(everything(), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )

distribution_plot <- function(data) {
  labels <- c(
    emotional_support = "Emotional support provided",
    collaborative_planning = "Collaborative planning (daily mean)",
    mvpa_solo = "Mean solo MVPA (minutes/day)",
    mvpa_joint = "Mean joint MVPA (minutes/day)",
    mvpa_total = "Mean total MVPA (minutes/day)"
  )

  data |>
    pivot_longer(-c(couple_id, person_id), names_to = "variable") |>
    mutate(variable = factor(variable, levels = names(labels), labels = labels)) |>
    ggplot(aes(value)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins = 16,
      fill = "#B88900",
      color = "white",
      linewidth = 0.25
    ) +
    geom_density(color = "#D94335", linewidth = 0.8, adjust = 1.1) +
    facet_wrap(vars(variable), scales = "free", ncol = 3) +
    labs(x = NULL, y = "Density") +
    theme_minimal(base_size = 15) +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold"),
      axis.text = element_text(color = "#222222")
    )
}
```

## Introduction

- Turu & Pascal
- Tell us about you and your dyadic data!
  - Name, Affiliation & Career stage
  - Your dyadic data and RQ
  
<!-- ca. 10 Minutes -->

## Today's Aims and Program

| Time | Focus |
|---|---|
| 11:00–12:45 | Introduction and different types of dyadic models |
| 12:45–14:00 | *Lunch break* |
| 14:00–15:30 | Estimating cross-sectional models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

<!-- ca. 5 Minutes -->

## A dyadic dataset

- 38 physically inactive romantic couples (oposite sex)
- both partners wanted to increase their physical activity jointly
- JITAIs and planning interventions
- Tracked with daily diaries over 55 days


## Pilot data: aggregated variables {.smaller}

```{r}
#| label: pilot-aggregated-distributions-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5.7
#| fig-alt: "Histograms of provided emotional support, collaborative planning, and person-level mean MVPA."

distribution_plot(pilot_aggregated)
```

## Measurement: social processes {.smaller}

### Emotional support provided

> **English translation of the original German item**<br>
> “How much did you emotionally support your partner today so that they would be physically active with greater enjoyment?”

**Translated response options:** 0 = *not at all today* to 5 = *very much today*<br>
**Source:** Adapted from @bolgerInvisibleSupportAdjustment2000.<br>
**Variable:** `ss_emo_pleasure`

### Collaborative planning

> **English translation of the original German item**<br>
> “We have already planned precisely for tomorrow when, where, and how we will be physically active together.”

**Translated response options:** 0 = *not at all true today*<br>
to 5 = *completely true today*

**Source:** Adapted from @scholzBeyondBehaviouralIntentions2008; Health Action Process Approach.<br>
**Variable:** `sc_actionplan`

## Measurement: physical activity {.smaller}

> **English translations of the original German items**<br>
> “Did you engage in moderate-to-vigorous physical activity today?”<br>
> If yes: “How many minutes did you spend engaging in moderate-to-vigorous physical activity today?”

- **Alone:** _____ minutes (`ss_pa_min_solo`)
- **Together with your partner:** _____ minutes (`ss_pa_min_collab`)

**Translated response options:** Yes/no screening item followed by numeric minute fields.<br>
**Source:** Screening item developed for the study; duration items adapted from @amireaultGodinShephardQuestionnaire2015.

Reported non-activity days are coded as **0 minutes**; missing diaries remain missing.

**Total MVPA = solo MVPA + joint MVPA.**
