---
title: "Dyadic Data Analysis with R"
subtitle: "Part I: Conceptual Foundations"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
csl: apa.csl
resources:
  - uzh-logo.svg
  - charite-logo.svg
  - js/residual-concepts.js
format:
  revealjs:
    embed-resources: true
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: false
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup-pilot-data
#| include: false

library(dplyr)
library(ggplot2)

diagram_helper_candidates <- c(
  "diagram-helpers.Rinc",
  file.path("vignettes", "diagram-helpers.Rinc"),
  file.path("..", "..", "vignettes", "diagram-helpers.Rinc")
)
diagram_helper <- diagram_helper_candidates[
  file.exists(diagram_helper_candidates)
][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(
  normalizePath(diagram_helper),
  envir = knitr::knit_global()
)


# FUNCTION TO FIT EARLY MODELS --------------------------------------

early_models_file <- if (file.exists(file.path("R", "fit-early-models.R"))) {
  file.path("R", "fit-early-models.R")
} else {
  file.path("dev", "workshop", "R", "fit-early-models.R")
}

sys.source(
  normalizePath(early_models_file),
  envir = knitr::knit_global()
)

dim_grid_candidates <- c(
  "dim-grid.Rinc",
  file.path("vignettes", "dim-grid.Rinc"),
  file.path("..", "..", "vignettes", "dim-grid.Rinc")
)
dim_grid_file <- dim_grid_candidates[file.exists(dim_grid_candidates)][1]

if (is.na(dim_grid_file)) {
  stop("Could not find vignettes/dim-grid.Rinc.")
}

sys.source(
  normalizePath(dim_grid_file),
  envir = knitr::knit_global()
)

apim_exchangeability_file <- if (
    file.exists(file.path("R", "draw-apim-exchangeability.R"))) {
  file.path("R", "draw-apim-exchangeability.R")
} else {
  file.path("dev", "workshop", "R", "draw-apim-exchangeability.R")
}

sys.source(
  normalizePath(apim_exchangeability_file),
  envir = knitr::knit_global()
)

residual_seesaw_file <- if (
    file.exists(file.path("R", "draw-residual-seesaw.R"))) {
  file.path("R", "draw-residual-seesaw.R")
} else {
  file.path("dev", "workshop", "R", "draw-residual-seesaw.R")
}

sys.source(
  normalizePath(residual_seesaw_file),
  envir = knitr::knit_global()
)

truth_bias_draw_file <- if (
    file.exists(file.path("R", "draw-truth-bias.R"))) {
  file.path("R", "draw-truth-bias.R")
} else {
  file.path("dev", "workshop", "R", "draw-truth-bias.R")
}

sys.source(
  normalizePath(truth_bias_draw_file),
  envir = knitr::knit_global()
)

data_file_candidates <- c(
  "dyadic-data.rds",
  file.path("dev", "workshop", "dyadic-data.rds")
)
data_file <- data_file_candidates[file.exists(data_file_candidates)][1]

if (is.na(data_file)) {
  stop("Could not find dev/workshop/dyadic-data.rds.")
}

pilot <- readRDS(data_file) |>
  transmute(
    couple_id = coupleID,
    person_id = personID,
    role = factor(gender, levels = c(1, 2), labels = c("female", "male")),
    provided_support,
    received_support,
    collaborative_planning,
    exerted_persuasion,
    experienced_persuasion,
    self_efficacy,
    solo_mvpa,
    joint_mvpa,
    total_mvpa
  )

distribution_plot <- function(data, variables, bin_widths) {
  labels <- c(
    provided_support = "Provided emotional support",
    received_support = "Received emotional support",
    collaborative_planning = "Collaborative planning (daily mean)",
    solo_mvpa = "Mean solo MVPA (minutes/day)",
    joint_mvpa = "Mean joint MVPA (minutes/day)",
    total_mvpa = "Mean MVPA (minutes/day)"
  )

  missing_widths <- setdiff(variables, names(bin_widths))
  if (length(missing_widths)) {
    stop("Missing bin widths for: ", paste(missing_widths, collapse = ", "))
  }

  plots <- Map(\(variable, bin_width) {
    values <- data[[variable]]
    values <- values[is.finite(values)]
    limits <- c(
      floor(min(values) / bin_width) * bin_width,
      ceiling(max(values) / bin_width) * bin_width
    )
    bandwidth <- density(values, adjust = 1.1)$bw
    curve <- density(
      c(values, -values), bw = bandwidth, from = 0, to = limits[2]
    )
    curve <- tibble(
      value = curve$x,
      count = curve$y * 2 * length(values) * bin_width
    )

    ggplot(tibble(value = values), aes(value)) +
      geom_histogram(
        binwidth = bin_width, boundary = limits[1], closed = "right",
        fill = "steelblue", color = "white", linewidth = 0.25
      ) +
      geom_line(
        data = curve, aes(value, count), inherit.aes = FALSE,
        color = "navy", linewidth = 1
      ) +
      labs(title = labels[[variable]], x = NULL, y = "Number of participants") +
      theme_minimal(base_size = 15) +
      theme(
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text = element_text(color = "#222222")
      )
  }, variables, unname(bin_widths[variables]))

  patchwork::wrap_plots(
    plots,
    ncol = min(3, length(plots)),
    axis_titles = "collect_y"
  )
}
```

## Introduction

- Turu & Pascal <!-- ca. 5 min -->
- Tell us about you and your dyadic data!
  - Name, affiliation & career stage
  - Your dyadic data and RQ
  
<!-- ca. 15 Minutes -->

## Today's Aims and Program

| Time | Focus |
|---|---|
| 11:00–12:45 | Introduction and types of dyadic models |
| 12:45–14:00 | *Lunch break* |
| 14:00–15:30 | Estimating cross-sectional models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

<!-- ca. 2.5 Minutes -->

# Introduction
Relevance and todays' example dataset

## Why dyadic data?

> “Much of what we think, do, and feel is determined not by who we are but by
> whom we are with.” [@kennyPartnerEffectsRelationship1999, p. 446]

::: {.incremental}
- **Partners matter:** An outcome may depend on both *my characteristics* and 
*my partner's characteristics*.
- **Some constructs and processes are dyadic:** Shared contexts and interdependence
processes can make partners' outcomes correlated.
- **The configuration matters:** Partners' similarity, or complementarity (differences) 
may predict outcomes beyond either individual score.
:::

[@kennyPartnerEffectsRelationship1999;
@iidaModelsInterdependentIndividuals2018]{.subtle}

<!-- ca. 2.5 Minutes -->

## Introducing today's dyadic dataset!

- 38 physically inactive opposite-sex romantic couples
- Both partners intended to increase their physical activity together
- Smartphone-based planning interventions and JITAIs
- Daily diaries over 55 days

[@kuengDailyAssociationsSocialControl2026;
@hoehenerEffectivenessSmartphoneDyadic2026]{.subtle}

<!-- ca. 1.5 Minutes -->

## Guiding question {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[Guiding question:]{.rq-label} [In romantic couples, how is provided health-related social support linked to physical activity?]{.rq-text}
:::
:::

:::{.notes}
This is intentionally broad, since we will look at it through different lenses
and analyze this with different models. 

Basically answering differt sub-RQs each time that I will introduce.
:::

<!-- ca. 0.5 Minutes -->

## Measurement - Predictor

### Provided emotional support

> **Rough English translation of the original German item**<br>
> “How much did you emotionally support your partner today so that they would
be physically active with greater enjoyment?”
>
> [Adapted from @bolgerInvisibleSupportAdjustment2000]{.subtle}

0 = *not at all today* <br>
5 = *very much today*<br>

<!-- ca. 1 Minute -->

## Measurement - Outcome

### Moderate to vigorous physical activity (MVPA)

> **Rough English translation of the original German item**<br>
> “How many minutes did you spend engaging in moderate-to-vigorous physical
activity today?” 
[Adapted from @amireaultGodinShephardQuestionnaire2015]{.subtle}

- **Alone:** _____ minutes
- **Together with your partner:** _____ minutes

[**Total MVPA = solo MVPA + joint MVPA.**]{.fragment}

<!-- ca. 1 Minute -->

## Aggregating for cross-sectional examples

To show simple dyadic models we aggregate the daily diary data to
obtain a "cross-sectional" dataset.

```{r}
library(dplyr)

pilot_aggregated <- pilot |>
  group_by(couple_id, person_id, role) |>
  summarise(
    across(where(is.numeric), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )
```

::: {.fragment}
The variables now represent person-level daily averages across the study:

- `provided_support`: average daily support provided
- `received_support`: average daily support received
- `total_mvpa`: average daily total MVPA (minutes)
:::

<!-- ca. 1 Minute -->

## Plotting provided support by partner role

```{r}
#| label: build-pilot-support-distributions-by-role
#| echo: true
#| code-line-numbers: false

support_distribution_plot <- ggplot(
  pilot_aggregated,
  aes(provided_support, y = after_stat(density))
) +
  geom_histogram(
    bins = 15, boundary = 0,
    fill = "steelblue", color = "white", alpha = 0.75
  ) +
  geom_density(
    bounds = c(0, Inf), color = "navy", linewidth = 1.1
  ) +
  facet_wrap(~role, nrow = 1) +
  labs(x = "Average provided support", y = "Density") +
  see::theme_modern(base_size = 16)
```

<!-- ca. 1 Minute -->

## Provided support by partner role

```{r}
#| label: pilot-support-distributions-by-role
#| echo: false
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.1
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Histograms and boundary-corrected density curves of average provided emotional support, shown separately for female and male partners."

support_distribution_plot
```

<!-- ca. 1 Minute -->

## MVPA by partner role

```{r}
#| label: pilot-mvpa-distributions-by-role
#| echo: false
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.1
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Histograms and boundary-corrected density curves of average daily MVPA, shown separately for female and male partners."

ggplot(pilot_aggregated,
       aes(total_mvpa, y = after_stat(density))) +
  geom_histogram(
    bins = 15, boundary = 0,
    fill = "steelblue", color = "white", alpha = 0.75
  ) +
  geom_density(
    bounds = c(0, Inf), color = "navy", linewidth = 1.1
  ) +
  facet_wrap(~role, nrow = 1) +
  labs(x = "Average MVPA (minutes/day)", y = "Density") +
  see::theme_modern(base_size = 16)
```

<!-- ca. 1 Minute -->

# Different types of dyadic models

## One model dominates dyadic research

```{r}
#| label: apim-use-plot
#| echo: false
#| dev: svglite
#| fig-alt: "A horizontal bar showing that more than 85% of 998 coded dyadic studies used an APIM or APIM variant, leaving fewer than 15% for all other model specifications combined."
#| out-width: 100%

model_use <- tibble(
  model = c("APIM or APIM variant", "All other specifications"),
  xmin = c(0, 85),
  xmax = c(85, 100),
  midpoint = c(42.5, 92.5),
  label = c("> 85%\nAPIM or APIM variant", "< 15%\nAll others")
)

ggplot(model_use) +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 1, fill = model),
    color = "white", linewidth = 1.2
  ) +
  geom_text(
    aes(x = midpoint, y = 0.5, label = label, color = model),
    lineheight = 0.95, fontface = "bold", size = c(9, 6.2)
  ) +
  scale_fill_manual(values = c(
    "APIM or APIM variant" = "#0028a5",
    "All other specifications" = "#e8eefb"
  )) +
  scale_color_manual(values = c(
    "APIM or APIM variant" = "white",
    "All other specifications" = "#001f73"
  )) +
  coord_cartesian(xlim = c(0, 100), ylim = c(-0.06, 1.06), expand = FALSE) +
  theme_void() +
  theme(legend.position = "none", plot.margin = margin(12, 6, 12, 6))
```

::: {.subtle}
Preliminary results from an ongoing systematic review (APIM use coded for
$n = 998$ eligible articles)

[@sakalukRenewalDyadicStructural2025, p. 5].
:::

<!-- ca. 1 Minute -->

## Many models to choose from

**Various dyadic models** can be used, each answering different questions!

::: {.incremental}
- **Actor-partner interdependence model (APIM)**: actor and partner associations between interdependent individuals
- **Dyad-individual model (DIM)** and **dyadic score model (DSM)**: associations involving dyad levels and within-dyad differences
- **Common fate model (CFM)**: latent dyad-level constructs reflected in both partners' reports
:::

::: {.fragment .text-smaller}
Other approaches include the **mutual-influence model**, for reciprocal influence
between partners [@iidaModelsInterdependentIndividuals2018, p. 61], and the
**Truth and Bias model**, for separating mean-level bias from correlational
accuracy in judgments of a partner [@kennyAccuracyBiasPerception2001;
@westTruthBiasModel2011].
:::

<!-- ca. 3 Minutes -->

## Many models to choose from

**Various dyadic models** can be used, each answering different questions!

- [**Actor-partner interdependence model (APIM)**: actor and partner associations between interdependent individuals]{.text-highlight}
- **Dyad-individual model (DIM)** and **dyadic score model (DSM)**: associations involving dyad levels and within-dyad differences
- **Common fate model (CFM)**: latent dyad-level constructs reflected in both partners' reports

:::{.text-smaller}
Other approaches include the **mutual-influence model**, for reciprocal influence
between partners [@iidaModelsInterdependentIndividuals2018, p. 61], and the
**Truth and Bias model**, for separating mean-level bias from correlational
accuracy in judgments of a partner [@kennyAccuracyBiasPerception2001;
@westTruthBiasModel2011].
:::

::: {.notes}
"For the guiding question, we first use the Actor-Partner Interdependence Model."
:::


<!-- ca. 1 Minute -->

# Actor-partner interdependence models
Distinguishable & Exchangeable APIM

## RQ1 · Actor and partner associations {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ1:]{.rq-label} [Is a person's provided emotional support associated with (a) their own MVPA and (b) their partner's MVPA?]{.rq-text}
:::
:::

<!-- ca. 0.5 Minutes -->


## The distinguishable APIM

```{r}
#| label: conceptual-distinguishable-apim-paths-only
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with actor and partner paths."

draw_apim_diagram(
  "distinguishable", show_residuals = FALSE,
  predictors_centered = TRUE
)
```

::: {.fragment}
Actor and partner effects together explain some of the covariance between partners' 
outcomes. 
:::


<!-- 
1. Explain actor and partner effects only and briefly. 

2. Reveal the final statement 
-->

<!-- ca. 2 Minutes -->

## The distinguishable APIM

```{r}
#| label: conceptual-distinguishable-apim_brief
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with role-specific actor and partner paths."

draw_apim_diagram("distinguishable", predictors_centered = TRUE)
```

Actor and partner effects together explain some of the covariance between partners' 
outcomes.  $\rightarrow$ But some covariance may remain!

:::{.notes}
But more on the residuals later!!!

-> Ask students what grand-mean centering is. 

:::

<!-- ca. 0.5 Minutes -->

## How to estimate actor and partner effects?

Conceptually, we are esimating these two equations in a single model:

$$
  \begin{gathered}
  \\[0.5em]
  Y_{\mathrm{male},i}
  = b_{0,\mathrm{M}} + b_{1,\mathrm{M}} \times X_{\mathrm{male},i}
  + b_{2,\mathrm{M}} \times X_{\mathrm{female},i} + \epsilon_{\mathrm{male},i}
  \\[2em]
  Y_{\mathrm{female},i}
  = b_{0,\mathrm{F}} + b_{1,\mathrm{F}} \times X_{\mathrm{female},i}
  + b_{2,\mathrm{F}} \times X_{\mathrm{male},i} + \epsilon_{\mathrm{female},i}
  \end{gathered}
$$
<br>

[Could we just use two separate models?]{.fragment}



<!-- ca. 2 Minutes -->

## Could we just use two separate models?

### Male model

```{r eval = FALSE}
male_model <- glmmTMB(
  
  outcome_male ~ X_male + X_female,
  
  data = data
)
```
<br>

### Female model

```{r eval = FALSE}
female_model <- glmmTMB(
  
  outcome_female ~ X_female + X_male,
  
  data = data
) 

```

<!-- ca. 0.5 Minutes -->

## Could we just use two separate models?

**Yes, we could! But then we: **

:::{.incremental}
- can't test whether the two actor effects or partner effects are equal
- can't pool equal paths into a shared estimate, which can increase power
- ignore unmodelled sources of interdependence between partners
  - joint modelling tells us how much interdependence remains after accounting for the predictors

:::

::: {.bottom-reference}
[@kennyPartnerEffectsRelationship1999, pp. 439–440;
@iidaModelsInterdependentIndividuals2018, pp. 68–69]
:::

:::{.notes}

The reduction in residual covariance indicates how much of the original 
within-dyad covariance is accounted for by the predictors.

:::

<!-- ca. 3 minutes -->


## Residual interdependence in the distinguishable APIM

```{r}
#| label: conceptual-distinguishable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with role-specific actor and partner paths."

draw_apim_diagram("distinguishable", predictors_centered = TRUE)
```

:::{.notes}

"This represents the two model formulas before, but in a single model with correlated
residuals." 

Map the preceding visual onto the APIM notation:

- Each outcome has a residual, which is what's not explained by the predictors.
- The residual standard deviations, $\sigma$, may differ between distinguishable
  partners. Their squares are the residual variances from the previous slide.
- The curved double-headed arrow is the residual correlation,
  $\rho_{\epsilon_F\epsilon_M}$.
  
:::

<!-- ca. 2 Minutes -->

## What remains after prediction?

```{r}
#| label: residual-variance-and-correlation
#| echo: false
#| results: asis
#| fig-width: 12
#| fig-height: 5.6
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Schematic residual plot showing a fixed prediction, smaller residual variance for female partners, larger residual variance for male partners, and a positive association between partners' residuals."

residual_helper_candidates <- c(
  file.path("R", "draw-residual-concepts.R"),
  file.path("dev", "workshop", "R", "draw-residual-concepts.R")
)
residual_helper <- residual_helper_candidates[
  file.exists(residual_helper_candidates)
][1]

if (is.na(residual_helper)) {
  stop("Could not find dev/workshop/R/draw-residual-concepts.R.")
}

sys.source(
  normalizePath(residual_helper),
  envir = knitr::knit_global()
)

if (knitr::is_html_output()) {
  residual_concepts_widget()
} else {
  print(draw_residual_concepts())
}
```

:::{.notes}

On the left:

- Everyone has the same predicted value in this simplified example.
- The x-axis identifies illustrative dyads in an arbitrary order.

- *A residual is the observed value minus that predicted value. *

On the right:

- Every numbered point is the same dyad as on the left, combining its two
  partners' raw residuals.
  
- **Move variance sliders** $\rightarrow$ raw residuals and paired
  points move, but the displayed correlation stays exactly the same.

- Residual **correlation describes interdependence** that remains after accounting
  for the predictors. It is separate from the size of either residual variance.

- **Move the correlation** while watching both sides: the residuals change, but
  both selected residual variances stay exactly the same.

:::

<!-- ca. 4 Minutes -->


## Fitting the distinguishable APIM [Interpretation →](#interpreting-the-distinguishable-apim-results){.results-forward-link} {#fitting-the-distinguishable-apim}

```{r}
#| label: fitted-distinguishable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted distinguishable APIM for provided support and MVPA."

distinguishable_apim <- fit_and_draw_distinguishable_apim(
  pilot_aggregated,
  predictor = provided_support,
  outcome = total_mvpa,
  labels = c(predictor = "Provided\nsupport", outcome = "MVPA")
)
```

:::{.notes}

**Ask participants to interpret** 

**MVPA: ** minutes of average daily total MVPA across the 55 days

- All associations are conditional on both partners' provided support.

- The intercepts, 35.28 for female partners and 31.24 for male partners, are the
  predicted daily MVPA when both partners' average provided support equals the
  pooled sample mean (0.92 on the original scale).
  
- Female actor effect, -4.61: for each one-unit higher support provided by a
  female partner, her own predicted daily MVPA is 4.61 minutes lower, holding
  her male partner's support constant.
  
- Female partner effect, 15.06: for each one-unit higher support provided by a
  male partner, the female partner's predicted daily MVPA is 15.06 minutes
  higher, holding her own support constant.
  
- Male actor effect, 17.92: for each one-unit higher support provided by a male
  partner, his own predicted daily MVPA is 17.92 minutes higher, holding his
  female partner's support constant.
  
- Male partner effect, -9.87: for each one-unit higher support provided by a
  female partner, the male partner's predicted daily MVPA is 9.87 minutes lower,
  holding his own support constant.
  
- The residual SDs, 21.40 minutes for female partners and 16.73 minutes for male
  partners, describe the remaining person-to-person spread around those
  predictions. An SD is a typical scale of residual deviation, not a variance
  and not the uncertainty of a coefficient.
  
- The residual correlation, .78, indicates strong positive remaining
  within-couple interdependence: couples in which the female partner is above
  her prediction also tend to have a male partner above his prediction, and
  likewise for being below prediction. This remains after accounting for both
  support scores.

These are associations in the aggregated observational data, not causal effects.

:::

<!-- ca. 5 Minutes -->

## The exchangeable APIM

**Sometimes:** 

- partners may not be theoretically meaningfully distinguishable
- no evidence of difference in effects $\rightarrow$ pool effects for greater power
- we want to formally test distinguishability

:::{.fragment}
$\rightarrow$ Then, we may fit an exchangeable model that imposes symmetry
:::

<!-- ca. 2 Minutes -->

## The exchangeable APIM

```{r}
#| label: conceptual-exchangeable-apim-paths-only
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual exchangeable APIM with pooled actor and partner paths."

draw_apim_diagram(
  "exchangeable", show_residuals = FALSE,
  predictors_centered = TRUE
)
```

:::{.notes}

Two members are now named arbitrarily. Swapping labels will NOT change the model

Pooled actor and partner effects

pooled intercepts pooling

:::

<!-- ca. 1 Minute -->


## The exchangeable APIM

```{r}
#| label: conceptual-exchangeable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual exchangeable APIM with pooled actor and partner paths."

draw_apim_diagram("exchangeable", predictors_centered = TRUE)
```

:::{.notes}

The two residual SDs are also constrained to be equal (e.g., pooled) 

This says that the unexplained outcome spread is the same for both arbitrary member positions.

The curved arrow remains: exchangeability does not assume that partners are
independent!!!! Their residual correlation is still estimated.

:::

<!-- ca. 2 Minutes -->

## Exchangeability constraints

```{r}
#| label: apim-exchangeability-comparison
#| echo: false
#| dev: svg
#| fig-width: 10.5
#| fig-height: 5.2
#| out-width: 95%
#| fig-align: center
#| fig-alt: "Distinguishable and exchangeable APIM equations, predictions, and residual constraints."

draw_apim_exchangeability_comparison()
```

:::{.notes}

In this example, imagine we have a few couples with exactly $X_F = 1$ and $X_M = 1$
on the GM-centered scale.
For these couples, the following would be true... 

On the left: 

- intercepts, actor effects, partner effects, and residual standard
  deviations may differ by role.
  
- Follow each prediction from its intercept through the actor contribution in
  blue and the partner contribution in purple. 
  
- The open circle is the conditional prediction. 
  - The nearby dots represent illustrative observed outcomes, 
  - and the arrow labelled $\epsilon$ isolates one residual as the distance from the prediction to an observation.
  
- The shaded area is supposed to visualize the residual variance  
  
On the right:

- the same intercept, actor effect, partner effect, and residual
  standard deviation are used for both arbitrary member positions.

- **Shared coefficients do not imply identical predictions!!**: the actor and partner
  predictor values are exchanged between the two equations.
  
- Equal residual variance is a constraint on the modeled residual distribution,
  not a claim that the two members have identical residual values.
  
- In both models, partners' residuals may still be correlated!

:::

<!-- ca. 4 Minutes -->

## Fitting the exchangeable APIM [Interpretation →](#interpreting-the-exchangeable-apim-results){.results-forward-link} {#fitting-the-exchangeable-apim}

```{r}
#| label: fitted-exchangeable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted exchangeable APIM for provided support and MVPA."

exchangeable_apim <- fit_and_draw_exchangeable_apim(
  pilot_aggregated,
  predictor = provided_support,
  outcome = total_mvpa,
  labels = c(predictor = "Provided\nsupport", outcome = "MVPA")
)
```

<!-- ca. 3 Minutes -->

:::{.notes}

-> Let students interpret

-> Ask if they think this model fits the data well, considering the distinguishable estimates. 

:::

## Testing exchangeability

```{r}
library(dyadMLM)

compare_nested_glmmTMB_models(
  exchangeable_apim,
  distinguishable_apim
)
```

# Level-and-difference models
Dyad-Individual Model & Dyadic Score Model

## Shifting the analytical lens

Sometimes, actor and partner associations are not the primary interest.

Instead, we may want to understand:

- how couples compare with other couples
- how partners stand relative to one another within a couple

::: {.fragment}

This shifts the focus to **dyadic levels and within-dyad differences**.

:::

::: {.fragment}

  - for exchangeable dyads, use the **dyad-individual model (DIM)**
  - for distinguishable dyads, use the **dyadic score model (DSM)**
  
:::

<!-- ca. 2 Minutes -->

## RQ2 · Dyad levels and differences {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ2:]{.rq-label} [How are couples' average levels and partner differences in provided emotional support associated with the couple-average MVPA and partner differences in MVPA?]{.rq-text}
:::
:::

<!-- ca. 0.5 Minutes -->


## The dyad-individual model (DIM)

**The DIM is:**

- A re-parametrization of the **exchangeable** APIM!
- We decompose predictors into two variables:
  - Dyad mean (GM-centered): $\bar{X} - \mu_{X}$
  - Within-dyad member deviation: $X_i - \bar{X}$

:::{.fragment}

In other words, these predictors express: 

- A couple's average level compared with the average of all couples
- A partner's level compared with the average of their couple

:::

:::{.notes}

So, you can see how this would answer different research questions than an
APIM, even though both model encode the exact same information and are
interchangeable.

:::

<!-- ca. 2 Minutes -->

## DIM predictors are APIM predictors in disguise!

```{r}
#| label: interactive-apim-dim-grid
#| echo: false
#| results: asis

dim_reparameterization_grid(
  exchangeable_apim,
  id = "dim-grid-introduction",
  predictor_label = "Provided support",
  outcome_label = "MVPA",
  compact = TRUE
)
```

<!-- ca. 4 Minutes -->


## The DIM

```{r}
#| label: conceptual-dim-paths-only
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DIM with a between-dyad path from the dyad mean and a within-dyad path from the within-dyad member deviation."

draw_dim_diagram(show_residuals = FALSE)
```

:::{.notes}

- Is a couple that is higher on X than others also higher on Y than others?

- is a person within a couple that is higher than their partner on X 
(or higher than the dyad-mean), also higher than their partner on Y?

:::

<!-- ca. 1 Minute -->

## The DIM

```{r}
#| label: conceptual-dim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DIM separating the between-dyad effect of the dyad mean from the within-dyad effect of the within-dyad member deviation."

draw_dim_diagram()
```

:::{.notes}

Ask people: where did the residual correlation between partners go? Are they no longer correlated?

$\rightarrow$ No, the residual correlation between individuals residuals are encoded in these two residuals.

The mean residual captures whether both partners are higher or lower than 
predicted $\rightarrow$ Can capture positive correlation between the partners!

The deviation residual captures whether one partner is higher while the
other is lower $\rightarrow$ can capture negative correlation between the partners!

$$
\operatorname{Cov}(e_1,e_2)
=\operatorname{Var}(M)-\operatorname{Var}(D).
$$

:::

<!-- ca. 2 Minutes -->


## Why are the mean and deviation residuals uncorrelated?

```{r}
#| label: dim-residual-seesaw
#| echo: false
#| dev: svg
#| fig-width: 10.8
#| fig-height: 3.15
#| out-width: 100%
#| fig-align: center
#| fig-alt: "An elevator represents the dyad mean residual, which moves both partners together; a seesaw represents the within-dyad residual, which moves partners in opposite directions. Swapping arbitrary member labels reverses only the seesaw, requiring zero covariance under exchangeability."

draw_residual_seesaw("dim")
```

::: {.seesaw-swap}
**Swap the arbitrary member labels:**
$r_{\mathrm{M}} \mapsto r_{\mathrm{M}} \qquad
r_{\mathrm{D}} \mapsto -r_{\mathrm{D}}$
:::

::: {.seesaw-equation}
Exchangeability requires the same covariance after relabelling:

$$
\operatorname{Cov}(r_{\mathrm{M}},r_{\mathrm{D}})
= \operatorname{Cov}(r_{\mathrm{M}},-r_{\mathrm{D}})
= -\operatorname{Cov}(r_{\mathrm{M}},r_{\mathrm{D}})
\quad\Longrightarrow\quad
\operatorname{Cov}(r_{\mathrm{M}},r_{\mathrm{D}})=0
$$
:::

::: {.seesaw-takeaway}
**Zero does not mean independent partners:**
$\operatorname{Cov}(e_1,e_2)
=\operatorname{Var}(r_{\mathrm{M}})-\operatorname{Var}(r_{\mathrm{D}})$

Their residual association is encoded by the relative elevator and seesaw variances.
:::

:::{.notes}

POTENTIALLY SKIP FOR TIME

Elevator: once the couples' mean value X has accounted for some variance in the
couples' outcome Y, there might still be some movement up or down that is unaccounted. $\rightarrow$ residual

Seesaw: once a persons' difference from the couple mean has accounted for some variance
in the persons' difference from the couple mean in Y, there might still be some additional deviation $\rightarrow$ residual

In the DIM, the member positions on the seasaw are arbitrary.
Swapping them leaves the elevator coordinate unchanged but reverses the sign of
the seesaw coordinate.

---

You might wonder, why the elevator and seasaw residuals are not correlated.

A positive mean–deviation residual correlation would mean:
  - couples above their prediction on average also tend to tilt toward the arbitrarily named Member 1. 
  - But Member 1 and Member 2 are assigned arbitrarily. Across couples, reversing these arbitrary assignments produces equally valid positive and negative signed deviations, which cancel in expectation. 
  - Therefore, the exchangeable DIM therefore constrains the population correlation to zero.

But again, this does not make the partners independent, and this model
IS exactly equivalent to the APIM

:::

<!-- ca. 4 Minutes -->

## Fitting the DIM [Interpretation →](#interpreting-the-fitted-dim-results){.results-forward-link} {#fitting-the-dim}

```{r}
#| label: fitted-dim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted DIM for provided support and MVPA."

fit_and_draw_dim(
  pilot_aggregated,
  predictor = provided_support,
  outcome = total_mvpa,
  labels = c(predictor = "support", outcome = "MVPA")
)
```

<!-- ca. 3 Minutes -->

:::{.notes}

Keep in mind that previously with the same data we have seen that 
the exchangeability constraints are not really supported by the data. Nonetheless
we fit the DIM here with the same data and variables to demonstrate. 

-> let students interpret

-> Why no intercept? for the deviation? 

  No intercept for the MVPA deviation, because that would break exchangeability. 
  if support deviation is zero, then the expected MVPA deviation is also zero. 
  
  -> We conceptually DO have an intercept, it is just constrained to be ZERO!
  
  -> just like the correlation between the residuals, they are not estimated, which 
  effectively constrains them to be zero. 
  
:::

## DIM to exchangeable APIM coefficients

- Once we fit an APIM, we have already fitted a DIM—and vice versa.

- We can transform values from a fitted APIM into DIM values.


## DIM to exchangeable APIM coefficients {.compact-derivation}

[The intercept is omitted; APIM predictors are grand-mean centered.]{.derivation-context}

For a focal person, let
$X_{\mathrm{self}}=1,\quad X_{\mathrm{partner}}=1$,

so that in DIM coordinates:
$X_{\mathrm{mean}}=1,\quad X_{\mathrm{dev,self}}=0.$

So,

$$
\hat Y_{\mathrm{self}}=1\times a+1\times p
$$

implies

$$
\hat Y_{\mathrm{self}}=1\times b_{\mathrm{mean}}+0\times b_{\mathrm{dev}}
=b_{\mathrm{mean}}.
$$

Therefore,

$$
b_{\mathrm{mean}}=a+p.
$$

## DIM to exchangeable APIM coefficients {.compact-derivation}

[The intercept is omitted; APIM predictors are grand-mean centered.]{.derivation-context}

For a focal person, let
$X_{\mathrm{self}}=1,\quad X_{\mathrm{partner}}=-1$,

so that in DIM coordinates:
$X_{\mathrm{mean}}=0,\quad X_{\mathrm{dev,self}}=1$.

So,

$$
\hat Y_{\mathrm{self}}=1\times a+(-1)\times p
$$

implies:

$$
\hat Y_{\mathrm{self}}=0\times b_{\mathrm{mean}}+1\times b_{\mathrm{dev}}
=b_{\mathrm{dev}}.
$$

Therefore,

$$
b_{\mathrm{dev}}=a-p.
$$


## DIM to exchangeable APIM coefficients

```{r}
#| label: interactive-apim-dim-grid-coefficients
#| echo: false
#| results: asis

dim_reparameterization_grid(
  exchangeable_apim,
  id = "dim-grid-coefficients",
  predictor_label = "Provided support",
  outcome_label = "MVPA",
  compact = TRUE
)
```

<!-- Practical quick demonstration and showing formulas -->

<!-- ca. 5 Minutes -->

## The dyadic score model (DSM)

What if we want DIM-style interpretation but have distinguishable dyads?

:::{.fragment}
**The DSM is:**

- the unconstrained distinguishable version of the DIM
- a re-parametrization of the distinguishable APIM
:::

:::{.fragment}
- **DIM**: deviation from the couple mean $X_i - \bar{X}$
  - opposite for each partner 
- **DSM**: meaningful directed differences $X_\mathrm{female} - X_\mathrm{male}$
  - repeated for both partners
:::


<!-- ca. 3 Minutes -->

## The DSM

```{r}
#| label: conceptual-dsm-paths-only
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DSM with paths between predictor and outcome scores."

draw_dsm_diagram(show_residuals = FALSE)
```


:::{.notes}

- a11 is interpreted the same way as b_mean in the DIM (and a10 too).

- a22: The DIM said: "if any of the partners is higher on X, is that same partner also higher on Y."
  - The DSM says: "if the female is higher (or lower) than the male on X, is the female also higher (or lower) than the male on Y". 

a21/a12:
Because we now have a stable direction that we defined with Female minus Male 
having the same meaning in every couple. These directional terms therefore no 
longer cancel (to zero) and may be estimated.

:::

<!-- ca. 3 Minutes -->

## The DSM

```{r}
#| label: conceptual-dsm
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DSM linking predictor mean and difference to outcome mean and difference."

draw_dsm_diagram()
```

:::{.notes}

Like the DIM, the DSM decomposes the two partner residuals into a residual mean 
and a residual difference. 

In the DIM, the positive direction of the residual difference is arbitrarily assigned. 
Positive and negative directional associations therefore cancel across arbitrary assignments.

In the DSM, the direction is fixed as Female minus Male. Positive has the same substantive meaning in every couple, so the residual mean and residual difference may correlate.

:::

<!-- ca. 3 minutes --> 

## Why may the DSM mean and difference residuals correlate?

```{r}
#| label: dsm-residual-seesaw
#| echo: false
#| dev: svg
#| fig-width: 10.8
#| fig-height: 3.15
#| out-width: 100%
#| fig-align: center
#| fig-alt: "An elevator represents the dyad mean residual, which moves both partners together; a seesaw with female and male seats represents the directed outcome-difference residual. Because these substantive labels cannot be swapped arbitrarily, the two residual components may correlate."

draw_residual_seesaw("dsm")
```

::: {.seesaw-swap}
**Substantive labels:**
$D_Y=Y_{\mathrm{F}}-Y_{\mathrm{M}}$

Swapping Female and Male changes the meaning and sign of the difference, so
label-swapping is not a symmetry of the DSM.
:::

::: {.seesaw-equation .seesaw-equation-dsm}
$$
\operatorname{Cov}(r_{\mathrm{M}},r_{\mathrm{D}})
\quad\text{is not constrained to zero.}
$$
:::

::: {.seesaw-takeaway}
**A positive $\rho_{r_\mathrm{M}r_\mathrm{D}}$ means:** couples whose outcomes are
higher than predicted on average also tend to have a larger female–male
difference than predicted.
:::

:::{.notes}

- The DSM uses the same elevator and seesaw idea, but the two seats now have
  substantively meaningful labels.

- Swapping Female and Male changes the meaning and sign of the directed
  difference. 

- The mean and difference residuals may consequently be correlated.

- A positive correlation means that, across couples, couples whose outcomes are
  higher than predicted on average also tend to have a more positive female minus male
  difference than predicted. 

:::

<!-- ca. 3 Minutes -->

## Fitting the DSM [Interpretation →](#interpreting-the-fitted-dsm-results){.results-forward-link} {#fitted-dsm-results}

```{r}
#| label: fitted-dsm
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted DSM for provided support and MVPA."

fit_and_draw_dsm(
  pilot_aggregated,
  predictor = provided_support,
  outcome = total_mvpa,
  labels = c(predictor = "support", outcome = "MVPA")
)
```


:::{.notes}

-> students interpret results, including correlation. 

**Intercepts**: For a couple with an average support mean and no female–male support difference:

- a10: predicted couple-mean MVPA is 33.26 minutes,
- a20: the female is predicted to be 4.04 min more active than the male.

Holding the couple-difference score at zero, 
**a 1 unit higher couple-mean support score** is associated with:

- a11: 9.25 additional minutes of couple-mean MVPA,
- a21: a 2.4-minute larger female-male MVPA difference (The female is 2.4 minutes + 4.04 minutes more active than the male). 

Holding the couple-mean score at the *sample-average*, 
**a one-unit larger female-male support difference** is associated with:

- a12: 11.86 fewer minutes of couple-mean MVPA
- a22: a 4.06-minute larger female-male MVPA difference (The female is 4.06 minutes + 4.04 minutes more active than the male)

**Residual correlations**

After accounting for the support mean and female–male support difference, 
couples whose average MVPA is higher than predicted also tend to have a more 
positive female–male MVPA difference than predicted. 

In other words, their unexplained MVPA tends to tilt toward the female partner.

<!-- TODO: after the workshop perhaps create a slide-deck that has the 
interpretations on a separate slide for each model -->

:::


## DSM to distinguishable APIM coefficients {.smaller}

- Once we fit a DSM, we have already fitted a distinguishable APIM—and vice versa.

- We can back-transform the DSM fixed effects into APIM coefficients:

$$
\begin{aligned}
b_{0,\mathrm F}
  &=a_{10}+\tfrac12a_{20},
&
b_{0,\mathrm M}
  &=a_{10}-\tfrac12a_{20},\\[0.4em]
b_{1,\mathrm F}
  &=\tfrac12a_{11}+a_{12}+\tfrac14a_{21}+\tfrac12a_{22},
&
b_{1,\mathrm M}
  &=\tfrac12a_{11}-a_{12}-\tfrac14a_{21}+\tfrac12a_{22},\\[0.4em]
b_{2,\mathrm F}
  &=\tfrac12a_{11}-a_{12}+\tfrac14a_{21}-\tfrac12a_{22},
&
b_{2,\mathrm M}
  &=\tfrac12a_{11}+a_{12}-\tfrac14a_{21}-\tfrac12a_{22}.
\end{aligned}
$$

Here, $b_1$ is the actor effect and $b_2$ is the partner effect. The APIM
predictors and DSM couple mean now use the same pooled grand-mean reference, so
no additional centering adjustment is needed for the intercepts.

# Common Fate Model
Modeling Dyadic Constructs

## When the construct is dyadic

Sometimes, a construct describes the **couple as a unit**, rather than either
partner individually.

- Both partners report on the same dyadic phenomenon.
- Differences between their reports are not the substantive focus.
- Interest lies in the couple's underlying dyadic level.

:::{.fragment}

**For our data-example:**

- Collaborative planning as a dyadic predictor
- Joint physical activity as a dyadic outcome

:::

<!-- ca. 1 Minute -->

:::{.notes}

-> Ask students if they had ideas on a dyadic construct

:::

## RQ3 · Shared dyadic constructs {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ3:]{.rq-label} [How is collaborative planning associated with joint physical activity in romantic couples?]{.rq-text}
:::
:::

## Measurement - CFM predictor

### Dyadic predictor: collaborative planning

> **Rough english translation of the original German item**<br>
> “We have already planned precisely for tomorrow when, where, and how we will
be physically active together.”
[Adapted from @scholzBeyondBehaviouralIntentions2008]{.subtle}

0 = *not at all true today*<br>
to 5 = *completely true today*

We aggregate each person's ratings into their **average daily collaborative
planning** across the 55 days.

## Do partners report similar dyadic levels?

```{r}
#| label: cfm-partner-agreement
#| echo: false
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.1
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Partner-agreement plots for collaborative planning and joint MVPA, with female reports on the horizontal axis, male reports on the vertical axis, and diagonal identity lines."

cfm_data <- pilot_aggregated |>
  group_by(couple_id) |>
  filter(n_distinct(role) == 2) |>
  ungroup() |>
  select(couple_id, role, collaborative_planning, joint_mvpa) |>
  tidyr::pivot_wider(
    names_from = role,
    values_from = c(collaborative_planning, joint_mvpa),
    names_sep = "_"
  )

planning_agreement_plot <- ggplot(
  cfm_data,
  aes(collaborative_planning_female, collaborative_planning_male)
) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey55") +
  geom_point(color = "steelblue", size = 2.4, alpha = 0.8) +
  coord_equal() +
  labs(
    title = "Collaborative planning (0–5)",
    x = "Female report",
    y = "Male report"
  ) +
  see::theme_modern(base_size = 16)

mvpa_agreement_plot <- ggplot(
  cfm_data,
  aes(joint_mvpa_female, joint_mvpa_male)
) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey55") +
  geom_point(color = "steelblue", size = 2.4, alpha = 0.8) +
  coord_equal() +
  labs(
    title = "Joint MVPA (minutes/day)",
    x = "Female report",
    y = "Male report"
  ) +
  see::theme_modern(base_size = 16)

patchwork::wrap_plots(planning_agreement_plot, mvpa_agreement_plot)
```

:::{.notes}

- Each point is one couple; the diagonal indicates identical partner reports.
- The partners need not report exactly the same value. The CFM treats both
  reports as imperfect indicators of an underlying dyadic level.
- Marginal distributions alone would not show the within-couple correspondence
  that motivates the CFM.

:::

<!-- ca. 1 Minute -->

## The distinguishable CFM

```{r}
#| label: conceptual-cfm-paths-only
#| echo: false
#| dev: svg
#| fig-width: 10
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable Common Fate Model with female and male reports as indicators of shared latent levels."

draw_cfm_diagram(
  show_residuals = FALSE,
  member_ids = c("F", "M"),
  member_names = c("Female", "Male")
)
```

## The distinguishable CFM

```{r}
#| label: conceptual-cfm
#| echo: false
#| dev: svg
#| fig-width: 10
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable Common Fate Model with female and male indicator intercepts, shared latent predictor and outcome levels, and symbolic standard deviations for all four indicator residuals."

draw_cfm_diagram(
  member_ids = c("F", "M"),
  member_names = c("Female", "Male")
)
```


:::{.notes}

- The intercepts are measurement intercepts. The expected values of each members' report
when the corresponding latent dyad level is zero. 
  - because latent means are fixed to zero (as a reference point) they represent the mean reported values.

- Member 1 and Member 2 may be meaningful, **distinguishable** roles, such as
  female and male.
- Alternatively, the positions may be assigned arbitrarily and the members
  treated as **exchangeable**.
- The next slide shows the equality constraints required so that swapping the
  arbitrary labels does not change the model.

:::

## Fitted distinguishable CFM

::: {.cfm-constraint}
**Model constraint:** The female and male joint-MVPA residual variances were
constrained equal to obtain admissible fit.
:::

```{r}
#| label: fitted-cfm
#| echo: false
#| dev: svg
#| fig-width: 10
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted Common Fate Model for collaborative planning and joint MVPA, including four observed indicator intercepts and unstandardized standard deviations for all four indicator residuals."

distinguishable_cfm_model <- '
  # Shared dyad levels: fix all four loadings to 1.
  planning_level =~
    1 * collaborative_planning_female +
    1 * collaborative_planning_male
  joint_mvpa_level =~
    1 * joint_mvpa_female +
    1 * joint_mvpa_male

  # Association between the two shared levels.
  joint_mvpa_level ~ b_level * planning_level

  # Same-member residual covariances.
  collaborative_planning_female ~~ c_female * joint_mvpa_female
  collaborative_planning_male ~~ c_male * joint_mvpa_male

  # Both partners report the same jointly performed activity.
  joint_mvpa_female ~~ e_mvpa * joint_mvpa_female
  joint_mvpa_male ~~ e_mvpa * joint_mvpa_male
'

workshop_distinguishable_cfm <- lavaan::sem(
  distinguishable_cfm_model,
  data = cfm_data,
  missing = "fiml",
  meanstructure = TRUE
)

draw_cfm_diagram(
  model = workshop_distinguishable_cfm,
  member_ids = c("F", "M"),
  member_names = c("Female", "Male"),
  predictor_name = "collaborative planning",
  outcome_name = "joint MVPA",
  show_standardized_effect = FALSE
)
```

:::{.notes}

- The unconstrained model converged but estimated a negative male joint-MVPA
  residual variance. Female and male reports of their joint activity correlate
  almost perfectly.
- We therefore constrain the two joint-MVPA residual variances to equality.
  This is defensible because both partners report the same jointly performed
  activity using the same measure.
- The shared-level association is observational and is estimated from a small
  sample.

:::

<!-- ca. 2 Minutes -->

## When should we model a CFM as exchangeable?

:::{.incremental}
- Partner positions are arbitrary, or theory and the data support treating the
  measurement process as invariant across roles.
- Swapping the partner labels should then leave the model unchanged.
- We constrain corresponding partner parameters to be equal:
  - **intercepts:** the same expected report level
  - **residual variances:** the same amount of indicator-specific variation
  - **same-member residual associations:** the same residual predictor–outcome
    association for both partners
  
:::

[@ledermannMediationDyadicData2009; @peughAnalyzingMixedDyadic2013]{.subtle}

:::{.notes}

- Equality constraints express the assumption that the measurement process does
  not depend on which arbitrary partner position receives a report.
- A well-fitting constrained model does not by itself make substantively
  meaningful roles exchangeable; the decision must also follow the design and
  research question.

:::

## Fitted CFM with partner-role equality constraints

```{r}
#| label: fitted-exchangeable-cfm
#| echo: false
#| dev: svg
#| fig-width: 10
#| fig-height: 5.35
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted exchangeable Common Fate Model for collaborative planning and joint MVPA. Female and male indicator intercepts, corresponding residual variances, and same-member residual covariances are constrained to equality."

exchangeable_cfm_model <- '
  # Shared dyad levels: fix all four loadings to 1.
  planning_level =~
    1 * collaborative_planning_female +
    1 * collaborative_planning_male
  joint_mvpa_level =~
    1 * joint_mvpa_female +
    1 * joint_mvpa_male

  # Association between the two shared levels.
  joint_mvpa_level ~ b_level * planning_level

  # Equal indicator intercepts across positions.
  collaborative_planning_female ~ i_planning * 1
  collaborative_planning_male ~ i_planning * 1
  joint_mvpa_female ~ i_mvpa * 1
  joint_mvpa_male ~ i_mvpa * 1

  # Equal residual variances across positions.
  collaborative_planning_female ~~ e_planning * collaborative_planning_female
  collaborative_planning_male ~~ e_planning * collaborative_planning_male
  joint_mvpa_female ~~ e_mvpa * joint_mvpa_female
  joint_mvpa_male ~~ e_mvpa * joint_mvpa_male

  # Equal same-member residual covariances across positions.
  collaborative_planning_female ~~ c_female * joint_mvpa_female
  collaborative_planning_male ~~ c_male * joint_mvpa_male
  c_female == c_male
'

workshop_exchangeable_cfm <- lavaan::sem(
  exchangeable_cfm_model,
  data = cfm_data,
  missing = "fiml",
  meanstructure = TRUE
)

cfm_model_comparison <- lavaan::lavTestLRT(
  workshop_distinguishable_cfm,
  workshop_exchangeable_cfm
)

cfm_chisq_difference <- cfm_model_comparison$`Chisq diff`[2]
cfm_df_difference <- cfm_model_comparison$`Df diff`[2]
cfm_comparison_p <- cfm_model_comparison$`Pr(>Chisq)`[2]

draw_cfm_diagram(
  model = workshop_exchangeable_cfm,
  parameter_labels = c(
    level = "b_level",
    residual_1 = "c_female",
    residual_2 = "c_male"
  ),
  member_ids = c("F", "M"),
  member_names = c("Female", "Male"),
  predictor_name = "collaborative planning",
  outcome_name = "joint MVPA",
  show_standardized_effect = FALSE
)

cfm_level_estimate <- lavaan::parameterEstimates(
  workshop_exchangeable_cfm
) |>
  filter(label == "b_level") |>
  pull(est)

cfm_level_r2 <- lavaan::lavInspect(
  workshop_exchangeable_cfm, "rsquare"
)[["joint_mvpa_level"]]
```

::: {.cfm-constraint}
**Likelihood-ratio test of the remaining partner-role equality constraints:**
$\Delta\chi^2(`r cfm_df_difference`) = `r sprintf("%.2f", cfm_chisq_difference)`$,
$p = `r sub("^0", "", sprintf("%.3f", cfm_comparison_p))`$; no detectable loss
of fit.
:::

:::{.notes}

- The four additional equality constraints fit the data very well. The
  likelihood-ratio test provides no evidence that they worsen fit.
- This is evidence that pooling the female and male measurement parameters is
  compatible with this small sample; it does not establish that gender is
  substantively irrelevant or that the partners were arbitrarily assigned.
- The fitted shared-level association remains observational.

:::

<!-- ca. 3 Minutes -->

## RQ3 result {.rq-slide}

::: {.rq-center}

> **RQ3:** In this small observational sample, a one-unit higher latent
> collaborative-planning level was associated with
> `r sprintf("%.2f", cfm_level_estimate)` more minutes of joint MVPA per day at
> the couple level ($R^2 = `r sprintf("%.2f", cfm_level_r2)`$).

:::


:::{.notes}

- Here, $R^2 = `r sprintf("%.2f", cfm_level_r2)`$ means that collaborative
  planning, the model's only predictor, accounts for
  `r sprintf("%.0f", 100 * cfm_level_r2)`% of the total model-implied variance
  across couples in the **latent joint-MVPA level**. It does not refer to the
  total variance in the two observed partner reports.
- This is an outcome-specific, model-based descriptive summary—not an overall
  fit measure or evidence of a causal effect.

:::
<!-- ca. 1 Minute -->

# Other interesting dyadic models to consider

## Other dyadic models and designs {.smaller}

- **Truth and Bias model:** Separates accuracy in judging a partner from
  directional bias and assumed similarity [@westTruthBiasModel2011].

- **Mutual-influence model:** Tests whether partners' outcomes directly and
  reciprocally influence one another [@kennyModelsNonIndependenceDyadic1996].

- **APIM mediation model:** Traces whether actor and partner associations operate
  through either partner's mediator [@ledermannAssessingMediationDyadic2011].

- **Dyadic response surface analysis:** Tests whether partner similarity,
  discrepancy, or particular combinations of scores relate to an outcome
  [@schonbrodtTestingSimilarityEffects2018].

- **One-with-many and Social Relations models:** Extend dyadic analysis to
  repeated relationships with several partners or round-robin groups, separating
  person and relationship components
  [@kenny2006dyadic; @kennySocialRelationsModel1984].

- **Network models:** Explain the formation and evolution of ties within an
  interconnected group, including reciprocity, clustering, and actor-driven
  network change [@hollandExponentialFamilyDirected1981;
  @snijdersStatisticalEvaluationNetwork2001].

<!-- ca. 1 minute -->

```{r}
#| label: fit-truth-bias
#| include: false

truth_bias_fit_file <- if (
    file.exists(file.path("R", "fit-truth-bias.R"))) {
  file.path("R", "fit-truth-bias.R")
} else {
  file.path("dev", "workshop", "R", "fit-truth-bias.R")
}

sys.source(
  normalizePath(truth_bias_fit_file),
  envir = knitr::knit_global()
)

truth_bias_result <- fit_truth_bias(pilot_aggregated)
```

## The Truth and Bias model [[@kennyAccuracyBiasPerception2001; @westTruthBiasModel2011]]{.heading-reference} {#the-truth-and-bias-model}

```{r}
#| label: fitted-truth-bias
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 90%
#| fig-align: center
#| fig-alt: "Fitted dyadic Truth and Bias model for provided and received emotional support. Estimates and significance stars annotate female- and male-specific assumed-similarity, direct-accuracy, and directional-bias paths. The partners' provided-support correlation represents actual similarity, and received-support judgment residuals covary."

draw_truth_bias_diagram(truth_bias_result)
```

::: {.text-smaller}

- The partner's report is the researcher-defined “truth”, not necessarily 
an objective truth. 

- After truth-mean centering, $b_0$ is **directional bias**.

:::

:::{.notes}

- Received support is the judgment; the partner's provided support is the
  operational “truth”; one's own provision is the assumed-similarity predictor.
- **Intercept:** At average provision by both partners, does the perceiver report
  receiving more or less than the partner reports providing?
- **Slopes:** Does perceived support track the partner's provision (direct
  accuracy) or one's own provision (assumed similarity)?
- The correlation between partners' provision is actual similarity, which lets
  assumed similarity contribute indirectly to accuracy.
- Here, assumed similarity was strong; direct accuracy was positive for female
  perceivers but not evident for male perceivers. Treat this small,
  cross-sectional example descriptively.

:::

<!-- ca. 2 minutes -->

```{r}
#| label: fit-dyadic-response-surface
#| include: false

drsa_fit_file <- if (
    file.exists(file.path("R", "fit-dyadic-response-surface.R"))) {
  file.path("R", "fit-dyadic-response-surface.R")
} else {
  file.path("dev", "workshop", "R", "fit-dyadic-response-surface.R")
}

drsa_draw_file <- if (
    file.exists(file.path("R", "draw-dyadic-response-surface.R"))) {
  file.path("R", "draw-dyadic-response-surface.R")
} else {
  file.path("dev", "workshop", "R", "draw-dyadic-response-surface.R")
}

sys.source(normalizePath(drsa_fit_file), envir = knitr::knit_global())
sys.source(normalizePath(drsa_draw_file), envir = knitr::knit_global())

drsa_result <- fit_dyadic_response_surface(pilot_aggregated)

```

## Dyadic response surface analysis [[@schonbrodtTestingSimilarityEffects2018]]{.heading-reference}

```{r}
#| label: fitted-dyadic-response-surface
#| echo: false
#| dev: svglite
#| fig-width: 12
#| fig-height: 4.25
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Two fitted response surfaces from one joint dyadic polynomial model. Each shows predicted female or male average daily total MVPA as a function of female-provided and male-provided emotional support centered at the pooled mean. Points and loops mark the observed couples and predictor region; the solid blue line is the line of congruence and the dashed purple line is the line of incongruence."

draw_dyadic_response_surface(drsa_result)
```

::: {.text-smaller}

- **[Line of incongruence]{style="color: #7B3FA1"}** curves upward; no clear slope or curvature along the **[line of congruence]{style="color: #1769AA"}**.

- **Interpretation:** No similarity-benefit pattern; instead, some evidence of a dissimilarity pattern.


:::

:::{.notes}

- Each surface predicts one partner's MVPA. Higher areas mean more MVPA; a
  slope shows change, and a bend shows that the association itself changes.
- **Blue line (congruence):** Both partners report the same support. Moving
  along it compares low-low with high-high couples.
- **Dashed purple line (incongruence):** One partner reports above-average and
  the other equally below-average support. Moving away from its centre means
  greater disagreement.
- Here, the purple line curves upward for both outcomes: larger discrepancies
  are associated with more MVPA. This is a dissimilarity pattern, not evidence
  that dissimilarity causes a benefit.
- Points and loops show the observed data region. With 36 couples, especially
  avoid interpreting the surface far outside it.

:::

<!-- ca. 2 minutes -->



<!----------------- APPLIED PART (plus MLM vs. SEM and software) 

TODO: Probably create a second slide deck!

- DHARMa plots!?! -> as required package for workshop -> link to materials or do in tutorial? Also check AR1 etc., etc. can dharma do this if resid is set to almost zero?

- glmmTMB -> as required 
- dyadMLM -> as required (via pak, see instructions; or should we already try a CRAN release now?? We may want to still add additional functions like rotating variance structs back, so probably not). -> maybe just require pak and we install newest version on workshop-day?
- dplyr -> required
- tibble -> required
- lavaan -> required
- ggplot2 -> required

-------------->

<!-- ILD-part: 

- see dev note on ar1 processes careful to consider degrees of freedom! relevant here?
  - TODO @codex: are degrees of freedom alreaedy important in the cross-sectional case and should we teach anything regarding those?
- see plots on: https://quantdev.ssri.psu.edu/tutorials/apa-ati-intensive-longitudinal-data-session-m-dyadic-multilevel-model

-->

<!-- Excercises part

Create exercises with data visalization, etc. Panel plots, histograms
correlations (wb corr)?

-->

## References

::: {#refs}
:::

## Continue with the applied tutorial {.rq-slide}

::: {.workshop-next}

Thank you for your attention!

[Continue to Part II: Applied Tutorial →](https://pascal-kueng.github.io/dyadMLM/workshop/applied/){.workshop-next-link}
:::



# Appendix

## Interpreting the distinguishable APIM [← Results](#fitting-the-distinguishable-apim){.appendix-back-link} {.smaller .interpretation-slide}

::: {.interpretation-footnote}
Red estimates are not statistically distinguishable from zero at
$\alpha = .05$. Their numerical translations are shown only to demonstrate how
APIM coefficients are interpreted; they should not be treated as evidence of a
nonzero effect.
:::

**Intercepts:** When both partners' support scores equal the pooled sample mean:

- $b_{0,\mathrm{F}}$: Predicted female MVPA is 35.28 minutes.
- $b_{0,\mathrm{M}}$: Predicted male MVPA is 31.24 minutes.

**Female MVPA**, holding the male partner's support constant:

- [$a_{\mathrm{F}}$: One-unit higher female support is associated with 4.61 fewer minutes of her own MVPA.]{.not-significant}
- $p_{\mathrm{F}}$: One-unit higher male support is associated with 15.06 additional minutes of female MVPA.

**Male MVPA**, holding the female partner's support constant:

- $a_{\mathrm{M}}$: One-unit higher male support is associated with 17.92 additional minutes of his own MVPA.
- [$p_{\mathrm{M}}$: One-unit higher female support is associated with 9.87 fewer minutes of male MVPA.]{.not-significant}

**Residual correlation ($\rho = .78$):** When the female's MVPA is higher than predicted, the male's MVPA also tends to be higher than predicted, and likewise for being lower. This is the remaining within-couple interdependence after accounting for both support scores.

## Interpreting the exchangeable APIM [← Results](#fitting-the-exchangeable-apim){.appendix-back-link} {.smaller .interpretation-slide}

::: {.interpretation-footnote}
Red estimates are not statistically distinguishable from zero at
$\alpha = .05$. Their numerical translations are shown only to demonstrate how
APIM coefficients are interpreted; they should not be treated as evidence of a
nonzero effect.
:::

The intercept, actor effect, partner effect, and residual SD are **pooled across members**.

**Pooled intercept:** When both partners' support scores equal the pooled sample mean:

- $b_0$: Predicted MVPA is 31.21 minutes for either member position.

**Pooled effects:** For either member, holding the other support score constant:

- $a$: One-unit higher own support is associated with 7.75 additional minutes of own MVPA.
- [$p$: One-unit higher partner support is associated with 1.97 additional minutes of own MVPA.]{.not-significant}

Pooling means that the same coefficients are used for both member positions. It does not require the two members to have identical predictions because their own and partner support scores may differ.

**Residual correlation ($\rho = .77$):** When one member's MVPA is higher than predicted, the partner's MVPA also tends to be higher than predicted, and likewise for being lower. The residual correlation remains freely estimated even though the other parameters are pooled.

## Interpreting the fitted DIM [← Results](#fitting-the-dim){.appendix-back-link} {.smaller .interpretation-slide}

::: {.interpretation-footnote}
Red estimates are not statistically distinguishable from zero at
$\alpha = .05$. Their numerical translations are shown only to demonstrate how
DIM coefficients are interpreted; they should not be treated as evidence of a
nonzero effect.
:::

**Intercepts:** For a couple with an average support mean:

- $b_0$: Predicted couple-mean MVPA is 31.21 minutes.
- The member-deviation intercept is constrained to zero: a member with no support deviation has no predicted MVPA deviation from the couple mean.

Holding the **members' support deviations fixed**, a one-unit higher couple-mean support score is associated with:

- $b_{\mathrm{mean}}$: 9.72 additional minutes of couple-mean MVPA.

Holding couple-mean support at the **sample average**, a one-unit larger member support deviation from the couple mean is associated with:

- [$b_{\mathrm{dev}}$: A 5.78-minute larger member MVPA deviation from the couple mean. Thus, a member whose support is one unit above the couple mean is predicted to have MVPA 5.78 minutes above the couple mean.]{.not-significant}

**Residual correlation:** Exchangeability constrains the correlation between the residual couple mean and residual member deviation to zero. Thus, couples whose MVPA is higher than predicted on average are not expected to tilt systematically toward either arbitrarily labeled member in residual MVPA. This is a model constraint, not an empirical finding, and it does not imply that the two partners' residuals are independent.

## Interpreting the fitted DSM [← Results](#fitted-dsm-results){.appendix-back-link} {.smaller .interpretation-slide}

::: {.interpretation-footnote}
Red estimates are not statistically distinguishable from zero at
$\alpha = .05$. Their numerical translations are shown only to demonstrate how
DSM coefficients are interpreted; they should not be treated as evidence of a
nonzero effect.
:::

**Intercepts:** For a couple with an average support mean and no female–male support difference:

- $a_{10}$: Predicted couple-mean MVPA is 33.26 minutes.
- [$a_{20}$: The female is predicted to be 4.04 minutes more active than the male.]{.not-significant}

Holding the **female–male support difference at zero**, a one-unit higher couple-mean support score is associated with:

- $a_{11}$: 9.25 additional minutes of couple-mean MVPA.
- [$a_{21}$: A 2.40-minute larger female–male MVPA difference. Thus, when couple-mean support is one unit above the sample average, the predicted difference is $4.04 + 2.40 = 6.44$ minutes.]{.not-significant}

Holding couple-mean support at the **sample average**, a one-unit larger female–male support difference is associated with:

- $a_{12}$: 11.86 fewer minutes of couple-mean MVPA.
- [$a_{22}$: A 4.06-minute larger female–male MVPA difference. Thus, when the female–male support difference is one unit, the predicted difference is $4.04 + 4.06 = 8.10$ minutes.]{.not-significant}

**Residual correlation ($\rho = .37$):** Couples with higher-than-predicted mean MVPA also tend to show a more positive-than-predicted female–male MVPA difference; residual MVPA therefore tilts toward the female.

## Exchangeable CFM: constraints and degrees of freedom {.smaller}

Positions 1 and 2 are arbitrary. The model must be unchanged when they are
swapped [@ledermannMediationDyadicData2009; @peughAnalyzingMixedDyadic2013].

$$
\begin{aligned}
\nu_{X_1} &= \nu_{X_2}, & \nu_{Y_1} &= \nu_{Y_2} \\
\sigma^2_{\epsilon_{X_1}} &= \sigma^2_{\epsilon_{X_2}}, &
\sigma^2_{\epsilon_{Y_1}} &= \sigma^2_{\epsilon_{Y_2}} \\
\operatorname{Cov}(\epsilon_{X_1},\epsilon_{Y_1})
&= \operatorname{Cov}(\epsilon_{X_2},\epsilon_{Y_2})
\end{aligned}
$$

- All four loadings remain fixed to 1.
- Ordinary wide-data **df = 6**; **df = 0** against the exchangeable saturated
  model.
- Usual exchangeability test: $\Delta df = 5$; here: $\Delta df = 4$
  ($\sigma^2_{\epsilon_Y}$ is already equal).

:::{.notes}

- Equality is imposed on the two same-member residual covariances. Because the
  corresponding residual variances are also equal, the residual correlations
  are equal as a consequence.
- `lavaan` reports 6 degrees of freedom because its default saturated reference
  model is fully unrestricted. Six observed-moment restrictions are required
  for exchangeability.
- The exchangeable CFM reproduces the exchangeable saturated model, so it has
  zero degrees of freedom relative to that appropriate reference model. Its
  ordinary chi-square is therefore a test of exchangeability, not an additional
  test of the basic CFM structure.
- An unconstrained distinguishable CFM has df = 1, giving the usual five-degree-
  of-freedom exchangeability test. Our distinguishable model already equates the
  two outcome residual variances and has df = 2, so the comparison tests the four
  remaining equality constraints.

:::

## Distinguishable APIM: residual covariance

### Within-dyad residual covariance block (shared across dyads):
$$
\operatorname{Cov}
\begin{pmatrix}
\epsilon_{Fi} \\
\epsilon_{Mi}
\end{pmatrix}
= \boldsymbol{\Sigma}_{\epsilon}
= \begin{bmatrix}
\sigma_{\epsilon_F}^{2}
& \rho_{\epsilon_F\epsilon_M}\sigma_{\epsilon_F}\sigma_{\epsilon_M} \\
\rho_{\epsilon_F\epsilon_M}\sigma_{\epsilon_F}\sigma_{\epsilon_M}
& \sigma_{\epsilon_M}^{2}
\end{bmatrix}
$$

### Residual covariance matrix for all dyads (first three shown):
$$
\boldsymbol{\Sigma}_{\mathrm{model}}
= \begin{bmatrix}
\boldsymbol{\Sigma}_{\epsilon} & \boldsymbol{0} & \boldsymbol{0} & \cdots \\
\boldsymbol{0} & \boldsymbol{\Sigma}_{\epsilon} & \boldsymbol{0} & \cdots \\
\boldsymbol{0} & \boldsymbol{0} & \boldsymbol{\Sigma}_{\epsilon} & \cdots \\
\vdots & \vdots & \vdots & \ddots
\end{bmatrix}
$$

<!-- ca. 1.5 minutes -->

## Exchangeable APIM: residual covariance

For arbitrary member positions 1 and 2, the residual variances must be equal:

$$
\operatorname{Cov}
\begin{pmatrix}
\epsilon_{1i} \\
\epsilon_{2i}
\end{pmatrix}
= \boldsymbol{\Sigma}_{\epsilon,\mathrm{exch}}
= \begin{bmatrix}
\sigma_{\epsilon}^{2}
& \rho_{\epsilon_1\epsilon_2}\sigma_{\epsilon}^{2} \\
\rho_{\epsilon_1\epsilon_2}\sigma_{\epsilon}^{2}
& \sigma_{\epsilon}^{2}
\end{bmatrix}
$$

- The equal diagonal elements express exchangeability.
- The off-diagonal element still allows residual interdependence.
- Across dyads, these blocks form the same block-diagonal model matrix as on
  the previous slide.

<!-- ca. 1.5 minutes -->

## DIM: residual covariance

$$
\begin{aligned}
r_{\mathrm{M}i}
&= \bar{\epsilon}_i
= \frac{\epsilon_{1i}+\epsilon_{2i}}{2}, \\
r_{\mathrm{D},mi}
&= \epsilon_{mi}-r_{\mathrm{M}i}
\quad (m=1,2),
\qquad
r_{\mathrm{D},1i}
= \frac{\epsilon_{1i}-\epsilon_{2i}}{2}
= -r_{\mathrm{D},2i}.
\end{aligned}
$$

Using the arbitrary orientation $r_{\mathrm{D}i}=r_{\mathrm{D},1i}$,
exchangeability gives:

$$
\operatorname{Cov}
\begin{pmatrix}
r_{\mathrm{M}i} \\
r_{\mathrm{D}i}
\end{pmatrix}
= \begin{bmatrix}
\sigma_{r_\mathrm{M}}^{2} & 0 \\
0 & \sigma_{r_\mathrm{D}}^{2}
\end{bmatrix}.
$$

Back-transform to the exchangeable APIM: partner covariance need not be zero.

::: {.covariance-backtransform .dim-backtransform}
$$
\begin{bmatrix}
\color{#1769AA}{\sigma_{r_\mathrm{M}}^{2}+\sigma_{r_\mathrm{D}}^{2}}
& \color{#7B3FA1}{\sigma_{r_\mathrm{M}}^{2}-\sigma_{r_\mathrm{D}}^{2}} \\[0.35em]
\color{#7B3FA1}{\sigma_{r_\mathrm{M}}^{2}-\sigma_{r_\mathrm{D}}^{2}}
& \color{#1769AA}{\sigma_{r_\mathrm{M}}^{2}+\sigma_{r_\mathrm{D}}^{2}}
\end{bmatrix}
=
\begin{bmatrix}
\color{#1769AA}{\sigma_{\epsilon}^{2}}
& \color{#7B3FA1}{\rho_{\epsilon_1\epsilon_2}\sigma_{\epsilon}^{2}} \\[0.35em]
\color{#7B3FA1}{\rho_{\epsilon_1\epsilon_2}\sigma_{\epsilon}^{2}}
& \color{#1769AA}{\sigma_{\epsilon}^{2}}
\end{bmatrix}.
$$
:::

<!-- ca. 2 minutes -->

## DSM: residual covariance

With distinguishable roles, the directed difference may covary with the mean:

$$
r_{\mathrm{M}i}=\frac{\epsilon_{Fi}+\epsilon_{Mi}}{2},
\qquad
r_{\mathrm{D}i}=\epsilon_{Fi}-\epsilon_{Mi}.
$$

$$
\operatorname{Cov}
\begin{pmatrix}
r_{\mathrm{M}i} \\
r_{\mathrm{D}i}
\end{pmatrix}
= \begin{bmatrix}
\sigma_{r_\mathrm{M}}^{2}
& \rho_{r_\mathrm{M}r_\mathrm{D}}
  \sigma_{r_\mathrm{M}}\sigma_{r_\mathrm{D}} \\
\rho_{r_\mathrm{M}r_\mathrm{D}}
  \sigma_{r_\mathrm{M}}\sigma_{r_\mathrm{D}}
& \sigma_{r_\mathrm{D}}^{2}
\end{bmatrix}.
$$

Back-transform to the distinguishable APIM, with
$c_{\mathrm{MD}}
=\rho_{r_\mathrm{M}r_\mathrm{D}}
\sigma_{r_\mathrm{M}}\sigma_{r_\mathrm{D}}$.

::: {.covariance-backtransform}
$$
\begin{bmatrix}
\color{#1769AA}{
  \sigma_{r_\mathrm{M}}^{2}+\frac{1}{4}\sigma_{r_\mathrm{D}}^{2}+c_{\mathrm{MD}}}
& \color{#7B3FA1}{
  \sigma_{r_\mathrm{M}}^{2}-\frac{1}{4}\sigma_{r_\mathrm{D}}^{2}} \\[0.35em]
\color{#7B3FA1}{
  \sigma_{r_\mathrm{M}}^{2}-\frac{1}{4}\sigma_{r_\mathrm{D}}^{2}}
& \color{#B45309}{
  \sigma_{r_\mathrm{M}}^{2}+\frac{1}{4}\sigma_{r_\mathrm{D}}^{2}-c_{\mathrm{MD}}}
\end{bmatrix}
=
\begin{bmatrix}
\color{#1769AA}{\sigma_{\epsilon_F}^{2}}
& \color{#7B3FA1}{
  \rho_{\epsilon_F\epsilon_M}\sigma_{\epsilon_F}\sigma_{\epsilon_M}} \\[0.35em]
\color{#7B3FA1}{
  \rho_{\epsilon_F\epsilon_M}\sigma_{\epsilon_F}\sigma_{\epsilon_M}}
& \color{#B45309}{\sigma_{\epsilon_M}^{2}}
\end{bmatrix}.
$$
:::

<!-- ca. 2 minutes -->
