---
title: "Dyadic Data Analysis with R"
subtitle: "From Cross-Sectional Models to Intensive Longitudinal Data"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
csl: apa.csl
resources:
  - uzh-logo.svg
  - charite-logo.svg
format:
  revealjs:
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: true
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup-pilot-data
#| include: false

library(dplyr)
library(ggplot2)

diagram_helper_candidates <- c(
  "diagram-helpers.Rinc",
  file.path("vignettes", "diagram-helpers.Rinc"),
  file.path("..", "..", "vignettes", "diagram-helpers.Rinc")
)
diagram_helper <- diagram_helper_candidates[
  file.exists(diagram_helper_candidates)
][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(
  normalizePath(diagram_helper),
  envir = knitr::knit_global()
)


# FUNCTION TO FIT EARLY MODELS --------------------------------------

early_models_file <- if (file.exists(file.path("R", "fit-early-models.R"))) {
  file.path("R", "fit-early-models.R")
} else {
  file.path("dev", "workshop", "R", "fit-early-models.R")
}

sys.source(
  normalizePath(early_models_file),
  envir = knitr::knit_global()
)

data_file <- if (file.exists("long.rds")) "long.rds" else "dev/workshop/long.rds"

pilot <- readRDS(data_file) |>
  select(
    coupleID, userID, gender, ss_pa, ss_pa_min_solo, ss_pa_min_collab,
    ss_emo_pleasure, sc_actionplan
  ) |>
  transmute(
    couple_id = coupleID,
    person_id = userID,
    role = factor(gender, levels = c(1, 2), labels = c("female", "male")),
    provided_emotional_support = ss_emo_pleasure,
    collaborative_planning = sc_actionplan,
    mvpa_solo = if_else(ss_pa == 0, 0, ss_pa_min_solo),
    mvpa_joint = if_else(ss_pa == 0, 0, ss_pa_min_collab),
    mvpa_total = mvpa_solo + mvpa_joint
  )

distribution_plot <- function(data, variables, bin_widths) {
  labels <- c(
    provided_emotional_support = "Provided emotional support",
    collaborative_planning = "Collaborative planning (daily mean)",
    mvpa_solo = "Mean solo MVPA (minutes/day)",
    mvpa_joint = "Mean joint MVPA (minutes/day)",
    mvpa_total = "Mean total MVPA (minutes/day)"
  )

  missing_widths <- setdiff(variables, names(bin_widths))
  if (length(missing_widths)) {
    stop("Missing bin widths for: ", paste(missing_widths, collapse = ", "))
  }

  plots <- Map(\(variable, bin_width) {
    values <- data[[variable]]
    values <- values[is.finite(values)]
    limits <- c(
      floor(min(values) / bin_width) * bin_width,
      ceiling(max(values) / bin_width) * bin_width
    )
    bandwidth <- density(values, adjust = 1.1)$bw
    curve <- density(
      c(values, -values), bw = bandwidth, from = 0, to = limits[2]
    )
    curve <- tibble(
      value = curve$x,
      count = curve$y * 2 * length(values) * bin_width
    )

    ggplot(tibble(value = values), aes(value)) +
      geom_histogram(
        binwidth = bin_width, boundary = limits[1], closed = "right",
        fill = "steelblue", color = "white", linewidth = 0.25
      ) +
      geom_line(
        data = curve, aes(value, count), inherit.aes = FALSE,
        color = "navy", linewidth = 1
      ) +
      labs(title = labels[[variable]], x = NULL, y = "Number of participants") +
      theme_minimal(base_size = 15) +
      theme(
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text = element_text(color = "#222222")
      )
  }, variables, unname(bin_widths[variables]))

  patchwork::wrap_plots(
    plots,
    ncol = min(3, length(plots)),
    axis_titles = "collect_y"
  )
}
```

## Introduction

- Turu & Pascal
- Tell us about you and your dyadic data!
  - Name, affiliation & career stage
  - Your dyadic data and RQ
  
<!-- ca. 15 Minutes -->

## Today's Aims and Program

| Time | Focus |
|---|---|
| 11:00–12:45 | Introduction and different types of dyadic models |
| 12:45–14:00 | *Lunch break* |
| 14:00–15:30 | Estimating cross-sectional models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

<!-- ca. 2.5 Minutes -->

## Why dyadic data?

> “Much of what we think, do, and feel is determined not by who we are but by
> whom we are with.” [@kennyPartnerEffectsRelationship1999, p. 446]

::: {.incremental}
- **Partners matter:** An outcome may depend on both *my characteristics* and 
*my partner's characteristics*.
- **Some constructs and processes are dyadic:** Shared contexts and interdependence
processes can make partners' outcomes correlated.
- **The configuration matters:** Partners' similarity, or complementarity (differences) 
may predict outcomes beyond either individual score.
:::

[@kennyPartnerEffectsRelationship1999;
@iidaModelsInterdependentIndividuals2018]{.subtle}

<!-- ca. 2.5 Minutes -->

## Introducing today's dyadic dataset!

- 38 physically inactive opposite-sex romantic couples
- Both partners intended to increase their physical activity together
- Smartphone-based planning interventions and JITAIs
- Daily diaries over 55 days

[@kuengDailyAssociationsSocialControl2026;
@hoehenerEffectivenessSmartphoneDyadic2026]{.subtle}

<!-- ca. 1 Minute -->

## Research Question 1 {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ1:]{.rq-label} [In romantic couples, how is provided health-related social support linked to physical activity?]{.rq-text}
:::
:::

:::{.notes}
This is intentionally broad, since we will look at it through different lenses
and analyze this with different models. 

Basically answering differt sub-RQs each time that I will introduce.
:::

## Measurement - Predictor

### Emotional support provided

> **Rough English translation of the original German item**<br>
> “How much did you emotionally support your partner today so that they would
be physically active with greater enjoyment?”
>
> [Adapted from @bolgerInvisibleSupportAdjustment2000]{.subtle}

0 = *not at all today* <br>
5 = *very much today*<br>

<!-- ca. 1 Minute -->

## Measurement - Outcome

### Moderate to vigorous physical activity (MVPA)

> **Rough English translation of the original German item**<br>
> “How many minutes did you spend engaging in moderate-to-vigorous physical
activity today?” 
[Adapted from @amireaultGodinShephardQuestionnaire2015]{.subtle}

- **Alone:** _____ minutes
- **Together with your partner:** _____ minutes

[**Total MVPA = solo MVPA + joint MVPA.**]{.fragment}

<!-- ca. 1 Minute -->

## Aggregating for cross-sectional examples

To show simple dyadic models we aggregate the daily diary data to
obtain a "cross-sectional" dataset.

```{r}
library(dplyr)

pilot_aggregated <- pilot |>
  group_by(couple_id, person_id, role) |>
  summarise(
    across(where(is.numeric), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )
```

::: {.fragment}
The variables now represent person-level daily averages across the study:

- `provided_emotional_support`: average daily support provided
- `mvpa_total`: average daily total MVPA (minutes)
:::

<!-- ca. 1 Minute -->

## Pilot data: APIM variables

```{r}
#| label: pilot-aggregated-distributions-plot
#| echo: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Histograms of provided emotional support and person-level mean total MVPA."

distribution_plot(
  pilot_aggregated,
  variables = c("provided_emotional_support", "mvpa_total"),
  bin_widths = c(
    provided_emotional_support = 0.25,
    mvpa_total = 7.5
  )
)
```

<!-- ca. 1 Minute -->

## Many models to choose from

**Various dyadic models** can be used, each answering different questions!

::: {.incremental}
- **Actor-partner interdependence model (APIM)**: interdependent individuals
- **Dyad-individual model (DIM)** and **dyadic score model (DSM)**: within- and between dyad associations
- **Common fate model (CFM)**: latent dyadic process
:::

::: {.fragment}
Other approaches include **mutual-influence models**, which represent reciprocal
influence between partners [@iidaModelsInterdependentIndividuals2018, p. 63].
:::

<!-- ca. 3 Minutes -->

::: {.notes}
"For RQ 1 we first go with the Actor-Partner interdependence model and 
see how we can use it to answer our research question (in what way). "
:::

## Many models to choose from

**Various dyadic models** can be used, each answering different questions!

- [**Actor-partner interdependence model (APIM)**: interdependent individuals]{.text-highlight}
- **Dyad-individual model (DIM)** and **dyadic score model (DSM)**: within- and between dyad associations
- **Common fate model (CFM)**: latent dyadic process

Other approaches include **mutual-influence models**, which represent reciprocal
influence between partners [@iidaModelsInterdependentIndividuals2018, p. 63].

<!-- ca. 3 Minutes -->

## The distinguishable APIM - Conceptually

```{r}
#| label: conceptual-distinguishable-apim-paths-only
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with actor and partner paths."

draw_apim_diagram("distinguishable", show_residuals = FALSE)
```

::: {.notes}

Explain actor and partner effects only and briefly. Not talking about residuals yet. 
-> to set up next slide

:::

<!-- ca. 2 Minutes -->

## Can we just estimate separate regression equations for each partner?

Do two separate models like this work?


$$
  \begin{gathered}
  \\[1em]
  Y_{male_i}
  = b_{0M} + b_{1M}X_{male_i} + b_{2M}X_{female_i} + \epsilon_{male_i}
  \\[2em]
  Y_{female_i}
  = b_{0F} + b_{1F}X_{female_i} + b_{2F}X_{male_i} + \epsilon_{female_i}
  \end{gathered}
$$


<!-- ca. 1 minute 

glmmTMB(
  outcome_male ~ 1 + X_male + X_female,
  
  data = data,
  family = gaussian()
) 

glmmTMB(
  outcome_female ~ 1 + X_male + X_female,
  
  data = data,
  family = gaussian()
) 

-->

## Can we just estimate separate regression equations for each partner?

**Yes! But then we: **

:::{.incremental}
- can't test whether the two actor effects or partner effects are equal
- can't pool equal paths into a shared estimate, which can increase power
- ignore unmodelled sources of interdependence between partners
  - joint modelling tells us how much interdependence remains after accounting for the predictors
  - we can find out how much covariance is explained by our predictors

:::

[@kennyPartnerEffectsRelationship1999, pp. 439–440;
@iidaModelsInterdependentIndividuals2018, pp. 68–69]{.subtle}

:::{.notes}

The reduction in residual covariance indicates how much of the original 
within-dyad covariance is accounted for by the predictors.
  
:::

<!-- ca. 3 minutes -->

:::{.notes}
Thus, we should model non-explained non-independence! 
E.g., correlated residuals.
:::

## The distinguishable APIM

```{r}
#| label: conceptual-distinguishable-apim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with role-specific actor and partner paths."

draw_apim_diagram("distinguishable")
```

:::{.notes}
Explain correlated residuals and explain residual variances and how they have 
a variance that can be different, and why. 
:::

<!-- ca. 2 Minutes 

Should we have a slide that explains residual covariances visually?
And correlated residuals?

-->

## Fitting the APIM

```{r}
#| label: fitted-distinguishable-apim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted distinguishable APIM for provided support and total MVPA."

fit_and_draw_distinguishable_apim(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total
)
```


<!------------- BELOW HERE IS JUST STRUCTURE  -------------------->

## The exchangeable APIM

Sometimes: 

- partners may not be theoretically distinguishable
- we may have mixed dyad types that we want to pool
- we want to test whether partners are meaningfully distinguishable

$\rightarrow$ Then, we may fit an exchangeable model that imposes symmetry

## The exchangeable APIM

```{r}
#| label: conceptual-exchangeable-apim-paths-only
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual exchangeable APIM with pooled actor and partner paths."

draw_apim_diagram("exchangeable", show_residuals = FALSE)
```

## The exchangeable APIM

```{r}
#| label: conceptual-exchangeable-apim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual exchangeable APIM with pooled actor and partner paths."

draw_apim_diagram("exchangeable")
```

## Interpreting the fit of the exchangeable APIM

```{r}
#| label: fitted-exchangeable-apim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted exchangeable APIM for provided support and total MVPA."

fit_and_draw_exchangeable_apim(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total
)
```

## The DIM: another parameterization of an exchangeable APIM

```{r}
#| label: conceptual-dim-paths-only
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DIM with dyad-mean and member-deviation paths."

draw_dim_diagram(show_residuals = FALSE)
```

## The DIM: another parameterization of an exchangeable APIM

```{r}
#| label: conceptual-dim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DIM separating dyad-mean and member-deviation effects."

draw_dim_diagram()
```

## Interpreting the fit of the DIM

```{r}
#| label: fitted-dim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted DIM for provided support and total MVPA."

fit_and_draw_dim(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total
)
```

## The DSM: a mean-and-difference parameterization of the distinguishable APIM

```{r}
#| label: conceptual-dsm-paths-only
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DSM with paths between predictor and outcome scores."

draw_dsm_diagram(show_residuals = FALSE)
```

## The DSM: a mean-and-difference parameterization of the distinguishable APIM

```{r}
#| label: conceptual-dsm
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DSM linking predictor mean and difference to outcome mean and difference."

draw_dsm_diagram()
```

## Interpreting the fit of the DSM

```{r}
#| label: fitted-dsm
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted DSM for provided support and total MVPA."

fit_and_draw_dsm(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total
)
```



## Research Question 2

RQ2: In romantic couples, how is collaborative planning linked to physical 
activity?

## Measurement - CFM predictor

### Dyadic predictor: collaborative planning

> **Rough english translation of the original German item**<br>
> “We have already planned precisely for tomorrow when, where, and how we will
be physically active together.”
[Adapted from @scholzBeyondBehaviouralIntentions2008]{.subtle}

0 = *not at all true today*<br>
to 5 = *completely true today*

We aggregate each person's ratings into their **average daily collaborative
planning** across the 55 days.

## Dyadic outcome: joint physical activity



```{r}
#| label: pilot-aggregated-distributions-cfm-plot
#| echo: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Histograms of provided collaborative planning and person-level mean joint MVPA."

distribution_plot(
  pilot_aggregated,
  variables = c("collaborative_planning", "mvpa_joint"),
  bin_widths = c(
    collaborative_planning = 0.25,
    mvpa_joint = 7.5
  )
)
```

## The CFM

```{r}
#| label: conceptual-cfm-paths-only
#| echo: false
#| fig-width: 10
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual Common Fate Model with shared latent levels."

draw_cfm_diagram(show_residuals = FALSE)
```

## The CFM

```{r}
#| label: conceptual-cfm
#| echo: false
#| fig-width: 10
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual Common Fate Model with shared latent predictor and outcome levels."

draw_cfm_diagram()
```



<!-- ILD-part: see dev note on ar1 processes -->


## References
