---
title: "Dyadic Data Analysis with R"
subtitle: "From Cross-Sectional Models to Intensive Longitudinal Data"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
csl: apa.csl
resources:
  - uzh-logo.svg
  - charite-logo.svg
format:
  revealjs:
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: true
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup-pilot-data
#| include: false

library(dplyr)
library(ggplot2)

diagram_helper_candidates <- c(
  "diagram-helpers.Rinc",
  file.path("vignettes", "diagram-helpers.Rinc"),
  file.path("..", "..", "vignettes", "diagram-helpers.Rinc")
)
diagram_helper <- diagram_helper_candidates[
  file.exists(diagram_helper_candidates)
][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(
  normalizePath(diagram_helper),
  envir = knitr::knit_global()
)

data_file <- if (file.exists("long.rds")) "long.rds" else "dev/workshop/long.rds"

pilot <- readRDS(data_file) |>
  select(
    coupleID, userID, ss_pa, ss_pa_min_solo, ss_pa_min_collab,
    ss_emo_pleasure, sc_actionplan
  ) |>
  transmute(
    couple_id = coupleID,
    person_id = userID,
    provided_emotional_support = ss_emo_pleasure,
    collaborative_planning = sc_actionplan,
    mvpa_solo = if_else(ss_pa == 0, 0, ss_pa_min_solo),
    mvpa_joint = if_else(ss_pa == 0, 0, ss_pa_min_collab),
    mvpa_total = mvpa_solo + mvpa_joint
  )

distribution_plot <- function(data, variables, bin_widths) {
  labels <- c(
    provided_emotional_support = "Provided emotional support",
    collaborative_planning = "Collaborative planning (daily mean)",
    mvpa_solo = "Mean solo MVPA (minutes/day)",
    mvpa_joint = "Mean joint MVPA (minutes/day)",
    mvpa_total = "Mean total MVPA (minutes/day)"
  )

  missing_widths <- setdiff(variables, names(bin_widths))
  if (length(missing_widths)) {
    stop("Missing bin widths for: ", paste(missing_widths, collapse = ", "))
  }

  plots <- Map(\(variable, bin_width) {
    values <- data[[variable]]
    values <- values[is.finite(values)]
    limits <- c(
      floor(min(values) / bin_width) * bin_width,
      ceiling(max(values) / bin_width) * bin_width
    )
    bandwidth <- density(values, adjust = 1.1)$bw
    curve <- density(
      c(values, -values), bw = bandwidth, from = 0, to = limits[2]
    )
    curve <- tibble(
      value = curve$x,
      count = curve$y * 2 * length(values) * bin_width
    )

    ggplot(tibble(value = values), aes(value)) +
      geom_histogram(
        binwidth = bin_width, boundary = limits[1], closed = "right",
        fill = "steelblue", color = "white", linewidth = 0.25
      ) +
      geom_line(
        data = curve, aes(value, count), inherit.aes = FALSE,
        color = "navy", linewidth = 1
      ) +
      labs(title = labels[[variable]], x = NULL, y = "Number of participants") +
      theme_minimal(base_size = 15) +
      theme(
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text = element_text(color = "#222222")
      )
  }, variables, unname(bin_widths[variables]))

  patchwork::wrap_plots(
    plots,
    ncol = min(3, length(plots)),
    axis_titles = "collect_y"
  )
}
```

## Introduction

- Turu & Pascal
- Tell us about you and your dyadic data!
  - Name, Affiliation & Career stage
  - Your dyadic data and RQ
  
<!-- ca. 10 Minutes -->

## Today's Aims and Program

| Time | Focus |
|---|---|
| 11:00–12:45 | Introduction and different types of dyadic models |
| 12:45–14:00 | *Lunch break* |
| 14:00–15:30 | Estimating cross-sectional models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

<!-- ca. 5 Minutes -->

## Why dyadic data?

> “Much of what we think, do, and feel is determined not by who we are but by
> whom we are with.” [@kennyPartnerEffectsRelationship1999, p. 446]

- **Partners influence each other:** An outcome may depend on both *my
characteristics* and *my partner's characteristics*.
- **Some experiences belong to the couple:** Shared contexts and relationship
processes can describe *us*, not either partner alone.
- **The configuration matters:** Partners' similarity, mismatch, or
complementarity may predict outcomes beyond either individual score.

[@kennyPartnerEffectsRelationship1999;
@iidaModelsInterdependentIndividuals2018]{.subtle}

## Introducing today's dyadic dataset!

- 38 physically inactive opposite-sex romantic couples
- Both partners intended to increase their physical activity together
- Smartphone-based planning interventions and JITAIs
- Daily diaries over 55 days

[@kuengDailyAssociationsSocialControl2026;
@hoehenerEffectivenessSmartphoneDyadic2026]{.subtle}

<!-- ca. 5 Minutes -->

## Measurement - Predictor

### Emotional support provided

> **Rough English translation of the original German item**<br>
> “How much did you emotionally support your partner today so that they would
be physically active with greater enjoyment?”
>
> [Adapted from @bolgerInvisibleSupportAdjustment2000]{.subtle}

0 = *not at all today* <br>
5 = *very much today*<br>

## Measurement - Outcome

### Moderate to vigorous physical activity (MVPA)

> **Rough English translation of the original German item**<br>
> “How many minutes did you spend engaging in moderate-to-vigorous physical
activity today?” 
[Adapted from @amireaultGodinShephardQuestionnaire2015]{.subtle}

- **Alone:** _____ minutes
- **Together with your partner:** _____ minutes

**Total MVPA = solo MVPA + joint MVPA.**

## Aggregating for cross-sectional examples

To show simple dyadic models we aggregate the daily diary data to
obtain a "cross-sectional" dataset.

```{r}
library(dplyr)

pilot_aggregated <- pilot |>
  group_by(couple_id, person_id) |>
  summarise(
    across(everything(), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )
```

The variables now represent person-level daily averages across the study:

- `provided_emotional_support`: average daily support provided
- `mvpa_total`: average daily total MVPA (minutes)

## Why separate regressions are not enough

- **Omitted-variable bias:** Own and partner predictors are often correlated;
leaving either out can bias the other effect.
- **No direct comparisons:** Separate equations cannot test whether actor and
partner effects differ across partners—or from each other.
- **Less power:** Equal paths cannot be pooled into a common, more precise
estimate.
- **Residual interdependence is lost:** The remaining association between
partners' outcomes is not modeled.
- **Other dyadic processes remain hidden:** Shared couple levels and
within-couple differences require a CFM or DSM.

[@kennyPartnerEffectsRelationship1999;
@iidaModelsInterdependentIndividuals2018;
@laurenceauIntroductionIntensiveLongitudinal2026]{.subtle}

## Pilot data: APIM variables

```{r}
#| label: pilot-aggregated-distributions-plot
#| echo: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Histograms of provided emotional support and person-level mean total MVPA."

distribution_plot(
  pilot_aggregated,
  variables = c("provided_emotional_support", "mvpa_total"),
  bin_widths = c(
    provided_emotional_support = 0.25,
    mvpa_total = 7.5
  )
)
```


## Measurement - CFM predictor

### Dyadic predictor: collaborative planning

> **Rough english translation of the original German item**<br>
> “We have already planned precisely for tomorrow when, where, and how we will
be physically active together.”
[Adapted from @scholzBeyondBehaviouralIntentions2008]{.subtle}

0 = *not at all true today*<br>
to 5 = *completely true today*

We aggregate each person's ratings into their **average daily collaborative
planning** across the 55 days.

## Dyadic outcome: joint physical activity



```{r}
#| label: pilot-aggregated-distributions-cfm-plot
#| echo: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Histograms of provided collaborative planning and person-level mean joint MVPA."

distribution_plot(
  pilot_aggregated,
  variables = c("collaborative_planning", "mvpa_joint"),
  bin_widths = c(
    collaborative_planning = 0.25,
    mvpa_joint = 7.5
  )
)
```

## References
