---
title: "Dyadic Data Analysis with R"
subtitle: "From Cross-Sectional Models to Intensive Longitudinal Data"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
csl: apa.csl
resources:
  - uzh-logo.svg
  - charite-logo.svg
format:
  revealjs:
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: true
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

<!--
!TODO: 

1. Why dyadic data?
2. Example dataset and substantive question
3. Why not analyze each partner separately?
4. APIM graphic
5. Minimal conceptual APIM equations
6. Actor and partner effect interpretations
7. Distinguishable versus indistinguishable dyads
8. DIM
9. DSM and CFM as shorter alternatives
10. Model-choice recap
  
-->

```{r}
#| label: setup-pilot-data
#| include: false

library(dplyr)
library(ggplot2)

diagram_helper_candidates <- c(
  "diagram-helpers.Rinc",
  file.path("vignettes", "diagram-helpers.Rinc"),
  file.path("..", "..", "vignettes", "diagram-helpers.Rinc")
)
diagram_helper <- diagram_helper_candidates[
  file.exists(diagram_helper_candidates)
][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(
  normalizePath(diagram_helper),
  envir = knitr::knit_global()
)

data_file <- if (file.exists("long.rds")) "long.rds" else "dev/workshop/long.rds"

pilot <- readRDS(data_file) |>
  select(
    coupleID, userID, ss_pa, ss_pa_min_solo, ss_pa_min_collab,
    ss_emo_pleasure, sc_actionplan
  ) |>
  transmute(
    couple_id = coupleID,
    person_id = userID,
    provided_emotional_support = ss_emo_pleasure,
    collaborative_planning = sc_actionplan,
    mvpa_solo = if_else(ss_pa == 0, 0, ss_pa_min_solo),
    mvpa_joint = if_else(ss_pa == 0, 0, ss_pa_min_collab),
    mvpa_total = mvpa_solo + mvpa_joint
  )

distribution_plot <- function(data, variables, bin_widths) {
  labels <- c(
    provided_emotional_support = "Provided emotional support",
    collaborative_planning = "Collaborative planning (daily mean)",
    mvpa_solo = "Mean solo MVPA (minutes/day)",
    mvpa_joint = "Mean joint MVPA (minutes/day)",
    mvpa_total = "Mean total MVPA (minutes/day)"
  )

  missing_widths <- setdiff(variables, names(bin_widths))
  if (length(missing_widths)) {
    stop("Missing bin widths for: ", paste(missing_widths, collapse = ", "))
  }

  plots <- Map(\(variable, bin_width) {
    values <- data[[variable]]
    values <- values[is.finite(values)]
    limits <- c(
      floor(min(values) / bin_width) * bin_width,
      ceiling(max(values) / bin_width) * bin_width
    )
    bandwidth <- density(values, adjust = 1.1)$bw
    curve <- density(
      c(values, -values), bw = bandwidth, from = 0, to = limits[2]
    )
    curve <- tibble(
      value = curve$x,
      count = curve$y * 2 * length(values) * bin_width
    )

    ggplot(tibble(value = values), aes(value)) +
      geom_histogram(
        binwidth = bin_width, boundary = limits[1], closed = "right",
        fill = "steelblue", color = "white", linewidth = 0.25
      ) +
      geom_line(
        data = curve, aes(value, count), inherit.aes = FALSE,
        color = "navy", linewidth = 1
      ) +
      labs(title = labels[[variable]], x = NULL, y = "Number of participants") +
      theme_minimal(base_size = 15) +
      theme(
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text = element_text(color = "#222222")
      )
  }, variables, unname(bin_widths[variables]))

  patchwork::wrap_plots(
    plots,
    ncol = min(3, length(plots)),
    axis_titles = "collect_y"
  )
}
```

## Introduction

- Turu & Pascal
- Tell us about you and your dyadic data!
  - Name, Affiliation & Career stage
  - Your dyadic data and RQ
  
<!-- ca. 15 Minutes -->

## Today's Aims and Program

| Time | Focus |
|---|---|
| 11:00–12:45 | Introduction and different types of dyadic models |
| 12:45–14:00 | *Lunch break* |
| 14:00–15:30 | Estimating cross-sectional models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

<!-- ca. 2.5 Minutes -->

## Why dyadic data?

> “Much of what we think, do, and feel is determined not by who we are but by
> whom we are with.” [@kennyPartnerEffectsRelationship1999, p. 446]

::: {.incremental}
- **Partners matter:** An outcome may depend on both *my characteristics* and 
*my partner's characteristics*.
- **Some constructs and processes are dyadic:** Shared contexts and interdependence
processes can make partners' outcomes correlated.
- **The configuration matters:** Partners' similarity, or complementarity (differences) 
may predict outcomes beyond either individual score.
:::

[@kennyPartnerEffectsRelationship1999;
@iidaModelsInterdependentIndividuals2018]{.subtle}

<!-- ca. 2.5 Minutes -->

## Introducing today's dyadic dataset!

- 38 physically inactive opposite-sex romantic couples
- Both partners intended to increase their physical activity together
- Smartphone-based planning interventions and JITAIs
- Daily diaries over 55 days

[@kuengDailyAssociationsSocialControl2026;
@hoehenerEffectivenessSmartphoneDyadic2026]{.subtle}

<!-- ca. 1 Minute -->

## Research Question 1 {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ1:]{.rq-label} [In romantic couples, how is provided health-related social support linked to physical activity?]{.rq-text}
:::
:::

:::{.notes}
This is intentionally broad, since we will look at it through different lenses
and analyze this with different models. 

Basically answering differt sub-RQs each time that I will introduce.
:::

## Measurement - Predictor

### Emotional support provided

> **Rough English translation of the original German item**<br>
> “How much did you emotionally support your partner today so that they would
be physically active with greater enjoyment?”
>
> [Adapted from @bolgerInvisibleSupportAdjustment2000]{.subtle}

0 = *not at all today* <br>
5 = *very much today*<br>

<!-- ca. 1 Minute -->

## Measurement - Outcome

### Moderate to vigorous physical activity (MVPA)

> **Rough English translation of the original German item**<br>
> “How many minutes did you spend engaging in moderate-to-vigorous physical
activity today?” 
[Adapted from @amireaultGodinShephardQuestionnaire2015]{.subtle}

- **Alone:** _____ minutes
- **Together with your partner:** _____ minutes

[**Total MVPA = solo MVPA + joint MVPA.**]{.fragment}

<!-- ca. 1 Minute -->

## Aggregating for cross-sectional examples

To show simple dyadic models we aggregate the daily diary data to
obtain a "cross-sectional" dataset.

```{r}
library(dplyr)

pilot_aggregated <- pilot |>
  group_by(couple_id, person_id) |>
  summarise(
    across(everything(), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )
```

::: {.fragment}
The variables now represent person-level daily averages across the study:

- `provided_emotional_support`: average daily support provided
- `mvpa_total`: average daily total MVPA (minutes)
:::

<!-- ca. 1 Minute -->

## Pilot data: APIM variables

```{r}
#| label: pilot-aggregated-distributions-plot
#| echo: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Histograms of provided emotional support and person-level mean total MVPA."

distribution_plot(
  pilot_aggregated,
  variables = c("provided_emotional_support", "mvpa_total"),
  bin_widths = c(
    provided_emotional_support = 0.25,
    mvpa_total = 7.5
  )
)
```

<!-- ca. 1 Minute -->

## Many models to choose from

**Various dyadic models** can be used, each answering different questions!

::: {.incremental}
- **Actor-partner interdependence model (APIM)**: interdependent individuals
- **Dyad-individual model (DIM)** and **dyadic score model (DSM)**: within- and between dyad associations
- **Common fate model (CFM)**: latent dyadic process
:::

::: {.fragment}
Other noteable models include the mutual influence model (iida?) and the 
truth-and bias model (ref?). !TODO: also add others? or omit?
:::

<!-- ca. 3 Minutes -->

::: {.notes}
"For RQ 1 we first go with the Actor-Partner interdependence model and 
see how we can use it to answer our research question (in what way). "
:::

## Many models to choose from

**Various dyadic models** can be used, each answering different questions!

- [**Actor-partner interdependence model (APIM)**: interdependent individuals]{.text-highlight}
- **Dyad-individual model (DIM)** and **dyadic score model (DSM)**: within- and between dyad associations
- **Common fate model (CFM)**: latent dyadic process

Other noteable models include the mutual influence model (iida?) and the 
truth-and bias model (ref?). !TODO: also add others?

<!-- ca. 3 Minutes -->

## The distinguishable APIM

<!-- probably add sub-RQ here -->

## The distinguishable APIM

```{r}
#| label: conceptual-distinguishable-apim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with role-specific actor and partner paths."

draw_apim_diagram("distinguishable")
```

::: {.notes}

Explain actor and partner effects only and briefly. Not talking about residuals yet. 
-> to set up next slide

:::

<!-- ca. 2 Minutes 


!TODO: 
  The current APIM diagram already shows residuals and their correlation, even though the notes say not to discuss them yet.
  That reveals the answer before the separate-regressions question.

  A cleaner progression would be:

  - APIM diagram with only actor and partner paths
  - “Could these simply be two separate regressions?”
  - Same APIM diagram, now adding correlated residuals
  - “This is what fitting both equations jointly adds”

  That would make the regression slides an effective problem–solution transition.

-->

## Can we just estimate separate regression equations for each partner?

Do two separate models like this work?

<br>

**Male model:** $\qquad Y_{male_i} = b_{0M} + b_{1M} \times X_{male_i} + b_{2M} \times  X_{female_i} + \epsilon_{male_i}$ 

<br>

**Female model:** $\quad Y_{female_i} = b_{0F} + b_{1F} \times X_{male_i} + b_{2F} \times  X_{female_i} + \epsilon_{female_i}$ 

<!-- ca. 2 minutes (ask students what issues) -->

## Why not just separate regression equations for each partner?

Yes, but there are downsides:

:::{.incremental}
- No formal tests of whether the two actor effects or partner effects are different for males and females.
- Equal paths cannot be pooled into a shared estimate, which can increase power.
- Ignores unmodelled sources of non-independence between partners. 
- A joint dyadic model can tell us how much of the interdependence is explained.
:::

!TODO: cite [bolgers workshop slides]{.subtle}

<!-- ca. 2 minutes -->

:::{.notes}
Thus, we need to properly account for non-explained non-independence! 
E.g., correlated residuals.
:::

## Interpreting the fit of the actor partner interdependence model

<!-- insert the fit model, discuss that in this model we have correlated residuals (and what that means)
And then interpret the actor path values and residual values and correlation -->


<!------------- BELOW HERE IS JUST STRUCTURE  -------------------->

## The exchangeable APIM

Sometimes: 
- partners may not be theoretically distinguishable
- we may have mixed dyad types that we want to pool
- we want to test whether partners are meaningfully distinguishable

Then, we may fit an exchangeable model that imposes symmetry

## The exchangeable APIM

```{r}
#| label: conceptual-exchangeable-apim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual exchangeable APIM with pooled actor and partner paths."

draw_apim_diagram("exchangeable")
```

## The DIM: another parameterization of an exchangeable APIM

```{r}
#| label: conceptual-dim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DIM separating dyad-mean and member-deviation effects."

draw_dim_diagram()
```

## The DSM: a mean-and-difference parameterization of the distinguishable APIM

```{r}
#| label: conceptual-dsm
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DSM linking predictor mean and difference to outcome mean and difference."

draw_dsm_diagram()
```



## Research Question 2

RQ2: In romantic couples, how is collaborative planning linked to physical 
activity?

## Measurement - CFM predictor

### Dyadic predictor: collaborative planning

> **Rough english translation of the original German item**<br>
> “We have already planned precisely for tomorrow when, where, and how we will
be physically active together.”
[Adapted from @scholzBeyondBehaviouralIntentions2008]{.subtle}

0 = *not at all true today*<br>
to 5 = *completely true today*

We aggregate each person's ratings into their **average daily collaborative
planning** across the 55 days.

## Dyadic outcome: joint physical activity



```{r}
#| label: pilot-aggregated-distributions-cfm-plot
#| echo: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Histograms of provided collaborative planning and person-level mean joint MVPA."

distribution_plot(
  pilot_aggregated,
  variables = c("collaborative_planning", "mvpa_joint"),
  bin_widths = c(
    collaborative_planning = 0.25,
    mvpa_joint = 7.5
  )
)
```

## The CFM

```{r}
#| label: conceptual-cfm
#| echo: false
#| fig-width: 10
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual Common Fate Model with shared latent predictor and outcome levels."

draw_cfm_diagram()
```




## References
