---
title: "Dyadic Data Analysis with R"
subtitle: "From Cross-Sectional Models to Intensive Longitudinal Data"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
csl: apa.csl
resources:
  - uzh-logo.svg
  - charite-logo.svg
  - js/residual-concepts.js
  - js/dim-grid.js
format:
  revealjs:
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: true
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup-pilot-data
#| include: false

library(dplyr)
library(ggplot2)

diagram_helper_candidates <- c(
  "diagram-helpers.Rinc",
  file.path("vignettes", "diagram-helpers.Rinc"),
  file.path("..", "..", "vignettes", "diagram-helpers.Rinc")
)
diagram_helper <- diagram_helper_candidates[
  file.exists(diagram_helper_candidates)
][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(
  normalizePath(diagram_helper),
  envir = knitr::knit_global()
)


# FUNCTION TO FIT EARLY MODELS --------------------------------------

early_models_file <- if (file.exists(file.path("R", "fit-early-models.R"))) {
  file.path("R", "fit-early-models.R")
} else {
  file.path("dev", "workshop", "R", "fit-early-models.R")
}

sys.source(
  normalizePath(early_models_file),
  envir = knitr::knit_global()
)

dim_grid_file <- if (file.exists(file.path("R", "dim-grid.R"))) {
  file.path("R", "dim-grid.R")
} else {
  file.path("dev", "workshop", "R", "dim-grid.R")
}

sys.source(
  normalizePath(dim_grid_file),
  envir = knitr::knit_global()
)

data_file <- if (file.exists("long.rds")) "long.rds" else "dev/workshop/long.rds"

pilot <- readRDS(data_file) |>
  select(
    coupleID, userID, gender, ss_pa, ss_pa_min_solo, ss_pa_min_collab,
    ss_emo_pleasure, sc_actionplan
  ) |>
  transmute(
    couple_id = coupleID,
    person_id = userID,
    role = factor(gender, levels = c(1, 2), labels = c("female", "male")),
    provided_emotional_support = ss_emo_pleasure,
    collaborative_planning = sc_actionplan,
    mvpa_solo = if_else(ss_pa == 0, 0, ss_pa_min_solo),
    mvpa_joint = if_else(ss_pa == 0, 0, ss_pa_min_collab),
    mvpa_total = mvpa_solo + mvpa_joint
  )

distribution_plot <- function(data, variables, bin_widths) {
  labels <- c(
    provided_emotional_support = "Provided emotional support",
    collaborative_planning = "Collaborative planning (daily mean)",
    mvpa_solo = "Mean solo MVPA (minutes/day)",
    mvpa_joint = "Mean joint MVPA (minutes/day)",
    mvpa_total = "Mean total MVPA (minutes/day)"
  )

  missing_widths <- setdiff(variables, names(bin_widths))
  if (length(missing_widths)) {
    stop("Missing bin widths for: ", paste(missing_widths, collapse = ", "))
  }

  plots <- Map(\(variable, bin_width) {
    values <- data[[variable]]
    values <- values[is.finite(values)]
    limits <- c(
      floor(min(values) / bin_width) * bin_width,
      ceiling(max(values) / bin_width) * bin_width
    )
    bandwidth <- density(values, adjust = 1.1)$bw
    curve <- density(
      c(values, -values), bw = bandwidth, from = 0, to = limits[2]
    )
    curve <- tibble(
      value = curve$x,
      count = curve$y * 2 * length(values) * bin_width
    )

    ggplot(tibble(value = values), aes(value)) +
      geom_histogram(
        binwidth = bin_width, boundary = limits[1], closed = "right",
        fill = "steelblue", color = "white", linewidth = 0.25
      ) +
      geom_line(
        data = curve, aes(value, count), inherit.aes = FALSE,
        color = "navy", linewidth = 1
      ) +
      labs(title = labels[[variable]], x = NULL, y = "Number of participants") +
      theme_minimal(base_size = 15) +
      theme(
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text = element_text(color = "#222222")
      )
  }, variables, unname(bin_widths[variables]))

  patchwork::wrap_plots(
    plots,
    ncol = min(3, length(plots)),
    axis_titles = "collect_y"
  )
}
```

## Introduction

- Turu & Pascal <!-- ca. 5 min -->
- Tell us about you and your dyadic data!
  - Name, affiliation & career stage
  - Your dyadic data and RQ
  
<!-- ca. 15 Minutes -->

## Today's Aims and Program

| Time | Focus |
|---|---|
| 11:00–12:45 | Introduction and different types of dyadic models |
| 12:45–14:00 | *Lunch break* |
| 14:00–15:30 | Estimating cross-sectional models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

<!-- ca. 2.5 Minutes -->

## Why dyadic data?

> “Much of what we think, do, and feel is determined not by who we are but by
> whom we are with.” [@kennyPartnerEffectsRelationship1999, p. 446]

::: {.incremental}
- **Partners matter:** An outcome may depend on both *my characteristics* and 
*my partner's characteristics*.
- **Some constructs and processes are dyadic:** Shared contexts and interdependence
processes can make partners' outcomes correlated.
- **The configuration matters:** Partners' similarity, or complementarity (differences) 
may predict outcomes beyond either individual score.
:::

[@kennyPartnerEffectsRelationship1999;
@iidaModelsInterdependentIndividuals2018]{.subtle}

<!-- ca. 2.5 Minutes -->

## Introducing today's dyadic dataset!

- 38 physically inactive opposite-sex romantic couples
- Both partners intended to increase their physical activity together
- Smartphone-based planning interventions and JITAIs
- Daily diaries over 55 days

[@kuengDailyAssociationsSocialControl2026;
@hoehenerEffectivenessSmartphoneDyadic2026]{.subtle}

<!-- ca. 1 Minute -->

## Research Question 1 {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ1:]{.rq-label} [In romantic couples, how is provided health-related social support linked to physical activity?]{.rq-text}
:::
:::

:::{.notes}
This is intentionally broad, since we will look at it through different lenses
and analyze this with different models. 

Basically answering differt sub-RQs each time that I will introduce.
:::

## Measurement - Predictor

### Emotional support provided

> **Rough English translation of the original German item**<br>
> “How much did you emotionally support your partner today so that they would
be physically active with greater enjoyment?”
>
> [Adapted from @bolgerInvisibleSupportAdjustment2000]{.subtle}

0 = *not at all today* <br>
5 = *very much today*<br>

<!-- ca. 1 Minute -->

## Measurement - Outcome

### Moderate to vigorous physical activity (MVPA)

> **Rough English translation of the original German item**<br>
> “How many minutes did you spend engaging in moderate-to-vigorous physical
activity today?” 
[Adapted from @amireaultGodinShephardQuestionnaire2015]{.subtle}

- **Alone:** _____ minutes
- **Together with your partner:** _____ minutes

[**Total MVPA = solo MVPA + joint MVPA.**]{.fragment}

<!-- ca. 1 Minute -->

## Aggregating for cross-sectional examples

To show simple dyadic models we aggregate the daily diary data to
obtain a "cross-sectional" dataset.

```{r}
library(dplyr)

pilot_aggregated <- pilot |>
  group_by(couple_id, person_id, role) |>
  summarise(
    across(where(is.numeric), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )
```

::: {.fragment}
The variables now represent person-level daily averages across the study:

- `provided_emotional_support`: average daily support provided
- `mvpa_total`: average daily total MVPA (minutes)
:::

<!-- ca. 1 Minute -->

## Pilot data: APIM variables

```{r}
#| label: pilot-aggregated-distributions-plot
#| echo: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Histograms of provided emotional support and person-level mean total MVPA."

distribution_plot(
  pilot_aggregated,
  variables = c("provided_emotional_support", "mvpa_total"),
  bin_widths = c(
    provided_emotional_support = 0.25,
    mvpa_total = 7.5
  )
)
```

<!-- ca. 1 Minute -->

<!-- TODO: implement plots for females and males separately! 

see for ref:
ggplot(data = BLdyads_doubleentry, aes(x = reldis)) +
  geom_histogram(fill = "white", color = "black") + 
  labs(x = "Relationship Dissatisfaction") +
  facet_grid(. ~ gender) # creating a separate plot for each gender

can we have a similar little code block so participants can re-use it and also print it on the slide?
With a very short code block, can we still style it a bit nicer and closer to what we have currently?
-->

## Many models to choose from

**Various dyadic models** can be used, each answering different questions!

::: {.incremental}
- **Actor-partner interdependence model (APIM)**: interdependent individuals
- **Dyad-individual model (DIM)** and **dyadic score model (DSM)**: within- and between dyad associations
- **Common fate model (CFM)**: latent dyadic process
:::

::: {.fragment}
Other approaches include the **mutual-influence model**, for reciprocal influence
between partners [@iidaModelsInterdependentIndividuals2018, p. 61], and the
**Truth and Bias model**, for separating mean-level bias from correlational
accuracy in judgments of a partner [@sternAssessingAccuracyClose2018].
:::

<!-- ca. 3 Minutes -->

::: {.notes}
"For RQ 1 we first go with the Actor-Partner interdependence model and 
see how we can use it to answer our research question (in what way). "
:::

## Many models to choose from

**Various dyadic models** can be used, each answering different questions!

- [**Actor-partner interdependence model (APIM)**: interdependent individuals]{.text-highlight}
- **Dyad-individual model (DIM)** and **dyadic score model (DSM)**: within- and between dyad associations
- **Common fate model (CFM)**: latent dyadic process

Other approaches include the **mutual-influence model**, for reciprocal influence
between partners [@iidaModelsInterdependentIndividuals2018, p. 61], and the
**Truth and Bias model**, for separating mean-level bias from correlational
accuracy in judgments of a partner [@sternAssessingAccuracyClose2018].

<!-- ca. 1 Minutes -->

<!-- potentially add a slide here showing how much the APIM is used vs. other models if we find a reference -->

## The distinguishable APIM - Conceptually

```{r}
#| label: conceptual-distinguishable-apim-paths-only
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with actor and partner paths."

draw_apim_diagram("distinguishable", show_residuals = FALSE)
```

::: {.notes}

Explain actor and partner effects only and briefly. Not talking about residuals yet. 
-> to set up next slide

:::

<!-- ca. 2 Minutes -->

## Can we just estimate separate regression equations for each partner?

Do two separate models like this work?


$$
  \begin{gathered}
  \\[1em]
  Y_{male_i}
  = b_{0M} + b_{1M}X_{male_i} + b_{2M}X_{female_i} + \epsilon_{male_i}
  \\[2em]
  Y_{female_i}
  = b_{0F} + b_{1F}X_{female_i} + b_{2F}X_{male_i} + \epsilon_{female_i}
  \end{gathered}
$$


<!-- ca. 1 minute 

glmmTMB(
  outcome_male ~ 1 + X_male + X_female,
  
  data = data,
  family = gaussian()
) 

glmmTMB(
  outcome_female ~ 1 + X_male + X_female,
  
  data = data,
  family = gaussian()
) 

-->

## Can we just estimate separate regression equations for each partner?

**Yes! But then we: **

:::{.incremental}
- can't test whether the two actor effects or partner effects are equal
- can't pool equal paths into a shared estimate, which can increase power
- ignore unmodelled sources of interdependence between partners
  - joint modelling tells us how much interdependence remains after accounting for the predictors
  - we can find out how much covariance is explained by our predictors

:::

[@kennyPartnerEffectsRelationship1999, pp. 439–440;
@iidaModelsInterdependentIndividuals2018, pp. 68–69]{.subtle}

:::{.notes}

The reduction in residual covariance indicates how much of the original 
within-dyad covariance is accounted for by the predictors.
  
:::

<!-- ca. 3 minutes -->

:::{.notes}
Thus, we should model non-explained non-independence! 
E.g., correlated residuals.
:::

## What remains after prediction?

```{r}
#| label: residual-variance-and-correlation
#| echo: false
#| results: asis
#| fig-width: 12
#| fig-height: 5.6
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Schematic residual plot showing a fixed prediction, smaller residual variance for female partners, larger residual variance for male partners, and a positive association between partners' residuals."

residual_helper_candidates <- c(
  file.path("R", "draw-residual-concepts.R"),
  file.path("dev", "workshop", "R", "draw-residual-concepts.R")
)
residual_helper <- residual_helper_candidates[
  file.exists(residual_helper_candidates)
][1]

if (is.na(residual_helper)) {
  stop("Could not find dev/workshop/R/draw-residual-concepts.R.")
}

sys.source(
  normalizePath(residual_helper),
  envir = knitr::knit_global()
)

if (knitr::is_html_output()) {
  residual_concepts_widget()
} else {
  print(draw_residual_concepts())
}
```

:::{.notes}

Start with the left-hand side:

- Everyone has the same predicted value in this simplified example.
- The x-axis identifies illustrative dyads in an arbitrary order; it is not
  time and has no substantive ordering.
- A residual is the observed value minus that predicted value. The vertical
  lines make the residuals visible.
- Move the female variance slider: only the female residual spread changes.
- Move the male variance slider: only the male residual spread changes.
- Reset to restore the intentionally unequal residual variances.

Then move to the right-hand side:

- Every numbered point is the same dyad as on the left, combining its two
  partners' residuals.
- Move either variance slider. The corresponding raw residuals and paired
  points move, but the displayed correlation stays exactly the same.
- A raw-score regression line would rotate because its slope also depends on
  the ratio of the two residual standard deviations. The small scale-free guide
  therefore represents correlation without being mistaken for that slope.
- Move the correlation toward zero, then negative, and finally positive.
- With a positive correlation, partners who are above their predictions tend
  to be above them together, and partners who are below tend to be below
  together.
- Residual correlation describes interdependence that remains after accounting
  for the predictors. It is separate from the size of either residual variance.
- Move the correlation while watching both sides: the residuals change, but
  both selected residual variances stay exactly the same.

:::

<!-- ca. 3 Minutes -->

## Residual interdependence in the distinguishable APIM

```{r}
#| label: conceptual-distinguishable-apim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with role-specific actor and partner paths."

draw_apim_diagram("distinguishable")
```

:::{.notes}

Map the preceding visual onto the APIM notation:

- Each outcome has a residual, epsilon.
- The residual standard deviations, sigma, may differ between distinguishable
  partners. Their squares are the residual variances from the previous slide.
- The curved double-headed arrow is the residual correlation, rho(FM).

:::

<!-- ca. 1 Minute -->

## Fitting the APIM

```{r}
#| label: fitted-distinguishable-apim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted distinguishable APIM for provided support and total MVPA."

fit_and_draw_distinguishable_apim(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total,
  labels = c(predictor = "Provided support", outcome = "Total MVPA")
)
```

:::{.notes}

Read the diagram from left to right. All outcome values are minutes of average
daily total MVPA, and all associations are conditional on both partners'
provided support.

- The intercepts, 25.68 for female partners and 23.85 for male partners, are the
  predicted daily MVPA when both partners' average provided support is zero.
- Female actor effect, -4.61: for each one-unit higher support provided by a
  female partner, her own predicted daily MVPA is 4.61 minutes lower, holding
  her male partner's support constant.
- Female partner effect, 15.06: for each one-unit higher support provided by a
  male partner, the female partner's predicted daily MVPA is 15.06 minutes
  higher, holding her own support constant.
- Male actor effect, 17.92: for each one-unit higher support provided by a male
  partner, his own predicted daily MVPA is 17.92 minutes higher, holding his
  female partner's support constant.
- Male partner effect, -9.87: for each one-unit higher support provided by a
  female partner, the male partner's predicted daily MVPA is 9.87 minutes lower,
  holding his own support constant.
- The residual SDs, 21.40 minutes for female partners and 16.73 minutes for male
  partners, describe the remaining person-to-person spread around those
  predictions. An SD is a typical scale of residual deviation, not a variance
  and not the uncertainty of a coefficient.
- The residual correlation, .78, indicates strong positive remaining
  within-couple interdependence: couples in which the female partner is above
  her prediction also tend to have a male partner above his prediction, and
  likewise for being below prediction. This remains after accounting for both
  support scores.

These are associations in the aggregated observational data, not causal effects.

:::

<!-- ca. 2 Minutes -->

## The exchangeable APIM

Sometimes: 

- partners may not be theoretically distinguishable
- we may have mixed dyad types that we want to pool
- we want to test whether partners are meaningfully distinguishable

$\rightarrow$ Then, we may fit an exchangeable model that imposes symmetry

<!-- ca. 1 Minutes -->

## The exchangeable APIM

```{r}
#| label: conceptual-exchangeable-apim-paths-only
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual exchangeable APIM with pooled actor and partner paths."

draw_apim_diagram("exchangeable", show_residuals = FALSE)
```

<!-- ca. 1 Minute -->


## The exchangeable APIM

```{r}
#| label: conceptual-exchangeable-apim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual exchangeable APIM with pooled actor and partner paths."

draw_apim_diagram("exchangeable")
```

:::{.notes}

Keep these points on this diagram rather than adding another slide; the residual
concepts were introduced visually just a few slides earlier.

- The two members are now labelled arbitrarily because their roles are not used
  to define different parameters.
- Exchangeability constrains the actor paths to be equal across members and the
  partner paths to be equal across members. The diagram therefore shows one
  pooled actor effect and one pooled partner effect.
- The two residual SDs are also constrained to be equal. This says that the
  unexplained outcome spread is the same for both arbitrary member positions.
- The curved arrow remains: exchangeability does not assume that partners are
  independent. Their residual correlation is still estimated.
- These are testable symmetry assumptions, not facts about the dyads. Compare
  the exchangeable and distinguishable models if substantive distinguishability
  is uncertain.

:::

<!-- ca. 3 Minutes -->


## Interpreting the fit of the exchangeable APIM

```{r}
#| label: fitted-exchangeable-apim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted exchangeable APIM for provided support and total MVPA."

workshop_exchangeable_apim <- fit_and_draw_exchangeable_apim(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total,
  labels = c(predictor = "Provided support", outcome = "Total MVPA")
)
```

<!-- ca. 3 Minutes -->


## The DIM

- No more actor and partner effects
- We decompose (center) predictors into between-couple and within-couple components
  - Between-couple variable: $\bar{X} - \mu_{X}$
  - Within-couple variable: $Xi - \bar{X}$
  
<!-- insert a table?  --> 
In other words, these predictors express: 

- A couples' average level compared to the average of all couples
- A partners' level compared to the average of their couple

<!-- ca. 2 Minutes -->

## The DIM

```{r}
#| label: interactive-apim-dim-grid
#| echo: false
#| results: asis

dim_reparameterization_grid(
  workshop_exchangeable_apim,
  predictor_label = "Provided support"
)
```

<!-- ca. 4 Minutes -->


## The DIM

```{r}
#| label: conceptual-dim-paths-only
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DIM with dyad-mean and member-deviation paths."

draw_dim_diagram(show_residuals = FALSE)
```

<!-- ca. 3 Minutes -->

## The DIM

```{r}
#| label: conceptual-dim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DIM separating dyad-mean and member-deviation effects."

draw_dim_diagram()
```

<!-- ca. 2 Minutes 
no details on why no correlations, just saying that info of both partners are in both resids
-->


## Interpreting the fit of the DIM

```{r}
#| label: fitted-dim
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted DIM for provided support and total MVPA."

fit_and_draw_dim(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total,
  labels = c(predictor = "Provided support", outcome = "Total MVPA")
)
```

<!-- ca. 3 Minutes -->



## The DSM: a mean-and-difference parameterization of the distinguishable APIM

```{r}
#| label: conceptual-dsm-paths-only
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DSM with paths between predictor and outcome scores."

draw_dsm_diagram(show_residuals = FALSE)
```

## The DSM: a mean-and-difference parameterization of the distinguishable APIM

```{r}
#| label: conceptual-dsm
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DSM linking predictor mean and difference to outcome mean and difference."

draw_dsm_diagram()
```

## Interpreting the fit of the DSM

```{r}
#| label: fitted-dsm
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted DSM for provided support and total MVPA."

fit_and_draw_dsm(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total,
  labels = c(predictor = "Provided support", outcome = "Total MVPA")
)
```

## Honorable mention: The truth and bias model 

<!-- add ref, basic idea, example graphic and basic interpretation of paths -->

<!-- ca. 2 minutes -->

## Research Question 2

RQ2: In romantic couples, how is collaborative planning linked to physical 
activity?

## Measurement - CFM predictor

### Dyadic predictor: collaborative planning

> **Rough english translation of the original German item**<br>
> “We have already planned precisely for tomorrow when, where, and how we will
be physically active together.”
[Adapted from @scholzBeyondBehaviouralIntentions2008]{.subtle}

0 = *not at all true today*<br>
to 5 = *completely true today*

We aggregate each person's ratings into their **average daily collaborative
planning** across the 55 days.

## Dyadic outcome: joint physical activity



```{r}
#| label: pilot-aggregated-distributions-cfm-plot
#| echo: false
#| out-width: 100%
#| out-height: 100%
#| fig-alt: "Histograms of provided collaborative planning and person-level mean joint MVPA."

distribution_plot(
  pilot_aggregated,
  variables = c("collaborative_planning", "mvpa_joint"),
  bin_widths = c(
    collaborative_planning = 0.25,
    mvpa_joint = 7.5
  )
)
```

## The CFM

```{r}
#| label: conceptual-cfm-paths-only
#| echo: false
#| fig-width: 10
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual Common Fate Model with shared latent levels."

draw_cfm_diagram(show_residuals = FALSE)
```

## The CFM

```{r}
#| label: conceptual-cfm
#| echo: false
#| fig-width: 10
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual Common Fate Model with shared latent predictor and outcome levels."

draw_cfm_diagram()
```

<!----------------- APPLIED PART (plus MLM vs. SEM and software) -------------->

TODO: Probably create a second slide deck!

- DHARMa plots!?! -> as required package for workshop -> link to materials or do in tutorial? Also check AR1 etc., etc. can dharma do this if resid is set to almost zero?

- glmmTMB -> as required 
- interdep -> as required (via pak, see instructions; or should we already try a CRAN release now?? We may want to still add additional functions like rotating variance structs back, so probably not). -> maybe just require pak and we install newest version on workshop-day?
- dplyr -> required
- tibble -> required
- lavaan -> required
- ggplot2 -> required



<!-- ILD-part: 

- see dev note on ar1 processes 
- see plots on: https://quantdev.ssri.psu.edu/tutorials/apa-ati-intensive-longitudinal-data-session-m-dyadic-multilevel-model

-->

<!-- Excercises part

Create exercises with data visalization, etc. Panel plots, histograms
correlations (wb corr)?

-->



## References
