---
title: "Dyadic Data Analysis with R"
subtitle: "Part I: Conceptual Foundations"
author:
  - "Pascal Küng"
  - "Turu Stadler"
bibliography: ../../vignettes/references.bib
csl: apa.csl
resources:
  - uzh-logo.svg
  - charite-logo.svg
  - js/residual-concepts.js
format:
  revealjs:
    embed-resources: true
    css: workshop.css
    include-after-body: branding.html
    footer: "Dyadic Data Analysis"
    slide-number: true
    chalkboard: false
    preview-links: auto
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup-pilot-data
#| include: false

library(dplyr)
library(ggplot2)

diagram_helper_candidates <- c(
  "diagram-helpers.Rinc",
  file.path("vignettes", "diagram-helpers.Rinc"),
  file.path("..", "..", "vignettes", "diagram-helpers.Rinc")
)
diagram_helper <- diagram_helper_candidates[
  file.exists(diagram_helper_candidates)
][1]

if (is.na(diagram_helper)) {
  stop("Could not find vignettes/diagram-helpers.Rinc.")
}

sys.source(
  normalizePath(diagram_helper),
  envir = knitr::knit_global()
)


# FUNCTION TO FIT EARLY MODELS --------------------------------------

early_models_file <- if (file.exists(file.path("R", "fit-early-models.R"))) {
  file.path("R", "fit-early-models.R")
} else {
  file.path("dev", "workshop", "R", "fit-early-models.R")
}

sys.source(
  normalizePath(early_models_file),
  envir = knitr::knit_global()
)

dim_grid_candidates <- c(
  "dim-grid.Rinc",
  file.path("vignettes", "dim-grid.Rinc"),
  file.path("..", "..", "vignettes", "dim-grid.Rinc")
)
dim_grid_file <- dim_grid_candidates[file.exists(dim_grid_candidates)][1]

if (is.na(dim_grid_file)) {
  stop("Could not find vignettes/dim-grid.Rinc.")
}

sys.source(
  normalizePath(dim_grid_file),
  envir = knitr::knit_global()
)

apim_exchangeability_file <- if (
    file.exists(file.path("R", "draw-apim-exchangeability.R"))) {
  file.path("R", "draw-apim-exchangeability.R")
} else {
  file.path("dev", "workshop", "R", "draw-apim-exchangeability.R")
}

sys.source(
  normalizePath(apim_exchangeability_file),
  envir = knitr::knit_global()
)

residual_seesaw_file <- if (
    file.exists(file.path("R", "draw-residual-seesaw.R"))) {
  file.path("R", "draw-residual-seesaw.R")
} else {
  file.path("dev", "workshop", "R", "draw-residual-seesaw.R")
}

sys.source(
  normalizePath(residual_seesaw_file),
  envir = knitr::knit_global()
)

data_file <- if (file.exists("long.rds")) "long.rds" else "dev/workshop/long.rds"

pilot <- readRDS(data_file) |>
  select(
    coupleID, userID, gender, ss_pa, ss_pa_min_solo, ss_pa_min_collab,
    ss_emo_pleasure, sc_actionplan
  ) |>
  transmute(
    couple_id = coupleID,
    person_id = userID,
    role = factor(gender, levels = c(1, 2), labels = c("female", "male")),
    provided_emotional_support = ss_emo_pleasure,
    collaborative_planning = sc_actionplan,
    mvpa_solo = if_else(ss_pa == 0, 0, ss_pa_min_solo),
    mvpa_joint = if_else(ss_pa == 0, 0, ss_pa_min_collab),
    mvpa_total = mvpa_solo + mvpa_joint
  )

distribution_plot <- function(data, variables, bin_widths) {
  labels <- c(
    provided_emotional_support = "Provided emotional support",
    collaborative_planning = "Collaborative planning (daily mean)",
    mvpa_solo = "Mean solo MVPA (minutes/day)",
    mvpa_joint = "Mean joint MVPA (minutes/day)",
    mvpa_total = "Mean MVPA (minutes/day)"
  )

  missing_widths <- setdiff(variables, names(bin_widths))
  if (length(missing_widths)) {
    stop("Missing bin widths for: ", paste(missing_widths, collapse = ", "))
  }

  plots <- Map(\(variable, bin_width) {
    values <- data[[variable]]
    values <- values[is.finite(values)]
    limits <- c(
      floor(min(values) / bin_width) * bin_width,
      ceiling(max(values) / bin_width) * bin_width
    )
    bandwidth <- density(values, adjust = 1.1)$bw
    curve <- density(
      c(values, -values), bw = bandwidth, from = 0, to = limits[2]
    )
    curve <- tibble(
      value = curve$x,
      count = curve$y * 2 * length(values) * bin_width
    )

    ggplot(tibble(value = values), aes(value)) +
      geom_histogram(
        binwidth = bin_width, boundary = limits[1], closed = "right",
        fill = "steelblue", color = "white", linewidth = 0.25
      ) +
      geom_line(
        data = curve, aes(value, count), inherit.aes = FALSE,
        color = "navy", linewidth = 1
      ) +
      labs(title = labels[[variable]], x = NULL, y = "Number of participants") +
      theme_minimal(base_size = 15) +
      theme(
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text = element_text(color = "#222222")
      )
  }, variables, unname(bin_widths[variables]))

  patchwork::wrap_plots(
    plots,
    ncol = min(3, length(plots)),
    axis_titles = "collect_y"
  )
}
```

## Introduction

- Turu & Pascal <!-- ca. 5 min -->
- Tell us about you and your dyadic data!
  - Name, affiliation & career stage
  - Your dyadic data and RQ
  
<!-- ca. 15 Minutes -->

## Today's Aims and Program

| Time | Focus |
|---|---|
| 11:00–12:45 | Introduction and types of dyadic models |
| 12:45–14:00 | *Lunch break* |
| 14:00–15:30 | Estimating cross-sectional models in R |
| 15:30–15:45 | *Short break* |
| 15:45–16:45 | Intensive longitudinal extensions |
| 16:45–17:00 | *Short break* |
| 17:00–18:00 | Exercises in R |

<!-- ca. 2.5 Minutes -->

# Introduction
Relevance and todays' example dataset

## Why dyadic data?

> “Much of what we think, do, and feel is determined not by who we are but by
> whom we are with.” [@kennyPartnerEffectsRelationship1999, p. 446]

::: {.incremental}
- **Partners matter:** An outcome may depend on both *my characteristics* and 
*my partner's characteristics*.
- **Some constructs and processes are dyadic:** Shared contexts and interdependence
processes can make partners' outcomes correlated.
- **The configuration matters:** Partners' similarity, or complementarity (differences) 
may predict outcomes beyond either individual score.
:::

[@kennyPartnerEffectsRelationship1999;
@iidaModelsInterdependentIndividuals2018]{.subtle}

<!-- ca. 2.5 Minutes -->

## Introducing today's dyadic dataset!

- 38 physically inactive opposite-sex romantic couples
- Both partners intended to increase their physical activity together
- Smartphone-based planning interventions and JITAIs
- Daily diaries over 55 days

[@kuengDailyAssociationsSocialControl2026;
@hoehenerEffectivenessSmartphoneDyadic2026]{.subtle}

<!-- ca. 1.5 Minutes -->

## Research Question 1 {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ1:]{.rq-label} [In romantic couples, how is provided health-related social support linked to physical activity?]{.rq-text}
:::
:::

:::{.notes}
This is intentionally broad, since we will look at it through different lenses
and analyze this with different models. 

Basically answering differt sub-RQs each time that I will introduce.
:::

<!-- ca. 0.5 Minutes -->

## Measurement - Predictor

### Emotional support provided

> **Rough English translation of the original German item**<br>
> “How much did you emotionally support your partner today so that they would
be physically active with greater enjoyment?”
>
> [Adapted from @bolgerInvisibleSupportAdjustment2000]{.subtle}

0 = *not at all today* <br>
5 = *very much today*<br>

<!-- ca. 1 Minute -->

## Measurement - Outcome

### Moderate to vigorous physical activity (MVPA)

> **Rough English translation of the original German item**<br>
> “How many minutes did you spend engaging in moderate-to-vigorous physical
activity today?” 
[Adapted from @amireaultGodinShephardQuestionnaire2015]{.subtle}

- **Alone:** _____ minutes
- **Together with your partner:** _____ minutes

[**Total MVPA = solo MVPA + joint MVPA.**]{.fragment}

<!-- ca. 1 Minute -->

## Aggregating for cross-sectional examples

To show simple dyadic models we aggregate the daily diary data to
obtain a "cross-sectional" dataset.

```{r}
library(dplyr)

pilot_aggregated <- pilot |>
  group_by(couple_id, person_id, role) |>
  summarise(
    across(where(is.numeric), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )
```

::: {.fragment}
The variables now represent person-level daily averages across the study:

- `provided_emotional_support`: average daily support provided
- `mvpa_total`: average daily total MVPA (minutes)
:::

<!-- ca. 1 Minute -->

## Plotting provided support by partner role

```{r}
#| label: build-pilot-support-distributions-by-role
#| echo: true
#| code-line-numbers: false

support_distribution_plot <- ggplot(
  pilot_aggregated,
  aes(provided_emotional_support, y = after_stat(density))
) +
  geom_histogram(
    bins = 15, boundary = 0,
    fill = "steelblue", color = "white", alpha = 0.75
  ) +
  geom_density(
    bounds = c(0, Inf), color = "navy", linewidth = 1.1
  ) +
  facet_wrap(~role, nrow = 1) +
  labs(x = "Average provided support", y = "Density") +
  see::theme_modern(base_size = 16)
```

<!-- ca. 1 Minute -->

## Provided support by partner role

```{r}
#| label: pilot-support-distributions-by-role
#| echo: false
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.1
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Histograms and boundary-corrected density curves of average provided emotional support, shown separately for female and male partners."

support_distribution_plot
```

<!-- ca. 1 Minute -->

## MVPA by partner role

```{r}
#| label: pilot-mvpa-distributions-by-role
#| echo: false
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.1
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Histograms and boundary-corrected density curves of average daily MVPA, shown separately for female and male partners."

ggplot(pilot_aggregated,
       aes(mvpa_total, y = after_stat(density))) +
  geom_histogram(
    bins = 15, boundary = 0,
    fill = "steelblue", color = "white", alpha = 0.75
  ) +
  geom_density(
    bounds = c(0, Inf), color = "navy", linewidth = 1.1
  ) +
  facet_wrap(~role, nrow = 1) +
  labs(x = "Average MVPA (minutes/day)", y = "Density") +
  see::theme_modern(base_size = 16)
```

<!-- ca. 1 Minute -->

# Different types of dyadic models

## One model dominates dyadic research

```{r}
#| label: apim-use-plot
#| echo: false
#| dev: svglite
#| fig-alt: "A horizontal bar showing that more than 85% of 998 coded dyadic studies used an APIM or APIM variant, leaving fewer than 15% for all other model specifications combined."
#| out-width: 100%

model_use <- tibble(
  model = c("APIM or APIM variant", "All other specifications"),
  xmin = c(0, 85),
  xmax = c(85, 100),
  midpoint = c(42.5, 92.5),
  label = c("> 85%\nAPIM or APIM variant", "< 15%\nAll others")
)

ggplot(model_use) +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 1, fill = model),
    color = "white", linewidth = 1.2
  ) +
  geom_text(
    aes(x = midpoint, y = 0.5, label = label, color = model),
    lineheight = 0.95, fontface = "bold", size = c(9, 6.2)
  ) +
  scale_fill_manual(values = c(
    "APIM or APIM variant" = "#0028a5",
    "All other specifications" = "#e8eefb"
  )) +
  scale_color_manual(values = c(
    "APIM or APIM variant" = "white",
    "All other specifications" = "#001f73"
  )) +
  coord_cartesian(xlim = c(0, 100), ylim = c(-0.06, 1.06), expand = FALSE) +
  theme_void() +
  theme(legend.position = "none", plot.margin = margin(12, 6, 12, 6))
```

::: {.subtle}
Preliminary results from an ongoing systematic review (APIM use coded for
$n = 998$ eligible articles)

[@sakalukRenewalDyadicStructural2025, p. 5].
:::

<!-- ca. 1 Minute -->

## Many models to choose from

**Various dyadic models** can be used, each answering different questions!

::: {.incremental}
- **Actor-partner interdependence model (APIM)**: interdependent individuals
- **Dyad-individual model (DIM)** and **dyadic score model (DSM)**: within- and between dyad associations
- **Common fate model (CFM)**: latent dyadic process
:::

::: {.fragment}
Other approaches include the **mutual-influence model**, for reciprocal influence
between partners [@iidaModelsInterdependentIndividuals2018, p. 61], and the
**Truth and Bias model**, for separating mean-level bias from correlational
accuracy in judgments of a partner [@kennyAccuracyBiasPerception2001;
@westTruthBiasModel2011].
:::

<!-- ca. 3 Minutes -->

## Many models to choose from

**Various dyadic models** can be used, each answering different questions!

- [**Actor-partner interdependence model (APIM)**: interdependent individuals]{.text-highlight}
- **Dyad-individual model (DIM)** and **dyadic score model (DSM)**: within- and between dyad associations
- **Common fate model (CFM)**: latent dyadic process

Other approaches include the **mutual-influence model**, for reciprocal influence
between partners [@iidaModelsInterdependentIndividuals2018, p. 61], and the
**Truth and Bias model**, for separating mean-level bias from correlational
accuracy in judgments of a partner [@kennyAccuracyBiasPerception2001;
@westTruthBiasModel2011].

::: {.notes}
"For broad RQ 1 we first use Actor-Partner interdependence model"
:::


<!-- ca. 1 Minute -->


## The distinguishable APIM

```{r}
#| label: conceptual-distinguishable-apim-paths-only
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with actor and partner paths."

draw_apim_diagram("distinguishable", show_residuals = FALSE)
```

::: {.fragment}
Actor and partner effects together explain some of the covariance between partners' 
outcomes. 
:::


<!-- 
1. Explain actor and partner effects only and briefly. 

2. Reveal the final statement 
-->

<!-- ca. 2 Minutes -->

## The distinguishable APIM

```{r}
#| label: conceptual-distinguishable-apim_brief
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with role-specific actor and partner paths."

draw_apim_diagram("distinguishable")
```

Actor and partner effects together explain some of the covariance between partners' 
outcomes.  $\rightarrow$ But some covariance may remain!

:::{.notes}
But more on the residuals later!!!
:::

<!-- ca. 0.5 Minutes -->

## How to estimate actor and partner effects?

Conceptually, we are esimating these two equations in a single model:

$$
  \begin{gathered}
  \\[0.5em]
  Y_{\mathrm{male},i}
  = b_{0,\mathrm{M}} + b_{1,\mathrm{M}} \times X_{\mathrm{male},i}
  + b_{2,\mathrm{M}} \times X_{\mathrm{female},i} + \epsilon_{\mathrm{male},i}
  \\[2em]
  Y_{\mathrm{female},i}
  = b_{0,\mathrm{F}} + b_{1,\mathrm{F}} \times X_{\mathrm{female},i}
  + b_{2,\mathrm{F}} \times X_{\mathrm{male},i} + \epsilon_{\mathrm{female},i}
  \end{gathered}
$$
<br>
[Could we just use two separate models?]{.fragment}

<!-- ca. 1 Minutes -->

## Could we just use two separate models?

### Male model

```{r eval = FALSE}
male_model <- glmmTMB(
  
  outcome_male ~ X_male + X_female,
  
  data = data
)
```
<br>

### Female model

```{r eval = FALSE}
female_model <- glmmTMB(
  
  outcome_female ~ X_female + X_male,
  
  data = data
) 

```

<!-- ca. 0.5 Minutes -->

## Could we just use two separate models?

**Yes, we could! But then we: **

:::{.incremental}
- can't test whether the two actor effects or partner effects are equal
- can't pool equal paths into a shared estimate, which can increase power
- ignore unmodelled sources of interdependence between partners
  - joint modelling tells us how much interdependence remains after accounting for the predictors

:::

::: {.bottom-reference}
[@kennyPartnerEffectsRelationship1999, pp. 439–440;
@iidaModelsInterdependentIndividuals2018, pp. 68–69]
:::

:::{.notes}

The reduction in residual covariance indicates how much of the original 
within-dyad covariance is accounted for by the predictors.

:::

<!-- ca. 3 minutes -->


## Residual interdependence in the distinguishable APIM

```{r}
#| label: conceptual-distinguishable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual distinguishable APIM with role-specific actor and partner paths."

draw_apim_diagram("distinguishable")
```

:::{.notes}

Map the preceding visual onto the APIM notation:

- Each outcome has a residual, which is what's not explained by the predictors.
- The residual standard deviations, $\sigma$, may differ between distinguishable
  partners. Their squares are the residual variances from the previous slide.
- The curved double-headed arrow is the residual correlation,
  $\rho_{\epsilon_F\epsilon_M}$.

:::

<!-- ca. 2 Minutes -->

## What remains after prediction?

```{r}
#| label: residual-variance-and-correlation
#| echo: false
#| results: asis
#| fig-width: 12
#| fig-height: 5.6
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Schematic residual plot showing a fixed prediction, smaller residual variance for female partners, larger residual variance for male partners, and a positive association between partners' residuals."

residual_helper_candidates <- c(
  file.path("R", "draw-residual-concepts.R"),
  file.path("dev", "workshop", "R", "draw-residual-concepts.R")
)
residual_helper <- residual_helper_candidates[
  file.exists(residual_helper_candidates)
][1]

if (is.na(residual_helper)) {
  stop("Could not find dev/workshop/R/draw-residual-concepts.R.")
}

sys.source(
  normalizePath(residual_helper),
  envir = knitr::knit_global()
)

if (knitr::is_html_output()) {
  residual_concepts_widget()
} else {
  print(draw_residual_concepts())
}
```

:::{.notes}

On the left:

- Everyone has the same predicted value in this simplified example.
- The x-axis identifies illustrative dyads in an arbitrary order.

- *A residual is the observed value minus that predicted value. *

On the right:

- Every numbered point is the same dyad as on the left, combining its two
  partners' raw residuals.
  
- **Move variance sliders** $\rightarrow$ raw residuals and paired
  points move, but the displayed correlation stays exactly the same.

- Residual **correlation describes interdependence** that remains after accounting
  for the predictors. It is separate from the size of either residual variance.

- **Move the correlation** while watching both sides: the residuals change, but
  both selected residual variances stay exactly the same.

:::

<!-- ca. 4 Minutes -->


## Fitting the distinguishable APIM

```{r}
#| label: fitted-distinguishable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted distinguishable APIM for provided support and MVPA."

distinguishable_apim <- fit_and_draw_distinguishable_apim(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total,
  labels = c(predictor = "Provided\nsupport", outcome = "MVPA")
)
```

:::{.notes}

**Ask participants to interpret** 

**MVPA: ** minutes of average daily total MVPA across the 55 days

- All associations are conditional on both partners' provided support.

- The intercepts, 25.68 for female partners and 23.85 for male partners, are the
  predicted daily MVPA when both partners' average provided support is zero.
  
- Female actor effect, -4.61: for each one-unit higher support provided by a
  female partner, her own predicted daily MVPA is 4.61 minutes lower, holding
  her male partner's support constant.
  
- Female partner effect, 15.06: for each one-unit higher support provided by a
  male partner, the female partner's predicted daily MVPA is 15.06 minutes
  higher, holding her own support constant.
  
- Male actor effect, 17.92: for each one-unit higher support provided by a male
  partner, his own predicted daily MVPA is 17.92 minutes higher, holding his
  female partner's support constant.
  
- Male partner effect, -9.87: for each one-unit higher support provided by a
  female partner, the male partner's predicted daily MVPA is 9.87 minutes lower,
  holding his own support constant.
  
- The residual SDs, 21.40 minutes for female partners and 16.73 minutes for male
  partners, describe the remaining person-to-person spread around those
  predictions. An SD is a typical scale of residual deviation, not a variance
  and not the uncertainty of a coefficient.
  
- The residual correlation, .78, indicates strong positive remaining
  within-couple interdependence: couples in which the female partner is above
  her prediction also tend to have a male partner above his prediction, and
  likewise for being below prediction. This remains after accounting for both
  support scores.

These are associations in the aggregated observational data, not causal effects.

:::

<!-- ca. 5 Minutes -->

## The exchangeable APIM

**Sometimes:** 

- partners may not be theoretically meaningfully distinguishable
- no evidence of difference in effects $\rightarrow$ pool effects for greater power
- we want to formally test distinguishability

$\rightarrow$ Then, we may fit an exchangeable model that imposes symmetry

<!-- ca. 2 Minutes -->

## The exchangeable APIM

```{r}
#| label: conceptual-exchangeable-apim-paths-only
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual exchangeable APIM with pooled actor and partner paths."

draw_apim_diagram("exchangeable", show_residuals = FALSE)
```

:::{.notes}

Two members are now named arbitrarily. Swapping labels will NOT change the model

Pooled actor and partner effects

pooled intercepts pooling

:::

<!-- ca. 1 Minute -->


## The exchangeable APIM

```{r}
#| label: conceptual-exchangeable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual exchangeable APIM with pooled actor and partner paths."

draw_apim_diagram("exchangeable")
```

:::{.notes}

The two residual SDs are also constrained to be equal (e.g., pooled) 

This says that the unexplained outcome spread is the same for both arbitrary member positions.

The curved arrow remains: exchangeability does not assume that partners are
independent!!!! Their residual correlation is still estimated.

:::

<!-- ca. 2 Minutes -->

## Exchangeability constraints

```{r}
#| label: apim-exchangeability-comparison
#| echo: false
#| dev: svg
#| fig-width: 10.5
#| fig-height: 5.2
#| out-width: 95%
#| fig-align: center
#| fig-alt: "Distinguishable and exchangeable APIM equations, predictions, and residual constraints."

draw_apim_exchangeability_comparison()
```

:::{.notes}

In this example, imagine we have a few couples with an exactly $X_F = 1$ and $X_M = 1$. 
For these couples, the following would be true... 

On the left: 

- intercepts, actor effects, partner effects, and residual standard
  deviations may differ by role.
  
- Follow each prediction from its intercept through the actor contribution in
  blue and the partner contribution in purple. 
  
- The open circle is the conditional prediction. 
  - The nearby dots represent illustrative observed outcomes, 
  - and the arrow labelled $\epsilon$ isolates one residual as the distance from the prediction to an observation.
  
- The shaded area is supposed to visualize the residual variance  
  
On the right:

- the same intercept, actor effect, partner effect, and residual
  standard deviation are used for both arbitrary member positions.

- **Shared coefficients do not imply identical predictions!!**: the actor and partner
  predictor values are exchanged between the two equations.
  
- Equal residual variance is a constraint on the modeled residual distribution,
  not a claim that the two members have identical residual values.
  
- In both models, partners' residuals may still be correlated!

:::

<!-- ca. 4 Minutes -->

## Fitting the exchangeable APIM

```{r}
#| label: fitted-exchangeable-apim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted exchangeable APIM for provided support and MVPA."

exchangeable_apim <- fit_and_draw_exchangeable_apim(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total,
  labels = c(predictor = "Provided\nsupport", outcome = "MVPA")
)
```

<!-- ca. 3 Minutes -->

:::{.notes}

-> Let students interpret

-> Ask if they think this model fits the data well, considering the distinguishable estimates. 

:::

## Testing exchangeability

```{r}
library(dyadMLM)

compare_dyad_models(
  exchangeable_apim,
  distinguishable_apim
)
```



## The DIM

- A re-parametrization of the **exchangeable** APIM!
- We decompose predictors into two variables:
  - Dyad mean (GM-centered): $\bar{X} - \mu_{X}$
  - Within-dyad member deviation: $X_i - \bar{X}$
  
In other words, these predictors express: 

- A couple's average level compared with the average of all couples
- A partner's level compared with the average of their couple

:::{.notes}

So, you can see how this would answer different research questions than an
APIM, even though both model encode the exact same information and are
interchangeable.

:::

<!-- ca. 2 Minutes -->

## DIM predictors are APIM predictors in disguise!

```{r}
#| label: interactive-apim-dim-grid
#| echo: false
#| results: asis

dim_reparameterization_grid(
  exchangeable_apim,
  id = "dim-grid-introduction",
  predictor_label = "Provided support",
  outcome_label = "MVPA",
  compact = TRUE
)
```

<!-- ca. 4 Minutes -->


## The DIM

```{r}
#| label: conceptual-dim-paths-only
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DIM with a between-dyad path from the dyad mean and a within-dyad path from the within-dyad member deviation."

draw_dim_diagram(show_residuals = FALSE)
```

:::{.notes}

- Is a couple that is higher on X than others also higher on Y than others?

- is a person within a couple that is higher than their partner on X 
(or higher than the dyad-mean), also higher than their partner on Y?
<!-- TODO @codex fact check these interpretations -->

:::

<!-- ca. 1 Minute -->

## The DIM

```{r}
#| label: conceptual-dim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DIM separating the between-dyad effect of the dyad mean from the within-dyad effect of the within-dyad member deviation."

draw_dim_diagram()
```

:::{.notes}

Ask people: where did the residual correlation between partners go? Are they no longer correlated?

$\rightarrow$ No, the residual correlation between individuals residuals are encoded in these two residuals.

The mean residual captures whether both partners are higher or lower than 
predicted $\rightarrow$ Can capture positive correlation between the partners!

The deviation residual captures whether one partner is higher while the
other is lower $\rightarrow$ can capture negative correlation between the partners!

$$
\operatorname{Cov}(e_1,e_2)
=\operatorname{Var}(M)-\operatorname{Var}(D).
$$

:::

<!-- ca. 2 Minutes -->


## Why are the DIM residuals uncorrelated?

```{r}
#| label: dim-residual-seesaw
#| echo: false
#| dev: svg
#| fig-width: 10.8
#| fig-height: 3.15
#| out-width: 100%
#| fig-align: center
#| fig-alt: "An elevator represents the dyad mean residual, which moves both partners together; a seesaw represents the within-dyad residual, which moves partners in opposite directions. Swapping arbitrary member labels reverses only the seesaw, requiring zero covariance under exchangeability."

draw_residual_seesaw("dim")
```

::: {.seesaw-swap}
**Swap the arbitrary member labels:**
$r_{\mathrm{M}} \mapsto r_{\mathrm{M}} \qquad
r_{\mathrm{D}} \mapsto -r_{\mathrm{D}}$
:::

::: {.seesaw-equation}
Exchangeability requires the same covariance after relabelling:

$$
\operatorname{Cov}(r_{\mathrm{M}},r_{\mathrm{D}})
= \operatorname{Cov}(r_{\mathrm{M}},-r_{\mathrm{D}})
= -\operatorname{Cov}(r_{\mathrm{M}},r_{\mathrm{D}})
\quad\Longrightarrow\quad
\operatorname{Cov}(r_{\mathrm{M}},r_{\mathrm{D}})=0
$$
:::

::: {.seesaw-takeaway}
**Zero does not mean independent partners:**
$\operatorname{Cov}(e_1,e_2)
=\operatorname{Var}(r_{\mathrm{M}})-\operatorname{Var}(r_{\mathrm{D}})$

Their residual association is encoded by the relative elevator and seesaw variances.
:::

:::{.notes}

POTENTIALLY SKIP FOR TIME

Elevator: once the couples' mean value X has accounted for some variance in the
couples' outcome Y, there might still be some movement up or down that is unaccounted. $\rightarrow$ residual

Seasaw: once a persons' difference from the couple mean has accounted for some variance
in the persons' difference from the couple mean in Y, there might still be some additional deviation $\rightarrow$ residual

In the DIM, the DIM, the member positions on the seasaw are arbitrary.
Swapping them leaves the elevator coordinate unchanged but reverses the sign of
the seesaw coordinate.

---

You might wonder, why the elevator and seasaw residuals are not correlated.

- A positive correlation for example would imply that higher Y1 + Y2 is linked to, Y1 being higher than Y2.
- If partners are exchangeable, across all partners, the differences in the seasaw cancel out to
zero. ! TODO @codex fact check this interpretation for correctness.

But again, this does not make the partners independent, and this model
IS exactly equivalent to the APIM

:::

<!-- ca. 4 Minutes -->

## Fitting the DIM

```{r}
#| label: fitted-dim
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted DIM for provided support and MVPA."

fit_and_draw_dim(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total,
  labels = c(predictor = "support", outcome = "MVPA")
)
```

<!-- ca. 3 Minutes -->

:::{.notes}

Keep in mind that previously with the same data we have seen that 
the exchangeability constraints are not really supported by the data. Nonetheless
we fit the DIM here with the same data and variables to demonstrate. 

-> let students interpret

-> Why no intercept? for the deviation? 

No intercept for the MVPA deviation, because that would break exchangeability. 
if support deviation is zero, then the expected MVPA deviation is also zero. 

-> We conceptually DO have an intercept, it is just constrained to be ZERO!

-> just like the correlation between the residuals, they are not estimated, which 
effectively constrains them to be zero. 

:::

## DIM to APIM coefficients

- Once we fit an APIM, we have already fitted a DIM—and vice versa.

- We can transform values from a fitted APIM into DIM values.


## DIM to APIM coefficients {.compact-derivation}

[For this example, the intercept is omitted.]{.derivation-context}

If our APIM predictor values for a focal person are:
$X_{\mathrm{self}}=1,\quad X_{\mathrm{partner}}=1$,

the same predictor profile in DIM coordinates is:
$X_{\mathrm{mean}}=1,\quad X_{\mathrm{dev,self}}=0.$

So,

$$
\hat Y_{\mathrm{self}}=1\times a+1\times p
$$

implies

$$
\hat Y_{\mathrm{self}}=1\times b_{\mathrm{mean}}+0\times b_{\mathrm{dev}}
=b_{\mathrm{mean}}.
$$

Therefore,

$$
b_{\mathrm{mean}}=a+p.
$$

## DIM to APIM coefficients {.compact-derivation}

[For this example, the intercept is omitted.]{.derivation-context}

If our APIM predictor values for a focal person are:
$X_{\mathrm{self}}=1,\quad X_{\mathrm{partner}}=-1$,

the same predictor profile in DIM coordinates is:
$X_{\mathrm{mean}}=0,\quad X_{\mathrm{dev,self}}=1$.

So,

$$
\hat Y_{\mathrm{self}}=1\times a+(-1)\times p
$$

implies for the focal person, whose $X_{\mathrm{self}}$ is 1 unit above the
couple mean:

$$
\hat Y_{\mathrm{self}}=0\times b_{\mathrm{mean}}+1\times b_{\mathrm{dev}}
=b_{\mathrm{dev}}.
$$

Therefore,

$$
b_{\mathrm{dev}}=a-p.
$$


## DIM to APIM coefficients

```{r}
#| label: interactive-apim-dim-grid-coefficients
#| echo: false
#| results: asis

dim_reparameterization_grid(
  exchangeable_apim,
  id = "dim-grid-coefficients",
  predictor_label = "Provided support",
  outcome_label = "MVPA",
  compact = TRUE
)
```

<!-- Practical quick demonstration and showing formulas -->

<!-- ca. 5 Minutes -->

## The DSM

### The dyadic score model (DSM) is: 

:::{.incremental}
- a re-parametrization of the distinguishable APIM
- the unconstrained distinguishable version of the DIM

- While the DIM uses deviation from the couple mean which is opposite for each
partners $X_i - \bar{X}$, the DSM uses meaningful directed differences, e.g.,
$X_\mathrm{female} - X_\mathrm{male}$, which is repeated for both partners. 
  - This allows the estimation of cross-paths, since both partners' values do not
  cancel out like in the DIM$ 
  
:::
  
<!-- TODO @coded verify the correctness and clarity of this slide, and how
layout could be improved-->

<!-- ca. 3 Minutes -->

## The DSM

```{r}
#| label: conceptual-dsm-paths-only
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DSM with paths between predictor and outcome scores."

draw_dsm_diagram(show_residuals = FALSE)
```

## The DSM

```{r}
#| label: conceptual-dsm
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual DSM linking predictor mean and difference to outcome mean and difference."

draw_dsm_diagram()
```

## In the DSM, the seats are labelled

```{r}
#| label: dsm-residual-seesaw
#| echo: false
#| dev: svg
#| fig-width: 10.8
#| fig-height: 3.15
#| out-width: 100%
#| fig-align: center
#| fig-alt: "A seesaw with female and male seats illustrates that the DSM outcome difference has a meaningful direction. Because the labels cannot be swapped arbitrarily, the outcome mean and difference residuals may correlate."

draw_residual_seesaw("dsm")
```

::: {.seesaw-equation .seesaw-equation-dsm}
$$
D_Y=Y_{\mathrm{F}}-Y_{\mathrm{M}}
\qquad\text{and}\qquad
\operatorname{Cov}(r_{\mathrm{M}},r_{\mathrm{D}})\ \text{may be non-zero.}
$$
:::


::: {.seesaw-takeaway}
**A positive $\rho_{r_m r_d}$ means:** couples whose outcomes are higher than predicted
on average also tend to have a larger female–male difference than predicted.
:::


::: {.seesaw-footnote}
The correlation describes a pattern across couples—not two residuals varying
within one couple.
:::

:::{.notes}

- The DSM uses the same elevator and seesaw idea, but the two seats now have
  substantively meaningful labels.

- If the outcome difference is defined as female minus male, reversing its sign
  changes the meaning of the variable. (different than before, where swapping seats did not change the model)

- The mean and difference residuals may consequently be correlated.
- A positive correlation means that, across couples, couples whose outcomes are
  higher than predicted on average also tend to have a larger female–male
  difference than predicted.
- This correlation describes a pattern across couples. It is not a correlation
  calculated from two changing residuals inside a single couple.

:::

<!-- ca. 2 Minutes -->

## Fitting the DSM

```{r}
#| label: fitted-dsm
#| echo: false
#| dev: svg
#| fig-width: 9
#| fig-height: 4.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted DSM for provided support and MVPA."

fit_and_draw_dsm(
  pilot_aggregated,
  predictor = provided_emotional_support,
  outcome = mvpa_total,
  labels = c(predictor = "support", outcome = "MVPA")
)
```


## Research Question 2 {.rq-slide}

::: {.rq-center}
::: {.rq-card}
[RQ2:]{.rq-label} [In romantic couples, how is collaborative planning linked to physical 
activity?]{.rq-text}
:::
:::

## Measurement - CFM predictor

### Dyadic predictor: collaborative planning

> **Rough english translation of the original German item**<br>
> “We have already planned precisely for tomorrow when, where, and how we will
be physically active together.”
[Adapted from @scholzBeyondBehaviouralIntentions2008]{.subtle}

0 = *not at all true today*<br>
to 5 = *completely true today*

We aggregate each person's ratings into their **average daily collaborative
planning** across the 55 days.

## Collaborative planning by partner role

```{r}
#| label: pilot-planning-distributions-by-role
#| echo: false
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.1
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Histograms and boundary-corrected density curves of average collaborative planning, shown separately for female and male partners."

ggplot(pilot_aggregated,
       aes(collaborative_planning, y = after_stat(density))) +
  geom_histogram(
    bins = 15, boundary = 0,
    fill = "steelblue", color = "white", alpha = 0.75
  ) +
  geom_density(
    bounds = c(0, Inf), color = "navy", linewidth = 1.1
  ) +
  facet_wrap(~role, nrow = 1) +
  labs(x = "Average collaborative planning", y = "Density") +
  see::theme_modern(base_size = 16)
```

:::{.notes}

- Each panel shows a partner's average daily collaborative planning across the
  study.
- The shared axes allow direct comparison of the two role-specific
  distributions.
- These are descriptive role differences, not evidence that the CFM parameters
  differ by role.

:::

<!-- ca. 1 Minute -->

## Joint MVPA by partner role

```{r}
#| label: pilot-joint-mvpa-distributions-by-role
#| echo: false
#| dev: svglite
#| fig-width: 12
#| fig-height: 5.1
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Histograms and boundary-corrected density curves of average daily joint MVPA, shown separately for female and male partners."

ggplot(pilot_aggregated,
       aes(mvpa_joint, y = after_stat(density))) +
  geom_histogram(
    bins = 15, boundary = 0,
    fill = "steelblue", color = "white", alpha = 0.75
  ) +
  geom_density(
    bounds = c(0, Inf), color = "navy", linewidth = 1.1
  ) +
  facet_wrap(~role, nrow = 1) +
  labs(x = "Average joint MVPA (minutes/day)", y = "Density") +
  see::theme_modern(base_size = 16)
```

:::{.notes}

- Each panel shows person-level average daily MVPA performed jointly with the
  partner.
- Both roles share the same axes and each dyad contributes one person to each
  panel.
- The distributions are descriptive inputs to the subsequent common fate
  model.

:::

<!-- ca. 1 Minute -->

## The CFM

```{r}
#| label: conceptual-cfm-paths-only
#| echo: false
#| dev: svg
#| fig-width: 10
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual Common Fate Model with shared latent levels."

draw_cfm_diagram(show_residuals = FALSE)
```

## The CFM

```{r}
#| label: conceptual-cfm
#| echo: false
#| dev: svg
#| fig-width: 10
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Conceptual Common Fate Model with four observed indicator intercepts, shared latent predictor and outcome levels, and symbolic standard deviations for all four indicator residuals."

draw_cfm_diagram()
```


:::{.notes}

- The intercepts are measurement intercepts. The expected values of each members' report
when the corresponding latent dyad level is zero. 
  - because latent means are fixed to zero (as a reference point) they represent the mean reported values.

- Member 1 and Member 2 may be meaningful, **distinguishable** roles, such as
  female and male.
- Alternatively, the positions may be assigned arbitrarily and the members
  treated as **exchangeable**.
- The next slide shows the equality constraints required so that swapping the
  arbitrary labels does not change the model.

:::

## Fitted distinguishable CFM

::: {.cfm-constraint}
**Constraint:**
$\sigma^2_{\epsilon_{Y_F}} = \sigma^2_{\epsilon_{Y_M}}$.
Without it, the estimated male outcome residual variance was negative.
:::

```{r}
#| label: fitted-cfm
#| echo: false
#| dev: svg
#| fig-width: 10
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted Common Fate Model for collaborative planning and joint MVPA, including four observed indicator intercepts and unstandardized standard deviations for all four indicator residuals."

cfm_data <- pilot_aggregated |>
  group_by(couple_id) |>
  filter(n_distinct(role) == 2) |>
  ungroup() |>
  select(couple_id, role, collaborative_planning, mvpa_joint) |>
  tidyr::pivot_wider(
    names_from = role,
    values_from = c(collaborative_planning, mvpa_joint),
    names_sep = "_"
  )

distinguishable_cfm_model <- '
  # Shared dyad levels: fix all four loadings to 1.
  planning_level =~
    1 * collaborative_planning_female +
    1 * collaborative_planning_male
  joint_mvpa_level =~
    1 * mvpa_joint_female +
    1 * mvpa_joint_male

  # Association between the two shared levels.
  joint_mvpa_level ~ b_level * planning_level

  # Same-member residual covariances.
  collaborative_planning_female ~~ c_female * mvpa_joint_female
  collaborative_planning_male ~~ c_male * mvpa_joint_male

  # Both partners report the same jointly performed activity.
  mvpa_joint_female ~~ e_mvpa * mvpa_joint_female
  mvpa_joint_male ~~ e_mvpa * mvpa_joint_male
'

workshop_distinguishable_cfm <- lavaan::sem(
  distinguishable_cfm_model,
  data = cfm_data,
  missing = "fiml",
  meanstructure = TRUE
)

draw_cfm_diagram(
  model = workshop_distinguishable_cfm,
  member_ids = c("F", "M"),
  member_names = c("Female", "Male"),
  predictor_name = "collaborative planning",
  outcome_name = "joint MVPA",
  show_standardized_effect = FALSE
)
```

:::{.notes}

- The unconstrained model converged but estimated a negative male joint-MVPA
  residual variance. Female and male reports of their joint activity correlate
  almost perfectly.
- We therefore constrain the two joint-MVPA residual variances to equality.
  This is defensible because both partners report the same jointly performed
  activity using the same measure.
- The shared-level association is observational and is estimated from a small
  sample.

:::

<!-- ca. 2 Minutes -->

## The exchangeable CFM

Positions 1 and 2 are arbitrary. The model must be unchanged when they are
swapped [@ledermannMediationDyadicData2009; @peughAnalyzingMixedDyadic2013].

$$
\begin{aligned}
\nu_{X_1} &= \nu_{X_2}, & \nu_{Y_1} &= \nu_{Y_2} \\
\sigma^2_{\epsilon_{X_1}} &= \sigma^2_{\epsilon_{X_2}}, &
\sigma^2_{\epsilon_{Y_1}} &= \sigma^2_{\epsilon_{Y_2}} \\
\operatorname{Cov}(\epsilon_{X_1},\epsilon_{Y_1})
&= \operatorname{Cov}(\epsilon_{X_2},\epsilon_{Y_2})
\end{aligned}
$$

- All four loadings remain fixed to 1.
- Ordinary wide-data **df = 6**; **df = 0** against the exchangeable saturated
  model.
- Usual exchangeability test: $\Delta df = 5$; here: $\Delta df = 4$
  ($\sigma^2_{\epsilon_Y}$ is already equal).

:::{.notes}

- Equality is imposed on the two same-member residual covariances. Because the
  corresponding residual variances are also equal, the residual correlations
  are equal as a consequence.
- `lavaan` reports 6 degrees of freedom because its default saturated reference
  model is fully unrestricted. Six observed-moment restrictions are required
  for exchangeability.
- The exchangeable CFM reproduces the exchangeable saturated model, so it has
  zero degrees of freedom relative to that appropriate reference model. Its
  ordinary chi-square is therefore a test of exchangeability, not an additional
  test of the basic CFM structure.
- An unconstrained distinguishable CFM has df = 1, giving the usual five-degree-
  of-freedom exchangeability test. Our distinguishable model already equates the
  two outcome residual variances and has df = 2, so the comparison on the next
  slide tests the four remaining equality constraints.
- A well-fitting constrained model does not make substantively meaningful roles
  exchangeable; that decision must also follow the research question.

:::

## Fitted exchangeable CFM

```{r}
#| label: fitted-exchangeable-cfm
#| echo: false
#| dev: svg
#| fig-width: 10
#| fig-height: 5.35
#| out-width: 100%
#| fig-align: center
#| fig-alt: "Fitted exchangeable Common Fate Model for collaborative planning and joint MVPA. Female and male indicator intercepts, corresponding residual variances, and same-member residual covariances are constrained to equality."

exchangeable_cfm_model <- '
  # Shared dyad levels: fix all four loadings to 1.
  planning_level =~
    1 * collaborative_planning_female +
    1 * collaborative_planning_male
  joint_mvpa_level =~
    1 * mvpa_joint_female +
    1 * mvpa_joint_male

  # Association between the two shared levels.
  joint_mvpa_level ~ b_level * planning_level

  # Equal indicator intercepts across positions.
  collaborative_planning_female ~ i_planning * 1
  collaborative_planning_male ~ i_planning * 1
  mvpa_joint_female ~ i_mvpa * 1
  mvpa_joint_male ~ i_mvpa * 1

  # Equal residual variances across positions.
  collaborative_planning_female ~~ e_planning * collaborative_planning_female
  collaborative_planning_male ~~ e_planning * collaborative_planning_male
  mvpa_joint_female ~~ e_mvpa * mvpa_joint_female
  mvpa_joint_male ~~ e_mvpa * mvpa_joint_male

  # Equal same-member residual covariances across positions.
  collaborative_planning_female ~~ c_female * mvpa_joint_female
  collaborative_planning_male ~~ c_male * mvpa_joint_male
  c_female == c_male
'

workshop_exchangeable_cfm <- lavaan::sem(
  exchangeable_cfm_model,
  data = cfm_data,
  missing = "fiml",
  meanstructure = TRUE
)

cfm_model_comparison <- lavaan::lavTestLRT(
  workshop_distinguishable_cfm,
  workshop_exchangeable_cfm
)

cfm_chisq_difference <- cfm_model_comparison$`Chisq diff`[2]
cfm_df_difference <- cfm_model_comparison$`Df diff`[2]
cfm_comparison_p <- cfm_model_comparison$`Pr(>Chisq)`[2]

cfm_distinguishable_aic <- lavaan::fitMeasures(
  workshop_distinguishable_cfm, "aic"
)
cfm_exchangeable_aic <- lavaan::fitMeasures(
  workshop_exchangeable_cfm, "aic"
)
cfm_distinguishable_bic <- lavaan::fitMeasures(
  workshop_distinguishable_cfm, "bic"
)
cfm_exchangeable_bic <- lavaan::fitMeasures(
  workshop_exchangeable_cfm, "bic"
)

draw_cfm_diagram(
  model = workshop_exchangeable_cfm,
  parameter_labels = c(
    level = "b_level",
    residual_1 = "c_female",
    residual_2 = "c_male"
  ),
  member_ids = c("F", "M"),
  member_names = c("Female", "Male"),
  predictor_name = "collaborative planning",
  outcome_name = "joint MVPA",
  show_standardized_effect = FALSE
)
```

::: {.cfm-constraint}
**No detectable loss of fit:**
$\Delta\chi^2(`r cfm_df_difference`) = `r sprintf("%.2f", cfm_chisq_difference)`$,
$p = `r sub("^0", "", sprintf("%.3f", cfm_comparison_p))`$.
**Parsimony favours the exchangeable model:**
AIC `r sprintf("%.2f", cfm_distinguishable_aic)` $\rightarrow$
`r sprintf("%.2f", cfm_exchangeable_aic)`;
BIC `r sprintf("%.2f", cfm_distinguishable_bic)` $\rightarrow$
`r sprintf("%.2f", cfm_exchangeable_bic)`.
:::

:::{.notes}

- The four additional equality constraints fit the data very well. The
  likelihood-ratio test provides no evidence that they worsen fit.
- AIC and BIC are lower for the exchangeable model because it achieves almost
  the same likelihood with four fewer free parameters.
- This is evidence that pooling the female and male measurement parameters is
  compatible with this small sample; it does not establish that gender is
  substantively irrelevant or that the partners were arbitrarily assigned.
- The fitted shared-level association remains observational.

:::

<!-- ca. 3 Minutes -->


## Honorable mention: The truth and bias model 

<!-- add ref, basic idea, example graphic and basic interpretation of paths -->

<!-- ca. 2 minutes -->



<!----------------- APPLIED PART (plus MLM vs. SEM and software) 

TODO: Probably create a second slide deck!

- DHARMa plots!?! -> as required package for workshop -> link to materials or do in tutorial? Also check AR1 etc., etc. can dharma do this if resid is set to almost zero?

- glmmTMB -> as required 
- dyadMLM -> as required (via pak, see instructions; or should we already try a CRAN release now?? We may want to still add additional functions like rotating variance structs back, so probably not). -> maybe just require pak and we install newest version on workshop-day?
- dplyr -> required
- tibble -> required
- lavaan -> required
- ggplot2 -> required

-------------->

<!-- ILD-part: 

- see dev note on ar1 processes careful to consider degrees of freedom! relevant here?
  - TODO @codex: are degrees of freedom alreaedy important in the cross-sectional case and should we teach anything regarding those?
- see plots on: https://quantdev.ssri.psu.edu/tutorials/apa-ati-intensive-longitudinal-data-session-m-dyadic-multilevel-model

-->

<!-- Excercises part

Create exercises with data visalization, etc. Panel plots, histograms
correlations (wb corr)?

-->

## References

::: {#refs}
:::

## Continue with the applied tutorial {.rq-slide}

::: {.workshop-next}

Thank you for your attention!

[Continue to Part II: Applied Tutorial →](https://pascal-kueng.github.io/dyadMLM/workshop/applied/){.workshop-next-link}
:::



# Appendix

## Distinguishable APIM: residual covariance

### Within-dyad residual covariance block (shared across dyads):
$$
\operatorname{Cov}
\begin{pmatrix}
\epsilon_{Fi} \\
\epsilon_{Mi}
\end{pmatrix}
= \boldsymbol{\Sigma}_{\epsilon}
= \begin{bmatrix}
\sigma_{\epsilon_F}^{2}
& \rho_{\epsilon_F\epsilon_M}\sigma_{\epsilon_F}\sigma_{\epsilon_M} \\
\rho_{\epsilon_F\epsilon_M}\sigma_{\epsilon_F}\sigma_{\epsilon_M}
& \sigma_{\epsilon_M}^{2}
\end{bmatrix}
$$

### Residual covariance matrix for all dyads (first three shown):
$$
\boldsymbol{\Sigma}_{\mathrm{model}}
= \begin{bmatrix}
\boldsymbol{\Sigma}_{\epsilon} & \boldsymbol{0} & \boldsymbol{0} & \cdots \\
\boldsymbol{0} & \boldsymbol{\Sigma}_{\epsilon} & \boldsymbol{0} & \cdots \\
\boldsymbol{0} & \boldsymbol{0} & \boldsymbol{\Sigma}_{\epsilon} & \cdots \\
\vdots & \vdots & \vdots & \ddots
\end{bmatrix}
$$

<!-- ca. 1.5 minutes -->

## Exchangeable APIM: residual covariance

For arbitrary member positions 1 and 2, the residual variances must be equal:

$$
\operatorname{Cov}
\begin{pmatrix}
\epsilon_{1i} \\
\epsilon_{2i}
\end{pmatrix}
= \boldsymbol{\Sigma}_{\epsilon,\mathrm{exch}}
= \begin{bmatrix}
\sigma_{\epsilon}^{2}
& \rho_{\epsilon_1\epsilon_2}\sigma_{\epsilon}^{2} \\
\rho_{\epsilon_1\epsilon_2}\sigma_{\epsilon}^{2}
& \sigma_{\epsilon}^{2}
\end{bmatrix}
$$

- The equal diagonal elements express exchangeability.
- The off-diagonal element still allows residual interdependence.
- Across dyads, these blocks form the same block-diagonal model matrix as on
  the previous slide.

<!-- ca. 1.5 minutes -->

## DIM: residual covariance

$$
\begin{aligned}
r_{\mathrm{M}i}
&= \bar{\epsilon}_i
= \frac{\epsilon_{1i}+\epsilon_{2i}}{2}, \\
r_{\mathrm{D},mi}
&= \epsilon_{mi}-r_{\mathrm{M}i}
\quad (m=1,2),
\qquad
r_{\mathrm{D},1i}
= \frac{\epsilon_{1i}-\epsilon_{2i}}{2}
= -r_{\mathrm{D},2i}.
\end{aligned}
$$

Using the arbitrary orientation $r_{\mathrm{D}i}=r_{\mathrm{D},1i}$,
exchangeability gives:

$$
\operatorname{Cov}
\begin{pmatrix}
r_{\mathrm{M}i} \\
r_{\mathrm{D}i}
\end{pmatrix}
= \begin{bmatrix}
\sigma_{r_\mathrm{M}}^{2} & 0 \\
0 & \sigma_{r_\mathrm{D}}^{2}
\end{bmatrix}.
$$

Back-transform to the exchangeable APIM: partner covariance need not be zero.

::: {.covariance-backtransform .dim-backtransform}
$$
\begin{bmatrix}
\color{#1769AA}{\sigma_{r_\mathrm{M}}^{2}+\sigma_{r_\mathrm{D}}^{2}}
& \color{#7B3FA1}{\sigma_{r_\mathrm{M}}^{2}-\sigma_{r_\mathrm{D}}^{2}} \\[0.35em]
\color{#7B3FA1}{\sigma_{r_\mathrm{M}}^{2}-\sigma_{r_\mathrm{D}}^{2}}
& \color{#1769AA}{\sigma_{r_\mathrm{M}}^{2}+\sigma_{r_\mathrm{D}}^{2}}
\end{bmatrix}
=
\begin{bmatrix}
\color{#1769AA}{\sigma_{\epsilon}^{2}}
& \color{#7B3FA1}{\rho_{\epsilon_1\epsilon_2}\sigma_{\epsilon}^{2}} \\[0.35em]
\color{#7B3FA1}{\rho_{\epsilon_1\epsilon_2}\sigma_{\epsilon}^{2}}
& \color{#1769AA}{\sigma_{\epsilon}^{2}}
\end{bmatrix}.
$$
:::

<!-- ca. 2 minutes -->

## DSM: residual covariance

With distinguishable roles, the directed difference may covary with the mean:

$$
r_{\mathrm{M}i}=\frac{\epsilon_{Fi}+\epsilon_{Mi}}{2},
\qquad
r_{\mathrm{D}i}=\epsilon_{Fi}-\epsilon_{Mi}.
$$

$$
\operatorname{Cov}
\begin{pmatrix}
r_{\mathrm{M}i} \\
r_{\mathrm{D}i}
\end{pmatrix}
= \begin{bmatrix}
\sigma_{r_\mathrm{M}}^{2}
& \rho_{r_\mathrm{M}r_\mathrm{D}}
  \sigma_{r_\mathrm{M}}\sigma_{r_\mathrm{D}} \\
\rho_{r_\mathrm{M}r_\mathrm{D}}
  \sigma_{r_\mathrm{M}}\sigma_{r_\mathrm{D}}
& \sigma_{r_\mathrm{D}}^{2}
\end{bmatrix}.
$$

Back-transform to the distinguishable APIM, with
$c_{\mathrm{MD}}
=\rho_{r_\mathrm{M}r_\mathrm{D}}
\sigma_{r_\mathrm{M}}\sigma_{r_\mathrm{D}}$.

::: {.covariance-backtransform}
$$
\begin{bmatrix}
\color{#1769AA}{
  \sigma_{r_\mathrm{M}}^{2}+\frac{1}{4}\sigma_{r_\mathrm{D}}^{2}+c_{\mathrm{MD}}}
& \color{#7B3FA1}{
  \sigma_{r_\mathrm{M}}^{2}-\frac{1}{4}\sigma_{r_\mathrm{D}}^{2}} \\[0.35em]
\color{#7B3FA1}{
  \sigma_{r_\mathrm{M}}^{2}-\frac{1}{4}\sigma_{r_\mathrm{D}}^{2}}
& \color{#B45309}{
  \sigma_{r_\mathrm{M}}^{2}+\frac{1}{4}\sigma_{r_\mathrm{D}}^{2}-c_{\mathrm{MD}}}
\end{bmatrix}
=
\begin{bmatrix}
\color{#1769AA}{\sigma_{\epsilon_F}^{2}}
& \color{#7B3FA1}{
  \rho_{\epsilon_F\epsilon_M}\sigma_{\epsilon_F}\sigma_{\epsilon_M}} \\[0.35em]
\color{#7B3FA1}{
  \rho_{\epsilon_F\epsilon_M}\sigma_{\epsilon_F}\sigma_{\epsilon_M}}
& \color{#B45309}{\sigma_{\epsilon_M}^{2}}
\end{bmatrix}.
$$
:::

<!-- ca. 2 minutes -->
