---
title: "Dyadic Data Analysis with R"
subtitle: "Workshop Exercises"
author:
  - "Pascal Küng"
  - "Turu Stadler"
format:
  html:
    embed-resources: true
    toc: true
    toc-title: "Exercises"
editor: source
---

# Exercises

The workshop exercises will be added here.

Install necessary packages (run this block):

```{r}
install.packages("dyadMLM", repos = c(
  "https://pascal-kueng.r-universe.dev",
  "https://cloud.r-project.org"
  )
)

# TODO: add all other packages they need, like easystats, DHARMa, and glmmTMB
```


# Getting started

Briefly explain how a .Rmd file works, and that they can run chunks, and how
to insert new chunks with ctrl + alt + i (how on Mac?)

```{r}

```


Load all required package:

```{r}
library(glmmTMB)
library(dyadMLM)
library(easystats)
library(DHARMa)
library(dplry)

```


# Loading, expecting, and preparing the data

Follow a simplified workflow:

1. Define dyads, distinguishability, and research questions
2. Prepare the data and validate the dyadic structure
3. Describe and visualize measures and observed relationships
4. Estimate the model
5. Verify model adequacy, distinguishability, and robustness
6. Visualize, interpret, and report results

# 1. Define dyads, distinguishability, and research questions

You may use any of the variables in the dataset as outcome or predictor. 

Inspect dataset 

```{r}
data
```


Choose your RQ: 

> [Insert your RQ]

# 2. Prepare the data and validate the dyadic structure

You may use `dyadMLM::prepare_dyad_data()`. Load the package:

```{r}
library(dyadMLM)
```

For help with the arguments and how to use the function, run: 

```{r}
?prepare_dyad_data()
```

Adding a `?` before any package function will produce a helpful help-page.

---

Prepare and validate the data for a **distinguishable** model now. 

Hint: if you set `include_arbitrary_member_contrast = TRUE` you can use the 
same data to fit both distinguishable and exchangeable models later.

```{r}


```

Inspect and see if columns make sense: 

```{r}

```


# 3. Describe and visualize measures and observed relationships

Here, should we provide the code already and only have them enter their
variables? And then ask them to interpret the output?


-> Describe what you can see and how you interpret the summaries and plots: 

[Insert your brief summary here]


# 4. Estimate the model

First, estimate a model for distinguishable dyads: 

```{r}

my_model <- glmmTMB(
  
  #[specify your model here]
  
  , data = # [insert your dataset]
  , family = gaussian()
)

summary(my_model)

```



# 5. Verify model adequacy, distinguishability, and robustness

### Adequacy:

Check if the model converged and if all estimates look plausible.

Then, use DHARMa for residual diagnostics via simulation:

```{r}
# Should we already provide the code?
```

Interpreting these residual diagnostics is not easy. In practice, refer to 
the vignette. For now, just acknowledge whether they seem acceptable or not. 


### Influential dyads

Next, check visually whether results could be driven by few dyads. You can use 
the following function (just run it to create it in your environment):

```{r}
# We should provide the full function
```

Use the function for each effect separately once: 

```{r}

```


Interpret the results. Should you refit with some dyads excluded? 

Optionally, refit the model here. To remove certain dyads from the dataset, 
you may use `easy call`: 

```{r}

```


### Distinguishability

Fit the equivalent model as exchangeable dyads. Hint: you need a 
`.arbitrary_member_contrast` column in the data. You may already have included
one earlier, or you may need to prepare the data freshly. You can check
your column names with `names(your_data)`. 


```{r}


```


Insert your brief conclusion, and choose whether you want to go forward with
the distinguishable or exchangeable model for the rest of the excercise:

[Insert conclusion]

[Insert decision and reasoning]


# 6. Visualize, interpret, and report results

Print a nice fixed effects table using model_parameters (remember to use `?` if you need help with arguments): 

```{r}

```

Print a nice random effects table: 

```{r}

```

Plot the fixed effects: 

```{r}

```


Interpret your results in practical terms. Remember that the interpretation
of the intercepts depends on whether you used grand-mean centred variables.

Your interpretation: 

[insert interpretation]


You have now completed the APIM-workflow!!


## Compute DSM or DIM estimates

Since the transformation to DIM estimate is much sipmler, let's try this. 
Check the output from your **exchangeable** model, and enter your coefficients
here: 

```{r}

b_actor <- # [insert from your model output]
b_partner <- # [insert from your model output]
  
```

Recover the DIM estimates:

```{r}

b_mean <- # [insert formula]
b_dev <- # [insert formula]
  
paste0("The between-dyad effect is: ", b_mean)
paste0("The within-dyad effect is: ", b_dev)

```


Interpret the effects: 

[Insert interpretation]


Optionally, if you want to fit a DIM or DSM from scratch, see: Vignette of dyadMLM


# Intensive longitudinal dyads

- Now use the dataset xyz...






