---
title: "Checking covariance pooling across dyad compositions"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
params:
  results_directory: "../../dev/diagnostic_checks/simulation-studies/report-data/covariance-pooling"
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 12, fig.height = 6, dpi = 160, out.width = "100%")
read_results <- function(filename) {
  read.csv(file.path(params$results_directory, filename))
}
conditions <- read_results("conditions.csv")
summary <- read_results("summary.csv") |> left_join(conditions, by = "condition")
fit_counts <- read_results("fits.csv")
scenario_labels <- c(null = "Correct pooling",
  variance_small = "Female/male SD ratio 1.25", variance_large = "Female/male SD ratio 1.50",
  correlation_small = "Correlations 0.10 / 0.30 / 0.50",
  correlation_large = "Correlations -0.10 / 0.30 / 0.70")
summary <- summary |> mutate(scenario = factor(scenario, levels = names(scenario_labels),
  labels = scenario_labels))
statistic_summary <- read_results("stats-summary.csv") |>
  left_join(conditions, by = "condition") |>
  mutate(scenario = factor(scenario, levels = names(scenario_labels), labels = scenario_labels))
false_alarm_results <- summary |> filter(allocation == "balanced", scenario == "Correct pooling")
false_alarm_rates_at_100 <- false_alarm_results |> filter(n_dyads == 100) |>
  select(method, rate) |> tibble::deframe()
plot_rates <- function(results, methods, y_label, y_maximum = 1) {
  results <- results |> filter(method %in% names(methods)) |>
    mutate(method = factor(method, levels = names(methods), labels = methods))
  ggplot(results, aes(n_dyads, rate, colour = method, linetype = method)) +
    geom_line() +
    geom_errorbar(aes(ymin = lower, ymax = upper), width = 5, show.legend = FALSE) +
    geom_point(size = 2, show.legend = FALSE) +
    scale_x_continuous(breaks = c(40, 60, 100, 200, 400)) +
    scale_y_continuous(labels = scales::label_percent(), limits = c(0, y_maximum)) +
    labs(x = "Total number of dyads", y = y_label, colour = NULL,
      linetype = NULL) +
    theme_bw(base_size = 12) +
    theme(legend.position = "bottom", legend.key.width = grid::unit(1.5, "cm"),
      panel.grid.minor = element_blank())
}
percentage_table <- function(results, methods, identifiers) {
  results |> filter(method %in% names(methods)) |>
    mutate(method = factor(method, levels = names(methods), labels = methods),
      percentage = ifelse(checked > 0, sprintf("%.1f", 100 * rate), "—")) |>
    arrange(scenario, n_dyads, method) |>
    select(all_of(identifiers), method, percentage) |>
    tidyr::pivot_wider(names_from = method, values_from = percentage) |>
    rename(Scenario = scenario, any_of(c("Total dyads" = "n_dyads", "Allocation" = "allocation"))) |>
    knitr::kable()
}
```

```{r preliminary, results='asis'}
if (any(conditions$repetitions < 1000)) {
  cat("> **Preliminary run.** These few datasets check that the study works. ",
    "The rates are too imprecise to interpret. The planned study uses 1,000 datasets per condition.\n\n")
}
```

## What the study asks

Can checking each dyad composition reveal covariance differences that an overall
check misses?
We simulate Gaussian outcomes for female-female, female-male, and male-male dyads.
Both models correctly allow different means and actor and partner effects for
each role and composition. **Only the covariance assumptions differ.**

The simpler model uses one exchangeable covariance across all dyads. The fuller
model estimates a separate covariance for each composition, allowing unequal role
variances in female-male dyads. We compare their fit and check whether the simpler
model reproduces the observed summaries, both pooled and by composition.

Each condition uses `r unique(conditions$repetitions)` generated datasets and
`r unique(conditions$reference_draws)` reference simulations per pooled model.

Composition checks can reveal differences hidden by pooling. They can also raise
more false alarms, so higher detection rates alone do not establish a better
formal test. Read the false-alarm and detection results together.

## False alarms when pooling is correct

A flag means at least one observed summary lies outside its middle 95% simulated
range. There are four pooled summaries and 14 composition summaries. Inspecting
more summaries creates more opportunities for a flag, so neither overall rate is
guaranteed to be 5%. The model comparison uses a likelihood-ratio test at 5%.

```{r false-alarms, fig.alt="False-alarm rates for pooled checks, composition checks, and the likelihood-ratio test, by total number of dyads."}
false_alarm_methods <- c(pooled_any = "Pooled check", composition_any = "Composition checks",
  likelihood_ratio = "Model comparison")
false_alarm_axis_maximum <- false_alarm_results |>
  filter(method %in% names(false_alarm_methods)) |>
  summarise(upper = max(upper, na.rm = TRUE)) |>
  pull(upper)
false_alarm_axis_maximum <- max(0.1, ceiling(false_alarm_axis_maximum * 10) / 10)
plot_rates(false_alarm_results, false_alarm_methods, "False-alarm rate", false_alarm_axis_maximum) +
  geom_hline(yintercept = 0.05, colour = "grey60", linetype = "dashed") +
  annotate("text", x = 400, y = 0.05, label = "5%", hjust = 1, vjust = 1.3, colour = "grey40") +
  scale_colour_manual(values = rep("black", 3)) +
  scale_linetype_manual(values = c("solid", "dotted", "longdash"))
```

At 100 total dyads, the composition checks raised a false alarm in
`r scales::percent(false_alarm_rates_at_100[["composition_any"]], accuracy = 0.1)`
of datasets. The model comparison did so in
`r scales::percent(false_alarm_rates_at_100[["likelihood_ratio"]], accuracy = 0.1)`.

Points use only datasets with a usable fit and check. Missing results do not count
as agreement. Bars show 95% intervals for uncertainty from the finite number of
study datasets. Fit counts are below.

## When covariance pooling is wrong

We vary either the female/male residual SD ratio within female-male dyads or the
partner correlations across compositions. These differences leave the pooled
covariance unchanged, so the pooled check has no population mismatch to detect.
Correlation labels list female-female, female-male, and male-male dyads, in that order.

For composition checks, detection requires a flag on a truly affected summary
**in the correct direction**. The model comparison tests the covariance restrictions
together and does not identify a direction. At 100 total dyads, each composition
contains about 33 dyads.

```{r alternatives, fig.height=8, fig.alt="Correct-direction detection by composition checks and rejection by the model comparison, across sample sizes and four covariance conditions."}
summary |> filter(allocation == "balanced", scenario != "Correct pooling") |>
  plot_rates(c(expected_direction = "Composition checks (correct direction)",
    likelihood_ratio = "Model comparison"), "Datasets with a detected mismatch") +
  facet_wrap(~scenario, ncol = 2) +
  scale_colour_manual(values = c("#D55E00", "#009E73")) +
  scale_linetype_manual(values = c("solid", "dashed"))
```

<details>
<summary>Direction and additional flags</summary>

Opposite-direction flags concern affected summaries but point the wrong way.
A dataset can have both correct- and opposite-direction flags.

The last two columns count any composition flag when the other method did not
flag, using datasets where both methods were available. All entries are percentages.

```{r directions}
summary |> filter(allocation == "balanced", scenario != "Correct pooling",
  n_dyads %in% c(100, 400)) |>
  percentage_table(c(expected_direction = "Correct direction",
    opposite_direction = "Opposite direction",
    composition_without_pooled = "Composition only: vs pooled",
    composition_without_lrt = "Composition only: vs comparison"), c("scenario", "n_dyads"))
```

</details>

### Which summaries are flagged?

This plot counts **any flag for each summary**, in either direction, at 100 total
dyads. Outlines mark quantities that truly differ from the pooled model.
A darker cell means more frequent flags.

```{r localization, fig.height=7, fig.alt="Any-direction flag rates for each composition summary at 100 total dyads. Outlined cells identify quantities that truly differ from the pooled model."}
localization_results <- statistic_summary |> filter(allocation == "balanced", n_dyads == 100,
  scenario != "Correct pooling", composition != "All dyads") |>
  mutate(statistic = sub(" \\((exchangeable|about zero|female minus male|female and male)\\)$", "", statistic),
    statistic = sub("Dyad-average/role-difference", "Average/difference", statistic))
ggplot(localization_results, aes(scenario, statistic, fill = rate)) + geom_tile(colour = "white") +
  geom_tile(data = filter(localization_results, expected_direction != 0),
    fill = NA, colour = "grey35", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.0f%%", 100 * rate), colour = rate > 0.6), size = 3.5) +
  facet_grid(composition ~ ., scales = "free_y", space = "free_y") +
  scale_x_discrete(labels = function(x) sub("(SD ratio |Correlations )", "\\1\n", x)) +
  scale_fill_gradient(low = "white", high = "#0072B2", limits = c(0, 1),
    labels = scales::label_percent(), name = "Any flag for each summary") +
  scale_colour_manual(values = c("black", "white"), guide = "none") +
  labs(x = NULL, y = NULL) + theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), legend.position = "bottom")
```

## Unequal composition sizes

At 100 total dyads, we also compare roughly equal groups with 20 female-female,
60 female-male, and 20 male-male dyads. Total sample size can hide how little
information is available for a particular composition.

<details>
<summary>Results with equal and unequal group sizes</summary>

Entries are percentages. A dash means that no rate is available. Correct-direction
flags do not apply when pooling is correct.

```{r allocation}
summary |> filter(n_dyads == 100) |>
  percentage_table(c(composition_any = "Any composition flag",
    expected_direction = "Correct direction", likelihood_ratio = "Model comparison"),
    c("scenario", "allocation"))
```

</details>

## Study details

<details>
<summary>Data, models, and interpretation</summary>

Every generated dataset is prepared with `prepare_dyad_data()`. Both models use
the package's covariance structures and `dispformula = ~0`. The pooled model has
one shared dyad effect and one arbitrary member-contrast effect. The fuller model
has separate shared and contrast effects for each same-role composition and an
unstructured female-male block. They estimate two and seven covariance parameters,
respectively. The likelihood-ratio test uses five degrees of freedom.

Predictors are standard normal, with partner correlation 0.3. Residual errors are
independent of predictors. The true mean coefficients are:

| Composition | Role | Intercept | Actor effect | Partner effect |
| --- | --- | ---: | ---: | ---: |
| Female-female | Female | 0.2 | 0.5 | 0.3 |
| Female-male | Female | 0.4 | 0.6 | 0.2 |
| Female-male | Male | -0.2 | 0.4 | 0.4 |
| Male-male | Male | -0.1 | 0.3 | 0.2 |

Balanced allocations use these counts:

```{r composition-counts}
conditions |> filter(allocation == "balanced") |>
  distinct(n_dyads, female_female_dyads, female_male_dyads, male_male_dyads) |>
  knitr::kable(col.names = c("Total dyads", "Female-female", "Female-male", "Male-male"))
```

All residual SDs are 1 and correlations are 0.3 under correct pooling. In the SD
conditions, female-male variances sum to 2 and their covariance remains 0.3.
Holding covariance fixed while changing the SDs also slightly changes their correlation.
Every condition retains a pooled population variance of 1 and covariance of 0.3.

Checks subtract fitted fixed-effect predictions. Pooled and composition checks use
the same reference simulations with parameters held fixed, without refitting.
The fully flexible covariance model is used for model comparison, not checked
against the same moments it estimates. This follows the rationale for choosing
informative summaries in the
[Stan guide](https://mc-stan.org/docs/stan-users-guide/posterior-predictive-checks.html#which-statistics-to-test).

These results assume a correct mean model. In practice, omitted predictors can
also cause mismatches. Agreement does not show that fixed effects can be pooled
or that the model fits well in every respect.

</details>

<details>
<summary>Fitting and saved results</summary>

```{r fitting}
fit_counts |> group_by(model) |>
  summarise(across(c(attempted, usable, with_warnings, check_failures,
    comparison_failures), sum), .groups = "drop") |>
  knitr::kable(col.names = c("Model", "Attempted", "Usable", "Warnings",
    "Check failures", "Comparison failures"))
```

Rates exclude failed fits, convergence or Hessian problems, and unavailable
checks. Warnings are counted separately and do not necessarily invalidate a fit.
The [study script](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/covariance-pooling/run.R) records per-dataset results and exports
the tables used here. Re-rendering this report changes no fitted models or results.

</details>
