---
title: "Covariance-pooling checks across outcome families"
description: "Composition checks and model comparison when one covariance is pooled across dyad compositions."
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
params:
  gaussian_directory: "../../dev/diagnostic_checks/simulation-studies/report-data/covariance-pooling"
  generalized_directory: "../../dev/diagnostic_checks/simulation-studies/report-data/generalized-covariance-pooling"
---

```{r setup, include=FALSE}
library(dplyr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 13, fig.height = 5.5, dpi = 180, fig.retina = 1, out.width = "100%")
source("../../dev/diagnostic_checks/simulation-studies/covariance-pooling/plot.R")
read_results <- function(directory, filename) read.csv(file.path(directory, filename))

conditions <- bind_rows(
  read_results(params$gaussian_directory, "conditions.csv") |> mutate(family = "gaussian"),
  read_results(params$generalized_directory, "conditions.csv"))
# Other families compare methods on the datasets where all of them were available.
summary <- bind_rows(
  read_results(params$gaussian_directory, "summary.csv") |> mutate(family = "gaussian",
    method = recode(method, composition_any = "checks", likelihood_ratio = "comparison",
      pooled_any = "overall")),
  read_results(params$generalized_directory, "summary.csv") |> mutate(method = recode(method,
    composition_paired = "checks", likelihood_ratio_paired = "comparison",
    full_paired = "full", pooled_any = "overall")))
scenario_labels <- c(null = "Correct pooling", variance_small = "SD ratio 1.25",
  variance_large = "SD ratio 1.50", correlation_small = "Correlations 0.10 / 0.30 / 0.50",
  correlation_large = "Correlations -0.10 / 0.30 / 0.70")
summary <- summary |> inner_join(conditions, by = c("family", "condition")) |>
  mutate(scenario = factor(scenario, names(scenario_labels), scenario_labels))
rates <- summary |> filter(allocation == "balanced", method %in% c("checks", "comparison", "full"))
fits <- bind_rows(read_results(params$gaussian_directory, "fits.csv") |> mutate(family = "gaussian"),
  read_results(params$generalized_directory, "fits.csv"))
response_settings <- read_results(params$generalized_directory, "settings.csv")
screening <- read_results(params$generalized_directory, "screening.csv")

families <- tibble::tribble(
  ~family, ~label,
  "gaussian", "Gaussian",
  "poisson", "Poisson",
  "compois", "COM-Poisson",
  "truncated_poisson", "Zero-truncated Poisson",
  "tweedie", "Tweedie",
  "bell", "Bell",
  "zi_poisson", "Zero-inflated Poisson"
)
percent_range <- function(values) paste0(paste(unique(round(100 * range(values))), collapse = "–"), "%")
correct_pooling <- filter(rates, scenario == "Correct pooling")
small_sample <- rates |> filter(family != "gaussian", method == "checks", n_dyads == min(n_dyads))
figure_descriptions <- as.vector(rbind(
  paste0(families$label, ": detection of incorrect covariance pooling by number of dyads."),
  paste0(families$label, ": false alarms when pooling is correct, by number of dyads.")
))
```

## What these plots compare

Two models allow different means and actor and partner effects for female-female,
female-male, and male-male dyads. The **pooled model** uses one exchangeable
covariance for all dyads. The **full model** estimates a separate covariance for
each composition. We make pooling wrong by changing the female/male SD ratio in
female-male dyads or the partner correlations across compositions. These changes
leave the pooled covariance unchanged. The
[partner-dependence study](partner-dependence-simulation.html) shows how often the
check detects an omitted partner correlation.

- **Orange:** composition checks of the pooled model,
  `check_partner_dependence(role = gender)`. A flag means at least one of 14
  summaries lies outside its middle 95% simulated range.
- **Green:** model comparison of the pooled and full models, a likelihood-ratio
  test at 5% with `compare_nested_models()`.
- **Grey (false-alarm plots of other families):** the same checks of the full
  model, which has the correct structure.

When pooling was correct, composition checks raised false alarms in
`r percent_range(correct_pooling$rate[correct_pooling$method == "checks"])` of
datasets, and model comparison in
`r percent_range(correct_pooling$rate[correct_pooling$method == "comparison"])`.
The checks flagged wrong pooling more often than model comparison in most
settings, partly because of these extra false alarms. Allowing for them, the
checks mostly did better for Gaussian outcomes with up to 200 dyads, and model
comparison usually did better for the other families with 400 dyads. For the
other families with `r min(small_sample$n_dyads)` dyads, the checks flagged `r percent_range(small_sample$rate[small_sample$scenario == "Correct pooling"])`
of correctly pooled datasets, and no more than
`r percent_range(summarise(filter(small_sample, scenario != "Correct pooling"), rate = max(rate), .by = family)$rate)`
of wrongly pooled ones, depending on the family. So a flag or its absence says little.

Use model comparison when both models can be fitted. The checks show where a
simpler model misfits and still work when the full model cannot be fitted.
Checking all dyads together (`role = NULL`) rarely flagged anything: at most
`r scales::percent(max(summary$rate[summary$method == "overall"], na.rm = TRUE), accuracy = 1)`
in any condition.

<details>
<summary>Study details</summary>

Gaussian outcomes have the covariance on the outcome scale, with residual SDs of 1
and partner correlations of 0.3 under correct pooling. For the other families,
each partner has a Gaussian latent effect on the log scale, with SDs of 0.6 and
correlations of 0.3 under correct pooling. Outcomes are then drawn from the family.
These families passed a
[fitting and recovery screen](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/generalized-covariance-pooling/screen-families.R)
of `r nrow(screening)` candidates.

Dyads are split roughly equally among the three compositions, so 100 total dyads
means about 33 per composition. Predictors are standard normal, with partner
correlation 0.3. Mean coefficients differ by role and composition, and the mean
model is correct throughout. In practice, omitted predictors can also cause
mismatches.

Checks subtract fitted fixed-effect predictions and use simulations from the
fitted model, without refitting. Failed fits, convergence or Hessian problems,
and unavailable checks are excluded. For the other families, all methods use the
same datasets: those where both checks and the model comparison were available.
The pooled model could be checked in nearly all datasets, but the full model often
had no usable fit with 40 or 100 dyads.

The report uses saved results. See the study code for
[Gaussian outcomes](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/covariance-pooling/run.R)
and [other families](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/generalized-covariance-pooling/run.R),
the [plot code](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/covariance-pooling/plot.R),
and the [study notes](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/README.md).

</details>

```{r family-results, results='asis', fig.alt=figure_descriptions}
for (family_index in seq_len(nrow(families))) {
  family_name <- families$family[family_index]
  family_label <- families$label[family_index]
  gaussian <- family_name == "gaussian"
  family_conditions <- filter(conditions, family == family_name)
  family_rates <- filter(rates, family == family_name)
  plot_arguments <- list(family_label = family_label,
    subtitle = if (gaussian) "outcomes, residual covariance" else
      "outcomes, latent covariance on the log scale",
    repetitions = unique(family_conditions$repetitions),
    reference_draws = unique(family_conditions$reference_draws), same_datasets = !gaussian)
  cat("\n\n## ", family_label, "\n\n", sep = "")
  print(do.call(plot_pooling_detection, c(list(
    filter(family_rates, scenario != "Correct pooling", method != "full")), plot_arguments)))
  print(do.call(plot_pooling_false_alarms, c(list(
    filter(family_rates, scenario == "Correct pooling")), plot_arguments)))

  cat("<details>\n<summary>Model settings and checks</summary>\n\n")
  family_fits <- filter(fits, family == family_name)
  if (gaussian) {
    cat("All ", format(sum(family_fits$attempted), big.mark = ","), " fits were usable. The table also counts flags on a summary that truly differed, ",
      "in the expected direction, and results with unequal group sizes (20 female-female, ",
      "60 female-male, and 20 male-male dyads).\n\n", sep = "")
    summary |> filter(family == "gaussian", scenario != "Correct pooling",
        (allocation == "balanced" & n_dyads %in% c(100, 400)) | allocation == "unbalanced",
        method %in% c("checks", "expected_direction", "comparison")) |>
      mutate(dyads = ifelse(allocation == "unbalanced", "100 (20 / 60 / 20)", as.character(n_dyads)),
        percentage = sprintf("%.0f%%", 100 * rate)) |>
      select(scenario, dyads, method, percentage) |>
      tidyr::pivot_wider(names_from = method, values_from = percentage) |>
      arrange(scenario, dyads) |>
      select(scenario, dyads, checks, expected_direction, comparison) |>
      knitr::kable(col.names = c("Setting", "Dyads", "Any flag", "Flag in expected direction",
        "Model comparison")) |> print()
  } else {
    settings <- filter(response_settings, family == family_name)
    cat("**Response:** ", settings$response_settings, "; log link.\n\n", sep = "")
    available <- family_rates |> filter(method == "checks") |>
      summarise(range = paste(unique(range(checked)), collapse = "–"), .by = n_dyads)
    cat("**Datasets with all methods available:** ",
      paste0(available$range, " of ", unique(family_conditions$repetitions), " at ",
        available$n_dyads, " dyads", collapse = "; "), ".\n\n", sep = "")
    if (family_name == "compois") cat("Simulations failed for ", sum(family_fits$check_failures),
      " fitted models (sampler overflow or non-finite draws).\n\n", sep = "")
    if (family_name == "tweedie") cat("Simulations used an equivalent, faster sampler ",
      "([details](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/README.md#tweedie-simulation-repair)).\n\n", sep = "")
    if (family_name == "zi_poisson") cat("Both models estimate a zero-inflation ",
      "intercept without random effects.\n\n", sep = "")
    unavailable <- sum(family_fits$comparison_failures)
    if (unavailable > 0) cat("Model comparison was unavailable for ", unavailable,
      if (unavailable == 1) " dataset" else " datasets",
      " because the full model had a lower log-likelihood than the pooled model.\n\n", sep = "")
  }
  cat("\n\n</details>\n\n")
}
```
