---
title: "Residual and outcome distribution checks"
description: "Flags under correct models, detection of selected mismatches, the effect of PIT centring, and a confirmation run."
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
params:
  results_directory: "../../dev/diagnostic_checks/simulation-studies/report-data/distribution-checks"
---

```{r setup, include=FALSE}
library(dplyr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 13, fig.height = 7, dpi = 180, fig.retina = 1, out.width = "100%")
source("../../dev/diagnostic_checks/simulation-studies/distribution-checks/plot.R")
read_results <- function(filename) read.csv(file.path(params$results_directory, filename))
conditions <- read_results("conditions.csv")
summary <- read_results("summary.csv")
paired <- read_results("paired-comparison.csv")
fits <- read_results("fits.csv")
default_summary <- summary |> filter(
  (check %in% c("residual", "both") & reference == "fitted-centred") |
    (check == "outcome" & reference == "fitted"),
  (view == "pooled" & role == "Pooled") | (view == "roles" & role == "All"))
# Checkpoints are read alphabetically (1, 10, 11, ...), so sort every table.
arrange_conditions <- function(data, ...) arrange(data, family, n_times, dyad_sd,
  match(scenario, distribution_scenario_order), n_dyads, ...)
format_rate <- function(rate, lower, upper, checked) {
  ifelse(checked == 0, "Unavailable", sprintf("%.1f%% (%.1f–%.1f); n = %s",
    100 * rate, 100 * lower, 100 * upper, checked))
}
rate_table <- function(data) {
  data <- arrange_conditions(data, view, match(check, c("residual", "outcome", "both")))
  data |> mutate(Family = unname(distribution_family_labels[family]),
    View = ifelse(view == "pooled", "Pooled", "Either role"),
    Check = recode(check, residual = "Residuals", outcome = "Outcomes", both = "Both checks"),
    Rate = format_rate(rate, rate_lower, rate_upper, checked)) |>
    select(Family, scenario_label, n_dyads, View, Check, Rate)
}
```

## Summary

```{r summary-numbers}
percent_range <- function(rate) paste0(paste(unique(sprintf("%.0f", 100 * range(rate))),
  collapse = "–"), "%")
current_panels <- setdiff(names(distribution_panel_labels), c("histogram", "outliers"))
panel_maximum <- summary |> filter(scenario == "correct", role %in% c("Pooled", "A", "B"),
    statistic %in% current_panels, reference %in% c("fitted-centred", "fitted")) |>
  pull(rate) |> max(na.rm = TRUE)
# Unions of both checks, with all recorded panels or only those the package still shows.
dropped <- read_results("dropped-panels.csv") |> filter(check == "both")
reduced <- filter(dropped, panel_set != "all")
reduced_range <- function(view_name, main_study = TRUE) with(filter(reduced,
  scenario == "correct", view == view_name, (condition %in% conditions$condition) == main_study),
  percent_range(rate))
detection <- filter(reduced, view == "roles", n_times == 1, dyad_sd == .6,
  condition %in% conditions$condition)
detected <- function(n, scenarios, keep = TRUE)
  detection$rate[detection$n_dyads == n & (detection$scenario %in% scenarios) == keep]
# Default references, either role, at 400 dyads; the add-on supplies the partner page.
flag_rate <- function(scenario_name, statistic_name) bind_rows(summary,
    read_results("severe-summary.csv")) |>
  filter(scenario == scenario_name, statistic == statistic_name, n_dyads == 400,
    view == "roles", role == "All", reference %in% c("fitted-centred", "fitted")) |>
  pull(rate) |> distribution_percent()
inference <- read_results("inference.csv")
coverage_range <- function(scenario_name) with(filter(inference, scenario == scenario_name,
  n_dyads == 400, n_times == 1, dyad_sd == .6), percent_range(coverage))
centring <- paired |> filter(check == "residual", statistic == "any", n_times == 1,
  dyad_sd == .6, (view == "pooled" & role == "Pooled") | (view == "roles" & role == "All"))
centring_change <- function(correct) with(filter(centring, (scenario == "correct") == correct),
  sprintf("%.0f", 100 * range(centred_minus_uncentred)))
confirmation <- read_results("confirmation-summary.csv")
unions <- confirmation |> filter(check == "both", statistic %in% c("core", "pages"),
  (view == "pooled" & role == "Pooled") | (view == "roles" & role == "All"))
union_range <- function(view_name, union) with(filter(unions, scenario == "correct",
  view == view_name, statistic == union), percent_range(rate))
```

The study uses `r unique(conditions$repetitions)` datasets per condition and
`r format(unique(conditions$reference_draws), big.mark = ",")` reference simulations per dataset.
Unless noted, this summary counts only the panels the package still shows.

- **Single panels rarely flag a correct model, but flags add up.** Each panel
  flagged at most `r distribution_percent(panel_maximum)` of datasets under
  correct models, pooled or in one role. In the settings studied, at least one
  flag appeared in `r reduced_range("pooled")` of datasets with all observations
  pooled, `r reduced_range("roles")` in either role, and
  `r reduced_range("roles", FALSE)` in either role with actor and partner
  predictor pages (dyad SD 1.2). On fresh seeds, the
  [confirmation run](#confirmation-run) gave `r union_range("pooled", "core")`,
  `r union_range("roles", "core")`, and `r union_range("roles", "pages")`
  (dyad SD 0.6, including ordinal models). So look for a consistent pattern,
  not any single flag.
- **Detection depends strongly on sample size.** With 40 dyads, most
  mismatches were hard to detect: apart from omitted role means of ±0.4
  (`r distribution_percent(detected(40, "mean_large"))`), mismatches were flagged
  in either role in `r percent_range(detected(40, c("correct", "mean_large"), FALSE))`
  of datasets, against `r percent_range(detected(40, "correct"))` under correct models.
  With 400 dyads, most were flagged in
  `r percent_range(detected(400, c("correct", "dispersion_small", "zeros_small"), FALSE))`,
  but NB2 size 10 fitted as Poisson only in
  `r distribution_percent(detected(400, "dispersion_small"))` and 10% extra zeros in
  `r distribution_percent(detected(400, "zeros_small"))`, against
  `r percent_range(detected(400, "correct"))` under correct models.
- **Each problem shows in a few panels**, listed in the help under "Reading
  flags". At 400 dyads, for example, 25% extra zeros flagged the QQ envelope in
  `r flag_rate("zeros_large", "qq")` of datasets but the zero count in
  `r flag_rate("zeros_large", "zeros")`, and an omitted partner effect (dyad SD
  1.2) flagged the PIT quartiles across the partner's predictor in
  `r flag_rate("omitted_partner", "partner_predictor_quantiles")` but the QQ
  envelope in `r flag_rate("omitted_partner", "qq")`. Panels suggest possible
  problems, not their cause.
- **A flag, or its absence, does not show whether estimates or SEs are wrong.**
  Harm depended on the kind of mismatch. At 400 dyads, omitted role means of ±0.4
  were flagged in `r distribution_percent(detected(400, "mean_large"))` of datasets,
  yet actor and partner intervals covered `r coverage_range("mean_large")`
  (predictors were distributed alike in both roles). NB2 size 10 fitted as
  Poisson lowered coverage to `r coverage_range("dispersion_small")` but was
  flagged in only `r distribution_percent(detected(400, "dispersion_small"))`.
  After a flag, fit the model the pattern suggests and compare the estimates and
  SEs you report.
- **Centring adds a few false alarms and helps detect some mismatches.** Median
  centring is applied automatically when the model has a freely estimated
  overall location (not for ordinal models). It raised the residual check's flag
  rate (all recorded panels) by `r paste(centring_change(TRUE), collapse = "–")`
  percentage points under correct models and by up to `r centring_change(FALSE)[2]`
  under mismatches, mainly omitted role means and extra zeros. Residuals then
  lose their overall median shift, so single panels can flag less often (see
  [Effect of median centring](#effect-of-median-centring)).

## What these plots compare

`check_dyad_residuals()` compares probability integral transform (PIT) residuals with
model simulations. `check_dyad_outcomes()` compares distributions and summaries on the
outcome scale. We study how often their numerical panels flag a departure under
correct Gaussian and negative-binomial models, and how often they detect selected
model mismatches. Both checks use the **same datasets, fitted models, and complete
reference simulations**.

We compare pooling all observations with checking the two partner roles
separately. We also compare the default residual check with the same calculation
without median centring. This isolates the adjustment's effect because both
versions use identical PIT randomization.

A **flag** means that an observed scalar falls strictly outside its simulated
95% limits, or a curve leaves its global envelope. An **any flag** rate counts
datasets with at least one flagged numerical panel. A flag in either role counts
once. Neither multiple panels nor multiple roles have a joint 5% error rate.
The ECDF display and count category ranges are excluded from these rates.

These are descriptive checks. Fitting can make their references conservative or
liberal, and more flags under a mismatch do not alone establish a better check.
Compare detection with the correct-model results. This focused study does not
validate other families, missing-data patterns, or every form of dependence.
The [partner-dependence study](partner-dependence-simulation.html) evaluates
omitted partner correlation, and the [covariance-pooling study](covariance-pooling.html)
evaluates differences between dyad compositions.

<details>
<summary>Study design and interpretation</summary>

The main conditions use 40, 100, or 400 complete dyads, each containing roles A
and B. Standard-normal predictors correlate 0.3 between partners. Actor and
partner effects are 0.5 and 0.3. A shared normal dyad intercept has SD 0.6.
Gaussian outcomes have intercept 1 and residual SD 1. Negative-binomial outcomes
use a log link, intercept log(3), and conditional variance
$\mu + \mu^2/k$, where $k$ is the size parameter.

Gaussian mismatches either omit role shifts of ±0.2 or ±0.4, or impose a common
residual SD when the true role SD ratio is 1.25 or 1.50. The variance conditions
keep the average residual variance at 1. Count mismatches either fit a Poisson
model to negative-binomial outcomes with size 10 or 3, or omit an extra-zero
probability of 0.10 or 0.25 from a negative-binomial model. One feature changes at
a time. The count mean before adding extra zeros is adjusted to preserve the
overall conditional expectation. The settings table below records the baseline
size and fitted family.

Two additional correct-model conditions increase the dyad SD to 1.2 at 100
dyads. Two Gaussian longitudinal conditions use 100 dyads and six occasions,
with either a correct mean model or omitted role shifts of ±0.2. They add an
independent stationary AR(1) process for each person, with SD 0.6 and lag-one
correlation 0.5, and independent residual noise with SD 0.8. The fitted model
includes the shared dyad intercept and the person-specific AR(1) process.

Reference simulations redraw all random effects while keeping predictors and
model parameters fixed. Models are not refitted within the reference simulations.
For correct-model conditions, a separate reference bank uses the true generating
parameters, without PIT centring. These known-parameter checks are run even when
fitting fails. They distinguish the behavior of the reference calculation from
the effects of estimating parameters.

Residual panels include the QQ envelope, histogram envelope, outlier count,
mean PIT distance, and the quartile and distance envelopes against predicted
outcomes. No additional predictor pages are requested. Outcome panels include
the SD after subtracting fixed-effect predictions, largest absolute deviation,
zero count where available, and count category frequencies. The package has
since removed the histogram, the outlier count, and count category displays (see
[Panels removed after the study](#panels-removed-after-the-study)).

Rates use available checks. Failed fits, convergence or Hessian problems, and
failed checks are recorded and excluded without retry. Known-parameter and
fitted-reference rates can therefore have different denominators. Bars show 95%
Wilson intervals for Monte Carlo uncertainty.
Counts and uncertainty are reported following the guidance of
[Morris, White, and Crowther (2019)](https://doi.org/10.1002/sim.8086).

The report reads saved CSV files and does not fit models. See the
[study code](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/distribution-checks/run.R),
[plot code](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/distribution-checks/plot.R),
and [study notes](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/README.md).

```{r settings}
conditions |> filter(n_times == 1, dyad_sd == .6) |> arrange_conditions() |>
  distinct(family, scenario_label, role_mean, sd_ratio, nb_size, zero_probability, fit_family) |>
  mutate(nb_size = ifelse(family == "gaussian", NA_real_, nb_size),
    family = unname(distribution_family_labels[family])) |>
  knitr::kable(col.names = c("Generated family", "Setting", "Role shift (±)",
    "SD ratio", "NB size", "Extra-zero probability", "Fitted family"), digits = 2)
```

</details>

## Correct models

Each row below is one numerical panel with all observations pooled. Grey points
use known parameters, blue points use the fitted model with the package defaults,
and orange points omit PIT centring. The dashed line marks 5% as a benchmark,
not a guaranteed fitted-model flag rate. The outcome summaries and original-scale
outlier counts do not depend on PIT centring.

```{r correct-panels, fig.height=8, fig.alt='Individual-panel flag rates under correct Gaussian and negative-binomial models, comparing known parameters, fitted defaults, and uncentred fitted PIT residuals.'}
plot_distribution_panels(summary)
known <- summary |> filter(reference == "known", role != "All",
  statistic %in% names(distribution_panel_labels), checked > 0)
percent <- function(rate) sprintf("%.1f%%", 100 * rate)
fitted_levels <- default_summary |> filter(scenario == "correct", view == "pooled",
    statistic %in% c("response_sd", "zeros", "mean_distance")) |>
  group_by(statistic) |> summarise(rate = max(rate, na.rm = TRUE))
fitted_level <- function(name) percent(fitted_levels$rate[fitted_levels$statistic == name])
correct_both <- default_summary |> filter(scenario == "correct", check == "both", statistic == "any")
both_range <- function(name) with(filter(correct_both, view == name),
  sprintf("%.0f–%.0f%%", 100 * min(rate), 100 * max(rate)))
```

With known parameters, the limits come from the model that generated the data,
so each panel should flag at most about 5% of datasets. Across all
`r nrow(known)` known-parameter rates, the average is `r percent(mean(known$rate))`.
This checks the implementation; it is not a finding about fitted references.

With fitted models and all observations pooled, the response SD, zero count, and
mean PIT distance flag in at most `r fitted_level("response_sd")`,
`r fitted_level("zeros")`, and `r fitted_level("mean_distance")` of datasets.
The fitted model already matches these overall levels.

Reading every panel of both checks for both roles, including the since-removed
histogram and outlier count, at least one flag appears in
`r both_range("roles")` of datasets under every correct model. With all
observations pooled, this is `r both_range("pooled")`.

<details>
<summary>Any flags under correct models</summary>

The table counts any numerical flag within each check, or across both checks.
The latter is a union within each dataset, not a sum of the two rates. Role-based
rates count either role, and their additional comparisons can increase the rate.
Each entry gives the percentage, its 95% Wilson interval, and the denominator.

```{r correct-any}
default_summary |> filter(scenario == "correct", n_times == 1, dyad_sd == .6,
    statistic == "any") |>
  rate_table() |> select(-scenario_label) |>
  knitr::kable(col.names = c("Family", "Dyads", "View", "Check", "Any flag; datasets checked"))
```

</details>

## Detecting model mismatches

The plots count any numerical flag for each default check. Solid lines count a
flag in either role, while dotted lines pool all observations. These are different
sets of comparisons, so the role-based check may flag more often even under a
correct model. Grey lines repeat the correct-model rate of the same check and
view, the false-alarm baseline. These rates include the since-removed PIT
histogram and outlier count.

```{r gaussian-detection, fig.alt='Gaussian mismatch detection by sample size, comparing pooled and role-based residual and outcome checks with the correct-model baseline.'}
plot_distribution_detection(summary, "gaussian")
pooled_outcome <- default_summary |> filter(family == "gaussian", n_times == 1,
    dyad_sd == .6, check == "outcome", view == "pooled", statistic == "any") |>
  group_by(correct = scenario == "correct") |>
  summarise(rate = max(rate, na.rm = TRUE))
```

Pooled outcome checks flag at most
`r distribution_percent(pooled_outcome$rate[!pooled_outcome$correct])` of datasets
under these Gaussian mismatches and at most
`r distribution_percent(pooled_outcome$rate[pooled_outcome$correct])` under the
correct model. The fitted model already reproduces the pooled SD, so pooled
outcome summaries have little to detect.

```{r count-detection, fig.alt='Negative-binomial mismatch detection by sample size, comparing pooled and role-based residual and outcome checks with the correct-model baseline.'}
plot_distribution_detection(summary, "nbinom2")
```

The tables show how often each panel flags each mismatch in either role, as
percentages of datasets at 40 / 100 / 400 dyads with the package defaults. The
correct-model column is the false-alarm baseline.

```{r panel-detection, results='asis'}
separator <- "\u202f/\u202f" # Narrow no-break spaces keep each entry on one line.
panel_rates <- function(rate, checked) if (all(checked == 0)) NA else
  paste(ifelse(checked > 0, round(100 * rate), "–"), collapse = separator)
for (family_name in names(distribution_family_labels)) {
  rates <- default_summary |> filter(family == family_name, n_times == 1, dyad_sd == .6,
      view == "roles", statistic %in% names(distribution_panel_labels)) |>
    arrange_conditions()
  settings <- unique(rates$scenario_label)
  rates |> group_by(scenario_label, statistic) |>
    summarise(Rates = panel_rates(rate, checked), .groups = "drop") |>
    filter(!is.na(Rates)) |>
    tidyr::pivot_wider(names_from = scenario_label, values_from = Rates) |>
    arrange(match(statistic, names(distribution_panel_labels))) |>
    mutate(statistic = distribution_panel_labels[statistic]) |>
    select(statistic, all_of(settings)) |>
    knitr::kable(col.names = c("Panel", settings),
      caption = paste(distribution_family_labels[family_name], "models")) |> print()
}
rates_at <- function(family_name, scenario_name, check_name, statistic_name, view_name = "roles")
  default_summary |> filter(family == family_name, scenario == scenario_name,
      check == check_name, statistic == statistic_name, view == view_name, n_times == 1,
      dyad_sd == .6, n_dyads %in% c(100, 400)) |>
    arrange(n_dyads) |> pull(rate) |> distribution_percent() |> paste(collapse = separator)
```

The main patterns, at 100 / 400 dyads in either role unless noted:

- **Omitted role means** show mainly in the role view of residual checks (±0.2:
  `r rates_at("gaussian", "mean_small", "residual", "any")`). The pooled residual
  view and the outcome checks stay near their correct-model rates
  (`r rates_at("gaussian", "mean_small", "residual", "any", "pooled")` versus
  `r rates_at("gaussian", "correct", "residual", "any", "pooled")`, and
  `r rates_at("gaussian", "mean_small", "outcome", "any")` versus
  `r rates_at("gaussian", "correct", "outcome", "any")`). The exception is the
  per-role response SD at ±0.4 (`r rates_at("gaussian", "mean_large", "outcome", "response_sd")`),
  which reacts to the inflated fitted residual SD.
- **Role SD differences** show most in the per-role response SD and mean PIT
  distance (ratio 1.50: `r rates_at("gaussian", "variance_large", "outcome", "response_sd")`
  and `r rates_at("gaussian", "variance_large", "residual", "mean_distance")`).
- **NB2 overdispersion fitted as Poisson** is partly absorbed by a larger fitted
  dyad SD (see below). Residual checks flag either role in
  `r rates_at("nbinom2", "dispersion_large", "residual", "any")` of datasets for
  size 3, compared with `r rates_at("nbinom2", "correct", "residual", "any")`
  under the correct model.
- **Omitted zero inflation** shows most in the PIT QQ envelope (25% extra zeros:
  `r rates_at("nbinom2", "zeros_large", "residual", "qq")`). The zero count stays
  weak (`r rates_at("nbinom2", "zeros_large", "outcome", "zeros")`) because a
  smaller fitted NB2 size absorbs the extra zeros (see below).

<details>
<summary>What the fitted models estimate</summary>

A fitted model can absorb part of a mismatch into its estimates (medians over
usable fits; data generated with dyad SD 0.6 and average residual variance 1).

```{r fitted-estimates}
fit_at_400 <- function(family_name, scenario_name, column) {
  row <- fits |> filter(reference == "fitted", family == family_name,
    scenario == scenario_name, n_dyads == 400, n_times == 1, dyad_sd == .6)
  sprintf("%.2f", row[[column]])
}
```

At 400 dyads, omitted role means of ±0.4 give a residual SD of
`r fit_at_400("gaussian", "mean_large", "median_dispersion")`, compared with
`r fit_at_400("gaussian", "correct", "median_dispersion")` under the correct model.
Poisson fits to NB2 data with size 3 give a dyad SD of
`r fit_at_400("nbinom2", "dispersion_large", "median_dyad_sd")`, compared with
`r fit_at_400("nbinom2", "correct", "median_dyad_sd")` under the correct count model.
With 25% omitted zero inflation, the NB2 size is
`r fit_at_400("nbinom2", "zeros_large", "median_dispersion")`, compared with
`r fit_at_400("nbinom2", "correct", "median_dispersion")` (a smaller size means
more overdispersion).

</details>

## Are mismatches that distort estimates or SEs detected?

Each fitted model gives 95% Wald intervals for the actor (0.5) and partner (0.3)
effects. **Coverage** is the percentage of intervals that contain the true value
and should be near 95%. The **SE ratio** divides the average SE by the SD of the
estimates; below 1, the SEs are too small. Low coverage with an SE ratio near 1
means the estimates are biased. **Flagged** is the percentage of datasets with any
flag in either check and either role, with the panels the package still shows.

Three severe mismatches were added to distort inference. Omitting the partner
effect biases the actor effect. A residual SD that rises with the actor predictor,
exp(actor − 1), and Poisson fits to NB2 outcomes with size 1 make the SEs too
small. These settings use 100 / 400 dyads, dyad SD 1.2, and a residual page for
each predictor. The correct model gives their false-alarm rate.

```{r consequences-severe}
flag_rates <- filter(reduced, view == "roles")
consequence_table <- function(selected) {
  collapse_sizes <- function(values) paste(ifelse(is.na(values), "–", values), collapse = separator)
  inference |> filter(condition %in% selected) |>
    transmute(condition, family, scenario, scenario_label, n_dyads, n_times, dyad_sd,
      term = sub("_predictor", "", term), coverage = round(100 * coverage),
      se_ratio = sprintf("%.2f", se_ratio)) |>
    tidyr::pivot_wider(names_from = term, values_from = c(coverage, se_ratio)) |>
    left_join(transmute(flag_rates, condition, flagged = round(100 * rate)), by = "condition") |>
    arrange_conditions() |>
    group_by(family, scenario, scenario_label) |>
    summarise(across(c(coverage_actor, coverage_partner, se_ratio_actor, se_ratio_partner,
      flagged), collapse_sizes), .groups = "drop") |>
    arrange(family, match(scenario, c(distribution_scenario_order, "omitted_partner"))) |>
    mutate(family = unname(distribution_family_labels[family])) |>
    select(-scenario) |>
    knitr::kable(col.names = c("Family", "Setting", "Actor coverage (%)",
      "Partner coverage (%)", "Actor SE ratio", "Partner SE ratio", "Flagged (%)"))
}
consequence_table(inference$condition[inference$severe])
```

The main mismatches, at 40 / 100 / 400 dyads:

```{r consequences-main}
consequence_table(conditions$condition[conditions$n_times == 1 & conditions$dyad_sd == .6])
```

## Effect of median centring

```{r centring-loss}
loss <- paired |> filter(scenario != "correct", n_times == 1, dyad_sd == .6,
    n_dyads == 400, check == "residual", statistic %in% names(distribution_panel_labels),
    (view == "pooled" & role == "Pooled") | (view == "roles" & role == "All")) |>
  slice_min(centred_minus_uncentred, n = 1, with_ties = FALSE)
loss_any <- summary |> filter(condition == loss$condition, view == loss$view,
    role == loss$role, check == "residual", statistic == "any") |>
  pull(rate, reference)
```

At 400 dyads, the largest loss of flags after centring is
`r sprintf("%.1f", -100 * loss$centred_minus_uncentred)` percentage points, for the
`r distribution_panel_labels[[loss$statistic]]` under
"`r loss$scenario_label`" (`r ifelse(loss$view == "pooled", "pooled", "either role")`).
Other panels flag more often there, so the residual any flag rate rises from
`r distribution_percent(loss_any[["fitted-uncentred"]])` without centring to
`r distribution_percent(loss_any[["fitted-centred"]])` with it.

## Additional designs and exclusions

The extra conditions check stronger dyad dependence and one longitudinal model.
They extend the baseline study without covering all dependence structures.

```{r additional-designs}
default_summary |> filter((n_times > 1 | dyad_sd > .6), statistic == "any") |>
  mutate(scenario_label = paste0(scenario_label,
    ifelse(n_times > 1, "; six occasions", "; dyad SD 1.2"))) |>
  rate_table() |>
  knitr::kable(col.names = c("Family", "Setting", "Dyads", "View", "Check", "Any flag; datasets checked"))
role_means <- default_summary |> filter(family == "gaussian", n_dyads == 100, dyad_sd == .6,
    check == "residual", view == "roles", statistic == "any")
role_means_rate <- function(scenario_name, occasions) distribution_percent(
  role_means$rate[role_means$scenario == scenario_name & role_means$n_times == occasions])
```

At 100 dyads, the role view of residual checks flags omitted role means of ±0.2
in `r role_means_rate("mean_small", 6)` of datasets with six occasions (correct model:
`r role_means_rate("correct", 6)`), compared with `r role_means_rate("mean_small", 1)`
with one occasion. Six occasions give each role six times as many rows.

```{r fit-failures, results='asis'}
failed <- fits |> filter(reference == "fitted",
    (fit_errors + fit_problems) / attempted > .02) |> arrange_conditions()
if (nrow(failed)) cat("More than 2% of fits had errors or problems in: ",
  paste0(distribution_family_labels[failed$family], ", ", failed$scenario_label, ", ",
    failed$n_dyads, " dyads", ifelse(failed$n_times > 1, ", six occasions", ""),
    ifelse(failed$dyad_sd > .6, ", dyad SD 1.2", ""), " (",
    failed$fit_errors + failed$fit_problems, " of ", failed$attempted, ")", collapse = "; "),
  ". Their rates use the remaining datasets.\n", sep = "")
longitudinal <- filter(fits, n_times > 1, scenario == "correct")
```

In the longitudinal correct model, all
`r longitudinal$fit_problems[longitudinal$reference == "fitted"]` fits with problems
estimated the residual SD at about zero: six occasions separate the person-level
AR(1) process from independent noise only weakly. The known-parameter rates there
still use all `r longitudinal$usable[longitudinal$reference == "known"]` datasets.

<details>
<summary>Complete saved results</summary>

The saved [summary table](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/report-data/distribution-checks/summary.csv)
contains every panel, category, role, and aggregate rate, with denominators and
Monte Carlo standard errors. The
[fit table](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/report-data/distribution-checks/fits.csv)
records availability and median fitted estimates for every condition. The
[paired table](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/report-data/distribution-checks/paired-comparison.csv)
records how often only the centred version, only the uncentred version, both, or
neither flagged the same dataset.

</details>

## Panels removed after the study

```{r dropped-panels}
categories <- default_summary |> filter(family == "nbinom2", scenario == "correct",
  n_times == 1, dyad_sd == .6, check == "outcome", statistic == "category_any")
cost <- dropped |> group_by(condition, scenario, scenario_label, n_dyads, view) |>
  summarise(cost = rate[panel_set == "all"] - rate[panel_set != "all"], .groups = "drop")
example <- dropped |> filter(scenario == "dispersion_small", n_dyads == 400,
  view == "roles") |> arrange(panel_set != "all") |> pull(rate)
```

After this study, the package removed three displays. The PIT histogram shows
the same residual distribution as the QQ plot, and the outlier count can
usually flag only an excess of outcomes beyond the simulated range. Together,
these two added `r paste(sprintf("%.1f", 100 * range(cost$cost[cost$scenario == "correct"])), collapse = "–")`
percentage points of flags under correct models (both checks, pooled or in
either role), while detection of mismatches without them fell by at most
`r sprintf("%.1f", 100 * max(cost$cost[cost$scenario != "correct"]))` points.
For example, NB2 size 10 fitted as Poisson at 400 dyads was flagged in either
role in `r distribution_percent(example[1])` of datasets with all panels and in
`r distribution_percent(example[2])` without them.

Count categories were taken from the observed values, so a rare high count got
its own category and often fell outside its range: at least one category
flagged in `r percent_range(filter(categories, view == "roles", checked > 0)$rate)`
of correct count models in either role, where shown. Counts now use the ECDF,
which has no flag rule and is excluded from the rates. Ordinal outcomes keep
bars with per-category ranges.

The summary uses the remaining panels; the
[dropped-panel table](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/report-data/distribution-checks/dropped-panels.csv)
gives the flag unions with and without the histogram and outlier count.

## Why not use DHARMa directly?

[DHARMa](https://CRAN.R-project.org/package=DHARMa) supports mixed models, but
its residual tests, such as the uniformity and outlier tests, assume
independent residuals. When random effects are redrawn, partners' residuals stay
correlated, so these tests' error rates need not hold, as DHARMa's help also
notes. With DHARMa's default since version 0.5.0, random effects keep their
fitted values; if they absorb all residual variation, the simulations reproduce
the data almost exactly. With random effects redrawn, grouping by dyad
(`recalculateResiduals(group = dyad)`) sums partners, which is valid but lets
role differences cancel. Selecting one role (`sel`) is valid for
cross-sectional dyads but not for repeated measures, and the rotation option
needs the full residual covariance, which for non-Gaussian models usually must
be estimated, imprecisely, from the simulations. dyadMLM instead builds each panel's
reference from whole simulated datasets, which keep the fitted partner and time
dependence; this study did not compare the two approaches.

## Limitations

- Predictors were distributed alike in both roles. When predictor means differ
  by role, omitted role means can bias actor and partner effects, so their good
  coverage here need not hold.
- Actor and partner effects were the same in both roles, so the study does not
  show how well the panels reveal role-specific effects.
- The panels the package still shows were chosen on this study's data, which
  can make their rates look better than they are. The
  [confirmation run](#confirmation-run) checks them on fresh seeds.
- No heavy-tailed residuals, gross outliers, or data errors were simulated, so
  the study does not show whether the remaining panels, such as the QQ plot and
  the largest absolute deviation, reveal them.

## Confirmation run

A confirmation run, whose design was fixed before it started (see the
[study notes](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/README.md#confirmation-run)),
checks the remaining panels on fresh seeds at 100 and 400 dyads with dyad SD
0.6: correct Gaussian, negative-binomial, and five-category ordinal models;
Gaussian models with a role SD ratio of 1.5, a residual SD that rises with the
actor predictor, or an omitted partner effect; and NB2 size 10 fitted as
Poisson. Each dataset is checked pooled and by role, with actor and partner
predictor pages. The *core* union covers QQ, mean PIT distance, PIT quartiles
and distance across predicted outcomes, response SD, largest absolute
deviation, and the zero count where shown; *with pages* adds the four
predictor-page panels. Ordinal category ranges enter no union.

```{r confirmation}
confirmation_fits <- read_results("confirmation-fits.csv")
family_labels <- c(distribution_family_labels, ordinal = "Ordinal")
# A no-break space and a word joiner keep each entry on one line.
unions |> arrange(condition, view, statistic) |>
  transmute(Family = unname(family_labels[family]), scenario_label, n_dyads,
    column = paste(view, statistic),
    rate = sprintf("%.0f\u00a0(%.0f–\u2060%.0f)", 100 * rate, 100 * rate_lower, 100 * rate_upper)) |>
  tidyr::pivot_wider(names_from = column, values_from = rate) |>
  knitr::kable(col.names = c("Family", "Setting", "Dyads", "Pooled: core",
    "Pooled: with pages", "Either role: core", "Either role: with pages"))
# Each mismatch against the correct model of the same family and sample size.
matched <- unions |> filter(view == "roles", statistic == "pages") |>
  group_by(family, n_dyads) |> mutate(baseline = rate[scenario == "correct"]) |>
  ungroup() |> filter(scenario != "correct")
weakest <- slice_min(matched, rate - baseline, n = 1, with_ties = FALSE)
ordinal_categories <- confirmation |> filter(family == "ordinal", statistic == "category_any",
  (view == "pooled" & role == "Pooled") | (view == "roles" & role == "All")) |>
  arrange(n_dyads)
ordinal_rates <- function(view_name) paste(distribution_percent(
  ordinal_categories$rate[ordinal_categories$view == view_name]), collapse = separator)
ties <- filter(read_results("confirmation-maximum-deviation-ties.csv"), family == "ordinal")
failed <- filter(confirmation_fits, usable < attempted)
fit_note <- if (!nrow(failed)) "All fits and checks succeeded." else paste0(
  "Fits or checks failed in ", knitr::combine_words(paste0(family_labels[failed$family],
    ", ", failed$scenario_label, ", ", failed$n_dyads, " dyads (",
    failed$attempted - failed$usable, " of ", failed$attempted, ")")),
  "; their rates use the remaining datasets.")
```

Entries give the percentage of datasets with at least one flag across both
checks, with its 95% Wilson interval, from `r unique(confirmation_fits$repetitions)`
datasets per setting and
`r format(unique(confirmation_fits$reference_draws), big.mark = ",")` reference
simulations per dataset. Under correct models, at least one core panel flagged
`r union_range("pooled", "core")` of datasets pooled and
`r union_range("roles", "core")` in either role; with predictor pages,
`r union_range("pooled", "pages")` and `r union_range("roles", "pages")`. In
either role with pages, mismatches were flagged in
`r percent_range(matched$rate[matched$n_dyads == 100])` of datasets at 100 dyads
and `r percent_range(matched$rate[matched$n_dyads == 400])` at 400; the smallest
difference from the matching correct model was for "`r weakest$scenario_label`" at
`r weakest$n_dyads` dyads (`r distribution_percent(weakest$rate)` versus
`r distribution_percent(weakest$baseline)`). For correct ordinal models, at least
one category fell outside its range in `r ordinal_rates("pooled")` of datasets
pooled and `r ordinal_rates("roles")` in either role (100 / 400 dyads), and the
largest absolute deviation had equal limits in `r percent_range(ties$rate)` of
checks (at least 95% of simulated datasets gave the same value). `r fit_note`
