---
title: "Residual and outcome distribution checks"
description: "Flags under correct models, detection of selected mismatches, and the effect of PIT centring."
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
params:
  results_directory: "../../dev/diagnostic_checks/simulation-studies/report-data/distribution-checks"
---

```{r availability, echo=FALSE, results='asis'}
required_files <- c("conditions.csv", "summary.csv", "paired-comparison.csv", "fits.csv")
if (!all(file.exists(file.path(params$results_directory, required_files)))) {
  cat("**Full results are not yet available.** The focused study is implemented, but its planned run has not been completed and exported. See the [study notes](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/README.md) for the design and running instructions.\n")
  knitr::knit_exit()
}
```

```{r setup, include=FALSE}
library(dplyr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 13, fig.height = 7, dpi = 180, fig.retina = 1, out.width = "100%")
source("../../dev/diagnostic_checks/simulation-studies/distribution-checks/plot.R")
read_results <- function(filename) read.csv(file.path(params$results_directory, filename))
conditions <- read_results("conditions.csv")
summary <- read_results("summary.csv")
paired <- read_results("paired-comparison.csv")
fits <- read_results("fits.csv")
fitted_counts <- filter(fits, reference == "fitted")
complete_run <- setequal(fitted_counts$condition, conditions$condition) &&
  all(fitted_counts$attempted == fitted_counts$repetitions)
default_summary <- summary |> filter(
  (check %in% c("residual", "both") & reference == "fitted-centred") |
    (check == "outcome" & reference == "fitted"),
  (view == "pooled" & role == "Pooled") | (view == "roles" & role == "All"))
format_rate <- function(rate, lower, upper, checked) {
  ifelse(checked == 0, "Unavailable", sprintf("%.1f%% (%.1f–%.1f); n = %s",
    100 * rate, 100 * lower, 100 * upper, checked))
}
rate_table <- function(data) {
  data |> mutate(Family = unname(distribution_family_labels[family]),
    View = ifelse(view == "pooled", "Pooled", "Either role"),
    Check = recode(check, residual = "Residuals", outcome = "Outcomes", both = "Both checks"),
    Rate = format_rate(rate, rate_lower, rate_upper, checked)) |>
    select(Family, scenario_label, n_dyads, View, Check, Rate)
}
```

```{r study-size, results='asis'}
if (!complete_run) {
  cat("**Incomplete results:** some datasets or conditions have not yet been run. The results below describe only the available checks.\n\n")
} else if (any(conditions$repetitions < 500 | conditions$reference_draws < 1000)) {
  cat("**Pilot results:** these runs use fewer than the planned 500 datasets or 1,000 reference simulations per condition. They check the study workflow and are too imprecise for general conclusions.\n\n")
}
cat(if (complete_run) "The saved run uses **" else "The run settings request **",
  paste(unique(conditions$repetitions), collapse = ", "),
  " datasets per condition** and **", paste(unique(conditions$reference_draws), collapse = ", "),
  " reference simulations per dataset**.\n", sep = "")
```

## What these plots compare

`check_dyad_residuals()` compares probability integral transform (PIT) residuals with
model simulations. `check_dyad_outcomes()` compares distributions and summaries on the
outcome scale. We study how often their numerical panels flag a departure under
correct Gaussian and negative-binomial models, and how often they detect selected
model mismatches. Both checks use the **same datasets, fitted models, and complete
reference simulations**.

We compare pooling all observations with checking the two partner roles
separately. We also compare the default residual check with the same calculation
without median centring. This isolates the adjustment's effect because both
versions use identical PIT randomization.

A **flag** means that an observed scalar falls strictly outside its simulated
95% limits, or a curve leaves its global envelope. An **any flag** rate counts
datasets with at least one flagged numerical panel. A flag in either role counts
once. Neither multiple panels nor multiple roles have a joint 5% error rate.
The ECDF display has no numerical limit and is excluded from these rates.

These are descriptive checks. Fitting can make their references conservative or
liberal, and more flags under a mismatch do not alone establish a better check.
Compare detection with the correct-model results. This focused study does not
validate other families, missing-data patterns, or every form of dependence.
The [partner-dependence study](partner-dependence-simulation.html) evaluates
omitted partner correlation, and the [covariance-pooling study](covariance-pooling.html)
evaluates differences between dyad compositions.

<details>
<summary>Study design and interpretation</summary>

The main conditions use 40, 100, or 400 complete dyads, each containing roles A
and B. Standard-normal predictors correlate 0.3 between partners. Actor and
partner effects are 0.5 and 0.3. A shared normal dyad intercept has SD 0.6.
Gaussian outcomes have intercept 1 and residual SD 1. Negative-binomial outcomes
use a log link, intercept log(3), and conditional variance
$\mu + \mu^2/k$, where $k$ is the size parameter.

Gaussian mismatches either omit role shifts of ±0.2 or ±0.4, or impose a common
residual SD when the true role SD ratio is 1.25 or 1.50. The variance conditions
keep the average residual variance at 1. Count mismatches either fit a Poisson
model to negative-binomial outcomes with size 10 or 3, or omit an extra-zero
probability of 0.10 or 0.25 from a negative-binomial model. One feature changes at
a time. The count mean before adding extra zeros is adjusted to preserve the
overall conditional expectation. The settings table below records the baseline
size and fitted family.

Two additional correct-model conditions increase the dyad SD to 1.2 at 100
dyads. Two Gaussian longitudinal conditions use 100 dyads and six occasions,
with either a correct mean model or omitted role shifts of ±0.2. They add an
independent stationary AR(1) process for each person, with SD 0.6 and lag-one
correlation 0.5, and independent residual noise with SD 0.8. The fitted model
includes the shared dyad intercept and the person-specific AR(1) process.

Reference simulations redraw all random effects while keeping predictors and
model parameters fixed. Models are not refitted within the reference simulations.
For correct-model conditions, a separate reference bank uses the true generating
parameters, without PIT centring. These known-parameter checks are run even when
fitting fails. They distinguish the behavior of the reference calculation from
the effects of estimating parameters.

Residual panels include the QQ envelope, histogram envelope, outlier count,
mean PIT distance, and the quartile and distance envelopes against predicted
outcomes. No additional predictor pages are requested. Outcome panels include
the SD after subtracting fixed-effect predictions, largest absolute deviation,
zero count where available, and displayed category frequencies. Counts use
category displays only when there are at most 20 distinct observed values.
Their categories are selected from the observations, so category-specific rates
are conditional on being displayed. The blue-only "Other" category has no
observed comparison and is excluded. Category intervals are separate, so a flag
in any category is not a single 5% check.

Rates use available checks. Failed fits, convergence or Hessian problems, and
failed checks are recorded and excluded without retry. Known-parameter and
fitted-reference rates can therefore have different denominators. Bars show 95%
Wilson intervals for Monte Carlo uncertainty. The paired centring comparison
reports the difference in rates and its paired Monte Carlo standard error.
Counts and uncertainty are reported following the guidance of
[Morris, White, and Crowther (2019)](https://doi.org/10.1002/sim.8086).

The PIT approach builds on
[Dunn and Smyth (1996)](https://doi.org/10.2307/1390802).
The global envelopes use extreme-rank-length ordering, described in
[Myllymäki and Mrkvička (2024)](https://doi.org/10.18637/jss.v111.i03).
Neither reference establishes calibration of the package's fitted-model
references or its median-centring adjustment. That is what this study examines
in the selected settings.

The report reads saved CSV files and does not fit models. See the
[study code](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/distribution-checks/run.R),
[plot code](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/distribution-checks/plot.R),
and [study notes](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/README.md).

```{r settings}
conditions |> filter(n_times == 1, dyad_sd == .6) |>
  distinct(family, scenario_label, role_mean, sd_ratio, nb_size, zero_probability, fit_family) |>
  mutate(nb_size = ifelse(family == "gaussian", NA_real_, nb_size),
    family = unname(distribution_family_labels[family])) |>
  knitr::kable(col.names = c("Generated family", "Setting", "Role shift (±)",
    "SD ratio", "NB size", "Extra-zero probability", "Fitted family"), digits = 2)
```

</details>

## Correct models

Each row below is one numerical panel with all observations pooled. Grey points
use known parameters, blue points use the fitted model with the package defaults,
and orange points omit PIT centring. The dashed line marks 5% as a benchmark,
not a guaranteed fitted-model flag rate. The outcome summaries and original-scale
outlier counts do not depend on PIT centring.

```{r correct-panels, fig.height=8, fig.alt='Individual-panel flag rates under correct Gaussian and negative-binomial models, comparing known parameters, fitted defaults, and uncentred fitted PIT residuals.'}
plot_distribution_panels(summary)
```

<details>
<summary>Any flags and category-frequency flags under correct models</summary>

The table counts any numerical flag within each check, or across both checks.
The latter is a union within each dataset, not a sum of the two rates. Role-based
rates count either role, and their additional comparisons can increase the rate.
Each entry gives the percentage, its 95% Wilson interval, and the denominator.

```{r correct-any}
default_summary |> filter(scenario == "correct", n_times == 1, dyad_sd == .6,
    statistic == "any") |>
  rate_table() |> select(-scenario_label) |>
  knitr::kable(col.names = c("Family", "Dyads", "View", "Check", "Any flag; datasets checked"))
```

```{r correct-categories}
default_summary |> filter(scenario == "correct", n_times == 1, dyad_sd == .6,
    check == "outcome", statistic == "category_any", checked > 0) |>
  rate_table() |> select(-scenario_label, -Check) |>
  knitr::kable(col.names = c("Family", "Dyads", "View", "Any displayed category flagged; datasets checked"))
```

</details>

## Detecting model mismatches

The plots count any numerical flag for each default check. Solid lines count a
flag in either role, while dotted lines pool all observations. These are different
sets of comparisons, so the role-based check may flag more often even under a
correct model.

```{r gaussian-detection, fig.alt='Gaussian mismatch detection by sample size, comparing pooled and role-based residual and outcome checks.'}
plot_distribution_detection(summary, "gaussian")
```

```{r count-detection, fig.alt='Negative-binomial mismatch detection by sample size, comparing pooled and role-based residual and outcome checks.'}
plot_distribution_detection(summary, "nbinom2")
```

## Effect of median centring

This comparison changes only median centring, leaving the fitted model,
simulations, and PIT randomization unchanged. Positive differences mean more
flags after centring. More correct-model flags may indicate less conservative
references or an excessive flag rate; interpret them alongside the absolute
rates above. Fewer flags under a mismatch may indicate a pattern the adjustment
hides. The plot shows 100 dyads; the saved table includes every sample size and
individual panel.

```{r centring, fig.height=7, fig.alt='Paired differences between centred and uncentred residual flag rates at 100 dyads, under correct and misspecified models.'}
plot_distribution_centring(paired)
```

## Additional designs and exclusions

The extra conditions check stronger dyad dependence and one longitudinal model.
They extend the baseline study without covering all dependence structures.

```{r additional-designs}
default_summary |> filter((n_times > 1 | dyad_sd > .6), statistic == "any") |>
  mutate(scenario_label = paste0(scenario_label,
    ifelse(n_times > 1, "; six occasions", "; dyad SD 1.2"))) |>
  rate_table() |>
  knitr::kable(col.names = c("Family", "Setting", "Dyads", "View", "Check", "Any flag; datasets checked"))
```

<details>
<summary>Fit availability and complete saved results</summary>

Warnings are counted whether or not they cause an exclusion. Known-parameter
checks do not require a fitted model. The rate denominators can be smaller than
the usable-fit counts when a particular summary is unavailable.

```{r fit-availability}
fits |> filter(!is.na(reference)) |> group_by(reference) |>
  summarise(across(c(attempted, usable, with_warnings, fit_errors, fit_problems, check_errors), sum),
    .groups = "drop") |>
  knitr::kable(col.names = c("Reference", "Attempted", "Usable",
    "With warnings", "Fit errors", "Fit problems", "Check errors"))
```

```{r fit-exclusions, results='asis'}
excluded <- fits |> filter(usable < attempted)
if (nrow(excluded)) {
  cat("\n\n**Conditions with unavailable checks**\n\n")
  excluded |> mutate(Setting = paste(distribution_family_labels[family], scenario_label,
      paste0(n_times, " occasion(s); dyad SD ", dyad_sd), sep = "; ")) |>
    select(Setting, n_dyads, reference, attempted, usable, fit_errors, fit_problems, check_errors) |>
    knitr::kable(col.names = c("Setting", "Dyads", "Reference", "Attempted", "Usable",
      "Fit errors", "Fit problems", "Check errors")) |> print()
}
```

The saved [summary table](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/report-data/distribution-checks/summary.csv)
contains every panel, category, role, and aggregate rate, with denominators and
Monte Carlo standard errors. The
[fit table](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/report-data/distribution-checks/fits.csv)
records availability for every condition. The
[paired table](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/report-data/distribution-checks/paired-comparison.csv)
records how often only the centred version, only the uncentred version, both, or
neither flagged the same dataset.

</details>
