---
title: "Partner-dependence checks for longitudinal data"
description: "False alarms, detection and fitting behavior for member-mean and same-occasion predictive checks."
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
params:
  output_directory: "../../dev/diagnostic_checks/simulation-studies/report-data/ild-partner-dependence"
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6, dpi = 180, fig.retina = 1, out.width = "100%")
read_results <- function(filename) read.csv(file.path(params$output_directory, filename))
cells <- read_results("cells.csv")
flags <- read_results("flags.csv") |> left_join(cells, by = "cell")
figures <- read_results("figures.csv") |> left_join(cells, by = "cell")
fits <- read_results("fits.csv") |> left_join(cells, by = "cell")
recovery <- read_results("recovery.csv") |> left_join(cells, by = "cell")
effect_sizes <- read_results("effect-sizes.csv") |> left_join(cells, by = "cell")
family_labels <- c(gaussian = "Gaussian", ar = "Gaussian + AR",
  poisson = "Poisson", ordinal = "Ordinal", lag_gaussian = "Gaussian + lags",
  lag_nbinom2 = "NB2 + lags")
model_labels <- c(no_occasion_covariance = "Omit occasion covariance",
  no_stable_covariance = "Omit stable covariance")
primary_flags <- flags |> filter(fits == "usable", rule == "rank")
primary_figures <- figures |> filter(fits == "usable", rule == "rank")
rate_interval <- function(rate, lower, upper) {
  sprintf("%.1f%% [%.1f, %.1f]", 100 * rate, 100 * lower, 100 * upper)
}
study_theme <- theme_bw(base_size = 12) + theme(legend.position = "bottom")
is_pilot <- any(cells$repetitions < 500 | cells$reference_draws < 1000)
```

## What the study checks

[`check_partner_dependence(member = person, time = occasion)`](../reference/check_partner_dependence.html) compares the model's
simulations with two summaries of the observed responses: **member means
(between)** and **deviations from those means on shared occasions (within)**.
This study asks how often those summaries flag under correct covariance
restrictions, and what they detect when a restriction is wrong.

The saved run targets `r unique(cells$repetitions)` datasets per cell and
`r unique(cells$reference_draws)` reference simulations per fitted model, across
`r nrow(cells)` cells. `r if (is_pilot) "**These are pilot results for checking the workflow; they do not establish calibration or detection rates.**" else "The results describe these simulated settings; they do not establish calibration for every supported model."`

A **row flag** means that one observed statistic lies outside the middle 95% of
its simulated values. A **figure flag** means that at least one row flags. These
are descriptive checks. The 5% line is a benchmark for individual rows, not a
promised rate or a figure-level threshold. Summaries freely estimated from the
same data may flag much less often.

The report uses rank limits based on ordered simulated values. These can change
flags compared with the earlier interpolated-quantile rule, shown below as a
sensitivity comparison. Both rules were recomputed from the saved simulations;
no new fits were needed.

<details>
<summary>Design and interpretation</summary>

All settings use female-male dyads and the generator's fixed-effect terms,
including role intercepts and actor/partner predictors. Gaussian settings vary the
number of dyads (50 or 200), occasions (3, 5 or 14), and missingness (0% or 20%).
Missing occasions are removed independently per member. A retained response
still contributes to its member's mean when the partner is missing. Non-lagged
actor/partner predictors are generated independently between partners at both levels.

Gaussian and Poisson models either retain both covariance blocks or set stable
or occasion partner covariance to zero. Ordinal outcomes have four categories,
a probit link and a shared dyad-day intercept instead of an unrestricted
occasion block. Separate settings make each covariance restriction true.
Additional Gaussian models impose equal occasion variances or omit AR(1)
dependence. Lagged-outcome Gaussian and NB2 fits are practical sensitivity
analyses: their predictors depend on earlier outcomes, so they are not clean
calibration controls.

Every reference simulation draws new random effects and responses, while keeping
fitted parameters and predictor values fixed. Models are not refitted. Checks
subtract the same fixed-effect predictions from observed and simulated responses;
each dataset then supplies its own member means. Nonlinear links and short
series can let a misfit at one level affect summaries at the other.

Plots and tables below use fits with optimizer convergence, a positive-definite
Hessian and a finite log-likelihood. Gaussian models use BFGS; other families use
`nlminb`. Failed fits are not retried. The saved tables also report all available
checks, including fits that fail those criteria, and preserve failures separately.

Error bars and brackets are 95% Wilson intervals for flag rates. Monte Carlo SEs
describe uncertainty from the number of generated datasets. Neither measures
uncertainty in an individual model's fitted parameters.

</details>

## False alarms under a true restriction

Here the fitted covariance restriction is true. Each panel shows the partner
correlation row at the corresponding level: zero stable covariance at the
between level, or zero occasion covariance at the within level. Other summaries
and full-model results remain in the saved tables. These true-restriction settings
use complete data and 5 or 14 occasions.

```{r true-restrictions, fig.alt="Partner-correlation flag rates under true zero-covariance restrictions, by family, level, number of occasions and dyad count."}
null_rows <- primary_flags |>
  filter(statistic == "partner_cor",
    (truth_case == "zero_stable" & model == "no_stable_covariance" & level == "between") |
    (truth_case == "zero_occasion" & model == "no_occasion_covariance" & level == "within")) |>
  mutate(family = factor(part, names(family_labels), family_labels), dyads = factor(n_dyads))
ggplot(null_rows, aes(n_occasions, rate, colour = dyads, group = dyads)) +
  geom_hline(yintercept = 0.05, linetype = 2, colour = "grey55") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.5,
    position = position_dodge(width = 0.8)) +
  geom_point(position = position_dodge(width = 0.8), size = 2.5) +
  facet_grid(family ~ level) +
  scale_x_continuous(breaks = c(5, 14)) +
  scale_y_continuous(labels = scales::label_percent(), limits = c(0, NA)) +
  scale_colour_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Occasions", y = "Datasets with a row flag", colour = "Dyads") + study_theme
```

## Detection at both levels

These panels use 200 dyads, 14 complete occasions and nonzero partner covariance
at both levels. They show how omitting either covariance affects the partner
correlation row at **both** levels. A flag does not by itself locate the source
of the mismatch.

```{r omitted-covariance, fig.alt="Partner-correlation flag rates at both levels when stable or occasion covariance is omitted, with 200 dyads and 14 complete occasions."}
detection_rows <- primary_flags |>
  filter(part %in% c("gaussian", "poisson", "ordinal"), truth_case == "correlated",
    n_dyads == 200, n_occasions == 14, missing_share == 0,
    model %in% names(model_labels), statistic == "partner_cor") |>
  mutate(family = factor(part, names(family_labels), family_labels),
    restriction = factor(model, names(model_labels), model_labels))
ggplot(detection_rows, aes(rate, restriction)) +
  geom_segment(aes(x = lower, xend = upper, yend = restriction), colour = "#0072B2") +
  geom_point(colour = "#0072B2", size = 2.5) +
  facet_grid(family ~ level) +
  scale_x_continuous(labels = scales::label_percent(), limits = c(0, 1)) +
  labs(x = "Datasets with a row flag", y = NULL) + study_theme
```

Response-scale correlations below describe the generated data after subtracting
their true fixed-effect means. They help distinguish response-scale dependence
from latent covariance parameters. These estimates use large complete samples;
they are not fitted-model results.

```{r response-effect-sizes}
effect_sizes |>
  filter(part %in% c("gaussian", "poisson", "ordinal"), truth_case == "correlated",
    n_dyads == 200, n_occasions == 14, missing_share == 0) |>
  transmute(Family = family_labels[part], Between = between_partner_cor,
    Within = within_partner_cor) |>
  knitr::kable(digits = 3)
```

```{r main-findings, results='asis'}
targeted_detection <- detection_rows |>
  filter((model == "no_stable_covariance" & level == "between") |
    (model == "no_occasion_covariance" & level == "within"))
if (!is_pilot) {
  null_figures <- primary_figures |>
    semi_join(null_rows, by = c("cell", "model", "response")) |> filter(figure == "either")
  full_figures <- primary_figures |> filter(model == "correct", figure == "either")
  gaussian_full <- full_figures |> filter(part == "gaussian")
  cat(sprintf("Under true zero-covariance restrictions, targeted correlation rows flagged **%.1f–%.1f%%** of datasets; a flag anywhere in either figure occurred in **%.1f–%.1f%%**. For full models, either-figure rates were **%.1f–%.1f%%** across families and **%.1f–%.1f%%** for Gaussian models.\n\n",
    100 * min(null_rows$rate), 100 * max(null_rows$rate),
    100 * min(null_figures$rate), 100 * max(null_figures$rate),
    100 * min(full_figures$rate), 100 * max(full_figures$rate),
    100 * min(gaussian_full$rate), 100 * max(gaussian_full$rate)))
  covariance_rows <- primary_flags |>
    filter(part %in% c("gaussian", "poisson", "ordinal"), truth_case == "correlated",
      model %in% names(model_labels), statistic == "partner_cor") |>
    mutate(targeted = (model == "no_stable_covariance" & level == "between") |
      (model == "no_occasion_covariance" & level == "within"))
  small_stable <- covariance_rows |> filter(targeted, n_dyads == 50, level == "between")
  small_occasion <- covariance_rows |> filter(targeted, n_dyads == 50, level == "within")
  short_series <- small_occasion |> filter(part == "gaussian", n_occasions == 3) |>
    arrange(missing_share)
  cat(sprintf("With 200 dyads and 14 complete occasions, targeted detection was **%.1f–%.1f%%**. At 50 dyads it was **%.1f–%.1f%%** for omitted stable covariance and **%.1f–%.1f%%** for omitted occasion covariance. In Gaussian data with 50 dyads and 3 occasions, occasion-covariance detection fell from **%.1f%%** with complete data to **%.1f%%** with 20%% missing member occasions.\n\n",
    100 * min(targeted_detection$rate), 100 * max(targeted_detection$rate),
    100 * min(small_stable$rate), 100 * max(small_stable$rate),
    100 * min(small_occasion$rate), 100 * max(small_occasion$rate),
    100 * short_series$rate[1], 100 * short_series$rate[2]))
  sd_rows <- primary_flags |> filter(part == "gaussian", model == "equal_occasion_variances",
    level == "within", statistic == "average_difference_cor")
  cat(sprintf("The partner-correlation row at the other level flagged at most **%.1f%%** across these covariance-omission settings. That result concerns this row, not every statistic or possible mismatch. In the main Gaussian study, incorrectly pooling occasion SDs with a true ratio of 1.5 was detected in **%.1f–%.1f%%** of datasets.\n\n",
    100 * max(covariance_rows$rate[!covariance_rows$targeted]),
    100 * min(sd_rows$rate), 100 * max(sd_rows$rate)))
}
```

## Serial dependence and lagged outcomes

The AR comparison holds total occasion variances equal. It checks the
within-level correlation between the partner average and half-difference when the fitted model imposes
that true equality but omits serial dependence. The φ = 0 setting is the control;
φ = 0.7 adds serial dependence to the generated data.

```{r ar-restriction}
ar_rows <- primary_flags |>
  filter(part == "ar", model == "no_ar_equal_occasion_variances", level == "within",
    statistic == "average_difference_cor") |>
  arrange(n_dyads, ar_phi)
ar_rows |>
  transmute(Dyads = n_dyads, `AR coefficient` = ar_phi, Checked = checked,
    `Flag rate [95% interval]` = rate_interval(rate, lower, upper),
    `Monte Carlo SE (%)` = round(100 * mcse, 2)) |>
  knitr::kable()
```

```{r serial-dependence-finding, results='asis'}
ar_unrestricted <- primary_figures |>
  filter(part == "ar", model == "no_ar", ar_phi == 0.7, figure == "either")
if (!is_pilot) {
  ar_changes <- ar_rows |> select(n_dyads, ar_phi, rate, mcse) |>
    tidyr::pivot_wider(names_from = ar_phi, values_from = c(rate, mcse)) |>
    mutate(change = .data[["rate_0.7"]] - rate_0,
      change_mcse = sqrt(.data[["mcse_0.7"]]^2 + mcse_0^2))
  cat(sprintf("From φ = 0 to φ = 0.7, row-flag rates rose by **%.1f and %.1f percentage points** at 50 and 200 dyads (Monte Carlo SEs **%.1f and %.1f**, independent cells): small and imprecise increases. The variance equality is true; the mismatch is the omitted serial dependence.\n\n",
    100 * ar_changes$change[1], 100 * ar_changes$change[2],
    100 * ar_changes$change_mcse[1], 100 * ar_changes$change_mcse[2]))
  if (all(ar_unrestricted$rate == 0))
    cat("The unrestricted covariance model without AR produced **no figure flags** at either sample size despite φ = 0.7. These summaries did not reveal the missing serial dependence. Autocorrelation needs a separate check.\n\n")
  if (file.exists(file.path(params$output_directory, "serial-recovery.csv"))) {
    serial_recovery <- read_results("serial-recovery.csv") |> left_join(cells, by = "cell") |>
      arrange(n_dyads)
    cat(sprintf("Despite those quiet checks, estimated stable correlation averaged **%.3f and %.3f** at 50 and 200 dyads, versus a true **%.2f** (Monte Carlo SEs **%.3f and %.3f**).\n\n",
      serial_recovery$mean[1], serial_recovery$mean[2], serial_recovery$true_value[1],
      serial_recovery$mcse[1], serial_recovery$mcse[2]))
  }
}
```

These checks can be misleading when the model includes lagged outcomes as predictors.
For these models, simulations keep the observed lag predictors fixed.
The following table compares raw and model-centred checks using the same
simulations. It reports a flag **anywhere in either figure**. High rates can
reflect this simulation limitation as well as fitting behavior; they are not
false-alarm estimates for a clean null model.

```{r lag-sensitivity}
primary_figures |>
  filter(startsWith(part, "lag_"), figure == "either") |>
  arrange(part, n_dyads, response) |>
  transmute(Family = family_labels[part], Dyads = n_dyads, Response = response,
    Checked = checked, `Any flag [95% interval]` = rate_interval(rate, lower, upper),
    `Monte Carlo SE (%)` = round(100 * mcse, 2)) |>
  knitr::kable()
```

Agreement in the Gaussian lagged-outcome case does not establish that keeping
outcome-derived predictors fixed gives a valid simulation reference.

## Fitting and sensitivity checks

The denominator of each rate is the number of datasets for which that check was
available in the stated fit subset. The table counts models across all cells in
each part; a single generated dataset may supply several fitted models.

```{r fit-availability}
fits |>
  summarise(Attempted = sum(attempted), Usable = sum(usable),
    `Optimizer failures` = sum(not_converged), `Hessian problems` = sum(not_positive_hessian),
    `Fit errors` = sum(fit_errors), `Check errors` = sum(check_errors), .by = part) |>
  mutate(Family = family_labels[part], .before = 1) |>
  select(-part) |>
  knitr::kable()
```

```{r rule-and-fit-sensitivity, results='asis'}
rule_comparison <- flags |>
  filter(fits == "usable") |>
  select(cell, model, response, level, statistic, rule, rate) |>
  tidyr::pivot_wider(names_from = rule, values_from = rate)
fit_comparison <- flags |>
  filter(rule == "rank") |>
  select(cell, model, response, level, statistic, fits, rate) |>
  tidyr::pivot_wider(names_from = fits, values_from = rate)
cat(sprintf("The largest per-row difference between the primary rank rule and the earlier quantile rule was **%.1f percentage points**. The largest difference between all available checks and usable fits was **%.1f percentage points**, among rows available in both subsets.\n\n",
  100 * max(abs(rule_comparison$quantile - rule_comparison$rank), na.rm = TRUE),
  100 * max(abs(fit_comparison$all - fit_comparison$usable), na.rm = TRUE)))
```

Parameter recovery is kept separate from check performance. The saved recovery
table contains estimates only for correctly specified models, excluding the
lagged-outcome sensitivity fits. This compact view gives the largest absolute
bias across cells for fixed predictor effects and latent partner correlations.

```{r parameter-recovery}
parameter_labels <- c(x_actor = "Actor effect", x_partner = "Partner effect",
  block1_cor = "Stable correlation", block2_cor = "Occasion correlation")
recovery |>
  filter(parameter %in% names(parameter_labels)) |>
  summarise(`Largest absolute bias` = max(abs(bias)),
    `Fewest usable estimates` = min(n), .by = c(part, parameter)) |>
  transmute(Family = family_labels[part], Parameter = parameter_labels[parameter],
    `Largest absolute bias`, `Fewest usable estimates`) |>
  knitr::kable(digits = 3)
```

The [full tables](https://github.com/Pascal-Kueng/dyadMLM/tree/main/dev/diagnostic_checks/simulation-studies/report-data/ild-partner-dependence)
retain all statistics, missingness settings, sample sizes,
figure-level rates and Monte Carlo SEs. See the
[study code](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/ild-partner-dependence/run.R),
[design](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/ild-partner-dependence-plan.md#7-validation-after-implementation-once)
and [run instructions](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/README.md#longitudinal-partner-dependence-checks).
