---
title: "Parameter recovery for mixed-composition APIMs"
description: "Fixed effects and member-level covariance recovery in correctly specified Gaussian models."
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
params:
  output_directory: "../../dev/diagnostic_checks/simulation-studies/report-data/mixed-apim-recovery"
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 10, dpi = 180, fig.retina = 1, out.width = "100%")
read_results <- function(name) read.csv(file.path(params$output_directory, paste0(name, ".csv")))
conditions <- read_results("conditions")
design_labels <- c(cross_intercept = "Cross-sectional: intercepts",
  cross_apim = "Cross-sectional: actor/partner effects",
  ild_intercept = "Longitudinal: random intercepts", ild_slopes = "Longitudinal: random slopes")
conditions <- conditions |>
  arrange(factor(design, levels = names(design_labels)),
    factor(pooling, levels = c("separate", "pooled")), n_dyads) |>
  mutate(Model = paste0(design_labels[design], ifelse(pooling == "pooled", " (pooled)", "")),
    Model = factor(Model, unique(Model)), Dyads = factor(n_dyads))
fits <- read_results("fits") |> left_join(conditions, by = "condition")
recovery <- read_results("recovery") |> left_join(conditions, by = "condition")
fixed <- filter(recovery, kind == "fixed")
random <- filter(recovery, kind != "fixed")
is_pilot <- any(conditions$repetitions != 500L) || nrow(fits) != nrow(conditions) ||
  any(fits$attempted != fits$repetitions)
study_theme <- theme_bw(base_size = 11) + theme(legend.position = "bottom",
  panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"))
sample_colours <- c("120" = "#0072B2", "360" = "#D55E00")
```

## What the study checks

Do the mixed-composition Gaussian models recover their true **fixed effects and
random covariance parameters**? Twelve conditions use 120 or 360 dyads, with equal
numbers of female–female, female–male and male–male dyads before pooling. Each
condition targets `r unique(conditions$repetitions)` independently generated
datasets.

The cross-sectional conditions fit intercept-only or actor–partner models.
Longitudinal conditions have 14 complete occasions and stable random intercepts,
with or without within-person actor and partner random slopes. Both retain
same-occasion partner covariance. Longitudinal conditions either separate all
compositions or correctly pool the two same-sex compositions. Every fitted model
matches its generating structure; there is no missingness or additional serial
dependence.
Pooled and separate conditions use different generating parameters, so their
comparison is descriptive.

The generating outcomes use manually computed observed predictor means and
deviations, then the models use package-prepared columns. This checks the
specification and coding, not latent centring. Exchangeable shared/difference
blocks are transformed back to member-level variances, covariances and
correlations before comparison with truth.

`r if (is_pilot) "**These are pilot or incomplete results. They check the workflow and do not establish parameter recovery.**" else "The completed run contains 500 datasets per condition (6,000 joint-model fitting procedures, plus fits used for starting values)."`

## Fitting and usable estimates

All fits use maximum likelihood and BFGS (`maxit = 1000`, `reltol = 1e-10`),
without retries. Cross-sectional and random-intercept models use profiling.
Random-slope models first fit the compositions sequentially with profiling and
unchanged prepared predictors. Their fitted coefficients and covariance parameters
provide named starting values for the literal joint model, fitted without
profiling. Starting values never use the generating parameters.

Recovery requires the final joint fit to converge, have a positive-definite
Hessian and return finite log-likelihood, estimates and fixed-effect SEs.
**Start issues** counts datasets with an initializer convergence or Hessian
problem; finite starts, including boundary estimates, are still used. A thrown
initialization error records a failed procedure. **Boundary** counts final fits
with a near-singular covariance block and alone does not exclude them. Warnings
include all stages. Counts can overlap.

```{r fit-table}
fits |>
  arrange(Model, Dyads) |>
  transmute(Model, Dyads, `Usable / attempted` = paste0(usable, " / ", attempted),
    `Hessian failures` = not_positive_hessian, Boundary = boundary,
    `Start issues` = initialization_failures, Errors = fit_errors) |>
  knitr::kable()
```

```{r availability-findings, results='asis'}
if (!is_pilot) {
  simple <- filter(fits, design != "ild_slopes")
  cat(sprintf("The cross-sectional and random-intercept models returned **%s/%s** usable fits.\n\n",
    sum(simple$usable), sum(simple$attempted)))
  for (size in c(120L, 360L)) {
    rates <- filter(fits, design == "ild_slopes", n_dyads == size)$usable_rate
    cat(sprintf("At **%s dyads**, only **%.1f–%.1f%%** of random-slope fits were usable.\n\n",
      size, 100 * min(rates), 100 * max(rates)))
  }
}
```

## Fixed effects

Each point below represents **one coefficient in one condition**; coefficients
with the same effect type remain separate. Bias is the mean estimate minus its
true value. Coverage is the share of available 95% Wald intervals containing the
true value. Bars show Monte Carlo uncertainty: bias ± 1.96 Monte Carlo SEs and
Wilson intervals for coverage. At 500 usable estimates, coverage near 95% has a
Monte Carlo SE of about one percentage point.

```{r fixed-recovery, fig.height=12, fig.alt="Bias and 95 percent interval coverage for every fixed coefficient, separated by effect type, model and number of dyads."}
effect_labels <- c(intercept = "Intercept", time = "Time", actor = "Actor",
  partner = "Partner", actor_between = "Actor: between", partner_between = "Partner: between",
  actor_within = "Actor: within", partner_within = "Partner: within")
fixed_plot <- fixed |>
  mutate(Effect = factor(sub(".*:", "", parameter),
    levels = rev(names(effect_labels)), labels = rev(effect_labels))) |>
  tidyr::uncount(2L, .id = "metric") |>
  mutate(Metric = factor(metric, labels = c("Bias", "95% interval coverage")),
    Value = ifelse(metric == 1L, bias, coverage),
    Lower = ifelse(metric == 1L, bias - 1.96 * bias_mcse, coverage_lower),
    Upper = ifelse(metric == 1L, bias + 1.96 * bias_mcse, coverage_upper))
benchmarks <- tibble::tibble(Metric = factor(c("Bias", "95% interval coverage"),
  levels = levels(fixed_plot$Metric)), value = c(0, .95))
dodge <- position_dodge(width = .65, orientation = "y")
ggplot(fixed_plot, aes(Value, Effect, colour = Dyads, group = interaction(parameter, Dyads))) +
  geom_vline(data = benchmarks, aes(xintercept = value), colour = "grey55", linetype = 2) +
  geom_errorbar(aes(xmin = Lower, xmax = Upper), orientation = "y",
    position = dodge, width = .3, linewidth = .3) +
  geom_point(position = dodge, size = 1.4) +
  facet_wrap(vars(Model, Metric), ncol = 2, scales = "free") +
  scale_colour_manual(values = sample_colours) +
  labs(x = "Bias / coverage proportion", y = NULL) + study_theme
```

```{r fixed-findings, results='asis'}
if (nrow(fixed)) {
  findings <- fixed |> summarise(min_bias = min(bias), max_bias = max(bias),
    min_coverage = min(coverage, na.rm = TRUE), max_coverage = max(coverage, na.rm = TRUE),
    .by = n_dyads) |> arrange(n_dyads)
  for (i in seq_len(nrow(findings))) {
    row <- findings[i, ]
    cat(sprintf("At **%s dyads**, fixed-effect biases ranged from **%.3f to %.3f** and coverage from **%.1f%% to %.1f%%** across coefficients and models.\n\n",
      row$n_dyads, row$min_bias, row$max_bias, 100 * row$min_coverage, 100 * row$max_coverage))
  }
}
```

```{asis, echo=!is_pilot}
Point estimates were generally close to truth. Some 95% Wald intervals
undercovered, including in simpler models; small mean bias alone does not
establish reliable interval estimates.
```

## Random covariance parameters

Each point compares one parameter's mean estimate with its true value. Points on
the diagonal show agreement in the mean; their spread across datasets is reported
in the CSV below. Variance, covariance and correlation panels use their own
scales. Signed covariances remain signed, including values close to zero. These
plots do not assess interval coverage for random parameters.

```{r covariance-recovery, fig.height=7, fig.alt="Mean estimated member-level variances, covariances and correlations plotted against their true values, separated by covariance level."}
random_plot <- random |>
  mutate(Component = factor(component, c("residual", "stable", "occasion"),
    c("Cross-sectional residual", "Stable dyad effects", "Same-occasion residual")),
    Kind = factor(kind, c("variance", "covariance", "correlation"),
      c("Variance", "Covariance", "Correlation")))
ggplot(random_plot, aes(true_value, mean, colour = Dyads)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey55", linetype = 2) +
  geom_point(size = 1.8, alpha = .55) +
  facet_wrap(vars(Component, Kind), ncol = 3, scales = "free") +
  scale_colour_manual(values = sample_colours) +
  labs(x = "True value", y = "Mean estimate") + study_theme
```

```{asis, echo=!is_pilot}
Simpler models recovered covariance parameters more closely. Their variance
estimates tended to be low, with smaller bias at 360 dyads. Random-slope models
had imprecise stable correlations and sometimes overestimated slope variances
among their usable fits. For example, the female partner-slope variance in
female–male dyads (true value 0.040) averaged 0.062–0.072 at 120 dyads.
```

The table shows the range of correlation RMSEs within each model and sample
size. RMSE measures error in individual datasets; small average bias can coexist
with imprecise estimates. Variance and covariance RMSEs remain in the full results.

```{r correlation-precision}
random |>
  filter(kind == "correlation") |>
  summarise(RMSE = sprintf("%.3f–%.3f", min(rmse), max(rmse)),
    .by = c(Model, Dyads)) |>
  tidyr::pivot_wider(names_from = Dyads, values_from = RMSE,
    names_glue = "{Dyads} dyads", names_sort = TRUE) |>
  arrange(Model) |>
  knitr::kable()
```

## Full results and scope

```{r csv-link, results='asis'}
cat(sprintf('<a download="mixed-apim-recovery.csv" href="%s">Full results for every parameter (CSV)</a>',
  base64enc::dataURI(file = file.path(params$output_directory, "recovery.csv"), mime = "text/csv")))
cat(sprintf(' · <a download="mixed-apim-conditions.csv" href="%s">Condition definitions (CSV)</a>',
  base64enc::dataURI(file = file.path(params$output_directory, "conditions.csv"), mime = "text/csv")))
cat(sprintf(' · <a download="mixed-apim-fits.csv" href="%s">Fit diagnostics (CSV)</a>',
  base64enc::dataURI(file = file.path(params$output_directory, "fits.csv"), mime = "text/csv")))
```

The table includes the true value, mean estimate, bias, Monte Carlo SE, empirical
SD, RMSE, mean absolute error and actual usable denominator. Fixed effects also
include mean model SEs, their ratio to empirical SD, and interval coverage.

Interpret recovery alongside fit availability: usable-fit summaries can look
better when difficult fits are excluded. The results concern these correctly
specified Gaussian designs and sample sizes. They do not establish reliable
estimation for arbitrary compositions, more complex temporal structures or other
fitting engines. Dyad-plus-singleton models are outside this study's scope.
