---
title: "Partner-dependence checks across outcome families"
description: "Detection and false alarms across sample sizes and response families."
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
params:
  results_directory: "../../dev/diagnostic_checks/simulation-studies/report-data/family-comparison"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 13, fig.height = 6, dpi = 180, fig.retina = 1, out.width = "100%")
source("../../dev/diagnostic_checks/simulation-studies/family-comparison/plot.R")
results_directory <- params$results_directory
settings <- read.csv(file.path(results_directory, "settings.csv")) |>
  dplyr::mutate(dplyr::across(c(mean_settings, dispersion_settings, zero_settings),
    ~ gsub("; ", ", ", .x, fixed = TRUE)))
figure_descriptions <- as.vector(rbind(
  paste0(settings$label, ": detection of omitted partner correlation by number of dyads."),
  paste0(settings$label, ": false alarms with no residual partner correlation by number of dyads.")
))
summary <- read.csv(file.path(results_directory, "summary.csv"))
calibration <- read.csv(file.path(results_directory, "calibration.csv"))
fit_exclusions <- read.csv(file.path(results_directory, "fit-exclusions.csv"))
```

## What these plots compare

The models include actor and partner effects but leave out residual partner
dependence. We compare checks of raw outcomes with checks that first subtract
fixed-effect predictions. Both use the **same datasets, fitted models, and reference simulations**.

- **Coloured lines:** how often the check detects omitted positive correlation.
  A mismatch is detected when the observed correlation exceeds the upper limit of
  the model's middle 95% simulated range. Flags below the lower limit are listed
  separately in each family's details.
- **Black lines:** false alarms when residual correlation is zero. Values below or
  above the middle 95% simulated range count as false alarms. The grey dashed line
  marks 5% as a benchmark, not a guaranteed rate.

The results apply to the settings studied and can differ with other model
settings. Rates include only usable fits. Many fits were excluded for skew-normal
models and small Student t samples, mainly because of numerical problems during
fitting.

Checks that subtract model predictions generally detected more omitted dependence
than raw checks. Their false-alarm rates ranged from 4.7% to 5.9% across settings
when pooled over sample sizes. Small residual correlations (0.10) were difficult
to detect even with several hundred dyads. At 1,000 dyads, detection ranged from
27% to 91% across the studied settings.

Ordinal checks use category scores 1, 2, and so on. Zero-inflated and hurdle checks
use the combined outcome, including zeros.

<details>
<summary>Study details</summary>

Predictors are standard normal with a partner correlation of 0.3. Actor and partner
effects are 0.5 and 0.3 on the model's link scale. We transform correlated normal
draws to each outcome distribution using a Gaussian copula. We adjust their
dependence to give residual correlations of 0, 0.10, 0.30, and 0.50 after subtracting
the true means given the predictors.

Reference simulations keep fitted parameters and predictors fixed, without
refitting the models. More simulations give more precise estimates of the 2.5th
and 97.5th percentile limits. More study datasets give more precise estimates of
the plotted rates. The bars and bands are 95% Wilson intervals for this Monte Carlo
uncertainty.

We exclude failed fits and checks without retry, as well as ordinal datasets
missing a category. Correlations must be defined to count as a valid check.
Undefined simulated correlations are omitted from reference ranges. Each family's
details give exclusion counts and reasons.

For 1,450 resumed Tweedie datasets, an isolated glmmTMB build used TMB's equivalent
single-Gamma sampler to avoid extremely slow simulation when fitted power was close
to 2. This preserved the fitted models and response distributions but changed the
random draws.
The [study notes](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/README.md#tweedie-simulation-repair) give the version, patch, and checks.

The report uses saved results. See the
[study code](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/family-comparison/run.R), [plot code](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/family-comparison/plot.R), and
[study notes](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/README.md).

</details>

```{r family-results, results='asis', fig.alt=figure_descriptions}
for (family_index in seq_len(nrow(settings))) {
  family_settings <- settings[family_index, ]
  family_summary <- summary[summary$family == family_settings$family, ]
  family_calibration <- calibration[calibration$family == family_settings$family, ]
  cat("\n\n## ", family_settings$label, "\n\n", sep = "")
  figure <- plot_family_comparison(family_summary, family_settings$label,
    family_settings$datasets_per_condition, family_settings$reference_draws)
  print(figure)
  print(plot_family_false_alarms(family_summary, family_settings$label,
    family_settings$datasets_per_condition, family_settings$reference_draws))

  cat("<details>\n<summary>Model settings and checks</summary>\n\n")
  cat("**Fitted model:** `", family_settings$formula, "`.\n\n", sep = "")
  cat("**Mean:** ", family_settings$mean_settings, "\n\n", sep = "")
  cat("**Dispersion:** ", family_settings$dispersion_settings, "\n\n", sep = "")
  if (!is.na(family_settings$zero_settings) && nzchar(family_settings$zero_settings)) {
    cat("**Zero component:** ", family_settings$zero_settings, "\n\n", sep = "")
  }

  cat("**Population correlations**\n\n")
  cat(paste(unique(gsub("; ", " and ", family_calibration$method, fixed = TRUE)),
    collapse = ". "), ".\n\n", sep = "")
  print(knitr::kable(family_calibration[c("target_correlation", "measured_residual_correlation", "verification_mcse")],
    col.names = c("Target residual correlation", "Verified correlation", "Monte Carlo SE"),
    digits = 3, row.names = FALSE))

  cat("\n\n**Opposite-direction flags with positive residual correlation**\n\n")
  opposite_direction <- family_summary |>
    dplyr::filter(target_correlation > 0) |>
    dplyr::group_by(response) |>
    dplyr::summarise(checked = sum(checked),
      opposite_direction = sum(opposite_direction), .groups = "drop")
  print(knitr::kable(opposite_direction,
    col.names = c("Method", "Datasets checked", "Opposite-direction flags")))

  # Fit counts repeat for the two methods. Count them only once.
  fit_counts <- family_summary[family_summary$response == "model-centred", ]
  cat("\n\n", sum(fit_counts$successful), " of ", sum(fit_counts$attempted),
      " datasets produced reference simulations.\n\n", sep = "")
  excluded_fits <- fit_exclusions |>
    dplyr::filter(family == family_settings$family) |>
    dplyr::select(status, Datasets)
  if (nrow(excluded_fits)) {
    excluded_fits$status <- c(fit_error = "Model fitting failed",
      fit_problem = "Convergence or Hessian problem", check_error = "Check failed",
      study_exclusion = "Ordinal category absent")[excluded_fits$status]
    print(knitr::kable(excluded_fits, col.names = c("Exclusion reason", "Datasets")))
  }
  reduced_counts <- family_summary |>
    dplyr::filter(checked < attempted) |>
    dplyr::group_by(n_dyads, target_correlation, attempted, checked) |>
    dplyr::summarise(methods = paste(response, collapse = ", "), .groups = "drop") |>
    dplyr::select(n_dyads, target_correlation, methods, attempted, checked)
  if (nrow(reduced_counts)) {
    cat("\n\n**Conditions with fewer usable checks**\n\n")
    print(knitr::kable(reduced_counts,
      col.names = c("Dyads", "Residual correlation", "Methods", "Generated", "Checked")))
  }
  cat("\n\nCounts for every plotted point are saved in the [summary table](https://github.com/Pascal-Kueng/dyadMLM/blob/main/dev/diagnostic_checks/simulation-studies/report-data/family-comparison/summary.csv).\n\n</details>\n\n")
}
```
