---
title: "APIM exploration: lavaan and glmmTMB"
author: "Pascal Küng"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    math_method: mathml
    df_print: kable
    css: explore-sem-mlm.css
---

This notebook shows the full workflow for the same cross-sectional Gaussian
APIM in lavaan and glmmTMB. Both give the same point estimates of the
covariance decomposition, and their bootstrap confidence intervals agree
within resampling noise. Each dyad has one parent and one teen, with complete data.

```{r table-style, include=FALSE}
# Each table includes its CSS when chunks are run separately.
table_css <- paste(readLines("explore-sem-mlm.css"), collapse = "\n")

print_table <- function(data) {
  htmltools::browsable(htmltools::tagList(
    htmltools::tags$style(htmltools::HTML(table_css)),
    htmltools::HTML(knitr::kable(
      data, format = "html", table.attr = 'class="apim-table"'
    ))
  ))
}
```

## Load and rename the data

```{r load-flashe}
library(lavaan)
flashe_data <- read.csv(
  file.path(
    "../references/explaining-interdependence-apim",
    "jp-materials-2026-09-13",
    "J-P&Niall-FLASHE-example",
    "flashe-lavaan.csv"
  ),
  header = TRUE
)

# Produce intake and efficacy refer to fruit and vegetable consumption.
# Keep the supplied scores centered at 7.6.
flashe_data <- dplyr::rename(
  flashe_data,
  produce_intake_parent = pftveg,
  produce_intake_teen = tftveg,
  efficacy_parent = peffv,
  efficacy_teen = teffv,
  efficacy_centered_parent = cpeffv,
  efficacy_centered_teen = cteffv,
  is_female_parent = pfem,
  is_female_teen = tfem
)
```

## 1. Lavaan: J-P and Niall's example

### Fit the model

```{r jp-niall-lavaan}
apim_model <- '
  # Intake intercepts: both centered efficacy predictors are zero.
  produce_intake_teen ~ intercept_teen * 1
  produce_intake_parent ~ intercept_parent * 1
  
  # Actor effects: own centered efficacy predicts own F&V intake.
  produce_intake_teen ~ actor_teen * efficacy_centered_teen
  produce_intake_parent ~ actor_parent * efficacy_centered_parent

  # Partner effects: parent efficacy -> teen intake, and vice versa.
  produce_intake_teen ~ partner_parent_to_teen * efficacy_centered_parent
  produce_intake_parent ~ partner_teen_to_parent * efficacy_centered_teen

  # Means of the centered efficacy predictors.
  efficacy_centered_teen ~ mean_efficacy_teen * 1
  efficacy_centered_parent ~ mean_efficacy_parent * 1

  # Predictor variances.
  efficacy_centered_teen ~~ var_efficacy_teen * efficacy_centered_teen
  efficacy_centered_parent ~~ var_efficacy_parent * efficacy_centered_parent

  # Residual variances of intake.
  produce_intake_teen ~~ residual_var_teen * produce_intake_teen
  produce_intake_parent ~~ residual_var_parent * produce_intake_parent

  # Covariance between parent and teen efficacy.
  efficacy_centered_parent ~~ cov_efficacy * efficacy_centered_teen

  # Residual covariance between parent and teen intake.
  produce_intake_parent ~~ residual_cov * produce_intake_teen

  # Path-traced contributions to the outcome covariance.
  actor_driven := actor_teen * cov_efficacy * actor_parent
  partner_driven := partner_parent_to_teen * cov_efficacy * partner_teen_to_parent
  teen_driven := actor_teen * var_efficacy_teen * partner_teen_to_parent
  parent_driven := actor_parent * var_efficacy_parent * partner_parent_to_teen
  residual := residual_cov
  explained := actor_driven + partner_driven + teen_driven + parent_driven
  total_covariance := explained + residual

  # Percentages of the total outcome covariance.
  percent_actor_driven := 100 * (actor_driven / total_covariance)
  percent_partner_driven := 100 * (partner_driven / total_covariance)
  percent_teen_driven := 100 * (teen_driven / total_covariance)
  percent_parent_driven := 100 * (parent_driven / total_covariance)
  percent_residual := 100 * (residual / total_covariance)
  percent_explained := 100 * (explained / total_covariance)
'

# Predictor means, variances and covariance are free parameters above.
# Bootstrap CIs: refit to 1999 samples of the dyads.
apim_fit <- sem(
  apim_model, fixed.x = FALSE, data = flashe_data, missing = "fiml",
  se = "bootstrap", bootstrap = 1999, iseed = 2026,
  parallel = "multicore", ncpus = 4
)

# Examine the model.
# summary(apim_fit, fit.measures = TRUE)
```

### Extract the model parameters from lavaan

First the usual estimates:

```{r extract-lavaan}
# Extract once, using our model labels and including 95% CIs.
lavaan_estimates <- broom::tidy(apim_fit, conf.int = TRUE) |>
  dplyr::select(
    parameter = label,
    op,
    estimate,
    ci_low = conf.low,
    ci_high = conf.high
  )

# Model parameters: exclude the quantities defined with :=.
lavaan_params <- lavaan_estimates |>
  dplyr::filter(op != ":=") |>
  dplyr::select(-op)

print_table(lavaan_params)
```

### Extract the covariance decomposition

Now the covariance decomposition estimates and CIs:

```{r lavaan-covariance-decomposition}
# Decomposition: keep only the quantities defined with :=.
lavaan_decomposition <- lavaan_estimates |>
  dplyr::filter(op == ":=") |>
  dplyr::select(-op)

# Display the five contributions and the explained subtotal, with CI and percent.
format_decomposition <- function(decomposition) {
  # Match each percentage row to its contribution by removing "percent_".
  percentages <- decomposition |>
    dplyr::filter(startsWith(parameter, "percent_")) |>
    dplyr::mutate(parameter = sub("^percent_", "", parameter)) |>
    dplyr::select(parameter, percent = estimate)

  # Format only the displayed CI; estimates keep their full precision.
  decomposition |>
    dplyr::filter(!startsWith(parameter, "percent_"),
                  parameter != "total_covariance") |>
    dplyr::mutate(
      `95% CI` = sprintf("[%.3f, %.3f]", ci_low, ci_high)
    ) |>
    dplyr::select(parameter, estimate, `95% CI`) |>
    dplyr::left_join(percentages, by = "parameter")
}

lavaan_decomposition_table <- format_decomposition(lavaan_decomposition)
print_table(lavaan_decomposition_table)
```


## 2. glmmTMB: reproduce the point estimates

`dyadMLM` prepares the actor and partner predictors.
We retain the supplied centering and use maximum likelihood to mirror lavaan.

### Reshape and prepare the data

Reshape to one row per person (long-format):

```{r reshape-long}
# Each original row is one dyad. Long format has one row per person.
flashe_long <- flashe_data |>
  dplyr::mutate(dyad_id = dplyr::row_number()) |>
  tidyr::pivot_longer(
    cols = -dyad_id,
    names_to = c(".value", "role"),
    names_pattern = "(.+)_(parent|teen)$"
  ) |>
  dplyr::mutate(person_id = dplyr::row_number())
```

Construct the actor/partner predictors:

```{r prepare-dyadmlm}
# Construct actor/partner columns without centering the scores again.
flashe_apim <- dyadMLM::prepare_dyad_data(
  flashe_long,
  dyad = dyad_id,
  member = person_id,
  role = role,
  predictors = efficacy_centered,
  model_types = "apim"
)

print(flashe_apim, n = 4)
```

### Fit the model

```{r fit-dyadmlm}
apim_mlm_fit <- glmmTMB::glmmTMB(
  produce_intake ~
    # Separate parent and teen intercepts.
    0 + .is_parent + .is_teen +

    # Actor effects: own efficacy predicts own intake.
    .is_parent:.efficacy_centered_actor +
    .is_teen:.efficacy_centered_actor +

    # Partner effects: partner efficacy predicts own intake.
    .is_parent:.efficacy_centered_partner +
    .is_teen:.efficacy_centered_partner +

    # Two residual variances and their covariance within each dyad.
    us(0 + .is_parent + .is_teen | dyad_id),

  # Represent the paired residual variation in the unstructured block.
  dispformula = ~ 0,
  family = gaussian(),
  data = flashe_apim,

  # Match lavaan's maximum-likelihood estimation.
  REML = FALSE
)

# summary(apim_mlm_fit)
```

### Extract the standard point estimates

We extract the needed coefficients, then add labels matching lavaan.
`dispformula = ~ 0` fixes glmmTMB's separate residual variance at about
1.5e-8, too small to matter. The `us()` block therefore holds the residual
variances and covariance.

```{r extract-glmmtmb-points}
# Extract all nine estimates with their original model names.
glmmtmb_points <- broom.mixed::tidy(
  apim_mlm_fit,
  effects = c("fixed", "ran_pars"),
  scales = c(NA, "vcov") # leaves fixed effects unchanged and requests variances/covariances for REs
) |>
  dplyr::select(effect, term, estimate)

# Rename the rows to match lavaan. Keep term for later model lookups.
glmmtmb_points <- glmmtmb_points |>
  dplyr::mutate(
    parameter = dplyr::recode(
      term,
      ".is_parent" = "intercept_parent",
      ".is_teen" = "intercept_teen",
      ".is_parent:.efficacy_centered_actor" = "actor_parent",
      ".is_teen:.efficacy_centered_actor" = "actor_teen",
      ".is_parent:.efficacy_centered_partner" = "partner_teen_to_parent",
      ".is_teen:.efficacy_centered_partner" = "partner_parent_to_teen",
      "var__.is_parent" = "residual_var_parent",
      "var__.is_teen" = "residual_var_teen",
      "cov__.is_parent..is_teen" = "residual_cov"
    )
  )

glmmtmb_params_point <- dplyr::select(glmmtmb_points, parameter, estimate)
print_table(glmmtmb_params_point)
```

### Calculate the contributions with simple products

We also need the two efficacy variances and their covariance. We calculate
these from one row per dyad, dividing by $n$ as in lavaan's ML estimation.

Note: dividing by n-1 (the default in cov()) might be valid too, but would
not match lavaan exactly.

We then collect all estimates we need for the computation
in a one-row table, `mlm_inputs`.

```{r}

predictor_covariance_matrix <- cov.wt(
  flashe_data[c("efficacy_centered_parent", "efficacy_centered_teen")],
  method = "ML"
)$cov


# Put each fitted estimate in its own column.
model_estimates <- glmmtmb_params_point |>
  tidyr::pivot_wider(names_from = parameter, values_from = estimate)

# Keep the four slopes and residual covariance; add the predictor moments.
mlm_inputs <- model_estimates |>
  dplyr::select(
    actor_parent,
    actor_teen,
    partner_parent_to_teen,
    partner_teen_to_parent,
    residual_cov
  ) |>
  dplyr::mutate(
    var_efficacy_parent = predictor_covariance_matrix[1, 1],
    var_efficacy_teen = predictor_covariance_matrix[2, 2],
    cov_efficacy = predictor_covariance_matrix[1, 2],
    .before = residual_cov
  )

print_table(mlm_inputs)
```


Computing the contributions with the columns directly: 

```{r glmmtmb-point-decomposition}
glmmtmb_decomposition_point <- mlm_inputs |>
  dplyr::mutate(
    actor_driven = actor_teen * cov_efficacy * actor_parent,
    partner_driven = partner_parent_to_teen * cov_efficacy * partner_teen_to_parent,
    teen_driven = actor_teen * var_efficacy_teen * partner_teen_to_parent,
    parent_driven = actor_parent * var_efficacy_parent * partner_parent_to_teen,
    residual = residual_cov,
    
    explained = actor_driven + partner_driven + teen_driven + parent_driven,
    
    total_covariance = explained + residual,
    
    .keep = "none"
  ) |>
  # Return to one row per contribution for the display table.
  tidyr::pivot_longer(
    dplyr::everything(), names_to = "parameter", values_to = "estimate"
  ) |>
  # Divide each estimate by the total covariance row, then multiply by 100.
  dplyr::mutate(
    percent = 100 * (estimate / estimate[parameter == "total_covariance"])
  )
print_table(glmmtmb_decomposition_point)
```

### Compare the point estimates with lavaan

The differences are below 1e-4 and come from the optimizers' stopping rules (tested).

```{r compare-parameter-points}
parameter_point_comparison <- dplyr::inner_join(
  dplyr::select(lavaan_params, parameter, estimate),
  glmmtmb_params_point,
  by = "parameter",
  suffix = c("_lavaan", "_glmmtmb")
) |>
  dplyr::mutate(difference = estimate_glmmtmb - estimate_lavaan)
print_table(parameter_point_comparison)
```

### Plot the point estimates

Visualize the contribution to the total covariance.

```{r prepare-decomposition-plots}
library(ggplot2)

route_labels <- c(
  actor_driven = "Actor-driven", partner_driven = "Partner-driven",
  teen_driven = "Teen-driven", parent_driven = "Parent-driven",
  residual = "Residual"
)
# Muted teal, ochre, blue, mauve and slate, in the same order as the routes.
route_colours <- c("#468C83", "#D3A064", "#6F91B7", "#AD8CA7", "#7B8794")
plot_data <- dplyr::bind_rows(
  lavaan = dplyr::select(lavaan_decomposition, parameter, estimate),
  glmmTMB = dplyr::select(glmmtmb_decomposition_point, parameter, estimate),
  .id = "model"
) |>
  dplyr::filter(parameter %in% names(route_labels)) |>
  dplyr::mutate(
    model = factor(model, levels = c("lavaan", "glmmTMB")),
    route = factor(parameter, levels = names(route_labels),
                   labels = unname(route_labels))
  ) |>
  dplyr::arrange(model, route) |>
  dplyr::group_by(model) |>
  dplyr::mutate(
    end = cumsum(estimate), start = end - estimate,
    percent = 100 * estimate / sum(estimate)
  )

plot_totals <- dplyr::summarise(plot_data, total = sum(estimate))
```

#### Waterfall: how the contributions add up

```{r decomposition-waterfall, dev="svg", fig.width=10, fig.height=4}
ggplot(plot_data, aes(fill = route)) +
  geom_rect(aes(
    xmin = as.numeric(route) - 0.4, xmax = as.numeric(route) + 0.4,
    ymin = pmin(start, end), ymax = pmax(start, end)
  )) +
  geom_text(
    aes(x = as.numeric(route), y = pmax(start, end),
        label = sprintf("%.1f%%", percent)),
    vjust = -0.6, colour = "#475569", size = 3.5
  ) +
  geom_hline(
    data = plot_totals, aes(yintercept = total),
    linetype = "dotted", colour = "#475569"
  ) +
  geom_text(
    data = plot_totals,
    aes(x = 0.6, y = total, label = sprintf("Total = %.3f", total)),
    inherit.aes = FALSE, hjust = 0, vjust = -0.6, colour = "#475569"
  ) +
  facet_wrap(~ model, nrow = 1) +
  scale_x_continuous(
    breaks = seq_along(route_labels), labels = gsub("-", "\n", route_labels)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  scale_fill_manual(values = route_colours) +
  labs(x = NULL, y = "Outcome covariance") +
  theme_minimal() +
  theme(legend.position = "none")
```

#### Stacked bars: compare the two fits

Here all five contributions are positive, so each bar's length is the total
covariance. With negative contributions, ggplot would place positive and
negative pieces on opposite sides of zero. The waterfall above might make
their net total easier to see.

```{r decomposition-stacked-bars, dev="svg", fig.width=10, fig.height=3}
ggplot(plot_data, aes(x = estimate, y = model, fill = route)) +
  geom_col(width = 0.5, position = position_stack(reverse = TRUE)) +
  # Put labels for tiny segments below the bar to avoid overlap.
  geom_text(
    aes(x = (start + end) / 2,
        y = as.numeric(model) + ifelse(abs(percent) < 1, -0.35, 0.35),
        label = sprintf("%.1f%%", percent)),
    colour = "#475569", size = 3.5
  ) +
  scale_fill_manual(values = route_colours) +
  labs(x = "Outcome covariance", y = NULL, fill = "Contribution") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3. Add confidence intervals for glmmTMB

We use the same bootstrap method as lavaan in section 1. We draw 1486 dyads
with replacement and refit the glmmTMB model from section 2 to them. We do
this 1999 times. For each estimate, the 2.5% and 97.5% quantiles of the 1999
values give the 95% CI. This does not assume normally distributed data.

### Functions for one bootstrap sample

`draw_dyads()` takes the drawn dyads from the long data of section 2.
`estimate_apim()` refits the section 2 model to these data and collects all
estimates as in section 2. `add_contributions()` contains the decomposition
formulas from section 2.

```{r bootstrap-functions}
# Take the drawn dyads from the long data, two rows each. Each drawn dyad gets
# a new ID, so a dyad drawn twice counts as two dyads.
draw_dyads <- function(long_data, dyads) {
  tibble::tibble(dyad_id = dyads, new_dyad_id = seq_along(dyads)) |>
    # many-to-many: a dyad can be drawn twice, and each dyad has two rows.
    dplyr::inner_join(
      long_data, by = "dyad_id", relationship = "many-to-many"
    ) |>
    dplyr::mutate(dyad_id = new_dyad_id) |>
    dplyr::select(-new_dyad_id)
}

# Refit the model to the long data, and compute the predictor variances and
# covariance (divided by n) from the wide data, as in section 2. Returns every
# estimate in one row, with lavaan's names.
estimate_apim <- function(model, long_data, wide_data) {
  # update() refits the model, with the same formula and settings, to new data.
  fit <- update(model, data = long_data)
  fixed_effects <- glmmTMB::fixef(fit)$cond
  residual_covariance_matrix <- glmmTMB::VarCorr(fit)$cond$dyad_id
  predictor_covariance_matrix <- cov.wt(
    wide_data[c("efficacy_centered_parent", "efficacy_centered_teen")],
    method = "ML"
  )$cov

  # In both matrices, parent comes first.
  tibble::tibble(
    intercept_parent = fixed_effects[[".is_parent"]],
    intercept_teen = fixed_effects[[".is_teen"]],
    actor_parent = fixed_effects[[".is_parent:.efficacy_centered_actor"]],
    actor_teen = fixed_effects[[".is_teen:.efficacy_centered_actor"]],
    partner_teen_to_parent = fixed_effects[[".is_parent:.efficacy_centered_partner"]],
    partner_parent_to_teen = fixed_effects[[".is_teen:.efficacy_centered_partner"]],
    residual_var_parent = residual_covariance_matrix[1, 1],
    residual_var_teen = residual_covariance_matrix[2, 2],
    residual_cov = residual_covariance_matrix[1, 2],
    var_efficacy_parent = predictor_covariance_matrix[1, 1],
    var_efficacy_teen = predictor_covariance_matrix[2, 2],
    cov_efficacy = predictor_covariance_matrix[1, 2]
  ) |>
    add_contributions()
}

# The decomposition formulas from section 2.
add_contributions <- function(estimates) {
  estimates |>
    dplyr::mutate(
      actor_driven = actor_teen * cov_efficacy * actor_parent,
      partner_driven = partner_parent_to_teen * cov_efficacy * partner_teen_to_parent,
      teen_driven = actor_teen * var_efficacy_teen * partner_teen_to_parent,
      parent_driven = actor_parent * var_efficacy_parent * partner_parent_to_teen,
      residual = residual_cov,
      explained = actor_driven + partner_driven + teen_driven + parent_driven,
      total_covariance = explained + residual
    )
}
```

### Bootstrap

```{r bootstrap-cache-key}
# knitr caches the next chunk. It reruns only if that chunk changes or if
# the data, the model fit or the three functions above change.
bootstrap_key <- digest::digest(list(
  flashe_data, logLik(apim_mlm_fit), deparse(draw_dyads),
  deparse(estimate_apim), deparse(add_contributions)
))
```

```{r glmmtmb-bootstrap, cache = TRUE, cache.extra = bootstrap_key}
# 1999 samples of 1486 dyads, drawn with replacement. A dyad's number is its
# row in flashe_data and its dyad_id in flashe_apim.
set.seed(2026)
bootstrap <- tibble::tibble(
  dyads = replicate(
    1999, sample(nrow(flashe_data), replace = TRUE), simplify = FALSE
  )
)

# Refit the model to each sample (about 5 to 15 minutes on 4 cores).
bootstrap <- bootstrap |>
  dplyr::mutate(estimates = parallel::mclapply(
    dyads,
    \(drawn) estimate_apim(
      model = apim_mlm_fit,
      long_data = draw_dyads(flashe_apim, drawn),
      wide_data = flashe_data[drawn, ]
    ),
    mc.cores = 4
  ))
```

```{r glmmtmb-cis}
# 95% CI for each column: its 2.5% and 97.5% quantiles.
percentile_cis <- function(draws) {
  draws |>
    tidyr::pivot_longer(
      dplyr::everything(), names_to = "parameter", values_to = "draw"
    ) |>
    dplyr::summarise(
      ci_low = quantile(draw, 0.025),
      ci_high = quantile(draw, 0.975),
      .by = parameter
    )
}

# One row per sample, one column per estimate.
glmmtmb_draws <- dplyr::bind_rows(bootstrap$estimates)

# Estimates from the full data, with their 95% CIs.
glmmtmb_cis <- estimate_apim(apim_mlm_fit, flashe_apim, flashe_data) |>
  tidyr::pivot_longer(
    dplyr::everything(), names_to = "parameter", values_to = "estimate"
  ) |>
  dplyr::left_join(percentile_cis(glmmtmb_draws), by = "parameter")
```

### Compare the CIs with lavaan

```{r compare-cis}
# Keep the estimates that both models have.
ci_comparison <- dplyr::inner_join(
  lavaan_estimates,
  glmmtmb_cis,
  by = "parameter",
  suffix = c("_lavaan", "_glmmtmb")
) |>
  dplyr::mutate(
    ci_lavaan = sprintf("[%.3f, %.3f]", ci_low_lavaan, ci_high_lavaan),
    ci_glmmtmb = sprintf("[%.3f, %.3f]", ci_low_glmmtmb, ci_high_glmmtmb)
  ) |>
  dplyr::select(parameter, ci_lavaan, ci_glmmtmb)
print_table(ci_comparison)
```

The CIs are very close. They differ a little because lavaan and glmmTMB
drew different random samples. 

## 4. Plot the confidence intervals

```{r decomposition-intervals, dev="svg", fig.width=10, fig.height=4}
ci_plot_data <- dplyr::bind_rows(
  lavaan = lavaan_decomposition, glmmTMB = glmmtmb_cis,
  .id = "model"
) |>
  dplyr::filter(parameter %in% names(route_labels)) |>
  dplyr::mutate(
    model = factor(model, levels = c("lavaan", "glmmTMB")),
    route = factor(parameter, levels = names(route_labels),
                   labels = unname(route_labels))
  )

ggplot(ci_plot_data, aes(x = estimate, y = route, colour = route)) +
  geom_pointrange(aes(xmin = ci_low, xmax = ci_high),
                  show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  facet_wrap(~ model) +
  # List the contributions from top to bottom, in the order of the tables.
  scale_y_discrete(limits = rev) +
  scale_colour_manual(values = route_colours) +
  labs(x = "Covariance contribution", y = NULL) +
  theme_minimal()
```


## 5. The routes as path diagrams

The numbers of the path diagrams are the glmmTMB estimates from section 2. 
The stars are bootstrap p-values from the draws in section 3. 

`apim-route-diagrams.R` draws the diagrams. It numbers the members: the teen
is member 1 and the parent member 2.

```{r route-diagram-inputs, echo=FALSE}
source("apim-route-diagrams.R")

# Two-sided bootstrap p-value: twice the share of draws on the far side of 0.
bootstrap_p_values <- glmmtmb_draws |>
  tidyr::pivot_longer(
    dplyr::everything(), names_to = "parameter", values_to = "draw"
  ) |>
  dplyr::summarise(
    p_value = min(1, 2 * min(mean(draw <= 0), mean(draw >= 0))),
    .by = parameter
  )

# Estimates and p-values with the diagram's names (teen = 1, parent = 2).
diagram_inputs <- glmmtmb_cis |>
  dplyr::left_join(bootstrap_p_values, by = "parameter") |>
  dplyr::mutate(diagram_name = dplyr::recode(
    parameter,
    actor_teen = "a1", actor_parent = "a2",
    partner_parent_to_teen = "p1", partner_teen_to_parent = "p2",
    var_efficacy_teen = "var_x1", var_efficacy_parent = "var_x2",
    cov_efficacy = "cov_x1_x2", residual_cov = "psi",
    teen_driven = "x1_driven", parent_driven = "x2_driven"
  ))
diagram_estimates <- tibble::deframe(
  dplyr::select(diagram_inputs, diagram_name, estimate)
)
# Variances are always positive, so they get no p-value.
diagram_p_values <- diagram_inputs |>
  dplyr::filter(!startsWith(diagram_name, "var_")) |>
  dplyr::select(diagram_name, p_value) |>
  tibble::deframe()
```

```{r route-diagrams, dev="svg", fig.width=12, fig.height=10.4, echo=FALSE}
draw_apim_route_grid(
  list(
    "1. Actor-driven" = "actor_driven", "2. Partner-driven" = "partner_driven",
    "3. Teen-driven" = "x1_driven", "4. Parent-driven" = "x2_driven",
    "5. Residual" = "residual", "All routes" = "all"
  ),
  formulas = list("All routes" = quote(Cov(Y[S1], Y[S2]))),
  labels = list(predictor = "Efficacy", outcome = "Intake"),
  roles = c("teen", "parent"),
  values = diagram_estimates, p_values = diagram_p_values,
  contributions = diagram_estimates
)
```

## 6. Example with a covariate

We add the parent's gender as a covariate (`parent_female`, 1 = mother,
0 = father). 

### Fit the model with the covariate

```{r covariate-fit}
# The parent's gender (1 = mother), on both rows of each dyad.
covariate_apim <- flashe_apim |>
  dplyr::mutate(parent_female = is_female[role == "parent"], .by = dyad_id)

# The model from section 2, plus the covariate's effect on each intake.
covariate_fit <- update(
  apim_mlm_fit,
  . ~ . + .is_teen:parent_female + .is_parent:parent_female,
  data = covariate_apim
)

fixed_effects <- glmmTMB::fixef(covariate_fit)$cond
residual_covariance_matrix <- glmmTMB::VarCorr(covariate_fit)$cond$dyad_id
# Rows and columns: teen efficacy, parent efficacy, parent gender.
predictor_moments <- cov.wt(
  flashe_data[c(
    "efficacy_centered_teen", "efficacy_centered_parent", "is_female_parent"
  )],
  method = "ML"
)$cov

# All estimates we need, in one row.
covariate_estimates <- tibble::tibble(
  actor_teen = fixed_effects[[".is_teen:.efficacy_centered_actor"]],
  actor_parent = fixed_effects[[".is_parent:.efficacy_centered_actor"]],
  partner_parent_to_teen = fixed_effects[[".is_teen:.efficacy_centered_partner"]],
  partner_teen_to_parent = fixed_effects[[".is_parent:.efficacy_centered_partner"]],
  gender_teen = fixed_effects[[".is_teen:parent_female"]],
  gender_parent = fixed_effects[[".is_parent:parent_female"]],
  var_efficacy_teen = predictor_moments[1, 1],
  var_efficacy_parent = predictor_moments[2, 2],
  var_gender = predictor_moments[3, 3],
  cov_efficacy = predictor_moments[1, 2],
  cov_teen_gender = predictor_moments[1, 3],
  cov_parent_gender = predictor_moments[2, 3],
  residual_cov = residual_covariance_matrix[1, 2]
)
print_table(covariate_estimates)
```

### Compute all ten routes

```{r covariate-routes}
covariate_routes <- covariate_estimates |>
  dplyr::transmute(
    # The four APIM routes, with slopes adjusted for the covariate.
    actor_driven = actor_teen * cov_efficacy * actor_parent,
    partner_driven = partner_parent_to_teen * cov_efficacy * partner_teen_to_parent,
    teen_driven = actor_teen * var_efficacy_teen * partner_teen_to_parent,
    parent_driven = partner_parent_to_teen * var_efficacy_parent * actor_parent,
    # The covariate as a common cause of both intakes.
    parent_gender_driven = gender_teen * var_gender * gender_parent,
    # One path from efficacy and one from the covariate.
    teen_actor_with_gender = actor_teen * cov_teen_gender * gender_parent,
    gender_with_parent_partner = gender_teen * cov_teen_gender * partner_teen_to_parent,
    gender_with_parent_actor = gender_teen * cov_parent_gender * actor_parent,
    teen_partner_with_gender = partner_parent_to_teen * cov_parent_gender * gender_parent,
    residual = residual_cov
  ) |>
  tidyr::pivot_longer(
    dplyr::everything(), names_to = "route", values_to = "estimate"
  )
print_table(covariate_routes)

# The routes add up to the observed covariance of the two intakes.
tibble::tibble(
  sum_of_routes = sum(covariate_routes$estimate),
  observed = cov.wt(
    flashe_data[c("produce_intake_teen", "produce_intake_parent")],
    method = "ML"
  )$cov[1, 2]
)
```

The first diagram shows all paths, variances (σ²) and covariances (σ) with
their estimates. C stands for the covariate.

```{r covariate-all-paths, dev="svg", fig.width=9, fig.height=5.4, echo=FALSE}
# Our names, and the matching names in the diagram.
covariate_diagram_values <- with(covariate_estimates, c(
  a1 = actor_teen, a2 = actor_parent,
  p1 = partner_parent_to_teen, p2 = partner_teen_to_parent,
  c1 = gender_teen, c2 = gender_parent,
  var_x1 = var_efficacy_teen, var_x2 = var_efficacy_parent, var_c = var_gender,
  cov_x1_x2 = cov_efficacy, cov_x1_c = cov_teen_gender,
  cov_x2_c = cov_parent_gender, psi = residual_cov
))
covariate_contributions <- covariate_routes |>
  dplyr::mutate(route = dplyr::recode(
    route,
    teen_driven = "x1_driven", parent_driven = "x2_driven",
    parent_gender_driven = "covariate_driven",
    teen_actor_with_gender = "x1_actor_with_c",
    gender_with_parent_partner = "c_with_x1_partner",
    gender_with_parent_actor = "c_with_x2_actor",
    teen_partner_with_gender = "x2_partner_with_c"
  )) |>
  tibble::deframe()

draw_apim_routes(
  "all", covariate = TRUE,
  labels = list(predictor = "Efficacy", outcome = "Intake", covariate = "Parent gender"),
  roles = c("teen", "parent"), values = covariate_diagram_values,
  contributions = covariate_contributions,
  formula = quote(Cov(Y[S1], Y[S2]))
)
```

```{r covariate-route-diagrams, dev="svg", fig.width=13, fig.height=19, echo=FALSE}
draw_apim_route_grid(
  list(
    "1. Actor-driven" = "actor_driven", "2. Partner-driven" = "partner_driven",
    "3. Teen-driven" = "x1_driven", "4. Parent-driven" = "x2_driven",
    "5. Parent-gender-driven" = "covariate_driven",
    "6. Teen actor with gender" = "x1_actor_with_c",
    "7. Gender with parent partner" = "c_with_x1_partner",
    "8. Gender with parent actor" = "c_with_x2_actor",
    "9. Teen partner with gender" = "x2_partner_with_c",
    "10. Residual" = "residual"
  ),
  covariate = TRUE,
  labels = list(predictor = "Efficacy", outcome = "Intake", covariate = "Parent gender"),
  roles = c("teen", "parent"), values = covariate_diagram_values,
  contributions = covariate_contributions, text_scale = 0.8
)
```

### Group the routes

We keep the four APIM routes separate, as without the covariate. The
parent-gender-driven route stays on its own. We add up the four mixed routes
(6 to 9) since each is small and hard to interpret alone. Together, they are the
part of the outcome covariance that comes from efficacy and parent gender
being correlated. The residual comes last. The table compares each group
with the model without the covariate.

```{r covariate-group-table}
route_groups <- c(
  actor_driven = "Actor-driven", partner_driven = "Partner-driven",
  teen_driven = "Teen-driven", parent_driven = "Parent-driven",
  parent_gender_driven = "Parent-gender-driven",
  teen_actor_with_gender = "Efficacy and parent gender together",
  gender_with_parent_partner = "Efficacy and parent gender together",
  gender_with_parent_actor = "Efficacy and parent gender together",
  teen_partner_with_gender = "Efficacy and parent gender together",
  residual = "Residual"
)

covariate_groups <- covariate_routes |>
  dplyr::mutate(group = factor(route_groups[route], unique(route_groups))) |>
  dplyr::summarise(estimate = sum(estimate), .by = group) |>
  dplyr::mutate(percent = 100 * estimate / sum(estimate))

# The same routes from the model without the covariate (section 3).
without_covariate <- glmmtmb_cis |>
  dplyr::filter(parameter %in% names(route_labels)) |>
  dplyr::transmute(
    group = route_labels[parameter], estimate_without_covariate = estimate
  )

covariate_group_table <- covariate_groups |>
  dplyr::left_join(without_covariate, by = "group")
print_table(covariate_group_table)
```

### Plot the groups

```{r covariate-waterfall, dev="svg", fig.width=10, fig.height=4}
covariate_plot_data <- covariate_groups |>
  dplyr::mutate(end = cumsum(estimate), start = end - estimate)

ggplot(covariate_plot_data, aes(fill = group)) +
  geom_rect(aes(
    xmin = as.numeric(group) - 0.4, xmax = as.numeric(group) + 0.4,
    ymin = pmin(start, end), ymax = pmax(start, end)
  )) +
  geom_text(
    aes(x = as.numeric(group), y = pmax(start, end),
        label = sprintf("%.1f%%", percent)),
    vjust = -0.6, colour = "#475569", size = 3.5
  ) +
  geom_hline(
    yintercept = sum(covariate_groups$estimate),
    linetype = "dotted", colour = "#475569"
  ) +
  scale_x_continuous(
    breaks = seq_along(levels(covariate_groups$group)),
    labels = scales::label_wrap(12)(levels(covariate_groups$group))
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  # The section 2 colours, plus two greens for the covariate groups.
  scale_fill_manual(values = c(
    route_colours[1:4], "#5E9E6E", "#A9CDB0", route_colours[5]
  )) +
  labs(x = NULL, y = "Outcome covariance") +
  theme_minimal() +
  theme(legend.position = "none")
```
