---
title: "APIM exploration: lavaan and glmmTMB"
author: "Pascal Küng"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    math_method: mathml
    df_print: kable
    css: explore-sem-mlm.css
---

This notebook shows the full workflow for the same cross-sectional Gaussian
APIM in lavaan and glmmTMB. Both give the same point estimates of the
covariance decomposition, and their bootstrap confidence intervals agree
within resampling noise. Each dyad has one parent and one teen, with complete
data. Further checks and other CI methods are in
`explore-sem-mlm-details.Rmd`.

```{r table-style, include=FALSE}
# Each table includes its CSS when chunks are run separately.
table_css <- paste(readLines("explore-sem-mlm.css"), collapse = "\n")

print_table <- function(data) {
  htmltools::browsable(htmltools::tagList(
    htmltools::tags$style(htmltools::HTML(table_css)),
    htmltools::HTML(knitr::kable(
      data, format = "html", table.attr = 'class="apim-table"'
    ))
  ))
}
```

## Load and rename the data

```{r load-flashe}
library(lavaan)
flashe_data <- read.csv(
  file.path(
    "../references/explaining-interdependence-apim",
    "jp-materials-2026-09-13",
    "J-P&Niall-FLASHE-example",
    "flashe-lavaan.csv"
  ),
  header = TRUE
)

# Produce intake and efficacy refer to fruit and vegetable consumption.
# Keep the supplied scores centered at 7.6.
flashe_data <- dplyr::rename(
  flashe_data,
  produce_intake_parent = pftveg,
  produce_intake_teen = tftveg,
  efficacy_parent = peffv,
  efficacy_teen = teffv,
  efficacy_centered_parent = cpeffv,
  efficacy_centered_teen = cteffv,
  is_female_parent = pfem,
  is_female_teen = tfem
)
```

## 1. Lavaan: J-P and Niall's example

### Fit the model

```{r jp-niall-lavaan}
apim_model <- '
  # Intake intercepts: both centered efficacy predictors are zero.
  produce_intake_teen ~ intercept_teen * 1
  produce_intake_parent ~ intercept_parent * 1
  
  # Actor effects: own centered efficacy predicts own F&V intake.
  produce_intake_teen ~ actor_teen * efficacy_centered_teen
  produce_intake_parent ~ actor_parent * efficacy_centered_parent

  # Partner effects: parent efficacy -> teen intake, and vice versa.
  produce_intake_teen ~ partner_parent_to_teen * efficacy_centered_parent
  produce_intake_parent ~ partner_teen_to_parent * efficacy_centered_teen

  # Means of the centered efficacy predictors.
  efficacy_centered_teen ~ mean_efficacy_teen * 1
  efficacy_centered_parent ~ mean_efficacy_parent * 1

  # Predictor variances.
  efficacy_centered_teen ~~ var_efficacy_teen * efficacy_centered_teen
  efficacy_centered_parent ~~ var_efficacy_parent * efficacy_centered_parent

  # Residual variances of intake.
  produce_intake_teen ~~ residual_var_teen * produce_intake_teen
  produce_intake_parent ~~ residual_var_parent * produce_intake_parent

  # Covariance between parent and teen efficacy.
  efficacy_centered_parent ~~ cov_efficacy * efficacy_centered_teen

  # Residual covariance between parent and teen intake.
  produce_intake_parent ~~ residual_cov * produce_intake_teen

  # Path-traced contributions to the outcome covariance.
  actor_driven := actor_teen * cov_efficacy * actor_parent
  partner_driven := partner_parent_to_teen * cov_efficacy * partner_teen_to_parent
  teen_driven := actor_teen * var_efficacy_teen * partner_teen_to_parent
  parent_driven := actor_parent * var_efficacy_parent * partner_parent_to_teen
  residual := residual_cov
  explained := actor_driven + partner_driven + teen_driven + parent_driven
  total_covariance := explained + residual

  # Percentages of the total outcome covariance.
  percent_actor_driven := 100 * (actor_driven / total_covariance)
  percent_partner_driven := 100 * (partner_driven / total_covariance)
  percent_teen_driven := 100 * (teen_driven / total_covariance)
  percent_parent_driven := 100 * (parent_driven / total_covariance)
  percent_residual := 100 * (residual / total_covariance)
  percent_explained := 100 * (explained / total_covariance)
'

# Predictor means, variances and covariance are free parameters above.
# Bootstrap CIs: refit to 1999 resamples of the dyads (15 to 30 s on 4 cores).
apim_fit <- sem(
  apim_model, fixed.x = FALSE, data = flashe_data, missing = "fiml",
  se = "bootstrap", bootstrap = 1999, iseed = 2026,
  parallel = "multicore", ncpus = 4
)

# Examine the model.
# summary(apim_fit, fit.measures = TRUE)
```

### Extract the model parameters

Lavaan provides both the fitted parameters and the quantities defined with
`:=`. We extract them once, then display the model parameters and decomposition
separately. The model parameters include the predictor means, variances and
covariance as well as the outcome coefficients and residual structure.

```{r extract-lavaan}
# Extract once, using our model labels and including 95% CIs.
lavaan_estimates <- broom::tidy(apim_fit, conf.int = TRUE) |>
  dplyr::select(
    parameter = label,
    op,
    estimate,
    ci_low = conf.low,
    ci_high = conf.high
  )

# Model parameters: exclude the quantities defined with :=.
lavaan_params <- lavaan_estimates |>
  dplyr::filter(op != ":=") |>
  dplyr::select(-op)

print_table(lavaan_params)
```

### Extract the covariance decomposition

The `:=` definitions above apply the Day 2 handout formulas. They calculate
contributions and percentages from the estimates without changing the fitted
model. With `se = "bootstrap"`, lavaan recalculates them in each of the 1999
resamples. The middle 95% of these values gives each 95% confidence interval
(CI). Each resample re-estimates all parameters, including the predictor
variances and covariance, so the CIs include all their uncertainty.

We select these estimates and CIs from `lavaan_estimates`:

```{r lavaan-covariance-decomposition}
# Decomposition: keep only the quantities defined with :=.
lavaan_decomposition <- lavaan_estimates |>
  dplyr::filter(op == ":=") |>
  dplyr::select(-op)

# Display the five contributions and the explained subtotal, with CI and percent.
format_decomposition <- function(decomposition) {
  # Match each percentage row to its contribution by removing "percent_".
  percentages <- decomposition |>
    dplyr::filter(startsWith(parameter, "percent_")) |>
    dplyr::mutate(parameter = sub("^percent_", "", parameter)) |>
    dplyr::select(parameter, percent = estimate)

  # Format only the displayed CI; estimates keep their full precision.
  decomposition |>
    dplyr::filter(!startsWith(parameter, "percent_"),
                  parameter != "total_covariance") |>
    dplyr::mutate(
      `95% CI` = sprintf("[%.3f, %.3f]", ci_low, ci_high)
    ) |>
    dplyr::select(parameter, estimate, `95% CI`) |>
    dplyr::left_join(percentages, by = "parameter")
}

lavaan_decomposition_table <- format_decomposition(lavaan_decomposition)
print_table(lavaan_decomposition_table)
```


## 2. glmmTMB: reproduce the point estimates

`dyadMLM` prepares the actor and partner predictors; `glmmTMB` fits the model.
We retain the supplied centering and use maximum likelihood, as in lavaan.

### Reshape and prepare the data

Reshape to one row per person, then construct the actor/partner predictors.

```{r prepare-dyadmlm}
# Each original row is one dyad. Long format has one row per person.
flashe_long <- flashe_data |>
  dplyr::mutate(dyad_id = dplyr::row_number()) |>
  tidyr::pivot_longer(
    cols = -dyad_id,
    names_to = c(".value", "role"),
    names_pattern = "(.+)_(parent|teen)$"
  ) |>
  dplyr::mutate(person_id = dplyr::row_number())

# Construct actor/partner columns without centering the scores again.
flashe_apim <- dyadMLM::prepare_dyad_data(
  flashe_long,
  dyad = dyad_id,
  member = person_id,
  role = role,
  predictors = efficacy_centered,
  model_types = "apim"
)

print(flashe_apim, n = 4)
```

### Fit the model

```{r fit-dyadmlm}
apim_mlm_fit <- glmmTMB::glmmTMB(
  produce_intake ~
    # Separate parent and teen intercepts.
    0 + .is_parent + .is_teen +

    # Actor effects: own efficacy predicts own intake.
    .is_parent:.efficacy_centered_actor +
    .is_teen:.efficacy_centered_actor +

    # Partner effects: partner efficacy predicts own intake.
    .is_parent:.efficacy_centered_partner +
    .is_teen:.efficacy_centered_partner +

    # Two residual variances and their covariance within each dyad.
    us(0 + .is_parent + .is_teen | dyad_id),

  # Represent the paired residual variation in the unstructured block.
  dispformula = ~ 0,
  family = gaussian(),
  data = flashe_apim,

  # Match lavaan's maximum-likelihood estimation.
  REML = FALSE
)

# summary(apim_mlm_fit)
```

### Extract the standard point estimates

We extract the needed coefficients, then add labels matching lavaan.
`dispformula = ~ 0` fixes glmmTMB's separate residual variance at about
1.5e-8, too small to matter. It does not affect the covariance. The `us()`
block therefore holds the residual variances and covariance.

```{r extract-glmmtmb-points}
# Extract all nine estimates with their original model names.
glmmtmb_points <- broom.mixed::tidy(
  apim_mlm_fit,
  effects = c("fixed", "ran_pars"),
  scales = c(NA, "vcov") # leaves fixed effects unchanged and requests variances/covariances for REs
) |>
  dplyr::select(effect, term, estimate)

# Rename the rows to match lavaan. Keep term for later model lookups.
glmmtmb_points <- glmmtmb_points |>
  dplyr::mutate(
    parameter = dplyr::recode(
      term,
      ".is_parent" = "intercept_parent",
      ".is_teen" = "intercept_teen",
      ".is_parent:.efficacy_centered_actor" = "actor_parent",
      ".is_teen:.efficacy_centered_actor" = "actor_teen",
      ".is_parent:.efficacy_centered_partner" = "partner_teen_to_parent",
      ".is_teen:.efficacy_centered_partner" = "partner_parent_to_teen",
      "var__.is_parent" = "residual_var_parent",
      "var__.is_teen" = "residual_var_teen",
      "cov__.is_parent..is_teen" = "residual_cov"
    )
  )

glmmtmb_params_point <- dplyr::select(glmmtmb_points, parameter, estimate)
print_table(glmmtmb_params_point)
```

### Calculate the contributions with simple products

We also need the two efficacy variances and their covariance. We calculate
these from one row per dyad, dividing by $n$ as in lavaan's ML estimation.

We then collect all estimates we need for the computation
in a one-row table, `mlm_inputs`.

```{r}

predictor_covariance_matrix <- cov.wt(
  flashe_data[c("efficacy_centered_parent", "efficacy_centered_teen")],
  method = "ML"
)$cov


# Put each fitted estimate in its own column.
model_estimates <- glmmtmb_params_point |>
  tidyr::pivot_wider(names_from = parameter, values_from = estimate)

# Keep the four slopes and residual covariance; add the predictor moments.
mlm_inputs <- model_estimates |>
  dplyr::select(
    actor_parent,
    actor_teen,
    partner_parent_to_teen,
    partner_teen_to_parent,
    residual_cov
  ) |>
  dplyr::mutate(
    var_efficacy_parent = predictor_covariance_matrix[1, 1],
    var_efficacy_teen = predictor_covariance_matrix[2, 2],
    cov_efficacy = predictor_covariance_matrix[1, 2],
    .before = residual_cov
  )

print_table(mlm_inputs)
```


Computing the contributions with the columns directly: 

```{r glmmtmb-point-decomposition}
glmmtmb_decomposition_point <- mlm_inputs |>
  dplyr::mutate(
    actor_driven = actor_teen * cov_efficacy * actor_parent,
    partner_driven = partner_parent_to_teen * cov_efficacy * partner_teen_to_parent,
    teen_driven = actor_teen * var_efficacy_teen * partner_teen_to_parent,
    parent_driven = actor_parent * var_efficacy_parent * partner_parent_to_teen,
    residual = residual_cov,
    
    explained = actor_driven + partner_driven + teen_driven + parent_driven,
    
    total_covariance = explained + residual,
    
    .keep = "none"
  ) |>
  # Return to one row per contribution for the display table.
  tidyr::pivot_longer(
    dplyr::everything(), names_to = "parameter", values_to = "estimate"
  ) |>
  # Divide each estimate by the total covariance row, then multiply by 100.
  dplyr::mutate(
    percent = 100 * (estimate / estimate[parameter == "total_covariance"])
  )
print_table(glmmtmb_decomposition_point)
```

### Compare the point estimates with lavaan

We compare the nine parameters estimated by both models, then the covariance
contributions. The predictor means, variances and covariance are fitted
parameters in lavaan. glmmTMB treats the predictors as observed values.
The differences are below 1e-4 and come from the optimizers' stopping rules.

```{r compare-parameter-points}
parameter_point_comparison <- dplyr::inner_join(
  dplyr::select(lavaan_params, parameter, estimate),
  glmmtmb_params_point,
  by = "parameter",
  suffix = c("_lavaan", "_glmmtmb")
) |>
  dplyr::mutate(difference = estimate_glmmtmb - estimate_lavaan)
print_table(parameter_point_comparison)
```

### Plot the point estimates

Visualize the contribution to the total covariance.

```{r prepare-decomposition-plots}
library(ggplot2)

route_labels <- c(
  actor_driven = "Actor-driven", partner_driven = "Partner-driven",
  teen_driven = "Teen-driven", parent_driven = "Parent-driven",
  residual = "Residual"
)
# Muted teal, ochre, blue, mauve and slate, in the same order as the routes.
route_colours <- c("#468C83", "#D3A064", "#6F91B7", "#AD8CA7", "#7B8794")
plot_data <- dplyr::bind_rows(
  lavaan = dplyr::select(lavaan_decomposition, parameter, estimate),
  glmmTMB = dplyr::select(glmmtmb_decomposition_point, parameter, estimate),
  .id = "model"
) |>
  dplyr::filter(parameter %in% names(route_labels)) |>
  dplyr::mutate(
    model = factor(model, levels = c("lavaan", "glmmTMB")),
    route = factor(parameter, levels = names(route_labels),
                   labels = unname(route_labels))
  ) |>
  dplyr::arrange(model, route) |>
  dplyr::group_by(model) |>
  dplyr::mutate(
    end = cumsum(estimate), start = end - estimate,
    percent = 100 * estimate / sum(estimate)
  )

plot_totals <- dplyr::summarise(plot_data, total = sum(estimate))
```

#### Waterfall: how the contributions add up

```{r decomposition-waterfall, dev="svg", fig.width=10, fig.height=4}
ggplot(plot_data, aes(fill = route)) +
  geom_rect(aes(
    xmin = as.numeric(route) - 0.4, xmax = as.numeric(route) + 0.4,
    ymin = pmin(start, end), ymax = pmax(start, end)
  )) +
  geom_text(
    aes(x = as.numeric(route), y = pmax(start, end),
        label = sprintf("%.1f%%", percent)),
    vjust = -0.6, colour = "#475569", size = 3.5
  ) +
  geom_hline(
    data = plot_totals, aes(yintercept = total),
    linetype = "dotted", colour = "#475569"
  ) +
  geom_text(
    data = plot_totals,
    aes(x = 0.6, y = total, label = sprintf("Total = %.3f", total)),
    inherit.aes = FALSE, hjust = 0, vjust = -0.6, colour = "#475569"
  ) +
  facet_wrap(~ model, nrow = 1) +
  scale_x_continuous(
    breaks = seq_along(route_labels), labels = gsub("-", "\n", route_labels)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  scale_fill_manual(values = route_colours) +
  labs(x = NULL, y = "Outcome covariance") +
  theme_minimal() +
  theme(legend.position = "none")
```

#### Stacked bars: compare the two fits

Here all five contributions are positive, so each bar's length is the total
covariance. With negative contributions, ggplot would place positive and
negative pieces on opposite sides of zero. The waterfall above might make
their net total easier to see.

```{r decomposition-stacked-bars, dev="svg", fig.width=10, fig.height=3}
ggplot(plot_data, aes(x = estimate, y = model, fill = route)) +
  geom_col(width = 0.5, position = position_stack(reverse = TRUE)) +
  # Put labels for tiny segments below the bar to avoid overlap.
  geom_text(
    aes(x = (start + end) / 2,
        y = as.numeric(model) + ifelse(abs(percent) < 1, -0.35, 0.35),
        label = sprintf("%.1f%%", percent)),
    colour = "#475569", size = 3.5
  ) +
  scale_fill_manual(values = route_colours) +
  labs(x = "Outcome covariance", y = NULL, fill = "Contribution") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3. Add confidence intervals for glmmTMB

### Coefficients

`broom.mixed::tidy()` gives Wald CIs for the intercepts and slopes. We join
the readable labels using the original term names, then compare with
lavaan's bootstrap CIs from part 1.

```{r extract-glmmtmb}
fixed_parameters <- broom.mixed::tidy(
  apim_mlm_fit,
  effects = "fixed",
  conf.int = TRUE,
  conf.method = "Wald"
) |>
  dplyr::inner_join(
    dplyr::select(glmmtmb_points, term, parameter),
    by = "term"
  ) |>
  dplyr::select(
    parameter, estimate,
    ci_low = conf.low,
    ci_high = conf.high
  )
```

```{r compare-parameters}
parameter_comparison <- dplyr::inner_join(
  lavaan_params,
  fixed_parameters,
  by = "parameter",
  suffix = c("_lavaan", "_glmmtmb")
) |>
  dplyr::mutate(
    ci_lavaan = sprintf("[%.3f, %.3f]", ci_low_lavaan, ci_high_lavaan),
    ci_glmmtmb = sprintf("[%.3f, %.3f]", ci_low_glmmtmb, ci_high_glmmtmb)
  ) |>
  dplyr::select(
    parameter, estimate_lavaan, ci_lavaan, estimate_glmmtmb, ci_glmmtmb
  )
print_table(parameter_comparison)
```

glmmTMB's Wald CIs are estimate $\pm 1.96\times\mathrm{SE}$. They are close
to lavaan's bootstrap CIs.

### Covariance contributions: bootstrap whole dyads

We bootstrap as lavaan did in part 1: draw 1486 dyads with replacement,
repeat part 2 on this sample, and do this 1999 times. The middle 95% of the
1999 results gives the 95% CI. It does not assume normal data, and it can be
asymmetric, which suits products of estimates.

`boot()` needs part 2 as a function of the data. `decompose_covariance()`
repeats its steps for any data with one row per dyad. It returns the five
contributions, the explained subtotal and the total.

```{r glmmtmb-bootstrap-function}
decompose_covariance <- function(data) {
  # Reshape and prepare as in part 2. Numbering the dyads anew makes a dyad
  # drawn twice count as two dyads.
  long <- data |>
    dplyr::mutate(dyad_id = dplyr::row_number()) |>
    tidyr::pivot_longer(
      cols = -dyad_id,
      names_to = c(".value", "role"),
      names_pattern = "(.+)_(parent|teen)$"
    ) |>
    dplyr::mutate(person_id = dplyr::row_number())
  prepared <- dyadMLM::prepare_dyad_data(
    long, dyad = dyad_id, member = person_id, role = role,
    predictors = efficacy_centered, model_types = "apim"
  )

  # Refit the part 2 model and take the same estimates.
  fit <- update(apim_mlm_fit, data = prepared)
  slopes <- glmmTMB::fixef(fit)$cond
  actor_parent <- slopes[[".is_parent:.efficacy_centered_actor"]]
  actor_teen <- slopes[[".is_teen:.efficacy_centered_actor"]]
  partner_teen_to_parent <- slopes[[".is_parent:.efficacy_centered_partner"]]
  partner_parent_to_teen <- slopes[[".is_teen:.efficacy_centered_partner"]]
  residual_cov <- glmmTMB::VarCorr(fit)$cond$dyad_id[".is_parent", ".is_teen"]

  # Predictor variances and covariance, dividing by n.
  S <- cov.wt(
    data[c("efficacy_centered_parent", "efficacy_centered_teen")],
    method = "ML"
  )$cov
  var_efficacy_parent <- S[1, 1]
  var_efficacy_teen <- S[2, 2]
  cov_efficacy <- S[1, 2]

  # The same formulas as in parts 1 and 2.
  contributions <- c(
    actor_driven = actor_teen * cov_efficacy * actor_parent,
    partner_driven = partner_parent_to_teen * cov_efficacy * partner_teen_to_parent,
    teen_driven = actor_teen * var_efficacy_teen * partner_teen_to_parent,
    parent_driven = actor_parent * var_efficacy_parent * partner_parent_to_teen,
    residual = residual_cov
  )
  c(
    contributions,
    explained = sum(contributions[1:4]),
    total_covariance = sum(contributions)
  )
}
```

```{r glmmtmb-bootstrap, cache = TRUE, cache.extra = digest::digest(list(logLik(apim_mlm_fit), model.frame(apim_mlm_fit), body(decompose_covariance)))}
# About 5 to 15 minutes on 4 cores. boot() draws all samples before fitting,
# so the seed fixes them, also when fitting in parallel.
# Cached until this chunk, the model fit or decompose_covariance() change.
set.seed(2026)
boot_decomposition <- boot::boot(
  flashe_data,
  function(data, rows) decompose_covariance(data[rows, ]),
  R = 1999, parallel = "multicore", ncpus = 4
)

# The 2.5% and 97.5% quantiles of the 1999 estimates give each 95% CI.
glmmtmb_decomposition <- tibble::tibble(
  parameter = names(boot_decomposition$t0),
  estimate = boot_decomposition$t0,
  ci_low = apply(boot_decomposition$t, 2, quantile, 0.025),
  ci_high = apply(boot_decomposition$t, 2, quantile, 0.975)
)
```

### Compare the CIs with lavaan

```{r compare-decompositions}
# Both bootstraps, one row per model and quantity.
decomposition_cis <- dplyr::bind_rows(
  lavaan = lavaan_decomposition, glmmTMB = glmmtmb_decomposition,
  .id = "model"
) |>
  dplyr::mutate(model = factor(model, levels = c("lavaan", "glmmTMB")))

decomposition_comparison <- decomposition_cis |>
  dplyr::filter(parameter %in% c(names(route_labels), "explained")) |>
  dplyr::mutate(ci = sprintf("[%.3f, %.3f]", ci_low, ci_high)) |>
  tidyr::pivot_wider(
    id_cols = parameter, names_from = model, values_from = ci
  )
print_table(decomposition_comparison)
```

The two bootstraps draw different resamples, so their CI limits differ
slightly by chance. Within this resampling noise, they agree.
`explore-sem-mlm-details.Rmd` refits lavaan to the same resamples, where the
contributions agree in every resample. It also compares other CI methods,
including a faster alternative for glmmTMB: a Monte Carlo CI with
`vcov(apim_mlm_fit, full = TRUE, sandwich = TRUE)`.

## 4. Plot the confidence intervals

Each panel shows one contribution on its own scale, with the 95% CIs from
both bootstraps.

```{r decomposition-intervals, dev="svg", fig.width=8, fig.height=5}
ci_plot_data <- decomposition_cis |>
  dplyr::filter(parameter %in% names(route_labels)) |>
  dplyr::mutate(
    route = factor(parameter, levels = names(route_labels),
                   labels = unname(route_labels))
  )

ggplot(ci_plot_data, aes(estimate, model, colour = route)) +
  geom_pointrange(aes(xmin = ci_low, xmax = ci_high),
                 show.legend = FALSE) +
  facet_wrap(~ route, ncol = 1, scales = "free_x") +
  scale_colour_manual(values = route_colours) +
  labs(x = "Covariance contribution", y = NULL) +
  theme_minimal() +
  # Room for the last tick label on the right.
  theme(plot.margin = margin(5.5, 15, 5.5, 5.5))
```
