---
title: "APIM exploration: checks and CI methods"
author: "Pascal Küng"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    math_method: mathml
    df_print: kable
    css: explore-sem-mlm.css
---

This notebook adds checks and other CI methods to `explore-sem-mlm.Rmd`. It
first knits that notebook in the background, so its data, fits and bootstrap
results are available here.

```{r run-main-notebook, include=FALSE}
# Use the main notebook's cache and figure folders, so knitr reuses its
# cached bootstrap. Without that cache, this takes 5 to 15 minutes.
invisible(knitr::knit_child(
  "explore-sem-mlm.Rmd",
  options = list(
    cache.path = "explore-sem-mlm_cache/html/",
    fig.path = "explore-sem-mlm_files/figure-html/"
  ),
  quiet = TRUE
))
```

## 1. glmmTMB and lavaan on the same samples

In the main notebook, the two bootstraps draw different samples, so they
agree only within resampling noise. Here we refit lavaan to the same 1999
samples that glmmTMB used (`bootstrap$dyads`). If the glmmTMB bootstrap is
correct, both models give the same estimates in every sample, up to the
precision of the optimizers.

```{r same-samples}
# Refit lavaan to any data; return all estimates as a one-row tibble.
# as.list() drops lavaan's own vector class.
estimate_lavaan <- function(data) {
  fit <- update(apim_fit, data = data, se = "none")
  tibble::as_tibble_row(as.list(coef(fit, type = "user")))
}

# One row per sample, one column per estimate, in the same order as
# glmmtmb_draws.
lavaan_draws <- parallel::mclapply(
  bootstrap$dyads, \(rows) estimate_lavaan(flashe_data[rows, ]), mc.cores = 4
) |>
  dplyr::bind_rows() |>
  dplyr::select(dplyr::all_of(names(glmmtmb_draws)))

# Largest difference between glmmTMB and lavaan, over all samples and
# estimates.
max(abs(glmmtmb_draws - lavaan_draws))
```

## 2. Other CI methods for the contributions

The main notebook uses the bootstrap. We compare it with three alternatives:

- **lavaan, delta method:** lavaan's default. It assumes normally distributed
  data and gives symmetric CIs.
- **lavaan, delta method with MLR:** the same, but with robust SEs that do not
  assume normal data.
- **glmmTMB, Monte Carlo:** draws many sets of estimates from their estimated
  sampling distribution and applies the formulas to each set (Preacher &
  Selig, 2012). `sandwich = TRUE` makes the uncertainty of the model
  estimates robust, clustered by dyad. The predictor variances and covariance
  are drawn assuming normal data.

```{r delta-method}
# Without se = "bootstrap", lavaan uses the delta method.
delta_fit <- sem(
  apim_model, fixed.x = FALSE, data = flashe_data, missing = "fiml"
)
mlr_fit <- sem(
  apim_model, fixed.x = FALSE, data = flashe_data, missing = "fiml",
  estimator = "MLR"
)
```

```{r monte-carlo}
# About 3 minutes, mostly for the sandwich uncertainty matrix.
set.seed(2026)
n_draws <- 20000
n_dyads <- nrow(flashe_data)

# Draw the coefficients and glmmTMB's internal variance parameters (theta)
# together: two log SDs and a transformed correlation.
theta <- glmmTMB::getME(apim_mlm_fit, "theta")
model_draws <- MASS::mvrnorm(
  n_draws,
  mu = c(
    glmmTMB::fixef(apim_mlm_fit)$cond,
    log_sd_parent = theta[1], log_sd_teen = theta[2], cor_theta = theta[3]
  ),
  Sigma = vcov(apim_mlm_fit, full = TRUE, sandwich = TRUE)
)

# Draw the predictor variances and covariance (normal theory). rWishart()
# returns a 2 x 2 x n_draws array: one covariance matrix per draw.
predictor_draws <- rWishart(
  n_draws, df = n_dyads, Sigma = predictor_covariance_matrix / n_dyads
)

# One row per draw, named as in estimate_apim(), then the same formulas.
monte_carlo_draws <- tibble::as_tibble(model_draws) |>
  dplyr::transmute(
    actor_parent = `.is_parent:.efficacy_centered_actor`,
    actor_teen = `.is_teen:.efficacy_centered_actor`,
    partner_teen_to_parent = `.is_parent:.efficacy_centered_partner`,
    partner_parent_to_teen = `.is_teen:.efficacy_centered_partner`,
    # Back-transform theta: SD parent * SD teen * correlation.
    residual_cov = exp(log_sd_parent) * exp(log_sd_teen) *
      cor_theta / sqrt(1 + cor_theta^2),
    var_efficacy_parent = predictor_draws[1, 1, ],
    var_efficacy_teen = predictor_draws[2, 2, ],
    cov_efficacy = predictor_draws[1, 2, ]
  ) |>
  add_contributions()

monte_carlo_cis <- percentile_cis(monte_carlo_draws)
```

```{r compare-ci-methods}
lavaan_cis <- function(fit) {
  broom::tidy(fit, conf.int = TRUE) |>
    dplyr::select(parameter = label, ci_low = conf.low, ci_high = conf.high)
}

ci_methods <- dplyr::bind_rows(
  `glmmTMB bootstrap` = glmmtmb_cis,
  `lavaan delta` = lavaan_cis(delta_fit),
  `lavaan delta, MLR` = lavaan_cis(mlr_fit),
  `glmmTMB Monte Carlo` = monte_carlo_cis,
  .id = "method"
) |>
  dplyr::filter(parameter %in% c(names(route_labels), "explained")) |>
  dplyr::mutate(ci = sprintf("[%.3f, %.3f]", ci_low, ci_high)) |>
  tidyr::pivot_wider(
    id_cols = parameter, names_from = method, values_from = ci
  )
print_table(ci_methods)
```

The bootstrap, the robust delta method and the Monte Carlo CIs are close.
The normal-theory delta method differs most for the residual covariance,
because the intake scores are not normally distributed.
