---
title: "APIM exploration: lavaan and glmmTMB"
author: "Pascal Küng"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    math_method: mathml
    df_print: kable
    css: explore-sem-mlm.css
---

This notebook fits the same cross-sectional Gaussian APIM in lavaan and
glmmTMB. We first compare and plot the point estimates, then add confidence
intervals for glmmTMB and plot them. Each dyad has one parent and one
teen, with complete data.

```{r table-style, include=FALSE}
# Each table includes its CSS when chunks are run separately.
table_css <- paste(readLines("explore-sem-mlm.css"), collapse = "\n")

print_table <- function(data) {
  htmltools::browsable(htmltools::tagList(
    htmltools::tags$style(htmltools::HTML(table_css)),
    htmltools::HTML(knitr::kable(
      data, format = "html", table.attr = 'class="apim-table"'
    ))
  ))
}
```

## Load and rename the data

```{r load-flashe}
library(lavaan)
flashe_data <- read.csv(
  file.path(
    "../references/explaining-interdependence-apim",
    "jp-materials-2026-09-13",
    "J-P&Niall-FLASHE-example",
    "flashe-lavaan.csv"
  ),
  header = TRUE
)

# Produce intake and efficacy refer to fruit and vegetable consumption.
# Keep the supplied scores centered at 7.6.
flashe_data <- dplyr::rename(
  flashe_data,
  produce_intake_parent = pftveg,
  produce_intake_teen = tftveg,
  efficacy_parent = peffv,
  efficacy_teen = teffv,
  efficacy_centered_parent = cpeffv,
  efficacy_centered_teen = cteffv,
  is_female_parent = pfem,
  is_female_teen = tfem
)
```

## 1. Lavaan: J-P and Niall's example

### Fit the model

```{r jp-niall-lavaan}
apim_model <- '
  # Intake intercepts: both centered efficacy predictors are zero.
  produce_intake_teen ~ intercept_teen * 1
  produce_intake_parent ~ intercept_parent * 1
  
  # Actor effects: own centered efficacy predicts own F&V intake.
  produce_intake_teen ~ actor_teen * efficacy_centered_teen
  produce_intake_parent ~ actor_parent * efficacy_centered_parent

  # Partner effects: parent efficacy -> teen intake, and vice versa.
  produce_intake_teen ~ partner_parent_to_teen * efficacy_centered_parent
  produce_intake_parent ~ partner_teen_to_parent * efficacy_centered_teen

  # Means of the centered efficacy predictors.
  efficacy_centered_teen ~ mean_efficacy_teen * 1
  efficacy_centered_parent ~ mean_efficacy_parent * 1

  # Predictor variances.
  efficacy_centered_teen ~~ var_efficacy_teen * efficacy_centered_teen
  efficacy_centered_parent ~~ var_efficacy_parent * efficacy_centered_parent

  # Residual variances of intake.
  produce_intake_teen ~~ residual_var_teen * produce_intake_teen
  produce_intake_parent ~~ residual_var_parent * produce_intake_parent

  # Covariance between parent and teen efficacy.
  efficacy_centered_parent ~~ cov_efficacy * efficacy_centered_teen

  # Residual covariance between parent and teen intake.
  produce_intake_parent ~~ residual_cov * produce_intake_teen

  # Path-traced contributions to the outcome covariance.
  actor_driven := actor_teen * cov_efficacy * actor_parent
  partner_driven := partner_parent_to_teen * cov_efficacy * partner_teen_to_parent
  teen_driven := actor_teen * var_efficacy_teen * partner_teen_to_parent
  parent_driven := actor_parent * var_efficacy_parent * partner_parent_to_teen
  residual := residual_cov
  explained := actor_driven + partner_driven + teen_driven + parent_driven
  total_covariance := explained + residual

  # Percentages of the total outcome covariance.
  percent_actor_driven := 100 * (actor_driven / total_covariance)
  percent_partner_driven := 100 * (partner_driven / total_covariance)
  percent_teen_driven := 100 * (teen_driven / total_covariance)
  percent_parent_driven := 100 * (parent_driven / total_covariance)
  percent_residual := 100 * (residual / total_covariance)
  percent_explained := 100 * (explained / total_covariance)
'

# Predictor means, variances and covariance are free parameters above.
apim_fit <- sem(
  apim_model, fixed.x = FALSE, data = flashe_data, missing = "fiml"
)

# Examine the model.
# summary(apim_fit, fit.measures = TRUE)
```

### Extract the model parameters

Lavaan provides both the fitted parameters and the quantities defined with
`:=`. We extract them once, then display the model parameters and decomposition
separately. The model parameters include the predictor means, variances and
covariance as well as the outcome coefficients and residual structure.

```{r extract-lavaan}
# Extract once, using our model labels and including 95% CIs.
lavaan_estimates <- broom::tidy(apim_fit, conf.int = TRUE) |>
  dplyr::select(
    parameter = label,
    op,
    estimate,
    ci_low = conf.low,
    ci_high = conf.high
  )

# Model parameters: exclude the quantities defined with :=.
lavaan_params <- lavaan_estimates |>
  dplyr::filter(op != ":=") |>
  dplyr::select(-op)

print_table(lavaan_params)
```

### Extract the covariance decomposition

The `:=` definitions above apply the Day 2 handout formulas. They calculate
contributions and percentages from the estimates without changing the fitted
model. Lavaan uses the delta method to calculate their standard errors (SEs)
and 95% confidence intervals (CIs). It accounts for uncertainty in all the
estimates used in each formula, including how those estimates vary together.
Because we estimate the predictor variances and covariance in the model,
their uncertainty is included too. 

We select these estimates and CIs from the table above:

```{r lavaan-covariance-decomposition}
# Decomposition: keep only the quantities defined with :=.
lavaan_decomposition <- lavaan_estimates |>
  dplyr::filter(op == ":=") |>
  dplyr::select(-op)

# Display the five contributions and the explained subtotal, with CI and percent.
format_decomposition <- function(decomposition) {
  # Match each percentage row to its contribution by removing "percent_".
  percentages <- decomposition |>
    dplyr::filter(startsWith(parameter, "percent_")) |>
    dplyr::mutate(parameter = sub("^percent_", "", parameter)) |>
    dplyr::select(parameter, percent = estimate)

  # Format only the displayed CI; estimates keep their full precision.
  decomposition |>
    dplyr::filter(!startsWith(parameter, "percent_"),
                  parameter != "total_covariance") |>
    dplyr::mutate(
      `95% CI` = sprintf("[%.3f, %.3f]", ci_low, ci_high)
    ) |>
    dplyr::select(parameter, estimate, `95% CI`) |>
    dplyr::left_join(percentages, by = "parameter")
}

lavaan_decomposition_table <- format_decomposition(lavaan_decomposition)
print_table(lavaan_decomposition_table)
```


## 2. glmmTMB: reproduce the point estimates

`dyadMLM` prepares the actor and partner predictors; `glmmTMB` fits the model.
We retain the supplied centering and use maximum likelihood, as in lavaan.

### Reshape and prepare the data

Reshape to one row per person, then construct the actor/partner predictors.

```{r prepare-dyadmlm}
# Use dyadMLM from this checkout.
library(dyadMLM)

# Each original row is one dyad. Long format has one row per person.
flashe_long <- flashe_data |>
  dplyr::mutate(dyad_id = dplyr::row_number()) |>
  tidyr::pivot_longer(
    cols = -dyad_id,
    names_to = c(".value", "role"),
    names_pattern = "(.+)_(parent|teen)$"
  ) |>
  dplyr::mutate(person_id = dplyr::row_number())

# Construct actor/partner columns without centering the scores again.
flashe_apim <- dyadMLM::prepare_dyad_data(
  flashe_long,
  dyad = dyad_id,
  member = person_id,
  role = role,
  predictors = efficacy_centered,
  model_types = "apim"
)

print(flashe_apim, n = 4)
```

### Fit the model

```{r fit-dyadmlm}
apim_mlm_fit <- glmmTMB::glmmTMB(
  produce_intake ~
    # Separate parent and teen intercepts.
    0 + .is_parent + .is_teen +

    # Actor effects: own efficacy predicts own intake.
    .is_parent:.efficacy_centered_actor +
    .is_teen:.efficacy_centered_actor +

    # Partner effects: partner efficacy predicts own intake.
    .is_parent:.efficacy_centered_partner +
    .is_teen:.efficacy_centered_partner +

    # Two residual variances and their covariance within each dyad.
    us(0 + .is_parent + .is_teen | dyad_id),

  # Represent the paired residual variation in the unstructured block.
  dispformula = ~ 0,
  family = gaussian(),
  data = flashe_apim,

  # Match lavaan's maximum-likelihood estimation.
  REML = FALSE
)

# summary(apim_mlm_fit)
```

### Extract the standard point estimates

We extract the needed coefficients, then add labels matching lavaan. 

```{r extract-glmmtmb-points}
# Extract all nine estimates with their original model names.
glmmtmb_points <- broom.mixed::tidy(
  apim_mlm_fit,
  effects = c("fixed", "ran_pars"),
  scales = c(NA, "vcov") # leaves fixed effects unchanged and requests variances/covariances for REs
) |>
  dplyr::select(effect, term, estimate)

# Rename the rows to match lavaan. Keep term for later model lookups.
glmmtmb_points <- glmmtmb_points |>
  dplyr::mutate(
    parameter = dplyr::recode(
      term,
      ".is_parent" = "intercept_parent",
      ".is_teen" = "intercept_teen",
      ".is_parent:.efficacy_centered_actor" = "actor_parent",
      ".is_teen:.efficacy_centered_actor" = "actor_teen",
      ".is_parent:.efficacy_centered_partner" = "partner_teen_to_parent",
      ".is_teen:.efficacy_centered_partner" = "partner_parent_to_teen",
      "var__.is_parent" = "residual_var_parent",
      "var__.is_teen" = "residual_var_teen",
      "cov__.is_parent..is_teen" = "residual_cov"
    ),
    # Include the tiny fixed dispersion in the two residual variances.
    estimate = estimate + dplyr::if_else(
      parameter %in% c("residual_var_parent", "residual_var_teen"),
      sigma(apim_mlm_fit)^2, 0 # Sigma returns an SD, so square it. 
    )
  )

glmmtmb_params_point <- dplyr::select(glmmtmb_points, parameter, estimate)
print_table(glmmtmb_params_point)
```

### Calculate the contributions with simple products

We also need the two efficacy variances and their covariance. We calculate
these from one row per dyad, dividing by $n$ as in lavaan's ML estimation.

We then collect all estimates we need for the computation
in a one-row table, `mlm_inputs`.

```{r}

n_dyads <- nrow(flashe_data)

predictor_covariance_matrix <- cov.wt(
  flashe_data[c("efficacy_centered_parent", "efficacy_centered_teen")],
  method = "ML"
)$cov


# Put each fitted estimate in its own column.
model_estimates <- glmmtmb_params_point |>
  tidyr::pivot_wider(names_from = parameter, values_from = estimate)

# Keep the four slopes and residual covariance; add the predictor moments.
mlm_inputs <- model_estimates |>
  dplyr::select(
    actor_parent,
    actor_teen,
    partner_parent_to_teen,
    partner_teen_to_parent,
    residual_cov
  ) |>
  dplyr::mutate(
    var_efficacy_parent = predictor_covariance_matrix[1, 1],
    var_efficacy_teen = predictor_covariance_matrix[2, 2],
    cov_efficacy = predictor_covariance_matrix[1, 2],
    .before = residual_cov
  )

print_table(mlm_inputs)
```


Computing the contributions with the columns directly: 

```{r glmmtmb-point-decomposition}
glmmtmb_decomposition_point <- mlm_inputs |>
  dplyr::mutate(
    actor_driven = actor_teen * cov_efficacy * actor_parent,
    partner_driven = partner_parent_to_teen * cov_efficacy * partner_teen_to_parent,
    teen_driven = actor_teen * var_efficacy_teen * partner_teen_to_parent,
    parent_driven = actor_parent * var_efficacy_parent * partner_parent_to_teen,
    residual = residual_cov,
    
    explained = sum(
      actor_driven, partner_driven, teen_driven, parent_driven
    ),
    
    total_covariance = explained + residual,
    
    .keep = "none"
  ) |>
  # Return to one row per contribution for the display table.
  tidyr::pivot_longer(
    dplyr::everything(), names_to = "parameter", values_to = "estimate"
  ) |>
  # Divide each estimate by the total covariance row, then multiply by 100.
  dplyr::mutate(
    percent = 100 * (estimate / estimate[parameter == "total_covariance"])
  )
print_table(glmmtmb_decomposition_point)
```

### Compare the point estimates with lavaan

We compare the nine parameters estimated by both models, then the covariance
contributions. The predictor means, variances and covariance are fitted
parameters in lavaan. glmmTMB treats the predictors as observed values.

```{r compare-parameter-points}
parameter_point_comparison <- dplyr::inner_join(
  dplyr::select(lavaan_params, parameter, estimate),
  glmmtmb_params_point,
  by = "parameter",
  suffix = c("_lavaan", "_glmmtmb")
)
print_table(parameter_point_comparison)
```

```{r compare-decomposition-points}
decomposition_point_comparison <- dplyr::inner_join(
  dplyr::select(lavaan_decomposition_table, parameter, estimate, percent),
  glmmtmb_decomposition_point,
  by = "parameter",
  suffix = c("_lavaan", "_glmmtmb")
)
print_table(decomposition_point_comparison)
```

### Plot the point estimates

Visualize the contribution to the total covariance.

```{r prepare-decomposition-plots}
library(ggplot2)

route_labels <- c(
  actor_driven = "Actor-driven", partner_driven = "Partner-driven",
  teen_driven = "Teen-driven", parent_driven = "Parent-driven",
  residual = "Residual"
)
# Muted teal, ochre, blue, mauve and slate, in the same order as the routes.
route_colours <- c("#468C83", "#D3A064", "#6F91B7", "#AD8CA7", "#7B8794")
plot_data <- dplyr::bind_rows(
  lavaan = dplyr::select(lavaan_decomposition, parameter, estimate),
  glmmTMB = dplyr::select(glmmtmb_decomposition_point, parameter, estimate),
  .id = "model"
) |>
  dplyr::filter(parameter %in% names(route_labels)) |>
  dplyr::mutate(
    model = factor(model, levels = c("lavaan", "glmmTMB")),
    route = factor(parameter, levels = names(route_labels),
                   labels = unname(route_labels))
  ) |>
  dplyr::arrange(model, route) |>
  dplyr::group_by(model) |>
  dplyr::mutate(
    end = cumsum(estimate), start = end - estimate,
    percent = 100 * estimate / sum(estimate)
  )

plot_totals <- dplyr::summarise(plot_data, total = sum(estimate))
```

#### Waterfall: how the contributions add up

```{r decomposition-waterfall, dev="svg", fig.width=10, fig.height=4}
ggplot(plot_data, aes(fill = route)) +
  geom_rect(aes(
    xmin = as.numeric(route) - 0.4, xmax = as.numeric(route) + 0.4,
    ymin = pmin(start, end), ymax = pmax(start, end)
  )) +
  geom_text(
    aes(x = as.numeric(route), y = pmax(start, end),
        label = sprintf("%.1f%%", percent)),
    vjust = -0.6, colour = "#475569", size = 3.5
  ) +
  geom_hline(
    data = plot_totals, aes(yintercept = total),
    linetype = "dotted", colour = "#475569"
  ) +
  geom_text(
    data = plot_totals,
    aes(x = 0.6, y = total, label = sprintf("Total = %.3f", total)),
    inherit.aes = FALSE, hjust = 0, vjust = -0.6, colour = "#475569"
  ) +
  facet_wrap(~ model, nrow = 1) +
  scale_x_continuous(
    breaks = seq_along(route_labels), labels = gsub("-", "\n", route_labels)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  scale_fill_manual(values = route_colours) +
  labs(x = NULL, y = "Outcome covariance") +
  theme_minimal() +
  theme(legend.position = "none")
```

#### Stacked bars: compare the two fits

Here all five contributions are positive, so each bar's length is the total
covariance. The two bars should look almost identical because the fits describe
the same model. With negative contributions, ggplot places positive and
negative pieces on opposite sides of zero; the waterfall makes their net
total easier to see.

```{r decomposition-stacked-bars, dev="svg", fig.width=10, fig.height=3}
ggplot(plot_data, aes(x = estimate, y = model, fill = route)) +
  geom_col(width = 0.5, position = position_stack(reverse = TRUE)) +
  # Put labels for tiny segments below the bar to avoid overlap.
  geom_text(
    aes(x = (start + end) / 2,
        y = as.numeric(model) + ifelse(abs(percent) < 1, -0.35, 0.35),
        label = sprintf("%.1f%%", percent)),
    colour = "#475569", size = 3.5
  ) +
  scale_fill_manual(values = route_colours) +
  labs(x = "Outcome covariance", y = NULL, fill = "Contribution") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3. Add confidence intervals for glmmTMB

### Add CIs for the standard parameters

We now add CIs to the point estimates above. `broom.mixed::tidy()` provides
the coefficient Wald CIs directly, in the same table format as before.
For the residual structure, `confint()` returns SDs and a correlation. We want
variances and a covariance, with symmetric Wald CIs like lavaan's. We use
`car::deltaMethod()` to transform glmmTMB's parameters and their joint
uncertainty; we do not need to derive the standard errors ourselves.

The helper `delta_table()` repeats this call for each formula and collects
the estimates, SEs and CIs into one table. We retain the SEs for the later
decomposition, but display only the estimates and CIs.

```{r glmmtmb-delta-helper}
# formulas is a named vector of expressions written as text.
# uncertainty is the matrix describing uncertainty in the input estimates.
# car differentiates each expression and calculates its SE and Wald CI.
delta_table <- function(estimates, uncertainty, formulas) {
  lapply(formulas, function(formula) {
    car::deltaMethod(estimates, formula, vcov. = uncertainty) |>
      tibble::as_tibble()
  }) |>
    dplyr::bind_rows(.id = "parameter") |>
    dplyr::select(
      parameter,
      estimate = Estimate,
      standard_error = SE,
      ci_low = `2.5 %`,
      ci_high = `97.5 %`
    )
}
```

We extract the CIs, then join the readable labels using the original term
names. Selecting the columns below keeps only the estimate and CI limits.

```{r extract-glmmtmb}
fixed_parameters <- broom.mixed::tidy(
  apim_mlm_fit,
  effects = "fixed",
  conf.int = TRUE,
  conf.method = "Wald"
) |>
  dplyr::inner_join(
    dplyr::select(glmmtmb_points, term, parameter),
    by = "term"
  ) |>
  dplyr::select(
    parameter, estimate,
    ci_low = conf.low,
    ci_high = conf.high
  )
```

For this two-member `us()` block, glmmTMB stores three internal parameters,
called `theta`: the log of each SD and a transformed correlation. We give them
readable names below. `cor_theta` is not the correlation itself; the formula
`cor_theta / sqrt(1 + cor_theta^2)` converts it to a correlation.

`vcov(..., full = TRUE)` provides the estimation uncertainty for all fitted
parameters. We select the three `theta` rows and columns to obtain their
joint uncertainty. This matrix describes how the parameter estimates vary
across samples, rather than how the parent and teen residuals vary.

The formulas then convert log-SDs to variances and multiply both SDs by the
correlation to obtain the covariance. `delta_table()` passes these formulas
and the uncertainty matrix to `car::deltaMethod()`.

```{r extract-glmmtmb-residuals}
# Internal parameters, in parent/teen order for the us() block above.
theta <- glmmTMB::getME(apim_mlm_fit, "theta")
names(theta) <- c("log_sd_parent", "log_sd_teen", "cor_theta")

# Keep their joint estimation uncertainty from the fitted model.
theta_vcov <- vcov(apim_mlm_fit, full = TRUE, include_nonest = FALSE)
theta_rows <- grep("^theta_", rownames(theta_vcov))
theta_vcov <- theta_vcov[theta_rows, theta_rows]
dimnames(theta_vcov) <- list(names(theta), names(theta))

residual_formulas <- c(
  residual_var_parent = "exp(2 * log_sd_parent)",
  residual_var_teen = "exp(2 * log_sd_teen)",
  residual_cov = paste(
    "exp(log_sd_parent + log_sd_teen) *",
    "cor_theta / sqrt(1 + cor_theta^2)"
  )
)
residual_parameters <- delta_table(
  theta, theta_vcov, residual_formulas
)

# ~0 fixes an additional, almost-zero dispersion variance.
# Add that constant to both variance estimates and their CI limits.
# across() applies the same adjustment to each of these three columns.
residual_parameters <- residual_parameters |>
  dplyr::mutate(dplyr::across(
    c(estimate, ci_low, ci_high),
    ~ .x + dplyr::if_else(
      parameter != "residual_cov", sigma(apim_mlm_fit)^2, 0
    )
  ))

glmmtmb_params <- dplyr::bind_rows(fixed_parameters, residual_parameters) |>
  dplyr::select(-standard_error)
print_table(glmmtmb_params)
```

### Add CIs for the covariance contributions

We keep the eight point estimates already collected in `mlm_inputs`.
`car::deltaMethod()` needs a named numeric vector, so `unlist()` converts the
one-row table while retaining the column names. We then add their joint
uncertainty.

```{r glmmtmb-ci-inputs}
mlm_estimates <- unlist(mlm_inputs)
```

Lavaan does this automatically for `:=` formulas because our model estimates
the predictor variances and covariance as well as the outcome parameters.
GlmmTMB fits the outcomes given the observed predictors. It provides uncertainty
for its fitted parameters, but does not estimate the predictor distribution
or automatically calculate CIs for our decomposition. We add the predictor
variances and covariance and their uncertainty to match the lavaan calculation.

### Prepare the uncertainty matrix

A contribution combines several estimates, so its CI needs uncertainty in
every input. `vcov(apim_mlm_fit)$cond` gives the uncertainty matrix for the
fixed-effect estimates: its diagonal contains squared SEs, and its other
entries describe how the coefficient estimates vary together across samples.
We obtain the decomposition's inputs from these sources:

| Inputs | Where the estimates come from | Where their uncertainty comes from |
| --- | --- | --- |
| Actor/partner slopes | `broom.mixed::tidy()` | `vcov(apim_mlm_fit)$cond` |
| Residual covariance | `broom.mixed::tidy()` | Its SE from the model's `theta_vcov` and the delta method |
| Predictor variances and covariance | `cov.wt(..., method = "ML")` | `covariance_uncertainty()` |

`VarCorr()` describes variation in parent and teen residuals. `vcov()`
describes uncertainty in the estimated model parameters. These are different
quantities, even though both are covariance matrices.

`mlm_vcov` combines the uncertainty in those eight estimates. It must use
the same order as `mlm_estimates`, so each estimate is paired with its own
uncertainty.

We still need a separate calculation for the predictor variances and
covariance, because glmmTMB does not estimate them. `covariance_uncertainty()`
takes their covariance matrix and the number of dyads, then returns the
uncertainty matrix for the two variances and covariance using Gaussian ML
formulas.
This helper assumes complete, independent Gaussian dyads; it cannot be used
unchanged for missing data or repeated observations within dyads.

```{r glmmtmb-predictor-uncertainty}
# S is in parent/teen order; n is the number of complete dyads.
# Rows and columns: parent variance, teen variance, predictor covariance.
covariance_uncertainty <- function(S, n) {
  v_parent <- S[1, 1]
  v_teen <- S[2, 2]
  c_parent_teen <- S[1, 2]
  rbind(
    c(2 * v_parent^2, 2 * c_parent_teen^2, 2 * v_parent * c_parent_teen),
    c(2 * c_parent_teen^2, 2 * v_teen^2, 2 * v_teen * c_parent_teen),
    c(2 * v_parent * c_parent_teen, 2 * v_teen * c_parent_teen,
      v_parent * v_teen + c_parent_teen^2)
  ) / n
}
```

The three uncertainty blocks are asymptotically uncorrelated under this
complete Gaussian ML model. `bdiag()` therefore fills the gaps between them
with zeros, while keeping the covariances within each block.

```{r glmmtmb-decomposition-inputs}
# Join the first four input labels to the original model names, in that order.
slope_names <- tibble::tibble(parameter = names(mlm_inputs)[1:4]) |>
  dplyr::left_join(
    dplyr::select(glmmtmb_points, parameter, term),
    by = "parameter"
  ) |>
  dplyr::pull(term)
slope_vcov <- vcov(apim_mlm_fit)$cond[slope_names, slope_names]

# Uncertainty in the predictor variances and covariance calculated in part 2.
predictor_vcov <- covariance_uncertainty(predictor_covariance_matrix, n_dyads)
residual_cov_parameter <- dplyr::filter(
  residual_parameters, parameter == "residual_cov"
)

# Join the three uncertainty blocks; the last is the residual covariance's SE^2.
mlm_vcov <- as.matrix(Matrix::bdiag(
  slope_vcov, predictor_vcov, residual_cov_parameter$standard_error^2
))
dimnames(mlm_vcov) <- list(names(mlm_estimates), names(mlm_estimates))
```

The table below shows the eight inputs and their SEs. The full matrix also
contains the covariances between estimates; these enter the delta calculation.

```{r show-decomposition-inputs}
decomposition_inputs <- tibble::tibble(
  parameter = names(mlm_estimates),
  estimate = unname(mlm_estimates),
  standard_error = sqrt(diag(mlm_vcov))
)
print_table(decomposition_inputs)
```

### Calculate the CI for one contribution

Take `actor_driven`. It multiplies the teen actor effect, the predictor
covariance, and the parent actor effect. Once the inputs above are ready,
this single call gives its estimate, SE and 95% Wald CI:

```{r actor-contribution-ci}
actor_driven_ci <- car::deltaMethod(
  mlm_estimates,
  "actor_teen * cov_efficacy * actor_parent",
  vcov. = mlm_vcov
)
print_table(actor_driven_ci)
```

The three arguments supply the **values**, the **formula**, and the
**uncertainty matrix**. Inside this call, the delta method:

1. Calculates how much the product changes when each input changes. These
   sensitivities provide the weights for the uncertainty calculation.
2. Combines those weights with the uncertainty matrix to calculate the
   product's SE, including how its input estimates vary together.
3. Forms the interval: estimate $\pm 1.96\times\mathrm{SE}$.

The matrix multiplication is handled by the package. We need to supply the
formula too, because different products have different sensitivities to the
same inputs. For a percentage, the formula also contains the total covariance
in the denominator; the same calculation accounts for its uncertainty.

### Calculate all contributions

We write the same formulas from part 2 as text for `car::deltaMethod()`.
`delta_table()` repeats the call for each contribution and percentage.

```{r glmmtmb-covariance-decomposition}
# The same five path products as in lavaan, written as text for car.
path_formulas <- c(
  actor_driven = "actor_teen * cov_efficacy * actor_parent",
  partner_driven = paste(
    "partner_parent_to_teen * cov_efficacy *", "partner_teen_to_parent"
  ),
  teen_driven = "actor_teen * var_efficacy_teen * partner_teen_to_parent",
  parent_driven = "actor_parent * var_efficacy_parent * partner_parent_to_teen",
  residual = "residual_cov"
)
# Join the five expressions with +; the first four give the explained subtotal.
total_formula <- paste(path_formulas, collapse = " + ")
path_formulas <- c(
  path_formulas, explained = paste(path_formulas[1:4], collapse = " + ")
)

# Each percentage is 100 * (contribution formula) / (total formula).
# sprintf() inserts both formulas; setNames() labels the resulting quantities.
decomposition_formulas <- c(
  path_formulas,
  total_covariance = total_formula,
  setNames(
    sprintf("100 * (%s) / (%s)", path_formulas, total_formula),
    paste0("percent_", names(path_formulas))
  )
)

glmmtmb_decomposition <- delta_table(
  mlm_estimates, mlm_vcov, decomposition_formulas
)
glmmtmb_decomposition_table <- format_decomposition(glmmtmb_decomposition)
print_table(glmmtmb_decomposition_table)
```

### Compare the model parameter CIs

```{r compare-parameters}
parameter_comparison <- dplyr::inner_join(
  lavaan_params,
  glmmtmb_params,
  by = "parameter",
  suffix = c("_lavaan", "_glmmtmb")
) |>
  dplyr::mutate(
    ci_lavaan = sprintf("[%.3f, %.3f]", ci_low_lavaan, ci_high_lavaan),
    ci_glmmtmb = sprintf("[%.3f, %.3f]", ci_low_glmmtmb, ci_high_glmmtmb)
  ) |>
  dplyr::select(
    parameter, estimate_lavaan, ci_lavaan, estimate_glmmtmb, ci_glmmtmb
  )

# Fixed effects: intercepts and actor/partner slopes.
fixed_effects_comparison <- parameter_comparison |>
  dplyr::filter(!startsWith(parameter, "residual_"))
print_table(fixed_effects_comparison)

# Residual structure: two variances and their covariance.
residual_structure_comparison <- parameter_comparison |>
  dplyr::filter(startsWith(parameter, "residual_"))
print_table(residual_structure_comparison)
```

Both tables use unstandardized estimates and 95% Wald confidence intervals.
The glmmTMB variance/covariance intervals transform the fitted model's joint
parameter uncertainty with the delta method. They use the same symmetric
Wald form as lavaan: estimate $\pm 1.96\times\mathrm{SE}$.

### Compare the contribution CIs

Both methods use full-precision estimates and 95% delta-method Wald intervals.
Formatting the intervals below only rounds their displayed values.
The CI refers to the covariance estimate; percent gives its share of the total.

```{r compare-decompositions}
decomposition_comparison <- dplyr::left_join(
  lavaan_decomposition_table,
  glmmtmb_decomposition_table,
  by = "parameter",
  suffix = c("_lavaan", "_glmmtmb")
)

print_table(decomposition_comparison)
```

## 4. Plot the confidence intervals

Each point is a contribution estimate; its horizontal line is the 95% CI.
The panels show lavaan and glmmTMB using the same scale.

```{r decomposition-intervals, dev="svg", fig.width=10, fig.height=4}
# Add the CI limits now that both decompositions include them.
ci_plot_data <- dplyr::bind_rows(
  lavaan = lavaan_decomposition, glmmTMB = glmmtmb_decomposition,
  .id = "model"
) |>
  dplyr::filter(parameter %in% names(route_labels)) |>
  dplyr::mutate(
    model = factor(model, levels = c("lavaan", "glmmTMB")),
    route = factor(parameter, levels = names(route_labels),
                   labels = unname(route_labels))
  )

ggplot(ci_plot_data, aes(estimate, route, colour = route)) +
  geom_pointrange(aes(xmin = ci_low, xmax = ci_high),
                 show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  facet_wrap(~ model) +
  scale_colour_manual(values = route_colours) +
  labs(x = "Covariance contribution", y = NULL) +
  theme_minimal()
```
