---
title: "APIM exploration: lavaan and glmmTMB"
author: "Pascal Küng"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    math_method: mathml
    df_print: kable
    css: explore-sem-mlm.css
---

This notebook shows the full workflow for the same cross-sectional Gaussian
APIM in lavaan and glmmTMB. Both give the same point estimates of the
covariance decomposition, and their bootstrap confidence intervals agree
within resampling noise. Each dyad has one parent and one teen, with complete data.

```{r table-style, include=FALSE}
# Each table includes its CSS when chunks are run separately.
table_css <- paste(readLines("explore-sem-mlm.css"), collapse = "\n")

print_table <- function(data) {
  htmltools::browsable(htmltools::tagList(
    htmltools::tags$style(htmltools::HTML(table_css)),
    htmltools::HTML(knitr::kable(
      data, format = "html", table.attr = 'class="apim-table"'
    ))
  ))
}
```

## Load and rename the data

```{r load-flashe}
library(lavaan)
flashe_data <- read.csv(
  file.path(
    "../references/explaining-interdependence-apim",
    "jp-materials-2026-09-13",
    "J-P&Niall-FLASHE-example",
    "flashe-lavaan.csv"
  ),
  header = TRUE
)

# Produce intake and efficacy refer to fruit and vegetable consumption.
# Keep the supplied scores centered at 7.6.
flashe_data <- dplyr::rename(
  flashe_data,
  produce_intake_parent = pftveg,
  produce_intake_teen = tftveg,
  efficacy_parent = peffv,
  efficacy_teen = teffv,
  efficacy_centered_parent = cpeffv,
  efficacy_centered_teen = cteffv,
  is_female_parent = pfem,
  is_female_teen = tfem
)
```

## 1. Lavaan: J-P and Niall's example

### Fit the model

```{r jp-niall-lavaan}
apim_model <- '
  # Intake intercepts: both centered efficacy predictors are zero.
  produce_intake_teen ~ intercept_teen * 1
  produce_intake_parent ~ intercept_parent * 1
  
  # Actor effects: own centered efficacy predicts own F&V intake.
  produce_intake_teen ~ actor_teen * efficacy_centered_teen
  produce_intake_parent ~ actor_parent * efficacy_centered_parent

  # Partner effects: parent efficacy -> teen intake, and vice versa.
  produce_intake_teen ~ partner_parent_to_teen * efficacy_centered_parent
  produce_intake_parent ~ partner_teen_to_parent * efficacy_centered_teen

  # Means of the centered efficacy predictors.
  efficacy_centered_teen ~ mean_efficacy_teen * 1
  efficacy_centered_parent ~ mean_efficacy_parent * 1

  # Predictor variances.
  efficacy_centered_teen ~~ var_efficacy_teen * efficacy_centered_teen
  efficacy_centered_parent ~~ var_efficacy_parent * efficacy_centered_parent

  # Residual variances of intake.
  produce_intake_teen ~~ residual_var_teen * produce_intake_teen
  produce_intake_parent ~~ residual_var_parent * produce_intake_parent

  # Covariance between parent and teen efficacy.
  efficacy_centered_parent ~~ cov_efficacy * efficacy_centered_teen

  # Residual covariance between parent and teen intake.
  produce_intake_parent ~~ residual_cov * produce_intake_teen

  # Path-traced contributions to the outcome covariance.
  actor_driven := actor_teen * cov_efficacy * actor_parent
  partner_driven := partner_parent_to_teen * cov_efficacy * partner_teen_to_parent
  teen_driven := actor_teen * var_efficacy_teen * partner_teen_to_parent
  parent_driven := actor_parent * var_efficacy_parent * partner_parent_to_teen
  residual := residual_cov
  explained := actor_driven + partner_driven + teen_driven + parent_driven
  total_covariance := explained + residual

  # Percentages of the total outcome covariance.
  percent_actor_driven := 100 * (actor_driven / total_covariance)
  percent_partner_driven := 100 * (partner_driven / total_covariance)
  percent_teen_driven := 100 * (teen_driven / total_covariance)
  percent_parent_driven := 100 * (parent_driven / total_covariance)
  percent_residual := 100 * (residual / total_covariance)
  percent_explained := 100 * (explained / total_covariance)
'

# Predictor means, variances and covariance are free parameters above.
# Bootstrap CIs: refit to 1999 samples of the dyads.
apim_fit <- sem(
  apim_model, fixed.x = FALSE, data = flashe_data, missing = "fiml",
  se = "bootstrap", bootstrap = 1999, iseed = 2026,
  parallel = "multicore", ncpus = 4
)

# Examine the model.
# summary(apim_fit, fit.measures = TRUE)
```

### Extract the model parameters from lavaan

First the usual estimates:

```{r extract-lavaan}
# Extract once, using our model labels and including 95% CIs.
lavaan_estimates <- broom::tidy(apim_fit, conf.int = TRUE) |>
  dplyr::select(
    parameter = label,
    op,
    estimate,
    ci_low = conf.low,
    ci_high = conf.high
  )

# Model parameters: exclude the quantities defined with :=.
lavaan_params <- lavaan_estimates |>
  dplyr::filter(op != ":=") |>
  dplyr::select(-op)

print_table(lavaan_params)
```

### Extract the covariance decomposition

Now the covariance decomposition estimates and CIs:

```{r lavaan-covariance-decomposition}
# Decomposition: keep only the quantities defined with :=.
lavaan_decomposition <- lavaan_estimates |>
  dplyr::filter(op == ":=") |>
  dplyr::select(-op)

# Display the five contributions and the explained subtotal, with CI and percent.
format_decomposition <- function(decomposition) {
  # Match each percentage row to its contribution by removing "percent_".
  percentages <- decomposition |>
    dplyr::filter(startsWith(parameter, "percent_")) |>
    dplyr::mutate(parameter = sub("^percent_", "", parameter)) |>
    dplyr::select(parameter, percent = estimate)

  # Format only the displayed CI; estimates keep their full precision.
  decomposition |>
    dplyr::filter(!startsWith(parameter, "percent_"),
                  parameter != "total_covariance") |>
    dplyr::mutate(
      `95% CI` = sprintf("[%.3f, %.3f]", ci_low, ci_high)
    ) |>
    dplyr::select(parameter, estimate, `95% CI`) |>
    dplyr::left_join(percentages, by = "parameter")
}

lavaan_decomposition_table <- format_decomposition(lavaan_decomposition)
print_table(lavaan_decomposition_table)
```


## 2. glmmTMB: reproduce the point estimates

`dyadMLM` prepares the actor and partner predictors.
We retain the supplied centering and use maximum likelihood to mirror lavaan.

### Reshape and prepare the data

Reshape to one row per person (long-format):

```{r reshape-long}
# Each original row is one dyad. Long format has one row per person.
flashe_long <- flashe_data |>
  dplyr::mutate(dyad_id = dplyr::row_number()) |>
  tidyr::pivot_longer(
    cols = -dyad_id,
    names_to = c(".value", "role"),
    names_pattern = "(.+)_(parent|teen)$"
  ) |>
  dplyr::mutate(person_id = dplyr::row_number())
```

Construct the actor/partner predictors:

```{r prepare-dyadmlm}
# Construct actor/partner columns without centering the scores again.
flashe_apim <- dyadMLM::prepare_dyad_data(
  flashe_long,
  dyad = dyad_id,
  member = person_id,
  role = role,
  predictors = efficacy_centered,
  model_types = "apim"
)

print(flashe_apim, n = 4)
```

### Fit the model

```{r fit-dyadmlm}
apim_mlm_fit <- glmmTMB::glmmTMB(
  produce_intake ~
    # Separate parent and teen intercepts.
    0 + .is_parent + .is_teen +

    # Actor effects: own efficacy predicts own intake.
    .is_parent:.efficacy_centered_actor +
    .is_teen:.efficacy_centered_actor +

    # Partner effects: partner efficacy predicts own intake.
    .is_parent:.efficacy_centered_partner +
    .is_teen:.efficacy_centered_partner +

    # Two residual variances and their covariance within each dyad.
    us(0 + .is_parent + .is_teen | dyad_id),

  # Represent the paired residual variation in the unstructured block.
  dispformula = ~ 0,
  family = gaussian(),
  data = flashe_apim,

  # Match lavaan's maximum-likelihood estimation.
  REML = FALSE
)

# summary(apim_mlm_fit)
```

### Extract the standard point estimates

We extract the needed coefficients, then add labels matching lavaan.
`dispformula = ~ 0` fixes glmmTMB's separate residual variance at about
1.5e-8, too small to matter. The `us()` block therefore holds the residual
variances and covariance.

```{r extract-glmmtmb-points}
# Extract all nine estimates with their original model names.
glmmtmb_points <- broom.mixed::tidy(
  apim_mlm_fit,
  effects = c("fixed", "ran_pars"),
  scales = c(NA, "vcov") # leaves fixed effects unchanged and requests variances/covariances for REs
) |>
  dplyr::select(effect, term, estimate)

# Rename the rows to match lavaan. Keep term for later model lookups.
glmmtmb_points <- glmmtmb_points |>
  dplyr::mutate(
    parameter = dplyr::recode(
      term,
      ".is_parent" = "intercept_parent",
      ".is_teen" = "intercept_teen",
      ".is_parent:.efficacy_centered_actor" = "actor_parent",
      ".is_teen:.efficacy_centered_actor" = "actor_teen",
      ".is_parent:.efficacy_centered_partner" = "partner_teen_to_parent",
      ".is_teen:.efficacy_centered_partner" = "partner_parent_to_teen",
      "var__.is_parent" = "residual_var_parent",
      "var__.is_teen" = "residual_var_teen",
      "cov__.is_parent..is_teen" = "residual_cov"
    )
  )

glmmtmb_params_point <- dplyr::select(glmmtmb_points, parameter, estimate)
print_table(glmmtmb_params_point)
```

### Calculate the contributions with simple products

We also need the two efficacy variances and their covariance. We calculate
these from one row per dyad, dividing by $n$ as in lavaan's ML estimation.

Note: dividing by n-1 (the default in cov()) might be valid too, but would
not match lavaan exactly.

We then collect all estimates we need for the computation
in a one-row table, `mlm_inputs`.

```{r}

predictor_covariance_matrix <- cov.wt(
  flashe_data[c("efficacy_centered_parent", "efficacy_centered_teen")],
  method = "ML"
)$cov


# Put each fitted estimate in its own column.
model_estimates <- glmmtmb_params_point |>
  tidyr::pivot_wider(names_from = parameter, values_from = estimate)

# Keep the four slopes and residual covariance; add the predictor moments.
mlm_inputs <- model_estimates |>
  dplyr::select(
    actor_parent,
    actor_teen,
    partner_parent_to_teen,
    partner_teen_to_parent,
    residual_cov
  ) |>
  dplyr::mutate(
    var_efficacy_parent = predictor_covariance_matrix[1, 1],
    var_efficacy_teen = predictor_covariance_matrix[2, 2],
    cov_efficacy = predictor_covariance_matrix[1, 2],
    .before = residual_cov
  )

print_table(mlm_inputs)
```


Computing the contributions with the columns directly: 

```{r glmmtmb-point-decomposition}
glmmtmb_decomposition_point <- mlm_inputs |>
  dplyr::mutate(
    actor_driven = actor_teen * cov_efficacy * actor_parent,
    partner_driven = partner_parent_to_teen * cov_efficacy * partner_teen_to_parent,
    teen_driven = actor_teen * var_efficacy_teen * partner_teen_to_parent,
    parent_driven = actor_parent * var_efficacy_parent * partner_parent_to_teen,
    residual = residual_cov,
    
    explained = actor_driven + partner_driven + teen_driven + parent_driven,
    
    total_covariance = explained + residual,
    
    .keep = "none"
  ) |>
  # Return to one row per contribution for the display table.
  tidyr::pivot_longer(
    dplyr::everything(), names_to = "parameter", values_to = "estimate"
  ) |>
  # Divide each estimate by the total covariance row, then multiply by 100.
  dplyr::mutate(
    percent = 100 * (estimate / estimate[parameter == "total_covariance"])
  )
print_table(glmmtmb_decomposition_point)
```

### Compare the point estimates with lavaan

The differences are below 1e-4 and come from the optimizers' stopping rules (tested).

```{r compare-parameter-points}
parameter_point_comparison <- dplyr::inner_join(
  dplyr::select(lavaan_params, parameter, estimate),
  glmmtmb_params_point,
  by = "parameter",
  suffix = c("_lavaan", "_glmmtmb")
) |>
  dplyr::mutate(difference = estimate_glmmtmb - estimate_lavaan)
print_table(parameter_point_comparison)
```

### Plot the point estimates

Visualize the contribution to the total covariance.

```{r prepare-decomposition-plots}
library(ggplot2)

route_labels <- c(
  actor_driven = "Actor-driven", partner_driven = "Partner-driven",
  teen_driven = "Teen-driven", parent_driven = "Parent-driven",
  residual = "Residual"
)
# Muted teal, ochre, blue, mauve and slate, in the same order as the routes.
route_colours <- c("#468C83", "#D3A064", "#6F91B7", "#AD8CA7", "#7B8794")
plot_data <- dplyr::bind_rows(
  lavaan = dplyr::select(lavaan_decomposition, parameter, estimate),
  glmmTMB = dplyr::select(glmmtmb_decomposition_point, parameter, estimate),
  .id = "model"
) |>
  dplyr::filter(parameter %in% names(route_labels)) |>
  dplyr::mutate(
    model = factor(model, levels = c("lavaan", "glmmTMB")),
    route = factor(parameter, levels = names(route_labels),
                   labels = unname(route_labels))
  ) |>
  dplyr::arrange(model, route) |>
  dplyr::group_by(model) |>
  dplyr::mutate(
    end = cumsum(estimate), start = end - estimate,
    percent = 100 * estimate / sum(estimate)
  )

plot_totals <- dplyr::summarise(plot_data, total = sum(estimate))
```

#### Waterfall: how the contributions add up

```{r decomposition-waterfall, dev="svg", fig.width=10, fig.height=4}
ggplot(plot_data, aes(fill = route)) +
  geom_rect(aes(
    xmin = as.numeric(route) - 0.4, xmax = as.numeric(route) + 0.4,
    ymin = pmin(start, end), ymax = pmax(start, end)
  )) +
  geom_text(
    aes(x = as.numeric(route), y = pmax(start, end),
        label = sprintf("%.1f%%", percent)),
    vjust = -0.6, colour = "#475569", size = 3.5
  ) +
  geom_hline(
    data = plot_totals, aes(yintercept = total),
    linetype = "dotted", colour = "#475569"
  ) +
  geom_text(
    data = plot_totals,
    aes(x = 0.6, y = total, label = sprintf("Total = %.3f", total)),
    inherit.aes = FALSE, hjust = 0, vjust = -0.6, colour = "#475569"
  ) +
  facet_wrap(~ model, nrow = 1) +
  scale_x_continuous(
    breaks = seq_along(route_labels), labels = gsub("-", "\n", route_labels)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  scale_fill_manual(values = route_colours) +
  labs(x = NULL, y = "Outcome covariance") +
  theme_minimal() +
  theme(legend.position = "none")
```

#### Stacked bars: compare the two fits

Here all five contributions are positive, so each bar's length is the total
covariance. With negative contributions, ggplot would place positive and
negative pieces on opposite sides of zero. The waterfall above might make
their net total easier to see.

```{r decomposition-stacked-bars, dev="svg", fig.width=10, fig.height=3}
ggplot(plot_data, aes(x = estimate, y = model, fill = route)) +
  geom_col(width = 0.5, position = position_stack(reverse = TRUE)) +
  # Put labels for tiny segments below the bar to avoid overlap.
  geom_text(
    aes(x = (start + end) / 2,
        y = as.numeric(model) + ifelse(abs(percent) < 1, -0.35, 0.35),
        label = sprintf("%.1f%%", percent)),
    colour = "#475569", size = 3.5
  ) +
  scale_fill_manual(values = route_colours) +
  labs(x = "Outcome covariance", y = NULL, fill = "Contribution") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3. Add confidence intervals for glmmTMB

We use the same bootstrap as lavaan in part 1. We draw 1486 dyads with
replacement and repeat part 2 on them. We do this 1999 times. For each
estimate, the 2.5% and 97.5% quantiles of the 1999 values give the 95% CI.
This does not assume normally distributed data.

### Part 2 as a function

`add_contributions()` contains the formulas from part 2. `estimate_apim()`
runs all of part 2 on any data with one row per dyad. It returns every
estimate in one row, with lavaan's names.

```{r bootstrap-functions}
# The decomposition formulas from part 2.
add_contributions <- function(estimates) {
  estimates |>
    dplyr::mutate(
      actor_driven = actor_teen * cov_efficacy * actor_parent,
      partner_driven = partner_parent_to_teen * cov_efficacy * partner_teen_to_parent,
      teen_driven = actor_teen * var_efficacy_teen * partner_teen_to_parent,
      parent_driven = actor_parent * var_efficacy_parent * partner_parent_to_teen,
      residual = residual_cov,
      explained = actor_driven + partner_driven + teen_driven + parent_driven,
      total_covariance = explained + residual
    )
}

estimate_apim <- function(data) {
  # As in part 2: reshape, prepare the predictors and refit. New dyad IDs
  # make a dyad that was drawn twice count as two dyads.
  long <- data |>
    dplyr::mutate(dyad_id = dplyr::row_number()) |>
    tidyr::pivot_longer(
      cols = -dyad_id,
      names_to = c(".value", "role"),
      names_pattern = "(.+)_(parent|teen)$"
    ) |>
    dplyr::mutate(person_id = dplyr::row_number())
  prepared <- dyadMLM::prepare_dyad_data(
    long, dyad = dyad_id, member = person_id, role = role,
    predictors = efficacy_centered, model_types = "apim"
  )
  fit <- update(apim_mlm_fit, data = prepared)

  # The estimates, and the predictor variances and covariance (divided by n,
  # as in part 2). In both matrices, parent comes first.
  fixed_effects <- glmmTMB::fixef(fit)$cond
  residual_covariance_matrix <- glmmTMB::VarCorr(fit)$cond$dyad_id
  predictor_covariance_matrix <- cov.wt(
    data[c("efficacy_centered_parent", "efficacy_centered_teen")],
    method = "ML"
  )$cov

  tibble::tibble(
    intercept_parent = fixed_effects[[".is_parent"]],
    intercept_teen = fixed_effects[[".is_teen"]],
    actor_parent = fixed_effects[[".is_parent:.efficacy_centered_actor"]],
    actor_teen = fixed_effects[[".is_teen:.efficacy_centered_actor"]],
    partner_teen_to_parent = fixed_effects[[".is_parent:.efficacy_centered_partner"]],
    partner_parent_to_teen = fixed_effects[[".is_teen:.efficacy_centered_partner"]],
    residual_var_parent = residual_covariance_matrix[1, 1],
    residual_var_teen = residual_covariance_matrix[2, 2],
    residual_cov = residual_covariance_matrix[1, 2],
    var_efficacy_parent = predictor_covariance_matrix[1, 1],
    var_efficacy_teen = predictor_covariance_matrix[2, 2],
    cov_efficacy = predictor_covariance_matrix[1, 2]
  ) |>
    add_contributions()
}
```

### Bootstrap

```{r bootstrap-cache-key}
# The next chunk is cached. knitr only reruns it if its code or this key
# changes, that is, if the data, the model or the two functions change.
bootstrap_key <- digest::digest(list(
  flashe_data, logLik(apim_mlm_fit),
  deparse(estimate_apim), deparse(add_contributions)
))
```

```{r glmmtmb-bootstrap, cache = TRUE, cache.extra = bootstrap_key}
# 1999 samples, each holding 1486 dyads (rows of flashe_data) drawn with
# replacement.
set.seed(2026)
bootstrap <- tibble::tibble(
  dyads = replicate(
    1999, sample(nrow(flashe_data), replace = TRUE), simplify = FALSE
  )
)

# Estimate everything in each sample (about 5 to 15 minutes on 4 cores).
bootstrap <- bootstrap |>
  dplyr::mutate(estimates = parallel::mclapply(
    dyads, \(rows) estimate_apim(flashe_data[rows, ]), mc.cores = 4
  ))
```

```{r glmmtmb-cis}
# 95% CI for each column: its 2.5% and 97.5% quantiles.
percentile_cis <- function(draws) {
  draws |>
    tidyr::pivot_longer(
      dplyr::everything(), names_to = "parameter", values_to = "draw"
    ) |>
    dplyr::summarise(
      ci_low = quantile(draw, 0.025),
      ci_high = quantile(draw, 0.975),
      .by = parameter
    )
}

# One row per sample, one column per estimate.
glmmtmb_draws <- dplyr::bind_rows(bootstrap$estimates)

# Estimates from the full data, with their 95% CIs.
glmmtmb_cis <- estimate_apim(flashe_data) |>
  tidyr::pivot_longer(
    dplyr::everything(), names_to = "parameter", values_to = "estimate"
  ) |>
  dplyr::left_join(percentile_cis(glmmtmb_draws), by = "parameter")
```

### Compare the CIs with lavaan

```{r compare-cis}
# Keep the estimates that both models have.
ci_comparison <- dplyr::inner_join(
  lavaan_estimates,
  glmmtmb_cis,
  by = "parameter",
  suffix = c("_lavaan", "_glmmtmb")
) |>
  dplyr::mutate(
    ci_lavaan = sprintf("[%.3f, %.3f]", ci_low_lavaan, ci_high_lavaan),
    ci_glmmtmb = sprintf("[%.3f, %.3f]", ci_low_glmmtmb, ci_high_glmmtmb)
  ) |>
  dplyr::select(parameter, ci_lavaan, ci_glmmtmb)
print_table(ci_comparison)
```

The CIs are very close. They differ a little because lavaan and glmmTMB
drew different random samples. In `explore-sem-mlm-details.Rmd`, lavaan is
refit to the same samples as glmmTMB, and then every estimate agrees. That
file also compares other CI methods. For glmmTMB, a Monte Carlo CI with
`vcov(apim_mlm_fit, full = TRUE, sandwich = TRUE)` is a faster option.

## 4. Plot the confidence intervals

One panel per contribution, each on its own scale.

```{r decomposition-intervals, dev="svg", fig.width=8, fig.height=5}
ci_plot_data <- dplyr::bind_rows(
  lavaan = lavaan_decomposition, glmmTMB = glmmtmb_cis,
  .id = "model"
) |>
  dplyr::filter(parameter %in% names(route_labels)) |>
  dplyr::mutate(
    model = factor(model, levels = c("lavaan", "glmmTMB")),
    route = factor(parameter, levels = names(route_labels),
                   labels = unname(route_labels))
  )

ggplot(ci_plot_data, aes(x = estimate, y = model, colour = route)) +
  geom_pointrange(aes(xmin = ci_low, xmax = ci_high),
                  show.legend = FALSE) +
  facet_wrap(~ route, ncol = 1, scales = "free_x") +
  scale_colour_manual(values = route_colours) +
  labs(x = "Covariance contribution", y = NULL) +
  theme_minimal() +
  # Leave room for the last x-axis label.
  theme(plot.margin = margin(5.5, 15, 5.5, 5.5))
```
