---
title: "APIM exploration: checks and CI methods"
author: "Pascal Küng"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    math_method: mathml
    df_print: kable
    css: explore-sem-mlm.css
---

This notebook checks the glmmTMB bootstrap from `explore-sem-mlm.Rmd` and
compares other ways to get CIs. It first runs the main notebook in the
background, so all its results are available here.

```{r run-main-notebook, include=FALSE}
# Use the main notebook's cache and figure folders, so the bootstrap is not
# rerun. Without the cache, this takes 5 to 15 minutes.
invisible(knitr::knit_child(
  "explore-sem-mlm.Rmd",
  options = list(
    cache.path = "explore-sem-mlm_cache/html/",
    fig.path = "explore-sem-mlm_files/figure-html/"
  ),
  quiet = TRUE
))
```

## 1. glmmTMB and lavaan on the same samples

In the main notebook, lavaan and glmmTMB drew different samples, so their
CIs differ a little. Here we refit lavaan to the same 1999 samples as glmmTMB
(`bootstrap$dyads`). If the glmmTMB bootstrap is right, both give the same
estimates in every sample, apart from tiny optimizer differences.

```{r same-samples}
# Refit lavaan to any data and return all estimates in one row. as.list()
# is needed because coef() returns a special lavaan class.
estimate_lavaan <- function(data) {
  fit <- update(apim_fit, data = data, se = "none")
  tibble::as_tibble_row(as.list(coef(fit, type = "user")))
}

# One row per sample and one column per estimate, as in glmmtmb_draws.
lavaan_draws <- parallel::mclapply(
  bootstrap$dyads, \(rows) estimate_lavaan(flashe_data[rows, ]), mc.cores = 4
) |>
  dplyr::bind_rows() |>
  dplyr::select(dplyr::all_of(names(glmmtmb_draws)))

# Largest difference between glmmTMB and lavaan.
max(abs(glmmtmb_draws - lavaan_draws))
```

## 2. Other CI methods for the contributions

The main notebook uses the bootstrap. Here we compare it with three other
methods:

- lavaan's default delta method. It assumes normally distributed data, and
  its CIs are symmetric.
- The same delta method with robust standard errors (`estimator = "MLR"`),
  which do not assume normal data.
- A Monte Carlo CI for glmmTMB (Preacher & Selig, 2012). It draws many sets
  of estimates from their estimated sampling distribution and computes the
  contributions for each set. With `sandwich = TRUE`, the model uncertainty is
  robust and clustered by dyad. The predictor variances and covariance are
  drawn assuming normal data.

```{r delta-method}
# Without se = "bootstrap", lavaan uses the delta method. MLR makes its
# standard errors robust.
delta_fit <- sem(
  apim_model, fixed.x = FALSE, data = flashe_data, missing = "fiml"
)
mlr_fit <- sem(
  apim_model, fixed.x = FALSE, data = flashe_data, missing = "fiml",
  estimator = "MLR"
)
```

```{r monte-carlo}
# Takes about 3 minutes, mostly for the sandwich matrix.
set.seed(2026)
n_draws <- 20000
n_dyads <- nrow(flashe_data)

# Draw the coefficients and glmmTMB's internal variance parameters together.
# glmmTMB stores the residual (co)variances as two log SDs and a transformed
# correlation (theta).
theta <- glmmTMB::getME(apim_mlm_fit, "theta")
model_draws <- MASS::mvrnorm(
  n_draws,
  mu = c(
    glmmTMB::fixef(apim_mlm_fit)$cond,
    log_sd_parent = theta[1], log_sd_teen = theta[2], cor_theta = theta[3]
  ),
  Sigma = vcov(apim_mlm_fit, full = TRUE, sandwich = TRUE)
)

# Draw the predictor variances and covariance, assuming normal data.
# rWishart() returns one 2 x 2 matrix per draw.
predictor_draws <- rWishart(
  n_draws, df = n_dyads, Sigma = predictor_covariance_matrix / n_dyads
)

# One row per draw, with the names from estimate_apim(). Then add the
# contributions as before.
monte_carlo_draws <- tibble::as_tibble(model_draws) |>
  dplyr::transmute(
    actor_parent = `.is_parent:.efficacy_centered_actor`,
    actor_teen = `.is_teen:.efficacy_centered_actor`,
    partner_teen_to_parent = `.is_parent:.efficacy_centered_partner`,
    partner_parent_to_teen = `.is_teen:.efficacy_centered_partner`,
    # Residual covariance = SD parent * SD teen * correlation.
    residual_cov = exp(log_sd_parent) * exp(log_sd_teen) *
      cor_theta / sqrt(1 + cor_theta^2),
    var_efficacy_parent = predictor_draws[1, 1, ],
    var_efficacy_teen = predictor_draws[2, 2, ],
    cov_efficacy = predictor_draws[1, 2, ]
  ) |>
  add_contributions()

monte_carlo_cis <- percentile_cis(monte_carlo_draws)
```

```{r compare-ci-methods}
# The CIs from a lavaan fit, with our labels.
lavaan_cis <- function(fit) {
  broom::tidy(fit, conf.int = TRUE) |>
    dplyr::select(parameter = label, ci_low = conf.low, ci_high = conf.high)
}

# One row per contribution, one column per method.
ci_methods <- dplyr::bind_rows(
  `glmmTMB bootstrap` = glmmtmb_cis,
  `lavaan delta` = lavaan_cis(delta_fit),
  `lavaan delta, MLR` = lavaan_cis(mlr_fit),
  `glmmTMB Monte Carlo` = monte_carlo_cis,
  .id = "method"
) |>
  dplyr::filter(parameter %in% c(names(route_labels), "explained")) |>
  dplyr::mutate(ci = sprintf("[%.3f, %.3f]", ci_low, ci_high)) |>
  tidyr::pivot_wider(
    id_cols = parameter, names_from = method, values_from = ci
  )
print_table(ci_methods)
```

The bootstrap, the robust delta method and the Monte Carlo CIs are close.
The default delta method differs most for the residual covariance. It
assumes normal data, but the intake scores are not normally distributed.
