---
title: "APIM exploration: checks and CI methods"
author: "Pascal Küng"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    math_method: mathml
    df_print: kable
    css: explore-sem-mlm.css
---

This notebook adds checks and other CI methods to `explore-sem-mlm.Rmd`. It
first runs that notebook's code, so its data, fits and bootstrap results are
available here.

```{r run-main-notebook, include=FALSE}
# About 5 to 15 minutes, mostly for the glmmTMB bootstrap.
source(knitr::purl("explore-sem-mlm.Rmd", output = tempfile(), quiet = TRUE))
```

## 1. glmmTMB and lavaan on the same resamples

In the main notebook, the two bootstraps draw different resamples, so they
agree only within resampling noise. Here we refit lavaan to the same 1999
samples that glmmTMB used. `boot.array()` returns the rows of `flashe_data`
that each sample drew. If the glmmTMB bootstrap is correct, both models give
the same contributions in every sample, up to the precision of the
optimizers.

```{r same-resamples}
# One row per sample, holding the rows of flashe_data that it drew.
resampled_rows <- boot::boot.array(boot_decomposition, indices = TRUE)

# Refit lavaan to each sample and keep the same seven quantities.
lavaan_draws <- parallel::mclapply(
  seq_len(nrow(resampled_rows)),
  function(b) {
    fit <- update(apim_fit, data = flashe_data[resampled_rows[b, ], ],
                  se = "none")
    coef(fit, type = "user")[names(boot_decomposition$t0)]
  },
  mc.cores = 4
)
lavaan_draws <- do.call(rbind, lavaan_draws)

# Largest difference between glmmTMB and lavaan, over all samples.
max(abs(boot_decomposition$t - lavaan_draws))
```

## 2. Other CI methods for the contributions

The main notebook uses the bootstrap. We compare it with three alternatives:

- **lavaan, delta method:** lavaan's default. It assumes normally distributed
  data and gives symmetric CIs.
- **lavaan, delta method with MLR:** the same, but with robust SEs that do not
  assume normal data.
- **glmmTMB, Monte Carlo:** draws many sets of estimates from their estimated
  sampling distribution and applies the formulas to each set (Preacher &
  Selig, 2012). `sandwich = TRUE` makes the uncertainty of the model
  estimates robust, clustered by dyad. The predictor variances and covariance
  are drawn assuming normal data.

```{r delta-method}
# Without se = "bootstrap", lavaan uses the delta method.
delta_fit <- sem(
  apim_model, fixed.x = FALSE, data = flashe_data, missing = "fiml"
)
mlr_fit <- sem(
  apim_model, fixed.x = FALSE, data = flashe_data, missing = "fiml",
  estimator = "MLR"
)
```

```{r monte-carlo}
# About 3 minutes, mostly for the sandwich uncertainty matrix.
set.seed(2026)
n_draws <- 20000

# Draw the coefficients and glmmTMB's internal variance parameters (theta)
# together: log SD parent, log SD teen, and a transformed correlation.
draws <- MASS::mvrnorm(
  n_draws,
  mu = c(
    glmmTMB::fixef(apim_mlm_fit)$cond,
    glmmTMB::getME(apim_mlm_fit, "theta")
  ),
  Sigma = vcov(apim_mlm_fit, full = TRUE, sandwich = TRUE)
)
theta <- draws[, 7:9]

# Draw the predictor variances and covariance (normal theory).
n_dyads <- nrow(flashe_data)
predictor_draws <- rWishart(
  n_draws, df = n_dyads, Sigma = predictor_covariance_matrix / n_dyads
)

# The same formulas as in the main notebook, once per draw.
actor_parent <- draws[, ".is_parent:.efficacy_centered_actor"]
actor_teen <- draws[, ".is_teen:.efficacy_centered_actor"]
partner_teen_to_parent <- draws[, ".is_parent:.efficacy_centered_partner"]
partner_parent_to_teen <- draws[, ".is_teen:.efficacy_centered_partner"]
var_efficacy_parent <- predictor_draws[1, 1, ]
var_efficacy_teen <- predictor_draws[2, 2, ]
cov_efficacy <- predictor_draws[1, 2, ]
residual_cov <- exp(theta[, 1] + theta[, 2]) * theta[, 3] /
  sqrt(1 + theta[, 3]^2)

mc_draws <- cbind(
  actor_driven = actor_teen * cov_efficacy * actor_parent,
  partner_driven = partner_parent_to_teen * cov_efficacy * partner_teen_to_parent,
  teen_driven = actor_teen * var_efficacy_teen * partner_teen_to_parent,
  parent_driven = actor_parent * var_efficacy_parent * partner_parent_to_teen,
  residual = residual_cov
)
mc_draws <- cbind(mc_draws, explained = rowSums(mc_draws[, 1:4]))

monte_carlo <- tibble::tibble(
  parameter = colnames(mc_draws),
  ci_low = apply(mc_draws, 2, quantile, 0.025),
  ci_high = apply(mc_draws, 2, quantile, 0.975)
)
```

```{r compare-ci-methods}
lavaan_cis <- function(fit) {
  broom::tidy(fit, conf.int = TRUE) |>
    dplyr::select(parameter = label, ci_low = conf.low, ci_high = conf.high)
}

ci_methods <- dplyr::bind_rows(
  `glmmTMB bootstrap` = glmmtmb_decomposition,
  `lavaan delta` = lavaan_cis(delta_fit),
  `lavaan delta, MLR` = lavaan_cis(mlr_fit),
  `glmmTMB Monte Carlo` = monte_carlo,
  .id = "method"
) |>
  dplyr::filter(parameter %in% c(names(route_labels), "explained")) |>
  dplyr::mutate(ci = sprintf("[%.3f, %.3f]", ci_low, ci_high)) |>
  tidyr::pivot_wider(
    id_cols = parameter, names_from = method, values_from = ci
  )
print_table(ci_methods)
```

The bootstrap, the robust delta method and the Monte Carlo CIs are close.
The normal-theory delta method differs most for the residual covariance,
because the intake scores are not normally distributed.
