---
title: "Explaining Interdependence in the Actor–Partner Interdependence Model"
subtitle: "Technical notes, worked examples, and reusable figures"
author: "Pascal Küng"
date: "`r format(Sys.Date(), '%d %B %Y')`"
bibliography:
  - ../../vignettes/references.bib
  - references.bib
link-citations: true
output:
  html_document:
    math_method: mathml
    toc: true
    toc_float: true
    number_sections: true
    fig_caption: true
    self_contained: true
---

<style>
math[display="block"] {
  font-size: 1.15em;
  margin: 1.1em 0;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  dpi = 144
)

helper_candidates <- c(
  "../../vignettes/diagram-helpers.Rinc",
  "../vignettes/diagram-helpers.Rinc",
  "vignettes/diagram-helpers.Rinc"
)
helper_file <- helper_candidates[file.exists(helper_candidates)][1]

if (is.na(helper_file)) {
  stop("The shared vignette diagram helpers could not be found.")
}
sys.source(helper_file, envir = knitr::knit_global())
diagram_device <- .vignette_diagram_device()
effect_colours <- .interdep_effect_colors(c("actor", "partner"))

knitr::opts_chunk$set(
  dev = diagram_device,
  fig.ext = diagram_device,
  dev.args = list(bg = "transparent")
)

colours <- c(
  actor = unname(effect_colours[["actor"]]),
  partner = unname(effect_colours[["partner"]]),
  mixed_1 = "#3F5EAE",
  mixed_2 = "#6250A7",
  residual = "#546E7A",
  covariance = "#334155",
  ink = "#263238",
  muted = "#64748B",
  faint = "#C6CFD8",
  surface = "#F8FAFC",
  white = "#FFFFFF"
)

fade <- function(colour, alpha = 0.22) {
  grDevices::adjustcolor(colour, alpha.f = alpha)
}

component_state <- function(component) {
  all_on <- identical(component, "TOTAL")
  c(
    actor_1 = all_on || component %in% c("AA", "AP"),
    actor_2 = all_on || component %in% c("AA", "PA"),
    partner_1 = all_on || component %in% c("PA", "PP"),
    partner_2 = all_on || component %in% c("AP", "PP"),
    predictor_covariance = all_on || component %in% c("AA", "PP"),
    residual_covariance = all_on || component == "R"
  )
}

on_colour <- function(on, colour) {
  if (isTRUE(on)) colour else colours[["faint"]]
}

draw_apim_paths <- function(component = "TOTAL", panel_title = NULL,
                            show_formula = FALSE, newpage = TRUE,
                            text_scale = 1) {
  component <- match.arg(component, c("TOTAL", "AA", "AP", "PA", "PP", "R"))
  state <- component_state(component)
  native <- function(value) grid::unit(value, "native")
  arrow_head <- grid::arrow(
    angle = 28, length = grid::unit(0.13 * text_scale, "inches"),
    type = "closed"
  )
  covariance_arrow <- grid::arrow(
    angle = 26, length = grid::unit(0.09 * text_scale, "inches"),
    ends = "both", type = "closed"
  )

  # Reuse the vignette palette, math labels, annotations, and rounded-node
  # layout. This paper-specific adaptation adds predictor covariance and
  # route highlighting without changing the vignette diagrams.
  draw_node <- function(x, y, label, active = TRUE, residual = FALSE) {
    node_colour <- on_colour(active, colours[["ink"]])
    node_gp <- grid::gpar(
      fill = if (residual) "white" else colours[["surface"]],
      col = node_colour, lwd = 1.8 * text_scale
    )
    if (residual) {
      grid::grid.circle(
        x = native(x), y = native(y), r = grid::unit(0.055, "snpc"),
        gp = node_gp
      )
    } else {
      grid::grid.roundrect(
        x = native(x), y = native(y),
        width = native(0.17), height = native(0.19),
        r = grid::unit(0.04, "snpc"), gp = node_gp
      )
    }
    grid::grid.text(
      .diagram_math_label(label), x = native(x), y = native(y),
      gp = grid::gpar(
        col = node_colour, fontsize = (if (residual) 17 else 21) * text_scale
      )
    )
  }

  draw_path <- function(x0, y0, x1, y1, active, colour) {
    path_colour <- on_colour(active, colour)
    grid::grid.segments(
      x0 = native(x0), y0 = native(y0),
      x1 = native(x1), y1 = native(y1), arrow = arrow_head,
      gp = grid::gpar(
        col = path_colour, fill = path_colour,
        lwd = if (active) 2.4 * text_scale else 1.2 * text_scale
      )
    )
  }

  path_label <- function(label, x, y, active, colour) {
    .draw_residual_annotation(
      label, x, y, fontsize = 15 * text_scale,
      colour = on_colour(active, colour)
    )
  }

  if (isTRUE(newpage)) grid::grid.newpage()
  grid::pushViewport(grid::viewport(xscale = c(0, 1), yscale = c(0, 1)))
  on.exit(grid::popViewport(), add = TRUE)

  if (!is.null(panel_title)) {
    grid::grid.text(
      panel_title, x = native(0.06), y = native(0.91), just = "left",
      gp = grid::gpar(
        col = colours[["ink"]], fontsize = 16 * text_scale, fontface = "bold"
      )
    )
  } else {
    grid::grid.text(
      c("Predictor", "Outcome", "Residual"),
      x = native(c(0.24, 0.65, 0.875)), y = native(0.89),
      gp = grid::gpar(col = colours[["muted"]], fontsize = 11 * text_scale)
    )
  }

  # Predictors on the left; actor and partner paths flow into outcomes on the
  # right. Partner subscripts always identify the outcome member.
  draw_path(0.325, 0.70, 0.565 - .diagram_arrow_clearance, 0.70,
            state[["actor_1"]], colours[["actor"]])
  draw_path(0.325, 0.30, 0.565 - .diagram_arrow_clearance, 0.30,
            state[["actor_2"]], colours[["actor"]])
  draw_path(0.325, 0.67, 0.565 - .diagram_arrow_clearance, 0.34,
            state[["partner_2"]], colours[["partner"]])
  draw_path(0.325, 0.33, 0.565 - .diagram_arrow_clearance, 0.66,
            state[["partner_1"]], colours[["partner"]])

  predictor_colour <- on_colour(
    state[["predictor_covariance"]], colours[["covariance"]]
  )
  grid::grid.curve(
    x1 = native(0.153), y1 = native(0.67),
    x2 = native(0.153), y2 = native(0.33),
    curvature = 0.32, angle = 90, ncp = 12, square = FALSE,
    arrow = covariance_arrow,
    gp = grid::gpar(
      col = predictor_colour, fill = predictor_colour, lwd = 1.8 * text_scale
    )
  )
  path_label(expression(sigma[X[12]]), 0.11, 0.50,
             state[["predictor_covariance"]], colours[["covariance"]])

  # Residual arrows still point into Y. The double-headed curve is covariance
  # psi_12, not the residual correlation shown in fitted vignette diagrams.
  residual_colour <- on_colour(
    state[["residual_covariance"]], colours[["residual"]]
  )
  for (y in c(0.70, 0.30)) {
    draw_path(0.840, y, 0.735 + .diagram_arrow_clearance, y,
              state[["residual_covariance"]], colours[["residual"]])
  }
  grid::grid.curve(
    x1 = native(0.905), y1 = native(0.66),
    x2 = native(0.905), y2 = native(0.34),
    curvature = -0.28, angle = 90, ncp = 12, square = FALSE,
    arrow = covariance_arrow,
    gp = grid::gpar(
      col = residual_colour, fill = residual_colour, lwd = 1.8 * text_scale
    )
  )
  path_label(expression(italic("\u03c8")[12]), 0.92, 0.50,
             state[["residual_covariance"]], colours[["residual"]])

  draw_node(0.24, 0.70, expression(X[1]),
            component %in% c("TOTAL", "AA", "AP", "PP"))
  draw_node(0.24, 0.30, expression(X[2]),
            component %in% c("TOTAL", "AA", "PA", "PP"))
  draw_node(0.65, 0.70, expression(Y[1]))
  draw_node(0.65, 0.30, expression(Y[2]))
  draw_node(0.875, 0.70, expression(epsilon[1]),
            state[["residual_covariance"]], residual = TRUE)
  draw_node(0.875, 0.30, expression(epsilon[2]),
            state[["residual_covariance"]], residual = TRUE)

  path_label(expression(a[1]), 0.44, 0.70,
             state[["actor_1"]], colours[["actor"]])
  path_label(expression(a[2]), 0.44, 0.30,
             state[["actor_2"]], colours[["actor"]])
  path_label(expression(p[2]), 0.40, 0.56,
             state[["partner_2"]], colours[["partner"]])
  path_label(expression(p[1]), 0.40, 0.44,
             state[["partner_1"]], colours[["partner"]])

  if (isTRUE(show_formula)) {
    formulas <- list(
      AA = expression(italic(C)[plain(AA)] == a[1] * a[2] * sigma[X[12]]),
      AP = expression(italic(C)[plain(AP)] == a[1] * p[2] * sigma[X[1]]^2),
      PA = expression(italic(C)[plain(PA)] == p[1] * a[2] * sigma[X[2]]^2),
      PP = expression(italic(C)[plain(PP)] == p[1] * p[2] * sigma[X[12]]),
      R = expression(italic(C)[plain(R)] == italic("\u03c8")[12]),
      TOTAL = expression(sum(italic(C)[italic(k)], italic(k)) ==
                           Cov(Y[1], Y[2]))
    )
    grid::grid.text(
      .diagram_math_label(formulas[[component]]),
      x = native(0.50), y = native(0.08),
      gp = grid::gpar(col = colours[["ink"]], fontsize = 15 * text_scale)
    )
  } else {
    .draw_diagram_legend(
      labels = c("Actor path", "Partner path"),
      colours = unname(effect_colours), y = 0.08,
      segment_starts = c(0.26, 0.53), text_x = c(0.40, 0.68),
      fontsize = 11 * text_scale
    )
  }
  invisible(NULL)
}

component_colours <- c(
  "Actor–actor" = colours[["actor"]],
  "Actor–partner" = colours[["mixed_1"]],
  "Partner–actor" = colours[["mixed_2"]],
  "Partner–partner" = colours[["partner"]],
  "Residual" = colours[["residual"]]
)
```

> **Technical notes (1 September 2026).** These notes develop the equations,
> worked examples, and figures. The current scope, manuscript structure, and
> validation decisions belong in [plan.md](plan.md); the
> [literature review](literature-review.md) records the evidence and remaining
> source-access gaps. Inference methods and simulation settings are not yet
> finalized.

> **Working claim.** An APIM should not only *account for* interdependence. It
> can also provide a model-based, descriptive explanation of where the
> covariance between partners' outcomes comes from.

Here and throughout, *explain* means attribute statistically within the fitted
model. It does not mean that the routes are automatically causal mechanisms.

The one-predictor example provides the teaching core. Covariates, multiple
predictors, and exchangeable dyads are also within Paper 1; later sections
record the corresponding algebra. These notes are not a second manuscript plan.

# The idea in one paragraph

The APIM is commonly introduced through actor effects, partner effects, and a
residual correlation [@kenny2006dyadic]. Work on dyadic patterns and compound
APIM processes has shown the value of turning individual coefficients into
interpretable derived quantities [@kennyDetectingMeasuringTesting2010;
@ledermannAssessingMediationDyadic2011]. That presentation nevertheless does
not by itself answer a natural dyadic question:

> Why are the two partners' outcomes correlated, and how much of that
> interdependence is associated with each modeled route?

Dyadic dependence can itself contain substantive information that is lost when
members are analyzed separately [@laurenceauUsingDiaryMethods2005]. Bolger and
Shrout [-@bolgerAccountingStatisticalDependency2007] treated covariance
structure as a starting point for studying shared events and interpersonal
influence. The present proposal asks a narrower question: within one fitted
cross-sectional APIM, which modeled routes carry the outcome covariance?

This is an existing APIM application of path tracing, not a newly discovered
decomposition. Dwyer and colleagues [-@dwyerAutonomousMotivationDyads2017]
applied the predictor-route decomposition, and the Bolger–Laurenceau webinar
[-@bolgerLaurenceauDyadicWebinar2016] teaches it with a numerical example.
Kenny's handout and APIM_MM documentation explicitly describe the partition
and signed contributions [@kennyExplainedNonindependence; @kennyAPIMMM2019];
Ferraris and colleagues [-@ferrarisDyadicSocialSupport2022] provide another
direct applied precedent with formulas. Paper 1 must build on these sources
through clarification, validation, and transparent implementation.

For one predictor construct measured on both partners, the familiar
path-tracing algebra gives an exact additive answer when predictors are
uncorrelated with outcome residuals [@bokerAlgorithmHierarchicalOrganization2002].
The four predictor-side path pairs are actor–actor, actor–partner,
partner–actor, and partner–partner. Together with residual covariance, these
five signed quantities sum to model-implied outcome covariance. Dividing every
component by the same two model-implied outcome standard deviations makes them
sum to model-implied outcome correlation. Signed path-weight decompositions
also exist for undirected Gaussian graphical models
[@jonesCovarianceDecompositionUndirected2005], but the direct APIM precedents
above govern the paper's novelty claims.

# The complete one-construct APIM

For distinguishable members 1 and 2, write

$$
\begin{aligned}
Y_1&=a_1X_1+p_1X_2+\varepsilon_1,\\
Y_2&=a_2X_2+p_2X_1+\varepsilon_2.
\end{aligned}
$$

Intercepts are omitted because they do not affect variances or covariances.
The subscript on an actor or partner effect identifies the **outcome member**:
$p_1$ is the effect of member 2's predictor on member 1's outcome; $p_2$ is the
effect of member 1's predictor on member 2's outcome.

Unless stated otherwise, $a_1,a_2,p_1,p_2$ are unstandardized coefficients and
all predictor moments are computed on the matching original scales. A fully
standardized presentation is possible, but coefficients, predictor moments,
residual covariance, and the outcome denominator must then be transformed
consistently.

```{r complete-apim, echo=FALSE, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Complete distinguishable APIM used for the decomposition, using the vignette's left-to-right layout and blue/purple path palette. Predictors are on the left and outcomes on the right. Curved double-headed paths show predictor covariance on the left and residual outcome covariance on the right; residual arrows point into the outcomes.", fig.alt="Left-to-right APIM with X1 above X2 on the left and Y1 above Y2 on the right. Blue actor paths are horizontal; purple partner paths cross diagonally. A curved path connects the predictors, and another connects the two residuals, which point into the outcomes."}
draw_apim_paths("TOTAL")
```

## Assumptions and what "exact" means

The proposed paper concerns independent dyads, continuous outcomes, and a
linear, fixed-slope, cross-sectional APIM. The variables are viewed as random
over a target population of dyads, so the result below decomposes **marginal
across-dyad covariance**. Intercepts are omitted and all variances and
covariances are taken around the appropriate role-specific means.

Assume finite second moments and

$$
\operatorname{E}(\boldsymbol\varepsilon\mid X_1,X_2)=\boldsymbol 0,
\qquad
\operatorname{Cov}(\boldsymbol\varepsilon\mid X_1,X_2)
=\boldsymbol\Psi,
$$

which implies that each predictor is uncorrelated with each outcome residual:

$$
\operatorname{Cov}(X_j,\varepsilon_k)=0
\qquad\text{for }j,k\in\{1,2\}.
$$

This is the usual exogeneity or residual-orthogonality assumption in the APIM.
It is part of the statistical model, not an identity that holds for any four
variables. Normality is not required for the covariance algebra. If
predictor–residual covariances are allowed, four additional terms enter the
outcome covariance and the five-part decomposition is incomplete.

Conditional on the observed predictors, the fixed-slope model implies
$\operatorname{Cov}(Y_1,Y_2\mid X_1,X_2)=\psi_{12}$. The predictor-side routes
appear when moving to marginal covariance across dyads whose predictor values
vary. Random slopes, nonlinear links, and longitudinal carryover paths create
additional terms and are outside the first paper's scope.

The derivation itself is short. By the linearity of covariance,

$$
\begin{aligned}
\operatorname{Cov}(Y_1,Y_2)
={}&\operatorname{Cov}\!\left(
a_1X_1+p_1X_2+\varepsilon_1,
a_2X_2+p_2X_1+\varepsilon_2
\right).
\end{aligned}
$$

Expanding this expression and applying the assumption above gives

$$
\begin{aligned}
\operatorname{Cov}(Y_1,Y_2)
={}&
\underbrace{a_1a_2\operatorname{Cov}(X_1,X_2)}_{\text{actor--actor}}
+\underbrace{a_1p_2\operatorname{Var}(X_1)}_{\text{actor--partner}}\\
&+\underbrace{p_1a_2\operatorname{Var}(X_2)}_{\text{partner--actor}}
+\underbrace{p_1p_2\operatorname{Cov}(X_1,X_2)}_{\text{partner--partner}}
+\underbrace{\operatorname{Cov}(\varepsilon_1,\varepsilon_2)}_{\text{residual}}.
\end{aligned}
$$

The four predictor–residual terms that disappear are
$a_1\operatorname{Cov}(X_1,\varepsilon_2)$,
$p_1\operatorname{Cov}(X_2,\varepsilon_2)$,
$a_2\operatorname{Cov}(X_2,\varepsilon_1)$, and
$p_2\operatorname{Cov}(X_1,\varepsilon_1)$. No more elaborate proof is needed:
the five remaining terms are precisely the five covariance-carrying routes in
the diagram.

# The five covariance-carrying routes

```{r path-components, echo=FALSE, fig.width=12, fig.height=9.9, out.width="100%", fig.cap="The five covariance-carrying routes, arranged in three rows of two left-to-right APIMs. Pale paths remain part of the model but do not belong to the highlighted component. The member-driven routes use one predictor's variance, not the covariance between predictors. The final panel restores all paths.", fig.alt="Six left-to-right APIM panels in three rows of two. In reading order, they highlight actor-actor, actor-partner from X1, partner-actor from X2, partner-partner, residual covariance, and all five components. Each panel includes its component formula."}
components <- c("AA", "AP", "PA", "PP", "R", "TOTAL")
panel_titles <- c(
  "1. Actor–actor", "2. Actor–partner",
  "3. Partner–actor", "4. Partner–partner",
  "5. Residual", "Their exact sum"
)
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(3, 2)))
for (index in seq_along(components)) {
  grid::pushViewport(grid::viewport(
    layout.pos.row = (index - 1L) %/% 2L + 1L,
    layout.pos.col = (index - 1L) %% 2L + 1L
  ))
  draw_apim_paths(
    components[[index]], panel_titles[[index]], show_formula = TRUE,
    newpage = FALSE, text_scale = 0.95
  )
  grid::popViewport()
}
grid::popViewport()
```

In the diagrams, $\sigma_{X_{12}}=\operatorname{Cov}(X_1,X_2)$,
$\sigma_{X_j}^2=\operatorname{Var}(X_j)$, and
$\psi_{12}=\operatorname{Cov}(\varepsilon_1,\varepsilon_2)$.

The four predictor-side combinations form a useful $2\times2$ table:

| Path into $Y_1$ | Path into $Y_2$ | Shared predictor quantity | Contribution |
|:--|:--|:--|:--|
| actor $a_1$ | actor $a_2$ | $\operatorname{Cov}(X_1,X_2)$ | $C_{\mathrm{AA}}=a_1a_2\operatorname{Cov}(X_1,X_2)$ |
| actor $a_1$ | partner $p_2$ | $\operatorname{Var}(X_1)$ | $C_{\mathrm{AP}}=a_1p_2\operatorname{Var}(X_1)$ |
| partner $p_1$ | actor $a_2$ | $\operatorname{Var}(X_2)$ | $C_{\mathrm{PA}}=p_1a_2\operatorname{Var}(X_2)$ |
| partner $p_1$ | partner $p_2$ | $\operatorname{Cov}(X_1,X_2)$ | $C_{\mathrm{PP}}=p_1p_2\operatorname{Cov}(X_1,X_2)$ |

The residual component is

$$
C_{\mathrm{R}}=\operatorname{Cov}(\varepsilon_1,\varepsilon_2).
$$

Therefore,

$$
\operatorname{Cov}(Y_1,Y_2)
=C_{\mathrm{AA}}+C_{\mathrm{AP}}+C_{\mathrm{PA}}+C_{\mathrm{PP}}+C_{\mathrm{R}}.
$$

## Plain-language interpretation

- **Actor–actor:** partners resemble each other on $X$, and each person's own
  $X$ predicts their own $Y$.
- **Actor–partner:** member 1's $X$ reaches both outcomes—through an actor path
  into $Y_1$ and a partner path into $Y_2$.
- **Partner–actor:** member 2's $X$ reaches both outcomes in the analogous way.
- **Partner–partner:** partners resemble each other on $X$, and each predictor
  crosses to the other member's outcome.
- **Residual:** outcome interdependence that remains after the modeled
  predictor routes.

## Female-driven and male-driven interdependence

For a female--male dyad, let member 1 be the female partner and member 2 the
male partner. The two routes carried by a single member's predictor variance
then have an especially intuitive interpretation:

$$
\underbrace{a_1p_2\operatorname{Var}(X_1)}_{\text{female-driven}}
\qquad\text{and}\qquad
\underbrace{p_1a_2\operatorname{Var}(X_2)}_{\text{male-driven}}.
$$

The first is female-driven because the female partner's predictor reaches both
outcomes: through her actor path into her own outcome and through her partner
path into the male partner's outcome. The second is male-driven in the
corresponding way. Here, "driven" describes a model-implied covariance route;
it is causal only when the design and assumptions support a causal
interpretation.

This yields two broader substantive families:

$$
\underbrace{C_{\mathrm{AP}}+C_{\mathrm{PA}}}_{\text{predictor-transmission routes}}
\qquad\text{and}\qquad
\underbrace{C_{\mathrm{AA}}+C_{\mathrm{PP}}}_{\text{cross-member predictor-covariance routes}}.
$$

"Predictor transmission" is an algebraic route label, not by itself a claim
that one partner causally transmits something to the other. The second family
can be called **predictor similarity** when
$\operatorname{Cov}(X_1,X_2)>0$; a negative predictor covariance instead
represents dissimilarity.

The phrase "actor-driven covariance" by itself is potentially misleading. One
person's predictor variance creates covariance between the two outcomes only
when that predictor reaches **both** outcomes—once through an actor path and
once through a partner path. Actor and partner paths work in pairs.

# From covariance to correlation points

The five components above are in the original units of $Y_1$ times the units
of $Y_2$. That scale is exact but often difficult to interpret. To express each
component on the familiar correlation scale, define the model-implied
denominator

$$
s_Y=\operatorname{SD}(Y_1)\operatorname{SD}(Y_2).
$$

For every component $C_k$, define a correlation-point contribution

$$
c_k=\frac{C_k}{s_Y}.
$$

These are signed additive contributions on the outcome-correlation scale. An
individual $c_k$ is not itself a correlation, a proportion, or a variance
share, and it need not lie between $-1$ and $1$. Percentages are especially
misleading when positive and negative routes cancel: they can be negative,
exceed 100%, or become unstable when the total covariance is near zero.

Because every component uses the same denominator,

$$
\rho^{\mathrm{model}}_{Y_1Y_2}
=c_{\mathrm{AA}}+c_{\mathrm{AP}}+c_{\mathrm{PA}}+c_{\mathrm{PP}}+c_{\mathrm{R}}.
$$

The required total outcome variances are

$$
\begin{aligned}
\operatorname{Var}(Y_1)
={}&a_1^2\operatorname{Var}(X_1)
+p_1^2\operatorname{Var}(X_2)
+2a_1p_1\operatorname{Cov}(X_1,X_2)
+\operatorname{Var}(\varepsilon_1),\\
\operatorname{Var}(Y_2)
={}&a_2^2\operatorname{Var}(X_2)
+p_2^2\operatorname{Var}(X_1)
+2a_2p_2\operatorname{Cov}(X_1,X_2)
+\operatorname{Var}(\varepsilon_2).
\end{aligned}
$$

The observed sample correlation, $r_{Y_1Y_2}$, should be reported alongside
the model-implied correlation as a descriptive reference. They are conceptually
different quantities and are equal only when the fitted model reproduces the
relevant sample moments. The five components should therefore not be rescaled
afterward merely to force them to sum to the observed correlation.

> **Important distinction.** The residual correlation
> $\operatorname{Cor}(\varepsilon_1,\varepsilon_2)$ divides residual covariance
> by the two **residual** standard deviations. The residual contribution $c_{\mathrm{R}}$
> divides the same covariance by the two **total model-implied outcome** standard
> deviations. Only $c_{\mathrm{R}}$ is on the same scale as the other pieces of
> the model-implied outcome correlation.

# Worked numerical example

Suppose

$$
\operatorname{Var}(X_1)=\operatorname{Var}(X_2)=1,
\qquad
\operatorname{Cov}(X_1,X_2)=0.40,
$$

$$
a_1=0.50,\quad a_2=0.40,\quad p_1=0.20,\quad p_2=0.30,
$$

and

$$
\boldsymbol\Psi=
\begin{pmatrix}
0.50 & 0.15\\
0.15 & 0.55
\end{pmatrix}.
$$

```{r calculate-example, include=FALSE}
example_parameters <- list(
  var_x1 = 1,
  var_x2 = 1,
  cov_x12 = 0.40,
  a1 = 0.50,
  a2 = 0.40,
  p1 = 0.20,
  p2 = 0.30,
  psi11 = 0.50,
  psi22 = 0.55,
  psi12 = 0.15
)

with(example_parameters, {
  example_components <<- c(
    "Actor–actor" = a1 * a2 * cov_x12,
    "Actor–partner" = a1 * p2 * var_x1,
    "Partner–actor" = p1 * a2 * var_x2,
    "Partner–partner" = p1 * p2 * cov_x12,
    "Residual" = psi12
  )

  example_var_y1 <<-
    a1^2 * var_x1 + p1^2 * var_x2 +
    2 * a1 * p1 * cov_x12 + psi11
  example_var_y2 <<-
    a2^2 * var_x2 + p2^2 * var_x1 +
    2 * a2 * p2 * cov_x12 + psi22
})

example_cov_y12 <- sum(example_components)
example_denominator <- sqrt(example_var_y1 * example_var_y2)
example_correlation_components <- example_components / example_denominator
example_rho_y12 <- sum(example_correlation_components)
example_residual_rho <- with(
  example_parameters,
  psi12 / sqrt(psi11 * psi22)
)

example_table <- data.frame(
  Route = names(example_components),
  `Covariance contribution` = unname(example_components),
  `Correlation-point contribution` = unname(example_correlation_components),
  check.names = FALSE
)
```

```{r example-table, echo=FALSE}
knitr::kable(
  example_table,
  digits = 3,
  align = c("l", "r", "r")
)
```

```{r example-contributions, echo=FALSE, fig.width=9.4, fig.height=5.4, out.width="92%", fig.cap="Waterfall decomposition of the model-implied outcome correlation. Each route starts where the preceding route ended; the five signed increments arrive exactly at the model-implied total.", fig.alt="Waterfall chart in which actor-actor, actor-partner, partner-actor, partner-partner, and residual correlation-point contributions accumulate to the model-implied outcome correlation."}
old_par <- graphics::par(mar = c(7.0, 4.8, 1.2, 1.0), xpd = NA)
increments <- unname(example_correlation_components)
starts <- c(0, head(cumsum(increments), -1))
ends <- cumsum(increments)
x <- seq_along(increments)

graphics::plot(
  NA,
  xlim = c(0.45, length(increments) + 0.75),
  ylim = c(0, example_rho_y12 * 1.16),
  xaxt = "n",
  xlab = "",
  ylab = "Outcome-correlation points",
  bty = "l"
)

for (index in seq_along(increments)) {
  graphics::rect(
    x[index] - 0.34,
    min(starts[index], ends[index]),
    x[index] + 0.34,
    max(starts[index], ends[index]),
    col = component_colours[index],
    border = NA
  )
  if (index < length(increments)) {
    graphics::segments(
      x[index] + 0.34, ends[index],
      x[index + 1] - 0.34, ends[index],
      col = colours[["muted"]], lty = 3, lwd = 1.4
    )
  }
  graphics::text(
    x[index], ends[index] + example_rho_y12 * 0.025,
    sprintf("%+.3f", increments[index]),
    cex = 0.82, col = colours[["ink"]]
  )
}

graphics::axis(
  1,
  at = x,
  labels = names(example_correlation_components),
  las = 2,
  tick = FALSE,
  line = -0.3,
  cex.axis = 0.85
)
graphics::abline(h = example_rho_y12, lty = 2, lwd = 2.1,
                 col = colours[["covariance"]])
graphics::text(
  0.55, example_rho_y12 + example_rho_y12 * 0.025,
  sprintf("total ρ = %.3f", example_rho_y12),
  pos = 4, offset = 0.25, cex = 0.92,
  col = colours[["covariance"]], font = 2
)
graphics::par(old_par)
```

Here,

$$
\operatorname{Cov}(Y_1,Y_2)=`r sprintf('%.3f', example_cov_y12)`,
\qquad
\rho^{\mathrm{model}}_{Y_1Y_2}=`r sprintf('%.3f', example_rho_y12)`.
$$

The residual correlation is
`r sprintf('%.3f', example_residual_rho)`, but its contribution to the total
outcome correlation is only
`r sprintf('%.3f', unname(example_correlation_components[['Residual']]))`.
These numbers answer different questions.

# Same interdependence, different route profiles

The decomposition is useful because the same total partner correlation can
hide very different model-implied route profiles. The second hypothetical
model below is chosen to have the **same** outcome variances and covariance as the first model, but
stronger actor paths, weaker partner paths, and more residual covariance.

```{r same-total-calculation, include=FALSE}
scenario_a <- example_components

scenario_b_parameters <- list(
  var_x1 = 1,
  var_x2 = 1,
  cov_x12 = 0.40,
  a1 = 0.70,
  a2 = 0.60,
  p1 = 0.05,
  p2 = 0.05,
  psi12 = 0.25
)

with(scenario_b_parameters, {
  scenario_b <<- c(
    "Actor–actor" = a1 * a2 * cov_x12,
    "Actor–partner" = a1 * p2 * var_x1,
    "Partner–actor" = p1 * a2 * var_x2,
    "Partner–partner" = p1 * p2 * cov_x12,
    "Residual" = psi12
  )
})

# The residual variances in scenario B can be set to preserve both total
# outcome variances from scenario A; they are positive and Psi remains valid.
scenario_b_psi11 <- example_var_y1 - with(
  scenario_b_parameters,
  a1^2 * var_x1 + p1^2 * var_x2 + 2 * a1 * p1 * cov_x12
)
scenario_b_psi22 <- example_var_y2 - with(
  scenario_b_parameters,
  a2^2 * var_x2 + p2^2 * var_x1 + 2 * a2 * p2 * cov_x12
)

stopifnot(
  isTRUE(all.equal(sum(scenario_a), sum(scenario_b))),
  scenario_b_psi11 > 0,
  scenario_b_psi22 > 0,
  scenario_b_psi11 * scenario_b_psi22 >
    scenario_b_parameters$psi12^2
)

scenario_matrix <- rbind(
  "A: stronger transmission" = scenario_a,
  "B: similarity + residual" = scenario_b
) / example_denominator
```

```{r same-total-figure, echo=FALSE, fig.width=10, fig.height=5.2, out.width="96%", fig.cap="Two APIMs with the same model-implied outcome correlation but different route profiles. Each horizontal stack has exactly the same length.", fig.alt="Two horizontal stacked bars of identical total length. Scenario A has larger actor-partner and partner-actor sections; scenario B has larger actor-actor and residual sections."}
old_par <- graphics::par(mar = c(7.2, 10.2, 1.0, 1.0), xpd = NA)
graphics::barplot(
  t(scenario_matrix),
  horiz = TRUE,
  beside = FALSE,
  col = component_colours,
  border = colours[["white"]],
  names.arg = rownames(scenario_matrix),
  las = 1,
  xlim = c(0, example_rho_y12 * 1.04),
  xlab = "Outcome-correlation points",
  cex.names = 0.9
)
graphics::legend(
  "bottom", inset = c(0, -0.36),
  legend = names(component_colours),
  fill = component_colours, border = NA,
  horiz = TRUE, bty = "n", cex = 0.78, xpd = NA
)
graphics::par(old_par)
```

This is the paper's most intuitive motivation: reporting only the total or
residual correlation cannot distinguish these model-implied route profiles.

# From a fitted APIM to estimated contributions

The formulas define population or model-implied quantities. In an empirical
analysis, estimate them by plugging the fitted APIM coefficients and predictor
moments into the same expressions. Let
$\widehat\sigma_{X1}^2$, $\widehat\sigma_{X2}^2$, and
$\widehat\sigma_{X12}$ denote the estimated predictor variances and covariance,
and let $\widehat\psi_{12}$ denote the fitted residual covariance. Then

$$
\begin{aligned}
\widehat C_{\mathrm{AA}}
&=\widehat a_1\widehat a_2\widehat\sigma_{X12}, &
\widehat C_{\mathrm{AP}}
&=\widehat a_1\widehat p_2\widehat\sigma_{X1}^2,\\
\widehat C_{\mathrm{PA}}
&=\widehat p_1\widehat a_2\widehat\sigma_{X2}^2, &
\widehat C_{\mathrm{PP}}
&=\widehat p_1\widehat p_2\widehat\sigma_{X12},\\
\widehat C_{\mathrm{R}}
&=\widehat\psi_{12}.&&
\end{aligned}
$$

The estimated model-implied outcome variances supply the denominator for the
correlation-point contributions. For example,

$$
\widehat c_{\mathrm{AP}}
=\frac{\widehat C_{\mathrm{AP}}}
{\sqrt{\widehat{\operatorname{Var}}(Y_1)
       \widehat{\operatorname{Var}}(Y_2)}}.
$$

In an SEM, predictor moments may be estimated as part of the joint model or
supplied from the observed exogenous variables, depending on how the model is
specified. In a multilevel or regression implementation that conditions on the
predictors, use their empirical dyad-level variance–covariance matrix. In both
cases, the reported decomposition should state where those moments came from.

Likewise, $\widehat\psi_{12}$ must represent the **total model-implied
covariance remaining between the two members** after the fixed predictor paths.
Depending on the mixed-model parameterization, that quantity may be represented
by a shared dyad effect, a covariance between member-specific random effects,
a residual covariance, or a sum of such components. It should not be replaced
uncritically by whichever single variance component is easiest to extract.

## Inference for estimated contributions

Compare a joint delta method with a whole-dyad bootstrap; neither is selected
in advance as the universally preferable method. Let
$\boldsymbol c=g(\boldsymbol\theta)$ collect the contributions, where
$\boldsymbol\theta$ includes the fitted paths, predictor and covariate moments,
and residual covariance parameters. The outcome-standardization denominator
is also a function of these parameters. A first-order delta approximation is

$$
\widehat{\operatorname{Var}}(\widehat{\boldsymbol c})
\approx
\boldsymbol J(\widehat{\boldsymbol\theta})
\widehat{\operatorname{Var}}(\widehat{\boldsymbol\theta})
\boldsymbol J(\widehat{\boldsymbol\theta})^{\mathsf T},
\qquad
\boldsymbol J(\boldsymbol\theta)
=\frac{\partial g(\boldsymbol\theta)}{\partial\boldsymbol\theta^{\mathsf T}}.
$$

For the population-marginal target, this requires the **joint** uncertainty of
all estimated inputs, including their cross-covariances. Propagating only slope
standard errors is insufficient. Treating predictor moments as fixed targets
conditional-on-design rather than population-marginal uncertainty; even then,
propagate the joint uncertainty of all fitted parameters entering a contribution
and its denominator. Defined-parameter delta and bootstrap inference already
exist in SEM software; see the
[lavaan documentation](https://lavaan.ugent.be/tutorial/mediation.html).
Their availability is not a new methodological contribution.

A whole-dyad bootstrap instead resamples paired rows with replacement,
preserves role labels, refits the APIM, and recomputes every moment, component,
and outcome denominator. Prespecify the interval construction and failed-refit
policy. Complete paired observations are the initial validation case; an
extension to incomplete dyads must explicitly align the estimation and
predictor-moment samples.

Products can have zero gradients when multiple constituent parameters are
zero, and covariance boundaries create additional difficulties. Ordinary
first-order delta intervals can then degenerate, and a conventional bootstrap
is not automatically reliable either. The focused simulation should evaluate
bias, coverage, interval width, and fit failures, including weak or zero paths
and cancellation. Prioritize intervals and a small set of prespecified
scientific contrasts over automatic significance tests for every route.

## Within-model attribution versus covariance reduction

Within the fitted predictor model, the exact identity is

$$
\Sigma^{\mathrm{model}}_{Y,12}-\Psi_{12}
=C_{\mathrm{AA}}+C_{\mathrm{AP}}+C_{\mathrm{PA}}+C_{\mathrm{PP}}.
$$

An intercept-only model attributes all model-implied outcome covariance to its
residual covariance. However, the reduction in residual covariance between two
separately fitted models equals the predictor-side sum above only when both
models reproduce the same marginal outcome covariance under compatible
estimation and sample-moment conventions. The within-model identity is the
safer primary result.

# Matrix form and extension to several predictors

The scalar derivation is easiest to teach, but the same result can be written
compactly in matrix form. For the one-construct APIM, let

$$
\boldsymbol Y=
\begin{pmatrix}Y_1\\Y_2\end{pmatrix},
\qquad
\boldsymbol X=
\begin{pmatrix}X_1\\X_2\end{pmatrix},
\qquad
\boldsymbol B=
\begin{pmatrix}
a_1 & p_1\\
p_2 & a_2
\end{pmatrix},
$$

and

$$
\boldsymbol\Sigma_X=
\begin{pmatrix}
\sigma_{X1}^2 & \sigma_{X12}\\
\sigma_{X12} & \sigma_{X2}^2
\end{pmatrix},
\qquad
\boldsymbol\Psi=
\begin{pmatrix}
\psi_{11} & \psi_{12}\\
\psi_{12} & \psi_{22}
\end{pmatrix}.
$$

With $\boldsymbol Y=\boldsymbol B\boldsymbol X+\boldsymbol\varepsilon$ and
$\operatorname{Cov}(\boldsymbol X,\boldsymbol\varepsilon)=\boldsymbol 0$,

$$
\boldsymbol\Sigma_Y
=\boldsymbol B\boldsymbol\Sigma_X\boldsymbol B^\mathsf{T}
+\boldsymbol\Psi.
$$

For the covariance between the two outcomes,

$$
\operatorname{Cov}(Y_1,Y_2)
=\sum_r\sum_s B_{1r}\Sigma_{X,rs}B_{2s}+\Psi_{12}.
$$

Thus, for a linear APIM with exogenous predictors, every contribution combines
one coefficient into $Y_1$, one predictor variance or covariance, and one
coefficient into $Y_2$.

Here $\boldsymbol B$ is a matrix of fixed regression coefficients, not a
covariance matrix of actor and partner effects. Correlated predictors and
residuals are represented by $\boldsymbol\Sigma_X$ and $\boldsymbol\Psi$;
sampling covariances between estimated coefficients belong in the separate
inference calculation above. Random-coefficient covariances belong to Paper 2.

## Covariates and cross-predictor terms

Multiple predictors and covariates are part of Paper 1. With focal predictor
vector $\boldsymbol X$ and covariate vector $\boldsymbol Z$, write

$$
\boldsymbol Y=\boldsymbol\alpha+
\boldsymbol B\boldsymbol X+\boldsymbol\Gamma\boldsymbol Z+
\boldsymbol\varepsilon,
\qquad
\operatorname{Cov}\!\left(\boldsymbol\varepsilon,
\begin{pmatrix}\boldsymbol X\\\boldsymbol Z\end{pmatrix}\right)
=\boldsymbol 0.
$$

The full marginal covariance matrix is

$$
\begin{aligned}
\boldsymbol\Sigma_Y
={}&\boldsymbol B\boldsymbol\Sigma_{XX}\boldsymbol B^{\mathsf T}
+\boldsymbol\Gamma\boldsymbol\Sigma_{ZZ}\boldsymbol\Gamma^{\mathsf T}
+\boldsymbol\Psi\\
&+\boldsymbol B\boldsymbol\Sigma_{XZ}\boldsymbol\Gamma^{\mathsf T}
+\boldsymbol\Gamma\boldsymbol\Sigma_{ZX}\boldsymbol B^{\mathsf T}.
\end{aligned}
$$

Its $(1,2)$ element gives the outcome covariance. Report focal-predictor,
covariate, cross-block, and residual terms explicitly; correlated constructs
within either block also generate cross-construct terms. Allocating a cross
term to one named construct requires a stated convention and is not a unique
share of covariance explained by that construct. Grouping and parameterization
therefore matter. For correlation-point reporting, use the common outcome
standard deviations implied by this full model.

Covariate-adjusted slopes do not by themselves make the target conditional on
the covariates. If $\operatorname{E}(\boldsymbol\varepsilon\mid
\boldsymbol X,\boldsymbol Z)=\boldsymbol 0$, then

$$
\operatorname{Cov}(\boldsymbol Y\mid\boldsymbol Z=\boldsymbol z)
=\boldsymbol B\operatorname{Cov}(\boldsymbol X\mid
\boldsymbol Z=\boldsymbol z)\boldsymbol B^{\mathsf T}
+\operatorname{Cov}(\boldsymbol\varepsilon\mid
\boldsymbol Z=\boldsymbol z).
$$

Using a constant $\boldsymbol\Psi$ for the last term requires an additional
homoscedasticity assumption. A conditional target needs the corresponding
conditional moments and denominator; linear residualization alone is not
generally equivalent to conditioning. Keep this target distinct from the
primary marginal decomposition.

# Exchangeable dyads

Paper 1 also includes exchangeable dyads. The one-construct case below shows
how label symmetry changes the interpretation and grouping of contributions
without changing the path-tracing rule.

In an exchangeable APIM parameterization, the member labels are arbitrary.
The covariance-relevant constraints are

$$
a_1=a_2=a,
\qquad
p_1=p_2=p,
$$

$$
\operatorname{Var}(X_1)=\operatorname{Var}(X_2)=\sigma_X^2,
\qquad
\operatorname{Cov}(X_1,X_2)=\sigma_{X12}.
$$

These are only the constraints needed to simplify the covariance components.
Second-order exchangeability also requires equal role-specific predictor
means, equal outcome intercepts, and equal residual variances. A full
distributional claim additionally requires a joint distribution that is
unchanged when the two member labels are swapped.

Keeping every path-tracing component visible gives

$$
\begin{aligned}
\operatorname{Cov}(Y_1,Y_2)
={}&
\underbrace{a^2\sigma_{X12}}_{\text{actor--actor}}
+\underbrace{ap\sigma_X^2}_{\text{member 1 driven}}\\
&+\underbrace{pa\sigma_X^2}_{\text{member 2 driven}}
+\underbrace{p^2\sigma_{X12}}_{\text{partner--partner}}
+\underbrace{\psi_{12}}_{\text{residual}}.
\end{aligned}
$$

The two member-driven contributions are now necessarily equal. Because
"member 1" and "member 2" are arbitrary seats, neither component has a stable
substantive identity. They should be combined:

$$
\underbrace{ap\sigma_X^2+ap\sigma_X^2}_{\text{member-driven contribution}}
=2ap\sigma_X^2.
$$

The complete compact form is therefore

$$
\operatorname{Cov}(Y_1,Y_2)
=\underbrace{2ap\sigma_X^2}_{\text{member-driven contribution}}
+\underbrace{(a^2+p^2)\sigma_{X12}}_{\text{cross-member predictor covariance}}
+\underbrace{\psi_{12}}_{\text{residual interdependence}}.
$$

Thus, exchangeability does not change the path-tracing logic. It changes what
can be named and compared: an exchangeable APIM estimates one symmetric
member-driven contribution, not separate female-driven and male-driven
contributions. Female–male dyads remain substantively distinguishable even if
some paths are constrained to equality. Such constraints test or impose
symmetry within a distinguishable model; they do not make the member labels
exchangeable. Under the full exchangeability constraints above, equality of
the two member-driven components follows from the model structure rather than
from evidence favoring one label assignment over another.

# Manuscript planning

The current scope, argument order, inference and validation plan, software
milestones, and later-paper programme are maintained in [plan.md](plan.md).
See the [literature review](literature-review.md) for source-level evidence,
applied examples, and remaining access gaps.

# References {-}
