---
title: "Explaining Interdependence in the Actor–Partner Interdependence Model"
subtitle: "A visual paper outline and covariance-decomposition prototype"
author: "Pascal Küng"
date: "`r format(Sys.Date(), '%d %B %Y')`"
bibliography: ../../vignettes/references.bib
link-citations: true
output:
  html_document:
    math_method: mathml
    toc: true
    toc_float: true
    number_sections: true
    fig_caption: true
    self_contained: true
---

<style>
math[display="block"] {
  font-size: 1.15em;
  margin: 1.1em 0;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  dpi = 144
)

helper_candidates <- c(
  "../../vignettes/diagram-helpers.Rinc",
  "../vignettes/diagram-helpers.Rinc",
  "vignettes/diagram-helpers.Rinc"
)
helper_file <- helper_candidates[file.exists(helper_candidates)][1]

if (is.na(helper_file)) {
  stop("The shared vignette diagram helpers could not be found.")
}
sys.source(helper_file, envir = knitr::knit_global())
diagram_device <- .vignette_diagram_device()
effect_colours <- .interdep_effect_colors(c("actor", "partner"))

knitr::opts_chunk$set(
  dev = diagram_device,
  fig.ext = diagram_device,
  dev.args = list(bg = "transparent")
)

colours <- c(
  actor = unname(effect_colours[["actor"]]),
  partner = unname(effect_colours[["partner"]]),
  mixed_1 = "#3F5EAE",
  mixed_2 = "#6250A7",
  residual = "#546E7A",
  covariance = "#334155",
  ink = "#263238",
  muted = "#64748B",
  faint = "#C6CFD8",
  surface = "#F8FAFC",
  white = "#FFFFFF"
)

fade <- function(colour, alpha = 0.22) {
  grDevices::adjustcolor(colour, alpha.f = alpha)
}

component_state <- function(component) {
  all_on <- identical(component, "TOTAL")
  c(
    actor_1 = all_on || component %in% c("AA", "AP"),
    actor_2 = all_on || component %in% c("AA", "PA"),
    partner_1 = all_on || component %in% c("PA", "PP"),
    partner_2 = all_on || component %in% c("AP", "PP"),
    predictor_covariance = all_on || component %in% c("AA", "PP"),
    residual_covariance = all_on || component == "R"
  )
}

on_colour <- function(on, colour) {
  if (isTRUE(on)) colour else colours[["faint"]]
}

draw_apim_paths <- function(component = "TOTAL", panel_title = NULL,
                            show_formula = FALSE, newpage = TRUE,
                            text_scale = 1) {
  component <- match.arg(component, c("TOTAL", "AA", "AP", "PA", "PP", "R"))
  state <- component_state(component)
  native <- function(value) grid::unit(value, "native")
  arrow_head <- grid::arrow(
    angle = 28, length = grid::unit(0.13 * text_scale, "inches"),
    type = "closed"
  )
  covariance_arrow <- grid::arrow(
    angle = 26, length = grid::unit(0.09 * text_scale, "inches"),
    ends = "both", type = "closed"
  )

  # Reuse the vignette palette, math labels, annotations, and rounded-node
  # layout. This paper-specific adaptation adds predictor covariance and
  # route highlighting without changing the vignette diagrams.
  draw_node <- function(x, y, label, active = TRUE, residual = FALSE) {
    node_colour <- on_colour(active, colours[["ink"]])
    node_gp <- grid::gpar(
      fill = if (residual) "white" else colours[["surface"]],
      col = node_colour, lwd = 1.8 * text_scale
    )
    if (residual) {
      grid::grid.circle(
        x = native(x), y = native(y), r = grid::unit(0.055, "snpc"),
        gp = node_gp
      )
    } else {
      grid::grid.roundrect(
        x = native(x), y = native(y),
        width = native(0.17), height = native(0.19),
        r = grid::unit(0.04, "snpc"), gp = node_gp
      )
    }
    grid::grid.text(
      .diagram_math_label(label), x = native(x), y = native(y),
      gp = grid::gpar(
        col = node_colour, fontsize = (if (residual) 17 else 21) * text_scale
      )
    )
  }

  draw_path <- function(x0, y0, x1, y1, active, colour) {
    path_colour <- on_colour(active, colour)
    grid::grid.segments(
      x0 = native(x0), y0 = native(y0),
      x1 = native(x1), y1 = native(y1), arrow = arrow_head,
      gp = grid::gpar(
        col = path_colour, fill = path_colour,
        lwd = if (active) 2.4 * text_scale else 1.2 * text_scale
      )
    )
  }

  path_label <- function(label, x, y, active, colour) {
    .draw_residual_annotation(
      label, x, y, fontsize = 15 * text_scale,
      colour = on_colour(active, colour)
    )
  }

  if (isTRUE(newpage)) grid::grid.newpage()
  grid::pushViewport(grid::viewport(xscale = c(0, 1), yscale = c(0, 1)))
  on.exit(grid::popViewport(), add = TRUE)

  if (!is.null(panel_title)) {
    grid::grid.text(
      panel_title, x = native(0.06), y = native(0.91), just = "left",
      gp = grid::gpar(
        col = colours[["ink"]], fontsize = 16 * text_scale, fontface = "bold"
      )
    )
  } else {
    grid::grid.text(
      c("Predictor", "Outcome", "Residual"),
      x = native(c(0.24, 0.65, 0.875)), y = native(0.89),
      gp = grid::gpar(col = colours[["muted"]], fontsize = 11 * text_scale)
    )
  }

  # Predictors on the left; actor and partner paths flow into outcomes on the
  # right. Partner subscripts always identify the outcome member.
  draw_path(0.325, 0.70, 0.565 - .diagram_arrow_clearance, 0.70,
            state[["actor_1"]], colours[["actor"]])
  draw_path(0.325, 0.30, 0.565 - .diagram_arrow_clearance, 0.30,
            state[["actor_2"]], colours[["actor"]])
  draw_path(0.325, 0.67, 0.565 - .diagram_arrow_clearance, 0.34,
            state[["partner_2"]], colours[["partner"]])
  draw_path(0.325, 0.33, 0.565 - .diagram_arrow_clearance, 0.66,
            state[["partner_1"]], colours[["partner"]])

  predictor_colour <- on_colour(
    state[["predictor_covariance"]], colours[["covariance"]]
  )
  grid::grid.curve(
    x1 = native(0.153), y1 = native(0.67),
    x2 = native(0.153), y2 = native(0.33),
    curvature = 0.32, angle = 90, ncp = 12, square = FALSE,
    arrow = covariance_arrow,
    gp = grid::gpar(
      col = predictor_colour, fill = predictor_colour, lwd = 1.8 * text_scale
    )
  )
  path_label(expression(sigma[X[12]]), 0.11, 0.50,
             state[["predictor_covariance"]], colours[["covariance"]])

  # Residual arrows still point into Y. The double-headed curve is covariance
  # psi_12, not the residual correlation shown in fitted vignette diagrams.
  residual_colour <- on_colour(
    state[["residual_covariance"]], colours[["residual"]]
  )
  for (y in c(0.70, 0.30)) {
    draw_path(0.840, y, 0.735 + .diagram_arrow_clearance, y,
              state[["residual_covariance"]], colours[["residual"]])
  }
  grid::grid.curve(
    x1 = native(0.905), y1 = native(0.66),
    x2 = native(0.905), y2 = native(0.34),
    curvature = -0.28, angle = 90, ncp = 12, square = FALSE,
    arrow = covariance_arrow,
    gp = grid::gpar(
      col = residual_colour, fill = residual_colour, lwd = 1.8 * text_scale
    )
  )
  path_label(expression(italic("\u03c8")[12]), 0.92, 0.50,
             state[["residual_covariance"]], colours[["residual"]])

  draw_node(0.24, 0.70, expression(X[1]),
            component %in% c("TOTAL", "AA", "AP", "PP"))
  draw_node(0.24, 0.30, expression(X[2]),
            component %in% c("TOTAL", "AA", "PA", "PP"))
  draw_node(0.65, 0.70, expression(Y[1]))
  draw_node(0.65, 0.30, expression(Y[2]))
  draw_node(0.875, 0.70, expression(epsilon[1]),
            state[["residual_covariance"]], residual = TRUE)
  draw_node(0.875, 0.30, expression(epsilon[2]),
            state[["residual_covariance"]], residual = TRUE)

  path_label(expression(a[1]), 0.44, 0.70,
             state[["actor_1"]], colours[["actor"]])
  path_label(expression(a[2]), 0.44, 0.30,
             state[["actor_2"]], colours[["actor"]])
  path_label(expression(p[2]), 0.40, 0.56,
             state[["partner_2"]], colours[["partner"]])
  path_label(expression(p[1]), 0.40, 0.44,
             state[["partner_1"]], colours[["partner"]])

  if (isTRUE(show_formula)) {
    formulas <- list(
      AA = expression(italic(C)[plain(AA)] == a[1] * a[2] * sigma[X[12]]),
      AP = expression(italic(C)[plain(AP)] == a[1] * p[2] * sigma[X[1]]^2),
      PA = expression(italic(C)[plain(PA)] == p[1] * a[2] * sigma[X[2]]^2),
      PP = expression(italic(C)[plain(PP)] == p[1] * p[2] * sigma[X[12]]),
      R = expression(italic(C)[plain(R)] == italic("\u03c8")[12]),
      TOTAL = expression(sum(italic(C)[italic(k)], italic(k)) ==
                           Cov(Y[1], Y[2]))
    )
    grid::grid.text(
      .diagram_math_label(formulas[[component]]),
      x = native(0.50), y = native(0.08),
      gp = grid::gpar(col = colours[["ink"]], fontsize = 15 * text_scale)
    )
  } else {
    .draw_diagram_legend(
      labels = c("Actor path", "Partner path"),
      colours = unname(effect_colours), y = 0.08,
      segment_starts = c(0.26, 0.53), text_x = c(0.40, 0.68),
      fontsize = 11 * text_scale
    )
  }
  invisible(NULL)
}

component_colours <- c(
  "Actor–actor" = colours[["actor"]],
  "Actor–partner" = colours[["mixed_1"]],
  "Partner–actor" = colours[["mixed_2"]],
  "Partner–partner" = colours[["partner"]],
  "Residual" = colours[["residual"]]
)
```

> **Draft status (31 August 2026).** The [literature review](literature-review.md)
> records direct APIM precedents found after this outline was drafted. Its
> evidence and caveats take precedence over the novelty framing below. The
> inference, simulation, and extension plans remain proposals for discussion.

> **Working claim.** An APIM should not only *account for* interdependence. It
> can also provide a model-based, descriptive explanation of where the
> covariance between partners' outcomes comes from.

Here and throughout, *explain* means attribute statistically within the fitted
model. It does not mean that the routes are automatically causal mechanisms.

This document first develops the argument as it could appear in the paper and
checks the underlying path tracing visually and numerically. The final section
collects the proposed contribution, study plan, manuscript structure, open
questions, and next steps.

# The idea in one paragraph

The APIM is commonly introduced through actor effects, partner effects, and a
residual correlation [@kenny2006dyadic]. Work on dyadic patterns and compound
APIM processes has shown the value of turning individual coefficients into
interpretable derived quantities [@kennyDetectingMeasuringTesting2010;
@ledermannAssessingMediationDyadic2011]. That presentation nevertheless does
not by itself answer a natural dyadic question:

> Why are the two partners' outcomes correlated, and how much of that
> interdependence is associated with each modeled route?

Dyadic dependence can itself contain substantive information that is lost when
members are analyzed separately [@laurenceauUsingDiaryMethods2005]. Bolger and
Shrout [-@bolgerAccountingStatisticalDependency2007] treated covariance
structure as a starting point for studying shared events and interpersonal
influence. The present proposal asks a narrower question: within one fitted
cross-sectional APIM, which modeled routes carry the outcome covariance?

For one predictor construct measured on both partners, established path-tracing
and covariance-component algebra gives an exact additive answer under the usual
APIM assumption that predictors are uncorrelated with the outcome residuals
[@bokerAlgorithmHierarchicalOrganization2002]. Outcome covariance can travel
through four predictor-side path pairs: actor–actor, actor–partner,
partner–actor, and partner–partner. A fifth component is the remaining residual
covariance. These five signed quantities sum exactly to the outcome covariance
implied by the model. Dividing every component by the same two model-implied
outcome standard deviations makes them sum to the model-implied outcome
correlation. Related signed path-weight decompositions on covariance and
correlation scales exist for undirected Gaussian graphical models
[@jonesCovarianceDecompositionUndirected2005]; the proposed contribution is the
APIM-specific estimand and workflow, not a new general tracing rule.

# The complete one-construct APIM

For distinguishable members 1 and 2, write

$$
\begin{aligned}
Y_1&=a_1X_1+p_1X_2+\varepsilon_1,\\
Y_2&=a_2X_2+p_2X_1+\varepsilon_2.
\end{aligned}
$$

Intercepts are omitted because they do not affect variances or covariances.
The subscript on an actor or partner effect identifies the **outcome member**:
$p_1$ is the effect of member 2's predictor on member 1's outcome; $p_2$ is the
effect of member 1's predictor on member 2's outcome.

Unless stated otherwise, $a_1,a_2,p_1,p_2$ are unstandardized coefficients and
all predictor moments are computed on the matching original scales. A fully
standardized presentation is possible, but coefficients, predictor moments,
residual covariance, and the outcome denominator must then be transformed
consistently.

```{r complete-apim, echo=FALSE, fig.width=9, fig.height=4.8, out.width="100%", fig.cap="Complete distinguishable APIM used for the decomposition, using the vignette's left-to-right layout and blue/purple path palette. Predictors are on the left and outcomes on the right. Curved double-headed paths show predictor covariance on the left and residual outcome covariance on the right; residual arrows point into the outcomes.", fig.alt="Left-to-right APIM with X1 above X2 on the left and Y1 above Y2 on the right. Blue actor paths are horizontal; purple partner paths cross diagonally. A curved path connects the predictors, and another connects the two residuals, which point into the outcomes."}
draw_apim_paths("TOTAL")
```

## Assumptions and what "exact" means

The proposed paper concerns independent dyads, continuous outcomes, and a
linear, fixed-slope, cross-sectional APIM. The variables are viewed as random
over a target population of dyads, so the result below decomposes **marginal
across-dyad covariance**. Intercepts are omitted and all variances and
covariances are taken around the appropriate role-specific means.

Assume finite second moments and

$$
\operatorname{E}(\boldsymbol\varepsilon\mid X_1,X_2)=\boldsymbol 0,
\qquad
\operatorname{Cov}(\boldsymbol\varepsilon\mid X_1,X_2)
=\boldsymbol\Psi,
$$

which implies that each predictor is uncorrelated with each outcome residual:

$$
\operatorname{Cov}(X_j,\varepsilon_k)=0
\qquad\text{for }j,k\in\{1,2\}.
$$

This is the usual exogeneity or residual-orthogonality assumption in the APIM.
It is part of the statistical model, not an identity that holds for any four
variables. Normality is not required for the covariance algebra. If
predictor–residual covariances are allowed, four additional terms enter the
outcome covariance and the five-part decomposition is incomplete.

Conditional on the observed predictors, the fixed-slope model implies
$\operatorname{Cov}(Y_1,Y_2\mid X_1,X_2)=\psi_{12}$. The predictor-side routes
appear when moving to marginal covariance across dyads whose predictor values
vary. Random slopes, nonlinear links, and longitudinal carryover paths create
additional terms and are outside the first paper's scope.

The derivation itself is short. By the linearity of covariance,

$$
\begin{aligned}
\operatorname{Cov}(Y_1,Y_2)
={}&\operatorname{Cov}\!\left(
a_1X_1+p_1X_2+\varepsilon_1,
a_2X_2+p_2X_1+\varepsilon_2
\right).
\end{aligned}
$$

Expanding this expression and applying the assumption above gives

$$
\begin{aligned}
\operatorname{Cov}(Y_1,Y_2)
={}&
\underbrace{a_1a_2\operatorname{Cov}(X_1,X_2)}_{\text{actor--actor}}
+\underbrace{a_1p_2\operatorname{Var}(X_1)}_{\text{actor--partner}}\\
&+\underbrace{p_1a_2\operatorname{Var}(X_2)}_{\text{partner--actor}}
+\underbrace{p_1p_2\operatorname{Cov}(X_1,X_2)}_{\text{partner--partner}}
+\underbrace{\operatorname{Cov}(\varepsilon_1,\varepsilon_2)}_{\text{residual}}.
\end{aligned}
$$

The four predictor–residual terms that disappear are
$a_1\operatorname{Cov}(X_1,\varepsilon_2)$,
$p_1\operatorname{Cov}(X_2,\varepsilon_2)$,
$a_2\operatorname{Cov}(X_2,\varepsilon_1)$, and
$p_2\operatorname{Cov}(X_1,\varepsilon_1)$. No more elaborate proof is needed:
the five remaining terms are precisely the five covariance-carrying routes in
the diagram.

# The five covariance-carrying routes

```{r path-components, echo=FALSE, fig.width=12, fig.height=9.9, out.width="100%", fig.cap="The five covariance-carrying routes, arranged in three rows of two left-to-right APIMs. Pale paths remain part of the model but do not belong to the highlighted component. The member-driven routes use one predictor's variance, not the covariance between predictors. The final panel restores all paths.", fig.alt="Six left-to-right APIM panels in three rows of two. In reading order, they highlight actor-actor, actor-partner from X1, partner-actor from X2, partner-partner, residual covariance, and all five components. Each panel includes its component formula."}
components <- c("AA", "AP", "PA", "PP", "R", "TOTAL")
panel_titles <- c(
  "1. Actor–actor", "2. Actor–partner",
  "3. Partner–actor", "4. Partner–partner",
  "5. Residual", "Their exact sum"
)
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(3, 2)))
for (index in seq_along(components)) {
  grid::pushViewport(grid::viewport(
    layout.pos.row = (index - 1L) %/% 2L + 1L,
    layout.pos.col = (index - 1L) %% 2L + 1L
  ))
  draw_apim_paths(
    components[[index]], panel_titles[[index]], show_formula = TRUE,
    newpage = FALSE, text_scale = 0.95
  )
  grid::popViewport()
}
grid::popViewport()
```

In the diagrams, $\sigma_{X_{12}}=\operatorname{Cov}(X_1,X_2)$,
$\sigma_{X_j}^2=\operatorname{Var}(X_j)$, and
$\psi_{12}=\operatorname{Cov}(\varepsilon_1,\varepsilon_2)$.

The four predictor-side combinations form a useful $2\times2$ table:

| Path into $Y_1$ | Path into $Y_2$ | Shared predictor quantity | Contribution |
|:--|:--|:--|:--|
| actor $a_1$ | actor $a_2$ | $\operatorname{Cov}(X_1,X_2)$ | $C_{\mathrm{AA}}=a_1a_2\operatorname{Cov}(X_1,X_2)$ |
| actor $a_1$ | partner $p_2$ | $\operatorname{Var}(X_1)$ | $C_{\mathrm{AP}}=a_1p_2\operatorname{Var}(X_1)$ |
| partner $p_1$ | actor $a_2$ | $\operatorname{Var}(X_2)$ | $C_{\mathrm{PA}}=p_1a_2\operatorname{Var}(X_2)$ |
| partner $p_1$ | partner $p_2$ | $\operatorname{Cov}(X_1,X_2)$ | $C_{\mathrm{PP}}=p_1p_2\operatorname{Cov}(X_1,X_2)$ |

The residual component is

$$
C_{\mathrm{R}}=\operatorname{Cov}(\varepsilon_1,\varepsilon_2).
$$

Therefore,

$$
\operatorname{Cov}(Y_1,Y_2)
=C_{\mathrm{AA}}+C_{\mathrm{AP}}+C_{\mathrm{PA}}+C_{\mathrm{PP}}+C_{\mathrm{R}}.
$$

## Plain-language interpretation

- **Actor–actor:** partners resemble each other on $X$, and each person's own
  $X$ predicts their own $Y$.
- **Actor–partner:** member 1's $X$ reaches both outcomes—through an actor path
  into $Y_1$ and a partner path into $Y_2$.
- **Partner–actor:** member 2's $X$ reaches both outcomes in the analogous way.
- **Partner–partner:** partners resemble each other on $X$, and each predictor
  crosses to the other member's outcome.
- **Residual:** outcome interdependence that remains after the modeled
  predictor routes.

## Female-driven and male-driven interdependence

For a female--male dyad, let member 1 be the female partner and member 2 the
male partner. The two routes carried by a single member's predictor variance
then have an especially intuitive interpretation:

$$
\underbrace{a_1p_2\operatorname{Var}(X_1)}_{\text{female-driven}}
\qquad\text{and}\qquad
\underbrace{p_1a_2\operatorname{Var}(X_2)}_{\text{male-driven}}.
$$

The first is female-driven because the female partner's predictor reaches both
outcomes: through her actor path into her own outcome and through her partner
path into the male partner's outcome. The second is male-driven in the
corresponding way. Here, "driven" describes a model-implied covariance route;
it is causal only when the design and assumptions support a causal
interpretation.

This yields two broader substantive families:

$$
\underbrace{C_{\mathrm{AP}}+C_{\mathrm{PA}}}_{\text{predictor-transmission routes}}
\qquad\text{and}\qquad
\underbrace{C_{\mathrm{AA}}+C_{\mathrm{PP}}}_{\text{cross-member predictor-covariance routes}}.
$$

"Predictor transmission" is an algebraic route label, not by itself a claim
that one partner causally transmits something to the other. The second family
can be called **predictor similarity** when
$\operatorname{Cov}(X_1,X_2)>0$; a negative predictor covariance instead
represents dissimilarity.

The phrase "actor-driven covariance" by itself is potentially misleading. One
person's predictor variance creates covariance between the two outcomes only
when that predictor reaches **both** outcomes—once through an actor path and
once through a partner path. Actor and partner paths work in pairs.

# From covariance to correlation points

The five components above are in the original units of $Y_1$ times the units
of $Y_2$. That scale is exact but often difficult to interpret. To express each
component on the familiar correlation scale, define the model-implied
denominator

$$
s_Y=\operatorname{SD}(Y_1)\operatorname{SD}(Y_2).
$$

For every component $C_k$, define a correlation-point contribution

$$
c_k=\frac{C_k}{s_Y}.
$$

These are signed additive contributions on the outcome-correlation scale. An
individual $c_k$ is not itself a correlation, a proportion, or a variance
share, and it need not lie between $-1$ and $1$. Percentages are especially
misleading when positive and negative routes cancel: they can be negative,
exceed 100%, or become unstable when the total covariance is near zero.

Because every component uses the same denominator,

$$
\rho^{\mathrm{model}}_{Y_1Y_2}
=c_{\mathrm{AA}}+c_{\mathrm{AP}}+c_{\mathrm{PA}}+c_{\mathrm{PP}}+c_{\mathrm{R}}.
$$

The required total outcome variances are

$$
\begin{aligned}
\operatorname{Var}(Y_1)
={}&a_1^2\operatorname{Var}(X_1)
+p_1^2\operatorname{Var}(X_2)
+2a_1p_1\operatorname{Cov}(X_1,X_2)
+\operatorname{Var}(\varepsilon_1),\\
\operatorname{Var}(Y_2)
={}&a_2^2\operatorname{Var}(X_2)
+p_2^2\operatorname{Var}(X_1)
+2a_2p_2\operatorname{Cov}(X_1,X_2)
+\operatorname{Var}(\varepsilon_2).
\end{aligned}
$$

The observed sample correlation, $r_{Y_1Y_2}$, should be reported alongside
the model-implied correlation as a descriptive reference. They are conceptually
different quantities and are equal only when the fitted model reproduces the
relevant sample moments. The five components should therefore not be rescaled
afterward merely to force them to sum to the observed correlation.

> **Important distinction.** The residual correlation
> $\operatorname{Cor}(\varepsilon_1,\varepsilon_2)$ divides residual covariance
> by the two **residual** standard deviations. The residual contribution $c_{\mathrm{R}}$
> divides the same covariance by the two **total model-implied outcome** standard
> deviations. Only $c_{\mathrm{R}}$ is on the same scale as the other pieces of
> the model-implied outcome correlation.

# Worked numerical example

Suppose

$$
\operatorname{Var}(X_1)=\operatorname{Var}(X_2)=1,
\qquad
\operatorname{Cov}(X_1,X_2)=0.40,
$$

$$
a_1=0.50,\quad a_2=0.40,\quad p_1=0.20,\quad p_2=0.30,
$$

and

$$
\boldsymbol\Psi=
\begin{pmatrix}
0.50 & 0.15\\
0.15 & 0.55
\end{pmatrix}.
$$

```{r calculate-example, include=FALSE}
example_parameters <- list(
  var_x1 = 1,
  var_x2 = 1,
  cov_x12 = 0.40,
  a1 = 0.50,
  a2 = 0.40,
  p1 = 0.20,
  p2 = 0.30,
  psi11 = 0.50,
  psi22 = 0.55,
  psi12 = 0.15
)

with(example_parameters, {
  example_components <<- c(
    "Actor–actor" = a1 * a2 * cov_x12,
    "Actor–partner" = a1 * p2 * var_x1,
    "Partner–actor" = p1 * a2 * var_x2,
    "Partner–partner" = p1 * p2 * cov_x12,
    "Residual" = psi12
  )

  example_var_y1 <<-
    a1^2 * var_x1 + p1^2 * var_x2 +
    2 * a1 * p1 * cov_x12 + psi11
  example_var_y2 <<-
    a2^2 * var_x2 + p2^2 * var_x1 +
    2 * a2 * p2 * cov_x12 + psi22
})

example_cov_y12 <- sum(example_components)
example_denominator <- sqrt(example_var_y1 * example_var_y2)
example_correlation_components <- example_components / example_denominator
example_rho_y12 <- sum(example_correlation_components)
example_residual_rho <- with(
  example_parameters,
  psi12 / sqrt(psi11 * psi22)
)

example_table <- data.frame(
  Route = names(example_components),
  `Covariance contribution` = unname(example_components),
  `Correlation-point contribution` = unname(example_correlation_components),
  check.names = FALSE
)
```

```{r example-table, echo=FALSE}
knitr::kable(
  example_table,
  digits = 3,
  align = c("l", "r", "r")
)
```

```{r example-contributions, echo=FALSE, fig.width=9.4, fig.height=5.4, out.width="92%", fig.cap="Waterfall decomposition of the model-implied outcome correlation. Each route starts where the preceding route ended; the five signed increments arrive exactly at the model-implied total.", fig.alt="Waterfall chart in which actor-actor, actor-partner, partner-actor, partner-partner, and residual correlation-point contributions accumulate to the model-implied outcome correlation."}
old_par <- graphics::par(mar = c(7.0, 4.8, 1.2, 1.0), xpd = NA)
increments <- unname(example_correlation_components)
starts <- c(0, head(cumsum(increments), -1))
ends <- cumsum(increments)
x <- seq_along(increments)

graphics::plot(
  NA,
  xlim = c(0.45, length(increments) + 0.75),
  ylim = c(0, example_rho_y12 * 1.16),
  xaxt = "n",
  xlab = "",
  ylab = "Outcome-correlation points",
  bty = "l"
)

for (index in seq_along(increments)) {
  graphics::rect(
    x[index] - 0.34,
    min(starts[index], ends[index]),
    x[index] + 0.34,
    max(starts[index], ends[index]),
    col = component_colours[index],
    border = NA
  )
  if (index < length(increments)) {
    graphics::segments(
      x[index] + 0.34, ends[index],
      x[index + 1] - 0.34, ends[index],
      col = colours[["muted"]], lty = 3, lwd = 1.4
    )
  }
  graphics::text(
    x[index], ends[index] + example_rho_y12 * 0.025,
    sprintf("%+.3f", increments[index]),
    cex = 0.82, col = colours[["ink"]]
  )
}

graphics::axis(
  1,
  at = x,
  labels = names(example_correlation_components),
  las = 2,
  tick = FALSE,
  line = -0.3,
  cex.axis = 0.85
)
graphics::abline(h = example_rho_y12, lty = 2, lwd = 2.1,
                 col = colours[["covariance"]])
graphics::text(
  0.55, example_rho_y12 + example_rho_y12 * 0.025,
  sprintf("total ρ = %.3f", example_rho_y12),
  pos = 4, offset = 0.25, cex = 0.92,
  col = colours[["covariance"]], font = 2
)
graphics::par(old_par)
```

Here,

$$
\operatorname{Cov}(Y_1,Y_2)=`r sprintf('%.3f', example_cov_y12)`,
\qquad
\rho^{\mathrm{model}}_{Y_1Y_2}=`r sprintf('%.3f', example_rho_y12)`.
$$

The residual correlation is
`r sprintf('%.3f', example_residual_rho)`, but its contribution to the total
outcome correlation is only
`r sprintf('%.3f', unname(example_correlation_components[['Residual']]))`.
These numbers answer different questions.

# Same interdependence, different route profiles

The decomposition is useful because the same total partner correlation can
hide very different model-implied route profiles. The second hypothetical
model below is chosen to have the **same** outcome variances and covariance as the first model, but
stronger actor paths, weaker partner paths, and more residual covariance.

```{r same-total-calculation, include=FALSE}
scenario_a <- example_components

scenario_b_parameters <- list(
  var_x1 = 1,
  var_x2 = 1,
  cov_x12 = 0.40,
  a1 = 0.70,
  a2 = 0.60,
  p1 = 0.05,
  p2 = 0.05,
  psi12 = 0.25
)

with(scenario_b_parameters, {
  scenario_b <<- c(
    "Actor–actor" = a1 * a2 * cov_x12,
    "Actor–partner" = a1 * p2 * var_x1,
    "Partner–actor" = p1 * a2 * var_x2,
    "Partner–partner" = p1 * p2 * cov_x12,
    "Residual" = psi12
  )
})

# The residual variances in scenario B can be set to preserve both total
# outcome variances from scenario A; they are positive and Psi remains valid.
scenario_b_psi11 <- example_var_y1 - with(
  scenario_b_parameters,
  a1^2 * var_x1 + p1^2 * var_x2 + 2 * a1 * p1 * cov_x12
)
scenario_b_psi22 <- example_var_y2 - with(
  scenario_b_parameters,
  a2^2 * var_x2 + p2^2 * var_x1 + 2 * a2 * p2 * cov_x12
)

stopifnot(
  isTRUE(all.equal(sum(scenario_a), sum(scenario_b))),
  scenario_b_psi11 > 0,
  scenario_b_psi22 > 0,
  scenario_b_psi11 * scenario_b_psi22 >
    scenario_b_parameters$psi12^2
)

scenario_matrix <- rbind(
  "A: stronger transmission" = scenario_a,
  "B: similarity + residual" = scenario_b
) / example_denominator
```

```{r same-total-figure, echo=FALSE, fig.width=10, fig.height=5.2, out.width="96%", fig.cap="Two APIMs with the same model-implied outcome correlation but different route profiles. Each horizontal stack has exactly the same length.", fig.alt="Two horizontal stacked bars of identical total length. Scenario A has larger actor-partner and partner-actor sections; scenario B has larger actor-actor and residual sections."}
old_par <- graphics::par(mar = c(7.2, 10.2, 1.0, 1.0), xpd = NA)
graphics::barplot(
  t(scenario_matrix),
  horiz = TRUE,
  beside = FALSE,
  col = component_colours,
  border = colours[["white"]],
  names.arg = rownames(scenario_matrix),
  las = 1,
  xlim = c(0, example_rho_y12 * 1.04),
  xlab = "Outcome-correlation points",
  cex.names = 0.9
)
graphics::legend(
  "bottom", inset = c(0, -0.36),
  legend = names(component_colours),
  fill = component_colours, border = NA,
  horiz = TRUE, bty = "n", cex = 0.78, xpd = NA
)
graphics::par(old_par)
```

This is the paper's most intuitive motivation: reporting only the total or
residual correlation cannot distinguish these model-implied route profiles.

# From a fitted APIM to estimated contributions

The formulas define population or model-implied quantities. In an empirical
analysis, estimate them by plugging the fitted APIM coefficients and predictor
moments into the same expressions. Let
$\widehat\sigma_{X1}^2$, $\widehat\sigma_{X2}^2$, and
$\widehat\sigma_{X12}$ denote the estimated predictor variances and covariance,
and let $\widehat\psi_{12}$ denote the fitted residual covariance. Then

$$
\begin{aligned}
\widehat C_{\mathrm{AA}}
&=\widehat a_1\widehat a_2\widehat\sigma_{X12}, &
\widehat C_{\mathrm{AP}}
&=\widehat a_1\widehat p_2\widehat\sigma_{X1}^2,\\
\widehat C_{\mathrm{PA}}
&=\widehat p_1\widehat a_2\widehat\sigma_{X2}^2, &
\widehat C_{\mathrm{PP}}
&=\widehat p_1\widehat p_2\widehat\sigma_{X12},\\
\widehat C_{\mathrm{R}}
&=\widehat\psi_{12}.&&
\end{aligned}
$$

The estimated model-implied outcome variances supply the denominator for the
correlation-point contributions. For example,

$$
\widehat c_{\mathrm{AP}}
=\frac{\widehat C_{\mathrm{AP}}}
{\sqrt{\widehat{\operatorname{Var}}(Y_1)
       \widehat{\operatorname{Var}}(Y_2)}}.
$$

In an SEM, predictor moments may be estimated as part of the joint model or
supplied from the observed exogenous variables, depending on how the model is
specified. In a multilevel or regression implementation that conditions on the
predictors, use their empirical dyad-level variance–covariance matrix. In both
cases, the reported decomposition should state where those moments came from.

Likewise, $\widehat\psi_{12}$ must represent the **total model-implied
covariance remaining between the two members** after the fixed predictor paths.
Depending on the mixed-model parameterization, that quantity may be represented
by a shared dyad effect, a covariance between member-specific random effects,
a residual covariance, or a sum of such components. It should not be replaced
uncritically by whichever single variance component is easiest to extract.

## Uncertainty without complicated mathematics

A dyad bootstrap provides a direct uncertainty procedure and avoids deriving a
large delta-method covariance matrix:

1. Resample whole dyads with replacement, keeping both partners together and
   preserving their role labels.
2. Refit the APIM in the bootstrap sample.
3. Recompute the predictor moments, all five covariance components, the
   model-implied outcome variances, and the correlation-point contributions.
4. Repeat and form intervals from the bootstrap distribution of each signed
   component.

Resampling whole dyads preserves the dependence that the model is designed to
represent. Recomputing both the fitted paths and predictor moments carries both
sources of sampling uncertainty forward. The analysis plan should prespecify
the interval type and how nonconverged bootstrap fits are handled.

Within the fitted predictor model, the exact identity is

$$
\Sigma^{\mathrm{model}}_{Y,12}-\Psi_{12}
=C_{\mathrm{AA}}+C_{\mathrm{AP}}+C_{\mathrm{PA}}+C_{\mathrm{PP}}.
$$

An intercept-only model attributes all model-implied outcome covariance to its
residual covariance. However, the reduction in residual covariance between two
separately fitted models equals the predictor-side sum above only when both
models reproduce the same marginal outcome covariance under compatible
estimation and sample-moment conventions. The within-model identity is the
safer primary result.

# Matrix form and extension to several predictors

The scalar derivation is easiest to teach, but the same result can be written
compactly in matrix form. For the one-construct APIM, let

$$
\boldsymbol Y=
\begin{pmatrix}Y_1\\Y_2\end{pmatrix},
\qquad
\boldsymbol X=
\begin{pmatrix}X_1\\X_2\end{pmatrix},
\qquad
\boldsymbol B=
\begin{pmatrix}
a_1 & p_1\\
p_2 & a_2
\end{pmatrix},
$$

and

$$
\boldsymbol\Sigma_X=
\begin{pmatrix}
\sigma_{X1}^2 & \sigma_{X12}\\
\sigma_{X12} & \sigma_{X2}^2
\end{pmatrix},
\qquad
\boldsymbol\Psi=
\begin{pmatrix}
\psi_{11} & \psi_{12}\\
\psi_{12} & \psi_{22}
\end{pmatrix}.
$$

With $\boldsymbol Y=\boldsymbol B\boldsymbol X+\boldsymbol\varepsilon$ and
$\operatorname{Cov}(\boldsymbol X,\boldsymbol\varepsilon)=\boldsymbol 0$,

$$
\boldsymbol\Sigma_Y
=\boldsymbol B\boldsymbol\Sigma_X\boldsymbol B^\mathsf{T}
+\boldsymbol\Psi.
$$

For the covariance between the two outcomes,

$$
\operatorname{Cov}(Y_1,Y_2)
=\sum_r\sum_s B_{1r}\Sigma_{X,rs}B_{2s}+\Psi_{12}.
$$

Thus, for a linear APIM with exogenous predictors, every contribution combines
one coefficient into $Y_1$, one predictor variance or covariance, and one
coefficient into $Y_2$.

However, several predictors introduce cross-predictor covariance terms. Their
allocation to a named construct may depend on how predictors are grouped or
parameterized. The first paper should therefore make the single dyadic
predictor pair the transparent core result, then present the matrix expression
and a prespecified grouping rule as the extension.

# Brief closing extension: exchangeable dyads

The paper's main case is cross-sectional and distinguishable: member 1 is the
female partner and member 2 is the male partner, so the female-driven and
male-driven components can differ. A short closing section can show that the
same logic also applies to exchangeable dyads without turning exchangeability
into a second focus of the paper.

For exchangeable dyads, the member labels are arbitrary. The
covariance-relevant constraints are

$$
a_1=a_2=a,
\qquad
p_1=p_2=p,
$$

$$
\operatorname{Var}(X_1)=\operatorname{Var}(X_2)=\sigma_X^2,
\qquad
\operatorname{Cov}(X_1,X_2)=\sigma_{X12}.
$$

These are only the constraints needed to simplify the covariance components.
Second-order exchangeability also requires equal role-specific predictor
means, equal outcome intercepts, and equal residual variances. A full
distributional claim additionally requires a joint distribution that is
unchanged when the two member labels are swapped.

Keeping every path-tracing component visible gives

$$
\begin{aligned}
\operatorname{Cov}(Y_1,Y_2)
={}&
\underbrace{a^2\sigma_{X12}}_{\text{actor--actor}}
+\underbrace{ap\sigma_X^2}_{\text{member 1 driven}}\\
&+\underbrace{pa\sigma_X^2}_{\text{member 2 driven}}
+\underbrace{p^2\sigma_{X12}}_{\text{partner--partner}}
+\underbrace{\psi_{12}}_{\text{residual}}.
\end{aligned}
$$

The two member-driven contributions are now necessarily equal. Because
"member 1" and "member 2" are arbitrary seats, neither component has a stable
substantive identity. They should be combined:

$$
\underbrace{ap\sigma_X^2+ap\sigma_X^2}_{\text{member-driven transmission}}
=2ap\sigma_X^2.
$$

The complete compact form is therefore

$$
\operatorname{Cov}(Y_1,Y_2)
=\underbrace{2ap\sigma_X^2}_{\text{member-driven transmission}}
+\underbrace{(a^2+p^2)\sigma_{X12}}_{\text{cross-member predictor covariance}}
+\underbrace{\psi_{12}}_{\text{residual interdependence}}.
$$

Thus, exchangeability does not change the path-tracing logic. It changes what
can be named and compared: an exchangeable APIM estimates one symmetric
member-driven contribution, not separate female-driven and male-driven
contributions. Female–male dyads remain substantively distinguishable even if
some paths are constrained to equality. Such constraints test or impose
symmetry within a distinguishable model; they do not make the member labels
exchangeable. Under the full exchangeability constraints above, equality of
the two member-driven components follows from the model structure rather than
from evidence favoring one label assignment over another.

# Plan for the paper

The sections above form a teaching-oriented prototype of the paper's argument.
This final section records the intended scope, contribution, evidence, and
remaining work.

## Working scope

The first paper should focus on:

- cross-sectional, independent, distinguishable dyads;
- female and male partners as the motivating distinguishable roles;
- one predictor construct measured on both partners;
- continuous outcomes and a linear, fixed-slope APIM;
- marginal, model-implied outcome covariance and correlation;
- descriptive statistical attribution rather than automatic causal claims.

Several predictors and exchangeable dyads can appear as compact extensions.
Longitudinal paths, random slopes, nonlinear outcome models, and formal causal
identification should remain outside the first paper.

## Central claim and contributions

The central claim is that dyadic interdependence need not be treated only as a
nuisance to accommodate. Within a fitted APIM, the model-implied covariance
between partners' outcomes can be decomposed into interpretable, signed route
contributions.

The paper should contribute the following package:

1. Present the exact five-part decomposition for a distinguishable APIM in one
   diagram, one table, and a short covariance expansion.
2. Name and interpret the female-driven and male-driven components while
   retaining the complete actor–actor, actor–partner, partner–actor,
   partner–partner, and residual breakdown.
3. Distinguish predictor-transmission routes from cross-member
   predictor-covariance routes, without treating either label as automatically
   causal.
4. Put all five components on a common signed correlation-point scale and show
   why they are not percentages or separate correlations.
5. Explain why the residual correlation is not the residual contribution to
   the total outcome correlation.
6. Provide a plug-in estimator, dyad-bootstrap uncertainty intervals, a
   focused `dyadMLM::decompose_apim_covariance()` implementation, and a
   reproducible table plus waterfall visualization.
7. Demonstrate that the same total interdependence can have different
   model-implied route profiles.
8. End with the compact exchangeable special case, where the two member-driven
   terms become equal and lose their role-specific interpretation.

## Planned software contribution

Existing tools already make APIMs more accessible and generic SEM path
components traceable [@stasGivingDyadicData2018;
@zhangUsingRPackageRAMpath2015]. The proposed software contribution is more
specific: a validated APIM post-estimation workflow that reconstructs the
fitted dyad sample, maps model terms to the five substantively named routes,
quantifies uncertainty, and produces a transparent result that can be checked
against the model-implied covariance.

The package contribution should be deliberately narrower than a universal
variance-partitioning function. The proposed public function is
`decompose_apim_covariance()`, not the broader `decompose_apim()`. Its first
supported contract should cover cross-sectional, two-member, linear Gaussian
APIMs with one dyadic predictor construct and fixed slopes.

The function should return:

1. the actor--actor, actor--partner, partner--actor, partner--partner, and
   residual covariance contributions;
2. the same five contributions in signed outcome-correlation points;
3. the two model-implied outcome variances, total outcome covariance, and
   total outcome correlation used to verify and scale the decomposition; and
4. the fitted sample, predictor moments, resolved coefficient names, residual
   covariance source, model backend, and assumptions needed to interpret the
   result.

The outcome variances are required for the common denominator, but the first
function should not relabel their correlated-predictor cross-products as
bounded actor, partner, overlap, or interaction shares. The decomposition is
signed and additive; it is not a percentage-of-covariance or percentage-of-
variance calculation.

Implementation should separate a small backend-neutral algebraic core from
model-specific extraction and APIM term mapping. Begin with the cross-sectional
distinguishable `glmmTMB` specification used in the package vignette. Add the
exchangeable special case by reusing the validated member-level covariance
recovery machinery. Add `brms` only after the point-estimate contract is stable,
retaining posterior draws rather than discarding their uncertainty. Require an
explicit term map whenever the actor, partner, role, predictor, or residual
covariance terms cannot be classified uniquely.

Predictor moments must be computed from the fitted paired analysis sample, not
naively from duplicated long-format design-matrix rows. The function should
report the number of complete fitted dyads used and reject unsupported or
ambiguous covariance structures rather than silently selecting one random-
effect block. A separate dyad-bootstrap workflow can add frequentist intervals
after its refitting and nonconvergence behavior have been validated.

## Novelty boundary and literature work

The five-term identity is a direct specialization of established SEM
path-tracing and covariance-component algebra
[@bokerAlgorithmHierarchicalOrganization2002]. Related work has decomposed
covariance into signed path weights and rescaled those weights to the
correlation scale in undirected Gaussian graphical models
[@jonesCovarianceDecompositionUndirected2005]. The `RAMpath` package can also
enumerate covariance components generically in SEMs
[@zhangUsingRPackageRAMpath2015]. The paper should therefore not claim a new
theorem, a new general tracing rule, a new generic correlation rescaling, or
the first software implementation of covariance path decomposition.

Prior dyadic work treats covariance as potentially arising from interpersonal
influence or shared events [@bolgerAccountingStatisticalDependency2007], and
recent random-covariance work treats the sign and magnitude of interdependence
as a dyad-varying estimand [@lawsRandomDyadicInterdependence2026]. Neither,
however, provides the proposed fixed-path decomposition of model-implied
cross-sectional APIM outcome covariance.

The potentially publishable contribution is the coherent APIM-specific
framework around the identity: defining the estimand, mapping fitted model
terms to named routes, handling the fitted dyad sample correctly, distinguishing
total from residual interdependence, quantifying uncertainty, providing
visualizations and software, and showing why equal total correlations need not
have equal model-implied explanations.

Before making a novelty claim, the literature review should establish:

- where the full distinguishable five-term expression has appeared;
- whether anyone has already estimated uncertainty for the individual terms;
- whether outcome-correlation-point contributions have been proposed;
- whether female-driven and male-driven interdependence have been named or
  compared explicitly;
- how previous work distinguishes residual covariance, explained covariance,
  and total interdependence.

The literature review should also establish how commonly APIM applications
report or decompose total outcome interdependence rather than only individual
paths and residual correlation.

## Mathematics in the manuscript

The main text needs only:

1. the two APIM equations;
2. the exogeneity and constant-residual-covariance assumptions;
3. the short covariance expansion producing the five terms;
4. the model-implied outcome-variance denominator;
5. the plug-in estimates and dyad-bootstrap recipe.

The expanded predictor–residual cross-terms, matrix generalization, and full
bootstrap implementation can go in an appendix or supplement. Delta-method
gradients, asymptotic proofs, elaborate matrix calculus, and longitudinal
derivations are not needed for the first paper.

## Proposed evidence

### Analytical and visual demonstration

- Derive the five components directly and by path tracing.
- Show how female-driven and male-driven covariance arise because one member's
  predictor reaches both outcomes.
- Contrast covariance contributions, correlation-point contributions,
  residual covariance, and residual correlation.
- Use the waterfall and same-total/different-profile figures as the primary
  intuitive demonstrations.

### Simulation study

The algebraic identity does not need simulation. Simulation is useful only to
study estimation and uncertainty. Vary sample size, predictor covariance,
unequal predictor variances, female–male coefficient differences, path signs,
and residual covariance. Evaluate:

- finite-sample bias of each plug-in component;
- coverage and width of dyad-bootstrap intervals;
- behavior when positive and negative components cancel;
- stability when total covariance is near zero;
- convergence and interval behavior in smaller samples.

Signed covariance and correlation-point contributions should be the primary
estimands. Percentage shares should not be a primary analysis.

### Empirical demonstration

Use one substantively clear predictor construct and outcome. Report:

1. raw partner outcome covariance and correlation;
2. the fitted model-implied covariance and correlation;
3. all five route contributions in covariance units;
4. all five correlation-point contributions;
5. the female-driven and male-driven comparison;
6. grouped predictor-transmission and predictor-covariance routes;
7. dyad-bootstrap uncertainty intervals;
8. one path diagram, one results table, and one signed waterfall plot.

The strongest example would compare two outcomes, predictors, or substantively
meaningful groups with similar total interdependence but different route
profiles.

## Broad manuscript structure

The target is a question-led methods paper of approximately 20--25 manuscript
pages, excluding references and supplement. The prose should introduce the
substantive problem and one running APIM before notation. Each displayed
formula should be followed immediately by a plain-language interpretation.

### Abstract: the complete argument in 200--250 words

- Begin with the missing question: APIMs estimate actor and partner paths, but
  researchers rarely ask what produces the model-implied covariance between
  the two outcomes.
- State the method: an exact five-route decomposition for a linear
  distinguishable APIM.
- State the reporting innovation narrowly: an APIM-specific table of signed
  covariance and outcome-correlation-point contributions.
- Mention the plug-in estimator, dyad-bootstrap intervals, simulation, one
  empirical demonstration, visualization, and `dyadMLM` implementation.
- End with the substantive payoff: equal total interdependence can conceal
  different model-implied route profiles.
- **Formulas:** none.
- **References:** normally none in the abstract.

### Manuscript section 1 — Introduction: the missing dyadic question (2--3 pages)

- Introduce interdependence as substantive information, not only a violation of
  independent errors [@laurenceauUsingDiaryMethods2005;
  @bolgerAccountingStatisticalDependency2007].
- Briefly review what a conventional APIM reports: actor effects, partner
  effects, and residual covariance or correlation [@kenny2006dyadic].
- Show that APIM methodology already derives interpretable combinations of
  paths, such as dyadic patterns and mediated effects
  [@kennyDetectingMeasuringTesting2010;
  @ledermannAssessingMediationDyadic2011].
- State the unresolved question: what explains the *total model-implied outcome
  covariance*, rather than only the remaining residual association?
- Give one motivating contrast: two dyads or fitted models can have the same
  outcome correlation but different route profiles.
- End with three concise contributions:
  - an APIM-specific five-route estimand and interpretation;
  - estimation, dyad-bootstrap uncertainty, and transparent software;
  - simulation and empirical evidence showing when the decomposition is useful.
- Define *explain* as descriptive, within-model attribution rather than causal
  explanation.
- **Formulas:** none. At most show the five route names in words.
- **Primary references:** Kenny et al.; Kenny and Ledermann; Ledermann et al.;
  Laurenceau and Bolger; Bolger and Shrout.

### Manuscript section 2 — One APIM, five covariance-carrying routes (about 2 pages)

- Introduce one running distinguishable APIM with the same predictor construct
  measured on both members.
- Present the complete path diagram before expanding the algebra.
- Define actor and partner coefficients by the outcome member so that the
  crossed paths cannot be confused.
- Explain visually how covariance reaches both outcomes through:
  - actor--actor;
  - actor--partner;
  - partner--actor;
  - partner--partner; and
  - residual covariance.
- Establish provenance: additive expected-covariance components follow standard
  SEM path tracing [@bokerAlgorithmHierarchicalOrganization2002].
- Contrast the proposed APIM mapping with generic covariance bridges and with
  compound dyadic effects [@zhangUsingRPackageRAMpath2015;
  @ledermannAssessingMediationDyadic2011].
- **Formula 1:** place the two APIM outcome equations together in one aligned
  display immediately after the path diagram.
- **Display:** Figure 1, a complete APIM with five highlighted route families.

### Manuscript section 3 — Exact decomposition and interpretation (about 3 pages)

- State the scope before deriving anything: independent dyads, linear Gaussian
  outcomes, fixed slopes, finite second moments, and predictors exogenous to
  the outcome residuals.
- Distinguish model-implied marginal outcome covariance from conditional
  residual covariance.
- Show the short covariance expansion and collect its five terms.
- Interpret each term twice:
  - as a product of model parameters and predictor moments; and
  - as a covariance-carrying route in substantive language.
- Introduce female-driven and male-driven contributions only for substantively
  distinguishable female--male dyads; otherwise retain neutral member labels.
- Show the grouped predictor-transmission and predictor-covariance summaries as
  optional secondary descriptions, not new estimands.
- Note the analogy to signed covariance path weights in graphical models while
  keeping the directed APIM derivation distinct
  [@jonesCovarianceDecompositionUndirected2005].
- **Formula 2:** exogeneity and residual covariance assumptions.
- **Formula 3:** the five-term covariance identity, displayed as the central
  equation of the paper.
- **Formulas 3a--3e or Table 1:** define the five named component symbols. Prefer
  a table over five separate displayed equations.
- **Display:** Table 1 with component name, formula, path, sign interpretation,
  and grouped route family.

### Manuscript section 4 — A common signed reporting scale (1.5--2 pages)

- Explain why raw covariance contributions are exact but scale-dependent.
- Derive both model-implied outcome variances using the same fitted predictor
  moments and residual variances.
- Divide every component by one common outcome-standard-deviation product.
- Emphasize that the resulting quantities:
  - sum exactly to the model-implied outcome correlation;
  - can be positive or negative;
  - can cancel one another;
  - are not percentages, bounded shares, or five separate correlations.
- Explain carefully why residual correlation differs from the residual
  contribution to total outcome correlation.
- Acknowledge that generic correlation-scale path weighting is not new; the
  contribution is its APIM-specific definition, naming, and reporting workflow
  [@jonesCovarianceDecompositionUndirected2005].
- **Formula 4:** the two model-implied outcome variances, preferably in one
  aligned display.
- **Formula 5:** component contribution
  $r_k=C_k/\sqrt{\operatorname{Var}(Y_1)\operatorname{Var}(Y_2)}$ and
  $\sum_k r_k=\rho_{Y_1Y_2}$.
- **Displays:** Figure 2, a signed waterfall; Figure 3, equal total
  interdependence with different route profiles.

### Manuscript section 5 — Estimation, uncertainty, and software (2--3 pages)

- Define plug-in estimates from a fitted APIM and explain exactly which fitted
  quantities are required.
- Reconstruct complete member pairs from the fitted analysis sample before
  estimating predictor moments.
- Require an explicit term map when role, predictor, or covariance terms are
  ambiguous.
- Describe the dyad bootstrap as a short algorithm:
  - resample complete dyads;
  - refit the APIM;
  - recompute predictor moments and all components;
  - record convergence failures under a prespecified rule;
  - form and report the chosen interval type.
- Cite bootstrap estimation of compound APIM effects as a nearby precedent,
  without implying that it estimated this decomposition
  [@ledermannAssessingMediationDyadic2011].
- Introduce `dyadMLM::decompose_apim_covariance()` as a focused post-estimation
  implementation, not a new fitting engine [@KungI2026DyadMLM]. The first
  backend uses APIMs fitted with `glmmTMB`
  [@brooksGlmmTMBBalancesSpeed2017].
- Position it relative to user-facing APIM software and generic SEM path
  tracing [@stasGivingDyadicData2018; @zhangUsingRPackageRAMpath2015].
- **Formula 6:** one compact plug-in notation statement, if needed. Do not give
  separate hatted versions of all five terms.
- **Algorithm box:** dyad-bootstrap workflow; no additional bootstrap formulas
  in the main text.

### Manuscript section 6 — Simulation study (3--4 pages)

- State explicitly that simulation evaluates estimation and interval behavior,
  not the truth of the algebraic identity.
- Organize the simulation around four questions:
  - Are the plug-in route estimates approximately unbiased?
  - Do dyad-bootstrap intervals attain useful coverage?
  - What happens when positive and negative routes cancel?
  - What happens when the total covariance is close to zero?
- Vary a focused set of design factors:
  - number of dyads;
  - predictor covariance and unequal predictor variances;
  - equal versus unequal role-specific coefficients;
  - positive, mixed-sign, and near-cancelling paths;
  - residual covariance.
- Prespecify primary outcomes: bias, empirical standard error, interval width,
  coverage, convergence rate, and component-sum reconstruction error.
- Use a small factorial core plus targeted edge cases rather than an exhaustive
  grid.
- Follow the question-to-simulation-to-recommendation style used in closely
  related APIM methods work [@gistelinckIndistinguishabilityTests2018].
- **Formulas:** none in the main simulation section. Put the data-generating
  equations and parameter grid in the supplement.
- **Display:** Figure 4 for coverage/bias in the most informative scenarios;
  Table 2 for the design and compact primary results.

### Manuscript section 7 — Empirical demonstration (about 3 pages)

- Use one substantively clear predictor and outcome measured on both partners.
- Explain why the example is useful for the decomposition, rather than merely
  convenient data availability.
- Report in a fixed sequence:
  - observed outcome covariance and correlation;
  - model-implied outcome covariance and correlation;
  - five contributions in covariance units;
  - five signed correlation-point contributions;
  - female-driven versus male-driven contributions when appropriate;
  - grouped predictor-transmission and predictor-covariance summaries;
  - dyad-bootstrap intervals and convergence information.
- Lead interpretation with the route profile, not a list of coefficients.
- If possible, compare two outcomes, predictors, or groups with similar total
  interdependence but different profiles.
- **Formulas:** none; refer back to Formulas 3 and 5.
- **Display:** Table 3 with estimates and intervals plus one empirical signed
  waterfall, either as a panel of Figure 2 or a separate figure.

### Manuscript section 8 — Extensions, boundaries, and assumptions (1.5--2 pages)

- Give the exchangeable special case as the only developed extension.
- Clarify that equality constraints within distinguishable dyads do not by
  themselves make member labels exchangeable
  [@gistelinckIndistinguishabilityTests2018].
- Briefly explain why several predictors require a visible allocation rule for
  cross-predictor covariance.
- Defer interactions, nonlinear links, intensive longitudinal data, random
  slopes, and generic variance shares to separately defined estimands.
- Use recent random-covariance work to motivate, but not claim to solve,
  heterogeneity of interdependence across dyads
  [@lawsRandomDyadicInterdependence2026].
- Reiterate that statistical route attribution is not causal mediation or
  interpersonal influence without additional assumptions and design.
- **Formula 7, optional:** the compact exchangeable identity. Move it to the
  supplement if the main text is already dense.

### Manuscript section 9 — Discussion and reporting recommendations (about 2 pages)

- Answer the opening question directly: APIM outcome interdependence can be
  represented as five signed, model-implied route contributions under stated
  assumptions.
- Summarize what the decomposition adds beyond actor effects, partner effects,
  and residual correlation.
- Give a minimum reporting recommendation:
  - model-implied total covariance and correlation;
  - all five signed contributions;
  - the common denominator;
  - uncertainty intervals;
  - fitted-dyad count and model assumptions.
- Reiterate the novelty boundary: APIM-specific estimand, interpretation,
  fitted-sample reconstruction, uncertainty, visualization, and software--not
  new general path algebra.
- End with the substantive message that similar total interdependence can arise
  through different modeled dyadic processes.
- **Formulas:** none.

### Supplement and reproducibility materials

- **Supplement A:** full scalar derivation, omitted predictor--residual
  cross-terms, and assumption checks.
- **Supplement B:** matrix formulation, several-predictor identity, and the
  exchangeable derivation if omitted from the main text.
- **Supplement C:** complete bootstrap algorithm, interval choice, failed-fit
  policy, and additional interval diagnostics.
- **Supplement D:** simulation data-generating equations, full parameter grid,
  Monte Carlo error, and secondary results.
- **Supplement E:** reproducible empirical code, model objects, component
  checks, and software session information.
- **Software tests:** verify that the five components reconstruct the
  model-implied covariance to numerical tolerance for every supported fit.

### Main-text formula budget

- **Formula 1:** paired APIM equations.
- **Formula 2:** exogeneity and residual covariance assumptions.
- **Formula 3:** central five-term covariance decomposition.
- **Formula 4:** paired model-implied outcome variances.
- **Formula 5:** correlation-point contribution and exact sum.
- **Formula 6:** optional compact plug-in notation.
- **Formula 7:** optional exchangeable special case.
- Keep the main text to approximately six displayed equation groups; Formula 7
  can move to the supplement. Use Table 1 instead of separate displays for the
  five component definitions.

## Planned displays

- **Figure 1:** complete distinguishable APIM with the five route families shown
  as coordinated panels or highlights.
- **Figure 2:** signed waterfall decomposition of one outcome correlation.
- **Figure 3:** equal total interdependence with different route profiles.
- **Figure 4:** focused simulation results for bias, coverage, cancellation, and
  the near-zero-total case.
- **Table 1:** component formulas, path interpretation, expected sign behavior,
  and grouped route family.
- **Table 2:** simulation design and compact primary performance results.
- **Table 3:** empirical estimates in covariance and correlation-point units,
  with dyad-bootstrap intervals.

## Candidate titles

**Preferred**

> Explaining Interdependence in the Actor–Partner Interdependence Model:  
> Decomposing Dyadic Outcome Covariance

Alternatives:

- *Where Does Dyadic Interdependence Come From? A Covariance Decomposition for
  the APIM*
- *Beyond the Residual Correlation: Explaining Interdependence in the APIM*
- *From Actor and Partner Paths to Dyadic Interdependence*
- *Partitioning Interdependence in the Actor–Partner Interdependence Model*

## Questions to resolve before drafting

- Confirm whether the substantive emphasis should remain on all five route
  components or foreground the two member-driven contributions.
- Decide whether the primary reporting target is covariance units, signed
  correlation-point contributions, or both. Reduction in residual covariance
  between separately fitted models should not be the primary identity.
- Complete a focused audit of applied APIM papers to establish how often total
  model-implied outcome interdependence is reported or decomposed.
- Ask relevant collaborators about any additional intellectual provenance,
  route terminology, or empirical examples without implying collaboration or
  authorship before those roles are agreed.
- Choose an example for which the five-route profile answers a real substantive
  question, ideally with a useful same-total/different-profile comparison.
- Choose the bootstrap interval type and failed-refit policy before finalizing
  the simulation design.

## Immediate next steps

1. Circulate the six-panel path figure and broad manuscript structure for
   conceptual feedback.
2. Complete the focused APIM application audit described above; the general
   path-tracing and covariance-decomposition provenance is now identified.
3. Draft manuscript Sections 1--4 around one running example and the six-formula
   main-text budget.
4. Prototype an internal `decompose_apim_covariance()` function that separates
   the algebraic core from `glmmTMB` extraction and explicit APIM term mapping.
5. Test the decomposition and extraction logic on the workshop APIM example.
6. Choose an empirical example with a clear substantive story.
7. Run a small bootstrap prototype before designing the full simulation.

## Deferred extensions and governing principles

The following extensions are worthwhile, but they should not enlarge the first
paper or the initial package function before the five-route estimand and its
uncertainty are stable:

- **Several predictors:** the matrix identity
  $\boldsymbol B\boldsymbol\Sigma_X\boldsymbol B^\mathsf{T}$ remains exact, but
  cross-predictor covariance terms require a prespecified, visible grouping or
  allocation rule. Results will depend on the chosen predictor basis and must
  not be presented as uniquely determined construct shares.
- **Interactions and nonlinear terms:** an interaction can be treated as an
  additional design-matrix column, but centering its constituent predictors
  does not generally make it orthogonal to their main effects. Do not create an
  automatic "synergistic" compartment without assumptions and a validated
  allocation rule; nonlinear links also require a separately defined response-
  or link-scale estimand.
- **Intensive longitudinal data:** within-person and between-person predictors
  suggest useful level-specific questions, but the package models may contain
  stable dyad effects, same-occasion covariance, random slopes, and serial
  residual processes. Define the target occasion weighting, fitted-pair sample,
  and marginalization level before calling any component Level 1 or Level 2.
- **Random slopes:** terms such as a fixed slope squared plus a random-slope
  variance are sufficient only under restrictive independence and centering
  conditions. A general extension must include intercept--slope and cross-member
  slope covariances and show how contributions change with predictor values.
- **Variance decompositions:** model-implied outcome variances should remain
  available as transparent checks and denominators. Any later actor/partner
  variance allocation with correlated predictors must name and justify its
  convention rather than implying that path tracing supplies unique bounded
  proportions.

These extensions can motivate later methodological work. The first paper and
software release should establish a small result that is exact under stated
assumptions, reproducible from supported fitted models, and difficult for users
to misinterpret.

# References {-}
