---
title: "APIM covariance decomposition"
subtitle: "Figures and paper structure"
output:
  html_document:
    math_method: mathml
    fig_caption: true
    self_contained: true
---

```{r load-figure-code, include=FALSE}
# Reuse the technical notes' figure code; do not render their prose or tables.
figure_code <- knitr::purl(
  "paper-idea.Rmd",
  output = tempfile(fileext = ".R"), quiet = TRUE
)
knitr::read_chunk(figure_code)
knitr::opts_chunk$set(echo = FALSE, out.width = "100%")
```

```{r prepare-figures, ref.label=c("setup", "calculate-example", "same-total-calculation"), include=FALSE}
```

## Complete APIM

```{r overview-apim, ref.label="complete-apim", fig.width=9, fig.height=4.8, fig.cap="The complete APIM. Curved paths represent predictor and residual covariances, not covariance between slopes.", fig.alt="Left-to-right APIM with two predictors, two outcomes, blue actor paths, purple partner paths, and predictor and residual covariance curves."}
```

## Five covariance-carrying routes

```{r overview-routes, ref.label="path-components", fig.width=12, fig.height=9.9, fig.cap="Five routes and their exact sum. Pale paths remain in the model but are not part of the highlighted component.", fig.alt="Six APIM panels highlighting actor-actor, actor-partner, partner-actor, partner-partner, residual, and all routes, with their component formulas."}
```

## Waterfall decomposition

```{r overview-waterfall, ref.label="example-contributions", fig.width=9.4, fig.height=5.4, fig.cap="Illustrative model: the five contributions add to the model-implied outcome correlation.", fig.alt="Waterfall chart accumulating five correlation-point contributions to the total outcome correlation."}
```

## Same total, different contributions

```{r overview-profiles, ref.label="same-total-figure", fig.width=10, fig.height=5.2, fig.cap="Two illustrative APIMs: identical total outcome correlation, different route contributions.", fig.alt="Two horizontal stacked bars with equal totals but different actor, partner, and residual contributions."}
```

## Paper structure — argument in order

Brief companion to the [current plan](plan.md); source details are in the
[literature review](literature-review.md). Paper 1 includes inference validation;
the recommended interval method has not yet been selected.

- **Introduction and prior work:** Ask what accounts for total interdependence;
  similar totals can conceal different route profiles. Build on the existing
  decomposition, signed reporting, and APIM software
  ([Dwyer et al., 2017](https://doi.org/10.1016/j.amepre.2017.01.011),
  [Kenny's handout](https://davidakenny.net/kkc/c7/Explained_Nonindependence.docx)).
- **Model and routes:** Introduce one distinguishable APIM, the complete diagram,
  and five highlighted routes; place the paired APIM equations here
  (Kenny, Kashy, & Cook, 2006).
- **Exact decomposition:** State the linear, cross-sectional, fixed-slope,
  independent-dyad and exogeneity assumptions, then show the five-term identity. Apply
  established path tracing, not new algebra ([Boker et al., 2002](https://doi.org/10.1207/S15328007SEM0902_2)).
- **Reporting scale:** Give the model-implied variances and common-denominator
  formula, then the waterfall and equal-total comparison. Contributions are
  signed correlation points, not percentages or residual correlations
  ([APIM_MM](https://davidakenny.net/doc/APIM_MM.pdf), direct reporting precedent).
- **Covariates, multiple predictors, and exchangeability:** Show the compact
  matrix identity, keep cross-block terms explicit, and group arbitrary
  member-driven terms under exchangeability. Distinguish marginal from
  conditional/adjusted covariance targets.
- **Estimation and software:** Define fitted-sample moments; compare joint
  delta-method and whole-dyad-bootstrap inference. Include the proposed
  `dyadMLM` functionality and transparent R/Mplus examples
  ([lavaan defined parameters](https://lavaan.ugent.be/tutorial/mediation.html)).
- **Simulation:** Evaluate bias, coverage, weak/zero products, correlated
  predictors, cancellation, and near-zero totals. Check joint uncertainty and
  failed fits; neither interval procedure is assumed to work everywhere.
- **Empirical example:** Report observed and model-implied totals, route
  contributions and uncertainty; explain what the route profile adds to the
  conventional APIM results.
- **Extensions and limits:** Motivate Paper 2 on Gaussian ILD levels/random
  slopes and a conditional Paper 3 on generalized observed-scale dependence.
  Keep those outside Paper 1; inference does not establish causal explanation.
- **Discussion:** Return to the opening question and recommend reporting totals,
  signed contributions, uncertainty and assumptions. State the clarification
  and validation achieved without claiming the first APIM decomposition.
- **Supplement:** Put full derivations, joint gradients, inference algorithms,
  simulation specifications and reproducible code here.
